### 2025-05-24 sandbox.qmd
swithing to new branch

---
title: "The illusory promise of the Aligned Rank Transform"
subtitle: "Appendix I. Additional experimental results"
author: 
  - name: Theophanis Tsandilas
    email: theophanis.tsandilas@inria.fr
    orcid: 0000-0002-0158-228X
    affiliations:
      - name: Université Paris-Saclay, CNRS, Inria, LISN
        country: France
  - name: Géry Casiez
    orcid: 0000-0003-1905-815X
    email: gery.casiez@univ-lille.fr
    affiliations:
      - name: Univ. Lille, CNRS, Inria, Centrale Lille, UMR 9189 CRIStAL
        country: France
bibliography: bibliography.bib

tbl-cap-location: top
---

Used for quick testing... eeee

### 2025-05-24 supplementary/answers/additional-example.Rmd
swithing to new branch

---
title: "Introduction - Illustrative example"
author: "Theophanis Tsandilas and Géry Casiez"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We detail our analysis for the illustrative example presented in the Introduction.

We use [data](example-data-2.csv) from a hypothetical experiment that evaluates the time performance of three user interface techniques (*A*, *B*, and *C*) that help users complete image editing tasks of four different difficulty levels: *Level1*, *Level2*, *Level3*, and *Level4*.

The experiment follows a fully balanced repeated-measures factorial design, where each participant ($N = 12$) performs 12 tasks in a unique order. The researcher measures the time that it takes participants to complete each task.

### Preliminary
Let us first import all required libraries.

```{r, echo = TRUE, message = FALSE}
library(lmerTest) # Lmer functions
library(ARTool) # ARTool library
library(effectsize) # To estimate effect sizes

# Additional libraries for plotting and reporting
library(kableExtra) 
library(gridExtra)
library(ggplot2) 
library(tidyverse)
```

### Reading the dataset

We read our dataset and fix the type of the factors:
```{r, warning=FALSE}
df <- read.csv("example-data-2.csv", sep=",", header=TRUE, strip.white=TRUE)

df$Participant <- factor(df$Participant)
df$Difficulty <- factor(df$Difficulty, ordered = TRUE) # We can consider it as ordinal
df$Technique <- factor(df$Technique)
```

We present our dataset with a table:
```{r, warning=FALSE}
kbl(df, booktabs = TRUE) %>% kable_styling(font_size=14) %>% kable_paper(position = "center") %>% scroll_box(height = "170px")
```

### Plotting the data

We then plot our data to observe the main trends.

```{r, warning=FALSE, fig.height=2.7, fig.width=8}
cbPalette <- c("#999999", "#E69F00", "#F15854")

p1 <- (df %>%  group_by(Participant,Technique) %>% 
summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Median Time (min)") +
        ylim(0, 8) + 
        geom_jitter(shape=20, position=position_jitter(0.1)) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette)

p2 <- (df %>%  group_by(Participant,Difficulty,Technique) %>% 
           summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Difficulty, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Median Time (min)") +
        ylim(0, 10) +
        geom_point(size = 0.7, position=position_jitterdodge(0.1)) +
        theme_bw() + theme(legend.position = "none")) +
        scale_fill_manual(values=cbPalette)

grid.arrange(p1, p2, nrow = 1, widths = c(1.5, 2.5))
```

### Statistical analysis

We conduct an analysis with five different methods: PAR, LOG, RNK, ART, and INT.
```{r, warning=FALSE}
# Implementation of the Inverse Normal Transform
int = function(x){
   qnorm((rank(x) - 0.5)/length(x))
}

mpar <- lmer(Time ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - no transformation
mlog <- lmer(log(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - Log-transformation
mart <- art(Time ~ Difficulty*Technique + (1|Participant), data=df) # ARTool
mrnk <- lmer(rank(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Rank transformation 
mint <- lmer(int(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Inverse Transform transformation

model1 <- glmer(Time ~ Difficulty*Technique + (1|Participant), family = gaussian(link = "log"), data = df)
model2 <- glmer(log(Time) ~ Difficulty*Technique + (1|Participant), family = gaussian(link = "identity"), data = df)

```

### Main and interaction effects
We finally report the results of our analysis:
```{r, warning=FALSE}
# Function for printing the results for each method
printAnalysis <- function(num, method, model) {
  cat(num, ". Analysis using ", method, "\n\n", sep="")

  print(anova(model))
  print(eta_squared(model))
  print(cohens_f(model))
}

printAnalysis(1, "Logarithmic Transformation", mlog)
printAnalysis(2, "ART", mart)
printAnalysis(3, "Rank Transformation", mrnk)
printAnalysis(4, "Inverse Normal Transformation", mint)
printAnalysis(5, "No Transformation", mpar)
```

### Contrasts
Finally, we conduct an analysis of pairwise differences between the three techniques *A*, *B*, and *C*, using a Bonferroni correction. For ART, we use the ```art.con``` function.

```{r, warning=FALSE}
library("emmeans")

# Function for printing the results for each method
printContrasts <- function(num, method, expr, model) {
  cat(num, ". Contrasts using ", method, "\n\n", sep="")

  if(method == "ART") art.con(model, expr, adjust = "bonferroni")
  else contrast(emmeans(model, expr), method="pairwise", adjust = "bonferroni")
}

printContrasts(1, "Logarithmic Transformation", ~Technique, mlog)
printContrasts(2, "ART", ~Technique, mart)
printContrasts(3, "Rank Transformation", ~Technique, mrnk)
printContrasts(4, "Inverse Normal Transformation", ~Technique, mint)
printContrasts(5, "No Transformation", ~Technique, mpar)
```

We observe again a significant divergence between the results of ART and the results of the other methods.

### 2025-05-24 supplementary/examples-case-studies/1-Introduction/illustrative-example.Rmd
swithing to new branch

--- 2025-05-24 supplementary/answers/additional-example.Rmd
@@
 
-We use [data](example-data-2.csv) from a hypothetical experiment that evaluates the time performance of three user interface techniques (*A*, *B*, and *C*) that help users complete image editing tasks of four different difficulty levels: *Level1*, *Level2*, *Level3*, and *Level4*.
+We use [data](example-data.csv) from a hypothetical experiment that evaluates the time performance of three user interface techniques (*A*, *B*, and *C*) that help users complete image editing tasks of four different difficulty levels: *Level1*, *Level2*, *Level3*, and *Level4*.
 
@@
 ```{r, warning=FALSE}
-df <- read.csv("example-data-2.csv", sep=",", header=TRUE, strip.white=TRUE)
+df <- read.csv("example-data-cut.csv", sep=",", header=TRUE, strip.white=TRUE)
 
@@
         ylab("Median Time (min)") +
-        ylim(0, 8) + 
+        ylim(0, 3) + 
         geom_jitter(shape=20, position=position_jitter(0.1)) +
@@
         ylab("Median Time (min)") +
-        ylim(0, 10) +
+        ylim(0, 6) +
         geom_point(size = 0.7, position=position_jitterdodge(0.1)) +
@@
 mint <- lmer(int(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Inverse Transform transformation
-
-model1 <- glmer(Time ~ Difficulty*Technique + (1|Participant), family = gaussian(link = "log"), data = df)
-model2 <- glmer(log(Time) ~ Difficulty*Technique + (1|Participant), family = gaussian(link = "identity"), data = df)
-
 ```

### 2025-05-24 text/20-intro.qmd
swithing to new branch

## Introduction {#intro}
In Human-Computer Interaction (HCI) and various fields within the behavioral sciences, researchers often gather data through user studies. Such data are typically messy, have small sample sizes, and may violate common statistical assumptions, such as the normality assumption. To address these challenges, researchers commonly employ nonparametric tests, which require fewer assumptions about the data. However, while nonparametric tests for simple one-factorial designs are well-established, researchers face challenges in selecting appropriate methods when dealing with multifactorial designs that require testing for both main effects and interactions. The Aligned Rank Transform or ART [@higgins:1990; @Salter:1993; @wobbrock:2011] addresses this problem by bridging the gap between nonparametric tests and ANOVAs. Its popularity in HCI research has surged, facilitated in part by the ARTool toolkit [@wobbrock:2011; @artool], which simplifies the use of the method.

Early Monte Carlo experiments in the 1990s [@Salter:1993;@Mansouri:1995] and more recent studies [@elkin:2021] suggested that ART is a robust alternative to ANOVA when normality assumptions are violated. These results have contributed to ART's reputation as a well-established method. However, other research [@luepsen:2017; @luepsen:2018] raised concerns about the robustness of the method, demonstrating that ART fails to control the Type I error rate in many scenarios, such as when data are ordinal or are drawn from skewed distributions. Unfortunately, these warnings have not received sufficient attention, and many authors still rely on Wobbrock et al.'s [-@wobbrock:2011] assertion that *"The ART is for use in circumstances similar to the parametric ANOVA, except that the response variable may be continuous or ordinal, and is not required to be normally distributed."*

Our goal is to clarify the severity of these issues and understand the potential risks of the method. We present the outcomes of a series of Monte Carlo experiments, following a distinctive simulation methodology grounded in latent variable modeling. This approach enables us to simulate effects consistently across a broad spectrum of distributions, both discrete and continuous, and allows us to address scaling issues in interpreting interactions. To ensure the clarity of our findings and facilitate their reproducibility, we divide the experimental process into multiple smaller experiments, where each experiment focuses on a distinct variable (e.g., sample size, experimental design, variance ratio) or measure (Type I error rate, power, precision of effect size estimates).

Our findings corroborate Lüpsen's alarming conclusions. We provide overwhelming evidence that ART confounds effects and raises Type I error rates at very high levels across a diverse array of non-normal distributions, including skewed, binomial, and ordinal distributions, as well as distributions with unequal variances. These issues persist for both main and interaction effects. Our results further show that simpler rank transformation methods outperform ART, while parametric ANOVA generally poses fewer risks than ART when distributions deviate from normal. Given these new insights, we conclude that ART is not a viable analysis method and advocate for its abandonment. We provide recommendations for alternative analysis methods, while we also raise warnings about the interpretation of interaction effects.

### Illustrative example 
We will begin with an illustrative example to demonstrate how the aligned rank transform can lead to an increase in false positives and a significant inflation of observed effects. This example will also serve as a brief introduction to the key concepts and methods employed throughout the paper.

Suppose an HCI researcher conducts an experiment to compare the performance of three user interface techniques (A, B, and C) that help users complete image editing tasks of four different difficulty levels. The experiment follows a fully balanced $4 \times 3$ repeated-measures factorial design, where each participant (N = 12) performs 12 tasks in a unique order. The researcher measures the time that it takes participants to complete each task. The following table presents the experimental results:

::: {#tbl-example}
```{r, echo=FALSE, warning=FALSE}
df <- read.csv("example_data.csv", sep=",", header=TRUE, strip.white=TRUE)
kbl(df) %>% kable_paper(position = "center") %>% scroll_box(width = "740px", height = "170px")
``` 
Example dataset: Time (in minutes) spent by 12 participants for four difficulty levels and three user interface techniques. Scroll down to see the full results.
:::

The experiment is hypothetical but has similarities with real-world experiments, e.g., see the experiments of @Fruchard:2023. Time performances have been randomly sampled from a population in which: (1) *Difficulty* has a large effect; (2) *Technique* has no effect; and (3) there is no interaction effect between the two factors. To generate time values, we drew samples from a log-normal distribution. The log-normal distribution is often a good fit for real-world measurements that are bounded by zero and have low means but large variance [@Limbert:2001]. Task-completion times are good examples of such measurements [@Sauro:2010].

@fig-example presents two boxplots that visually summarize the main effects observed through the experiment. We plot medians to account for the fact that distributions are skewed. We observe that differences in the overall time performance of the three techniques are not visually clear, although the overall median time is somewhat higher for Technique B. In contrast, time performance clearly deteriorates as task difficulty increases. 
We also observe that for the most difficult tasks (Level 4), the median time for Technique C is lower than the median time for Techniques A and B, so we may suspect that *Difficulty* interacts with *Technique*. However, since the spread of observed values is extremely large and the number of data points is small, such differences could result from random noise.

::: {#fig-example}
```{r, echo=FALSE, warning=FALSE, fig.height=2.7, fig.width=8}
cbPalette <- c("#999999", "#E69F00", "#F15854")

p1 <- (df %>%  group_by(Participant,Technique) %>% summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Median Time (min)") +
        ylim(0, 3) + 
        geom_jitter(shape=20, position=position_jitter(0.1)) +
        theme_bw() + theme(legend.position = "none") + scale_fill_manual(values=cbPalette))

p2 <- (df %>%  group_by(Participant,Difficulty,Technique) %>% summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Difficulty, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Median Time (min)") +
        ylim(0, 6) +
        #geom_jitter(shape=20, position=position_jitter(0.1)) +
        geom_point(size = 0.7, position=position_jitterdodge(0.1)) +
        theme_bw() + theme(legend.position = "none") +
        #scale_fill_brewer()
        scale_fill_manual(values=cbPalette))

grid.arrange(p1, p2, nrow = 1, widths = c(1.5, 2.5))
```
Boxplots summarizing the results of our illustrative example. Dots represent the median time performance of each individual participant. 
:::

We opt for a multiverse analysis [@Dragicevic:2019] to analyze the data, where we conduct a repeated-measures ANOVA with four different data-transformation methods:

1. *Log transformation (LOG).* Data are transformed with the logarithmic function. For our data, this is the most appropriate method as we drew samples from a log-normal distribution.

2. *Aligned rank transformation (ART).* Data are transformed and analyzed with the ARTool [@wobbrock:2011;@elkin:2021].

3. *Pure rank transformation (RNK).* Data are transformed with the original rank transformation [@conover:1981], which does not perform any data alignment.

4. *Inverse normal transformation (INT).* The data are transformed by using their normal scores. This rank-based method is simple to implement and has been commonly used in some disciplines. However, it has also received criticism [@Beasley:2009].

For comparison, we also report the results of the regular parametric ANOVA with no transformation (*PAR*). For each ANOVA analysis, we use a linear mixed-effects model, treating the participant identifier as a random effect. To simplify our analysis and like @elkin:2021, we consider random intercepts but no random slopes. For example, we use the following R code to create the model for the log-transformed response:

```{r, echo=TRUE, warning=FALSE}
m.log <- lmer(log(Time) ~ Difficulty*Technique + (1|Participant), data = df)
```

The table below presents the *p*-values for the main effects of the two factors and their interaction:

|        | PAR  | LOG  | ART | RNK | INT |
|--------|------|------|-----|-----|-----|
| Difficulty  | $1.8 \times 10^{-26}$ | $8.1 \times 10^{-47}$  |  $9.0 \times 10^{-43}$ | $4.3 \times 10^{-46}$ | $4.4 \times 10^{-44}$ |
| Technique   | $.10$ | $.18$  | $.00061$ | $.38$ | $.17$ |
| Difficulty $\times$ Technique | $.056$ | $.10$ | $.0017$ | $.24$ | $.23$ |
: *p*-values for main and interaction effects {.sm}

The disparity in findings between ART and the three alternative transformation methods is striking. ART suggests that all three effects are statistically significant. What adds to the intrigue is the fact that ART's *p*-values for *Technique* and its interaction with *Difficulty* are orders of magnitude lower than the *p*-values obtained from all other methods. We will observe similar discrepancies if we conduct contrast tests with the ART procedure [@elkin:2021], though we leave this as an exercise for the reader.

We also examine effect size measures, which are commonly reported in scientific papers. The table below presents results for partial $\eta^2$, which describes the ratio of variance explained by a variable or an interaction:

|        | PAR  | LOG  | ART | RNK | INT |
|--------|------|------|-----|-----|-----|
| Difficulty  | $.64\ [.55, 1.0]$  | $.83\ [.79, 1.0]$  |  $.80\ [.76, 1.0]$ | $.83\ [.79, 1.0]$ | $.81\ [.77, 1.0]$  |
| Technique   | $.04\ [.00, 1.0]$ | $.03\ [.00, 1.0]$  | $.11\ [.03, 1.0]$ | $.02\ [.00, 1.0]$ | $.03\ [.00, 1.0]$ |
| Difficulty $\times$ Technique | $.10\ [.00, 1.0]$ | $.08\ [.00, 1.0]$  | $.16\ [.04, 1.0]$| $.06\ [.00, 1.0]$| $.06\ [.00, 1.0]$ |
: partial $\eta^2$ and its 95\% confidence interval {.sm}

We observe that ART exaggerates both the effect of *Technique* and its interaction with *Difficulty*.

### Overview
The above example does not capture a rare phenomenon. We will show that ART's error inflation is systematic for a range of distributions that deviate from normality, both continuous and discrete. We will explain how the problem emerges. We will also see that ART often performs worse than simpler methods that ART is widely considered to repair or improve, such as the pure rank transformation or no transformation at all.

The paper is structured as follows. [Section 2](#background) introduces nonparametric tests and rank transformations, and explains how ART is constructed. It also provides a summary of previous experimental results regarding the robustness of ART and other related rank-based procedures, along with a survey of recent studies using ART. [Section 3](#interpretation) investigates issues regarding effect interpretation and introduce our distribution simulation approach. [Section 4](#methodology) outlines our experimental methodology, while [Section 5](#findings) presents our findings. [Section 6](#casestudies) revisits results of previous studies employing ART and  illustrates how its application can lead to erroneous conclusions. [Section 7](#recommendations) offers recommendations for researchers. Finally, [Section 8](#conclusions) concludes the paper.

In addition to the main sections of the paper, we provide an [appendix](appendix.html) with results from additional Monte Carlo experiments.

### 2025-05-24 supplementary/answers/Main-1-within-4x3-Testing.R

@@ -1,0 +1,95 @@
+# Author: Theophanis Tsandilas, 2023-2024
+# Inria & Université Paris-Saclay
+
+# Experiment evaluating the Type I error rates of PAR, ART, RNK, and INT on ratio scales (distributions below) for a 4x3 within-subjects design
+# See Fig. 9, Fig. 11, Fig. 12
+
+# By default, this code will generate a data file for n = 20 with 5000 iteration. 
+# You can edit the experimental parameters in the code to test different conditions.
+
+rm(list=ls())
+
+source("data-generators.R") # Import the data-generation functions
+
+plotCellLatent <- function(df, x1s, x2 = "B1", min = -8, max = 8) {
+	colors = c("#666666", "#E69F00", "#009E73", "#FF5E00")
+
+	createText <- function(values){
+		mean <- mean(values)
+		sd <- sd(values)
+		paste("m=", round(mean,digits=2), ", sd=", round(sd, digits = 2), sep = "")
+	}
+
+	values <- df[df$x1 == x1s[1] & df$x2 == x2, ]$dv
+	plot(density(values), col = colors[1], lwd = 2, ylim = c(0, .4), xlim = c(min, max), main = paste("x2 =", x2), xlab = NA, bty="l")
+	text(median(values) - 1, max(density(values)$y - .05), createText(values), pos = 2, col = colors[1])
+
+	values <- df[df$x1 == x1s[2] & df$x2 == x2, ]$dv
+	lines(density(values), col = colors[2], lwd = 2, ylim = c(0, .4), xlim = c(min, max))
+	text(median(values)+.3, max(density(values)$y + .03), createText(values), pos = 2, col = colors[2])
+
+	values <- df[df$x1 == x1s[3] & df$x2 == x2, ]$dv
+	lines(density(values), col = colors[3], lwd = 2, ylim = c(0, .4), xlim = c(min, max))
+	text(median(values)-.3, max(density(values)$y + .03), createText(values), pos = 4, col = colors[3])
+
+	values <- df[df$x1 == x1s[4] & df$x2 == x2, ]$dv
+	lines(density(values), col = colors[4], lwd = 2, ylim = c(0, .4), xlim = c(min, max))
+	text(median(values)+1, max(density(values)$y -.05), createText(values), pos = 4, col = colors[4])
+	
+}
+
[…]

### 2025-05-24 supplementary/answers/data-generators.R

@@ -1,0 +1,345 @@
+# Author: Theophanis Tsandilas, 2023-2024
+# Inria & Université Paris-Saclay
+
+# Several functions used by our eperimental code, mainly for data generation and transformation
+
+# A great library for simulating data for a large range of desings: https://debruine.github.io/faux
+library(faux)
+
+library(dplyr)
+library(tidyr)
+
+
+# https://yingji15.github.io/2019-08-17-inverse-normal-transformation/
+# https://www.biostars.org/p/80597/
+# use rankit version where c = 0.5, based on https://www.researchgate.net/publication/286889581_Impact_of_Rank-Based_Normalizing_Transformations_on_the_Accuracy_of_Test_Scores
+inverseNormalTransform = function(x){
+	qnorm((rank(x,na.last="keep") - 0.5)/sum(!is.na(x)))
+}
+
+# Transforms a set of values to ordinal based on the given thresholds
+# See: https://doi.org/10.1016/j.jesp.2018.08.009
+toOrdinal = function(values, thresholds = c(-2,-1, 1,2)) {
+	discretizeValue = function(x){
+		for(i in 1:length(thresholds)) {
+			if(x <= thresholds[i]) return(i)
+		}
+		return(length(thresholds) + 1)
+	}
+
+	sapply(values, discretizeValue)
+}
+
+
+# Transforms a set of values to ordinal based on the given thresholds
+# See: https://doi.org/10.1016/j.jesp.2018.08.009
+# Here, the thresholds take into account the variance of the data, trying to better cover the full range of values.
+toOrdinalDistr = function(values, nlevels = 5, equidistant = TRUE) {
+	lim <- 2*sd(values) # I take 2 standard deviations that should contain the largest portion of the data (around 95% of the data values)
+
+	thresholds <- NA
+	if(equidistant) {
+		thresholds <- seq(from=-lim, to=lim, by=2*lim/nlevels)[c(-1, -(nlevels + 1))]
+	} else {
+		thresholds <- sort(runif(nlevels - 1))*2*lim-lim
+	}
+
+	toOrdinal(values, thresholds)
+}
+
+
+# Additional support for other distributions with not conversion support at faux
[…]

### 2025-05-24 supplementary/answers/equal-var-model.R

@@ -1,0 +1,19 @@
+rm(list=ls())
+
+
+sdlog <- function(mean, sd) {
+	sqrt(log(1 + sd^2/mean^2))	
+}
+
+meanlog <- function(mean, sd) {
+	log(mean^2/sqrt(sd^2 + mean^2))
+}
+
+cbPalette = c("#888888", "#E69F00", "#009E73", "#FF5E00")
+
+sd <- 0.8
+
+curve(dlnorm(x-0.2, meanlog(1.2, sd), sdlog(1.2, sd)), xlim=c(0,4.5),col=cbPalette[1], lwd = 2, frame.plot = FALSE, yaxt='n', ylab="", xlab="Time (sec)")
+curve(dlnorm(x-0.2, meanlog(1.5, sd), sdlog(1.5, sd)), xlim=c(0,4.5),col=cbPalette[2], lwd = 2, yaxt='n', ylab="", xlab="Time (sec)", add = TRUE)
+curve(dlnorm(x-0.2, meanlog(1.8, sd), sdlog(1.8, sd)), xlim=c(0,4.5),col=cbPalette[3], lwd = 2, yaxt='n', ylab="", xlab="Time (sec)", add = TRUE)
+curve(dlnorm(x-0.2, meanlog(2.1, sd), sdlog(2.1, sd)), xlim=c(0,4.5),col=cbPalette[4], lwd = 2, yaxt='n', ylab="", xlab="Time (sec)", add = TRUE)

### 2025-05-24 supplementary/answers/example-interactions.R

@@ -1,0 +1,45 @@
+rm(list=ls())
+library(tidyverse)
+library(ggplot2)
+library(gridExtra)
+
+source("data-generators.R") # Import the data-generation functions
+
+design <- c(4,3)
+
+effects <- c(2, 0, 0)
+distr <- "lnorm"
+n <- 12
+
+plotInteraction <- function(){
+	df <- sim.within.2f(n, design, effects) %>% toDistribution(distr)
+	df <- df[,c(1,2,3,9)] %>% 
+		rename(Participant=s, Difficulty=x1, Technique=x2,Time=y) %>%
+		mutate(Difficulty=recode(Difficulty,"A1" = "Level1", "A2"="Level2","A3"="Level3","A4"="Level4")) %>%
+		mutate(Technique=recode(Technique,"B1" = "Technique A", "B2" = "Technique B", "B3" = "Technique C"))
+
+	dfmeans <- aggregate(df, Time ~ Difficulty + Technique, FUN = mean)
+
+
+	cbPalette <- c("#999999", "#E69F00", "#F15854")
+
+	p <- dfmeans %>% ggplot(aes(x = Difficulty, y = Time, color = Technique, group=Technique)) + 
+		geom_line(linewidth=1.2) +
+		geom_point(size=2) +
+		ylab("Mean Time (min)") +
+		ylim(0, 4) + 
+		theme_bw() + theme(legend.position = "none") + 
+		scale_color_manual(values=cbPalette)
+}
+
+set.seed(3000)
+p1 <- plotInteraction()
+p2 <- plotInteraction()
+p3 <- plotInteraction()
+p4 <- plotInteraction()
+p5 <- plotInteraction()
+p6 <- plotInteraction()
+
+grid.arrange(p1, p2, p3, p4, p5, p6, ncol=3, widths=c(2.3, 2.3, 2.3))
+
+

### 2025-05-24 supplementary/answers/higgins-model.R

@@ -1,0 +1,10 @@
+rm(list=ls())
+
+cbPalette = c("#888888", "#E69F00", "#009E73", "#FF5E00")
+
+w <- 2
+
+curve(dlnorm(x-0.2, sd=0.5), xlim=c(0,4.5),col=cbPalette[1], lwd = w, frame.plot = FALSE, yaxt='n', ylab="", xlab="Time (sec)")
+curve(dlnorm(x-0.5, sd=0.5), xlim=c(0,4.5),col=cbPalette[2], lwd = w, yaxt='n', ylab="", xlab="Time (sec)", add = TRUE)
+curve(dlnorm(x-0.8, sd=0.5), xlim=c(0,4.5),col=cbPalette[3], lwd = w, yaxt='n', ylab="", xlab="Time (sec)", add = TRUE)
+curve(dlnorm(x-1.1, sd=0.5), xlim=c(0,4.5),col=cbPalette[4], lwd = w, yaxt='n', ylab="", xlab="Time (sec)", add = TRUE)

### 2025-05-24 supplementary/answers/p-values.R

@@ -1,0 +1,124 @@
+# Author: Theophanis Tsandilas, 2023-2024
+# Inria & Université Paris-Saclay
+
+# Experiment evaluating the effect size estimates of PAR, ART, RNK, and INT.
+# You can edit the experimental parameters in the code to test different conditions.
+
+
+rm(list=ls())
+library(lmerTest)
+library(effectsize) # Effect size functions
+
+# The ARTool
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+source("data-generators.R") # Import the data-generation functions
+
+eta_values = function(model) {
+	return(round(eta_squared(anova(model))[, 2], digits=4))
+}
+
+getLmerPValues <- function(model, vars) {
+	return(anova(model)[vars, 6])
+}
+
+getARTPValues <- function(model, vars) {
+	return(anova(model)[vars, 5]) # For within-subjects
+}
+
+# Repeated measures with two factors
+test.within.2f <- function(n, design, effects, distr){ 
+	df <- sim.within.2f(n, design, effects) %>% toDistribution(distr)
+
+	model = suppressMessages(lmer(dv ~ x1*x2 + (1|s), data=df)) # This is the reference groundtruth model
+
+	model0 <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
+	model1 <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
+	model2 <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
+	model3 <- suppressMessages(lmer(inverseNormalTransform(y) ~ x1*x2 + (1|s), data=df)) # INT
+
+	vars <- c("x1", "x2", "x1:x2")
+	c(effects, 
+		getLmerPValues(model, vars), getLmerPValues(model0, vars), getLmerPValues(model1, vars), getARTPValues(model2, vars), getLmerPValues(model3, vars),
+		eta_values(model), eta_values(model0), eta_values(model1), eta_values(model2), eta_values(model3))
+}
+
+# Performs repetitive tests for the given configuration
+test <- function(repetitions, n, design, distr, effects) {
+	results <- foreach(rid = 1:repetitions, .combine=rbind) %dopar% {
[…]

### 2025-05-24 supplementary/answers/sim-data.R

@@ -1,0 +1,114 @@
+# Author: Theophanis Tsandilas, 2023-2024
+# Inria & Université Paris-Saclay
+
+# Experiment evaluating the Type I error rates of PAR, ART, RNK, and INT on ratio scales (distributions below) for a 4x3 within-subjects design
+# See Fig. 9, Fig. 11, Fig. 12
+
+# By default, this code will generate a data file for n = 20 with 5000 iteration. 
+# You can edit the experimental parameters in the code to test different conditions.
+
+rm(list=ls())
+library(lmerTest)
+
+# The ARTool
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+source("data-generators.R") # Import the data-generation functions
+
+getLmerPValues <- function(model, vars) {
+	return(anova(model)[vars, 6])
+}
+
+
+getARTPValues <- function(model, vars) {
+	return(anova(model)[vars, 5]) # For within-subjects
+}
+
+# Repeated measures with two factors
+test.within.2f <- function(n, design, effects, distr){ 
+	df <- sim.within.2f(n, design, effects) %>% toDistribution(distr)
+
+	model0 <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
+	model1 <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
+	model2 <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
+	model3 <- suppressMessages(lmer(inverseNormalTransform(y) ~ x1*x2 + (1|s), data=df)) # INT
+	vars <- c("x1", "x2", "x1:x2")
+	c(getLmerPValues(model0, vars), getLmerPValues(model1, vars), getARTPValues(model2, vars), getLmerPValues(model3, vars))
+}
+
+# Performs repetitive tests for the given configuration
+test <- function(repetitions, n, design, effects, distr) {
+	results <- foreach(rid = 1:repetitions, .combine=rbind) %dopar% {
+		tryCatch(
+			{
+				test.within.2f(n, design, effects, distr)
+			}, 
+			error = function(cond) {
+				# do nothing
+			}, finally = {
+				# do nothing
+			}
+		)
+	}
+
[…]

### 2025-05-24 supplementary/examples-case-studies/1-Introduction/example-cut.R

@@ -1,0 +1,33 @@
+rm(list=ls())
+library(lmerTest) # Lmer functions
+library(ARTool) # ARTool library
+library(effectsize) # To estimate effect sizes 
+
+library(tidyverse)
+library("emmeans")
+
+
+df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)
+#df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)
+
+df$Participant <- factor(df$Participant)
+df$Difficulty <- factor(df$Difficulty, ordered = TRUE) # We can consider it as ordinal
+df$Technique <- factor(df$Technique)
+
+
+# Implementation of the Inverse Normal Transform
+int = function(x){
+   qnorm((rank(x) - 0.5)/length(x))
+}
+ 
+#df[df$Technique == "A" & df$Difficulty == "Level4", ]$Time = df[df$Technique == "A" & df$Difficulty == "Level4", ]$Time  + .13
+df$Time <- ifelse(df$Difficulty == "Level4" & df$Technique == "A", df$Time + .13, df$Time)
+
+#df <- df[df$Difficulty != "Level4" | (df$Time < 9 & df$Time > 1.07), ]
+
+mpar <- lmer(Time ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - no transformation
+mlog <- lmer(log(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - Log-transformation
+mart <- art(Time ~ Difficulty*Technique + (1|Participant), data=df) # ARTool
+mrnk <- lmer(rank(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Rank transformation 
+mint <- lmer(int(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Inverse Transform transformation
+

### 2025-05-24 supplementary/examples-case-studies/1-Introduction/example.R

@@ -1,0 +1,31 @@
+rm(list=ls())
+library(lmerTest) # Lmer functions
+library(ARTool) # ARTool library
+library(effectsize) # To estimate effect sizes 
+
+library(tidyverse)
+library("emmeans")
+
+
+df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)
+
+df$Participant <- factor(df$Participant)
+df$Difficulty <- factor(df$Difficulty, ordered = TRUE) # We can consider it as ordinal
+df$Technique <- factor(df$Technique)
+
+
+# Implementation of the Inverse Normal Transform
+int = function(x){
+   qnorm((rank(x) - 0.5)/length(x))
+}
+
+mpar <- lmer(Time ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - no transformation
+mlog <- lmer(log(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - Log-transformation
+mart <- art(Time ~ Difficulty*Technique + (1|Participant), data=df) # ARTool
+mrnk <- lmer(rank(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Rank transformation 
+mint <- lmer(int(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Inverse Transform transformation
+
+
+
+
+

### 2025-05-24 supplementary/experimental-code/Main-1-within-4x3-ratio.R

@@ -105,7 +105,7 @@ effects <- matrix(c(0, 0, 0,
 Ns <- c(20) 
 
 # 5000 iterations
-R <- 5000
+R <- 1000
 
 filename = "1_test_4x3_Ratio"
 

### 2025-05-24 supplementary/experimental-code/Test-p-values-scatterplots.R

@@ -1,0 +1,135 @@
+# Author: Theophanis Tsandilas, 2023-2024
+# Inria & Université Paris-Saclay
+
+# Experiment evaluating the effect size estimates of PAR, ART, RNK, and INT.
+# See Fig. 28, Fig. 29
+
+# You can edit the experimental parameters in the code to test different conditions.
+
+
+rm(list=ls())
+library(lmerTest)
+library(effectsize) # Effect size functions
+
+# The ARTool
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+source("data-generators.R") # Import the data-generation functions
+
+
+eta_values = function(model) {
+	return(eta_squared(anova(model))[, 2])
+}
+
+getLmerPValues <- function(model, vars, lmer = TRUE) {
+	return(suppressMessages(anova(model)[vars, ifelse(lmer, 6, 5)])) # For aov models the table is different
+}
+
+getARTPValues <- function(model, vars, lmer = TRUE) {
+	return(suppressMessages(anova(model)[vars, ifelse(lmer, 5, 7)])) # For aov models the ART's table is different
+}
+
+
+# Repeated measures with two factors
+test.within.2f <- function(n, design, effects, distr){ 
+	df <- sim.within.2f(n, design, effects) %>% toDistribution(distr)
+
+	model = suppressMessages(lmer(dv ~ x1*x2 + (1|s), data=df)) # This is the reference groundtruth model
+
+	model0 <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
+	model1 <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
+	model2 <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
+	model3 <- suppressMessages(lmer(inverseNormalTransform(y) ~ x1*x2 + (1|s), data=df)) # INT
+
+#	c(effects, eta_values(model), eta_values(model0), eta_values(model1), eta_values(model2), eta_values(model3))
+
+	vars <- c("x1", "x2", "x1:x2")
+	c(effects, getLmerPValues(model, vars), getLmerPValues(model0, vars), getLmerPValues(model1, vars), getARTPValues(model2, vars), getLmerPValues(model3, vars))
+}
+
+# Performs repetitive tests for the given configuration
[…]

### 2025-05-24 supplementary/kay/1_binary/test-independent-1000.R

@@ -1,0 +1,83 @@
+# author: Theophanis Tsandilas
+# This code reads a template of a 4x3 repeated-measures design and makes the response variable take random binary values 
+# by assuming a between-subjects design 
+# It then evaluates the Type I error rate of ART
+
+rm(list=ls())
+
+library(dplyr)
+library(tidyr)
+
+library(lmerTest)
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+
+# This method takes the template and randomly assign 0s or 1s to the response variable y 
+# The parameter prob determines the occurence probability of 1s (by default, we test equal probabilities)
+createRandomSample <- function(df, prob = 0.5){
+	df$y <- rbinom(nrow(df), size = 1, prob)
+	
+	df
+}
+
+# Analysis with two methods: PAR and ART (note that RNK and INT are identical to PAR for binary responses)
+analyze <- function(df) {
+	m.par <- suppressMessages(aov(y ~ x1*x2, data=df)) # Parametric
+	m.art <- suppressMessages(art(y ~ x1*x2, data=df)) # ARTool
+
+	vars <- c("x1", "x2", "x1:x2")
+	c(# Return the p-values for the three effects 
+		suppressMessages(anova(m.par)[vars, 5]), 
+		suppressMessages(anova(m.art)[vars, 7])
+	) 
+}
+
+# Performs repetitive tests and assess Type I error rates 
+test <- function(template, repetitions = 3000) {
+	results <- foreach(rid = 1:repetitions, .combine=rbind) %dopar% {
+		tryCatch(
+			{
+				analyze(createRandomSample(template)) # Analyze a random sample
+			}, 
+			error = function(cond) { }, 
+			finally = { }
+		)
+	}
+
+	# From p-values to Type I error rates (false positives)
+	res.05 <- round(colMeans(results<.05, na.rm = TRUE), digits = 4)
+
+	# Split the results into separate rows 
+	return(tribble(~method, ~rates,
+			"PAR", res.05[1:3], 
+			"ART", res.05[4:6]
+		)
+	)
+}
+
+
+#Parallel: https://nceas.github.io/oss-lessons/parallel-computing-in-r/parallel-computing-in-r.html
+CoresNum <- 4
+registerDoParallel(CoresNum)  # use multicore, set to the number of our cores
+
+# Read the data template and fix the type of the factors, as required by the ARTool
[…]

### 2025-05-24 supplementary/kay/1_binary/test-independent-20.R

@@ -1,0 +1,83 @@
+# author: Theophanis Tsandilas
+# This code reads a template of a 4x3 repeated-measures design and makes the response variable take random binary values
+# by assuming a between-subjects design 
+# It then evaluates the Type I error rate of ART
+
+rm(list=ls())
+
+library(dplyr)
+library(tidyr)
+
+library(lmerTest)
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+
+# This method takes the template and randomly assign 0s or 1s to the response variable y 
+# The parameter prob determines the occurence probability of 1s (by default, we test equal probabilities)
+createRandomSample <- function(df, prob = 0.5){
+	df$y <- rbinom(nrow(df), size = 1, prob)
+	
+	df
+}
+
+# Analysis with two methods: PAR and ART (note that RNK and INT are identical to PAR for binary responses)
+analyze <- function(df) {
+	m.par <- suppressMessages(aov(y ~ x1*x2, data=df)) # Parametric
+	m.art <- suppressMessages(art(y ~ x1*x2, data=df)) # ARTool
+
+	vars <- c("x1", "x2", "x1:x2")
+	c(# Return the p-values for the three effects 
+		suppressMessages(anova(m.par)[vars, 5]), 
+		suppressMessages(anova(m.art)[vars, 7])
+	) 
+}
+
+# Performs repetitive tests and assess Type I error rates 
+test <- function(template, repetitions = 3000) {
+	results <- foreach(rid = 1:repetitions, .combine=rbind) %dopar% {
+		tryCatch(
+			{
+				analyze(createRandomSample(template)) # Analyze a random sample
+			}, 
+			error = function(cond) { }, 
+			finally = { }
+		)
+	}
+
+	# From p-values to Type I error rates (false positives)
+	res.05 <- round(colMeans(results<.05, na.rm = TRUE), digits = 4)
+
+	# Split the results into separate rows 
+	return(tribble(~method, ~rates,
+			"PAR", res.05[1:3], 
+			"ART", res.05[4:6]
+		)
+	)
+}
+
+
+#Parallel: https://nceas.github.io/oss-lessons/parallel-computing-in-r/parallel-computing-in-r.html
+CoresNum <- 4
+registerDoParallel(CoresNum)  # use multicore, set to the number of our cores
+
+# Read the data template and fix the type of the factors, as required by the ARTool
[…]

### 2025-05-24 supplementary/kay/2_binomial/test-1000.R

@@ -1,0 +1,102 @@
+# author: Theophanis Tsandilas
+# This code reads a template of a 4x3 repeated-measures design and makes the response variable take random values from a binomial distribution 
+# It then evaluates the Type I error rate of ART
+
+rm(list=ls())
+
+library(dplyr)
+library(tidyr)
+
+library(lmerTest)
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+
+# This method takes the template and randomly assigns values drawn from a binomial distribution to the response variable y 
+# The parameter prob determines the probability of error (or succes), and size the number of Bernoulli trials
+# shape1 is a parameter for the beta distribution for individual participant differences
+createRandomSample <- function(df, size = 10, prob = 0.05, shape1 = 2){
+	# I add subject-level effects. For each subject, I randomly pick a different probabily drawn from a beta distribution with shape1 = 2
+	# See: https://en.wikipedia.org/wiki/Beta_distribution
+	# I derive the second parameter of the beta distribution from the desired mean probability
+	shape2 = shape1/prob - shape1
+	df <- df %>% group_by(s) %>% mutate(probs = rbeta(1, shape1 = 1, shape2))
+
+	# from that, generate responses from a binomial distribution
+	df$y <- rbinom(nrow(df), size, df$probs)
+	df
+}
+
+# INT implementation
+INT <- function(x){
+    qnorm((rank(x) - 0.5)/length(x))
+}
+
+
+# Analysis with two methods: PAR and ART
+analyze <- function(df) {
+	m.par <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
+	m.art <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
+	m.rnk <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
+	m.int <- suppressMessages(lmer(INT(y) ~ x1*x2 + (1|s), data=df)) # INT
+
+	vars <- c("x1", "x2", "x1:x2")
+	c(# Return the p-values for the three effects 
+		suppressMessages(anova(m.par)[vars, 6]), 
+		suppressMessages(anova(m.art)[vars, 5]),
+		suppressMessages(anova(m.rnk)[vars, 6]), 
+		suppressMessages(anova(m.int)[vars, 6])
+	) 
+}
+
+
[…]

### 2025-05-24 supplementary/kay/2_binomial/test-20.R

@@ -1,0 +1,102 @@
+# author: Theophanis Tsandilas
+# This code reads a template of a 4x3 repeated-measures design and makes the response variable take random values from a binomial distribution 
+# It then evaluates the Type I error rate of ART
+
+rm(list=ls())
+
+library(dplyr)
+library(tidyr)
+
+library(lmerTest)
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+
+# This method takes the template and randomly assigns values drawn from a binomial distribution to the response variable y 
+# The parameter prob determines the probability of error (or succes), and size the number of Bernoulli trials
+# shape1 is a parameter for the beta distribution for individual participant differences
+createRandomSample <- function(df, size = 10, prob = 0.05, shape1 = 2){
+	# I add subject-level effects. For each subject, I randomly pick a different probabily drawn from a beta distribution with shape1 = 2
+	# See: https://en.wikipedia.org/wiki/Beta_distribution
+	# I derive the second parameter of the beta distribution from the desired mean probability
+	shape2 = shape1/prob - shape1
+	df <- df %>% group_by(s) %>% mutate(probs = rbeta(1, shape1 = 1, shape2))
+
+	# from that, generate responses from a binomial distribution
+	df$y <- rbinom(nrow(df), size, df$probs)
+	df
+}
+
+# INT implementation
+INT <- function(x){
+    qnorm((rank(x) - 0.5)/length(x))
+}
+
+
+# Analysis with two methods: PAR and ART
+analyze <- function(df) {
+	m.par <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
+	m.art <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
+	m.rnk <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
+	m.int <- suppressMessages(lmer(INT(y) ~ x1*x2 + (1|s), data=df)) # INT
+
+	vars <- c("x1", "x2", "x1:x2")
+	c(# Return the p-values for the three effects 
+		suppressMessages(anova(m.par)[vars, 6]), 
+		suppressMessages(anova(m.art)[vars, 5]),
+		suppressMessages(anova(m.rnk)[vars, 6]), 
+		suppressMessages(anova(m.int)[vars, 6])
+	) 
+}
+
+
[…]

### 2025-05-24 supplementary/kay/3_likert/test-20.R

@@ -1,0 +1,132 @@
+# author: Theophanis Tsandilas
+# This code reads a template of a 4x3 repeated-measures design and makes the response variable take values from an ordinal scale
+# It then evaluates the Type I error rate of ART
+
+rm(list=ls())
+
+library(dplyr)
+library(tidyr)
+
+library(lmerTest)
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+
+# Transforms a set of values to ordinal based on the given thresholds
+# See: https://doi.org/10.1016/j.jesp.2018.08.009
+toOrdinal = function(values, thresholds = c(-2,-1, 1,2)) {
+	discretizeValue = function(x){
+		for(i in 1:length(thresholds)) {
+			if(x <= thresholds[i]) return(i)
+		}
+		return(length(thresholds) + 1)
+	}
+
+	sapply(values, discretizeValue)
+}
+
+# Transforms a set of values to ordinal based on the given thresholds
+# See: https://doi.org/10.1016/j.jesp.2018.08.009
+# Here, the thresholds take into account the variance of the data, trying to better cover the full range of values.
+toOrdinalDistr = function(values, nlevels, equidistant = FALSE) {
+	lim <- 2*sd(values) # I take 2 standard deviations that should contain the largest portion of the data (around 95% of the data values)
+
+	thresholds <- NA
+	if(equidistant) {
+		thresholds <- seq(from=-lim, to=lim, by=2*lim/nlevels)[c(-1, -(nlevels + 1))]
+	} else {
+		thresholds <- sort(runif(nlevels - 1))*2*lim-lim
+	}
+
+	toOrdinal(values, thresholds)
+}
+
+
+# This method takes the template and randomly assigns ordinal values to the response variable y 
+# We set the number of levels in the ordinal scale
+createRandomSample <- function(df, nlevels = 5){
+	# We first create some random subject-level intercepts (drawn from normal ditribution)
+	# You can vary the sd expressing individual variability
+	df <- df %>% group_by(s) %>% mutate(intercept = rnorm(1, mean = 0, sd = 0.3))
+
+	# We then generate responses from a normal again distribution
+	df$y <- rnorm(nrow(df), mean = df$intercept, sd = 1) 
+
[…]

### 2025-05-24 supplementary/kay/3_likert/test-independent-1000.R

@@ -1,0 +1,129 @@
+# author: Theophanis Tsandilas
+# This code reads a template of a 4x3 repeated-measures design and makes the response variable take values from an ordinal scale
+# by assuming a between-subjects design
+# It then evaluates the Type I error rate of ART
+
+rm(list=ls())
+
+library(dplyr)
+library(tidyr)
+
+library(lmerTest)
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+
+# Transforms a set of values to ordinal based on the given thresholds
+# See: https://doi.org/10.1016/j.jesp.2018.08.009
+toOrdinal = function(values, thresholds = c(-2,-1, 1,2)) {
+	discretizeValue = function(x){
+		for(i in 1:length(thresholds)) {
+			if(x <= thresholds[i]) return(i)
+		}
+		return(length(thresholds) + 1)
+	}
+
+	sapply(values, discretizeValue)
+}
+
+# Transforms a set of values to ordinal based on the given thresholds
+# See: https://doi.org/10.1016/j.jesp.2018.08.009
+# Here, the thresholds take into account the variance of the data, trying to better cover the full range of values.
+toOrdinalDistr = function(values, nlevels, equidistant = FALSE) {
+	lim <- 2*sd(values) # I take 2 standard deviations that should contain the largest portion of the data (around 95% of the data values)
+
+	thresholds <- NA
+	if(equidistant) {
+		thresholds <- seq(from=-lim, to=lim, by=2*lim/nlevels)[c(-1, -(nlevels + 1))]
+	} else {
+		thresholds <- sort(runif(nlevels - 1))*2*lim-lim
+	}
+
+	toOrdinal(values, thresholds)
+}
+
+
+# This method takes the template and randomly assigns ordinal values to the response variable y 
+# We set the number of levels in the ordinal scale
+createRandomSample <- function(df, nlevels = 5){
+	# We first draw values from a standard normal distribution
+	df$y <- rnorm(nrow(df), mean = 0, sd = 1) 
+
+	# We then descretize these values usign either equidistant or flexible thresholds	
+	df$y <- toOrdinalDistr(df$y, nlevels, equidistant = FALSE)
+
+	df
+}
+
+# INT implementation
+INT <- function(x){
+    qnorm((rank(x) - 0.5)/length(x))
+}
+
+
[…]

### 2025-05-24 supplementary/kay/3_likert/test-independent-20.R

@@ -1,0 +1,129 @@
+# author: Theophanis Tsandilas
+# This code reads a template of a 4x3 repeated-measures design and makes the response variable take values from an ordinal scale
+# by assuming a between-subjects design
+# It then evaluates the Type I error rate of ART
+
+rm(list=ls())
+
+library(dplyr)
+library(tidyr)
+
+library(lmerTest)
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+
+# Transforms a set of values to ordinal based on the given thresholds
+# See: https://doi.org/10.1016/j.jesp.2018.08.009
+toOrdinal = function(values, thresholds = c(-2,-1, 1,2)) {
+	discretizeValue = function(x){
+		for(i in 1:length(thresholds)) {
+			if(x <= thresholds[i]) return(i)
+		}
+		return(length(thresholds) + 1)
+	}
+
+	sapply(values, discretizeValue)
+}
+
+# Transforms a set of values to ordinal based on the given thresholds
+# See: https://doi.org/10.1016/j.jesp.2018.08.009
+# Here, the thresholds take into account the variance of the data, trying to better cover the full range of values.
+toOrdinalDistr = function(values, nlevels, equidistant = FALSE) {
+	lim <- 2*sd(values) # I take 2 standard deviations that should contain the largest portion of the data (around 95% of the data values)
+
+	thresholds <- NA
+	if(equidistant) {
+		thresholds <- seq(from=-lim, to=lim, by=2*lim/nlevels)[c(-1, -(nlevels + 1))]
+	} else {
+		thresholds <- sort(runif(nlevels - 1))*2*lim-lim
+	}
+
+	toOrdinal(values, thresholds)
+}
+
+
+# This method takes the template and randomly assigns ordinal values to the response variable y 
+# We set the number of levels in the ordinal scale
+createRandomSample <- function(df, nlevels = 5){
+	# We first draw values from a standard normal distribution
+	df$y <- rnorm(nrow(df), mean = 0, sd = 1) 
+
+	# We then descretize these values usign either equidistant or flexible thresholds	
+	df$y <- toOrdinalDistr(df$y, nlevels, equidistant = FALSE)
+
+	df
+}
+
+# INT implementation
+INT <- function(x){
[…]

### 2025-05-24 supplementary/kay/5_lognormal-homoscedastic/test-20-x1.R

@@ -1,0 +1,125 @@
+# author: Theophanis Tsandilas
+# This code reads a template of a 4x3 repeated-measures design and makes the response variable take values from lognormal ditributions
+# where all distributions have equal variances on the scale of the actual observations
+# It then evaluates the Type I error rate of ART for the effect on x2 and the interaction effect when there is a strong effect on x1 
+
+rm(list=ls())
+
+library(dplyr)
+library(tidyr)
+
+library(lmerTest)
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+
+# These methods enable us to control the mean and standard deviation of the lognormal distributions on the scale of the actual observations
+# See: https://en.wikipedia.org/wiki/Log-normal_distribution#Generation_and_parameters
+sdlog <- function(mean, sd) {
+	sqrt(log(1 + sd^2/mean^2))	
+}
+meanlog <- function(mean, sd) {
+	log(mean^2/sqrt(sd^2 + mean^2))
+}
+
+
+# This method takes the template and assigns lognormal values to the response variable y 
+# For each of the four levels of x1 we draw samples from a differet lognormal distribution
+# mean: vector containing the means of these distributions (on the scale of the actual observations)
+# sd: the common standard deviation of the distributions (on the scale of the actual observations)
+# Here, we fix the meanslog parameters to specific values, but you can try different combinations: 
+# (ART's errors will further increase if you choose a larger sd) 
+createRandomSample <- function(df, means = c(0.5, 1, 1.5, 2), sd = 1.5){
+	# We first create some random subject-level intercepts (drawn from normal ditribution)
+	# You can vary the sd expressing individual variability
+	df <- df %>% group_by(s) %>% mutate(intercept = rnorm(1, mean = 0, sd = 0.3))
+
+	#  I know that x1 has four levels: A1, A2, A3, A4
+	ncell <- nrow(df) / 4 # This is the number of observations for each level of x1
+
+	# Make sure that the response is double precision	
+	df <- df %>% mutate(y = as.double(y))
+
[…]

### 2025-05-24 supplementary/kay/5_lognormal-homoscedastic/test-20-x2.R

@@ -1,0 +1,124 @@
+# author: Theophanis Tsandilas
+# This code reads a template of a 4x3 repeated-measures design and makes the response variable take values from lognormal ditributions
+# where all distributions have equal variances on the scale of the actual observations
+# It then evaluates the Type I error rate of ART for the effect on x1 and the interaction effect when there is a strong effect on x2 
+
+rm(list=ls())
+
+library(dplyr)
+library(tidyr)
+
+library(lmerTest)
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+
+# These methods enable us to control the mean and standard deviation of the lognormal distributions on the scale of the actual observations
+# See: https://en.wikipedia.org/wiki/Log-normal_distribution#Generation_and_parameters
+sdlog <- function(mean, sd) {
+	sqrt(log(1 + sd^2/mean^2))	
+}
+meanlog <- function(mean, sd) {
+	log(mean^2/sqrt(sd^2 + mean^2))
+}
+
+
+# This method takes the template and assigns lognormal values to the response variable y 
+# For each of the three levels of x2 we draw samples from a differet lognormal distribution
+# mean: vector containing the means of these distributions (on the scale of the actual observations)
+# sd: the common standard deviation of the distributions (on the scale of the actual observations)
+# Here, we fix the meanslog parameters to specific values, but you can try different combinations: 
+# (ART's errors will further increase if you choose a larger sd) 
+createRandomSample <- function(df, means = c(0.5, 1, 2), sd = 1.5){
+	# We first create some random subject-level intercepts (drawn from normal ditribution)
+	# You can vary the sd expressing individual variability
+	df <- df %>% group_by(s) %>% mutate(intercept = rnorm(1, mean = 0, sd = 0.3))
+
+	#  I know that x2 has three levels: B1, B2, B3
+	ncell <- nrow(df) / 3 # This is the number of observations for each level of x2
+
+	# Make sure that the response is double precision	
+	df <- df %>% mutate(y = as.double(y))
+
[…]

### 2025-05-24 supplementary/kay/5_lognormal-homoscedastic/test-independent-20-x1.R

@@ -1,0 +1,121 @@
+# author: Theophanis Tsandilas
+# This code reads a template of a 4x3 repeated-measures design and makes the response variable take values from lognormal ditributions
+# where all distributions have equal variances on the scale of the actual observations
+# It then evaluates the Type I error rate of ART for the effect on x2 and the interaction effect when there is a strong effect on x1 
+# We assume here a between-subjects design.
+
+rm(list=ls())
+
+library(dplyr)
+library(tidyr)
+
+library(lmerTest)
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+
+# These methods enable us to control the mean and standard deviation of the lognormal distributions on the scale of the actual observations
+# See: https://en.wikipedia.org/wiki/Log-normal_distribution#Generation_and_parameters
+sdlog <- function(mean, sd) {
+	sqrt(log(1 + sd^2/mean^2))	
+}
+meanlog <- function(mean, sd) {
+	log(mean^2/sqrt(sd^2 + mean^2))
+}
+
+
+# This method takes the template and assigns lognormal values to the response variable y 
+# For each of the four levels of x1 we draw samples from a differet lognormal distribution
+# mean: vector containing the means of these distributions (on the scale of the actual observations)
+# sd: the common standard deviation of the distributions (on the scale of the actual observations)
+# Here, we fix the meanslog parameters to specific values, but you can try different combinations: 
+# (ART's errors will further increase if you choose a larger sd) 
+createRandomSample <- function(df, means = c(0.5, 1, 1.5, 2), sd = 1.5){
+	#  I know that x1 has four levels: A1, A2, A3, A4
+
+	ncell <- nrow(df) / 4 # This is the number of observations for each level of x1
+
+	# Let's now draw ncell1 observations for each level but from a different lognormal distribution.  
+	# This sampling process does not make any distinction among the different levels of x2
+	df[df$x1 == "A1",]$y <- rlnorm(ncell, meanlog = meanlog(means[1], sd), sdlog(means[1], sd))
[…]

### 2025-05-24 supplementary/kay/5_lognormal-homoscedastic/test-independent-20-x2.R

@@ -1,0 +1,118 @@
+# author: Theophanis Tsandilas
+# This code reads a template of a 4x3 repeated-measures design and makes the response variable take values from lognormal ditributions
+# where all distributions have equal variances on the scale of the actual observations
+# It then evaluates the Type I error rate of ART for the effect on x1 and the interaction effect when there is a strong effect on x2 
+
+rm(list=ls())
+
+library(dplyr)
+library(tidyr)
+
+library(lmerTest)
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+
+# These methods enable us to control the mean and standard deviation of the lognormal distributions on the scale of the actual observations
+# See: https://en.wikipedia.org/wiki/Log-normal_distribution#Generation_and_parameters
+sdlog <- function(mean, sd) {
+	sqrt(log(1 + sd^2/mean^2))	
+}
+meanlog <- function(mean, sd) {
+	log(mean^2/sqrt(sd^2 + mean^2))
+}
+
+
+# This method takes the template and assigns lognormal values to the response variable y 
+# For each of the three levels of x2 we draw samples from a differet lognormal distribution
+# mean: vector containing the means of these distributions (on the scale of the actual observations)
+# sd: the common standard deviation of the distributions (on the scale of the actual observations)
+# Here, we fix the meanslog parameters to specific values, but you can try different combinations: 
+# (ART's errors will further increase if you choose a larger sd) 
+createRandomSample <- function(df, means = c(0.5, 1, 2), sd = 1.5){
+	#  I know that x2 has three levels: B1, B2, B3
+
+	ncell <- nrow(df) / 3 # This is the number of observations for each level of x2
+
+	# Let's now draw ncell1 observations for each level but from a different lognormal distribution.  
+	# This sampling process does not make any distinction among the different levels of x1
+	df[df$x2 == "B1",]$y <- rlnorm(ncell, meanlog = meanlog(means[1], sd), sdlog(means[1], sd))
+	df[df$x2 == "B2",]$y <- rlnorm(ncell, meanlog = meanlog(means[2], sd), sdlog(means[2], sd))
[…]

### 2025-05-24 supplementary/kay/data-template/create-template.R

@@ -1,0 +1,23 @@
+# author: Theophanis Tsandilas
+# This code create a 4x3 repeated-measures design, where all response values are zero
+
+rm(list=ls())
+
+library(dplyr)
+library(tidyr)
+library(faux) # A great library for simulating data for a large range of desings: https://debruine.github.io/faux
+
+# Number of subjects  -- We propose testing both a small sample size a large one n = 1000 (to assess the asymptotic behavior of the methods)
+n <- 20
+# n <- 1000
+
+design <- c(4,3) # Focus on a 4x3 design
+
+data <- add_random(s = n) %>%
+	add_within("s", x1 = paste('A', 1:design[1], sep=""), x2 = paste('B', 1:design[2], sep="")) %>% 
+	mutate(y = 0) # All responses are set to 0
+data$s <- factor(data$s)
+
+# Store the data template
+csvfile <- paste("template-", n, ".csv", sep="")
+write.csv(data, file = csvfile, row.names=FALSE, quote=F)

### 2025-05-24 appendix.qmd
updated repository from jovi

---
title: "The illusory promise of the Aligned Rank Transform"
subtitle: "Appendix. Additional experimental results"
author: 
  - name: Theophanis Tsandilas
    orcid: 0000-0002-0158-228X
    email: theophanis.tsandilas@inria.fr
    affiliations:
      - name: Université Paris-Saclay, CNRS, Inria, LISN
        country: France
  - name: Géry Casiez
    orcid: 0000-0003-1905-815X
    email: gery.casiezuniv-lille.fr
    affiliations:
      - name: Univ. Lille, CNRS, Inria, Centrale Lille, UMR 9189 CRIStAL
        country: France
bibliography: bibliography.bib
csl: canadian-journal-of-philosophy.csl

tbl-cap-location: top
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Useful libraries
library(crosstalk)
library(kableExtra)
library(gridExtra)
library(lmerTest)
library(tidyverse)
library(plotly)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Source code for reading data from experimental results and for plotting
source("dataReaders.R")
source("plotting.R")

library(dplyr)
library(tidyverse)
source("plotlying.R")
```

```{=html}
<style>
.math.inline .MathJax  {
  font-size: 105% !important;
}
</style>
```

We present complementary results and new experiments that investigate additional scenarios. We also compare INT and RNK with other nonparametric methods. Unless explicitly mentioned in each section, we follow the experimental methodology presented in the main article. At the end of each section, we summarize our conclusions.

## Results for $\alpha = .01$ {#alpha}
Although we only presented results for $\alpha = .05$ in the main article, we observe the same trends for other significance levels. The following figures show the Type I error rates of the methods for the $4 \times 3$ within-subjects design when $\alpha = .01$. Note that error rates are not proportional to the $\alpha$ level.

::: {#fig-ratio-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix1 <- "1_test_4x3_Ratio"
alpha <- .01
distributions1 <- c("norm", "lnorm", "exp", "poisson", "binom")
dnames1 <- c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial")
df1 <- readlyData(prefix1, alpha, 1, distributions1, dnames1)

prefix2 <- "2_test_4x3_Ordinal"
distributions2 <- c("likert5B")
dnames2 <- c("Ordinal (5 levels)")
df2 <- readlyData(prefix2, alpha, 1, distributions2, dnames2)

df <- rbind(df1,df2)

plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 27.1)
```
Type I error rates ($\alpha = .01$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
:::

::: {#fig-ratio-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyError(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 27.1)
```
Type I error rates ($\alpha = .01$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
:::

::: {#fig-ratio-interaction-2}
```{r, echo=FALSE, message=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df1 <- readlyData(prefix1, alpha, 0, distributions1, dnames1)
df2 <- readlyData(prefix2, alpha, 0, distributions2, dnames2)
df <- rbind(df1,df2)

plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 104)
```
Type I error rates ($\alpha = .01$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ 
:::

For results from different experiments, we refer readers to our raw result files.

### Conclusion
Results for $\alpha = .01$ confirm that ART is not a robust method. It only behaves correctly when distributions are normal.

## Main effects in the presence of interactions {#interactions}
In all experiments assessing Type I error rates reported in our article, we assumed no interaction effects. However, we also need to understand whether weak or strong interaction effects could affect the sensitivity of the methods in detecting main effects. This experiment evaluates the Type I error rate of the methods in the presence of an interaction effect alone, or alternatively, in the presence of a simultaneous main effect. We focus again on the three two-factor experimental designs we evaluated for our previous experiment, setting the sample size to $n = 20$.

To simulate populations in which interactions emerge in the absence of main effects, we examine perfectly symmetric cross-interactions. To this end, we slightly change the method we use to encode the levels of each factor, such that levels are uniformly positioned around 0. For a factor with three levels, we numerically encode the levels as $\{-0.5, 0, 0.5\}$. For a factor with four levels, we encode them as $\{-0.5, -0.1667, 0.1667, 0.5\}$.

**Interaction effect only**. We first test how the interaction effect alone influences the Type I error rate on $X_2$. @fig-interactions-main-2 presents our results. We observe that both PAR and ART fail for many configurations. Error rates are especially high in the case of the $4 \times 3$ within-subjects design and the $2\times 4$ mixed design under the log-normal and exponential distribution. ART exhibits the worst performance. In contrast, RNK and INT keep error rates close to nominal levels. However, when interaction effects become sufficiently large ($a_{12} > 4$), we observe that under the binomial and ordinal scale, they also start inflating errors.

::: {#fig-interactions-main-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
alpha <- .05

prefix <- "Appendix_test_growing-interactions"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha, effectType = 0, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX1", xvar = "effectX1X2", max = 104)

```
Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$ ($n=20$)
:::

::: {#fig-interactions-main-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX2", xvar = "effectX1X2", max = 104)

```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$ ($n=20$)
:::

**Interaction effect combined with main effect**. We also evaluate the Type I error rate on $X_2$ when the interaction effect is combined with a main effect on $X_1$. @fig-interactions-main-3 presents our results. The error rates of ART and PAR now explode for all non-normal distributions and all three designs. But the performance of RNK and INT is also affected. Their error rates become extremely low under continuous distributions, which suggests a lack of power in detecting small main effects when strong effects of other factors are combined with strong interactions. In contrast, the Type I error of the two methods explodes under the binomial and ordinal scale. Interestingly, RNK exhibits the best performance in these tests.

::: {#fig-interactions-main-3}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha, effectType = 6, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of effects", var = "rateX2", xvar = "effectX1X2", max = 104)

```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitudes $a_1 = a_{12}$ of a combined main effect on $X_1$ and an interaction effect $X_1 \times X_2$ ($n=20$)
:::

### Conclusion
When distributions are skewed, the presence of an interaction effect can cause ART to detect a non-existent main effect. ART is more sensitive to such problems than parametric ANOVA. The performance of INT and RNK can also be affected by the presence of interaction effects, especially when the interaction effect is combined with a main effect, and distributions are either binomial or ordinal. More generally, main effects should be interpreted with caution when strong interactions exist.

## Missing data {#missing}
We evaluate how missing data can affect the performance of the four methods. Specifically, we study a scenario, where a random sample of $10\%$ of the observations is missing. Missing observations lead to unbalanced designs. However, we emphasize that our scenario does not cover systematic imbalances due to missing data for specific levels of a factor.

**Main effects**. @fig-missing-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases. We observe that missing data make the error rate for ART to increase even further, with a larger increase for the mixed design. This is also the case for the normal distribution. In contrast, the accuracy of the three other methods does not seem to be affected.

::: {#fig-missing-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test_missing"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha = .05, effectType = 1, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 98)

```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
:::

**Interaction effects**. @fig-missing-interaction1 and @fig-missing-interaction2 present Type I error rates for the interaction effect in the presence of a single main effect or two parallel main effects. The error levels for all methods, including ART, are now very similar to the ones observed with no missing data.

::: {#fig-missing-interaction1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 98)

```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
:::

::: {#fig-missing-interaction2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha = .05, effectType = 0, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 103)

```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitudes $a_{1}$ and $a_{2}$ of the main effects ($n=20$)
:::

### Conclusion
ART is sensitive to the presence of missing data when at least $10\%$ of the observations are missing. Specifically, we found that its Type I error rate for main effects increases further, even with normal distributions. The other methods do not seem to be affected by missing data. However, we emphasize that we only tested missing data that are randomly drawn from the complete set of observations. Results might be different if there were systematic bias in the data imbalance structure.

## Log-normal distributions {#log-normal}
In a different experiment, we evaluate log-normal distributions with a wider range of $\sigma$ parameters (see @fig-lognormal-distributions), in particular distributions with less variance, which exhibit a lower degree of skew.

::: {#fig-lognormal-distributions}
```{r, echo=FALSE, fig.height=1.3, fig.width = 9, warning=FALSE,  message=FALSE}
source("effects_plot.R")
plotEffectsLognormal(nlevels = 2, effectSize = 2)
```
Log-normal distributions that we evaluate, shown here for a factor with two levels and a magnitude of effect $a_{i} = 2$. 
:::

**Main effects**. @fig-lognormal-main presents our results on Type I error rates for main effects. As expected, ART's inflation of error rates is less serious when distributions are closer to normal, while the problem becomes worse as distributions are more skewed.

::: {#fig-lognormal-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test_lognormal"
distributions = c("lnorm-0.2", "lnorm-0.4", "lnorm-0.6", "lnorm-0.8", "lnorm-1.0", "lnorm-1.2")
dnames = c("sd = 0.2", "sd = 0.4", "sd = 0.6", "sd = 0.8", "sd = 1.0", "sd = 1.2")

df <- readData(prefix, n = 20, alpha = .05, effectType = 1, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 69)

```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
:::

**Interaction effects**. We observe similar patterns for the Type I error rate of the interaction effect in the presence of a single main effect (@fig-lognormal-interaction1) or two parallel main effects (@fig-lognormal-interaction2). As shown in @fig-lognormal-interaction2, any advantage of ART over RNK and INT for testing interaction disappears even when distributions exhibit light skew levels. We also observe again that the performance of RNK and INT remains identical across all skew levels.

::: {#fig-lognormal-interaction1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 69)

```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
:::

::: {#fig-lognormal-interaction2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha = .05, effectType = 0, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 103)

```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitudes $a_{1}$ and $a_{2}$ of the main effects ($n=20$)
:::

### Conclusion
ART's robustness issues become less severe as log-normal distributions become less skewed and thus closer to normal. However, even under distributions with light skew, ART's Type I error rates for interactions reach higher levels than those of both INT and RNK when parallel main effects are present.

## Binomial distributions {#binomial}
We also evaluate a wider range of parameters for the binomial distribution. We focus on the lower range of probabilities $p$. However, we expect results to be identical for their symmetric probabilities $1-p$. Specifically, we test $p=.05$, $.1$, and $.2$, and for each, we consider $k=5$ and $10$ task repetitions (Bernoulli trials).

**Main effects**. We present our results for the main effect in @fig-binom-main. We observe that ART's Type I error rates increase as the number of repetitions decreases and the probability of success approaches zero, reaching very high levels when $k=5$ and $p=.05$. This trend is consistent across designs. The other methods maintain low error rates. However, their error rates fall below nominal levels when the magnitude of the effect on $X_1$ grows beyond a certain threshold, indicating a loss of power in these cases.

::: {#fig-binom-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test_binom"
distributions = c("binom-5-05", "binom-5-10", "binom-5-20", "binom-10-05", "binom-10-10", "binom-10-20")
dnames = c("k=5, p=5%", "k=5, p=10%", "k=5, p=20%", "k=10, p=5%", "k=10, p=10%", "k=10, p=20%")

df <- readData(prefix, n = 20, alpha = .05, effectType = 1, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 98)

```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
:::

**Interaction effects**. @fig-binom-interaction1 shows similar patterns for the Type I error rate of the interaction effect in the presence of a single main effect. When both main effects increase beyond a certain level (see @fig-binom-interaction2), all methods seem to fail to control the error rate. ART again demonstrates the worst behavior, systematically inflating error rates even when main effects are absent. In several cases, RNK performs better than INT.

::: {#fig-binom-interaction1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 98)

```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
:::

::: {#fig-binom-interaction2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha = .05, effectType = 0, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 103)

```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitudes $a_{1}$ and $a_{2}$ of the main effects ($n=20$)
:::

### Conclusion
ART is extremely problematic under binomial distributions, raising Type I error at very high levels even in the absence of other effects. Testing interactions in the presence of parallel main effects can be problematic for all other methods.

## Ordinal data {#ordinal}
Given the frequent use of ART with ordinal data, we evaluate our complete set of ordinal scales, based on both equidistant and flexible thresholds, with additional experimental designs.

**Main effects**. @fig-ordinal-designs-main presents Type I error rates for the main effect. ART preserves error rates at nominal levels under the $2 \times 3$ between-subjects design and the $2 \times 3$ mixed design, as long as thresholds are equidistant. Under the two within-subject designs, it inflates error rates, especially when there are fewer ordinal levels with flexible thresholds.

::: {#fig-ordinal-designs-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test_ordinal"
alpha = 0.05

distributions <- c("likert5", "likert5B", "likert7", "likert7B", "likert11", "likert11B")
dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")

df <- readData(prefix, n = 20, alpha, effectType = 1, distributions)
df <- reshapeByDesign(df, dnames)

plotlyErrorByDesign(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 62)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
:::

**Interaction effects**. @fig-missing-interaction1 and @fig-missing-interaction2 present Type I error rates for the interaction effect in the presence of a single main effect or two parallel main effects. These results lead to similar conclusions. Even in cases where ART keeps error rates close to nominal levels (e.g., under the between-subjects design with equidistance thresholds), the performance of parametric ANOVA is constantly better.

::: {#fig-ordinal-designs-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 62)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$) 
:::

::: {#fig-ordinal-designs-interaction-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha, effectType = 0, distributions)
df <- reshapeByDesign(df, dnames)
plotlyErrorByDesign(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$)
:::

### Conclusion
ART's inflation of Type I error rates with ordinal data is confirmed across a range of designs. For the between-subjects and mixed designs, the problem primarily concerns ordinal scales with flexible thresholds. However, for within-subjects designs, ART also inflates error rates for scales with equidistant thresholds, particularly when the number of levels is as low as five or seven. Again, all methods may fail to correctly infer interactions when parallel main effects exceed a certain threshold.

## ART with median alignment {#median-alignment}
We evaluate a modified implementation of ART (ART-MED), where we use medians instead of means to align ranks. This approach draws inspiration from results by @Salter:1993, showing that median alignment corrects ART's instable behavior under the Cauchy distribution. We only test the $4 \times 3$ within-participants design for sample sizes $n=10$, $20$, and $30$. For this experiment, we omit the RNK method and only present results for non-normal distributions.

We emphasize that @Salter:1993 only apply mean and median alignment to interactions. Our implementation for main effects is based on the alignment approach of @wobbrock:2011, where we simply replace means by medians.

**Main effects**. Our results presented in @fig-median-main demonstrate that median alignment (ART-MED), or at least our implementation of the method, is not appropriate for testing main effects. Although Type I error rates are now lower for the Cauchy distribution compared to the original method, they are still above nominal levels. In addition, they are significantly higher for all other distributions.

::: {#fig-median-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test_4x3_ART-Median"
alpha <- .05

methods = c("PAR", "INT", "ART", "ART-MED")
palette = c("#888888", "#009E73", "#FF5E00", "#FFA27F")
symbols = c("asterisk",  "star-diamond", "star-triangle-up", "star-triangle-down")

distributions <- c("lnorm", "exp", "cauchy", "poisson", "binom", "likert5B")
dnames <- c("Log-normal", "Exponential", "Cauchy", "Poisson", "Binomial", "Ordinal (5 levels)")
df <- readlyData(prefix, alpha, 1, distributions, dnames, methods = methods)

plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 64, cbPalette = palette, symbols = symbols)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
:::

**Interaction effects**. In contrast, median alignment works surprisingly well for interactions, correcting deficiencies of ART, especially when main effects are absent or weak. @fig-median-interaction-1 and @fig-median-interaction-2 present our results. Despite this improved performance, we cannot recommend using the method because it still cannot compete with INT. Additionally, its advantages over parametric ANOVA are only apparent for the Cauchy distribution.

::: {#fig-median-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}

plotlyError(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 64, cbPalette = palette, symbols = symbols)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
:::

::: {#fig-median-interaction-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readlyData(prefix, alpha, 0, distributions, dnames, methods = methods)
plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105, cbPalette = palette, symbols = symbols)

```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ 
:::

### Conclusion
Using median instead of mean alignment with ART significantly improves the method's performance in testing interactions across all the distributions we tested. However, we cannot recommend it, as the method is still less robust than INT. Furthermore, it is unclear how to apply median alignment for testing main effects --- using medians with the alignment method of @wobbrock:2011 results in extremely high error rates.

## Nonparametric tests in single-factor designs {#nonparametric-tests}
We compare PAR, RNK, and INT to nonparamatric tests for within- and between-subjects single-factor designs, where the factor has two, three, or four levels. Depending on the design, we use different nonparametric tests. For within-subjects designs, we use the Wilcoxon sign-rank test if the factor has two levels (*2 within*) and the Friedman test if the factor has three (*3 within*) or four (*4 within*) levels. For between-subjects designs (*2 between*, *3 between*, and *4 between*), we use the Kruskal–Wallis test.

**Power**. @fig-nonparametric-power compares the power of the various methods as the magnitude of the main effect increases, where we use the abbreviation *NON* to designate a nonparametric test. We observe that primarily INT, but also RNK, generally exhibit better power than the nonparametric tests. Differences are more pronounced for within-subjects designs, corroborating Conover's [-@Conover:2012] observation that the rank transformation results in a test that is superior to the Friedman test under certain conditions.

::: {#fig-nonparametric-power}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
source("plotlying-v3.R")

readSingle <- function(prefix, n, alpha, distributions, methods, POWER = FALSE) {
  	df <- read.csv(paste("data/", prefix, "_", n, ".csv" , sep=""), sep=",", header=TRUE, strip.white=TRUE)
	  df$distr <- factor(df$distr, levels=distributions)
	  df$method <- factor(df$method, levels=methods)

    if(POWER) {
      df <- df[df$effect > 0,]
    }
    else {
      df <- df[df$effect == 0,]
    }

    if(is.na(alpha)) return(df)
    else return(df[df$alpha == alpha,])
}

prefix <- "Appendix_test-One-Factor"
alpha = 0.05

distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
methods = c("PAR", "RNK", "INT", "NON")
palette = c("#888888", "#E69F00", "#009E73", "#FF70AB")

df <- readSingle(prefix, n = 20, alpha, distributions, methods = methods, POWER = TRUE)
#df <- reshapeByDesign(df, dnames, effectvars = c("effect"))
df <- df %>% arrange(design,distr,effect,rates)  %>% group_by(design,distr,effect) %>% mutate(rank = rank(rates))

df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effect"))
plotlyPowerByDesign_v3(df, xlab = "magnitude of effect", var = "rank", xvar = "effect", max = 4.3, cbPalette = palette)
```
Ranking of methods comparing their power ($\alpha = .05$) as a function of the magnitude of effect ($n = 20$). 
:::

We expect that the accuracy of ANOVA on rank-transformed values will decrease with smaller samples. However, our tests with smaller samples of $n=10$ show that INT remains robust and still outperforms other nonparametric methods. Although it is possibe to couple INT with permutation testing for higher accuracy [@Beasley:2009], we have not explored this possibility here.

**Type I error rate under equal and unequal variances**. @fig-nonparametric-unequal-var presents the rate of positives under conditions of equal ($r_{sd} = 0$) and unequal variances ($r_{sd} > 0$). While this rate can be considered a Type I error rate when variances are equal, interpreting it under other conditions requires special attention because the hypothesis of interest may differ among methods. Parametric ANOVA is particularly sensitive to unequal variances when distributions are skewed because it tests differences among means. While the normal distributions of the latent space have the same means, this is not the case with the skewed distributions of the transformed variable, which have the same median but different means. All nonparametric methods we tested use ranks, which preserve medians and mitigate this problem. However, their rate of positives can still exceed $5\%$ under certain conditions.

For between-subjects designs, we observe that the Kruskal–Wallis test and RNK yield very similar results. This is not surprising, as RNK is known to be a good approximation of the Kruskal–Wallis test [@Conover:2012]. INT's positive rates are similar, although slightly higher under the binomial distribution. For within-subjects designs, differences among methods are more pronounced. The Wilcoxon sign-rank test (*2 within*) inflates rates well above $5\%$, demonstrating that the test is not a pure test of medians. In contrast, the Friedman test (*3 within* and *4 within*) provides the best control among all methods.

::: {#fig-nonparametric-unequal-var}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test-One-Factor-Unequal-Var"

df <- readSingle(prefix, n = 20, alpha, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("sd_ratio", "effect"))

plotlyErrorByDesign_v3(df, xlab = "max ratio of standard deviations", var = "rates", xvar = "sd_ratio", ytitle = 'Positives (%)', max = 24.2, nticks=8, cbPalette = palette)
```
Positives rates ($\alpha = .05$) as a function of the maximum ratio $r_{sd}$ of standard deviations between levels ($n = 20$). 
:::

@fig-nonparametric-unequal-var-alpha01 presents the same results but for $\alpha = .01$. Discrepancies among different methods are now more pronounced, and we notice again that the Friedman test keeps rates closer to the nominal level of $1\%$ compared to INT and RNK. Nevertheless, in addition to their greater power compared to the Friedman test, RNK or INT present other advantages, such as the possibility to use common ANOVA-based procedures to partly correct issues associated with unequal variances. For instance, we ran an experiment, where we applied a Greenhouse–Geisser correction following sphericity violations detected with the Mauchly's sphericity test ($\alpha = .05$). We found that this correction brings the error rates of INT close to nominal levels for continuous distributions, such as the normal and log-normal distributions. In the case of the binomial and ordinal distributions, error rates are also significantly reduced, well below those of the Friedman test, although not reaching nominal levels.

::: {#fig-nonparametric-unequal-var-alpha01}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readSingle(prefix, n = 20, alpha= 0.01, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("sd_ratio", "effect"))

plotlyErrorByDesign_v3(df, xlab = "max ratio of standard deviations", var = "rates", xvar = "sd_ratio", ytitle = 'Positives (%)', max = 7.2, nticks=8, cbPalette = palette)
```
Positives rates ($\alpha = .01$) as a function of the maximum ratio $r_{sd}$ of standard deviations between levels ($n = 20$). 
:::

### Conclusion
We do not see significant benefits in using dedicated nonparametric tests over RNK or INT. INT can replace nonparametric tests even for single-factor designs. If, after transforming the data, the assumptions of homoscedasticity or sphericity are still not met, applying common correction procedures (e.g., a Greenhouse–Geisser correction for sphericity violations) on the transformed data can reduce the risk of Type I errors.

## ANOVA-type statistic (ATS) {#ATS}
We compare PAR, RNK, and INT to the ANOVA-type statistic (ATS) [@Brunner_ATS:2001] for two-factor designs. We use its implementation in the R package *nparLD* [@nparLD], which does not support between-subjects designs. Thus, we only evaluate it for the $4 \times 3$ within-subjects and the $2 \times 4$ mixed designs.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("plotlying-v2.R")
```

**Type I error rates: Main effects**. @fig-ATS-designs-main presents Type I error rates for the main effect of $X_2$. Under the mixed design, RNK, INT, and ATS exhibit very similar error rates, which are close to nominal levels. In the within-subjects design, the error rates of ATS tend to be slightly above $5\%$. Additionally, unlike the other methods whose error rates drop significantly below $5\%$ when the effect of $X_1$ becomes stronger under binomial and ordinal scales, the power of ATS does not seem to be affected in these cases.

::: {#fig-ATS-designs-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test-ATS"
alpha = 0.05

distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
methods = c("PAR", "RNK", "INT", "ATS")
palette = c("#888888", "#E69F00", "#009E73", "#FF70AB")

df <- readData(prefix, n = 20, alpha, effectType = 1, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotlyErrorByDesign_v2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 7.1, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
:::

@fig-ATS-designs-main-2 presents results for the main effect of $X_1$. The error rates of ATS are now slightly inflated under the mixed design. The other methods exhibit the same trends as for the other factor.

::: {#fig-ATS-designs-main-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df2 <- readData(prefix, n = 20, alpha, effectType = 2, distributions, methods = methods)
df2 <- reshapeByDesign(df2, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotlyErrorByDesign_v2(df2, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX2", max = 7.1, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as a function of the magnitude $a_2$ of the main effect of $X_2$ ($n = 20$)
:::

**Type I error rates: Interactions**. @fig-ATS-designs-interaction-1 presents Type I error rates for the interaction in the presence of a single main effect. Results are again very similar for all three nonparametric methods under the mixed design. In contrast, the error rates of ATS tend to be lower than nominal levels under the within-subjects design, often falling below $4\%$.

::: {#fig-ATS-designs-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign_v2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 7.1, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$) 
:::

 When two parallel main effects are present, ATS and RNK lead to very similar trends (see @fig-ATS-designs-interaction-2). Overall, INT appears to be a more robust method with the exception of the binomial distribution, for which error rates are higher for this method.

::: {#fig-ATS-designs-interaction-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha, effectType = 0, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign_v2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 104, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$)
:::

**Power: Main effects**. As shown in @fig-ATS-power-main-1 and @fig-ATS-power-main-2, ATS appears as the most powerful method for detecting effects of $X_1$ for the mixed design. However, in all other situations, it has less power than INT and offer less power than RNK.

::: {#fig-ATS-power-main-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
alpha <- .05

prefix <- "Appendix_test-ATS-Power"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha, effectType = 3, distributions, methods = methods)

df <- df %>% arrange(design,distr,effectX1,rateX1)  %>% group_by(design,distr,effectX1) %>% mutate(rank = rank(rateX1))
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyPowerByDesign_v2(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
```
Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_1$** as a function of the magnitude of effect $a_1$ ($n=20$). Hover over the charts to inspect their power ($\%$).
:::

::: {#fig-ATS-power-main-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha, effectType = 4, distributions, methods = methods)

df <- df %>% arrange(design,distr,effectX2,rateX2)  %>% group_by(design,distr,effectX2) %>% mutate(rank = rank(rateX2))
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyPowerByDesign_v2(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
```
Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_2$** as a function of the magnitude of effect $a_2$ ($n=20$). Hover over the charts to inspect their power ($\%$).
:::

**Power: Interactions**. @fig-ATS-power-interaction shows results on power for interactions. INT emerges again as the most powerful method. The power of ATS is particularly low under the within-subjects design.

::: {#fig-ATS-power-interaction}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha, effectType = 5, distributions, methods = methods)

df <- df %>% arrange(design,distr,effectX1X2,rateX1X2)  %>% group_by(design,distr,effectX1X2) %>% mutate(rank = rank(rateX1X2))
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyPowerByDesign_v2(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
```
Ranking of methods for their power ($\alpha = .05$) to detect the **interaction effect $X_1 \times X_2$** as a function of the magnitude of effect $a_{12}$ ($n=20$). Hover over the charts to inspect their power ($\%$).
:::

### Conclusion
Although ATS appears to be a valid alternative, it does not offer clear performance advantages over INT, which is also simpler and more versatile.

## Generalizations of nonparametric tests {#generalized}

Finally, we evaluate the generalizations of nonparametric tests recommended by Lüpsen [-@luepsen:2018; -@luepsen:2023] as implemented in his *np.anova* function [@luepsen_R]. Specifically, we evaluate the generalization of the van der Waerden test (VDW), and the generalization of the Kruskal-Wallis and Friedman test (KWF). Their implementation uses R's *aov* function and requires considering random slopes in the error term of the model, that is, using ``Error(s/(x1*x2))`` for the two-factor within-participants design and ``Error(s/x2)`` for the mixed design, where ``s`` is the subject identifier variable. We also used the *aov* function and the same formulation of the error term for all other methods.

**Type I error rates: Main effects**. @fig-vdWaerden-designs-main presents Type I error rates for the main effect of $X_2$. While all methods perform well under the within-subjects and mixed designs, the error rates of VDW and KWF drop drastically as the effect of $X_1$ grows under the between-subjects design. We will see below that the power of these methods evaporates in these cases.

::: {#fig-vdWaerden-designs-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test-vdWaerden"
alpha = 0.05

distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
methods = c("RNK", "INT", "VDW", "KWF")
palette = c("#E69F00", "#009E73", "#FF70AB", "#888888")

df <- readData(prefix, n = 20, alpha, effectType = 1, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 7.1, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
:::

**Type I error rates: Interactions**. @fig-vdWaerden-designs-interaction-1 presents the Type I error rates for the interaction, in the presence of a single main effect. Once again, the error rates of VDW and KWF decrease rapidly in both the between-subjects and mixed designs. @fig-vdWaerden-designs-interaction-2 shows the results when both main effects are present. Under the within-subjects design, KWF is the poorest-performing technique. While VDW outperforms RNK, it remains inferior to INT. For the between-subjects and mixed designs, KFW and VDW yield low or near-zero error rates when main effects are large, likely due to their extreme loss of power in these scenarios.

::: {#fig-vdWaerden-designs-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha, effectType = 2, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX2", max = 7.1, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_2$ of the main effect of $X_2$ ($n = 20$) 
:::

::: {#fig-vdWaerden-designs-interaction-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha, effectType = 0, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 104, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ ($n = 20$)
:::

**Power: Main effects**. @fig-vdWaerden-power-main-1 presents power results for detecting the main effect of $X_1$. For the within-subjects and mixed design, KFW and VDW exhibit a lower power than RNK and INT, but differences are generally small. Although VDW appears as powerful as INT under the between-subjects design, this only occurs when the effect of the second factor is zero. As shown in @fig-vdWaerden-power-main-1B, the power of both KFD and VDW drops drastically as the effect of $X_2$ increases for this design.

::: {#fig-vdWaerden-power-main-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
alpha <- .05

prefix <- "Appendix_test-vdWaerden-Power"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha, effectType = 3, distributions, methods = methods)

df <- df %>% arrange(design,distr,effectX1,rateX1)  %>% group_by(design,distr,effectX1) %>% mutate(rank = rank(rateX1))
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyPowerByDesign(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
```
Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_1$** as a function of the magnitude of effect $a_1$ ($n=20$). Hover over the charts to inspect their power ($\%$).
:::

::: {#fig-vdWaerden-power-main-1B}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
alpha <- .05
prefix <- "Appendix_test-vdWaerden-Power-Multieffect"

df <- readData(prefix, n = 20, alpha, effectType = 3, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX2", ytitle ="Power (%)", max = 104, nticks=6, cbPalette = palette)
```
Power ($\alpha = .05$) to detect the **main effect of $X_1$** ($a_1 = 0.8$) as a function of the magnitude of effect $a_2$ of the main effect of $X_2$ ($n=20$).
:::

@fig-vdWaerden-power-main-2 also presents results for the main effect of $X_1$, where we observe once again that the generalized tests cannot compete with the more powerful INT, or even with RNK.

::: {#fig-vdWaerden-power-main-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test-vdWaerden-Power"
df <- readData(prefix, n = 20, alpha, effectType = 4, distributions, methods = methods)

df <- df %>% arrange(design,distr,effectX2,rateX2)  %>% group_by(design,distr,effectX2) %>% mutate(rank = rank(rateX2))
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyPowerByDesign(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
```
Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_2$** as a function of the magnitude of effect $a_2$ ($n=20$). Hover over the charts to inspect their power ($\%$).
:::

**Power: Interactions**. We also present results on the power of methods to detect interactions in @fig-vdWaerden-power-interaction, confirming the advantage of INT across all designs.
@fig-vdWaerden-power-interactionB provides a clearer picture of how power is affected by the presence of a main effect. We observe that the power of all methods drops as the main effect of $X_2$ increases, but this trend is more pronounced for KFD and VDW, particularly for the between-subjects and mixed designs.

::: {#fig-vdWaerden-power-interaction}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha, effectType = 5, distributions, methods = methods)

df <- df %>% arrange(design,distr,effectX1X2,rateX1X2)  %>% group_by(design,distr,effectX1X2) %>% mutate(rank = rank(rateX1X2))
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyPowerByDesign(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
```
Ranking of methods for their power ($\alpha = .05$) to detect the **interaction effect $X_1 \times X_2$** as a function of the magnitude of effect $a_{12}$ ($n=20$). Hover over the charts to inspect their power ($\%$).
:::

::: {#fig-vdWaerden-power-interactionB}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
alpha <- .05
prefix <- "Appendix_test-vdWaerden-Power-Multieffect"

df <- readData(prefix, n = 20, alpha, effectType = 5, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX2", ytitle ="Power (%)", max = 104, nticks=6, cbPalette = palette)
```
Power ($\alpha = .05$) to detect the **interaction effect** ($a_{12} = 1.5$) as a function of the magnitude of effect $a_2$ of the main effect of $X_2$ ($n=20$).
:::

### Conclusion
Our results do not support Lüpsen's [-@luepsen:2018; -@luepsen:2023] conclusions. The behavior of the generalized nonparametric tests presents issues in numerous scenarios. While these tests exhibit lower error rates under specific conditions, this is due to a significant loss of power when other effects are at play. Therefore, we advise against the use of these methods.

## References {.unnumbered}

::: {#refs}
:::

### 2025-05-24 index.qmd
updated repository from jovi

---
title: "The illusory promise of the Aligned Rank Transform"
subtitle: "A systematic study of rank transformations"
author: 
  - name: Theophanis Tsandilas
    orcid: 0000-0002-0158-228X
    email: theophanis.tsandilas@inria.fr
    affiliations:
      - name: Université Paris-Saclay, CNRS, Inria, LISN
        country: France
  - name: Géry Casiez
    orcid: 0000-0003-1905-815X
    email: gery.casiezuniv-lille.fr
    affiliations:
      - name: Univ. Lille, CNRS, Inria, Centrale Lille, UMR 9189 CRIStAL
        country: France
bibliography: bibliography.bib
csl: canadian-journal-of-philosophy.csl

tbl-cap-location: top
---

::: {.callout-important appearance="simple"}
## Under Review {.unnumbered}
This paper is [under review](https://www.journalovi.org/under-review.html) on the experimental track of
the [Journal of Visualization and Interaction](https://www.journalovi.org/). See the 
[reviewing process](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues).
:::

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Useful libraries
library(crosstalk)
library(kableExtra)
library(gridExtra)
library(lmerTest)
library(tidyverse)
library(plotly)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Source code for reading data from experimental results and for plotting
source("dataReaders.R")
source("plotting.R")
source("effects_plot.R")
source("interactions_plot.R")
```

```{=html}
<style>
.math.inline .MathJax  {
  font-size: 105% !important;
}
</style>
```

::: {.callout-note appearance="simple" icon=false collapse=false}

{{< include text/10-abstract.qmd >}}

{{< include text/20-intro.qmd >}}

{{< include text/30-background.qmd >}}

{{< include text/40-interpretingEffects.qmd >}}

{{< include text/50-ExperimentalMethod.qmd >}}

## Results {#findings}
Each experiment concentrated on a unique combination of distributions, experimental designs, and evaluation measures. Consequently, we organize our results into several subsections, addressing both main and interaction effects.

{{< include text/60-Results-RatioScales.qmd >}}

{{< include text/61-Results-OrdinalScales.qmd >}}

{{< include text/62-Results-AcrossExpDesigns.qmd >}}

{{< include text/63-Results-UnequalVariance.qmd >}}

{{< include text/64-Results-StatPower.qmd >}}

{{< include text/65-Results-EffectSize.qmd >}}

{{< include text/70-CaseStudies.qmd >}}

{{< include text/80-Recommandations.qmd >}}

{{< include text/90-Conclusions.qmd >}}

## Acknowledgements {.unnumbered}
We express our gratitude to Pierre Dragicevic for introducing us to the concept of removable interactions [@Loftus:1978;@Wagenmakers:2012]. Additionally, we extend our thanks to Haiko Lüpsen for his invaluable assistance in correctly using his implementation of the multifactorial generalizations of the van der Waerden test, and the Kruskal-Wallis and Friedman tests. Finally, we are grateful to Sophie Siestrup and Daniel Martin for their responsiveness and for generously sharing their data with us.

## References {.unnumbered}

::: {#refs}
:::

### 2025-05-24 supplementary/examples-case-studies/1-Introduction/illustrative-example.Rmd
updated repository from jovi

@@
 ```{r, warning=FALSE}
-df <- read.csv("example-data-cut.csv", sep=",", header=TRUE, strip.white=TRUE)
+df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)
 

### 2025-05-24 supplementary/examples-case-studies/6-Case-studies/Siestrup-and-Schubotz-23/Siestrup-and-Schubotz-23.Rmd
updated repository from jovi

---
title: "Ordinal and binomial scales - Siestrup and Schubotz (2023)"
author: "Theophanis Tsandilas and Géry Casiez"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We detail here our analysis of the data of [Siestrup and Schubotz (2023)](https://doi.org/10.1162/jocn_a_02047) partly presented in our article as a case study. This dataset is not publicly available, but the authors have included a data availability statement: *"The data sets generated for this study are available on request to the corresponding author."* While the authors gave us access to the data, we don't have permission to share them online.

### Context
Siestrup and Schubotz (2023) investigated how different types of modifications in video sequences describing a story affect people’s episodic memory. During a first session (*encoding*), participants encoded episodes (filmed stories acted by manipulating PLAYMOBIL toys) by watching each video several times. On the day after the encoding, participants went through two experimental parts:

1. An incomplete reminder of each video, interrupted at the moment preceding the modification of interest. 
2. An fMRI session, where videos were presented either in their original version, or modified.

A post-fMRI memory test on the day after evaluated participants' recall of which of the videos seen was the original episode encoded during the first session. The authors studied the effects of two factors:

- *Version*. The modified version of the video. It was either *surface*, i.e., the modification did not affect the storyline of the episode, or *gist*, i.e., modification affected the storyline (or gist) of the episode.

- *Modification*. It was either *yes* or *no*, depending on whether a modified version of the eposide was presented during the fMRI session.

The authors used ART for the analysis of various measures that we revisit below. 
Although the authors have shared their data with us, the data are not publicly available. Thus, we do not include them in our supplementary materials.

### Rating task
The experiment included a rating task, asking participants to rate how much the storyline of modified episodes deviated from an original version in a scale from 1 (0% different) to 6 (100% different). The authors used ART to analyze aggregated scores from multiple ratings per condition. We replicate their analysis but also conduct the analysis with alternative methods.

We start by reading and reformatting the data:

```{r, warning=FALSE, message=FALSE}
library(tidyverse)

# We don't provide this file as we don't have the authors' permission to share it online.
ratings <- read.csv("ratings.csv", sep=",", header = TRUE)

df <- gather(ratings, modversion, rating, no_surface:yes_gist, factor_key=TRUE)
df$modification <- ifelse(df$modversion %in% c('yes_surface', 'yes_gist'), 'yes', 'no')
df$version <- ifelse(df$modversion %in% c('yes_surface', 'no_surface'), 'surface', 'gist')
df <- df[c(1,4,5,3)]

df$participant_id <- factor(df$participant_id)
df$version <- factor(df$version)
df$modification <- factor(df$modification)
```

The following plot shows the distributions of individual ratings per group, where the thick lines represent group means.

```{r, warning=FALSE, message = FALSE}
library("yarrr")
pirateplot(rating ~ modification+version, data = df, theme = 1,            
    inf.f.o = 0, # Turn off inf fill
    inf.b.o = 0, # Turn off inf border
    bean.b.o = 0,
    avg.line.fun = mean,
    bar.f.col = gray(.8), # bar filling color
    bar.b.col = gray(.4), # bar border color
    bar.f.o = .5, # bar filling opacity
    bar.b.o = .7 # bar border opacity
)
```

We conduct an analysis with four different methods: ART, PAR, RNK, and INT.
```{r, warning=FALSE, message=FALSE}
library(ARTool)
library(effectsize)

# Implementation of the Inverse Normal Transform
int = function(x){
   qnorm((rank(x) - 0.5)/length(x))
}

# We use here the aov function to be consistent with the authors' analysis, also considering slope random effects 
mr_art <- art(rating ~ 1 + modification*version + Error(participant_id/(1+modification*version)), data = df)
mr_par <- aov(rating ~ 1 + modification*version + Error(participant_id/(1+modification*version)), data = df)
mr_rnk <- aov(rank(rating) ~ 1 + modification*version + Error(participant_id/(1+modification*version)), data = df)
mr_int <- aov(int(rating) ~ 1 + modification*version + Error(participant_id/(1+modification*version)), data = df)
```

The results obtained with ART are as follows:
```{r, warning=FALSE}
# Function for printing the results for each method
printAnalysis <- function(num, method, model) {
  cat(num, ". Analysis using ", method, "\n\n", sep="")

  if(method == "ART") print(anova(model))
  else print(summary(model))

  print(eta_squared(model))
  print(cohens_f(model))
}

printAnalysis(1, "ART", mr_art)
```

We observe that in addition to a clear strong effect of the ```version``` factor, ART shows a statistically significant effect ($p = .034$) effect of the ```modification``` factor, which leads the authors to conclude that *"modified videos that had already been presented during the fMRI session received lower story-change ratings than those that had been presented in the original version before"* [(Siestrup and Schubotz, 2023)](https://doi.org/10.1162/jocn_a_02047).

The results of the other methods, however, do not support this conclusion:

```{r, warning=FALSE}
printAnalysis(2, "PAR", mr_par)
printAnalysis(3, "RNK", mr_rnk)
printAnalysis(4, "INT", mr_int)
```

### Rate of false alarms
The authors also measured the rate of false alarms for modified videos, which can be thought off as a variable following a binomial distribution. 
For this dataset, the average probabilty of a false alarm was approximately 15%.

We start again by reading the data:

```{r, warning=FALSE, message=FALSE}
# We don't provide this file as we don't have the authors' permission to share it online.
alarm_rates <- read.csv("False_Alarm_Rate.csv", sep=",", header = TRUE)

df <- gather(alarm_rates, modversion, arate, no_surface:yes_gist, factor_key=TRUE)
df$modification <- ifelse(df$modversion %in% c('yes_surface', 'yes_gist'), 'yes', 'no')
df$version <- ifelse(df$modversion %in% c('yes_surface', 'no_surface'), 'surface', 'gist')
df <- df[c(1,4,5,3)]

df$participant_id <- factor(df$participant_id)
df$version <- factor(df$version)
df$modification <- factor(df$modification)
```

The authors used again ART for the analysis of the false alarm rate.
Here, we compare the results of ART with the results of the three other methods.

```{r, warning=FALSE, message=FALSE}
# We use here the aov function to be consistent with the authors' analysis, also considering slope random effects 
ma_art <- art(arate ~ 1 + modification*version + Error(participant_id/(1+modification*version)), data = df)
ma_par <- aov(arate ~ 1 + modification*version + Error(participant_id/(1+modification*version)), data = df)
ma_rnk <- aov(rank(arate) ~ 1 + modification*version + Error(participant_id/(1+modification*version)), data = df)
ma_int <- aov(int(arate) ~ 1 + modification*version + Error(participant_id/(1+modification*version)), data = df)

printAnalysis(1, "ART", ma_art)
printAnalysis(2, "PAR", ma_par)
printAnalysis(3, "RNK", ma_rnk)
printAnalysis(4, "INT", ma_int)
```

We notice that all methods show statistically significant effects for both factors as well as we their interaction. However, ART results in small $p$-values and finds larger effect sizes.

The following figure summarizes the results.

```{r, warning=FALSE, message = FALSE}
pirateplot(arate ~ modification+version, data = df, theme = 2,            
     inf.f.o = 0, # Turn off inf fill
     inf.b.o = 0, # Turn off inf border
     point.o = .2,   # Turn up points
     bar.f.o = .3, # Turn up bars
     bean.f.o = 0, # Light bean filling
     bean.b.o = 0, # Light bean border
     avg.line.o = 1, # Turn off average line
     point.col = "black"
)
```

As we discuss in the paper, statistically significant interactions in the presence of strong main effects may not be trusted. Although the difference between the *yes* and *no* modification conditions is clearly larger for surface modifications, this trend does not necessarily imply an intrinsic interaction effect. We observe that the interaction does not cross over, and the fact that the difference is smaller for gist modifications can be simply attributed to the fact that the measure is bounded by zero.

### AUC measure
Another measure of memory performance that the authors studied using ART is the *Area Under the ROC Curve*, or AUC, which takes values between 0 and 1.
We first read and plot the data:

```{r, warning=FALSE, message=FALSE}
# We don't provide this file as we don't have the authors' permission to share it online.
auc_data <- read.csv("AUC.csv", sep=",", header = TRUE)
df <- gather(auc_data, modversion, auc, no_surface:yes_gist, factor_key=TRUE)
df$modification <- ifelse(df$modversion %in% c('yes_surface', 'yes_gist'), 'yes', 'no')
df$version <- ifelse(df$modversion %in% c('yes_surface', 'no_surface'), 'surface', 'gist')
df <- df[c(1,4,5,3)]

df$participant_id <- factor(df$participant_id)
df$version <- factor(df$version)
df$modification <- factor(df$modification)

pirateplot(auc ~ modification+version, data = df, theme = 1,            
    inf.f.o = 0, # Turn off inf fill
    inf.b.o = 0, # Turn off inf border
    bean.b.o = 0,
    avg.line.fun = mean,
    bar.f.col = gray(.8), # bar filling color
    bar.b.col = gray(.4), # bar border color
    bar.f.o = .5, # bar filling opacity
    bar.b.o = .7 # bar border opacity
)
```

We then present our analysis with the four different methods. We observe again that altough all methods lead to the same conclusions, ART finds larger effect sizes. 
Discrepencies are more pronounced for the interaction effect, which should again be interpreted with caution.

```{r, warning=FALSE, message=FALSE}
# We use here the aov function to be consistent with the authors' analysis, also considering slope random effects 
mauc_art <- art(auc ~ 1 + modification*version + Error(participant_id/(1+modification*version)), data = df)
mauc_par <- aov(auc ~ 1 + modification*version + Error(participant_id/(1+modification*version)), data = df)
mauc_rnk <- aov(rank(auc) ~ 1 + modification*version + Error(participant_id/(1+modification*version)), data = df)
mauc_int <- aov(int(auc) ~ 1 + modification*version + Error(participant_id/(1+modification*version)), data = df)

printAnalysis(1, "ART", mauc_art)
printAnalysis(2, "PAR", mauc_par)
printAnalysis(3, "RNK", mauc_rnk)
printAnalysis(4, "INT", mauc_int)
```

### 2025-05-24 supplementary/rebuttals/higgins/homoscedastic/example/additional-example.Rmd
updated repository from jovi

--- 2025-05-24 supplementary/answers/additional-example.Rmd
@@
 ---
-title: "Introduction - Illustrative example"
+title: "Rebuttal - Complimentary example (homoscedastic log-normal data)"
 author: "Theophanis Tsandilas and Géry Casiez"
@@
 
-We detail our analysis for the illustrative example presented in the Introduction.
+We detail the analysis for a complimentary example where data were drawn from homoscedastic log-normal populations.
+We present this example in our [first answer](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues/2#issuecomment-2483020115) to the commentary of Prof.
+James Higgins.
 
-We use [data](example-data-2.csv) from a hypothetical experiment that evaluates the time performance of three user interface techniques (*A*, *B*, and *C*) that help users complete image editing tasks of four different difficulty levels: *Level1*, *Level2*, *Level3*, and *Level4*.
+As in our original illustrative example, we use [data](example-data-2.csv) from a hypothetical experiment that evaluates the time performance of three user interface techniques (*A*, *B*, and *C*) that help users complete image editing tasks of four different difficulty levels: *Level1*, *Level2*, *Level3*, and *Level4*.
 

### 2025-05-24 text/20-intro.qmd
updated repository from jovi

@@
 The experiment is hypothetical but has similarities with real-world experiments, e.g., see the experiments of @Fruchard:2023.
-Time performances have been randomly sampled from a population in which: (1) *Difficulty* has a large effect; (2) *Technique* has no effect; and (3) there is no interaction effect between the two factors.
+Time performances have been randomly sampled from a population in which: (1) *Difficulty* has a large main effect; (2) *Technique* has no main effect; and (3) there is no interaction effect between the two factors.
 To generate time values, we drew samples from a log-normal distribution.

### 2025-05-24 text/30-background.qmd
updated repository from jovi

## Background {#background}
In this section, we provide background on the history, scope, construction, and use of rank transformation methods, with a particular focus on ART. Additionally, we summarize the results of previous evaluation studies and position our own work within this context.

### Nonparametric statistical tests
A common assumption of ANOVAs and linear regression is that the distribution of residuals is normal. More technically, the normality assumption concerns the sampling distribution of the mean. Therefore, when sample sizes are sufficiently large, these methods are robust to departures from the normality assumption because the sampling distribution of the mean is asymptotically normal, regardless of the shape of the population distributions (Central Limit Theorem). However, their accuracy drops if samples are relatively small and populations markedly deviate from normal.

The goal of *nonparametric* procedures is to address this problem by making fewer or no assumptions about the underlying distributions. The original idea of replacing the data values by their ranks belongs to @spearman1904. The key advantage of ranks is that they preserve the monotonic relationship of values while also allowing the derivation of an exact probability distribution. This probability distribution can then be used for statistical inference, such as calculating a *p*-value. The idea of using ranks was adopted by other researchers, leading to the development of various nonparametric tests commonly used today. Representative examples include the Mann-Whitney U test [@Mann-Whitney-U] and the Kruskal–Wallis test [@Kruskal-Wallis] for independent samples, and the Wilcoxon sign-rank test [@Wilcoxon] and the Friedman test [@Friedman] for paired samples and repeated measures.

### Rank transformations
Unfortunately, the scope of nonparametric tests is limited, as they only support simple experimental designs with a single independent variable. Rank transformations aim to address this limitation by bridging the gap between nonparametric statistics and ANOVA.

**The rank transform.** @Conover:2012 provides a comprehensive introduction to the simple rank transform, which consists of sorting the raw observations and replacing them by their ranks. In case of ties, tied observations are assigned *fractional ranks* equal to their average order position in the ascending sequence of values. For example, the two $0.3$ instances of the $Y$ responses in @fig-ART-explained (a), which are the lowest values in the dataset, receive a rank of $1.5$ (see $rank(Y)$), while the next value (a $0.4$) in the sequence receives a rank of $3$.

@conover:1981 showed that for one-factor designs, using ANOVAs on these ranks produces tests that are asymptotically equivalent or good replacements of traditional nonparametric tests (see [Section 3](#interpretation)). However, a series of studies in the 80s raised caution flags on the use of the method, showing that it may confound main and interaction effects in two-factor and three-factor experimental designs. For an extensive review of these studies, we refer readers to @Sawilowsky:1990.

**The aligned rank transform.** Given such negative results, many researchers turned to aligned (or adjusted) rank transformations. @Sawilowsky:1990 discusses several variations of aligned rank-based transformations (ART) and tests, while @higgins:1990 detail the ART method that we evaluate in this paper for two-way designs. @wobbrock:2011 generalize it to more than two factors, while 10 years later, @elkin:2021 show how to apply it to multifactor contrast tests. @fig-ART-explained explains the general method for a design with two factors, $A$ and $B$.

![Example showing the use of the rank, ART, and INT transformation methods (a). Construction of ART for a two-factor design (b,c)](images/art-explained.svg){#fig-ART-explained width=80%}

The key intuition behind the transform is that responses are ranked after *aligning* them, such that effects of interest are separated from other effects. This means that for each effect of interest, either main or interaction, we produce a separate ranking. This is clearly demonstrated through the example dataset in @fig-ART-explained (a), where each aligned ranking ($ART_A$, $ART_B$, and $ART_{AB}$) is for testing a different effect ($A$, $B$, and $A \times B$, respectively).

Notice that even for this very small dataset ($n = 2$), these three rankings are distinct from each other, as well as very different from the ranking of the simple rank transform. Interestingly, equal responses (the two $0.3$ in our example dataset) may be assigned to very different ranks, while very different responses (such as $0.3$ and $3.0$ in our dataset) can be assigned to the same rank. Furthermore, differences between the two groups $a_1$ and $a_2$ within factor $A$ are significantly exaggerated by ART (see $ART_A$ ranks). We will later see that this behavior can lead to unstable results and conclusions, and is the source of the main problems of the method. For example, if we run a repeated-measures ANOVA on these ranks, we obtain a *p*-value $= .044$ for the effect of $A$. For comparison, if we run repeated-measures ANOVA on the ranks of simple rank transformation, we obtain a *p*-value $= .46$, while if we use a log-transformation that is more appropriate for these data, we obtain a *p*-value $= .85$.

@fig-ART-explained (b-c) details the calculation of the transformation, where we highlight the following terms: (i) residuals (in yellow) that represent the unexplained error (due to individual differences in our example); (ii) main effects (in green and pink) estimated from the observed means of the individual levels of the two factors; and (iii) interaction effect estimates (in blue). Observe that the estimates of the two main effects are subtracted from the interaction term. The objective of this approach is to eliminate the influence of main effects when estimating interaction effects.

This is not the only alignment technique found in the research literature. @Sawilowsky:1990 suggests that at least for balanced designs, interactions could be removed when aligning main effects, in the same way main effects are removed when aligning interactions. This approach is also taken by @Leys:2010, who derived a common ranking for both main effects after subtracting the interaction term. We do not test these alternative alignment methods in this paper since to the best of our knowledge, they are not commonly used.

**The inverse normal transform.** 
A third transformation method that we evaluate is the rank-based inverse normal transformation (INT). INT is more than 70 years old [@VanDerWaerden:1952] and appears in several variations [@Beasley:2009;@Solomon2009]. Its general formulation is as follows:

$$ 
INT(Y) = \Phi^{-1}(\frac{rank(Y) - c}{N - 2c + 1})
$${#eq-int}

where $N$ is the total number of observations and $\Phi^{-1}$ is the standard normal quantile function, which transforms a uniform distribution of ranks into a normal distribution. Different authors have used a different parameter $c$. In our experiments, we use the *Rankit* formulation [@Bliss:1956], where $c = 0.5$, since past simulation studies [@Solomon2009] have shown that it is more accurate than alternative formulations. However, as @Beasley:2009 report, the choice of $c$ is of minor importance. For our experiments, we implement the INT method in R as follows:

```{r}
INT <- function(x){
	qnorm((rank(x) - 0.5)/length(x))
}
```

@fig-ART-explained (a) shows how this function transforms the responses for our example dataset.

**Other non-parametric rank-based methods.** There are other rank-based statistical tests that deal with interactions, the ANOVA-type statistic (ATS) [@Brunner_ATS:2001] being the most representative one. @Kaptein:2010 introduced this method to the HCI community, advocating its application in the analysis of Likert-type data as a viable alternative to parametric ANOVA. In addition to ATS, Lüpsen [-@luepsen:2017; -@luepsen:2018; -@luepsen:2023] investigated several other multifactorial nonparametric methods. In particular, the author evaluated the hybrid ART+INT technique proposed by @Mansouri:1995, which applies INT on the ranks of ART. He also tested multifactorial generalizations of the van der Waerden test [@VanDerWaerden:1952] and the Kruskal-Wallis and Friedman tests [@Kruskal-Wallis;@Friedman]. The former is based on INT, but instead of using F-tests on the transformed values as part of ANOVA, it computes $\chi^2$ ratios over sums of squares. These two methods are not widely available. Their implementation in R can be downloaded from Lüpsen's web page [@luepsen_R].

### Experimental evaluations
Previous studies have assessed ART and related procedures through various Monte Carlo experiments, yielding conflicting results and conclusions.

**Results in support of ART**. The outcomes from numerous experiments conducted in the 80s and 90s suggested that ART was robust in testing interaction effects. Noteworthy instances include studies, such as those by @Salter:1993, which compared the method to parametric ANOVA. The authors found that ART remains robust even in the presence of outliers or specific non-normal distributions, such as the logistic, exponential, and double exponential distributions. Their findings indicated only a marginal increase in error rates (ranging from 6.0% to 6.3% instead of the expected 5%) when applied to the exponential distribution. Furthermore, ART demonstrated superior statistical power compared to parametric ANOVA. @Mansouri:1995 evaluated ART under a different set of non-normal distributions (normal, uniform, log-normal, exponential, double exponential, and Cauchy) in the presence of increasing main effects. Except for the Cauchy distribution, ART maintained Type I error rates close to nominal levels across all scenarios, irrespective of the magnitude of main effects. In contrast, the error rates of the rank transformation reached very high levels (up to $100\%$) as the magnitude of main effects increased, even under the normal distribution. ART only failed under the Cauchy distribution, which is well-known to be pathological.

More recently, @elkin:2021 compared ART to parametric t-tests for testing multifactor contrasts under six distributions: normal, log-normal, exponential, double exponential, Cauchy, and Student's t-distribution ($\nu=3$). Their results confirmed that ART keeps Type I error rates close to nominal levels across all distributions, except for the Cauchy distribution. In addition, they found that ART exhibits a higher power than the t-test. We emphasize, however, that the experimental methodology of @elkin:2021 did not allow for detecting the influence of a factor's main effect on the effect of a different factor or their interaction.

While most evaluation studies have focused on continuous distributions, @Payton:2006 have also studied how various transformations (rank, ART, log-transform, and squared-root transform) perform under the Poisson distribution, focusing again on interaction effects when main effects were present. The authors found that ART and parametric ANOVA (no transformation) performed best, keeping Type I error rates close to nominal levels. All other transformations inflated error rates.

**Warnings**. While the above results indicate that ART is a robust method, other studies have identified some serious issues. The second author of this paper has observed that, in certain cases, ART seems to detect spurious effects that alternative methods fail to identify [@casiez:2022]. Such informal observations, conducted with both simulated and real datasets, motivated us to delve deeper into the existing literature.

@Carletti:2005 report that *"aligned rank transform methods are more affected by unequal variances than analysis of variance especially when sample sizes are large."* Years later, @luepsen:2018 conducted a series of Monte Carlo experiments, comparing a range of rank-based transformations, including the rank transformation, ART, INT, a combination of ART and INT (ART+INT), and ATS. His experiments focused on a $2 \times 4$ balanced between-subjects design and a $4 \times 5$ severely unbalanced design and tested normal, uniform, discrete uniform (integer responses from 1 to 5), log-normal, and exponential distributions, with equal or unequal variances. Furthermore, they tested both interaction and main effects when the magnitude of other effects increased. The results revealed that ART inflates error rates beyond acceptable levels in several configurations: right-skewed distributions (log-normal and exponential), discrete responses, unequal variances, and unbalanced designs. @luepsen:2018 also found that using INT in combination with ART (ART+INT) is preferable to the pure ART technique. However, as the method still severely inflated error rates in many settings, @luepsen:2018 concluded that both ART and ART+INT are *"not recommendable."*

Another interesting finding of @luepsen:2018 was that the simple rank transformation *"appeared not as bad as it is often described"* [@luepsen:2018], outperforming ART in certain configurations, such as discrete and skewed distributions, or unequal variances. These results are in full contradiction with the findings of @Mansouri:1995.

The same author conducted an additional series of experiments [@luepsen:2017], focusing on two discrete distributions (uniform and exponential) with a varying number of discrete levels: 2, 4, and 7 levels with integer values for the uniform distribution, and 5, 10, and 18 levels with integer values for the exponential distribution. Again, the Type I error rates of both ART and ART+INT reached very high levels, but error rates were especially pronounced when the number of discrete levels became small and the sample size increased. The author provides a detailed analysis of why ART fails in these cases, summarized as follows:

> *"There is an explanation for the increase of the type I error rate when the number of distinct values gets smaller or the sample size larger: due to the subtraction of the other effects --- a linear combination of the means --- from the observed values even tiny differences between the means lead to large differences in the ranking"* [@luepsen:2017].

Given these results, the author's conclusion was the following:

> *"the ART as well as the ART+INT cannot be applied to Likert and similar metric or ordinal scaled variables, e.g. frequencies like the number of children in a family or the number of goals, or ordinal scales with ranges from 1 to 5"* [@luepsen:2017].

**Results on other rank-based methods.** We are also interested in understanding how ART compares with INT, which is frequently used in some research domains, such as genetics research [@Beasley:2009]. @Beasley:2009 conducted an extensive evaluation of the method and reached the conclusion that *"INTs do not necessarily maintain proper control over Type 1 error rates relative to the use of untransformed data unless they are coupled with permutation testing."* @luepsen:2018 included INT in his evaluation and found that in most cases, it maintained better control of Type I error rates compared to ART and the pure rank transformation, while also presenting a higher statistical power. However, he also identified several cases where INT failed to sufficiently control for Type I errors, such as design configurations with unequal variances in unbalanced designs or skewed distributions, and when testing interactions in the presence of non-null main effects.

In addition to INT, @luepsen:2018 evaluated the ATS method but found it to suffer from low power while presenting similar challenges as the rank transformation with regard to Type I error rates. Amongst all evaluated methods, the author identified the generalized van der Waerden test as the method that provided the best overall control of Type I error rates. More recently, @luepsen:2023 conducted a new series of experiments testing rank-based nonparametric methods on split-plot designs with two factors. While the author reported on various tradeoffs of the methods, he concluded that overall, the generalized van der Waerden test and the generalized Kruskal-Wallis and Friedman tests were the best performing methods.

### The use of ART in experimental research {#ART-use}
To get a sense of how frequently ART is used in experimental research, we examined the citations of ARTool [@wobbrock:2011] indexed by Google Scholar. As shown in @fig-ARTool-citations), the rate of citations to the original CHI 2011 paper introducing the method to the HCI community is steadily increasing, with over 400 citations in 2023 alone.

![Citations of the CHI 2011 paper on ARTool [@wobbrock:2011] as shown in Google Scholar on 28 April 2024. We analyzed the 439 citing bibliographic entries of 2023 in more detail.](images/ARTool-citations-28-april-2024.png){width=80% #fig-ARTool-citations}

@fig-ART-citations-venues presents the most frequently citing venues for 2023. We observe that the HCI, Augmented Reality (AR), and Virtual Reality (VR) venues dominate these citations. However, we found that ARTool is used in other scientific disciplines. For example, among the papers published in 2023, we counted nine Nature Scientific Reports and a total of 25 articles appearing in journals with the words "neuro" or "brain" in their title, including the Journal of Neuropsychology (three citations), the Journal of Neuroscience (two citations), the Journal of Cognitive Neuroscience (two citations), Neuropsychologia (two citations), Nature Neuroscience, Neuropharmacology, Brain, and NeuroImage.

::: {#fig-ART-citations-venues}
```{r, echo=FALSE, message=FALSE, fig.height=4.5, fig.width = 8, warning=FALSE}
library(ggplot2)
library(bib2df)
library(stringr)

df <- bib2df("2023-papers/citations-ART-2024-no-theses.bib")

df$venue <- tools::toTitleCase(ifelse(!is.na(df$JOURNAL), df$JOURNAL, df$BOOKTITLE))
 venues <- df %>%
  filter(!is.na(venue)) %>%
  group_by(venue) %>%
  summarize(n = n()) %>%
  arrange(desc(n))

venues <- venues %>% slice(1:17)
venues$venue <- factor(venues$venue, levels = venues$venue[order(venues$n)])

plot <- ggplot(venues, aes(x = venue, y = n)) +
  geom_col(fill = "gray70") +
  geom_text(
    aes(label = n), 
    ## make labels left-aligned
    hjust = 1, nudge_y = -.4, 
  ) + coord_flip()  +
  theme_void() + theme(axis.title = element_blank()) + 
  theme(axis.text.y = element_text(size = 10, hjust = 1, family = "Arial"),
        plot.margin = margin(rep(15, 4)))

plot
```
Publication venues with the highest number of citations to ARTool [@wobbrock:2011] in 2023. Note that among the articles published in the IEEE Transactions on Visualization and Computer Graphics, eight articles appeared at the IEEE International Symposium on Mixed and Augmented Reality (IEEE ISMAR'23), and three articles appeared at the IEEE Conference on Virtual Reality and 3D User Interfaces (IEEE VR'23). So the total number of citing articles appearing at these two venues is 19 for IEEE ISMAR'23 and 10 for IEEE VR'23. 
:::

To gain insights into the prevalent experimental designs using ART, we examined a subset of citing papers comprising the 39 most recent English publications as of November 2023. Among these, 25 report using a within-participants design, three papers report using a between-participants design, while 11 papers report using a mixed design that involved both between- and within-participants factors. The number of factors ranges from one to five, with two factors representing approximately 64% of the experimental designs. Participant counts in these studies vary between 12 and 468, with a median of 24 participants. Within-cell sample sizes (n) range from 8 to 48, with a median of 20.

Furthermore, we explored the types of data analyzed using ART. Among the 39 papers, 21 use ART for analyzing ordinal data, often in conjunction with other data types. Ordinal data include responses to Likert scales or other scales of subjective assessment, such as the NASA Task Load Index (NASA TXL) [@hart1988development]. Furthermore, several authors apply ART to individual ordinal items, including Likert items with five levels (two papers), seven levels (seven papers), and 11 levels (three papers). Other types of data for which the method is used include task-completion or response times, counts, and measures of accuracy, such as target hit rates or recall scores. A common rationale for employing ART for such measures is the detection of departures from normality assumptions, often identified through tests like the Shapiro-Wilk test. Interestingly, several authors use ART only for ratio data, opting for nonparametric tests such as the Friedman test when analyzing ordinal data.

Out of the 39 papers examined, 38 identify at least one statistically significant main effect using ART. Additionally, 30 papers conduct tests for interactions, with 24 of them reporting at least one statistically significant interaction effect. Only five papers have publicly available data. We revisit the results of three of these papers in [Section 6](#casestudies).

<!----
with subgroups ranging from 8 to 48 (median = 20.0). For dependent variables, 51.3% of the studies used ratio variables and the remaining ordinal variables.  94.9% of the studies found at least one significant main effect in the results, using the ART. 80.0% of the studies studying the interaction between factors using the ART found at least one significant interaction. Only 12.8% of the papers made the data of their studies publicly available. --->

### Positioning
We observe that ART is frequently used for the analysis of experimental results. Interestingly, the cautions raised by Lüpsen [-@luepsen:2017; -@luepsen:2018] have been widely overlooked, and ART is commonly applied to datasets, including Likert data with five or seven levels, where the method has been identified as particularly problematic. Given the contradictory findings in the existing literature, researchers grapple with significant dilemmas regarding which past recommendations to rely on.

Our goal is to verify Lüpsen's findings by employing an alternative set of experimental configurations and a distinctly different methodology. In particular, we adopt a latent variable modeling approach that establishes a unified framework for data generation across all distributions. This approach allows us to demonstrate the shortcomings of rank-based approaches in effectively addressing interactions through the lens of removable interactions [@Loftus:1978]. Moreover, it facilitates the assessment of the methods through more suitable generation procedures for ordinal data, as suggested by @Liddell:2018.

With a focus on clarity, we chose to exclusively examine the three rank-based transformations presented earlier: the pure rank transformation (RNK), ART, and INT. While we do not elaborate on the performance of ATS in the main paper, additional experimental results are available in the [appendix](appendix.html). These findings indicate that its performance is comparable to the rank transformation but seems to be inferior to the simpler and more versatile INT. The appendix features additional results regarding the performance of the generalized van der Waerden test, as well as the generalized Kruskal-Wallis and Friedman tests. Our findings show that, at least in balanced designs, these methods suffer from low power issues without demonstrating any clear benefits compared to INT.

Finally, we chose to limit our investigation to balanced experimental designs. The rationale behind this decision is that we have not encountered any prior claims about ART's suitability for unbalanced data, and the ARTool [@artool] issues a warning in such situations. However, the [appendix](appendix.html) presents additional results on experimental designs with missing data, which confirm that in such situations, ART's accuracy may deteriorate further.

### 2025-05-24 text/50-ExperimentalMethod.qmd
updated repository from jovi

## Experimental method {#methodology}
We can now detail our experimental method. We evaluate the standard parametric approach (*PAR*) and the three rank-transformation methods (*RNK*, *INT*, and *ART*) that we introduced earlier. We conduct a series of Monte Carlo experiments that assess their performance under a variety of experimental configurations:

1. We evaluate *ratio* and *ordinal* data. For ratio data, we examine four representative *continuous* distributions (normal, log-normal, exponential, and Cauchy distribution) and two *discrete* distributions (Poisson and binomial distribution). For ordinal data, we examine distributions for 5-level, 7-level, and 11-level Likert item responses.

2. We present results for five experimental designs. To simplify our presentation, we start with a 4 $\times$ 3 repeated-measures factorial design. We then show how our conclusions generalize to four additional designs: (i) a 2 $\times$ 3 between-subjects design; (ii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a repeated-measures factor; (iii) a 2 $\times$ 2 $\times$ 2 repeated-measures design, and (iv) a 3 $\times$ 3 $\times$ 3 repeated-measures design.

3. We evaluate three sample sizes, $n=10$, $n=20$, and $n=30$, where $n$ represents the cell size in an experimental design. For within-subjects designs where all factors are treated as repeated measures, $n$ coincides with the number $N$ of subjects (or commonly human participants in HCI research). In contrast, for a 2 $\times$ 3 between-subjects design, if the cell size is $n = 20$, then the number of subjects is $N = 120$.

4. We test the robustness of the methods under unequal variances.

5. In addition to Type I error rates, we compare the statistical power of the methods and also evaluate the quality of their effect size estimates.

6. We assess the above measures for both main and interaction effects and examine how growing effects on one or two factors affect the Type I error rate on other factors and their interactions.

<!--
Furthermore, we assess how a growing interaction effect, either in isolation or in combination with a main effect, can affect the inference of main effects. -->

Previous evaluations of rank transformation methods [@Beasley:2009; @luepsen:2018] have also tested unbalanced designs, where the cell size is not constant across all levels of a factor. When combined with unequal variances, unbalanced designs are often problematic for both parametric procedures [@blanca:2018] and rank transformation methods [@Beasley:2009; @luepsen:2018]. As we mentioned earlier, we do not evaluate unbalanced designs here. However, we provide additional experimental results on missing data in the [appendix](appendix.html).

### Statistical modeling
To model the latent variable $Y$, we use a two-way (two factors) or a three-way (three factors) mixed-effects model. For simplicity, we explain here the model for two factors. Its extension to three factors is straightforward. The model has the following form:

$$ 
y_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j} + \epsilon_{ijk}
$${#eq-linear-model}

 - $\mu$ is the grand mean

 - $s_k$ is the random intercept effect of the *k*-th subject, where $k = 1..n$

 - $x_{1i}$ is a numerical encoding of the *i*-th level of factor $X_1$, where $i = 1..m_1$

 - $x_{2j}$ is a numerical encoding of the *j*-th level of factor $X_2$, where $j = 1..m_2$

 - $a_1$, $a_2$, and $a_{12}$ express the magnitude of main and interaction effects

 - $\epsilon_{ijk}$ is the experimental error effect

To encode the levels of the two factors $x_{1i} \in X_1$ and $x_{2j} \in X_2$ we proceed as follows:

1. We normalize the distance between their first and their second levels such that $x_{12} - x_{11} = 1$ and $x_{22} - x_{21} = 1$. This approach enables us to conveniently control for the main and interaction effects by simply varying the parameters $a_1$, $a_2$, and $a_{12}$.

2. For the remaining levels, we randomly sample from a uniform distribution that spans the range between these two extreme levels, i.e., between $x_{11}$ and $x_{12}$ for $X_1$, and between $x_{21}$ and $x_{22}$ for $X_2$. This approach allows us to generate and evaluate a substantial variety of configurations, each representing different relative effects between levels.

3. We require all levels to sum up to 0, or $\sum\limits_{i=1}^{m_1}  x_{1i} = 0$ and $\sum\limits_{j=1}^{m_2}  x_{2j} = 0$, which ensures that the grand mean is $\mu$.

For example, we can encode a factor with four levels as $\{-.6, .4, .1, .1\}$ or as $\{-.5, .5, .3, -.3\}$.

While random slope effects can have an impact on real experimental data [@barr:2013], we do not consider them here for two main reasons: (1) to be consistent with previous evaluations of the ART procedure [@elkin:2021]; and (2) because mixed-effects procedures with random slope effects are computationally demanding, adding strain to simulation resources. However, there is no good reason to believe that adding random slope effects would impact our findings and conclusions.

### Population control and distribution conversions 
To simplify our simulations, we fix the following population parameters: $\mu = 0$, $\sigma = 1$, and $\sigma_s = 1$. We then control the magnitude of effects by varying $a_1$, $a_2$, and, for some experiments $a_{12}$. @fig-effects presents the range of values that we test for $a_1$ and $a_2$. We also visualize their effects on the latent variable for factors with two, three, and four categorical levels.

::: {#fig-effects}
```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
source("effects_plot.R")
plotEffects()
```
Varying $a_1$ or $a_2$ to control the magnitude of main effects. The plots show examples of population distributions on the latent variable $Y$ for factors ($X_1$ or $X_2$) with two (top), three (middle), or four categorical levels (bottom).    
:::

We follow the approach of @DeBruine:2021 and use the R package *faux* v1.2.1 [@faux] to simulate data for our mixed-effects models. We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\epsilon_{ijk} \sim N(0,\sigma)$.
To simulate the observed variable $Y'$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach). A key advantage of this method is that we can generate responses for any distribution, while we control effects on the latent variable in the exact same way.

### Implementation of rank transformation methods 
For the aligned rank transformation (ART), we use the R implementation of ARTool v0.11.1 [@artool]. For the pure rank transformation (RNK), we use R's *rank()* function. We use the *Rankit* formulation [@Bliss:1956] for the inverse normal transformation (INT), as explained earlier. Unless explicitly mentioned otherwise, we use the formula ``Y' ~ X1*X2 + (1|S)`` with R's *lmer* function, except for the between-subjects design where we use the formula ``Y' ~ X1*X2 + Error(S)`` with R's *aov* function.

### Evaluation measures
Significance tests have two types of errors. *Type I errors*, or false positives, are mistaken rejections of the null hypothesis. Type II errors, of false negatives, are failures to reject a null hypothesis that is actually true. In our illustrative example in @fig-example, a Type I error is finding that there is an effect of the choice of the technique on time performance. A *Type II error*, in turn, is finding that the task difficulty has no effect on time performance.

Statistical significance testing requires setting a significance threshold known as significance or $\alpha$ (alpha) level, with typical values $\alpha = .05$ and $\alpha = .01$. The Type I error rate of a well-behaved significance test should be close to this nominal alpha level. An error rate clearly above this level suggests that the significance test is too liberal, while an error rate clearly below this level suggests that the test is too conservative. Four of our experiments specifically assess the Type I error rate of the methods. We test two significance levels: $\alpha = .05$ and $\alpha = .01$. For brevity, we only report results for $\alpha = .05$ in the main paper and include additional results in our supplementary material.

We do not directly evaluate Type II errors. Instead, we report on statistical *power* defined as $Power = 1 - \beta$, where $\beta$ is the rate of Type II errors. Significance tests do not provide any power guarantees. However, we can compare the power of different methods to evaluate their relative performance. In addition to power, we assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted on the latent variable $Y$. While partial $\eta^2$ is the most commonly used effect size measure, we chose to base our analysis on Cohen's $f$, as its expected value is proportional to the real magnitude of effect. However, $\eta^2$ can be directly derived from Cohen's $f$ as follows:

$$ 
\eta^2 = \frac{f^2}{1 + f^2}
$${#eq-cohensf}

As we explained earlier, interaction effects are deformed when we transform the latent variable $Y$ to obtain the observed responses. Such transformations also affect the Type I error rates and power that we observe. Here, we are interested in evaluating these measures with respect to the effects we apply to the latent variable $Y$. We make this distinction whenever it is relevant in our analysis.

### Hardware platform and iterations
Our experimental R code is available in our supplementary material. We ran our experiments separately in a cluster of 8 machines Dell R640 Intel Xeon Silver 4112 2.6GHz with 4 cores and 64 GB memory. Our R code was parallelized to use all four cores of each machine. Some experiments took a few hours to complete, while others took several days.

To estimate the power and Type I error rates of the four methods with enough precision, we ran $5000$ iterations for each population configuration and each sample size.

<!--
We evaluate the Type I error rate, that is the rate of false positives of each method. To evaluate the Type I error rate for the interaction effect, we set $a_{12} = 0$. Likewise, to evaluate the Type I error rate for the second factor $X_2$, we set $a_{2} = 0$. For each population configuration and sample size $n$, we run $5000$ iterations and test two significance levels: $\alpha = .05$ and $\alpha = .01$. For brevity, we only report results for $\alpha = .05$ in the main paper, while additional results are presented in supplementary materialsxa
-->

### 2025-05-24 text/60-Results-RatioScales.qmd
updated repository from jovi

### Type I error rates in ratio scales {#ratio}
Our first experiments evaluate Type I error rates for ratio scales. We test a 4 $\times$ 3 repeated-measures design, where we refer to the 4-level factor as $X_1$ and the 3-level factor as $X_2$. For the observed response variable $Y' = \mathcal{T}(Y)$, we evaluate transformations to four continuous and two discrete distributions:

1. No transformation, or $Y' = Y$. Distributions are normal with a grand mean $\mu = 0$ (e.g., see @fig-effects).

2. Log-normal distribution $LogN(\mu, \sigma)$ with global parameters $\mu = 0$ and $\sigma = 1$. As discussed in [the introduction](#intro), the log-normal distribution is a good model for various measures bounded by zero, such as task-completion times.

3. Exponential distribution $Exp(\lambda)$ with a global parameter $\lambda = 2$. The exponential distribution naturally emerges when describing the time elapsed between events. For example, we could use it to model the time a random person spends with a public display, or the waiting time before a new person approaches to interact with the display, when the average waiting time is $\frac{1}{\lambda}$.

4. Cauchy distribution $Cauchy(x_0,\gamma)$ with global parameters $x_0 = 0$ and $\gamma = 1$. The Cauchy distribution is the distribution of the ratio of two independent normally distributed random variables. It rarely emerges in practice. However, it is commonly used in statistics to test the robustness of statistical procedures because both its mean and variance are undefined. As we discussed earlier, past evaluations of ART [@Mansouri:1995; @elkin:2021] show that the method fails under the Cauchy distribution.

5. Poisson distribution $Pois(\lambda)$ with a single parameter $\lambda = 3$. It expresses the probability of a given number of events in a fixed interval of time. For example, we could use it to model the number of people who interact with a public display in an hour, when the average rate is $\lambda = 3$ people per hour.

6. Binomial distribution $B(k,p)$ with parameters $k = 10$ and $p=.1$. It frequently appears in HCI research, as it can model the number of successes and failures in a series of experimental tasks. For example, we could use it to model the number of errors that participants make in a series of $k = 10$ repetitions of a memorization task, when the average error rate is $10\%$, thus the average error probability is $p=.1$.

Specifically for the log-normal and binomial distributions, we present results for a wider range of their parameters in the [appendix](appendix.html).

**Main effects**. @fig-ratio-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases. The results show a very good behavior for RNK and INT across all distributions. The regular parametric ANOVA (PAR) keeps error rates below $5\%$. However, error rates become extremely low for some distributions, suggesting a loss in statistical power. We confirm previous results that ART fails to control the Type I error rate under the Cauchy distribution [@Mansouri:1995; @elkin:2021]. However, we also show that the method can be problematic with other non-normal distributions. As the main effect on the first factor $X_1$ increases, Type I errors on the second factor $X_2$ grow and reach high levels. This pattern is particularly pronounced under the log-normal distribution. We also observe that for the binomial distribution, error rates are high ($\approx 11\%$ for $n = 20$) even when effects on $X_1$ are zero. In addition, error rates further grow when the sample size increases.

::: {#fig-ratio-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
library(dplyr)
library(tidyverse)
source("dataReaders.R")
source("plotlying.R")

prefix <- "1_test_4x3_Ratio"
alpha <- .05
distributions <- c("norm", "lnorm", "exp", "cauchy", "poisson", "binom")
dnames <- c("Normal", "Log-normal", "Exponential", "Cauchy", "Poisson", "Binomial")
df <- readlyData(prefix, alpha, 1, distributions, dnames)
plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 64)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
:::

**Contrasts**. The same problems appear when we run ART's procedure for contrasts [@elkin:2021]. @fig-ratio-contrasts shows our results, where we report average error rates for three pairwise comparisons (since $X_2$ has three levels). In the rest of the paper, we only show results for overall effects, since results for contrasts exhibit the same patterns.

::: {#fig-ratio-contrasts}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "1_test_4x3_Contrasts"

df <- readlyData(prefix, alpha, 1, distributions, dnames)
plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 52)
```
Average Type I error rates ($\alpha = .05$) for **contrasts on factor $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ 
:::

**Interaction effects**. @fig-ratio-interaction-1 presents Type I error rates for the interaction effect $X_1 \times X_2$, when the main effect on $X_2$ is zero while the main effect on $X_1$ increases. Overall, we observe the same trends as for main effects. Again, ART fails in similar ways, although its error rates are now slightly lower.

::: {#fig-ratio-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "1_test_4x3_Ratio"
df <- readlyData(prefix, alpha, 1, distributions, dnames)

plotlyError(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 52)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
:::

Finally, @fig-ratio-interaction-2 presents our results when the main effects on the two factors increase in parallel. Error rates become exceptionally high in some cases, growing up to $100\%$.

::: {#fig-ratio-interaction-2}
```{r, echo=FALSE, message=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readlyData(prefix, alpha, 0, distributions, dnames)

#xlab <- TeX("\\text{main effect of factors }X_1\\text{ and }X_2\\text{ }(a_1 = a_2)")
plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ 
:::

However, these results require special attention since interaction interpretation issues come now into place. Let us examine the performance of each method:

- *PAR*. Error rates are high for all non-normal distributions, and become even higher when the sample size increases. As we explained in [Section 3](#interpretation), the parametric method cannot deal with scale differences that emerge when effects are transformed in a non-linear fashion. If we interpreted interactions based on absolute mean differences, disregarding their different scales (i.e., if we decided that the trend in @fig-interactions (left) is indeed a valid interaction effect), we would not observe these high error rates.

- *RNK.* Error rates explode when both main effects exceed a certain level (e.g., when $a_1, a_2 \ge 4$ and $n = 20$). As shown in @fig-interactions-rank, the problem is due to the way the rank transformation deforms interactions.

- *INT.* The method exhibits a better behavior than RNK. Errors start again increasing for all distributions, but not until main effects become significantly large. An exception is the binomial distribution, for which the error rates of INT and RNK are similar. We also observe that for continuous distributions, both INT and RNK are scale invariant, since their performance is not affected by the choice of scale.

- *ART.* It keeps error rates at correct levels as long as population distributions are normal. For all other distributions, its error rates grow rapidly as effect sizes increase. We distinguish between two sources of errors: (i) lack of statistical robustness as observed in our previous tests; and (ii) interaction interpretation issues (see [Section 3](#interpretation)). Our results confirm that ART is not scale invariant. On the contrary, it is extremely sensitive to the scale of the observed data.

### 2025-05-24 text/61-Results-OrdinalScales.qmd
updated repository from jovi

### Type I error rates in ordinal scales {#ordinal}
Our second experiment evaluates Type I error rates for ordinal scales. We test again a $4 \times 3$ repeated-measures design. We focus on individual Likert items levels and implement an ordered-probit method [@Liddell:2018] to discretize the latent variable $Y$ into 5, 7, or 11 ordinal levels. To derive the discretization thresholds, we first consider the range $[-2SD, 2SD]$ , where $SD$ is the overall standard deviation of the responses $y_{ijk}$. We then divide this range into 5, 7, or 11 intervals, following two different strategies: (i) setting thresholds to be *equidistant*; or (ii) considering *flexible* thresholds, randomly drawing their position in the above range. @fig-ordinal presents examples of equidistant and flexible thresholds for a 5-level scale when the magnitude of main effect of $X_1$ is either $a_1 = 2$ or $a_1 = 8$, while all other effects are zero.

::: {#fig-ordinal}
```{r, echo=FALSE, message=FALSE, fig.height=2.5, fig.width = 9, warning=FALSE}
plotThresholds()
```
The four vertical lines in each plot represent thresholds defining 5-level ordinal scales. Thresholds are either equidistant (top) or flexible (bottom) within a range of $\pm 2$ standard deviations around the grand mean $\mu = 0$.  
:::

**Main effects**. @fig-ordinal-main present Type I errors for $X_2$'s main effect, as we vary $X_1$'s magnitude of main effect. Our results indicate that PAR, RNK, and INT consistently maintain error rates close to $5\%$ across all tested ordinal scales. In contrast, ART exhibits a significant inflation of error rates, although this issue becomes less severe when the number of levels increases. Notably, error rates are more pronounced for flexible thresholds and tend to increase with sample size.

::: {#fig-ordinal-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "2_test_4x3_Ordinal"

distributions <- c("likert5", "likert5B", "likert7", "likert7B", "likert11", "likert11B")
dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
df <- readlyData(prefix, alpha, 1, distributions, dnames)

plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 52)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
:::

**Interaction effects**. @fig-ordinal-interaction-1 displays error rates for the interaction $X_1 \times X_2$ with a single main effect applied to $X_1$. The observed patterns align with our previous findings; ART consistently inflates error rates, although to a lesser extent now.

::: {#fig-ordinal-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}

plotlyError(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 45)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
:::

Finally, @fig-ordinal-interaction-2 presents our results when main effects are applied to both factors. None of the methods successfully maintains low error rates across all conditions. Intriguingly, we note that ART consistently performs worse than PAR. INT performs worse than ART and PAR in one specific case ($a_2, a_2 = 8$ in a scale with 11 equidistant levels), but overall, it exhibits a better behavior than all other methods.

::: {#fig-ordinal-interaction-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readlyData(prefix, alpha, 0, distributions, dnames)
plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)

```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ 
:::

### 2025-05-24 text/65-Results-EffectSize.qmd
updated repository from jovi

### Effect size estimates {#effect-sizes}
Our final experiment assesses the precision of the effect size estimates (Cohen's $f$) of each method.

**Main effects**. We first evaluate estimates of the main effect of $X_1$ as the magnitude of the two main effects varies. @fig-effect-main-scatterplots-1 illustrates the relationship between the Cohen's $f$ obtained with each method and the $f$ of the ground-truth method, when $a_1$ and $a_2$ vary within the range $[-2, 2]$. The proximity of data points to the black diagonal indicates the closeness of estimates to the ground truth. Points above the line are likely to overestimate the effect size, while points under the line are likely to underestimate it.

::: {#fig-effect-main-scatterplots-1}
```{r, echo=FALSE, fig.height=3.4, message=FALSE, warning=FALSE}
source("effect_sizes_scatterplots.R")
prefix <- "6_scatter-Effect_Size-g2"
df <- readlyDataPoints(prefix,  distributions=c("norm", "lnorm", "likert5B"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)")) %>% toCohensf()
fig <- plotlyScatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "fX1", max = 1.05)

fig
```
Scatterplots showing the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground-truth method for the main effect of $X_1$, when $a_1$ and $a_2$ vary within the range $[-2, 2]$. 
:::

We observe that PAR tends to underestimate effect sizes under log-normal distributions, in consistency with its low power (see @fig-power-main-1). ART's estimates are instead spread across both sides. We invite the reader to zoom in on the lower range of values ($f < 0.4$), where the method frequently exaggerates effects. INT demonstrates the highest precision, followed by RNK. However, their performance deteriorates under the ordinal scale. Still, ART performs worse than all other methods.

@fig-effect-main-scatterplots-2 shows the same relationship but for a larger range of effects $[-8, 8]$. We observe that all three rank-based methods struggle to detect large effects, even when the underlying distribution is normal. Interestingly, INT's performance declines when Cohen's $f$ exceeds a certain level. For the ordinal distribution, estimates are extremely noisy across all methods. We emphasize, however, that this is at least partly due to the low precision of the ordinal scale itself, since its five discrete levels cannot capture the full range of variations of the latent variable. The fact that all methods underestimate large effects can be simply explained by the limited range of the ordinal scale's extremes.

::: {#fig-effect-main-scatterplots-2}
```{r, echo=FALSE, fig.height=3.4, message=FALSE, warning=FALSE}
#source("effect_sizes_scatterplots.R")
prefix <- "6_scatter-Effect_Size-g8"

df <- readlyDataPoints(prefix,  distributions=c("norm", "lnorm", "likert5B"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)")) %>% toCohensf()
fig <- plotlyScatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "fX1", max = 4.05)

fig
```
Scatterplots showing the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground-truth method for the main effect of $X_1$, when $a_1$ and $a_2$ vary within the range $[-8, 8]$. 
:::

To systematically evaluate the precision of effect size estimates across different methods, we employ the coefficient of determination, denoted as $R^2$ ($R$ squared). This metric quantifies the proportion of variation in the $f$ estimates of each method that is predictable from the estimate provided by the reference method, considered as the ground truth. We compute $R^2$ over 1000 iterations. Within each iteration, we randomly sample the magnitude of the two main effects ($a_1$ and $a_2$) from a range spanning $[-g, g]$, where $g$ takes values from the set ${0, 0.5, 1, 2, 4, 8}$. It is worth noting that the $R^2$ measure assesses the proximity of estimates to the ground truth, without distinguishing between underestimates and overestimates. We present these results based on a sample size of $n=20$.

@fig-effect-main-1 depicts trends for the first factor $X_1$. INT performs exceptionally well across all continuous distributions, but its precision declines when the range of effects becomes large ($g > 4$). PAR's performance is strikingly low under the log-normal distribution and, to a lesser degree, under the exponential distribution. ART emerges as the second-worst method for these distributions. Finally, we observe low $R^2$ values for all methods under the binomial and ordinal distributions. ART's performance is particularly problematic when effects sizes are low.

::: {#fig-effect-main-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "6_test-Effect-Size-Main"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha= NA, effectType = -1, distributions)

df <- df %>% arrange(design,distr,effectX1,cohensfX1)  %>% group_by(design,distr,effectX1)
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"), groupvars = c("distr","method","n"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "range of effects (g)", var = "cohensfX1", xvar = "effectX1", min = -11, max = 103, ytitle = 'R squared')
```
$R^2$ scores measuring the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground-truth method for the main effect on $X_1$. We express these scores as a function of the range of the magnitude of effects $g$.
:::

**Interaction effects**. We extend our analysis to examine trends regarding the interaction effect, as depicted in Figure @fig-effect-interaction. We now also vary $a_{12}$ within the same range $[-g, g]$. $R^2$ approaches zero or becomes negative under certain conditions, indicating that effect size estimates cannot be trusted in these cases. RNK and INT emerge as the most precise methods under non-normal distributions, but their performance declines rapidly when $g > 2$. Finally, PAR outperforms ART across all discrete distributions, as well as the normal distribution.

::: {#fig-effect-interaction}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "6_test-Effect-Size"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha= NA, effectType = -1, distributions)

df <- df %>% arrange(design,distr,effectX1X2,cohensfX1X2)  %>% group_by(design,distr,effectX1X2)
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"), groupvars = c("distr","method","n"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "range of effects (g)", var = "cohensfX1X2", xvar = "effectX1X2", min = -51, max = 105, ytitle = 'R squared')
```
$R^2$ scores measuring the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground truth method for the interaction effect $X_1 \times X_2$. We express these scores as a function of the range of the magnitude of effects $g$.
:::

### 2025-05-24 text/70-CaseStudies.qmd
updated repository from jovi

## Case studies {#casestudies}
In this section, we examine some specific types of data through concrete examples. We focus on recent work (published in 2023) that has used ART and made data publicly available. We identify problems and evaluate possible risks. We also present alternative methods for conducting the analysis and compare their outcomes.

### Binary responses

@Martin:2023 ran two experiments to investigate the psychological phenomenon of change blindness, that is, how humans fail to detect certain changes in a scene. We focus here on their first experiment, which evaluated how 22 participants succeeded in observing object manipulations in an immersive environment. The authors studied the influence of four different factors, namely the Type of manipulation (4 levels), its Distance from the observer (2 levels), the Complexity of the manipulated item (2 levels), and the field of view (not examined here). To analyze the effect of the first three factors, the authors used a $4 \times 2 \times 2$ repeated-measures ANOVA using ART.

Their design has two particularities. First, the response variable is binary (Detected vs. Not Detected). Second, the design is unbalanced; each participant did not complete the same tasks, so their types and complexities did not appear with the same frequency among participants. The authors pointed to these issues to justify the use of ART:

> *"Since we are dealing with **nonparametric, unbalanced data** and multiple factors, we apply the Aligned Rank Transform (ART) for nonparametric factorial ANOVA"* [@Martin:2023].

In their ART model, the authors treated the participant identifier and the task identifier as random effects. We could re-run their analysis to replicate the results presented in Table 1 [@Martin:2023]. However, we extended the analysis by fitting a linear mixed-effects model (LMER function) without any transformation (PAR). When responses are binary, RNK and INT do not alter the data, leading to results that coincide with those of PAR.

Below, we present the *p*-values obtained from each method:

|  | ART | PAR |
|------|-----|-----|
| Type | $.92$ | $.016$ |
| Distance | $.031$ | $.12$ |
| Complexity | $.023$ | $.0027$ |
| Type $\times$ Distance | $.71$ | $.13$ |
| Type $\times$ Complexity | $.00077$ | $.0039$ |
| Distance $\times$ Complexity | $.0011$ | $.0029$ | 
| Type $\times$ Distance $\times$ Complexity | $.00055$ | $.00016$ |
: *p*-values keeping two significant digits - Type III hypothesis tests {.sm}

There are notable differences between the results obtained from the two methods, but it is important to note that the aforementioned *p*-values are outcomes of Type III hypothesis tests. While these tests are deemed more suitable for unbalanced data when interaction effects emerge, their use remains a subject of debate [@Smith:2022], and some experts argue that interpreting main effects in the presence of interactions may not be meaningful [@Hector:2010]. Thus, the presence of interactions in the above unbalanced design complexifies the detection and interpretation of main effects. In this specific case, the authors decided to create partial models with two factors each time to further evaluate contrasts, and interestingly, they now used t-tests instead of ART.

We emphasize that the ARTool raises a warning for such unbalanced designs:

> *F values of ANOVAs on aligned responses not of interest are not all ~0. ART may not be appropriate.*

Our tests in the [appendix](appendix.html) show that ART's Type I error rates for main effects increase further in unbalanced designs with missing data.

**Monte Carlo simulation.** What if the experimental design were balanced, and such issues did not emerge? What risks would arise from using ART with such binary responses? To delve into this issue, we conduct a new Monte Carlo experiment.

Like @Martin:2023, we investigate a $4 \times 2 \times 2$ repeated-measures design and set the number of participants to $n=22$. However, we now test a perfectly balanced design. We transform the latent variable to a Bernoulli distribution with a success probability parameter $p=.46$ ---average success rate found by @Martin:2023--- and then assess the Type I error rate of ART in the absence of any effect. We also assess the Type I error rate of regular ANOVA (LMER model) with no transformation (PAR). Our results over 5000 iterations are as follows:

|  | $X_1$ | $X_2$ | $X_3$ | $X_1 \times X_2$ | $X_1 \times X_3$ | $X_2 \times X_3$ | $X_1 \times X_2 \times X_3$ |
|------:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|
| PAR/LMER | $4.4$ | $4.5$ | $4.6$ | $5.5$ | $5.1$ | $5.2$ | $4.9$ |
| ART | $22.4$ | $19.9$ | $20.2$ | $22.6$ | $21.8$ | $19.8$ | $21.1$ |
: Type I error rates (%) for $\alpha = .05$ (nominal error rate $=5\%$) and $p=.46$ {.sm}

|  | $X_1$ | $X_2$ | $X_3$ | $X_1 \times X_2$ | $X_1 \times X_3$ | $X_2 \times X_3$ | $X_1 \times X_2 \times X_3$ |
|------:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|
| PAR/LMER | $0.7$ | $1.1$ | $0.9$ | $1.1$ | $1.0$ | $1.1$ | $1.0$ |
| ART | $10.0$ | $8.8$ | $9.7$ | $10.6$ | $9.7$ | $9.6$ | $9.9$ |
: Type I error rates (%) for $\alpha = .01$ (nominal error rate $=1\%$) and $p=.46$ {.sm}

We observe that ART exhibits significant challenges with such binary data, as its error rates escalate to very high levels for all main and interaction effects. In contrast, PAR (LMER) consistently maintains error rates close to nominal levels. We extend our simulation to Bernoulli distributions with a low success (or error) probability parameter $p=.06$. The results are now as follows:

|  | $X_1$ | $X_2$ | $X_3$ | $X_1 \times X_2$ | $X_1 \times X_3$ | $X_2 \times X_3$ | $X_1 \times X_2 \times X_3$ |
|------:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|
| PAR/LMER | $4.9$ | $4.8$ | $6.0$ | $4.4$ | $4.0$ | $4.6$ | $4.6$ |
| ART | $78.1$ | $68.6$ | $68.7$ | $71.9$ | $71.3$ | $68.8$ | $70.5$ |
: Type I error rates (%) for $\alpha = .05$ (nominal error rate $=5\%$) and $p=.06$ {.sm}

|  | $X_1$ | $X_2$ | $X_3$ | $X_1 \times X_2$ | $X_1 \times X_3$ | $X_2 \times X_3$ | $X_1 \times X_2 \times X_3$ |
|------:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|
| PAR/LMER | $1.0$ | $1.0$ | $1.2$ | $0.8$ | $0.7$ | $1.1$ | $0.9$ |
| ART | $68.6$ | $60.6$ | $60.2$ | $61.6$ | $60.1$ | $60.6$ | $59.9$ |
: Type I error rates (%) for $\alpha = .01$ (nominal error rate $=1\%$) and $p=.06$ {.sm}

Clearly, ART is completely inappropriate when responses are binary. In contrast, parametric ANOVA demonstrates an excellent behavior even when the success (or error) rate is as low as $6\%$.

A more adapted approach for this type of data would be to use a generalized linear model (GLMER) with a *binomial link function*, which essentially performs a logistic regression. Such models are in line with the latent variable approach, where the link function establishes a connection between a normally distributed latent psychological variable (e.g., the human ability to detect changes in a virtual environment) and the observed variable (e.g., successes or failures in sequential change detection tests). Consequently, we expect them to exhibit stronger predictive power than regular linear models and support a more appropriate interpretation of interaction effects. However, inference with such models requires additional steps, such as comparing alternative models using various criteria [@Bolker2009glm], so we do not include them in our evaluation here.

### Likert items

@Rosso:2023 conducted a body-swap experiment, where pairs participants ("dyads") performed a joint-finger tapping task under four different conditions, structured under two different factors:

- *Coupling*. Participants saw either their own hand (*Coupled*) or their partner's hand (*Uncoupled*).

- *Perspective*. Participants saw the hand from a first person (*1P*) perspective or from a second person perspective (*2P*).

Overall, the experiment was structured as a $2 \times 2$ within-participants design with 38 participants (19 dyads). The authors investigated numerous measures. Here, we only revisit their analysis of the sense of ownership of the visually perceived hand, for which they used ART (Page 10). The variable was measured through a self-reported scale with five levels (1 to 5). The authors report:

> *Aligned rank transform (ART) ANOVA revealed significant main effects of Coupling (Df residual = 147, F = 104.353, p < 0.001) and Perspective (Df residual = 147, F = 8.983, p < 0.01) on the self-reported ownership ratings. The former indicates that participants were capable of telling apart their own hand from the partner’s regardless of the visual perspective, whilst the latter indicates that perceiving a hand in 1st person generally resulted in a stronger sense of ownership. Crucially, the interaction effect between Coupling and Perspective (Df residual = 147, F = 5.232, p < 0.05) revealed that the increase in ownership relative to the 2nd person perspective was significantly stronger when participants were coupled.* [@Rosso:2023]

Thus, the authors found supportive statistical evidence for both main effects and the interaction of the two factors. We replicated their analysis and found that the author did not conduct a repeated-measures ANOVA, i.e., they did not treat ratings from the same participant as independent. We re-analyzed the data with a mixed-effects *lmer* model, treating the participant identifier as a random effect nested under the identifier of participant pairs (dyads). Conclusions do not change since our analysis is more powerful, so $p$-values are now even smaller for all three effects. However, we also conducted the analysis using the three other methods (PAR, RNK, INT), and unfortunately, conclusions are very different:

|  | ART | PAR | RNK | INT |
|------|-----|-----|-----|-----|
| Coupling | $3.2 \times 10^{-19}$ | $4.9 \times 10^{-18}$ | $9.5 \times 10^{-20}$ | $5.4 \times 10^{-19}$ |
| Perspective | $.0016$ | $.20$ | $.25$ | $.29$ |
| Coupling $\times$ Perspective | $.015$ | $.78$ | $.90$  | $.95$ |
: *p*-values keeping two significant digits {.sm}

All four methods agree that there is strong evidence about the effect of Coupling. Nevertheless, the results of PAR, RNK, and INT for the effects of Perspective and Coupling $\times$ Perspective do not support the authors' conclusions. The discrepancy between the $p$-values returned by these methods and the $p$-values of ART is striking.

One may notice that a single entry (out of 152 entries) in the above dataset is missing, so the design is slightly unbalanced, and the ARTool again outputs a warning. Could this imbalance explain the above differences? The answer is negative. If we complete the missing row with a neutral value (e.g., a 3 in the 5-level ordinal scale) to dismiss the warning, the resulting *p*-values will still be very similar to the ones reported above.

**Ordered probit models.** We conducted complementary analyses with ordered probit models, where we used two different methods: (1) ordinal regression as implemented in the *ordinal* R package [@Christensen2023ordinal]; and (2) a Bayesian statistics framework with the *brms* R package [@Burkner:2019].

For the first (frequentist) analysis, we use the following R code:

```{r, eval=FALSE}
library(ordinal)
# We make sure that the response variable is treated as ordinal
ownership_data$ownership <- factor(ownership_data$ownership, ordered = TRUE)

fit_frequentist <- clmm(
  ownership ~ Coupling + Perspective + Coupling:Perspective + (1|Dyad/Participant),
  link = "probit", 
  threshold = "flexible",
  data = ownership_data
)
```

We can build a simpler regression model by setting ```threshold = "equidistant"``` and compare the two models by conducting a likelihood-ratio test. For this specific example, the flexible thresholds do not improve the model's predictive power, so the simpler model is more promising. However, both models lead to almost identical parameter estimates.

For the Bayesian formulation, we use the following code in R:

```{r, eval=FALSE}
library(brms)
fit_ownership <- brm(
  formula = ownership ~ Coupling + Perspective + Coupling:Perspective 
    + (1|Dyad/Participant),
  data = ownership_data,
  family = cumulative("probit")
)
```

Notice that we do not use informative priors in the above model, although this is possible. For an extensive discussion about why using a Bayesian framework for this type of analysis, we refer interested readers to @Liddell:2018. @Burkner:2019 provide an excellent introductory tutorial to this approach. Its main shortcoming is the substantial computational resources it demands, whereas fitting a model with the frequentist approach is extremely fast.

Below, we present parameter estimates from the two methods, where brackets represent either $95\%$ confidence intervals (frequentist method), or $95\%$ credible intervals (Bayesian method):

| | Frequentist [@Christensen2023ordinal] | Bayesian [@Burkner:2019] |
|-----|-----|------|
| Coupling (Coupled $-$ Uncoupled) | $-1.85$ $[-2.46, -1.24]$ | $-1.95$ $[-2.58, -1.35]$  |
| Perspective (1P $-$ 2P) |  $\quad 0.29$ $[-0.37, 0.94]$  | $\quad 0.28$ $[-0.39, 0.95]$ |
| Coupling $\times$ Perspective | $-0.11$ $[-0.93, 0.70]$| $-0.09$ $[-0.89, 0.75]$ |
: Effect estimates and their $95\%$ confidence or credible intervals. They are expressed as standard deviations of the latent variable {.sm}

Interpreting these results requires special attention, as the above estimates do not refer to the observed ordinal scale of measurements. All effects are expressed in standard deviation units over the latent variable, so they are analogous to Cohen's $d$ standardized effect sizes. The above results suggest that participants' sense of ownership was $1.85$ (or $1.95$) standard deviations lower in the Coupled condition, where standard deviations refer to the continuous latent scale of sense of ownership.

The whole $95\%$ CI for both methods is far below zero, which means that there is overwhelming evidence that this effect is strong. In contrast, we observe that the intervals for Perspective and the interaction term extend from negative to positive values. Thus, there is no sufficient statistical evidence for these effects. If we subsequently conduct a likelihood-ratio test with alternative models, the simpler model with ```formula = ownership ~ Coupling + (1|Dyad/Participant)``` will be the winning one.

These results are consistent with the results of our *lmer* models using PAR, RNK, and INT. In conclusion, using ART to analyze these data is problematic. Here, the presence of a strong effect on the first factor seems to make ART sensitive in detecting other non-existent effects (or perhaps in amplifying tiny effects).

### Likert scales

Other authors have used ART to analyze Likert scales that group several items together. For example, @Karolus:2023 investigate different methods (referred to as "gamification types") for communicating feedback on users' task proficiency, such as during a writing task. The authors used a $3 \times 2$ between-participants design to structure the study, testing the following two factors:

- *Gamification type*. Each participant was presented with one of the following elements providing feedback: (i) a *progress bar* (ii) an *Emoji*, or (iii) *none*.

- *Feedback type*. Feedback was either *continuous* or *revision*-based (provided at the end of the task to help participants revise).

The authors collected responses from 147 participants through the Amazon Mechanical Turk Service. They analyzed a range of measures, but here we focus on their analysis of participants' responses to the Situational Motivation Scale (SIMS) [@Blanchard_SIMS:2000]. The scale consists of four subscales (intrinsic motivation, identified regulation, external regulation, and amotivation), where each contains four 7-level Likert items. For each subscale, the authors conducted an ANOVA using ART on the average score. Below, we present the results of the same analysis with all four methods for intrinsic motivation and identified regulation:

|  | ART | PAR | RNK | INT |
|------|-----|-----|-----|-----|
| Gamification | $.0057$ $(0.07)$| $.010$ $(0.06)$| $.0058$ $(0.07)$| $.0062$ $(0.07)$|
| Feedback | $.77$ | $.59$ | $.56$ | $.55$ |
| Gamification $\times$ Feedback | $.94$ | $.69$ | $.94$  | $.86$ |
: Intrinsic motivation: *p*-values keeping two significant digits (partial $\eta^2$ in parentheses when $p < .05$) {.sm}

|  | ART | PAR | RNK | INT |
|------|-----|-----|-----|-----|
| Gamification | $.019$ $(0.05)$ | $.011$ $(0.06)$ | $.013$ $(0.06)$ | $.0085$ $(0.07)$|
| Feedback | $.55$ | $.74$ | $.42$ | $.55$ |
| Gamification $\times$ Feedback | $.99$ | $.86$ | $.98$  | $.93$ |
: Identified regulation: *p*-values keeping two significant digits (partial $\eta^2$ in parentheses when $p < .05$) {.sm}

Overall, differences between the results of the four methods are reasonably small and support the authors' conclusions. We omit results for external regulation and amotivation for which all *p*-values are greater than $.05$ --- again, results are consistent among the four methods.

In many situations, the above methods will behave correctly and produce similar results. However, it is difficult to know in advance when problems will arise. In a different experiment, @Siestrup:2023 investigated how different types of modifications in video sequences describing a story affect people's episodic memory. The experiment included a rating task, asking participants to rate how much the storyline of modified episodes deviated from an original version (from $1 = 0\%$ different to $6 = 100\%$ different). The authors used ART to analyze aggregated scores from multiple ratings per condition. Below, we present the results of our re-analysis with the four different methods:

|  | ART | PAR | RNK | INT |
|------|-----|-----|-----|-----|
| Version | $4.5 \times 10^{-20}$ | $2.7 \times 10^{-30}$ | $4.7 \times 10^{-20}$ | $2.0 \times 10^{-17}$ |
| Modification | $.034$ | $.11$ | $.23$ | $.40$ |
| Version $\times$ Modification | $.19$ | $.17$ | $.18$  | $.058$ |
: *p*-values keeping two significant digits {.sm}

We see now that results from different methods can lead to different conclusions. In particular, the authors' statement that *"modified videos that had already been presented during the fMRI session received lower story-change ratings than those that had been presented in the original version before"* [@Siestrup:2023], based on the observed effect of the second factor (Modification), is only supported by their analysis with ART. Effect size estimates for this factor also vary among methods, although their wide confidence intervals indicate that those estimates are highly uncertain:

|  | ART | PAR | RNK | INT |
|------|-----|-----|-----|-----|
| partial $\eta^2$| $0.12$ $[.01, 1.0]$ | $0.07$ $[.00, 1.0]$| $0.04$ $[.00, 1.0]$| $0.02$ $[.00, 1.0]$|
: effect size estimates for Modification and their $95\%$ confidence intervals {.sm}

**Ordered probit models.** Alternatively, we can use ordered probit models [@Burkner:2019;@Christensen2023ordinal], as we did for the analysis of individual Likert items. To this end, we do not average the individual Likert items of the scale but treat instead items and participants as a random effects. For example, to analyze participants' responses for intrinsic motivation [@Karolus:2023], we can use the following formula:

```formula = intrinsic_motivation ~ 1 + Gamification:Feedback + (1|Item) + (1|Participant)```

Interested readers can refer to our supplementary material for additional details.

### 2025-05-24 text/80-Recommandations.qmd
updated repository from jovi

## Recommendations {#recommendations}
Reflecting on our findings, we formulate a set of recommendations that we hope will be helpful to researchers.

### Abandoning ART
Our results corroborate Lüpsen's [-@luepsen:2017; -@luepsen:2018] warnings, overwhelmingly demonstrating that the ART method is defective. ART's advantages over the simple rank transformation (RNK) are marginal and only applicable to normal distributions with equal variances. However, the method introduces additional risks for a range of non-normal distributions. Specifically, it inflates the Type I error rate of both main effects and interactions. Furthermore, the risks associated with ART are more severe compared to those associated with parametric ANOVA. While skewed distributions lead to a loss of statistical power with parametric ANOVA, ART consistently inflates Type I error rates. It is also important to emphasize that parametric ANOVA is consistently superior to ART for the analysis of ordinal data, irrespective of whether the thresholds of the ordinal scale are equidistant or not. In light of these findings, we conclude that ART is an outdated procedure and urge researchers to discontinue its use.

Some researchers may be tempted to seek alternative formulations of ART. In particular, @Salter:1993 report that employing ART with median-based alignment (i.e., using medians instead of means to align values) for interactions corrects the method's instability under the Cauchy distribution. We present results regarding the method's performance for the $4 \times 3$ repeated-measures design in our [appendix](appendix.html). We observe that indeed, using a median-based alignment for interactions brings error rates close to nominal levels, provided that main effects remain below a certain threshold. This trend holds across all distributions tested. However, despite this improvement, the method still lags behind INT in performance, and compared to parametric ANOVA, its advantages are evident only for the Cauchy distribution. Additionally, median-based alignment proves inappropriate for main effects, as in most cases, it yields even higher Type I error rates than mean-based alignment.

### Adopting other nonparametric methods
Given the widespread adoption of ART, many researchers may feel perplexed about which alternative nonparametric method to use. Our results indicate that INT performs well across various configurations and emerges as the top performer among rank transformations. However, like other rank-based methods, INT faces challenges in accurately inferring interactions when parallel main effects are present. It may also lead to inflated error rates or reduced power in cases of unequal variances, and tends to underestimate large effect sizes. These issues are exacerbated when dealing with discrete distributions. @Beasley:2009 identify additional situations where INT may fail, such as in highly imbalanced designs.

Despite these limitations, INT outperforms parametric ANOVA across various scenarios that we investigated. 
This is especially evident when distributions are skewed. In these cases, parametric ANOVA experiences a significant loss of power and becomes sensitive to scale interpretation issues, which typically emerge when testing interactions or when variances are unequal. 
Finally, while it is still feasible to use common one-factor nonparametric tests (e.g., the Kruskal–Wallis test, the Wilcoxon sign-rank test, and the Friedman test) independently on each factor of a multifactorial design, this approach necessitates data aggregation, which compromises the statistical power of these tests. 
Nevertheless, we found that even for one-factor designs, INT turns out to be more powerful than these nonparametric tests.

Consequently, our recommendation aligns with Gelman's [-@gelman2015int_vs_wilcoxon] advice to use INT as a general nonparametric method, with careful consideration of potential cases of failure and a cautious interpretation of interaction effects.

### Priviledging parametric methods
The scope of rank-based nonparametric statistics is very limited. @baguley2012 summarizes a fundamental concern regarding their use as follows:

> *"rank transformations dont's just transform the data, they do so in a way that makes it impossible to recover the original scores ... This won't matter if the sole aim is to construct a significance test, but if you want to make predictions or obtains CIs, use of ranks introduces a potentially unnecessary obstacle."* [@baguley2012, pp. 356]

We also explained that nonparametric tests are not free of assumptions, while rank transformations complicate the interpretation of the null hypothesis. Therefore, there are valid reasons to prioritize parametric methods and reserve rank transformations as last-resort solutions. Merely obtaining a $p$-value $< .05$ from a normality test (or a test of the homogeneity of variance assumption) is insufficient justification for switching to a nonparametric procedure. According to @baguley2012, using a significance test to evaluate the statistical assumptions *"is asking completely the wrong question"* because *"It is not the presence of a violation, but the degree of departure from the assumption (i.e., effect size) that matters"* [@baguley2012, pp. 324].

Several papers we reviewed, citing ARTool [@wobbrock:2011], justify their use of ART based on the presence of *"nonparametric data."* 
However, it is important to clarify that terms like "parametric" or "nonparametric" refer to properties of the statistical model, not inherent characteristics of the collected data.
Even when ANOVA is clearly unsuitable, there are often alternative parametric methods for addressing violations of common statistical assumptions and still ensure that results can be easily communicated and interpreted. 
For example, one can log-transform positively skewed data following log-normal distributions or use generalized mixed-effects models, which provide support for a broader set of data distributions. 
Finally, employing simulations, as we did in this article, is an excellent method for evaluating the impact and risks of assumption violations and determine which data analysis procedures to use.

### Analyzing Likert-type data
The analysis of Likert-type data requires special attention due to the contradicting research literature on this topic. 20 years ago, @Jamieson:2004likert_abuse alarmed researchers regarding the frequent use of analyses that treat Likert scales as interval rather than ordinal data. Jamieson's article provoked a number of vivid responses. @Carifio:2008 replied that *"the root of many of the logical problems with the ordinalist position"* is the fact that it makes no distinction between *"a Likert response format, a Likert (graded valence) question (or stem) or a Likert scale (collection of items)."* Based on *"a wide array of additional supporting empirical evidence"*, the authors concluded that:

> *"It is, therefore, as the intervalists contend, perfectly appropriate to summarise the ratings generated from Likert scales using means and standard deviations, and it is perfectly appropriate to use parametric techniques like Analysis of Variance to analyse Likert scales"* [@Carifio:2008].

However, @Carifio:2008 also stated that analyzing individual Likert items *"is a practice that should only occur very rarely."* This statement was criticized by @Norman:2010, who argued that it does not convincingly refute the *"ordinalist"* position. According to the author, since ANOVA and other parametric statistics are robust enough to deal with Likert data, even with small sizes and unequal variances, there is *"no fear of coming to the wrong conclusion."* In the same line, @Harper:2015 came up with a number of recommendations about these issues, including the following two: (1) *"Individual rating items with numerical response formats at least five categories in length may generally be treated as continuous data;"* and (2) *"Consider nonparametric or categorical data analysis approaches for individual rating items with numerical response formats containing four or fewer categories."* However, @Liddell:2018 identified several situations in which standard parametric methods fail, even when applied to averages of multiple items in Likert scales. In particular, the authors demonstrated how they can deform interaction effects or inflate errors when assumptions of equal variances are not met.

Our experimental results on individual Likert items with 5, 7, and 11 levels show that parametric ANOVA is robust for the analysis of main effects, as long as variances are equal. On the other hand, we observed that parametric ANOVA method often fails to correctly infer interactions when main effects become large, and the problem becomes more apparent when ordinal thresholds are not equidistant. INT appears to reduce the impact of this problem but without offering a satisfying solution. We also found that at least for 5-level items with flexible thresholds, all methods may lead to unprecise effect size estimates. In conclusion, while ANOVA seems to be a valid method for hypothesis testing even based on the analysis of individual Likert-item responses, we identified situations that require special attention. Unfortunately, nonparametric approaches may not necessarily work well in these situations.

In our prior research [@fages:2022], we applied ordered probit models [@Burkner:2019;@Christensen2023ordinal] for analyzing ordinal responses in two experimental studies. Since the HCI community is not familiar with these models, we found it necessary to provide motivation for their use and offer additional explanations to reviewers and readers on how to interpret results. These models offer numerous advantages, including their ability to support interval estimation and prediction in addition to null hypothesis testing. However, researchers are required to acquaint themselves with concepts of generalized mixed-effects models or Bayesian inference in order to use them correctly. We hope that our examples in the paper and the supplementary materials can inspire the community to give these methods more thoughtful consideration.

### Interpreting interactions
 We discussed how the scale of the data affects the interpretation of interactions [@Loftus:1978]. In particular, when distributional assumptions are violated, a positive test for interaction may not indicate a genuine interaction. We demonstrated that nonparametric methods do not eliminate this problem, as they also struggle to control the Type I error rate of interactions in the presence of strong parallel main effects. Larger sample sizes do not mitigate this issue; instead, the ratio of false positives increases with sample size because the test becomes more sensitive.

A useful method for diagnosing such issues is to plot interactions (e.g., see @fig-removable-interactions) and visually inspect the trends of the effects. @Loftus:1978 notes that interactions that do not cross over are removable under monotonic transformations, and therefore they cannot be interpreted. In constrast, interactions that cross over are non-removable. @fig-removable-interactions presents examples of a removable and a non-removable interaction. It is noteworthy that the crossover of the non-removable interaction is only discernible on the second chart (in <span style="color:#FF5E00"> **orange-red**</span>), where each line shows the effect of Technique.

::: {#fig-removable-interactions}
```{r, echo=FALSE, message=FALSE, fig.height=3.5, fig.width = 9, warning=FALSE}
df1 <- data.frame(Difficulty = c("easy","hard","easy","hard"), Technique=c("Tech A","Tech A","Tech B","Tech B"),
   Time=c(0.3, 1.1, 1.5, 3.8))

df2 <- data.frame(Difficulty = c("easy","hard","easy","hard"), Technique=c("Tech A","Tech A","Tech B","Tech B"),
   Time=c(1.5, 0.4, 2.5, 3.8))

annotations = list( 
  list( 
    x = 0.2,  
    y = 1.0,  
    text = "removable interaction",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  ),  
  list( 
    x = 0.8,  
    y = 1,  
    text = "nonremovable interaction",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  ))

fig1 <- createInteractionPlot(df1, palette = c("#888888", "#888888"))
fig2 <- createInteractionPlotInv(df1, palette = c("#FF5E00", "#FF5E00"))
fig3 <- createInteractionPlot(df2, offx = -0.3, palette = c("#888888", "#888888"))
fig4 <- createInteractionPlotInv(df2, offx = -0.65, offy = 1.5, palette = c("#FF5E00", "#FF5E00"))

figA <- subplot(fig1, fig2, titleY = TRUE, shareY = TRUE, titleX = TRUE, margin = 0.04)
figB <- subplot(fig3, fig4, titleY = TRUE, shareY = TRUE, titleX = TRUE, margin = 0.04)

fig <- subplot(figA, figB, titleY = TRUE, margin = 0.06) %>%   
  layout(annotations = annotations) %>%
  config(displayModeBar = FALSE, scrollZoom = FALSE, displaylogo = FALSE) %>% layout(hovermode = 'x', dragmode='pan')

fig
```
Examples of a removable (left) and a non-removable interaction (right). A removable interaction does not cross over, regardless of how it is plotted. 
:::

### Research planning and transparency
Planning analysis methods before data collection is a sensible practice, as it helps focus on measures that are easy to analyze and communicate, and anticipate the types of responses they may generate. We also encourage authors to open their datasets so that other researchers can replicate their findings and evaluate the robustness of their analysis methods. Although ART has been used in a large number of studies, very few datasets from these studies are openly available. Therefore, the extent to which the use of this method has contaminated research outcomes remains unclear and unfortunately difficult to evaluate.

### 2025-05-24 supplementary/answers/Main-1-within-4x3-Testing.R

@@ -1,95 +1,0 @@
-# Author: Theophanis Tsandilas, 2023-2024
-# Inria & Université Paris-Saclay
-
-# Experiment evaluating the Type I error rates of PAR, ART, RNK, and INT on ratio scales (distributions below) for a 4x3 within-subjects design
-# See Fig. 9, Fig. 11, Fig. 12
-
-# By default, this code will generate a data file for n = 20 with 5000 iteration. 
-# You can edit the experimental parameters in the code to test different conditions.
-
-rm(list=ls())
-
-source("data-generators.R") # Import the data-generation functions
-
-plotCellLatent <- function(df, x1s, x2 = "B1", min = -8, max = 8) {
-	colors = c("#666666", "#E69F00", "#009E73", "#FF5E00")
-
-	createText <- function(values){
-		mean <- mean(values)
-		sd <- sd(values)
-		paste("m=", round(mean,digits=2), ", sd=", round(sd, digits = 2), sep = "")
-	}
-
-	values <- df[df$x1 == x1s[1] & df$x2 == x2, ]$dv
-	plot(density(values), col = colors[1], lwd = 2, ylim = c(0, .4), xlim = c(min, max), main = paste("x2 =", x2), xlab = NA, bty="l")
-	text(median(values) - 1, max(density(values)$y - .05), createText(values), pos = 2, col = colors[1])
-
-	values <- df[df$x1 == x1s[2] & df$x2 == x2, ]$dv
-	lines(density(values), col = colors[2], lwd = 2, ylim = c(0, .4), xlim = c(min, max))
-	text(median(values)+.3, max(density(values)$y + .03), createText(values), pos = 2, col = colors[2])
-
-	values <- df[df$x1 == x1s[3] & df$x2 == x2, ]$dv
-	lines(density(values), col = colors[3], lwd = 2, ylim = c(0, .4), xlim = c(min, max))
-	text(median(values)-.3, max(density(values)$y + .03), createText(values), pos = 4, col = colors[3])
-
-	values <- df[df$x1 == x1s[4] & df$x2 == x2, ]$dv
-	lines(density(values), col = colors[4], lwd = 2, ylim = c(0, .4), xlim = c(min, max))
-	text(median(values)+1, max(density(values)$y -.05), createText(values), pos = 4, col = colors[4])
-	
-}
-
[…]

### 2025-05-24 supplementary/answers/Main-1-within-4x3-ratio.R

@@ -1,103 +1,0 @@
-# Author: Theophanis Tsandilas, 2023-2024
-# Inria & Université Paris-Saclay
-
-# Experiment evaluating the Type I error rates of PAR, ART, RNK, and INT on ratio scales (distributions below) for a 4x3 within-subjects design
-# See Fig. 9, Fig. 11, Fig. 12
-
-# By default, this code will generate a data file for n = 20 with 5000 iteration. 
-# You can edit the experimental parameters in the code to test different conditions.
-
-rm(list=ls())
-library(lmerTest)
-
-# The ARTool
-library(ARTool)
-
-# Parallel computation
-library(foreach)
-library(doParallel)
-
-source("data-generators.R") # Import the data-generation functions
-
-getLmerPValues <- function(model, vars) {
-	return(anova(model)[vars, 6])
-}
-
-
-getARTPValues <- function(model, vars) {
-	return(anova(model)[vars, 5]) # For within-subjects
-}
-
-# Repeated measures with two factors
-test.within.2f <- function(n, design, effects, distr){ 
-	df <- sim.within.2f(n, design, effects) %>% toDistribution(distr)
-
-	model0 <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
-	model1 <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
-	model2 <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
-	model3 <- suppressMessages(lmer(inverseNormalTransform(y) ~ x1*x2 + (1|s), data=df)) # INT
-	vars <- c("x1", "x2", "x1:x2")
-	c(getLmerPValues(model0, vars), getLmerPValues(model1, vars), getARTPValues(model2, vars), getLmerPValues(model3, vars))
-}
-
-# Performs repetitive tests for the given configuration
-test <- function(repetitions, n, design, effects, distr) {
-	results <- foreach(rid = 1:repetitions, .combine=rbind) %dopar% {
-		tryCatch(
-			{
-				test.within.2f(n, design, effects, distr)
-			}, 
-			error = function(cond) {
-				# do nothing
-			}, finally = {
-				# do nothing
-			}
-		)
-	}
-
[…]

### 2025-05-24 supplementary/answers/example-interactions.R

@@ -1,45 +1,0 @@
-rm(list=ls())
-library(tidyverse)
-library(ggplot2)
-library(gridExtra)
-
-source("data-generators.R") # Import the data-generation functions
-
-design <- c(4,3)
-
-effects <- c(2, 0, 0)
-distr <- "lnorm"
-n <- 12
-
-plotInteraction <- function(){
-	df <- sim.within.2f(n, design, effects) %>% toDistribution(distr)
-	df <- df[,c(1,2,3,9)] %>% 
-		rename(Participant=s, Difficulty=x1, Technique=x2,Time=y) %>%
-		mutate(Difficulty=recode(Difficulty,"A1" = "Level1", "A2"="Level2","A3"="Level3","A4"="Level4")) %>%
-		mutate(Technique=recode(Technique,"B1" = "Technique A", "B2" = "Technique B", "B3" = "Technique C"))
-
-	dfmeans <- aggregate(df, Time ~ Difficulty + Technique, FUN = mean)
-
-
-	cbPalette <- c("#999999", "#E69F00", "#F15854")
-
-	p <- dfmeans %>% ggplot(aes(x = Difficulty, y = Time, color = Technique, group=Technique)) + 
-		geom_line(linewidth=1.2) +
-		geom_point(size=2) +
-		ylab("Mean Time (min)") +
-		ylim(0, 4) + 
-		theme_bw() + theme(legend.position = "none") + 
-		scale_color_manual(values=cbPalette)
-}
-
-set.seed(3000)
-p1 <- plotInteraction()
-p2 <- plotInteraction()
-p3 <- plotInteraction()
-p4 <- plotInteraction()
-p5 <- plotInteraction()
-p6 <- plotInteraction()
-
-grid.arrange(p1, p2, p3, p4, p5, p6, ncol=3, widths=c(2.3, 2.3, 2.3))
-
-

### 2025-05-24 supplementary/answers/p-values.R

@@ -1,124 +1,0 @@
-# Author: Theophanis Tsandilas, 2023-2024
-# Inria & Université Paris-Saclay
-
-# Experiment evaluating the effect size estimates of PAR, ART, RNK, and INT.
-# You can edit the experimental parameters in the code to test different conditions.
-
-
-rm(list=ls())
-library(lmerTest)
-library(effectsize) # Effect size functions
-
-# The ARTool
-library(ARTool)
-
-# Parallel computation
-library(foreach)
-library(doParallel)
-
-source("data-generators.R") # Import the data-generation functions
-
-eta_values = function(model) {
-	return(round(eta_squared(anova(model))[, 2], digits=4))
-}
-
-getLmerPValues <- function(model, vars) {
-	return(anova(model)[vars, 6])
-}
-
-getARTPValues <- function(model, vars) {
-	return(anova(model)[vars, 5]) # For within-subjects
-}
-
-# Repeated measures with two factors
-test.within.2f <- function(n, design, effects, distr){ 
-	df <- sim.within.2f(n, design, effects) %>% toDistribution(distr)
-
-	model = suppressMessages(lmer(dv ~ x1*x2 + (1|s), data=df)) # This is the reference groundtruth model
-
-	model0 <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
-	model1 <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
-	model2 <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
-	model3 <- suppressMessages(lmer(inverseNormalTransform(y) ~ x1*x2 + (1|s), data=df)) # INT
-
-	vars <- c("x1", "x2", "x1:x2")
-	c(effects, 
-		getLmerPValues(model, vars), getLmerPValues(model0, vars), getLmerPValues(model1, vars), getARTPValues(model2, vars), getLmerPValues(model3, vars),
-		eta_values(model), eta_values(model0), eta_values(model1), eta_values(model2), eta_values(model3))
-}
-
-# Performs repetitive tests for the given configuration
-test <- function(repetitions, n, design, distr, effects) {
-	results <- foreach(rid = 1:repetitions, .combine=rbind) %dopar% {
[…]

### 2025-05-24 supplementary/answers/sim-data.R

@@ -1,114 +1,0 @@
-# Author: Theophanis Tsandilas, 2023-2024
-# Inria & Université Paris-Saclay
-
-# Experiment evaluating the Type I error rates of PAR, ART, RNK, and INT on ratio scales (distributions below) for a 4x3 within-subjects design
-# See Fig. 9, Fig. 11, Fig. 12
-
-# By default, this code will generate a data file for n = 20 with 5000 iteration. 
-# You can edit the experimental parameters in the code to test different conditions.
-
-rm(list=ls())
-library(lmerTest)
-
-# The ARTool
-library(ARTool)
-
-# Parallel computation
-library(foreach)
-library(doParallel)
-
-source("data-generators.R") # Import the data-generation functions
-
-getLmerPValues <- function(model, vars) {
-	return(anova(model)[vars, 6])
-}
-
-
-getARTPValues <- function(model, vars) {
-	return(anova(model)[vars, 5]) # For within-subjects
-}
-
-# Repeated measures with two factors
-test.within.2f <- function(n, design, effects, distr){ 
-	df <- sim.within.2f(n, design, effects) %>% toDistribution(distr)
-
-	model0 <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
-	model1 <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
-	model2 <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
-	model3 <- suppressMessages(lmer(inverseNormalTransform(y) ~ x1*x2 + (1|s), data=df)) # INT
-	vars <- c("x1", "x2", "x1:x2")
-	c(getLmerPValues(model0, vars), getLmerPValues(model1, vars), getARTPValues(model2, vars), getLmerPValues(model3, vars))
-}
-
-# Performs repetitive tests for the given configuration
-test <- function(repetitions, n, design, effects, distr) {
-	results <- foreach(rid = 1:repetitions, .combine=rbind) %dopar% {
-		tryCatch(
-			{
-				test.within.2f(n, design, effects, distr)
-			}, 
-			error = function(cond) {
-				# do nothing
-			}, finally = {
-				# do nothing
-			}
-		)
-	}
-
[…]

### 2025-05-24 supplementary/examples-case-studies/1-Introduction/example-cut.R

@@ -1,33 +1,0 @@
-rm(list=ls())
-library(lmerTest) # Lmer functions
-library(ARTool) # ARTool library
-library(effectsize) # To estimate effect sizes 
-
-library(tidyverse)
-library("emmeans")
-
-
-df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)
-#df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)
-
-df$Participant <- factor(df$Participant)
-df$Difficulty <- factor(df$Difficulty, ordered = TRUE) # We can consider it as ordinal
-df$Technique <- factor(df$Technique)
-
-
-# Implementation of the Inverse Normal Transform
-int = function(x){
-   qnorm((rank(x) - 0.5)/length(x))
-}
- 
-#df[df$Technique == "A" & df$Difficulty == "Level4", ]$Time = df[df$Technique == "A" & df$Difficulty == "Level4", ]$Time  + .13
-df$Time <- ifelse(df$Difficulty == "Level4" & df$Technique == "A", df$Time + .13, df$Time)
-
-#df <- df[df$Difficulty != "Level4" | (df$Time < 9 & df$Time > 1.07), ]
-
-mpar <- lmer(Time ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - no transformation
-mlog <- lmer(log(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - Log-transformation
-mart <- art(Time ~ Difficulty*Technique + (1|Participant), data=df) # ARTool
-mrnk <- lmer(rank(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Rank transformation 
-mint <- lmer(int(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Inverse Transform transformation
-

### 2025-05-24 supplementary/examples-case-studies/1-Introduction/example.R

@@ -1,31 +1,0 @@
-rm(list=ls())
-library(lmerTest) # Lmer functions
-library(ARTool) # ARTool library
-library(effectsize) # To estimate effect sizes 
-
-library(tidyverse)
-library("emmeans")
-
-
-df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)
-
-df$Participant <- factor(df$Participant)
-df$Difficulty <- factor(df$Difficulty, ordered = TRUE) # We can consider it as ordinal
-df$Technique <- factor(df$Technique)
-
-
-# Implementation of the Inverse Normal Transform
-int = function(x){
-   qnorm((rank(x) - 0.5)/length(x))
-}
-
-mpar <- lmer(Time ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - no transformation
-mlog <- lmer(log(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - Log-transformation
-mart <- art(Time ~ Difficulty*Technique + (1|Participant), data=df) # ARTool
-mrnk <- lmer(rank(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Rank transformation 
-mint <- lmer(int(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Inverse Transform transformation
-
-
-
-
-

### 2025-05-24 supplementary/experimental-code/Main-1-within-4x3-ratio.R

@@ -105,7 +105,7 @@ effects <- matrix(c(0, 0, 0,
 Ns <- c(20) 
 
 # 5000 iterations
-R <- 1000
+R <- 5000
 
 filename = "1_test_4x3_Ratio"
 

### 2025-05-24 supplementary/experimental-code/Main-6-effect-sizes.R

@@ -2,7 +2,7 @@
 # Inria & Université Paris-Saclay
 
 # Experiment evaluating the effect size estimates of PAR, ART, RNK, and INT.
-# See Fig. 30, Fig. 31
+# See Fig. 31, Fig. 32
 
 # You can edit the experimental parameters in the code to test different conditions.
 

### 2025-05-24 supplementary/experimental-code/Test-p-values-scatterplots.R

@@ -1,135 +1,0 @@
-# Author: Theophanis Tsandilas, 2023-2024
-# Inria & Université Paris-Saclay
-
-# Experiment evaluating the effect size estimates of PAR, ART, RNK, and INT.
-# See Fig. 28, Fig. 29
-
-# You can edit the experimental parameters in the code to test different conditions.
-
-
-rm(list=ls())
-library(lmerTest)
-library(effectsize) # Effect size functions
-
-# The ARTool
-library(ARTool)
-
-# Parallel computation
-library(foreach)
-library(doParallel)
-
-source("data-generators.R") # Import the data-generation functions
-
-
-eta_values = function(model) {
-	return(eta_squared(anova(model))[, 2])
-}
-
-getLmerPValues <- function(model, vars, lmer = TRUE) {
-	return(suppressMessages(anova(model)[vars, ifelse(lmer, 6, 5)])) # For aov models the table is different
-}
-
-getARTPValues <- function(model, vars, lmer = TRUE) {
-	return(suppressMessages(anova(model)[vars, ifelse(lmer, 5, 7)])) # For aov models the ART's table is different
-}
-
-
-# Repeated measures with two factors
-test.within.2f <- function(n, design, effects, distr){ 
-	df <- sim.within.2f(n, design, effects) %>% toDistribution(distr)
-
-	model = suppressMessages(lmer(dv ~ x1*x2 + (1|s), data=df)) # This is the reference groundtruth model
-
-	model0 <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
-	model1 <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
-	model2 <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
-	model3 <- suppressMessages(lmer(inverseNormalTransform(y) ~ x1*x2 + (1|s), data=df)) # INT
-
-#	c(effects, eta_values(model), eta_values(model0), eta_values(model1), eta_values(model2), eta_values(model3))
-
-	vars <- c("x1", "x2", "x1:x2")
-	c(effects, getLmerPValues(model, vars), getLmerPValues(model0, vars), getLmerPValues(model1, vars), getARTPValues(model2, vars), getLmerPValues(model3, vars))
-}
-
-# Performs repetitive tests for the given configuration
[…]

### 2025-05-24 supplementary/experimental-code/data-generators.R

@@ -14,7 +14,7 @@ library(tidyr)
 # https://www.biostars.org/p/80597/
 # use rankit version where c = 0.5, based on https://www.researchgate.net/publication/286889581_Impact_of_Rank-Based_Normalizing_Transformations_on_the_Accuracy_of_Test_Scores
 inverseNormalTransform = function(x){
-	qnorm((rank(x,na.last="keep") - 0.5)/sum(!is.na(x)))
+	qnorm((rank(x) - 0.5)/length(x))
 }
 
 # Transforms a set of values to ordinal based on the given thresholds

### 2025-05-24 supplementary/experimental-code/np.anova.R

@@ -1,3 +1,7 @@
+# This code was downloaded from http://www.uni-koeln.de/~luepsen/R/
+# As we only call the method with pseudo = F, the pseudorank library is not loaded.
+# Note that this library was removed from CRAN but can be obtained from the archive: https://cran.r-project.org/web/packages/pseudorank/index.html
+
 np.anova <- function(formel,data,method=0,compact=T,pseudo=F)
 {                        # data  :  dataframe 
                          # formel:  anova-Modell
@@ -10,7 +14,7 @@ np.anova <- function(formel,data,method=0,compact=T,pseudo=F)
                          #          F: seperate tables for each error term (repeated measures only)
                          # pseudo   T: use pseudoranks instead of ordinary ranks (for unequal ni)
 
-if (pseudo) library(pseudorank)
+if (pseudo) library(pseudorank) # See: https://cran.r-project.org/web/packages/pseudorank/index.html
                                                # check formula
 if (mode(formel)!="call") stop("invalid formula")
 if (class(data[,as.character(formel)[2]])=="factor") stop("invalid type of dependent variable")

### 2025-05-24 supplementary/rebuttals/effect-size/plot-results-lognormal.R

@@ -1,0 +1,27 @@
+rm(list=ls())
+
+data <- read.csv("log/effects-20.csv", sep=",", header=TRUE, strip.white=TRUE)
+data$effectX2 <- factor(data$effectX2)
+
+ylab = "X1 partial eta squared"
+xlab = "X2 effect (a2)"
+
+par(mfrow = c(4, 2), mar = c(4, 4, 2, 1))
+
+df <- data[data$effectX1 == 0,]
+min = 0
+max = 0.21
+
+plot(df$effectX2, df$X1, ylim = c(min, max), ylab = ylab, xlab = xlab, main = "Groundtruth", col = "white")
+plot(df$effectX2, df$parX1, ylim = c(min, max), ylab = ylab, xlab = xlab, main = "PAR", col = "grey")
+plot(df$effectX2, df$intX1, ylim = c(min, max), ylab = ylab, xlab = xlab, main = "INT", col = "green3")
+plot(df$effectX2, df$artX1, ylim = c(min, max), ylab = ylab, xlab = xlab, main = "ART", col = "tomato1")
+
+df <- data[data$effectX1 == 1,]
+min = 0
+max = 0.41
+
+plot(df$effectX2, df$X1, ylim = c(min, max), ylab = ylab, xlab = xlab, main = "Groundtruth", col = "white")
+plot(df$effectX2, df$parX1, ylim = c(min, max), ylab = ylab, xlab = xlab, main = "PAR", col = "grey")
+plot(df$effectX2, df$intX1, ylim = c(min, max), ylab = ylab, xlab = xlab, main = "INT", col = "green3")
+plot(df$effectX2, df$artX1, ylim = c(min, max), ylab = ylab, xlab = xlab, main = "ART", col = "tomato1")

### 2025-05-24 supplementary/rebuttals/effect-size/plot-results-ordinal.R

@@ -1,0 +1,18 @@
+rm(list=ls())
+
+data <- read.csv("log/effects-ordinal.csv", sep=",", header=TRUE, strip.white=TRUE)
+data$n <- factor(data$n)
+
+ylab = "X1 partial eta squared"
+xlab = "n"
+
+par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
+
+df <- data[data$effectX1 == 0,]
+min = 0
+max = 0.21
+
+plot(df$n, df$X1, ylim = c(min, max), ylab = ylab, xlab = xlab, main = "Groundtruth", col = "white")
+plot(df$n, df$parX1, ylim = c(min, max), ylab = ylab, xlab = xlab, main = "PAR", col = "grey")
+plot(df$n, df$intX1, ylim = c(min, max), ylab = ylab, xlab = xlab, main = "INT", col = "green3")
+plot(df$n, df$artX1, ylim = c(min, max), ylab = ylab, xlab = xlab, main = "ART", col = "tomato1")

### 2025-05-24 supplementary/rebuttals/effect-size/test-effect-sizes-lognormal.R

@@ -1,0 +1,122 @@
+# Author: Theophanis Tsandilas, 2023-2024
+# Inria & Université Paris-Saclay
+
+# Experiment evaluating the effect size estimates of PAR, ART, RNK, and INT.
+# You can edit the experimental parameters in the code to test different conditions.
+
+
+rm(list=ls())
+library(lmerTest)
+library(effectsize) # Effect size functions
+
+# The ARTool
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+source("data-generators.R") # Import the data-generation functions
+
+
+eta_values = function(model) {
+	return(round(eta_squared(anova(model))[, 2], digits=4))
+}
+
+# Repeated measures with two factors
+test.within.2f <- function(n, design, effects, distr){ 
+	df <- sim.within.2f(n, design, effects) %>% toDistribution(distr)
+
+	model = suppressMessages(lmer(dv ~ x1*x2 + (1|s), data=df)) # This is the reference groundtruth model
+
+	model0 <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
+	model1 <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
+	model2 <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
+	model3 <- suppressMessages(lmer(inverseNormalTransform(y) ~ x1*x2 + (1|s), data=df)) # INT
+
+	c(effects, eta_values(model), eta_values(model0), eta_values(model1), eta_values(model2), eta_values(model3))
+}
+
+# Performs repetitive tests for the given configuration
+test <- function(repetitions, n, design, distr, effects) {
+	results <- foreach(rid = 1:repetitions, .combine=rbind) %dopar% {
+	#	effects = c(runif(1, min=-maxeffects[1], max=maxeffects[1]), maxeffects[2], maxeffects[3])
+	effects = c(effects[1], effects[2], effects[3])
+
+		tryCatch(
+			{
+				test.within.2f(n, design, effects, distr)
+			}, 
+			error = function(cond) {
+				# do nothing
+			}, finally = {
+				# do nothing
+			}
+		)
+	}
+	
+	designStr <- paste(design, collapse="x")
+
+	# Split the results into separate rows 
[…]

### 2025-05-24 supplementary/rebuttals/higgins/art-plotting/example-art-plots.R

@@ -1,0 +1,19 @@
+library(ARTool) # ARTool library
+
+df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)
+
+df$Difficulty <- factor(df$Difficulty, ordered = TRUE) # We can consider it as ordinal
+df$Technique <- factor(df$Technique)
+
+mart <- art(Time ~ Difficulty*Technique + (1|Participant), data=df) # ARTool
+
+par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
+plot(df$Time,rank(df$Time), bty="l", xlab = "Time", ylab = "rank (Time)")
+#text(7,40, paste("Spearman correlation = ", round(cor(df$Time,rank(df$Time), method='spearman'), digits=2), sep=""))
+
+plot(df$Time,mart$aligned.ranks[,1], bty="l", xlab = "Time", ylab = "ART-Difficulty (Time)") 
+#text(7,40, paste("Spearman correlation = ", round(cor(df$Time,mart$aligned.ranks[,1], method='spearman'), digits=2), sep=""))
+
+plot(df$Time,mart$aligned.ranks[,2], bty="l", xlab = "Time", ylab="ART-Technique (Time)")
+#text(7,40, paste("Spearman correlation = ", round(cor(df$Time,mart$aligned.ranks[,2], method='spearman'), digits=2), sep=""))
+

### 2025-05-24 supplementary/rebuttals/higgins/challenge/analysis-ratings.R

@@ -1,0 +1,46 @@
+rm(list=ls())
+
+
+library(lmerTest) # Lmer functions
+library(ARTool) # ARTool library
+#library(effectsize) # To estimate effect sizes 
+
+options(scipen = 999) # To show full decimal digits
+
+# Implementation of the Inverse Normal Transform
+int = function(x){
+   qnorm((rank(x) - 0.5)/length(x))
+}
+
+printAnalysis <- function(num, method, model) {
+  cat(num, ". Analysis using ", method, "\n\n", sep="")
+
+  print(anova(model))
+  #print(eta_squared(model))
+  #print(cohens_f(model))
+}
+
+df <- read.csv("large-sample-ratings.csv", sep=",", header=TRUE, strip.white=TRUE)
+df$Participant <- factor(df$Participant)
+df$Difficulty <- factor(df$Difficulty, ordered = TRUE) # We can consider it as ordinal
+df$Technique <- factor(df$Technique)
+
+mpar <- lmer(Rating ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - no transformation
+mart <- art(Rating ~ Difficulty*Technique + (1|Participant), data=df) # ARTool
+
+
+df$rnk <- rank(df$Rating)
+df$int <- int(df$Rating)
+
+mrnk <- lmer(rnk ~ Difficulty*Technique + (1|Participant), data=df) # Rank transformation 
+mint <- lmer(int ~ Difficulty*Technique + (1|Participant), data=df) # Inverse Transform transformation
+mrnkart <- art(rnk ~ Difficulty*Technique + (1|Participant), data=df) # RNK then ART
+mintart <- art(int ~ Difficulty*Technique + (1|Participant), data=df) # INT then ART
+
+printAnalysis(1, "PAR", mpar)
+printAnalysis(2, "ART", mart)
+printAnalysis(3, "RNK", mrnk)
+printAnalysis(4, "INT", mint)
+
+printAnalysis(5, "RNK -> ART", mrnkart)
+printAnalysis(6, "INT -> ART", mintart)

### 2025-05-24 supplementary/rebuttals/higgins/challenge/analysis-time.R

@@ -1,0 +1,40 @@
+rm(list=ls())
+
+
+library(lmerTest) # Lmer functions
+library(ARTool) # ARTool library
+#library(effectsize) # To estimate effect sizes 
+
+options(scipen = 999) # To show full decimal digits
+
+# Implementation of the Inverse Normal Transform
+int = function(x){
+   qnorm((rank(x) - 0.5)/length(x))
+}
+
+printAnalysis <- function(num, method, model) {
+  cat(num, ". Analysis using ", method, "\n\n", sep="")
+
+  print(anova(model))
+  #print(eta_squared(model))
+  #print(cohens_f(model))
+}
+
+df <- read.csv("large-sample-times.csv", sep=",", header=TRUE, strip.white=TRUE)
+df$Participant <- factor(df$Participant)
+df$Difficulty <- factor(df$Difficulty, ordered = TRUE) # We can consider it as ordinal
+df$Technique <- factor(df$Technique)
+
+mpar <- lmer(Time ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - no transformation
+mart <- art(Time ~ Difficulty*Technique + (1|Participant), data=df) # ARTool
+
+mlog <- lmer(log(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - Log-transformation
+mrnk <- lmer(rank(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Rank transformation 
+mint <- lmer(int(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Inverse Transform transformation
+
+printAnalysis(1, "PAR", mpar)
+printAnalysis(2, "ART", mart)
+printAnalysis(5, "LOG", mlog)
+printAnalysis(4, "RNK", mrnk)
+printAnalysis(5, "INT", mint)
+

### 2025-05-24 supplementary/rebuttals/higgins/homoscedastic/experiment/data-generators.R

@@ -1,0 +1,90 @@
+# Author: Theophanis Tsandilas, Inria & Université Paris-Saclay
+
+# A great library for simulating data for a large range of desings: https://debruine.github.io/faux
+library(faux)
+
+library(dplyr)
+library(tidyr)
+
+
+# https://yingji15.github.io/2019-08-17-inverse-normal-transformation/
+# https://www.biostars.org/p/80597/
+# use rankit version where c = 0.5, based on https://www.researchgate.net/publication/286889581_Impact_of_Rank-Based_Normalizing_Transformations_on_the_Accuracy_of_Test_Scores
+inverseNormalTransform = function(x){
+	qnorm((rank(x) - 0.5)/length(x))
+}
+
+# Additional support for other distributions with not conversion support at faux
+# Based on https://github.com/debruine/faux/blob/master/R/distribution_convertors.R
+
+# To Cauchy distribution 
+norm2cauchy <- function(x, location = 0, scale = 1, mu = mean(x), sd = stats::sd(x), ...) {
+  p <- stats::pnorm(x, mu, sd)
+  stats::qcauchy(p, location, scale, ...)
+}
+
+norm2lnorm <- function(x, meanlog = 0, sdlog = 1, mu = mean(x), sd = stats::sd(x), ...) {
+	 p <- stats::pnorm(x, mu, sd)
+ 	 stats::qlnorm(p, meanlog, sdlog, ...)
+}
+
+
+# We control the standard deviation of the generated data so that it is equal
+sdlog <- function(mean, sd) {
+	sqrt(log(1 + sd^2/mean^2))	
+}
+
+meanlog <- function(mean, sd) {
+	log(mean^2/sqrt(sd^2 + mean^2))
+}
+
+# Here, we ensure that the reponses come from log-normal populations with equal variances
+toDistributionLNormHomo <- function(data, means = c(0.5, 1, 1.5, 2), sd = 1.5) {
+	data$y <- 0
+
+	data[data$x1 == "A1",]$y <- norm2lnorm(data[data$x1 == "A1",]$dv, meanlog = meanlog(means[1], sd), sdlog = sdlog(means[1], sd))
+	data[data$x1 == "A2",]$y <- norm2lnorm(data[data$x1 == "A2",]$dv, meanlog = meanlog(means[2], sd), sdlog = sdlog(means[2], sd))
[…]

### 2025-05-24 supplementary/rebuttals/higgins/homoscedastic/experiment/homoscedastic-lnorm-4x3.R

@@ -1,0 +1,139 @@
+# Author: Theophanis Tsandilas, Inria & Université Paris-Saclay, 2024
+# An experiment to test homoscedastic log-normal distributions
+
+rm(list=ls())
+library(lmerTest)
+
+# The ARTool
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+source("data-generators.R") # Import the data-generation functions
+
+getLmerPValues <- function(model, vars) {
+	return(anova(model)[vars, 6])
+}
+
+
+getARTPValues <- function(model, vars) {
+	return(anova(model)[vars, 5]) # For within-subjects
+}
+
+# Repeated measures with two factors
+test.within.2f <- function(n, design, sd){ 
+	df <- sim.within.2f(n, design, effects = c(2,0,0)) %>% toDistributionLNormHomo(sd = sd)
+
+	model0 <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
+	model1 <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
+	model2 <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
+	model3 <- suppressMessages(lmer(inverseNormalTransform(y) ~ x1*x2 + (1|s), data=df)) # INT
+	vars <- c("x1", "x2", "x1:x2")
+	c(getLmerPValues(model0, vars), getLmerPValues(model1, vars), getARTPValues(model2, vars), getLmerPValues(model3, vars))
+}
+
+
+repeating <- function(n=20, design = c(4,3), sd) {
+	keep <- TRUE
+
+	counter <- 0
+	df <- NA
+	while(keep) {
+		df <- sim.within.2f(n, design, effects = c(2,0,0)) %>% toDistributionLNormHomo(sd = sd)
+
+		model0 <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
+		model1 <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
+		model2 <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
+		model3 <- suppressMessages(lmer(inverseNormalTransform(y) ~ x1*x2 + (1|s), data=df)) # INT
+		vars <- c("x1", "x2", "x1:x2")
[…]

### 2025-05-24 supplementary/rebuttals/higgins/p-values/analysis-2.R

@@ -1,0 +1,30 @@
+rm(list=ls())
+
+
+df <- read.csv("p-values-noeffect-12.csv", sep=",", header=TRUE, strip.white=TRUE)
+
+min = min(df$artX1,df$parX1)
+max = max(df$artX1,df$parX1)
+
+#min = min(df$pX1,df$artX1,df$parX1)
+#max = max(df$pX1,df$artX1,df$parX1)
+
+par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
+
+plot(df$parX1, df$artX1, ylim=c(min, max), xlim=c(min,max), col="blue", log='xy', frame.plot=F, ylab="ART p-value", xlab="PAR p-value", pch=4, main = "n = 12")
+par(new=TRUE)
+curve(1*x, add = T,col="grey")
+curve(.05+0*x, add = T,col="red", lty =2)
+text(min, .07, "alpha = .05", col = "red", pos = 4)
+abline(v=0.05, col="red", lty =2)
+
+
+df <- read.csv("p-values-noeffect-1000.csv", sep=",", header=TRUE, strip.white=TRUE)
+min = min(df$pX1,df$artX1,df$intX1,df$parX1)
+max = max(df$pX1,df$artX1,df$intX1,df$parX1)
+plot(df$parX1,df$artX1, ylim=c(min, max), xlim=c(min,max), col="blue", log='xy', frame.plot=F, ylab="ART p-value", xlab="PAR p-value", pch=4, main = "n = 1000")
+par(new=TRUE)
+curve(1*x, add = T,col="grey")
+curve(.05+0*x, add = T,col="red", lty =2)
+text(min, .09, "alpha = .05", col = "red", pos = 4)
+abline(v=0.05, col="red", lty =2)

### 2025-05-24 supplementary/rebuttals/higgins/p-values/analysis-interaction.R

@@ -1,0 +1,40 @@
+rm(list=ls())
+
+
+df <- read.csv("p-values-12-interaction.csv", sep=",", header=TRUE, strip.white=TRUE)
+
+min = min(df$pX1X2,df$artX1X2,df$intX1X2,df$parX1X2)
+max = max(df$pX1X2,df$artX1X2,df$intX1X2,df$parX1X2)
+
+#min = min(df$pX1,df$artX1,df$parX1)
+#max = max(df$pX1,df$artX1,df$parX1)
+
+par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
+
+plot(df$pX1X2,df$artX1X2, ylim=c(min, max), xlim=c(min,max), col="blue", log='xy', frame.plot=F, ylab="p-value", xlab="reference p-value (log)", pch=4, main = "n = 12")
+par(new=TRUE)
+plot(df$pX1X2,df$parX1X2, ylim=c(min, max), xlim=c(min,max),  col = rgb(.6,.6,.6, 0.8), log='xy', bty ="n", axes=F, frame.plot=F, xaxt='n', yaxt='n', xlab=NA, ylab=NA, pch=5)
+par(new=TRUE)
+#plot(df$pX1,df$intX1,  ylim=c(min, max), xlim=c(min,max), log='xy', col = "blue", bty ="n", axes=F, frame.plot=F, xaxt='n', yaxt='n', xlab=NA, ylab=NA)
+curve(1*x, add = T,col="grey")
+
+text(.0000001, max, "PAR", col = rgb(.6,.6,.6), pos = 4)
+text(.0000001, 0.2, "ART", col = "blue", pos = 4)
+
+
+#df <- read.csv("p-values-noeffe.csv", sep=",", header=TRUE, strip.white=TRUE)
+min = min(df$pX1X2, df$rnkX1X2,df$intX1X2, df$artX1X2)
+max = max(df$pX1X2, df$rnkX1X2,df$intX1X2, df$artX1X2)
+
+#min = min(df$pX1X2, df$rnkX1X2,df$intX1X2)
+#max = max(df$pX1X2, df$rnkX1X2,df$intX1X2)
+
+plot(df$pX1X2,df$rnkX1X2, ylim=c(min, max), xlim=c(min,max), col="orange", log='xy', frame.plot=F, ylab="p-value", xlab="reference p-value (log)", pch=4, main = "n = 12")
+par(new=TRUE)
[…]

### 2025-05-24 supplementary/rebuttals/higgins/permutation-tests/simulation-test-data-1-birthseason.R

@@ -1,0 +1,110 @@
+# Author: Theophanis Tsandilas, 2024
+# This code demonstrates failures of ART when permutating the data of our illustrative example in Fig 1.
+# It shows that when we randomly assign Techniques to observations, ART still finds a main and interaction effect with a very high error rate  
+# But we now also add spurious variable Birthseason, producing a partial imbalance in the data
+
+rm(list=ls())
+
+library(lmerTest) # Lmer functions
+library(ARTool) # ARTool library
+library(tidyverse)
+
+# This function randomly assigns techniques (A,B,C) to each Difficulty level
+# We also add a spurious variable, the Birthseason of participants (1:spring, 2:summer, 3:automn, 4:winter) randomly assigned to independent observations
+createSample <- function(df, nlevels=4, n=12, toreplace){
+	df[df$Difficulty == "Level1",]$Time <- sample(df[df$Difficulty == "Level1",]$Time, replace = toreplace)
+	df[df$Difficulty == "Level2",]$Time <- sample(df[df$Difficulty == "Level2",]$Time, replace = toreplace)
+	df[df$Difficulty == "Level3",]$Time <- sample(df[df$Difficulty == "Level3",]$Time, replace = toreplace)
+	df[df$Difficulty == "Level4",]$Time <- sample(df[df$Difficulty == "Level4",]$Time, replace = toreplace)
+
+	df$Technique <- rep(c("A","B","C"), nlevels*n)
+	df$Technique <- factor(df$Technique)
+	df$Birthseason <- sample(c(1,2,3,4),144,replace=TRUE)
+	df$Birthseason <- factor(df$Birthseason)
+
+	df
+}
+
+
+# INT implementation
+inverseNormalTransform <- function(x){
+    qnorm((rank(x) - 0.5)/length(x))
+}
+
+# Functions to extract the p-values
+getLmerPValues <- function(model, vars, lmer = TRUE) {
+	return(suppressMessages(anova(model)[vars, ifelse(lmer, 6, 5)])) # For aov models the table is different
+}
+
+getARTPValues <- function(model, vars, lmer = TRUE) {
+	return(suppressMessages(anova(model)[vars, ifelse(lmer, 5, 7)])) # For aov models the ART's table is different
+}
+
+# Analysis with all five methods: PAR, RNK, ART, INT, and LOG
[…]

### 2025-05-25 sandbox.qmd
corrections in intro and background

@@
 title: "The illusory promise of the Aligned Rank Transform"
-subtitle: "Appendix I. Additional experimental results"
+subtitle: "A systematic study of rank transformations"
 author: 
   - name: Theophanis Tsandilas
+    orcid: 0000-0002-0158-228X
     email: theophanis.tsandilas@inria.fr
-    orcid: 0000-0002-0158-228X
     affiliations:
@@
     orcid: 0000-0003-1905-815X
-    email: gery.casiez@univ-lille.fr
+    email: gery.casiezuniv-lille.fr
     affiliations:
@@
 bibliography: bibliography.bib
+csl: canadian-journal-of-philosophy.csl
 
@@
 
-Used for quick testing... eeee
+::: {.callout-important appearance="simple"}
+## Under Review {.unnumbered}
+This paper is [under review](https://www.journalovi.org/under-review.html) on the experimental track of
+the [Journal of Visualization and Interaction](https://www.journalovi.org/). See the 
+[reviewing process](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues).
+:::
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+# Useful libraries
+library(crosstalk)
+library(kableExtra)
+library(gridExtra)
+library(lmerTest)
+library(tidyverse)
+library(plotly)
+```
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+# Source code for reading data from experimental results and for plotting
+source("dataReaders.R")
+source("plotting.R")
+source("effects_plot.R")
+source("interactions_plot.R")
+```
+
+```{=html}
+<style>
+.math.inline .MathJax  {
+  font-size: 105% !important;
+}
+</style>
+```
+
+::: {.callout-note appearance="simple" icon=false collapse=false}
+
+{{< include text/10-abstract.qmd >}}
+
+::: {#tbl-example}
+```{r, echo=FALSE, warning=FALSE}
+df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)
+kbl(df) %>% kable_paper(position = "center") %>% scroll_box(width = "740px", height = "170px")
+```
+
+::: {#fig-example}
+```{r, echo=FALSE, warning=FALSE, fig.height=2.7, fig.width=8}
+cbPalette <- c("#999999", "#E69F00", "#F15854")
+
+p1 <- (df %>%  group_by(Participant,Technique) %>% summarise(Time = median(Time), .groups="drop") %>%   
+        ggplot(aes(x = Technique, y = Time, fill = Technique)) + 
+        geom_boxplot(outlier.shape = NA) +
+        ylab("Median Time (min)") +
+        ylim(0, 3) + 
+        geom_jitter(shape=20, position=position_jitter(0.1, 0)) +
+        theme_bw() + theme(legend.position = "none") + scale_fill_manual(values=cbPalette))
+
+p2 <- (df %>%  group_by(Participant,Difficulty,Technique) %>% summarise(Time = median(Time), .groups="drop") %>%   
+        ggplot(aes(x = Difficulty, y = Time, fill = Technique)) + 
+        geom_boxplot(outlier.shape = NA) +
+        ylab("Median Time (min)") +
+        ylim(0, 6) +
+        #geom_jitter(shape=20, position=position_jitter(0.1)) +
+        geom_point(size = 0.7, position=position_jitterdodge(0.1)) +
+        theme_bw() + theme(legend.position = "none") +
+        #scale_fill_brewer()
+        scale_fill_manual(values=cbPalette))
+
+grid.arrange(p1, p2, nrow = 1, widths = c(1.5, 2.5))
+```
+Boxplots summarizing time performance results for our illustrative example. Dots represent the median time for each individual participant. 
+:::
+
+::: {#fig-example-ordinal}
+```{r, echo=FALSE, warning=FALSE, fig.height=3, fig.width=9}
+df$Perceived_Efficiency <- ordered(df$Perceived_Efficiency, levels = c("inefficient", "somewhat inefficient", "neutral", "somewhat efficient", "efficient"))
+
+p3 <- (df %>%  group_by(Participant,Technique) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop") %>%   
+        ggplot(aes(x = Technique, y = Perceived_Efficiency, fill = Technique)) + 
+        geom_boxplot(outlier.shape = NA) +
+        ylab("Perceived Efficiency") +
+        scale_y_continuous(breaks = 1:5, limits = c(1,5), labels=levels(df$Perceived_Efficiency)) +
+        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
+        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette) +  
+        theme(axis.text.y=element_text(angle=50, hjust=0.7))
+
+#diffPalette <- c("#f9f9f9", "#e2e2e2", "#b9b9b9", "#999999")
+diffPalette <- c("#f9f9fd", "#e2e6fa", "#bacafb", "#99b9f9")
+p4 <- (df %>%  group_by(Participant,Difficulty) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop") %>%   
+        ggplot(aes(x = Difficulty, y = Perceived_Efficiency, fill = Difficulty)) + 
+        geom_boxplot(outlier.shape = NA) + ylim(1,5)+
+        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
+        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=diffPalette) +  
+        theme(axis.title.y=element_blank(), axis.text.y=element_blank())
+
+grid.arrange(p3, p4, nrow = 1, widths = c(3, 3))
+```
+Boxplots summarizing the results on efficiency as perceived by participants. Circles represent the mean rating of each individual participant. 
+:::

### 2025-05-25 supplementary/examples-case-studies/1-Introduction/illustrative-example.Rmd
corrections in intro and background

@@
 The experiment follows a fully balanced repeated-measures factorial design, where each participant ($N = 12$) performs 12 tasks in a unique order.
-The researcher measures the time that it takes participants to complete each task.
+The researchers measure the time that it takes participants to complete each task, but participants also evaluate the perceived efficiency, using an ordinal scale with five levels.
 
@@
 df$Technique <- factor(df$Technique)
+
+df$Perceived_Efficiency <- ordered(df$Perceived_Efficiency, levels = c("inefficient", "somewhat inefficient", "neutral", "somewhat efficient", "efficient"))
 ```
@@
 
+# Plots for the time measure 
 p1 <- (df %>%  group_by(Participant,Technique) %>% 
@@
 grid.arrange(p1, p2, nrow = 1, widths = c(1.5, 2.5))
+
+# Plots for the perceived efficiency measure
+dd1 <- df %>%  group_by(Participant,Technique) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop")
+
+p3 <- (dd1 %>%   
+        ggplot(aes(x = Technique, y = Perceived_Efficiency, fill = Technique)) + 
+        geom_boxplot(outlier.shape = NA) +
+        ylab("Perceived Efficiency (means)") +
+        scale_y_continuous(breaks = 1:5, limits = c(1,5), labels=levels(df$Perceived_Efficiency)) +
+        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
+        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette) +  
+        theme(axis.text.y=element_text(angle=50, hjust=0.7))
+
+dd2 <- df %>%  group_by(Participant,Difficulty) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop")
+diffPalette <- c("#f9f9fd", "#e2e6fa", "#bacafb", "#99b9f9")
+p4 <- (dd2 %>%   
+        ggplot(aes(x = Difficulty, y = Perceived_Efficiency, fill = Difficulty)) + 
+        geom_boxplot(outlier.shape = NA) + ylim(1,5)+
+        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
+        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=diffPalette) +  
+        theme(axis.title.y=element_blank(), axis.text.y=element_blank())
+
+grid.arrange(p3, p4, nrow = 1, widths = c(3, 3))
 ```
 
-### Statistical analysis
+### Statistical analysis of time performance
 
-We conduct an analysis with five different methods: PAR, LOG, RNK, ART, and INT.
+For task-completion times, we conduct an analysis with five different methods: PAR, LOG, RNK, ART, and INT.
 ```{r, warning=FALSE}
@@
 
-### Main and interaction effects
-We finally report the results of our analysis:
+#### Main and interaction effects
+We report the results of our analysis.
 ```{r, warning=FALSE}
@@
 
-### Contrasts
+We observe that only ART shows statistical evidence for a main effect of Technique, as well as an interaction effect.
+
+#### Contrasts
 Finally, we conduct an analysis of pairwise differences between the three techniques *A*, *B*, and *C*, using a Bonferroni correction. For ART, we use the ```art.con``` function.
@@
 We observe again a significant divergence between the results of ART and the results of the other methods.
+
+### Statistical analysis of perceived efficiency
+We conduct an analysis with PAR, RNK, ART, and INT.
+```{r, warning=FALSE}
+df$Perceived_Efficiency <- as.integer(df$Perceived_Efficiency)
+
+mpar <- lmer(Perceived_Efficiency ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - no transformation
+mart <- art(Perceived_Efficiency ~ Difficulty*Technique + (1|Participant), data=df) # ARTool
+mrnk <- lmer(rank(Perceived_Efficiency) ~ Difficulty*Technique + (1|Participant), data=df) # Rank transformation 
+mint <- lmer(int(Perceived_Efficiency) ~ Difficulty*Technique + (1|Participant), data=df) # Inverse Transform transformation
+```
+
+We report the results of our analysis.
+```{r, warning=FALSE}
+printAnalysis(2, "ART", mart)
+printAnalysis(3, "Rank Transformation", mrnk)
+printAnalysis(4, "Inverse Normal Transformation", mint)
+printAnalysis(5, "No Transformation", mpar)
+```
+
+ART is the only method to detect an main effect for both Difficulty ($p=.0011$, partial $\eta^2 = .12$, $95\%$ CI $[.03, 1.0]$) and Technique ($p=.00016$, partial $\eta^2 = .13$, $95\%$ CI $[.05, 1.0]$).
+Its *p-*values are substantially lower than the *p-*values of the other methods, and its effect size estimates are surprisingly large.

### 2025-05-25 text/20-intro.qmd
corrections in intro and background

--- 2025-05-24 text/20-intro.qmd
@@
 
-Suppose an HCI researcher conducts an experiment to compare the performance of three user interface techniques (A, B, and C) that help users complete image editing tasks of four different difficulty levels.
+Suppose a team of HCI researchers conduct an experiment to compare the performance of three user interface techniques (A, B, and C) that help users complete image editing tasks of four different difficulty levels.
 The experiment follows a fully balanced $4 \times 3$ repeated-measures factorial design, where each participant (N = 12) performs 12 tasks in a unique order.
-The researcher measures the time that it takes participants to complete each task.
+The researchers measure the time that it takes participants to complete each task.
+In addition, the participants rate their perceived efficiency completing each task, using an ordinal scale with five levels: (1) *inefficient*, (2) *somewhat inefficient*, (3) *neutral*, (4) *somewhat efficient*, and (5) *efficient*.
+
 The following table presents the experimental results:
@@
 ```{r, echo=FALSE, warning=FALSE}
-df <- read.csv("example_data.csv", sep=",", header=TRUE, strip.white=TRUE)
+df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)
 kbl(df) %>% kable_paper(position = "center") %>% scroll_box(width = "740px", height = "170px")
 ``` 
-Example dataset: Time (in minutes) spent by 12 participants for four difficulty levels and three user interface techniques. Scroll down to see the full results.
+Example dataset: Time (in minutes) spent by 12 participants and their perceived efficiency (ordinal scale) for four difficulty levels and three user interface techniques. Scroll down for full results.
 :::
@@
 The experiment is hypothetical but has similarities with real-world experiments, e.g., see the experiments of @Fruchard:2023.
-Time performances have been randomly sampled from a population in which: (1) *Difficulty* has a large effect; (2) *Technique* has no effect; and (3) there is no interaction effect between the two factors.
+
+**Time performance.** Time measurements have been randomly sampled from a population in which: (1) *Difficulty* has a large main effect; (2) *Technique* has no main effect; and (3) there is no interaction effect between the two factors.
 To generate time values, we drew samples from a log-normal distribution.
@@
         ylim(0, 3) + 
-        geom_jitter(shape=20, position=position_jitter(0.1)) +
+        geom_jitter(shape=20, position=position_jitter(0.1, 0)) +
         theme_bw() + theme(legend.position = "none") + scale_fill_manual(values=cbPalette))
@@
 ```
-Boxplots summarizing the results of our illustrative example. Dots represent the median time performance of each individual participant. 
+Boxplots summarizing time performance results for our illustrative example. Dots represent the median time for each individual participant. 
 :::
@@
 We observe that ART exaggerates both the effect of *Technique* and its interaction with *Difficulty*.
+This example demonstrates a more general problem with ART’s alignment mechanism under long-tailed distributions.
+ART tends to confound effects: a strong effect on a single factor often leads the method to detect spurious effects on other factors or to identify non-existent interactions.
+
+**Perceived efficiency.**  The perceived efficiency ratings were randomly sampled from a population with no main effect on any of the two factors and no interaction effect.
+@fig-example-ordinal shows that participants' ratings are concentrated at the higher end of the scale but with no clear trends.
+
+::: {#fig-example-ordinal}
+```{r, echo=FALSE, warning=FALSE, fig.height=2.7, fig.width=8}
+df$Perceived_Efficiency <- ordered(df$Perceived_Efficiency, levels = c("inefficient", "somewhat inefficient", "neutral", "somewhat efficient", "efficient"))
+
+p3 <- (df %>%  group_by(Participant,Technique) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop") %>%   
+        ggplot(aes(x = Technique, y = Perceived_Efficiency, fill = Technique)) + 
+        geom_boxplot(outlier.shape = NA) +
+        ylab("Perceived Efficiency") +
+        scale_y_continuous(breaks = 1:5, limits = c(1,5), labels=levels(df$Perceived_Efficiency)) +
+        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
+        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette) +  
+        theme(axis.text.y=element_text(angle=50, hjust=0.7))
+
+#diffPalette <- c("#f9f9f9", "#e2e2e2", "#b9b9b9", "#999999")
+diffPalette <- c("#f9f9fd", "#e2e6fa", "#bacafb", "#99b9f9")
+p4 <- (df %>%  group_by(Participant,Difficulty) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop") %>%   
+        ggplot(aes(x = Difficulty, y = Perceived_Efficiency, fill = Difficulty)) + 
+        geom_boxplot(outlier.shape = NA) + ylim(1,5)+
+        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
+        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=diffPalette) +  
+        theme(axis.title.y=element_blank(), axis.text.y=element_blank())
+
+grid.arrange(p3, p4, nrow = 1, widths = c(3, 3))
+```
+Boxplots summarizing the results on efficiency as perceived by participants. Circles represent the mean rating of each individual participant. 
+:::
+
+For our analysis, we compare the results of PAR, ART, RNK, and INT, since LOG is not relevant for this type of data.
+We also report the results of the Friedman test [@Friedman] for each factor, where we first aggregate the data by taking the mean over the other factor.
+Results are as follows:
+
+|        | PAR  | ART | RNK | INT | Friedman Test |
+|--------|------|-----|-----|-----|-----|
+| Difficulty  | $.93$ | $.0011$  |  $.86$ | $.88$ | $.83$ |
+| Technique   | $.18$ | $.00016$  | $.32$ | $.23$ | $.24$ |
+| Difficulty $\times$ Technique | $.27$ | $.10$ | $.33$ | $.34$ | |
+: *p*-values for main and interaction effects {.sm}
+
+ART's *p*-values for the two main effects are surprisingly low, while no other methods shows any substantial evidence for such effects. We also presents the methods' effect size estimates:
+
+|        | PAR  | ART | RNK | INT |
+|--------|------|-----|-----|-----|
+| Difficulty  | $.004\ [.00, 1.0]$  | $.12\ [.03, 1.0]$  |  $.006\ [.00, 1.0]$ | $.005\ [.00, 1.0]$ |
+| Technique   | $.03\ [.00, 1.0]$ | $.13\ [.05, 1.0]$  | $.02\ [.00, 1.0]$ | $.02\ [.00, 1.0]$ |
+| Difficulty $\times$ Technique | $.06\ [.00, 1.0]$ | $.04\ [.00, 1.0]$  | $.05\ [.00, 1.0]$| $.05\ [.00, 1.0]$|
+: partial $\eta^2$ and its 95\% confidence interval {.sm}
+
+The discrepancies between ART's estimates for the two main effects and those of the other methods are large.
+As we demonstrate in this paper, such discrepancies stem from ART's inability to handle discrete distributions.
+The issue becomes more pronounced with larger sample sizes.
+@luepsen:2017 previously warned researchers about this problem, but his warnings have largely been ignored.
 
 ### Overview
-The above example does not capture a rare phenomenon.
-We will show that ART's error inflation is systematic for a range of distributions that deviate from normality, both continuous and discrete.
-We will explain how the problem emerges.
-We will also see that ART often performs worse than simpler methods that ART is widely considered to repair or improve, such as the pure rank transformation or no transformation at all.
+The above example does not reflect a rare phenomenon. We will show that ART's error inflation is systematic for a range of distributions that deviate from normality, both continuous and discrete.
 

### 2025-05-25 text/30-background.qmd
corrections in intro and background

@@
 ## Background {#background}
-In this section, we provide background on the history, scope, construction, and use of rank transformation methods, with a particular focus on ART.
+We provide background on the history, scope, construction, and use of rank transformation methods, with a particular focus on ART.
 Additionally, we summarize the results of previous evaluation studies and position our own work within this context.
@@
 ### Rank transformations
-Unfortunately, the scope of nonparametric tests is limited, as they only support simple experimental designs with a single independent variable.
+ The scope of nonparametric tests is limited, as they only support simple experimental designs with a single independent variable.
 Rank transformations aim to address this limitation by bridging the gap between nonparametric statistics and ANOVA.
@@
 
-**The aligned rank transform.** Given such negative results, many researchers turned to aligned (or adjusted) rank transformations.
-@Sawilowsky:1990 discusses several variations of aligned rank-based transformations (ART) and tests, while @higgins:1990 detail the ART method that we evaluate in this paper for two-way designs.
-@wobbrock:2011 generalize it to more than two factors, while 10 years later, @elkin:2021 show how to apply it to multifactor contrast tests.
+**The aligned rank transform.** In response to these negative results, many researchers turned to aligned (or adjusted) rank transformations.
+@Sawilowsky:1990 discusses several variations of aligned rank-based transformations (ART) and tests, while @higgins:1990 describe the specific ART method that we evaluate in this paper for two-way designs.
+@wobbrock:2011 generalize it to more than two factors, and a decade later, @elkin:2021 show how to apply it to multifactor contrast tests.
 @fig-ART-explained explains the general method for a design with two factors, $A$ and $B$.
@@
 
-The key intuition behind the transform is that responses are ranked after *aligning* them, such that effects of interest are separated from other effects.
-This means that for each effect of interest, either main or interaction, we produce a separate ranking.
-This is clearly demonstrated through the example dataset in @fig-ART-explained (a), where each aligned ranking ($ART_A$, $ART_B$, and $ART_{AB}$) is for testing a different effect ($A$, $B$, and $A \times B$, respectively).
+The key intuition behind the transform is that responses are ranked after they are *aligned*, such that effects of interest are separated from other effects.
+This implies that for each effect, either main or interaction, a separate ranking is produced.
+This is clearly illustrated in the example dataset shown in @fig-ART-explained (a), where each aligned ranking ($ART_A$, $ART_B$, and $ART_{AB}$) is for testing a different effect ($A$, $B$, and $A \times B$, respectively).
 
-Notice that even for this very small dataset ($n = 2$), these three rankings are distinct from each other, as well as very different from the ranking of the simple rank transform.
-Interestingly, equal responses (the two $0.3$ in our example dataset) may be assigned to very different ranks, while very different responses (such as $0.3$ and $3.0$ in our dataset) can be assigned to the same rank.
-Furthermore, differences between the two groups $a_1$ and $a_2$ within factor $A$ are significantly exaggerated by ART (see $ART_A$ ranks).
-We will later see that this behavior can lead to unstable results and conclusions, and is the source of the main problems of the method.
-For example, if we run a repeated-measures ANOVA on these ranks, we obtain a *p*-value $= .044$ for the effect of $A$.
-For comparison, if we run repeated-measures ANOVA on the ranks of simple rank transformation, we obtain a *p*-value $= .46$, while if we use a log-transformation that is more appropriate for these data, we obtain a *p*-value $= .85$.
+Notice that even for this very small dataset ($n = 2$), the three ART rankings are distinct from one another, and also markedly different from the ranking produced by a simple rank transformation.
+Interestingly, identical responses (e.g., the two $0.3$ in our example dataset) may be assigned very different ranks, while responses that differ greatly (e.g., $0.3$ and $3.0$) can be assigned the same rank.
+Furthermore, ART significantly exaggerates the differences between the two groups $a_1$ and $a_2$ within factor $A$ (as seen in the $ART_A$ ranks).
+
+As we will show later, this behavior can lead to unstable results and misleading conclusions, and it represents a central flaw of the method.
+For example, running a repeated-measures ANOVA on these ART ranks yields a *p*-value of $.044$ for the effect of $A$.
+In contrast, using a repeated-measures ANOVA on the ranks of simple rank transformation produces a *p*-value of $.46$, while applying a log-transformation --- more appropriate for this type of data --- yields a *p*-value of $.85$.
 
@@
 
-This is not the only alignment technique found in the research literature.
-@Sawilowsky:1990 suggests that at least for balanced designs, interactions could be removed when aligning main effects, in the same way main effects are removed when aligning interactions.
+This is not the only alignment technique discussed in the literature.
+@Sawilowsky:1990 suggests that, at least for balanced designs, interactions could also be removed when aligning main effects, in the same way main effects are removed when aligning interactions.
 This approach is also taken by @Leys:2010, who derived a common ranking for both main effects after subtracting the interaction term.
-We do not test these alternative alignment methods in this paper since to the best of our knowledge, they are not commonly used.
+We do not evaluate these alternative alignment methods in this paper, as they are, to the best of our knowledge, not commonly used in practice.
 
 **The inverse normal transform.** 
-A third transformation method that we evaluate is the rank-based inverse normal transformation (INT).
-INT is more than 70 years old [@VanDerWaerden:1952] and appears in several variations [@Beasley:2009;@Solomon2009].
+A third transformation method we evaluate is the rank-based inverse normal transformation (INT).
+INT has been in use for over 70 years [@VanDerWaerden:1952] and exists in several variations [@Beasley:2009;@Solomon2009].
 Its general formulation is as follows:
@@
 
-**Other non-parametric rank-based methods.** There are other rank-based statistical tests that deal with interactions, the ANOVA-type statistic (ATS) [@Brunner_ATS:2001] being the most representative one.
-@Kaptein:2010 introduced this method to the HCI community, advocating its application in the analysis of Likert-type data as a viable alternative to parametric ANOVA.
+**Other non-parametric rank-based methods.** Several other rank-based statistical tests handle interactions, with the ANOVA-type statistic (ATS) [@Brunner_ATS:2001] being the most representative one.
+@Kaptein:2010 introduced this method to the HCI community, advocating its use for analyzing Likert-type data as a viable alternative to parametric ANOVA.
 In addition to ATS, Lüpsen [-@luepsen:2017; -@luepsen:2018; -@luepsen:2023] investigated several other multifactorial nonparametric methods.
@@
 The former is based on INT, but instead of using F-tests on the transformed values as part of ANOVA, it computes $\chi^2$ ratios over sums of squares.
-These two methods are not widely available.
-Their implementation in R can be downloaded from Lüpsen's web page [@luepsen_R].
+These two methods are not widely available, but implementations in R can be downloaded from Lüpsen's website [@luepsen_R].
 
 ### Experimental evaluations
-Previous studies have assessed ART and related procedures through various Monte Carlo experiments, yielding conflicting results and conclusions.
+Previous studies have evaluated ART and related procedures using various Monte Carlo simulations, often producing conflicting results and conclusions.
 
 **Results in support of ART**.
-The outcomes from numerous experiments conducted in the 80s and 90s suggested that ART was robust in testing interaction effects.
+A number of experiments conducted during the 80s and 90s suggested that ART is robust for testing interaction effects.
 Noteworthy instances include studies, such as those by @Salter:1993, which compared the method to parametric ANOVA.

### 2025-05-31 text/20-background.qmd
starting new section

--- 2025-05-25 text/30-background.qmd
@@
 
-Another interesting finding of @luepsen:2018 was that the simple rank transformation *"appeared not as bad as it is often described"* [@luepsen:2018], outperforming ART in certain configurations, such as discrete and skewed distributions, or unequal variances.
+Another notable finding by @luepsen:2018 was that the simple rank transformation *"appeared not as bad as it is often described"* [@luepsen:2018], outperforming ART in many scenarios, such as discrete and skewed distributions, or distributions with unequal variances.
 These results are in full contradiction with the findings of @Mansouri:1995.

### 2025-05-31 text/30-assumptions.qmd
starting new section

## Re-examining ART's assumptions {#assumptions}
Although ART is commonly thought as a nonparametric method, its construction relies on the assumption of a linear model with very specific characteristics. For a two-way experimental design, @higgins:1990 assume the following model for responses:

$Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \epsilon_{ijk}$

where $\mu$ is the grand mean, $\alpha_i$ is the main effect of the first factor for level $i$, $\beta_j$ is the main effect of the second factor for level $j$, and $(\alpha \beta)_{ij}$ is the interaction effect. The term $\epsilon_{ijk}$ is the experimental error effect (where $k = 1, ..., n$ is the subject number) and is assumed to have a zero mean and a common standard deviation. @Mansouri:1995 provide a proof of the limiting distributional properties of ART for interactions, where they assume the same model structure, as well as continuous distributions.

### 2025-05-31 text/abstract.qmd
starting new section

## Abstract {.unnumbered}

###### Introduction
The Aligned Rank Transform (ART) is a popular statistical method, commonly used when the statistical assumptions of parametric procedures, such as normality assumptions, are not met. Its popularity stems from several Monte Carlo experiments demonstrating that the method is robust and powerful. However, contrary experimental findings suggest that ART may be less robust compared to simpler rank transformation methods, failing to adequately control Type I error rates across various scenarios, including discrete data, skewed distributions, and distributions with unequal variances. These results have not gained attention within the Human-Computer Interaction community, where ART is routinely used.

###### Objectives
This research aims to clarify the contradictory findings from previous studies by reevaluating the robustness and effectiveness of ART. Its complementary goal is to investigate the performance of alternative nonparametric methods.

###### Methods
We conducted a series of Monte Carlo experiments to systematically evaluate ART across various data distributions and experimental designs. Our comparative analysis includes two alternative transformation methods: the simple rank transformation and the inverse normal transformation, along with parametric ANOVA without transformation. Additionally, we present supplementary experiments involving other competitive nonparametric methods in the [appendix](appendix.html).

Our experimental methodology builds upon a latent variable modeling approach, enabling us to simulate main and interaction effects across diverse distributions and address interpretation issues related to these effects. In addition to Monte Carlo experiments, we explore the behavior of ART through illustrative examples and reanalyze data from previous user studies.

###### Results
Our results indicate that ART systematically confounds effects across all the experimental designs we examined. Specifically, an effect on one factor may cause ART to incorrectly detect effects on other factors or their interactions. We show that the method fails under various conditions, including skewed distributions, binomial distributions, Likert-type data with five or seven ordinal levels, distributions with unequal variances, and experimental designs with missing data.

Surprisingly, ART operates as expected only under normal distributions with equal variances. In all other scenarios, it performs less effectively than simpler rank transformations and, in many situations, such as with ordinal data, it is significantly worse than parametric ANOVA. Our analysis of real-world data further demonstrates that using ART often leads to incorrect conclusions. Conversely, the inverse normal transformation emerges as the most versatile and robust nonparametric method, although we identify situations in which this method can also inflate errors or distort effects.

In addition, our results show that testing interactions in the presence of multiple main effects is challenging for all methods. We raise awareness about previous warnings regarding the interpretation of *removable interactions*.

###### Conclusion
We conclude that ART is a flawed statistical method and should be abandoned. We recommend prioritizing parametric methods, including linear models and their generalizations. While we propose the inverse normal transformation as a generic nonparametric alternative, we caution researchers about risks and limitations of rank transformation methods.

###### Materials
We provide [supplementary materials](https://github.com/journalovi/2024-tsandilas-ranktransforms/tree/main/supplementary) with experimental code, raw experimental results, and detailed analyses of examples and case studies presented throughout the article. Source files of our article, appendix, and supplementary materials can also be found on OSF at [https://osf.io/hs7j8](https://osf.io/hs7j8). 
:::

::: {.callout-note appearance="simple" icon=false collapse=true}
## Materials, Authorship, License, Conflicts

###### Research materials

We provide [supplementary materials](https://github.com/journalovi/2024-tsandilas-ranktransforms/tree/main/supplementary) with experimental code, raw experimental results, and detailed analyses of examples and case studies presented throughout the article. Source files of our article, appendix, and supplementary materials can also be found on OSF at [https://osf.io/hs7j8](https://osf.io/hs7j8).

###### Authorship

**Theophanis Tsandilas:** Conceptualization, Methodology, Resources, Software, Formal Analysis, Investigation, Validation, Data Curation, Vizualization, Writing - Original Draft, Writing - Review & Editing.
**Géry Casiez:** Conceptualization, Investigation, Writing - Review & Editing.

###### License

This work is licensed under a [Creative Commons Attribution 4.0 International License](https://creativecommons.org/licenses/by/4.0/).

###### Conflicts of interest

The authors declare that there are no competing interests.

:::

### 2025-06-06 text/20-background.qmd
Added section on ART's assumptions....

@@
 
-Notice that even for this very small dataset ($n = 2$), the three ART rankings are distinct from one another, and also markedly different from the ranking produced by a simple rank transformation.
-Interestingly, identical responses (e.g., the two $0.3$ in our example dataset) may be assigned very different ranks, while responses that differ greatly (e.g., $0.3$ and $3.0$) can be assigned the same rank.
-Furthermore, ART significantly exaggerates the differences between the two groups $a_1$ and $a_2$ within factor $A$ (as seen in the $ART_A$ ranks).
+Even in this very small dataset ($n = 2$), the three ART rankings are distinct from one another and also differ significantly from the ranking produced by a simple rank transformation.
+Interestingly, identical responses (e.g., the two values of $0.3$) can be assigned very different ranks, while responses that differ substantially (e.g., $0.3$ and $3.0$) may receive the same rank.
+More surprisingly, larger values can receive lower ranks —-- revealing that ART is a non-monotonic transformation.
+For instance, for subject $s_1$ under condition $b_1$, the responses are $0.4$ and $0.8$ for $a_1$ and $a_2$, yet their respective $ART_A$ ranks are $7.5$ and $3.0$.
+In this case, ART significantly distorts the actual difference between the two groups, $a_1$ and $a_2$.
 
@@
 In addition, they found that ART exhibits a higher power than the t-test.
-We emphasize, however, that the experimental methodology of @elkin:2021 did not allow for detecting the influence of a factor's main effect on the effect of a different factor or their interaction.
 
@@
 Again, the Type I error rates of both ART and ART+INT reached very high levels, but error rates were especially pronounced when the number of discrete levels became small and the sample size increased.
-The author provides a detailed analysis of why ART fails in these cases, summarized as follows:
-
-> *"There is an explanation for the increase of the type I error rate when the number of distinct values gets smaller or the sample size larger: due to the subtraction of the other effects --- a linear combination of the means --- from the observed values even tiny differences between the means lead to large differences in the ranking"* [@luepsen:2017].
-
 Given these results, the author's conclusion was the following:

### 2025-06-06 text/30-assumptions.qmd
Added section on ART's assumptions....

@@
-## Re-examining ART's assumptions {#assumptions}
-Although ART is commonly thought as a nonparametric method, its construction relies on the assumption of a linear model with very specific characteristics.
-For a two-way experimental design, @higgins:1990 assume the following model for responses:
+## Revisiting ART's distributional assumptions {#assumptions}
+Although ART is often regarded as a nonparametric method, its alignment mechanism relies on a set of strict assumptions.
+For a two-way experimental design with independent observation, @higgins:1990 assume the following model for responses:
 
-$Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \epsilon_{ijk}$
+$$ 
+Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \epsilon_{ijk}
+$${#eq-higgins-model}
 
-where $\mu$ is the grand mean, $\alpha_i$ is the main effect of the first factor for level $i$, $\beta_j$ is the main effect of the second factor for level $j$, and $(\alpha \beta)_{ij}$ is the interaction effect.
-The term $\epsilon_{ijk}$ is the experimental error effect (where $k = 1, ..., n$ is the subject number) and is assumed to have a zero mean and a common standard deviation.
-@Mansouri:1995 provide a proof of the limiting distributional properties of ART for interactions, where they assume the same model structure, as well as continuous distributions.
+where $\mu$ is the grand mean, $\alpha_i$ is the main effect of the first factor for level $i$, $\beta_j$ is the main effect of the second factor for level $j$, $(\alpha \beta)_{ij}$ is the interaction effect, and $\epsilon_{ijk}$ is the residual error term, assumed to have a common variance.
+The index $k = 1, ..., n$ denotes the subject number.
+
+This formulation makes several key assumptions:
+
+1. A linear relationship between effects and responses;
+2. A common variance among all experimental conditions; and 
+3. Continuous responses.
+
+The proof by @Mansouri:1995 assessing the limiting distributional properties of ART for interaction effects is based on these same assumptions.
+
+It is worth noting that ANOVA relies on similar assumptions, with the added requirement that errors be normally distributed.
+One might therefore argue that ART imposes fewer assumptions than ANOVA.
+Additionally, since ART is rank-based, it may appear reasonable to expect some degree of robustness to violations of the linear model.
+
+Unfortunately, this line of reasoning is not correct.
+ART's assumptions pertain specifically to its alignment mechanism, which is unique to this method.
+The rank transformation is applied only after alignment, and the authors of ART have not provided any theoretical or empirical justification for their alignment method under non-linear models, non-equal variances, or discrete data.
+We explain the severity of these assumptions in more detail.
+
+<!--
+Thus, the critical question we must address is: how does ART perform when its core assumptions are violated?
+-->
+
+### Violations of the linearity assumption
+Suppose we studied image-editing tasks (as in our illustrative example), measuring the time participants needed to complete them for two difficulty levels: *easy* and *hard*.
+If we assume that observed times follow log-normal distributions, the model of @higgins:1990 will generate distributions as the ones in @fig-higgins-model.
+We observe that the distribution for the hard task is simply shifted to the right, while their shape conveniently remains identical.
+@Mansouri:1995 have shown that at least for interaction effects, ART remains robust under such distributions.
+However, how realistic are they?
+
+::: {#fig-higgins-model}
+```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
+library(plotly)
+
+palette <- c("#E69F00", "#009E73")
+
+logsd <- 0.5
+logm <- -0.3
+
+xs <- seq(0, 10, length.out = 1000)
+y1s <- dlnorm(xs, logm, logsd)
+y2s <- dlnorm(xs - 0.5, logm, logsd)
+
+med1 <- exp(logm)
+med2 <- exp(logm) + 0.5
+
+annot1 <- paste("Log-normal: meanlog = ", logm, ", sdlog = ", logsd)
+annot2 <- paste("Log-normal: meanlog = ", logm, ", sdlog = ", logsd)
+
+fig <- plot_ly() %>%  
+      add_lines(x = xs, y = y1s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot1, hoverinfo = 'text', line=list(color=palette[1])) %>%
+      add_lines(x = xs, y = y2s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot2, hoverinfo = 'text', line=list(color=palette[2])) %>%      
+      add_annotations(x = med1-0.06, y = max(y1s), text = "easy", xref = "x", yref = "y", showarrow = F, font = list(color = palette[1]), xanchor = 'left') %>% 
+      add_annotations(x = med2-0.06, y = max(y2s), text = "hard", xref = "x", yref = "y", showarrow = F, font = list(color = palette[2]), xanchor = 'left') %>%   
+      layout(
+        showlegend = FALSE,
+        xaxis = list(title = "Time (min)", range = c(0, 4.5), showgrid = F, showticks = T, ticks="outside", zeroline = F), 
+        yaxis = list(showgrid = F, showticklabels = F, showline = F, range=c(0, max(y1s) + .05), fixedrange=T)
+      ) %>% 
+	      config(displayModeBar = TRUE, scrollZoom = FALSE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale", "hoverclosest", 'hoverCompare'))  %>% layout(dragmode='pan')
+
+fig
+
+```
+The model of @higgins:1990 assumes that distributions have an identical shape across all conditions.
+Here, the log-normal distribution of the time measure is simply shifted to the right as the task difficulty increases.
+:::
+
+Heavy-tailed distributions, such as the log-normal distribution here, commonly arise in nature when measurements cannot be negative or fall below a certain threshold (Limpert et al., 2001), e.g., the minimum time needed to react to a visual stimulus.
+In most experimental research, however, this threshold does not shift across conditions while preserving the distribution’s shape.
+Instead, distributions are more likely to resemble the ones in @fig-realistic-model.
+
+::: {#fig-realistic-model}
+```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
+library(plotly)
+
+palette <- c("#E69F00", "#009E73")
+
+logsd <- 0.5
+logm1 <- -0.3
+logm2 <- 0.3
+
+xs <- seq(0, 10, length.out = 1000)
+y1s <- dlnorm(xs, logm1, logsd)
+y2s <- dlnorm(xs, logm2, logsd)
+
+med1 <- exp(logm1)
+med2 <- exp(logm2)
+
+annot1 <- paste("Log-normal: meanlog = ", logm1, ", sdlog = ", logsd)
+annot2 <- paste("Log-normal: meanlog = ", logm2, ", sdlog = ", logsd)
+
+fig <- plot_ly() %>%  
+      add_lines(x = xs, y = y1s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot1, hoverinfo = 'text', line=list(color=palette[1])) %>%
+      add_lines(x = xs, y = y2s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot2, hoverinfo = 'text', line=list(color=palette[2])) %>%      
+      add_annotations(x = med1, y = max(y1s), text = "easy", xref = "x", yref = "y", showarrow = F, font = list(color = palette[1]), xanchor = 'left') %>% 
+      add_annotations(x = med2, y = max(y2s), text = "hard", xref = "x", yref = "y", showarrow = F, font = list(color = palette[2]), xanchor = 'left') %>%   
+      layout(
+        showlegend = FALSE,
+        xaxis = list(title = "Time (min)", range = c(0, 4.5), showgrid = F, showticks = T, ticks="outside", zeroline = F), 
+        yaxis = list(showgrid = F, showticklabels = F, showline = F, range=c(0, max(y1s) + .05), fixedrange=T)
+      ) %>% 
+	      config(displayModeBar = TRUE, scrollZoom = FALSE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale", "hoverclosest", 'hoverCompare'))  %>% layout(dragmode='pan')
+
+fig
+
+```
+Log-normal distributions produced using a generalized linear model. As the mean of the distributions increases, their overall shape changes. 
+:::
+
+In these distributions, the mean for had tasks also increases, but this increase is not reflected as a simple global shift in the distribution.
+Instead, the overall shape of the distribution changes.
+The model for these distributions is structured as follows:
+
+$$ 
+log(Y_{ijk} - \theta) = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \epsilon_{ijk}
+$${#eq-generalized-model}
+
+where $\epsilon_{ijk}$ is normally distributed, and $\theta$ represents a threshold below which response values cannot occur.
+Unfortunately, ART's alignment mechanism is not designed to handle such distributions.
+The alignment calculations shown in @fig-ART-explained include various mean terms that are sensitive to distribution differences across the levels of factors, and --- as our results show --- they confounds effects.
+
+We acknowledge that @elkin:2021 diverged from the modeling approach described by @higgins:1990 and @Mansouri:1995 and used generalized linear models as in @eq-generalized-model to evaluate ART.
+However, their experiments failed to observe that ART confounds effect, simply because they assessed Type I error rates only under scenarios where main effects for all factors were null.
+
+### Heteroscedasticity issues
+@myers2012 explain that *"if the underlying distribution of the response variable is not normal, but is a continuous, skewed distribution, such as the lognormal, gamma, or Weibull distribution, we often find that the constant variance assumption is violated,* and in such cases, *"the variance is a function of the mean"* (Pages 54-55).
+This pattern frequently occurs in studies measuring task-completion times, whether the task is visual, motor, or cognitive.
+As tasks become more difficult and prolonged, variance tends to increase.
+Similarly, slower participants generally exhibit greater variance across tasks compared to faster, more practiced users.
+This suggests that ART's assumption of constant variance in skewed distributions is unrealistic.
+
+@Wagenmakers:2007 conducted an analysis of nine independent experiments to empircally investigate this pattern in response time distributions.
+They confirmed that standard deviations were proportional to the mean.
+The model we presented earlier (see @eq-generalized-model) is in line with these observations.
+Even when the standard deviation of the error term $\epsilon_{ijk}$ is constant, the standard deviation of observed log-normal distributions increases linearly with their mean.
+@fig-example (right) shows this pattern in our illustrative example, where the standard deviation of time responses increases proportionally with task difficulty.
+
+However, a question remains: Is ART’s failure in handling such models due solely to heteroscedasticity issues?
+For example, consider the two time distributions in @fig-homoscedastic-model, which --- despite having different shapes --- share the same standard deviation.
+Would ART perform correctly in this case?
+While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html) showing that as long as distribution shapes differ across levels, ART tends to confound effects.
+Heteroscedasticity may exacerbate the problem, but it is not its sole cause.
+
+::: {#fig-homoscedastic-model}
+```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
+library(plotly)
+
+palette <- c("#E69F00", "#009E73")
+
+sdlog <- function(mean, sd) {
+	sqrt(log(1 + sd^2/mean^2))	
+}
+
+meanlog <- function(mean, sd) {
+	log(mean^2/sqrt(sd^2 + mean^2))
+}
+
+xs <- seq(0, 10, length.out = 1000)
+
+logm1 <- meanlog(0.8, 0.4)
+logm2 <- meanlog(1.4, 0.4)
+
+logsd1 <- sdlog(0.8, 0.4)
+logsd2 <- sdlog(1.4, 0.4)
+
+y1s <- dlnorm(xs, logm1, logsd1)
+y2s <- dlnorm(xs, logm2, logsd2)
+
+med1 <- 0.68
+med2 <- 1.40
+
+annot1 <- paste("Log-normal: meanlog = ", round(logm1, 2), ", sdlog = ", round(logsd1, 2))
+annot2 <- paste("Log-normal: meanlog = ", round(logm2, 2), ", sdlog = ",  round(logsd1, 2))
+
+fig <- plot_ly() %>%  
+      add_lines(x = xs, y = y1s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot1, hoverinfo = 'text', line=list(color=palette[1])) %>%
+      add_lines(x = xs, y = y2s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot2, hoverinfo = 'text', line=list(color=palette[2])) %>%      
+      add_annotations(x = med1, y = max(y1s), text = "easy", xref = "x", yref = "y", showarrow = F, font = list(color = palette[1]), xanchor = 'left') %>% 
+      add_annotations(x = med2, y = max(y2s), text = "hard", xref = "x", yref = "y", showarrow = F, font = list(color = palette[2]), xanchor = 'left') %>%   
+      layout(
+        showlegend = FALSE,
+        xaxis = list(title = "Time (min)", range = c(0, 4.5), showgrid = F, showticks = T, ticks="outside", zeroline = F), 
+        yaxis = list(showgrid = F, showticklabels = F, showline = F, range=c(0, max(y1s) + .05), fixedrange=T)
+      ) %>% 
+	      config(displayModeBar = TRUE, scrollZoom = FALSE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale", "hoverclosest", 'hoverCompare'))  %>% layout(dragmode='pan')
+
+fig
+
+```
+Log-normal distributions with equal variances but different means. Notice that their shapes are distinct. 
+:::
+
+### Discrete distributions
+As discussed earlier, @luepsen:2017 found that ART frequently fails when reponses are discrete.
+The problem becomes more pronounced for variables with few discrete levels and as the sample size increases.
+The author provides a detailed analysis of why this happens, summarized as follows:
+
+> *"There is an explanation for the increase of the type I error rate when the number of distinct values gets smaller or the sample size larger: due to the subtraction of the other effects --- a linear combination of the means --- from the observed values even tiny differences between the means lead to large differences in the ranking"* [@luepsen:2017].
+
+@fig-art-binary illustrates this problem using a simple dataset with a binary variable.
+In our example, all but one value are zero.
+Notice how dramatically ART alters the ranks and exaggerates the difference between the two levels of $B$ ($b_1$ vs.
+$b_2$).
+This issue persists even when a large number of participants are added and there is only a single value of 1 among thousands of 0s, causing the method to produce arbitrarily low *p*-values.
+Then, a single value variation (e.g., turning this 1 to 0) will cause ART’s results to drastically change.
+
+::: {#fig-art-binary}
+```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=3.2, fig.width=8}
+library(gridExtra)
+#library(ggplot2)
+
+df <- read.csv("data-binary.csv", sep=",", header=TRUE, strip.white=TRUE)
+df$S <- factor(df$S)
+df$A <- factor(df$A)
+df$B <- factor(df$B)
+
+df$rnk <- rank(df$Y)
+
+m <- mean(df$Y)
+Ai <- aggregate(Y ~ A, data = df, mean)
+Bj <- aggregate(Y ~ B, data = df, mean) 
+AiBj <- aggregate(Y ~ A+B, data = df, mean) 
+cmean <- aggregate(Y ~ A+B, data = df, mean)
+
+library(dplyr)
+df <- df %>% group_by(A) %>% mutate(Ai = mean(Y)) %>% 
+  group_by(B) %>% mutate(Bj = mean(Y)) %>% 
+  group_by(A,B) %>% mutate(AiBj = mean(Y)) %>% mutate(Cell = mean(Y))
+
+df$m <- mean(df$Y)
+
+df$artA <- df$Y - df$Cell + df$Ai - df$m
+df$artB <- df$Y - df$Cell + df$Bj - df$m
+df$artAB <- df$Y - df$Cell + df$AiBj - df$Ai - df$Bj + df$m
+
+df$rnkA <- rank(round(df$artA, digits = 7))
+df$rnkB <- rank(round(df$artB, digits = 7))
+df$rnkAB <- rank(round(df$artAB, digits = 7))
+
+mytheme <- gridExtra::ttheme_default(
+    core = list(fg_params=list(cex = 2.0)),
+    colhead = list(fg_params=list(cex = 1.0)),
+    rowhead = list(fg_params=list(cex = 1.0)))
+
+myTable <- tableGrob(df[,1:4], theme = ttheme_minimal(base_size = 10) )
+
+diffPalette <- c("#e2e6fa", "#99b9f9")
+p1 <- (df %>%   
+        ggplot(aes(x = B, y = rnk, fill = B)) + ggtitle("RNK") +
+        geom_boxplot(outlier.shape = NA) +
+        geom_jitter(shape=1, position=position_jitter(0.3,0), size = 2) +
+        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) + scale_fill_manual(values=diffPalette) +  
+        ylab("ranking") + scale_y_discrete(limits=as.character(1:nrow(df)))
+
+p2 <- (df %>%   
+        ggplot(aes(x = B, y = rnkB, fill = B)) + ggtitle("ART_B") +
+        geom_boxplot(outlier.shape = NA) + 
+        geom_jitter(shape=1, position=position_jitter(0.3,0), size = 2) +
+        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) + scale_fill_manual(values=diffPalette) +  
+        theme(axis.title.y=element_blank(), axis.text.y=element_blank()) + scale_y_discrete(limits=as.character(1:nrow(df)))
+
+grid.arrange(myTable, p1, p2, nrow = 1)
+```
+Results from an experiment following a $3 \times 2$ design, where the response variable $Y$ is binary and indicates the presence of an error.
+The plots show the rankings produced by the simple rank transformation (RNK) and by ART for the second factor.
+The solid lines represent group medians.
+:::

### 2025-06-13 text/10-intro.qmd
updates in Section 3... still ongoing

--- 2025-05-25 text/20-intro.qmd
@@
 The Aligned Rank Transform or ART [@higgins:1990; @Salter:1993; @wobbrock:2011] addresses this problem by bridging the gap between nonparametric tests and ANOVAs.
-Its popularity in HCI research has surged, facilitated in part by the ARTool toolkit [@wobbrock:2011; @artool], which simplifies the use of the method.
+Its popularity in HCI research has surged, in part due to the ARTool toolkit [@wobbrock:2011; @artool], which makes the method easy to use.
+@Oppenlaender:2025 identify the paper by @wobbrock:2011 as major milestone in HCI research, ranking third in terms of citations in the ACM CHI Proceedings (1981-2024), behind Braun and Clark's [-@Braun:2006] article on thematic analysis and Hart and Staveland's [-@hart1988development] article on the NASA-TXL index.
+
+However, given the significant influence of such milestone methods on our community's research outcomes, they warrant a closer scrutiny.
+Our results suggest that ART's popularity step largely from an ignorance of the method's assumptions.
+Researchers often treat ART as a general-purpose nonparametric alternative when ANOVA's assumptions are violated, overlooking the fact that ART's alignment mechanism introduces stricter assumptions than ANOVA itself, including a linear relationship between the response variable and the effects, equal variances, and continuous responses.
 
@@
 
-Our goal is to clarify the severity of these issues and understand the potential risks of the method.
+This paper aims to clarify the severity of these issues and understand the potential risks of the method.
 We present the outcomes of a series of Monte Carlo experiments, following a distinctive simulation methodology grounded in latent variable modeling.
-This approach enables us to simulate effects consistently across a broad spectrum of distributions, both discrete and continuous, and allows us to address scaling issues in interpreting interactions.
+This approach enables us to simulate effects consistently across a broad spectrum of distributions, both discrete and continuous, and allows us to address issues in interpreting main and interaction effects.
 To ensure the clarity of our findings and facilitate their reproducibility, we divide the experimental process into multiple smaller experiments, where each experiment focuses on a distinct variable (e.g., sample size, experimental design, variance ratio) or measure (Type I error rate, power, precision of effect size estimates).
@@
 
-ART's *p*-values for the two main effects are surprisingly low, while no other methods shows any substantial evidence for such effects. We also presents the methods' effect size estimates:
+ART's *p*-values for the two main effects are surprisingly low, while no other method shows any substantial evidence for such effects. We also present the methods' effect size estimates:
 
@@
 It also provides a summary of previous experimental results regarding the robustness of ART and other related rank-based procedures, along with a survey of recent studies using ART.
-[Section 3](#interpretation) investigates issues regarding effect interpretation and introduce our distribution simulation approach.
-[Section 4](#methodology) outlines our experimental methodology, while [Section 5](#findings) presents our findings.
-[Section 6](#casestudies) revisits results of previous studies employing ART and  illustrates how its application can lead to erroneous conclusions.
-[Section 7](#recommendations) offers recommendations for researchers.
-Finally, [Section 8](#conclusions) concludes the paper.
+[Section 3](#assumptions) clarifies the method's distributional assumptions.
+[Section 4](#interpretation) investigates issues regarding effect interpretation and introduce our distribution simulation approach.
+[Section 5](#methodology) outlines our experimental methodology, while [Section 6](#findings) presents our findings.
+[Section 7](#casestudies) revisits results of previous studies employing ART and  illustrates how its application can lead to erroneous conclusions.
+[Section 8](#recommendations) offers recommendations for researchers.
+Finally, [Section 9](#conclusions) concludes the paper.
 

### 2025-06-13 text/20-background.qmd
updates in Section 3... still ongoing

@@
 In particular, we adopt a latent variable modeling approach that establishes a unified framework for data generation across all distributions.
-This approach allows us to demonstrate the shortcomings of rank-based approaches in effectively addressing interactions through the lens of removable interactions [@Loftus:1978].
+This approach allows us to identify interpretation issues regarding the definition of the null hypothesis, especially for interaction effects, and better explain when and why each method fails.
 Moreover, it facilitates the assessment of the methods through more suitable generation procedures for ordinal data, as suggested by @Liddell:2018.
+Our simulation strategy is similar to that of @elkin:2021 but employs a simpler method for generating data from a latent parameter space.
+In addition, we evaluate critical scenarios that @elkin:2021 did not consider, particularly the influence of no-null effects and discrete data.
 
@@
 The appendix features additional results regarding the performance of the generalized van der Waerden test, as well as the generalized Kruskal-Wallis and Friedman tests.
-Our findings show that, at least in balanced designs, these methods suffer from low power issues without demonstrating any clear benefits compared to INT.
+Our findings show that, at least in balanced designs, these methods suffer from low power issues in certain situations without demonstrating any clear benefits compared to INT.
 

### 2025-06-13 text/30-assumptions.qmd
updates in Section 3... still ongoing

@@
 @Mansouri:1995 have shown that at least for interaction effects, ART remains robust under such distributions.
-However, how realistic are they?
+However, how realistic are these distributions?
 
@@
 Instead, the overall shape of the distribution changes.
-The model for these distributions is structured as follows:
+The model we used to generate these distributions is structured as follows:
 
@@
 where $\epsilon_{ijk}$ is normally distributed, and $\theta$ represents a threshold below which response values cannot occur.
-Unfortunately, ART's alignment mechanism is not designed to handle such distributions.
-The alignment calculations shown in @fig-ART-explained include various mean terms that are sensitive to distribution differences across the levels of factors, and --- as our results show --- they confounds effects.
+Unfortunately, ART's alignment mechanism is not designed to handle such cases.
+The alignment calculations shown in @fig-ART-explained include various mean terms that are sensitive to distribution differences across the levels of factors.
+In particular, the values of a random sample that appear in the tail of the right distribution can disproportionately influence the within-cell mean.
+This, in turn, leads to the exaggeration of all ranks within that cell, causing the method --- as our results demonstrate --- to confound effects.
 
-We acknowledge that @elkin:2021 diverged from the modeling approach described by @higgins:1990 and @Mansouri:1995 and used generalized linear models as in @eq-generalized-model to evaluate ART.
-However, their experiments failed to observe that ART confounds effect, simply because they assessed Type I error rates only under scenarios where main effects for all factors were null.
+We acknowledge that @elkin:2021 diverged from the modeling approach described by @higgins:1990 and @Mansouri:1995 and considered nonlinear models to evaluate ART.
+Yet, their experiments did not reveal that ART confounds effects, simply because they assessed Type I error rates only in scenarios where all main effects were null.
 
@@
 Similarly, slower participants generally exhibit greater variance across tasks compared to faster, more practiced users.
-This suggests that ART's assumption of constant variance in skewed distributions is unrealistic.
+This suggests that ART's assumption of constant variance in skewed distributions is largely unrealistic.
 
@@
 While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html) showing that as long as distribution shapes differ across levels, ART tends to confound effects.
-Heteroscedasticity may exacerbate the problem, but it is not its sole cause.
+Naturally, as the shapes become more similar, the problem diminishes.
+In conclusion, heteroscedasticity may exacerbate the problem, but it is not its sole cause.
 
@@
 Then, a single value variation (e.g., turning this 1 to 0) will cause ART’s results to drastically change.
+We will show that this unstable behavior makes ART inflate error rates even for variables with a larger range of discrete values.
 

### 2025-06-13 text/40-interpretingEffects.qmd
updates in Section 3... still ongoing

## Defining the null hypothesis of interest {#interpretation}
Before comparing different statistical methods, it is essential to assess whether they are actually comparable. If two methods are not designed to test the same null hypothesis, then direct comparisons between them can be misleading. In what follows, we clarify the interpretation of main and interaction effects and explain how we address potential issues in our analysis.

### Interpreting main effects
The traditional ANOVA is used to test differences between two or more means. However, nonparametric tests often target other population parameters. For example, the Wilcoxon sign-rank test is commonly described as a test of medians for paired samples [@McDonald:2014] and is used when population means are not of interest, e.g., when population distributions are skewed. The Mann-Whitney U and the Kruskal–Wallis tests are used, instead, to assess whether two or more independent samples come from the same population, or more technically, whether the mean ranks of the groups are the same. They can be only interpreted as tests of medians under the strict assumption that the population distributions of all groups have identical shapes and scales [@Divine:2018].

Defining the null hypothesis of interest of a rank transformation is more challenging. @conover:1981 show that the simple rank transformation procedure (RNK) is equivalent to the Mann-Whitney U and Kruskal–Wallis tests for independent samples. For paired samples, however, it results in a new test, which is different from the Wilcoxon sign-rank test and the Friedman test. Defining the null hypothesis of interest of ART is even more challenging because of the hybrid nature of the method. In particular, while ART is a rank-based transformation procedure, it aligns data with respect to means, where alignment is performed independently for each group.

**Dealing with interpretation issues.** To avoid such interpretation issues, we focus --- unless otherwise stated --- on effects that apply monotonic transformations to population distributions. This also ensures a monotonic relationship between different measures of central tendency such as medians and means (with the exception of the Cauchy distribution, where the mean is undefined). In other words, if a treatment increases the population mean, it will also increase the population median. We present an example in @fig-distributions. The figure shows two population distributions corresponding to the two intermediate levels of difficulty of our illustrative example (see @fig-example). We observe that the increased difficulty of the task translates both the population mean and the median to the right. In this case, we expect a statistical test to reject the null hypothesis, no matter whether it tests the population mean, the median, or the overall distribution shape.

::: {#fig-distributions}
```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
library(plotly)

palette <- c("#E69F00", "#009E73")

logsd <- 0.5
logm1 <- -0.3
logm2 <- 0.3

xs <- seq(0, 10, length.out = 1000)
y1s <- dlnorm(xs, logm1, logsd)
y2s <- dlnorm(xs, logm2, logsd)

m1 <- exp(logm1 + (logsd^2)/2)
m2 <- exp(logm2 + (logsd^2)/2)
med1 <- exp(logm1)
med2 <- exp(logm2)

anot_m1 <- list(x = m1, y = 0, text = "mean", xref = "x", yref = "y",
  showarrow = TRUE, arrowhead = 6, arrowsize = 1, arrowcolor = palette[1],
  ax = 0, ay = -40
)

anot_med1 <- list(x = med1, y = 0, text = "median", xref = "x", yref = "y", 
  showarrow = TRUE, arrowhead = 7, arrowsize = 1, arrowcolor = palette[1],
  ax = 0, ay = 30
)

anot_m2 <- list(x = m2, y = 0, text = "mean", xref = "x", yref = "y",
  showarrow = TRUE, arrowhead = 6, arrowsize = 1, arrowcolor = palette[2],
  ax = 0, ay = -43
)

anot_med2 <- list(x = med2, y = 0, text = "median", xref = "x", yref = "y", 
  showarrow = TRUE, arrowhead = 7, arrowsize = 1, arrowcolor = palette[2],
  ax = 0, ay = 30
)

annot1 <- paste("Log-normal: meanlog = ", logm1, ", sdlog = ", logsd)
annot2 <- paste("Log-normal: meanlog = ", logm2, ", sdlog = ", logsd)

fig <- plot_ly() %>%  
      add_lines(x = xs, y = y1s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot1, hoverinfo = 'text', line=list(color=palette[1])) %>%
      add_lines(x = xs, y = y2s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot2, hoverinfo = 'text', line=list(color=palette[2])) %>%      
      add_annotations(x = med1, y = max(y1s), text = "Difficulty Level 2", xref = "x", yref = "y", showarrow = F, font = list(color = palette[1]), xanchor = 'left') %>% 
      add_annotations(x = med2, y = max(y2s), text = "Difficulty Level 3", xref = "x", yref = "y", showarrow = F, font = list(color = palette[2]), xanchor = 'left') %>% 
      layout(annotations = anot_m1) %>% layout(annotations = anot_med1) %>%      
      layout(annotations = anot_m2) %>% layout(annotations = anot_med2) %>%    
      layout(
        showlegend = FALSE,
        xaxis = list(title = "Time (min)", range = c(0, 4.5), showgrid = F, showticks = T, ticks="outside", zeroline = F), 
        yaxis = list(showgrid = F, showticklabels = F, showline = F, range=c(0, max(y1s) + .05), fixedrange=T)
      ) %>% 
	      config(displayModeBar = TRUE, scrollZoom = FALSE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale", "hoverclosest", 'hoverCompare'))  %>% layout(dragmode='pan')

fig

```
Time distributions for two task populations with difficulty levels 2 and 3 (see @fig-example). 
:::

### Interpreting interaction effects
The ART procedure was proposed as an alternative to the rank transformation [@conover:1981] for testing interactions. As @higgins:1990 explained, the rank transformation is non-linear and, as a result, it changes the structure of interactions. Therefore, *"interaction may exist in the transformed data but not in the original data, or vice versa"* [@higgins:1990]. @fig-interactions-rank demonstrates the problem. In this example, the data have been sampled from perfectly normal distributions with equal variances. We observe that while no interaction effect appears in the original data (lines are parallel), the rank transformation deforms the trend. In particular, differences are more pronounced for the middle points of the three-level factor ("medium difficulty"). The figure also shows that the inverse normal transformation also deforms the interaction but to a lesser extent. Note that the problem emerges when the main effect is strong on all interacting factors.

::: {#fig-interactions-rank}
```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 7, warning=FALSE}
source("interactions_plot.R")
library("ARTool")

#library(plotly)

INT <- function(x){
	qnorm((rank(x) - 0.5)/length(x))
}

df <- read.csv("interactions_rank.csv", sep=",", header=TRUE, strip.white=TRUE)
df$Difficulty <- ordered(df$Difficulty, levels = c("easy", "medium", "hard"))

df_aggr <- aggregate(Time ~ Difficulty+Technique, data = df, mean)
df_rank <- aggregate(rank(Time) ~ Difficulty+Technique, data = df, mean)
df_int <- aggregate(INT(Time) ~ Difficulty+Technique, data = df, mean)
colnames(df_rank) <- c("Difficulty", "Technique", "Time")
colnames(df_int) <- c("Difficulty", "Technique", "Time")

fig1 <- createRankInteractionPlot(df_aggr) %>% layout( yaxis = list(hoverformat = '.2f'))
fig2 <- createRankInteractionPlot(df_rank, rnkscale = TRUE) %>% layout( yaxis = list(hoverformat = '.2f'))
fig3 <- createRankInteractionPlot(df_int) %>% layout( yaxis = list(title = 'INT(Time)', range = c(-2.1,2.1), hoverformat = '.2f'))

fig <- subplot(fig1, fig2, fig3, titleY = TRUE, titleX = TRUE, margin = 0.05, widths = c(0.32, 0.32, 0.32)) %>% 
	        layout(margin = list(l = 40, r = 0, b = 0, t = 50, pad = 0)) %>% layout(hovermode = 'x', dragmode='pan')

fig
```
Visualization of interaction effect for a 3 $\times$ 2 experimental design before and after applying the rank transformation and the inverse normal transformation (INT) on an [example dataset](interactions_rank.csv) (within-participants design, $n = 20$). All data points represent means.
:::

ART aims to correct this problem. However, non-linear transformations come into place in various ways in experimental designs [@Loftus:1978; @Wagenmakers:2012]. They can deform distributions, making the interpretation of observed effects especially challenging. Before presenting our experimental method, we discuss these problems and explain how our approach takes them into consideration.

**Removable interactions.** Let us take a different [dataset](removable_interactions.csv) from a fictional experiment (within-participants design with $n = 24$) that evaluates the performance of two techniques (*Tech A* and *Tech B*) under two task difficulty levels (*easy* vs. *hard*). The experiment, for example, could test a mobile typing task, where the levels of difficulty correspond to texts of different lengths (*short* vs. *long*) under two typing techniques (*with* vs. *without auto-completion*). We assume that the researchers measure two dependent variables: task-completion time and perceived performance, which is measured through a five-level ordinal scale (from "very quick" to "very slow"). In this example, the main effects of task difficulty and technique are large. It is less clear, however, whether there is also an interaction between the two factors.

@fig-interactions visualizes the means for each combination of the levels of the factors and highlights the possible interactions. Let us first concentrate on the first two plots that present results for the time measure. The trends in the left plot indicate an interaction effect, since the two lines seem to diverge as the task difficulty increases.

::: {#fig-interactions}
```{r, echo=FALSE, message=FALSE, fig.height=3.5, fig.width = 9, warning=FALSE}
df <- read.csv("removable_interactions.csv", sep=",", header=TRUE, strip.white=TRUE)
dftime <- aggregate(Time ~ Difficulty+Technique, data = df, mean)
dfpref <- aggregate(PerceivedPerformance ~ Difficulty+Technique, data = df, mean)

fig1 <- createInteractionPlot(dftime) %>% layout( yaxis = list(hoverformat = '.2f'))
fig2 <- createInteractionPlot(dftime, logscale = TRUE) %>% layout( yaxis = list(hoverformat = '.2f'))
fig3 <- createInteractionPlot(dfpref, likert = TRUE) %>% layout( yaxis = list(hoverformat = '.1f'))
#fig <- subplot(fig1, fig2, fig3, titleY = TRUE, titleX = TRUE, margin = 0.08, widths = c(0.32, 0.36, 0.32))

fig <- subplot(fig1, fig2, titleY = TRUE, titleX = TRUE, margin = 0.06) %>% subplot(fig3, titleY = TRUE, titleX = TRUE, margin = 0.07, widths = c(0.66, 0.33)) %>% layout(hovermode = 'x', dragmode='pan')

fig
```
The line charts visualize the effects of task difficulty (*easy* vs. *hard*) and technique (*Tech* A vs. *Tech B*) for two measures: task completion time (left and middle) and perceived performance (right). All data points represent group means. 
:::

But how meaningful is this interpretation of interaction? As we dicussed earlier, time measurements are often taken from skewed distributions where the variance is not constant. Therefore, large effects are harder to observe in quick tasks than in slow ones. However, such trends do not necessarily reveal any real interactions, because they are simply due to observations at different time scales. @fig-interactions (middle) visualizes the effects using a logarithmic scale. Notice that the lines in the plot are now almost parallel, suggesting no interaction effect.

The concept of *removable* or *uninterpretable* interactions, that is, interactions that disappear after applying a monotonic non-linear transformation, was introduced by @Loftus:1978.  Over three decades later, @Wagenmakers:2012 revisited this work and found that psychology researchers are largely unaware of the concept, drawing incorrect conclusions about psychological effects on the basis of meaningless interactions.

This issue also extends to data collected from questionnaires. The right plot in @fig-interactions shows results for perceived performance. Again, the line trends suggest an interaction effect. Unfortunately, the scale is ordinal, which means that distances between the five levels of the scale may not be perceived as equal by people. Furthermore, the scale is bounded, so the reason that the two lines are not parallel might be simply due to the absence of additional levels beyond the extreme "very slow" ranking. Concluding that there is a meaningful interaction here could be incorrect. @Liddell:2018 extensively discuss how ordinal scales deform interactions.

**Formal testing.** We now formally test the above interactions by using ANOVA with different transformation methods. Below, we present the *p*-value returned by each method for task-completion time:

| PAR  | LOG  | ART | RNK | INT |
|------|------|-----|-----|-----|
| $.023$ | $.67$ | $.00073$ | $.66$ | $.67$ |
: *p*-values for interaction effect on task-completion time {.sm}

We observe that *RNK* and *INT* lead to *p*-values very close to the *p*-value of *LOG*, which suggests a similar interpretation of interaction effects. In contrast, ART returns a very low *p*-value (lower than the *p*-value of the regular ANOVA), indicating a different interpretation.

We also test the interaction effect on the ordinal dependent variable:

| PAR  | ART | RNK | INT | ATS |
|------|-----|-----|-----|-----|
| $.0020$ | $.00075$ | $.0067$ | $.0037$ | $.0081$ |
: *p*-values for interaction effect on perceived performance {.sm}

Notice that we omit the log-transformation method (*LOG*), as it is not relevant in this context. Instead, we conduct an analysis with the nonparametric ATS method [@Brunner_ATS:2001], as implemented in the R package *nparLD* [@nparLD]. All *p*-values are low, suggesting that an interaction effect exists. However, if we conduct a more appropriate analysis using an ordered probit model [@Burkner:2019;@Christensen2023ordinal], we find no supportive evidence for such an effect (check our analysis in the supplementary material).

In conclusion, focusing solely on the issues with the rank tranformation illustrated in @fig-interactions-rank is akin to missing the forest for the trees. When parallel main effects are present, interpreting interactions can be challenging for all methods --- and ART offers no solution to this problem.

### Debate on the interpretation of interactions
We further discuss interpetation issues in the broader context of generalized liner models (GLMs).

GLMs are defined by a link function $g$ that connects a variable —- linearly defined by the predictors -— to a response variable $Y$, which follows an arbitrary distribution. For an experimental design with two factors, $x_1$ and $x_2$, the expected value (or mean) of $Y$ conditional on $x_1$ and $x_2$ is expressed as follows:

$$
E(Y|x_1, x_2) = g^{-1}(a_0 + a_1 x_1 + a_2 x_2 + a_{12}x1 x2)
$${#eq-glm}

where $g^{-1}$ is the inverse of the link function.

Statistical procedures based on GLMs define interaction in terms of the interaction coefficient $a_{12}$ in the linear component of the model. That is, the null hypothesis should be rejected when the coefficient is non-zero. However, there is no consensus among researchers on this approach. We clarify the debate and explain how we address it.

**Defining interactions on the scale of the response variable.** @Ai:2003 and, more recently, @McCabe:2022 argue that the interaction coefficient alone is not sufficient to describe the actual interaction trend on the natural scale of the response variable. They argue that an interaction effect should instead be defined as the observed change in the marginal effect of $x_1$ as a function of $x_2$. For continuous predictors, this can be formalized as the second-order cross partial derivative of the expected value (which is $\frac{\partial^2 E(Y|x_1, x_2)}{\partial x_1 \partial x_2}$), while for discrete predictors, it can be expressed using discrete differences between levels of the predictors.

Using this definition, the authors show that a zero interaction coefficient ($a_{12} = 0$) does not necessarily imply the absence of an interaction effect. In fact, for continuous predictors, they demonstrate that  when $a_{12} = 0$, the interaction effect is proportional to $a_1$ and $a_2$.

**Responding to @Ai:2003.** @Greene:2010 responds that the interpretation of @Ai:2003 *“produces generally uninformative and sometimes contradictory and misleading results”* and instead advocates for using the model’s coefficients as the basis for inference:

*“Statistical testing about the model specification is done at this step. Hypothesis tests are about model coefficients and about the structural aspects of the model specifications. Partial effects are neither coefficients nor elements of the specification of the model. They are implications of the specified and estimated model”* [@Greene:2010, Page 295].

Our position is fully aligned with Greene's argument. As our earlier example (see @fig-interactions) demonstrates, interactions assessed on the scale of the response variable can be misleading about the true underlying effects that generated the data. How meaningful is it to declare interaction effects that arise solely from parallel main effects and lack any theoretical interpretation? It is worth noting that @Ai:2003, as well as @McCabe:2022, overlook Loftus' [-@Loftus:1978] critique of uninterpretable interactions.

**Clarifying interpretation boundaries.** Interaction interpretation issues only arise when parallel main effects are present, that is, when the coefficients of all interacting factors are non-zero. But interpretation issues can also affect main effects. In particular, if the interaction coefficient $a_{12}$ is non-zero, we may observe a change in the marginal effect of a factor (e.g., $x_1$) despite the fact that its corresponding model coefficient (e.g., $a_1$) is zero.

Throughout the rest of the article, we make every effort to clarify when the interpretation of an effect is ambiguous. Because different methods may rely on different assumptions about the null hypothesis, we analyze such situations separately to avoid confusion. Importantly, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* and no interaction. Similarly, in the majority of our experiments demonstrating ART's deficient behavior, we avoid interpretation issues by varying the model coefficient of a single factor while keeping all other coefficients at zero.

### Our approach {#approach}
<!-- Our analysis shows that inference errors are not simply due to the lack of robustness of a statistical procedure. In the case of interaction effects, errors will also emerge when the procedure makes inference on the wrong scale. 
As @Wagenmakers:2012 explain, *"the dependent variable reflects merely the output of a latent psychological process",* and unfortunately, *"in most experimental paradigms the exact relationship between unobserved process and observed behavior is unknown ..."*

Ideally, a statistical procedure should lead to conclusions that capture the true effects on the latent variable of interest. But as we discussed above, this might not be the case for the rank transformation methods that we study. For example, all four methods (*PAR*, *RNK*, *ART*, and *INT*) suggest that task difficulty interacts with use of technique on perceived performance because they disregard the fact that the observed data are simply projections on a discrete ordinal scale. Our goal is to understand how ART and the other methods deal with such problems. 
-->
The responses in a generalized linear model can follow arbitrary distributions. A key challenge is how to simulate such distributions while maintaining control over the parameters of the linear submodel. We address this by adopting a latent variable modeling approach, in which a normally distributed latent variable is transformed into the observed response variable.

**Latent variables.** We assume here is a single latent variable $\mathcal{Y}$ that is distinct from the variable $Y$ we observe. For example, the latent variable may represent a population's performance potential, working memory capacity, the perceived utility of a new technology, or quality of life. For convenience, we assume that this latent variable is linear, continuous, and normally distributed. This is a common assumption in latent variable modeling, e.g., in diffusion-decision models that predict response time and error in two-choice decision tasks [@Ratcliff_diffusion:2008], and in ordinal models [@Liddell:2018]. This assumption allows us to define null hypotheses within a shared, normalized space, regardless of the scale of the observed data.

**Observed variables.** Then, the response variable $Y$ is derived from this latent variable through a monotonic transformation, thus $Y = \mathcal{T}(\mathcal{Y})$, where $\mathcal{T}(y_1) \le \mathcal{T}(y_2)$ if and only if $y_1 \le y_2$. A transformation for example occurs when study participants perform a selection task or repond to a Likert-scale item through a questionnaire. @fig-conversions shows how we transform normal distributions to log-normal, binomial, and ordinal scales.

::: {#fig-conversions}
```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
source("conversions_plot.R")

n <- 2000

xs <- c(rnorm(n, mean = 1.5, sd = 1), rnorm(n, mean = -1.5, sd = 1))

ys1 <- norm2lnorm(xs, meanlog = 0, sdlog = 1, mu = 0, sd = sd(xs))
plot1 <- plotConversion(xs, ys1, ylab = "Log-normal scale", ymax = 10)

ys2 <- norm2binom(xs, size = 10, prob=.1)
plot2 <- plotConversion(xs, ys2, ymin = - 0.3, ylab = "Binomial scale", as.density = F)

ys3 <- toOrdinal(xs, thresholds = c(-2.5, -1, 0, 2.5))
plot3 <- plotConversion(xs, ys3, ymin = 0.7, ymax = 5.3, ylab = "Ordinal scale", as.density = F)

fig <- cowplot::plot_grid(plot1, plot2, plot3, ncol= 3)

fig
```
Transformation of normal latent variables to other continuous or discrete scales: log-normal (left), binomial (middle), and ordinal (right).  
:::

To transform the latent variable to a ratio scale (e.g., a log-normal and binomial scale), we adopt the distribution conversion approach of *faux* v1.2.1 [@faux], an R package for experimental simulations. We first derive the cumulative density distribution of the latent variable. We then use the inverse quantile function of the target distribution to derive the observed variable. For example, in @fig-conversions (left), where we transform the latent variable to a log-normal scale with parameters $\mu = 0$ and $\sigma = 1$, we use the following R function:

```{r}
norm2lnorm <- function(x, meanlog = 0, sdlog = 1, mu = mean(x), sd = sd(x), ...) {
	 p <- pnorm(x, mu, sd) 
 	 qlnorm(p, meanlog, sdlog, ...) 
}
```
For the binomial scale of @fig-conversions (left), we use instead the inverse quantile function of the binomial distribution ```qbinom(p,size,prob)``` with parameters ```size = 10``` and ```prob = .1```, which respectively represent the number of Bernoulli trials and their success probability.

To transform the latent variable to an ordinal scale, we implement an ordered-probit model, as explained by @Liddell:2018. According to this model, we discretize the latent variable with thresholds that determine the ordinal levels of interest. For our example in @fig-conversions (right), we use as threshold the values $(-2.5, -1, 0, 2.5)$, defining an ordinal scale of five levels. Observe that these thresholds are not equidistant.

**Interpreting effects.** Our approach allows us to simulate main and interaction effects on the latent variable $\mathcal{Y}$ and observe how these effects manifest on the transformed variable $Y = \mathcal{T}(Y)$.

As discussed earlier, we make clear distinction between the following two scenarios:

1. The definition of the null hypothesis remains the same regardless of whether we evaluate effects on the latent variable $\mathcal{Y}$ or the response variable $Y$.

2. The definition of the null hypothesis changes with the scale —- a null effect on the latent variable $\mathcal{Y}$ may not remain null when evaluated on the response variable $Y$.

For the first scenarios, we ensure that the interaction coefficient of the linear model is zero (e.g., $a_{12} = 0$ for the model in @eq-glm) when assessing main effects. When assessing interaction effects, we vary a single only coefficient that corresponds to a predictor (e.g., either $a_1$ or $a_2$), while keeping the others zero.

<!--
As discussed earlier, how to infer main effects is straightforward since we only study monotonic transformations here. Specifically, if we observe a main effect on the observed variable $Y'$, we can also conclude that there is a main effect on the latent variable $Y$. Notice, however, that if the observed variable is bounded or takes only discrete values, a main effect on $Y$ may not be observable.

The interpretation of interactions is more challenging. Depending on how the statistical procedure we use deals with different scales, observations of interactions among two or more factors on the transformed variable $Y'$ may lead to incorrect conclusions about the presence of interactions on the latent space of $Y$. Such errors emerge when the main effect of all interacting factors on the latent (or observed) variable is non-zero. In our analysis, we count them as statistical inference errors (Type I and II errors). Nevertheless, we try to distinguish them from typical inference errors, for which interaction interpretation issues do not come into place.
-->

### 2025-06-22 sandbox.qmd
many changes

--- 2025-05-24 sandbox.qmd
@@
 
-Used for quick testing... eeee
+::: {.callout-important appearance="simple"}
+## Under Review {.unnumbered}
+This paper is [under review](https://www.journalovi.org/under-review.html) on the experimental track of
+the [Journal of Visualization and Interaction](https://www.journalovi.org/). See the 
+[reviewing process](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues).
+:::
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+# Useful libraries
+library(crosstalk)
+library(kableExtra)
+library(gridExtra)
+library(lmerTest)
+library(tidyverse)
+library(plotly)
+```
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+# Source code for reading data from experimental results and for plotting
+source("dataReaders.R")
+source("plotting.R")
+source("effects_plot.R")
+source("interactions_plot.R")
+```
+
+```{=html}
+<style>
+.math.inline .MathJax  {
+  font-size: 105% !important;
+}
+</style>
+```
+
+::: {#fig-eta-squared}
+```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
+data <- read.csv("data/effect-sizes-lognormal.csv", sep=",", header=TRUE, strip.white=TRUE)
+data$effectX2 <- factor(data$effectX2)
+
+ylab = "X1 - partial eta squared"
+xlab = "X2 - magnitude of main effect (a2)"
+
+margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
+cbPalette = c("#8888ff", "#888888", "#E69F00", "#009E73", "#FF5E00")
+cbPalette2 = c("#8888ff88", "#88888888", "#E69F0088", "#009E7388", "#FF5E0088")
+
+df <- data[data$effectX1 == 0,]
+
+fig1 <- plot_ly(df, x = ~effectX2, y = ~X1, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
+fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
+fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
+fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))
+
+fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
+  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
+	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab)) %>% layout(
+    hovermode = "x unified",  # Shared hover on the x-axis
+    yaxis = list(spikemode = "across", spikesnap = "cursor")
+  )
+
+fig
+
+df <- data[data$effectX1 == 1,]
+```
+Boxplots summarizing time performance results for our illustrative example. Dots represent the median time for each individual participant. 
+:::
+
+::: {#fig-eta-squared-2}
+```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
+df <- data[data$effectX1 == 1,]
+
+fig1 <- plot_ly(df, x = ~effectX2, y = ~X1, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
+fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
+fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
+fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))
+
+fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
+  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
+	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab)) %>% layout(
+    hovermode = "x unified",  # Shared hover on the x-axis
+    yaxis = list(spikemode = "across", spikesnap = "cursor")
+  )
+
+fig
+
+df <- data[data$effectX1 == 1,]
+```
+Boxplots summarizing time performance results for our illustrative example. Dots represent the median time for each individual participant. 
+:::

### 2025-06-22 text/40-interpretingEffects.qmd
many changes

@@
 
-**Dealing with interpretation issues.** To avoid such interpretation issues, we focus --- unless otherwise stated --- on effects that apply monotonic transformations to population distributions.
+**Dealing with interpretation issues.** To avoid such interpretation issues, we focus -- unless otherwise stated -- on effects that apply monotonic transformations to population distributions.
 This also ensures a monotonic relationship between different measures of central tendency such as medians and means (with the exception of the Cauchy distribution, where the mean is undefined).
@@
 In conclusion, focusing solely on the issues with the rank tranformation illustrated in @fig-interactions-rank is akin to missing the forest for the trees.
-When parallel main effects are present, interpreting interactions can be challenging for all methods --- and ART offers no solution to this problem.
+When parallel main effects are present, interpreting interactions can be challenging for all methods -- and ART offers no solution to this problem.
 
-### Debate on the interpretation of interactions
+### Controversy about the interpretation of interactions
 We further discuss interpetation issues in the broader context of generalized liner models (GLMs).
@@
 $$
-E(Y|x_1, x_2) = g^{-1}(a_0 + a_1 x_1 + a_2 x_2 + a_{12}x1 x2)
+E(Y|x_1, x_2) = g^{-1}(a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2)
 $${#eq-glm}
@@
 
+We emphasize that the existing literature does not clarify how ART is expected to interpret main and interaction effects in these scenarios.
+Early evaluations of ART did not consider nonlinear models, so the question was not even posed.
+@elkin:2021 who do study such models, use log-transformed values as their baseline for log-normal distributions (see their example on Page 4).
+So it seems that Elkin et al.
+assume ART's conclusions should align with the coefficients of the linear submodel.
+Unfortunately, their evaluations focus on contrasts and do not account for parallel effects, avoiding interpretation issues.
+
 Throughout the rest of the article, we make every effort to clarify when the interpretation of an effect is ambiguous.
@@
 ### Our approach {#approach}
-<!-- Our analysis shows that inference errors are not simply due to the lack of robustness of a statistical procedure.
-In the case of interaction effects, errors will also emerge when the procedure makes inference on the wrong scale.
-As @Wagenmakers:2012 explain, *"the dependent variable reflects merely the output of a latent psychological process",* and unfortunately, *"in most experimental paradigms the exact relationship between unobserved process and observed behavior is unknown ..."*
-
-Ideally, a statistical procedure should lead to conclusions that capture the true effects on the latent variable of interest.
-But as we discussed above, this might not be the case for the rank transformation methods that we study.
-For example, all four methods (*PAR*, *RNK*, *ART*, and *INT*) suggest that task difficulty interacts with use of technique on perceived performance because they disregard the fact that the observed data are simply projections on a discrete ordinal scale.
-Our goal is to understand how ART and the other methods deal with such problems.
--->
-The responses in a generalized linear model can follow arbitrary distributions.
+Responses in a generalized linear model can follow arbitrary distributions.
 A key challenge is how to simulate such distributions while maintaining control over the parameters of the linear submodel.
@@
 
-**Latent variables.** We assume here is a single latent variable $\mathcal{Y}$ that is distinct from the variable $Y$ we observe.
+**Latent variables.** We assume there is a single latent variable $\mathcal{Y}$ that is distinct from the variable $Y$ we observe.
 For example, the latent variable may represent a population's performance potential, working memory capacity, the perceived utility of a new technology, or quality of life.
@@
 
-**Interpreting effects.** Our approach allows us to simulate main and interaction effects on the latent variable $\mathcal{Y}$ and observe how these effects manifest on the transformed variable $Y = \mathcal{T}(Y)$.
+**Interpreting effects.** Our approach allows us to simulate main and interaction effects on the latent variable $\mathcal{Y}$ and observe how these effects manifest on the transformed variable $Y = \mathcal{T}(\mathcal{Y})$.
 
@@
 
-2. The definition of the null hypothesis changes with the scale —- a null effect on the latent variable $\mathcal{Y}$ may not remain null when evaluated on the response variable $Y$.
+2. The definition of the null hypothesis changes with the scale: a null effect on the latent variable $\mathcal{Y}$ may not remain null when evaluated on the response variable $Y$.
 
-For the first scenarios, we ensure that the interaction coefficient of the linear model is zero (e.g., $a_{12} = 0$ for the model in @eq-glm) when assessing main effects.
-When assessing interaction effects, we vary a single only coefficient that corresponds to a predictor (e.g., either $a_1$ or $a_2$), while keeping the others zero.
-
-<!--
-As discussed earlier, how to infer main effects is straightforward since we only study monotonic transformations here.
-Specifically, if we observe a main effect on the observed variable $Y'$, we can also conclude that there is a main effect on the latent variable $Y$.
-Notice, however, that if the observed variable is bounded or takes only discrete values, a main effect on $Y$ may not be observable.
-
-The interpretation of interactions is more challenging.
-Depending on how the statistical procedure we use deals with different scales, observations of interactions among two or more factors on the transformed variable $Y'$ may lead to incorrect conclusions about the presence of interactions on the latent space of $Y$.
-Such errors emerge when the main effect of all interacting factors on the latent (or observed) variable is non-zero.
-In our analysis, we count them as statistical inference errors (Type I and II errors).
-Nevertheless, we try to distinguish them from typical inference errors, for which interaction interpretation issues do not come into place.
--->
+Unless we explicitely mention interpretation issues, we focus on the first scenario.
+Note that in the case of discrete response variables, the function $\mathcal{T}$ is monotonic but not $strictly$ monotonic, since it involves a loss of information.
+Consequently, an effect on the latent variable may not be observable in the responses, or the observed magnitude of the effect may appear reduced.

### 2025-06-22 text/50-ExperimentalMethod.qmd
many changes

@@
 ### Statistical modeling
-To model the latent variable $Y$, we use a two-way (two factors) or a three-way (three factors) mixed-effects model.
+To model the latent variable $\mathcal{Y}$, we use a two-way (two factors) or a three-way (three factors) mixed-effects model.
 For simplicity, we explain here the model for two factors.
@@
 Varying $a_1$ or $a_2$ to control the magnitude of main effects.
-The plots show examples of population distributions on the latent variable $Y$ for factors ($X_1$ or $X_2$) with two (top), three (middle), or four categorical levels (bottom).
+The plots show examples of population distributions on the latent variable $\mathcal{Y}$ for factors ($X_1$ or $X_2$) with two (top), three (middle), or four categorical levels (bottom).
 :::
@@
 We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\epsilon_{ijk} \sim N(0,\sigma)$.
-To simulate the observed variable $Y'$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach).
+To simulate the observed variable $Y$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach).
 A key advantage of this method is that we can generate responses for any distribution, while we control effects on the latent variable in the exact same way.
@@
 However, we can compare the power of different methods to evaluate their relative performance.
-In addition to power, we assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted on the latent variable $Y$.
+In addition to power, we assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted on the latent variable $\mathcal{Y}$.
 While partial $\eta^2$ is the most commonly used effect size measure, we chose to base our analysis on Cohen's $f$, as its expected value is proportional to the real magnitude of effect.
@@
 
-As we explained earlier, interaction effects are deformed when we transform the latent variable $Y$ to obtain the observed responses.
+As we explained earlier, interaction effects are deformed when we transform the latent variable $\mathcal{Y}$ to obtain the observed responses.
 Such transformations also affect the Type I error rates and power that we observe.
-Here, we are interested in evaluating these measures with respect to the effects we apply to the latent variable $Y$.
+Here, we are interested in evaluating these measures with respect to the effects we apply to the latent variable $\mathcal{Y}$.
 We make this distinction whenever it is relevant in our analysis.

### 2025-06-22 text/60-Results-RatioScales.qmd
many changes

@@
 ### Type I error rates in ratio scales {#ratio}
-Our first experiments evaluate Type I error rates for ratio scales.
+We first evaluate Type I error rates for ratio scales.
 We test a 4 $\times$ 3 repeated-measures design, where we refer to the 4-level factor as $X_1$ and the 3-level factor as $X_2$.
-For the observed response variable $Y' = \mathcal{T}(Y)$, we evaluate transformations to four continuous and two discrete distributions:
+For the observed response variable $Y = \mathcal{T}(\mathcal{Y})$, we evaluate transformations to four continuous and two discrete distributions:
 
-1. No transformation, or $Y' = Y$. Distributions are normal with a grand mean $\mu = 0$ (e.g., see @fig-effects).
+1. No transformation, or $Y = \mathcal{Y}$. Distributions are normal with a grand mean $\mu = 0$ (e.g., see @fig-effects).
 
@@
 
-**Main effects**.
-@fig-ratio-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases.
-The results show a very good behavior for RNK and INT across all distributions.
-The regular parametric ANOVA (PAR) keeps error rates below $5\%$.
-However, error rates become extremely low for some distributions, suggesting a loss in statistical power.
-We confirm previous results that ART fails to control the Type I error rate under the Cauchy distribution [@Mansouri:1995; @elkin:2021].
-However, we also show that the method can be problematic with other non-normal distributions.
-As the main effect on the first factor $X_1$ increases, Type I errors on the second factor $X_2$ grow and reach high levels.
-This pattern is particularly pronounced under the log-normal distribution.
-We also observe that for the binomial distribution, error rates are high ($\approx 11\%$ for $n = 20$) even when effects on $X_1$ are zero.
-In addition, error rates further grow when the sample size increases.
+**Main effects**. @fig-ratio-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases.
 
@@
 :::
+
+ The results show a very good behavior for RNK and INT across all distributions.
+The regular parametric ANOVA (PAR) keeps error rates below $5\%$.
+However, error rates become extremely low for some distributions, suggesting a loss in statistical power.
+We confirm previous results that ART fails to control the Type I error rate under the Cauchy distribution [@Mansouri:1995; @elkin:2021] but we also show that the method is problematic with other non-normal distributions.
+As the main effect on the first factor $X_1$ increases, Type I errors on the second factor $X_2$ grow and reach high levels.
+This pattern is particularly pronounced under the log-normal distribution.
+We also observe that for the binomial distribution, error rates are high ($\approx 11\%$ for $n = 20$) even when effects on $X_1$ are zero.
+In addition, error rates further grow when the sample size increases.
 
@@
 
-Finally, @fig-ratio-interaction-2 presents our results when the main effects on the two factors increase in parallel. Error rates become exceptionally high in some cases, growing up to $100\%$.
-
-::: {#fig-ratio-interaction-2}
-```{r, echo=FALSE, message=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- readlyData(prefix, alpha, 0, distributions, dnames)
-
-#xlab <- TeX("\\text{main effect of factors }X_1\\text{ and }X_2\\text{ }(a_1 = a_2)")
-plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
-```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ 
-:::
-
-However, these results require special attention since interaction interpretation issues come now into place. Let us examine the performance of each method:
-
-- *PAR*.
-Error rates are high for all non-normal distributions, and become even higher when the sample size increases.
-As we explained in [Section 3](#interpretation), the parametric method cannot deal with scale differences that emerge when effects are transformed in a non-linear fashion.
-If we interpreted interactions based on absolute mean differences, disregarding their different scales (i.e., if we decided that the trend in @fig-interactions (left) is indeed a valid interaction effect), we would not observe these high error rates.
-
-- *RNK.* Error rates explode when both main effects exceed a certain level (e.g., when $a_1, a_2 \ge 4$ and $n = 20$).
-As shown in @fig-interactions-rank, the problem is due to the way the rank transformation deforms interactions.
-
-- *INT.* The method exhibits a better behavior than RNK.
-Errors start again increasing for all distributions, but not until main effects become significantly large.
-An exception is the binomial distribution, for which the error rates of INT and RNK are similar.
-We also observe that for continuous distributions, both INT and RNK are scale invariant, since their performance is not affected by the choice of scale.
-
-- *ART.* It keeps error rates at correct levels as long as population distributions are normal.
-For all other distributions, its error rates grow rapidly as effect sizes increase.
-We distinguish between two sources of errors: (i) lack of statistical robustness as observed in our previous tests; and (ii) interaction interpretation issues (see [Section 3](#interpretation)).
-Our results confirm that ART is not scale invariant.
-On the contrary, it is extremely sensitive to the scale of the observed data.
+The scenarios above do not capture the failure cases of the rank transformation discussed earlier (see @fig-interactions-rank), as those arise only when both coefficients $a_1$ and $a_2$ are non-zero.
+Since there can be different possible definitions of an interaction effect in these cases, we address them in a separate subsection.

### 2025-06-22 text/61-Results-OrdinalScales.qmd
many changes

@@
 ### Type I error rates in ordinal scales {#ordinal}
-Our second experiment evaluates Type I error rates for ordinal scales.
+We also evaluate Type I error rates for ordinal scales.
 We test again a $4 \times 3$ repeated-measures design.
@@
 
-**Main effects**.
-@fig-ordinal-main present Type I errors for $X_2$'s main effect, as we vary $X_1$'s magnitude of main effect.
-Our results indicate that PAR, RNK, and INT consistently maintain error rates close to $5\%$ across all tested ordinal scales.
-In contrast, ART exhibits a significant inflation of error rates, although this issue becomes less severe when the number of levels increases.
-Notably, error rates are more pronounced for flexible thresholds and tend to increase with sample size.
+**Main effects**. @fig-ordinal-main present Type I errors for $X_2$'s main effect, as we vary $X_1$'s magnitude of main effect.
 
@@
 
+Our results indicate that PAR, RNK, and INT consistently maintain error rates close to $5\%$ across all tested ordinal scales.
+In contrast, ART exhibits a significant inflation of error rates, although this issue becomes less severe when the number of levels increases.
+Notably, error rates are more pronounced for flexible thresholds and tend to increase with sample size.
+
 **Interaction effects**.
 @fig-ordinal-interaction-1 displays error rates for the interaction $X_1 \times X_2$ with a single main effect applied to $X_1$.
-The observed patterns align with our previous findings; ART consistently inflates error rates, although to a lesser extent now.
+The observed patterns align with our previous findings; ART consistently inflates error rates.
 
@@
 
-Finally, @fig-ordinal-interaction-2 presents our results when main effects are applied to both factors.
-None of the methods successfully maintains low error rates across all conditions.
-Intriguingly, we note that ART consistently performs worse than PAR.
-INT performs worse than ART and PAR in one specific case ($a_2, a_2 = 8$ in a scale with 11 equidistant levels), but overall, it exhibits a better behavior than all other methods.
+**Analysis for large samples.** Let us further explore ART's error rates when sample sizes become larger.
+As shown in @fig-ordinal-large-samples, error rates explode even when all population effects are null.
 
-::: {#fig-ordinal-interaction-2}
+::: {#fig-ordinal-large-samples}
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- readlyData(prefix, alpha, 0, distributions, dnames)
-plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
+prefix <- "2_test_4x3_Ordinal_sample_size"
 
+distributions <- c("likert5", "likert5B", "likert7", "likert7B", "likert11", "likert11B")
+dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
+
+df <- readlyData1(prefix, alpha, 0, distributions, dnames)
+plotlyErrorByEffect(df, xlab = "n", max = 100)
 ```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ 
+Type I error rates ($\alpha = .05$) as a function of the size $n$ of the sample, where all population effects are null.
 :::
+
+We will observe similar trends if we test other discrete distributions, such as binomial distributions. These results confirm that ART is inappropriate for discrete data.

### 2025-06-22 text/62-Results-AcrossExpDesigns.qmd
many changes

### Type I errors across experimental designs {#extra}
We investigates the generalizability of the aforementioned results to other experimental designs involving two or three factors. We assess five out of the six ratio scales examined earlier. We exclude the Cauchy distribution and replace it by a 5-level ordinal scale with flexible thresholds.

**Main effects**. @fig-designs-main illustrates Type I errors for the main effect of $X_2$ while varying the magnitude of the main effect of $X_1$, with a focus on $n = 20$. Across all cases, RNK and INT consistently maintain error rates close to $5\%$. PAR's error rate remains overall close to $5\%$ but reaches notably low levels for specific combinations of designs ($2 \times 2 \times 2$ and $3 \times 3 \times 3$ repeated-measures) and distributions (log-normal and exponential). In contrast, ART inflates error rates for all non-normal distributions, with discrepancies across different designs. We observe that ART is particularly problematic in discrete distributions when applied to the $2 \times 2 \times 2$ and $3 \times 3 \times 3$ designs.

::: {#fig-designs-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "3_test-Designs"
alpha = 0.05

distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha, effectType = 1, distributions)
df <- reshapeByDesign(df, dnames)

plotlyErrorByDesign(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 62)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
:::

We provide additional results for the between-subjects design in @fig-2x3-main, where we vary the effect of $X_2$ and measure Type I error rates on $X_1$. ART's error rates for the three discrete distributions appear higher than those observed in @fig-designs-main, indicating a potential dependence on the number of levels of the factors. Again, we observe a consistent trend where error rates in all discrete distributions increase with the sample size.

::: {#fig-2x3-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix2 <- "3_test-Designs_2x3"

df2x3 <- readlyData2(prefix2, alpha, effectType = 2, distributions, dnames)

plotlyError(df2x3, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX2", max = 62)
```
Additional results for the $2 \times 3$ between-subjects design, where we measure Type I error rates ($\alpha = .05$) for the **main effect of $X_1$.** We now vary the magnitude $a_2$ of the main effect of $X_2$.
:::

**Interaction effects**. @fig-designs-interaction-1 displays Type I error rates for the interaction effect $X_1 \times X_2$ while varying the effect of $X_1$ ($n = 20$). We observe consistent trends in line with our previous findings. For additional insights, we direct readers to our raw experimental data, which demonstrate that these trends persist across other interaction terms, namely $X_1 \times X_3$, $X_2 \times X_3$, and $X_1 \times X_2 \times X_3$, within the two 3-factor designs.

::: {#fig-designs-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 62)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$) 
:::

### 2025-06-22 text/63-Results-UnequalVariance.qmd
many changes

### Type I errors under unequal variances
Many statistical procedures assume equal variances among all levels of each independent variable, or more strictly, among all possible pairs of independent variable levels (known as sphericity assumption in repeated-measures ANOVA). Nonparametric tests are often mistakenly considered to be free of such as assumptions, but as we discussed earlier, this is generally incorrect.

Our next experiment evaluates the behavior of the four methods when populations on the latent variable $\mathcal{Y}$ have unequal variances. As we previously discussed, even when variances are equal at the latent space, they may appear as unequal on the response scale if the response distributions are skewed. This is the case, for example, with the log-normal and the exponential distributions. So, in this experiment, we explore more extreme violations of the equal-variance assumption for these distributions.

We examine three 2-factor designs: (i) a $4 \times 3$ within-subjects design; (ii) a $2 \times 3$ between-subjects design; and (iii) a $2 \times 4$ mixed design. We set all effects to zero and then vary the ratio $r_{sd}$ of standard deviations between the levels of the first factor $X_1$. @fig-unequal shows the five ratios that we test when $X_1$ has two levels. When $X_1$ has additional levels ($4 \times 3$ design), their standard deviation is randomly drawn in a range between the standard deviations of the first two levels.

::: {#fig-unequal}
```{r, echo=FALSE, message=FALSE, fig.height=1.5, fig.width = 9, warning=FALSE}
plotVariances()
```
Varying the ratio $r_{sd}$ of standard deviations between the levels of $X_1$. The plots show examples of population distributions for factors with two categorical levels.    
:::

**Main effects**. We begin by investigating how the four methods detect main effects on $X_1$. The interpretation of results in this scenario is challenging due to differences in the original populations -- although they share the same means (and medians), their variances differ. Depending on the null hypothesis of interest, conclusions may vary. Complicating matters further, non-linear transformations can result in distributions with differing means and medians. Consequently, the choice of statistical method may lead to different outcomes; one method may be sensitive to mean differences while others to median differences.

@fig-unequal-main-1 presents the percentage of times each method rejects the null hypothesis ($\%$ positives) at a significance level of $\alpha = .05$. If the null hypothesis of interest posits that *"$X_1$ has no effect on the mean of the latent variable $Y$,"* then these instances should be considered as Type I errors. Under normal distributions, all methods either moderately inflate or deflate the positives ratio as $r_{sd}$ increases. For the $4 \times 3$ within-subjects design ($r_{sd} \ge 2.5$), the ratio for PAR, RNK, and ART reaches levels of around $7-8\%$, while for INT, it drops below $3\%$ under the $2 \times 4$ mixed design ($r_{sd} \ge 2.5$).

::: {#fig-unequal-main-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "4_test_Unequal_Variances"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha, effectType = 1, distributions)
df <- reshapeByDesign(df, dnames, effectvars = c("sd_ratio", "effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1", xvar = "sd_ratio", max = 82, ytitle = 'Positives (%)')
```
The percentage of positives (i.e., rejecting the null hypothesis) for the **main effect of $X_1$** as the ratio of standard deviations $r_{sd}$ on $X_1$ increases ($n=20$). Depending on the hypothesis of interest, these percentages can be interpreted as Type I error rates ($\alpha = .05$).
:::

The performance of RNK and INT remains unchanged across continuous distributions, a result expected due to the preservation of medians in such scales. Conversely, mean differences grow with increasing $r_{sd}$, which explains why PAR's ratio of positives reaches high levels under these distributions. However, this trend is not consistent across all three designs. The behavior of ART is less clear, further supporting our argument that its null hypothesis of interest is ill-defined. Under discrete distributions, the trends of positives vary across designs, with PAR and ART exhibiting the highest rates.

We also investigate the influence of unequal variances among the levels of $X_1$ on positive rates for $X_2$, as depicted in @fig-unequal-main-2. These can be reliably considered as Type I errors because source populations defined by $X_2$ are identical. The error rates of PAR, RNK, and INT do not seem to be affected by an increase of $r_{sd}$. In contrast, ART yields high error rates, surpassing acceptable levels even in the case of normal distributions (check results for the $2 \times 4$ mixed design). Thus, the problem of unequal variance is more serious for ART than for other methods, as the method's sensitivity propagates across factors. This further demonstrates that ART's alignment procedure causes the method to confound effects.

::: {#fig-unequal-main-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX2", xvar = "sd_ratio", max = 52)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as the ratio of standard deviations $r_{sd}$ on $X_1$ increases ($n=20$)
:::

**Interaction effects**. Finally, we measure Type I error rates for the interaction effect $X_1 \times X_2$, shown in @fig-unequal-interaction. The error rates of ART exhibit similar trends, with a somewhat lower level under the $4 \times 3$ within-subjects design (compared to @fig-unequal-main-2). Additionally, for this design, we observe error inflation for the three other methods, albeit to a significantly lesser degree.

::: {#fig-unequal-interaction}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1X2", xvar = "sd_ratio", max = 52)
```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as the ratio of standard deviations $r_{sd}$ on $X_1$ increases ($n=20$)
:::

### 2025-06-22 text/66-Results-Interactions.qmd
many changes

### Interactions under parallel main effects {#interactions}
Let us now examine interaction effects when both coefficients $a_1$ and $a_2$ increase in parallel.
In this scenario, the interpretation of an interaction effect can become ambiguous. Here, we define the null hypothesis in terms of the coefficients of the latent linear model, following the recommendation of @Greene:2010 -- that is, $a_{12} = 0$.

**Ratio scales**. @fig-ratio-interaction-2 presents Type I error rates for ratio scales.

::: {#fig-ratio-interaction-2}
```{r, echo=FALSE, message=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
distributions <- c("norm", "lnorm", "exp", "cauchy", "poisson", "binom")
dnames <- c("Normal", "Log-normal", "Exponential", "Cauchy", "Poisson", "Binomial")
alpha <- .05
prefix <- "1_test_4x3_Ratio"
df <- readlyData(prefix, alpha, 0, distributions, dnames)

#xlab <- TeX("\\text{main effect of factors }X_1\\text{ and }X_2\\text{ }(a_1 = a_2)")
plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$. *Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$. A different definition may lead to different results.* 
:::

Error rates become exceptionally high in some cases, reaching up to 100%. However, these results require careful interpretation. As discussed in [Section 4](#interpretation), interaction effects in nonlinear models can be interpreted in two distinct ways, and the results from the different methods are not consistent with our definition. Let us examine the outcomes for each method:

- *PAR*. Error rates are high for all non-normal distributions and become even higher when the sample size increases. However, this behavior clearly stems from the method's sensitivity to the scale of the observations, rather than to the model parameters.

- *RNK.* Error rates explode when both main effects exceed a certain level (e.g., when $a_1, a_2 \ge 4$ and $n = 20$). This problem arises from the way the rank transformation distorts interaction effects (see @fig-interactions-rank). For all continuous distributions, we observe that error rates exhibit the exact same patterns, since ranks are scale-invariant.

- *INT.* It performs better than RNK. Errors begin to increase across all distributions only when the main effects become large. An exception is the binomial distribution, for which the error rates of INT and RNK are similar.

- *ART.* It maintains correct error rates as long as population distributions are normal. For all other distributions, error rates increase rapidly as effect sizes grow. Two factors explain this behavior: (i) a lack of statistical robustness to violation of the method's assumptions, as also observed in earlier tests; and (ii) interaction interpretation issues. These results confirm that ART is not scale-invariant, interpreting interactions relative to the scale of the observed data.

**Ordinal scales.** @fig-ordinal-interaction-2 presents our results for ordinal data. No method maintains low error rates under all conditions. We note that ART consistently performs worse than PAR. INT performs worse than ART and PAR in one specific case ($a_2, a_2 = 8$ in a scale with 11 equidistant levels), but overall, it exhibits a better behavior than all other methods. Once again, the high error rates can be partly explained by insufficient statistical robustness and ambiguities in how the methods interpret interaction effects.

::: {#fig-ordinal-interaction-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "2_test_4x3_Ordinal"
distributions <- c("likert5", "likert5B", "likert7", "likert7B", "likert11", "likert11B")
dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")

df <- readlyData(prefix, alpha, 0, distributions, dnames)
plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)

```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$. *Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$. A different definition may lead to different results.*
:::

**Other designs.** Our findings in @fig-designs-interaction-2 further demonstrate that all methods struggle to accurately detect interactions for different reasons. Interestingly, INT shows unusually low error rates in certain configurations (e.g., $2 \times 3$ and $2 \times 2 \times 2$), indicating potential difficulty in detecting subtle interaction effects when main effects are large. RNK follows a similar trend for the $2 \times 2 \times 2$ design. Both RNK and INT also exhibit higher error rates under ordinal and binomial scales, though these remain lower than those observed for PAR and ART.

::: {#fig-designs-interaction-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "3_test-Designs"
distributions <- c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames <- c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha, effectType = 0, distributions)
df <- reshapeByDesign(df, dnames)
plotlyErrorByDesign(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$). *Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$. A different definition may lead to different results.*
:::

Some readers may argue that evaluating PAR and ART using an interaction definition that accounts for the scale of the observed variable might be more appropriate. We offer three responses to this criticism:

1. It is unclear how to reliably control the null hypothesis at the level of the response scale when both main effects are varied.

2. Our earlier experiments demonstrate that ART fails to maintain nominal Type I error rates for interactions even when only a single main effect is present. Introducing a second main effect could only exacerbate the problem.

3. The purpose of these results is not to provide precise benchmarks, but rather to illustrate the complexity and pitfalls of interpreting interaction effects under assumption violations -- regardless of the method used. Many researchers are unaware of these issues and often misinterpret the $p$-value when testing interactions. As previously noted, @elkin:2021 compared ART to ANOVA on log-transformed data under the belief that ART tests null hypotheses about parameters of the latent linear model. Our findings clearly show that this is not the case.

### 2025-06-22 plotlying.R

@@ -1,8 +1,7 @@
 
 library(plotly)
 
-# If byDesign is TRUE, then the menu will choose among three experimental designs 
-plotlyError <- function(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 100, 
+plotlyError <- function(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", ytitle = 'Type I errors (%)', max = 100, 
 	cbPalette = c("#888888", "#E69F00", "#009E73", "#FF5E00"), symbols = c("asterisk", "x", "star-diamond", "star-triangle-up")){
 	# aesthetics
 	
@@ -22,7 +21,7 @@ plotlyError <- function(df, xlab = "magnitude of main effects", var = "rateX1X2"
 	  layout(
 	    legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5),
 	    xaxis = list(title = NA, showline=T, mirror = F, fixedrange=T, ticks="outside",tickangle=60, tickfont = list(size = 11)),
-	    yaxis = list(title = 'Type I errors (%)', font = list(size = 13), zeroline = F, showline=T, linewidth=1, mirror = F,  nticks=6, ticks="inside", tickfont = list(size = 11), range=c(0, max))
+	    yaxis = list(title = ytitle, font = list(size = 13), zeroline = F, showline=T, linewidth=1, mirror = F,  nticks=6, ticks="inside", tickfont = list(size = 11), range=c(0, max))
 	  ) 
 
 	  p
@@ -49,8 +48,52 @@ plotlyError <- function(df, xlab = "magnitude of main effects", var = "rateX1X2"
 	fig
 }
 
+plotlyErrorByEffect <- function(df, xlab = "n", max = 100, 
+	cbPalette = c("#888888", "#E69F00", "#009E73", "#FF5E00"), symbols = c("asterisk", "x", "star-diamond", "star-triangle-up")){
+	# aesthetics
+	
+	margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
 
-plotlyErrorByDesign <- function(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 100){
+	dnames <- levels(df$distr)
+	# To be used as labels on the subplots
[…]

### 2025-06-22 supplementary/experimental-code/Main-2-within-4x3-ordinal-sample-size.R

@@ -1,0 +1,127 @@
+# Author: Theophanis Tsandilas, 2023-2024
+# Inria & Université Paris-Saclay
+
+# Experiment evaluating the Type I error rates of PAR, ART, RNK, and INT on ordinal scales for a 4x3 within-subjects design 
+# See Fig. 14, Fig. 15, Fig. 16
+
+# By default, this code will generate a data file for n = 20 with 5000 iteration. 
+# You can edit the experimental parameters in the code to test different conditions.
+
+rm(list=ls())
+library(lmerTest)
+
+# The ARTool
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+source("data-generators.R") # Import the data-generation functions
+
+getLmerPValues <- function(model, vars) {
+	return(anova(model)[vars, 6])
+}
+
+
+getARTPValues <- function(model, vars) {
+	return(anova(model)[vars, 5]) # For within-subjects
+}
+
+# Repeated measures with two factors
+test.within.2f <- function(n, design, effects, distr){ 
+	df <- sim.within.2f(n, design, effects) %>% toDistribution(distr)
+
+	model0 <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
+	model1 <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
+	model2 <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
+	model3 <- suppressMessages(lmer(inverseNormalTransform(y) ~ x1*x2 + (1|s), data=df)) # INT
+	vars <- c("x1", "x2", "x1:x2")
+	c(getLmerPValues(model0, vars), getLmerPValues(model1, vars), getARTPValues(model2, vars), getLmerPValues(model3, vars))
+}
+
+# Performs repetitive tests for the given configuration
+test <- function(repetitions, n, design, effects, distr) {
+	results <- foreach(rid = 1:repetitions, .combine=rbind) %dopar% {
+		tryCatch(
+			{
+				test.within.2f(n, design, effects, distr)
+			}, 
+			error = function(cond) {
+				# do nothing
+			}, finally = {
+				# do nothing
+			}
+		)
+	}
+
+	# From p-values to positive rates that represent (depending on whether there is a true effect) either Type I error rates (false positives) or power (true positives) 
[…]

### 2025-06-24 sandbox.qmd
effect sizes

--- 2025-05-24 sandbox.qmd
@@
 
-Used for quick testing... eeee
+::: {.callout-important appearance="simple"}
+## Under Review {.unnumbered}
+This paper is [under review](https://www.journalovi.org/under-review.html) on the experimental track of
+the [Journal of Visualization and Interaction](https://www.journalovi.org/). See the 
+[reviewing process](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues).
+:::
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+# Useful libraries
+library(crosstalk)
+library(kableExtra)
+library(gridExtra)
+library(lmerTest)
+library(tidyverse)
+library(plotly)
+```
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+# Source code for reading data from experimental results and for plotting
+source("dataReaders.R")
+source("plotting.R")
+source("plotlying.R")
+source("effects_plot.R")
+source("interactions_plot.R")
+```
+
+```{=html}
+<style>
+.math.inline .MathJax  {
+  font-size: 105% !important;
+}
+</style>
+```
+
+::: {#fig-effect-main-scatterplots-1}
+```{r, echo=FALSE, fig.height=3.4, message=FALSE, warning=FALSE}
+source("effect_sizes_scatterplots.R")
+prefix <- "6_scatter-Effect_Size-g2"
+df <- readlyDataPoints(prefix,  distributions=c("norm", "lnorm", "likert5B"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)"), effect = "X1X2") %>% toCohensf2()
+
+fig <- plotlyScatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "fX1X2", max = 0.6)
+
+fig
+```
+Scatterplots showing the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground-truth method for the main effect of $X_1$, when $a_1$ and $a_2$ vary within the range $[-2, 2]$.
+:::

### 2025-06-24 text/64-Results-StatPower.qmd
effect sizes

### Statistical power {#power}
Our sixth experiment compares the statistical power of the four methods on the three 2-factor designs. Because there is a tradeoff between Type I and Type II errors, high power can simply be the result of a high Type I error rate. Since parallel effects can inflate errors (see our previous results), we focus here on single effects, both main or interaction effects. We only report results for $n=20$ (see supplementary materials for additional results).

**Main effects**. We individually vary the magnitude of effect on $X_1$ and $X_2$ and observe the power of each method to detect this effect ($\alpha = .05$). Specifically, we manipulate the parameters $a_1$ or $a_2$ within the range of $0.4$ to $1.0$. Depending on the experimental design, this range allows us to simulate both low-power experiments (e.g., with less than $30\%$ power) and high-power experiments (e.g., with more than $80\%$ power).

@fig-power-main-1 presents our results for $X_1$. In several configurations, the differences in power among the methods are marginal. To enhance clarity in our comparisons, we also present the method rankings. Readers are encouraged to interact with the graphs for precise power values. PAR exhibits the highest power under the normal distribution, closely followed by INT. INT is also consistently the winning method across all non-normal distributions. PAR's power is notably low under the log-normal and exponential distribution.

ART initially appears advantageous under the three discrete distributions for small effects, likely due to inflated Type I errors even without an effect on $X_2$ (as observed in @fig-2x3-main). However, as effects grow larger, its relative power diminishes, and ART eventually becomes the least effective method.

::: {#fig-power-main-1}
<div id="plot-power-1.1">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "5_test-Power"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

data <- readData(prefix, n = 20, alpha = .05, effectType = 3, distributions)
df <- reshapeByDesign(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX1", max = 104,  ytitle = 'Power (%)')
```
</div>

<div id="plot-power-1.2">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- data %>% arrange(design,distr,effectX1,rateX1)  %>% group_by(design,distr,effectX1) %>% mutate(rank = rank(rateX1))
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyPowerByDesign(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power - ranking')
```
</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-1.2").classList.add("hidden"); }); </script>

<button id="btn1" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
  onclick="const plot = document.getElementById('plot-power-1.1');
  plot.classList.toggle('hidden');
  document.getElementById('plot-power-1.2').classList.toggle('hidden');
  btn1.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
">
  show ranking
</button>

Power ($\alpha = .05$) for detecting the **main effect of $X_1$** as a function of its magnitude $a_1$, where $n=20$.
:::

@fig-power-main-2 presents our results for $X_2$. Given the different number of levels for this factor, we now assess a different range of power for each design. Nevertheless, we observe similar patterns. INT stands out as the most powerful method. Although ART surpasses PAR under the log-normal and exponential distributions, it does not demonstrate any other notable advantages compared to the other methods.

::: {#fig-power-main-2}
<div id="plot-power-2.1">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
data <- readData(prefix, n = 20, alpha, effectType = 4, distributions) # CHANGE n TO 20

df <- reshapeByDesign(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX2", max = 104,  ytitle = 'Power (%)')
```
</div>

<div id="plot-power-2.2">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- data %>% arrange(design,distr,effectX2,rateX2)  %>% group_by(design,distr,effectX2) %>% mutate(rank = rank(rateX2))
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyPowerByDesign(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power - ranking')
```
</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-2.2").classList.add("hidden"); }); </script>

<button id="btn2" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
  onclick="const plot = document.getElementById('plot-power-2.1');
  plot.classList.toggle('hidden');
  document.getElementById('plot-power-2.2').classList.toggle('hidden');
  btn2.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
">
  show ranking
</button>

Power ($\alpha = .05$) for detecting the **main effect of $X_2$** as a function of its magnitude $a_2$, where $n=20$. 
:::

**Interaction effects**. We also vary the magnitude of the interaction effect $X_1 \times X_2$ by manipulating the parameter $a_{12}$ across the range of $0.5$ to $2.0$. We then measure the power to detect this interaction effect. Our results are summarized in @fig-power-interaction, aligning with findings for main effects. ART's relative advantage diminishes later under the binomial and ordinal distributions. Once again, the method's inflated Type I error rates could account for this observed trend.

::: {#fig-power-interaction}
<div id="plot-power-3.1">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
data <- readData(prefix, n = 20, alpha, effectType = 5, distributions)

df <- reshapeByDesign(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX1X2", xvar = "effectX1X2", max = 104,  ytitle = 'Power (%)')
```
</div>

<div id="plot-power-3.2">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- data %>% arrange(design,distr,effectX1X2,rateX1X2)  %>% group_by(design,distr,effectX1X2) %>% mutate(rank = rank(rateX1X2))
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyPowerByDesign(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power - ranking')
```
</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-3.2").classList.add("hidden"); }); </script>

<button id="btn3" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
  onclick="const plot = document.getElementById('plot-power-3.1');
  plot.classList.toggle('hidden');
  document.getElementById('plot-power-3.2').classList.toggle('hidden');
  btn3.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
">
  show ranking
</button>

Power ($\alpha = .05$) for detecting the **interaction effect $X_1 \times X_2$** as a function of its magnitude $a_{12}$, where $n=20$.
:::

### 2025-06-24 text/65-Results-EffectSize.qmd
effect sizes

@@
 ### Effect size estimates {#effect-sizes}
+
+We expect that discrepancies in the *p*-values produced by different methods will also be reflected in their corresponding effect size estimates.
+We presents results from various experiments, concentrating on either partial $\eta^2$ estimates or Cohen's $f$ estimates.
+
+**Effect sizes estimates in log-normal distributions.** Consider a $4 \times 3$ repeated-measures design with $n = 20$, where the effect of the first factor $X_1$ is null ($a_1 = 0$) and there is no interaction ($a_{12} = 0$).
+We assume that the responses follow log-normal distributions.
+@fig-eta-squared displays the partial $\eta^2$ estimated by each method for the effect of $X_1$ as the magnitude of the effect on the second factor increases.
+Each cell contains 300 data points.
+The ground truth uses ANOVA directly on the latent variable, where distributions are normal.
+
+::: {#fig-eta-squared}
+```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
+data <- read.csv("data/effect-sizes-lognormal.csv", sep=",", header=TRUE, strip.white=TRUE)
+data$effectX2 <- factor(data$effectX2)
+
+ylab = "X1 - partial eta squared"
+xlab = "X2 - magnitude of main effect (a2)"
+
+margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
+cbPalette = c("#8888ff", "#888888", "#E69F00", "#009E73", "#FF5E00")
+cbPalette2 = c("#8888ff88", "#88888888", "#E69F0088", "#009E7388", "#FF5E0088")
+
+df <- data[data$effectX1 == 0,]
+
+fig1 <- plot_ly(df, x = ~effectX2, y = ~X1, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
+fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
+fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
+fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))
+
+fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
+  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
+	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab)) %>% layout(
+    hovermode = "x unified",  # Shared hover on the x-axis
+    yaxis = list(spikemode = "across", spikesnap = "cursor")
+  )
+
+fig
+
+df <- data[data$effectX1 == 1,]
+```
+Boxplots summarizing the $\eta^2$ estimates obtained with each method for 300 data points.
+The population effect of $X1$ is null, and responses follow log-normal distributions.
+Hover over the plots to compare the methods.
+:::
+
+ART is the only method whose effect size estimates for $X_1$ are affected by the magnitude of the effect on $X_2$.
+The results clearly demonstrate that the method's alignment mechanism fails in nonlinear models with skewed distributions and confounds effects.
+
+@fig-eta-squared-2 shows the $\eta^2$ estimates when the effect of $X_1$ is no longer null ($a_1 = 1$).
+In this scenario, PAR consistently underestimates effect sizes, indicating its unsuitability for such data.
+RNK also slightly underestimates effects.
+ART tends to underestimate effects when $a_2 = 0$, but as $a_2$ increases, its estimates become more variable and exhibit an upward bias.
+INT provides the most accurate and stable estimates, with only a slight deviation when $a_2 = 8$.
+
+@fig-eta-squared-2 presents $\eta^2$ estimates when the effect of $X_1$ is no longer null ($a_1 = 1$).
+We observe that PAR is not ideal for such data, as it consistently underestimates effect sizes.
+RNK also slightly underestimates effects.
+ART tends to underestimate effects when $a_2 = 0$, but as $a_2$ increases, its estimates become more variable and show an overall tendency to increase.
+INT provides the most precise estimates but is also slightly affected when $a_2 = 8$.
+
+::: {#fig-eta-squared-2}
+```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
+df <- data[data$effectX1 == 1,]
+
+fig1 <- plot_ly(df, x = ~effectX2, y = ~X1, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
+fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
+fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
+fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))
+
+fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
+  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
+	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab)) %>% layout(
+    hovermode = "x unified",  # Shared hover on the x-axis
+    yaxis = list(spikemode = "across", spikesnap = "cursor")
+  )
+
+fig
+
+df <- data[data$effectX1 == 1,]
+```
+Boxplots summarizing the $\eta^2$ estimates obtained with each method for 300 data points.
+The population effect of $X1$ is now non-null ($a_1 = 1$).
+Responses follow again log-normal distributions.
+Hover over the plots to compare the methods.
+:::
+
 Our final experiment assesses the precision of the effect size estimates (Cohen's $f$) of each method.
 
-**Main effects**.
-We first evaluate estimates of the main effect of $X_1$ as the magnitude of the two main effects varies.
+**Relationship to ground truth estimates**.
+We evaluate the precision of effect size estimates for wider range of effects.
+We now use Cohen's $f$ as a measure of effect size, which is expected to be proportional to the real effect.
+Without loss of generality, we focus on the main effect of $X_1$.
+We also limit our analysis to three distributions: normal, log-normal, and ordinal with five flexible levels.
+
 @fig-effect-main-scatterplots-1 illustrates the relationship between the Cohen's $f$ obtained with each method and the $f$ of the ground-truth method, when $a_1$ and $a_2$ vary within the range $[-2, 2]$.
@@
 @fig-effect-main-scatterplots-2 shows the same relationship but for a larger range of effects $[-8, 8]$.
-We observe that all three rank-based methods struggle to detect large effects, even when the underlying distribution is normal.
-Interestingly, INT's performance declines when Cohen's $f$ exceeds a certain level.
-For the ordinal distribution, estimates are extremely noisy across all methods.
-We emphasize, however, that this is at least partly due to the low precision of the ordinal scale itself, since its five discrete levels cannot capture the full range of variations of the latent variable.
-The fact that all methods underestimate large effects can be simply explained by the limited range of the ordinal scale's extremes.
 
@@
 
+We observe that all three rank-based methods struggle to detect large effects, even when the underlying distribution is normal.
+This result reflects a fundamental limitation of ranks.
+Consider, for example, two sets of numbers $A= \{1, 2, 3\}$ and $B= \{4, 5, 6\}$.
+The mean difference between them is $\overline{B} - \overline{A} = 3$, whether calculated from the row values or their ranks.
+However, if the set $B$ becomes ${9, 10, 11\}$, the mean difference of raw values increases to $8$, while the mean difference in ranks remains $3$.
+As effects become larger, this problem becomes more apparent.
+
+In the case of ordinal data, all methods yield highly variable estimates.
+This variability is largely due to the limited resolution of the ordinal scale itself: with only five discrete levels, it cannot adequately represent the full range of variation in the latent variable.
+The systematic underestimation of large effects across methods is also due to the limited range of the ordinal scale's extremes, which masks the true magnitude of underlying differences.
+
+<!--
 To systematically evaluate the precision of effect size estimates across different methods, we employ the coefficient of determination, denoted as $R^2$ ($R$ squared).
@@
 :::
+
+-->

### 2025-06-24 effect_sizes_scatterplots.R

@@ -122,6 +122,14 @@ toCohensf <- function(df){
     df
 }
 
+toCohensf2 <- function(df){
+    df[,5:10] <- sqrt(df[,5:10]/(1-df[,5:10]))
+    names(df)[5:10] <- c("fX1X2.2x3", "fX1X2_.2x3", "fX1X2.2x4", "fX1X2_.2x4", "fX1X2.4x3", "fX1X2_.4x3")
+
+    df
+}
+
+
 
 #prefix <- "6_scatter-Effect_Size-Main"
 #df <- readlyDataPoints(prefix,  distributions=c("norm", "lnorm", "likert5B"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)"))

### 2025-06-27 sandbox.qmd
new version

@@
 
-::: {#fig-effect-main-scatterplots-1}
-```{r, echo=FALSE, fig.height=3.4, message=FALSE, warning=FALSE}
-source("effect_sizes_scatterplots.R")
-prefix <- "6_scatter-Effect_Size-g2"
-df <- readlyDataPoints(prefix,  distributions=c("norm", "lnorm", "likert5B"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)"), effect = "X1X2") %>% toCohensf2()
+::: {#fig-art-binary}
+```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=3.2, fig.width=8}
+library(gridExtra)
 
-fig <- plotlyScatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "fX1X2", max = 0.6)
+df <- read.csv("data-binary.csv", sep=",", header=TRUE, strip.white=TRUE)
+df$S <- factor(df$S)
+df$A <- factor(df$A)
+df$B <- factor(df$B)
 
-fig
+df$rnk <- rank(df$Y)
+
+m <- mean(df$Y)
+Ai <- aggregate(Y ~ A, data = df, mean)
+Bj <- aggregate(Y ~ B, data = df, mean) 
+AiBj <- aggregate(Y ~ A+B, data = df, mean) 
+cmean <- aggregate(Y ~ A+B, data = df, mean)
+
+library(dplyr)
+df <- df %>% group_by(A) %>% mutate(Ai = mean(Y)) %>% 
+  group_by(B) %>% mutate(Bj = mean(Y)) %>% 
+  group_by(A,B) %>% mutate(AiBj = mean(Y)) %>% mutate(Cell = mean(Y))
+
+df$m <- mean(df$Y)
+
+df$artA <- df$Y - df$Cell + df$Ai - df$m
+df$artB <- df$Y - df$Cell + df$Bj - df$m
+df$artAB <- df$Y - df$Cell + df$AiBj - df$Ai - df$Bj + df$m
+
+df$rnkA <- rank(round(df$artA, digits = 7))
+df$rnkB <- rank(round(df$artB, digits = 7))
+df$rnkAB <- rank(round(df$artAB, digits = 7))
+
+mytheme <- gridExtra::ttheme_default(
+    core = list(fg_params=list(cex = 2.0)),
+    colhead = list(fg_params=list(cex = 1.0)),
+    rowhead = list(fg_params=list(cex = 1.0)))
+
+myTable <- tableGrob(df[,1:4], theme = ttheme_minimal(base_size = 10) )
+
+diffPalette = c("#E69F00", "#FF5E00")
+p1 <- (df %>%   
+        ggplot(aes(x = B, y = rnk, fill = B)) + ggtitle("RNK") +
+        geom_boxplot(outlier.shape = NA) +
+        geom_jitter(position=position_jitter(0.3,0), size = 2, color=diffPalette[1], alpha=0.6) +
+        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) +  
+        ylab("ranking") + scale_y_discrete(limits=as.character(1:nrow(df)))
+
+p2 <- (df %>%   
+        ggplot(aes(x = B, y = rnkB, fill = B)) + ggtitle("ART_B") +
+        geom_boxplot(outlier.shape = NA) + 
+        geom_jitter(position=position_jitter(0.3,0), size = 2, color=diffPalette[2], alpha=0.6) +
+        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) +  
+        theme(axis.title.y=element_blank(), axis.text.y=element_blank()) + scale_y_discrete(limits=as.character(1:nrow(df)))
+
+grid.arrange(myTable, p1, p2, nrow = 1)
 ```
-Scatterplots showing the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground-truth method for the main effect of $X_1$, when $a_1$ and $a_2$ vary within the range $[-2, 2]$.
+Results from an experiment following a $3 \times 2$ design, where the response variable $Y$ is binary and indicates the presence of an error.
+The plots show the rankings produced by the simple rank transformation (RNK) and by ART for the second factor.
+The solid lines represent group medians.
 :::

### 2025-06-27 text/30-assumptions.qmd
new version

@@
 library(gridExtra)
-#library(ggplot2)
 
@@
 
-diffPalette <- c("#e2e6fa", "#99b9f9")
+diffPalette = c("#E69F00", "#FF5E00")
 p1 <- (df %>%   
@@
         geom_boxplot(outlier.shape = NA) +
-        geom_jitter(shape=1, position=position_jitter(0.3,0), size = 2) +
-        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) + scale_fill_manual(values=diffPalette) +  
+        geom_jitter(position=position_jitter(0.3,0), size = 2, color=diffPalette[1], alpha=0.6) +
+        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) +  
         ylab("ranking") + scale_y_discrete(limits=as.character(1:nrow(df)))
@@
         geom_boxplot(outlier.shape = NA) + 
-        geom_jitter(shape=1, position=position_jitter(0.3,0), size = 2) +
-        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) + scale_fill_manual(values=diffPalette) +  
+        geom_jitter(position=position_jitter(0.3,0), size = 2, color=diffPalette[2], alpha=0.6) +
+        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) +  
         theme(axis.title.y=element_blank(), axis.text.y=element_blank()) + scale_y_discrete(limits=as.character(1:nrow(df)))

### 2025-06-27 text/40-interpretingEffects.qmd
new version

@@
 
-GLMs are defined by a link function $g$ that connects a variable —- linearly defined by the predictors -— to a response variable $Y$, which follows an arbitrary distribution.
+GLMs are defined by a link function $g$ that connects a variable -- linearly defined by the predictors -- to a response variable $Y$, which follows an arbitrary distribution.
 For an experimental design with two factors, $x_1$ and $x_2$, the expected value (or mean) of $Y$ conditional on $x_1$ and $x_2$ is expressed as follows:
@@
 Unless we explicitely mention interpretation issues, we focus on the first scenario.
-Note that in the case of discrete response variables, the function $\mathcal{T}$ is monotonic but not $strictly$ monotonic, since it involves a loss of information.
+Note that in the case of discrete response variables, the function $\mathcal{T}$ is monotonic but not *strictly* monotonic, since it involves a loss of information.
 Consequently, an effect on the latent variable may not be observable in the responses, or the observed magnitude of the effect may appear reduced.

### 2025-06-27 text/50-ExperimentalMethod.qmd
new version

@@
 3.
-We evaluate three sample sizes, $n=10$, $n=20$, and $n=30$, where $n$ represents the cell size in an experimental design.
-For within-subjects designs where all factors are treated as repeated measures, $n$ coincides with the number $N$ of subjects (or commonly human participants in HCI research).
-In contrast, for a 2 $\times$ 3 between-subjects design, if the cell size is $n = 20$, then the number of subjects is $N = 120$.
+We focus on three sample sizes, $n=10$, $n=20$, and $n=30$, where $n$ represents the cell size in an experimental design.
+In within-subjects designs, where all factors are treated as repeated measures, $n$ corresponds to the number of subjects $N$ (commonly human participants in HCI research).
+In contrast, in a 2 $\times$ 3 between-subjects design, a cell size of $n = 20$, implies a total of $N = 120$ subjects.
+For ordinal data only, we also report results for a larger range of sample sizes, up to $n = 512$.
 
-4. We test the robustness of the methods under unequal variances.
+4. We test the robustness of the methods when the variances on the latent variable are unequal.
 
-5. In addition to Type I error rates, we compare the statistical power of the methods and also evaluate the quality of their effect size estimates.
+5. In addition to Type I error rates, we compare the statistical power of the methods and assess the accuracy of their effect size estimates.
 
-6. We assess the above measures for both main and interaction effects and examine how growing effects on one or two factors affect the Type I error rate on other factors and their interactions.
+6. We analyze both main and interaction effects, examining how increasing the effect size of one or two factors influences the Type I error rates of other factors and their interactions.
 
@@
 
-Previous evaluations of rank transformation methods [@Beasley:2009; @luepsen:2018] have also tested unbalanced designs, where the cell size is not constant across all levels of a factor.
-When combined with unequal variances, unbalanced designs are often problematic for both parametric procedures [@blanca:2018] and rank transformation methods [@Beasley:2009; @luepsen:2018].
-As we mentioned earlier, we do not evaluate unbalanced designs here.
+Previous evaluations of rank transformation methods [@Beasley:2009; @luepsen:2018] have also examined unbalanced designs, where cell sizes vary across the levels of a factor.
+When combined with unequal variances, such designs often pose challenges for both parametric procedures [@blanca:2018] and rank transformation methods [@Beasley:2009; @luepsen:2018].
+As noted earlier, we do not include unbalanced designs in our evaluation.
 However, we provide additional experimental results on missing data in the [appendix](appendix.html).
@@
 In addition to power, we assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted on the latent variable $\mathcal{Y}$.
-While partial $\eta^2$ is the most commonly used effect size measure, we chose to base our analysis on Cohen's $f$, as its expected value is proportional to the real magnitude of effect.
+While partial $\eta^2$ is the most commonly used effect size measure, we also evaluate Cohen's $f$, as its expected value is proportional to the real magnitude of effect.
 However, $\eta^2$ can be directly derived from Cohen's $f$ as follows:
@@
 
-As we explained earlier, interaction effects are deformed when we transform the latent variable $\mathcal{Y}$ to obtain the observed responses.
-Such transformations also affect the Type I error rates and power that we observe.
-Here, we are interested in evaluating these measures with respect to the effects we apply to the latent variable $\mathcal{Y}$.
-We make this distinction whenever it is relevant in our analysis.
+As explained earlier, interaction effects are distorted when the latent variable $\mathcal{Y}$ is transformed to produce the observed responses.
+These transformations also influence the Type I error rates and statistical power we observe.
+In this analysis, we focus on evaluating these measures with respect to the effects applied to the latent variable $\mathcal{Y}$.
+We explicitly make this distinction whenever it is relevant to our discussion.
 

### 2025-06-27 text/65-Results-EffectSize.qmd
new version

@@
     hovermode = "x unified",  # Shared hover on the x-axis
-    yaxis = list(spikemode = "across", spikesnap = "cursor")
+    yaxis = list(spikemode = "across", spikesnap = "cursor"),
+    xaxis = list(fixedrange=T)
   )
@@
 fig
-
-df <- data[data$effectX1 == 1,]
 ```
-Boxplots summarizing the $\eta^2$ estimates obtained with each method for 300 data points.
-The population effect of $X1$ is null, and responses follow log-normal distributions.
+Boxplots summarizing the $\eta^2$ estimates obtained with each method for 300 data points ($n=20$).
+The population effect of $X1$ is null, and responses follow **log-normal distributions**.
 Hover over the plots to compare the methods.
@@
 fig
-
-df <- data[data$effectX1 == 1,]
 ```
-Boxplots summarizing the $\eta^2$ estimates obtained with each method for 300 data points.
+Boxplots summarizing the $\eta^2$ estimates obtained with each method for 300 data points ($n=20$).
 The population effect of $X1$ is now non-null ($a_1 = 1$).
-Responses follow again log-normal distributions.
+Responses follow again **log-normal distributions**.
 Hover over the plots to compare the methods.
@@
 
+**Interactions**.
+We also provide additional results illustrating how rank-based methods can fail to accurately estimate interaction effects.
+@fig-eta-squared-interaction shows the estimated values of partial $\eta^2$ for the interaction term under normal distributions, as both main effects increase.
+Once again, we observe that RNK can substantially distort the estimated interaction.
+While INT reduces this distortion, the problem reappears as the main effects grow larger.
+
+::: {#fig-eta-squared-interaction}
+```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
+data <- read.csv("data/effect-sizes-normal-interactions.csv", sep=",", header=TRUE, strip.white=TRUE)
+data$effectX2 <- factor(data$effectX2)
+
+ylab = "interaction - partial eta squared"
+xlab = "magnitude of main effects: a1 and a2"
+
+margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
+cbPalette = c("#8888ff", "#888888", "#E69F00", "#009E73", "#FF5E00")
+cbPalette2 = c("#8888ff88", "#88888888", "#E69F0088", "#009E7388", "#FF5E0088")
+
+df <- data
+
+fig1 <- plot_ly(df, x = ~effectX2, y = ~X1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1X2, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
+fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
+fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
+fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))
+
+fig <- subplot(fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
+  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
+	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab), xaxis = list(fixedrange=T)) %>% layout(
+    hovermode = "x unified",  # Shared hover on the x-axis
+    yaxis = list(spikemode = "across", spikesnap = "cursor")
+  )
+
+fig
+```
+Boxplots summarizing the $\eta^2$ estimates for interactions obtained with each method for 300 data points ($n=20$).
+The population interaction effect is null, and responses follow **normal distributions**.
+Hover over the plots to compare the methods.
+:::
+
+If the data take a discrete form, effect size estimates become even more distorted across all methods. @fig-eta-squared-ordinal presents results for an ordinal scale with 5 flexible levels.
+
+::: {#fig-eta-squared-ordinal}
+```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
+data <- read.csv("data/effect-sizes-ordinal-interactions.csv", sep=",", header=TRUE, strip.white=TRUE)
+data$effectX2 <- factor(data$effectX2)
+
+ylab = "interaction - partial eta squared"
+xlab = "magnitude of main effects: a1 and a2"
+
+margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
+cbPalette = c("#8888ff", "#888888", "#E69F00", "#009E73", "#FF5E00")
+cbPalette2 = c("#8888ff88", "#88888888", "#E69F0088", "#009E7388", "#FF5E0088")
+
+df <- data
+#df <- data[data$effectX1 == 0,]
+
+fig1 <- plot_ly(df, x = ~effectX2, y = ~X1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1X2, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
+fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
+fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
+fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))
+
+fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
+  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
+	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab), xaxis = list(fixedrange=T)) %>% layout(
+    hovermode = "x unified",  # Shared hover on the x-axis
+    yaxis = list(spikemode = "across", spikesnap = "cursor")
+  )
+
+fig
+```
+Boxplots summarizing the $\eta^2$ estimates for interactions obtained with each method for 300 data points ($n=20$).
+The population interaction effect is null, and responses are **ordinal with 5 flexible levels**.
+Hover over the plots to compare the methods.
+:::
+
+As discussed extensively earlier, such ordinal scales can distort interaction trends when main effects are non-zero, leading to interpretation issues.
+Although the methods may detect strong interaction effects under these conditions, the effects may not be intrinsic or meaningful.
+
 <!--

### 2025-06-27 supplementary/rebuttals/effect-size/test-effect-sizes-normal-interaction.R

@@ -1,0 +1,118 @@
+# Author: Theophanis Tsandilas, 2023-2024
+# Inria & Université Paris-Saclay
+
+# Experiment evaluating the effect size estimates of PAR, ART, RNK, and INT.
+# You can edit the experimental parameters in the code to test different conditions.
+
+
+rm(list=ls())
+library(lmerTest)
+library(effectsize) # Effect size functions
+
+# The ARTool
+library(ARTool)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+source("data-generators.R") # Import the data-generation functions
+
+
+eta_values = function(model) {
+	return(round(eta_squared(anova(model))[, 2], digits=4))
+}
+
+# Repeated measures with two factors
+test.within.2f <- function(n, design, effects, distr){ 
+	df <- sim.within.2f(n, design, effects) %>% toDistribution(distr)
+
+	model = suppressMessages(lmer(dv ~ x1*x2 + (1|s), data=df)) # This is the reference groundtruth model
+
+	model0 <- suppressMessages(lmer(y ~ x1*x2 + (1|s), data=df)) # Parametric
+	model1 <- suppressMessages(lmer(rank(y) ~ x1*x2 + (1|s), data=df)) # RNK
+	model2 <- suppressMessages(art(y ~ x1*x2 + (1|s), data=df)) # ARTool
+	model3 <- suppressMessages(lmer(inverseNormalTransform(y) ~ x1*x2 + (1|s), data=df)) # INT
+
+	c(effects, eta_values(model), eta_values(model0), eta_values(model1), eta_values(model2), eta_values(model3))
+}
+
+# Performs repetitive tests for the given configuration
+test <- function(repetitions, n, design, distr, effects) {
+	results <- foreach(rid = 1:repetitions, .combine=rbind) %dopar% {
+	#	effects = c(runif(1, min=-maxeffects[1], max=maxeffects[1]), maxeffects[2], maxeffects[3])
+	effects = c(effects[1], effects[2], effects[3])
+
+		tryCatch(
+			{
+				test.within.2f(n, design, effects, distr)
+			}, 
+			error = function(cond) {
+				# do nothing
+			}, finally = {
+				# do nothing
+			}
+		)
+	}
+	
+	designStr <- paste(design, collapse="x")
+
+	# Split the results into separate rows 
[…]

### 2025-06-27 text/20-background.qmd
small correction

@@
 Interestingly, identical responses (e.g., the two values of $0.3$) can be assigned very different ranks, while responses that differ substantially (e.g., $0.3$ and $3.0$) may receive the same rank.
-More surprisingly, larger values can receive lower ranks —-- revealing that ART is a non-monotonic transformation.
+More surprisingly, larger values can receive lower ranks —- revealing that ART is a non-monotonic transformation.
 For instance, for subject $s_1$ under condition $b_1$, the responses are $0.4$ and $0.8$ for $a_1$ and $a_2$, yet their respective $ART_A$ ranks are $7.5$ and $3.0$.
@@
 For example, running a repeated-measures ANOVA on these ART ranks yields a *p*-value of $.044$ for the effect of $A$.
-In contrast, using a repeated-measures ANOVA on the ranks of simple rank transformation produces a *p*-value of $.46$, while applying a log-transformation --- more appropriate for this type of data --- yields a *p*-value of $.85$.
+In contrast, using a repeated-measures ANOVA on the ranks of simple rank transformation produces a *p*-value of $.46$, while applying a log-transformation -- more appropriate for this type of data -- yields a *p*-value of $.85$.
 

### 2025-06-30 text/10-intro.qmd
many changes
preparing new version

@@
 
-However, given the significant influence of such milestone methods on our community's research outcomes, they warrant a closer scrutiny.
-Our results suggest that ART's popularity step largely from an ignorance of the method's assumptions.
-Researchers often treat ART as a general-purpose nonparametric alternative when ANOVA's assumptions are violated, overlooking the fact that ART's alignment mechanism introduces stricter assumptions than ANOVA itself, including a linear relationship between the response variable and the effects, equal variances, and continuous responses.
+However, given their substantial impact on research outcomes in our community, these widely adopted methods deserve closer scrutiny.
+As the first author previously argued [@tsandilas:2018], the HCI community often lacks the necessary expertise to critically assess the validity of statistical techniques.
+Our findings suggest that the widespread use of ART largely stems from a lack of awareness regarding its underlying assumptions.
+Many researchers treat ART as a general-purpose nonparametric alternative when ANOVA's assumptions are violated, overlooking the fact that ART’s alignment mechanism imposes even stricter assumptions -- including a linear relationship between the response and predictors, equal variances, and continuous data.
 

### 2025-06-30 text/20-background.qmd
preparing new version

@@
 ![Citations of the CHI 2011 paper on ARTool [@wobbrock:2011] as shown in Google Scholar on 28 April 2024.
-We analyzed the 439 citing bibliographic entries of 2023 in more detail.](images/ARTool-citations-28-april-2024.png){width=80% #fig-ARTool-citations}
+We analyzed the 439 citing bibliographic entries of 2023 in more detail.
+*Note: Citations have continued to increase throughout the 2024–2025 period.*](images/ARTool-citations-28-april-2024.png){width=80% #fig-ARTool-citations}
 

### 2025-06-30 text/80-Recommandations.qmd
preparing new version

@@
 
-Several papers we reviewed, citing ARTool [@wobbrock:2011], justify their use of ART based on the presence of *"nonparametric data."* 
-However, it is important to clarify that terms like "parametric" or "nonparametric" refer to properties of the statistical model, not inherent characteristics of the collected data.
+Several papers we reviewed, citing ARTool [@wobbrock:2011], justify their use of ART based on the presence of *"nonparametric data."* However, it is important to clarify that terms like "parametric" or "nonparametric" refer to properties of the statistical model, not inherent characteristics of the collected data.
 Even when ANOVA is clearly unsuitable, there are often alternative parametric methods for addressing violations of common statistical assumptions and still ensure that results can be easily communicated and interpreted.
@@
 INT appears to reduce the impact of this problem but without offering a satisfying solution.
-We also found that at least for 5-level items with flexible thresholds, all methods may lead to unprecise effect size estimates.
+We also found that at least for 5-level items with flexible thresholds, all methods may lead to imprecise effect size estimates.
 In conclusion, while ANOVA seems to be a valid method for hypothesis testing even based on the analysis of individual Likert-item responses, we identified situations that require special attention.
@@
 Since the HCI community is not familiar with these models, we found it necessary to provide motivation for their use and offer additional explanations to reviewers and readers on how to interpret results.
-These models offer numerous advantages, including their ability to support interval estimation and prediction in addition to null hypothesis testing.
+These models offer numerous advantages, including support for interval estimation and prediction in addition to null hypothesis testing.
 However, researchers are required to acquaint themselves with concepts of generalized mixed-effects models or Bayesian inference in order to use them correctly.
+They also come with certain limitations -- for example, the need to carefully specify priors to ensure convergence in certain scenarios, or their reliance on strict assumptions about populations variances and thresholds.
+
 We hope that our examples in the paper and the supplementary materials can inspire the community to give these methods more thoughtful consideration.

### 2025-06-30 text/90-Conclusions.qmd
preparing new version

## Conclusions {#conclusions}
The Align Rank Transformation (ART) was introduced as a remedy for interaction effect distortions caused by the simple rank transformation [@conover:1981]. Early simulation experiments demonstrated ART's robustness for 2-factorial designs, while subsequent tools such as ARTool [@wobbrock:2011;@elkin:2021] extended its application to multifactorial designs, for testing main effects, interactions, and contrasts. The ARTool has gained significant traction, particularly within the fields of Human-Computer Interaction, Augmented Reality, and Virtual Reality, and has been used in the analysis of over a thousand user studies.

However, our experimental findings and examples reveal that ART is defective, performing adequately only under the strict assumptions of linearity, continuous responses with equal variances among groups, and balanced designs. These results corroborate previous warnings regarding ART's performance [@luepsen:2017;@luepsen:2018], leading us to conclude that the method should be abandoned. We recommend prioritizing parametric methods. While we propose the inverse normal transformation (INT) as a generic nonparametric alternative, we identify situations in which this method may also fail. Finally, we caution researchers about the interpretation of interactions.

### 2025-06-30 text/abstract.qmd
preparing new version

@@
 ###### Results
-Our results indicate that ART systematically confounds effects across all the experimental designs we examined.
-Specifically, an effect on one factor may cause ART to incorrectly detect effects on other factors or their interactions.
-We show that the method fails under various conditions, including skewed distributions, binomial distributions, Likert-type data with five or seven ordinal levels, distributions with unequal variances, and experimental designs with missing data.
+Our results indicate that ART fails under a range of discrete distributions, including binomial distributions, and Likert-type data with 5, 7, and 11 ordinal levels.
+In addition, ART systematically confounds effects in skewed distributions or distributions with unequal variances.
+We report results for various experimental designs, including between-subjects, within-subjects, and mixed designs with two or three factors.
 
-Surprisingly, ART operates as expected only under normal distributions with equal variances.
-In all other scenarios, it performs less effectively than simpler rank transformations and, in many situations, such as with ordinal data, it is significantly worse than parametric ANOVA.
+Surprisingly, ART functions as intended only under the strict assumption of a linear model with continuous responses, equal variances, and no missing data.
+In all other scenarios, it performs worse than simpler rank transformation methods and, in many cases -- such as with ordinal data -- significantly worse than parametric ANOVA.
 Our analysis of real-world data further demonstrates that using ART often leads to incorrect conclusions.
-Conversely, the inverse normal transformation emerges as the most versatile and robust nonparametric method, although we identify situations in which this method can also inflate errors or distort effects.
+By contrast, the inverse normal transformation proves to be the most versatile and robust nonparametric method, though we also identify scenarios where it inflates errors or distort effects.
 
-In addition, our results show that testing interactions in the presence of multiple main effects is challenging for all methods.
+In addition, our results show that testing interactions in the presence of multiple main effects remains a challenge for all methods.
 We raise awareness about previous warnings regarding the interpretation of *removable interactions*.

### 2025-06-30 sandbox.qmd
minor corrections

--- 2025-06-24 sandbox.qmd
@@
 
-::: {#fig-effect-main-scatterplots-1}
-```{r, echo=FALSE, fig.height=3.4, message=FALSE, warning=FALSE}
-source("effect_sizes_scatterplots.R")
-prefix <- "6_scatter-Effect_Size-g2"
-df <- readlyDataPoints(prefix,  distributions=c("norm", "lnorm", "likert5B"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)"), effect = "X1X2") %>% toCohensf2()
+::: {#fig-higgins-model}
+```{r, echo=FALSE, message=FALSE, fig.height=2.5, fig.width = 9, warning=FALSE}
+library(plotly)
 
-fig <- plotlyScatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "fX1X2", max = 0.6)
+palette <- c("#E69F00", "#009E73")
+
+logsd <- 0.5
+logm <- -0.3
+
+xs <- seq(0, 10, length.out = 1000)
+y1s <- dlnorm(xs, logm, logsd)
+y2s <- dlnorm(xs - 0.5, logm, logsd)
+
+med1 <- exp(logm)
+med2 <- exp(logm) + 0.5
+
+annot1 <- paste("Log-normal: meanlog = ", logm, ", sdlog = ", logsd)
+annot2 <- paste("Log-normal: meanlog = ", logm, ", sdlog = ", logsd)
+
+fig <- plot_ly() %>%  
+      add_lines(x = xs, y = y1s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot1, hoverinfo = 'text', line=list(color=palette[1])) %>%
+      add_lines(x = xs, y = y2s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot2, hoverinfo = 'text', line=list(color=palette[2])) %>%      
+      add_annotations(x = med1-0.06, y = max(y1s), text = "easy", xref = "x", yref = "y", showarrow = F, font = list(color = palette[1]), xanchor = 'left') %>% 
+      add_annotations(x = med2-0.06, y = max(y2s), text = "hard", xref = "x", yref = "y", showarrow = F, font = list(color = palette[2]), xanchor = 'left') %>%   
+      layout(
+        showlegend = FALSE,
+        xaxis = list(title = "Time (min)", range = c(0, 4.5), showgrid = F, showticks = T, ticks="outside", zeroline = F), 
+        yaxis = list(showgrid = F, showticklabels = F, showline = F, range=c(0, max(y1s) + .05), fixedrange=T)
+      ) %>% 
+	      config(displayModeBar = TRUE, scrollZoom = FALSE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale", "hoverclosest", 'hoverCompare'))  %>% layout(dragmode='pan')
 
 fig
+
 ```
-Scatterplots showing the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground-truth method for the main effect of $X_1$, when $a_1$ and $a_2$ vary within the range $[-2, 2]$.
+The model of @higgins:1990 assumes that distributions have an identical shape across all conditions.
+Here, the log-normal distribution of the time measure is simply shifted to the right as the task difficulty increases.
 :::

### 2025-06-30 text/30-assumptions.qmd
minor corrections

@@
 ::: {#fig-higgins-model}
-```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
+```{r, echo=FALSE, message=FALSE, fig.height=2.5, fig.width = 9, warning=FALSE}
 library(plotly)
@@
 
-Heavy-tailed distributions, such as the log-normal distribution here, commonly arise in nature when measurements cannot be negative or fall below a certain threshold (Limpert et al., 2001), e.g., the minimum time needed to react to a visual stimulus.
+Heavy-tailed distributions, such as the log-normal distribution here, commonly arise in nature when measurements cannot be negative or fall below a certain threshold [@Limbert:2001], e.g., the minimum time needed to react to a visual stimulus.
 In most experimental research, however, this threshold does not shift across conditions while preserving the distribution’s shape.
@@
 ::: {#fig-realistic-model}
-```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
+```{r, echo=FALSE, message=FALSE, fig.height=2.5, fig.width = 9, warning=FALSE}
 library(plotly)
@@
 
-In these distributions, the mean for had tasks also increases, but this increase is not reflected as a simple global shift in the distribution.
+In these distributions, the mean for hard tasks also increases, but this increase is not reflected as a simple global shift in the distribution.
 Instead, the overall shape of the distribution changes.
@@
 ::: {#fig-homoscedastic-model}
-```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
+```{r, echo=FALSE, message=FALSE, fig.height=2.5, fig.width = 9, warning=FALSE}
 library(plotly)

### 2025-07-01 text/30-assumptions.qmd
tiny change

@@
 Then, a single value variation (e.g., turning this 1 to 0) will cause ART’s results to drastically change.
-We will show that this unstable behavior makes ART inflate error rates even for variables with a larger range of discrete values.
+Although this example illustrates an extreme case, ART's instability can manifest in various forms, inflating error rates even when responses span a wider range of discrete values.
 

### 2025-07-04 sandbox.qmd
final corrections on interpretation issues

--- 2025-06-24 sandbox.qmd
@@
 
-::: {#fig-effect-main-scatterplots-1}
-```{r, echo=FALSE, fig.height=3.4, message=FALSE, warning=FALSE}
-source("effect_sizes_scatterplots.R")
-prefix <- "6_scatter-Effect_Size-g2"
-df <- readlyDataPoints(prefix,  distributions=c("norm", "lnorm", "likert5B"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)"), effect = "X1X2") %>% toCohensf2()
+**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects -- that is, when the coefficients of all interacting factors are non-zero.
+Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero.
+In such cases, we may observe a change in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.
 
-fig <- plotlyScatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "fX1X2", max = 0.6)
+By contrast, interpretation is unambiguous in these situations:
 
-fig
-```
-Scatterplots showing the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground-truth method for the main effect of $X_1$, when $a_1$ and $a_2$ vary within the range $[-2, 2]$.
-:::
+- *Main effects*: when the interaction coefficient $a_{12}$ is zero.
+
+- *Interaction effect*: when either $a_{1}$ or $a_{2}$ is zero.
+
+For example, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* ($a_{1} = 0$) and no interaction ($a_{12}=0$).
+
+Throughout the remainder of the article, we make every effort to clarify when the interpretation of an effect is ambiguous.
+Because different methods may rely on different assumptions about the null hypothesis, we analyze such situations separately.
+However, it is important to note that the existing literature offers no guidance on how ART is expected to interpret main and interaction effects under these conditions.
+Early evaluations of ART did not consider nonlinear models, and therefore this issue was not raised.
+While @elkin:2021 does examine such models, they avoid these scenarios and do not address these problems.

### 2025-07-04 text/40-interpretingEffects.qmd
final corrections on interpretation issues

@@
 
-**Defining interactions on the scale of the response variable.** @Ai:2003 and, more recently, @McCabe:2022 argue that the interaction coefficient alone is not sufficient to describe the actual interaction trend on the natural scale of the response variable.
+**Defining interactions on the scale of the response variable.** @Ai:2003 and, more recently, @McCabe:2022 show that the interaction coefficient alone is not sufficient to describe the actual interaction trend on the natural scale of the response variable.
 They argue that an interaction effect should instead be defined as the observed change in the marginal effect of $x_1$ as a function of $x_2$.
-For continuous predictors, this can be formalized as the second-order cross partial derivative of the expected value (which is $\frac{\partial^2 E(Y|x_1, x_2)}{\partial x_1 \partial x_2}$), while for discrete predictors, it can be expressed using discrete differences between levels of the predictors.
+Thus, the trends shown in @fig-interactions do represent interactions, since the difference between *Tech* A and *Tech B* changes as a function of *Difficulty* on the original reponse scale (time or perceived performance).
 
 Using this definition, the authors show that a zero interaction coefficient ($a_{12} = 0$) does not necessarily imply the absence of an interaction effect.
-In fact, for continuous predictors, they demonstrate that  when $a_{12} = 0$, the interaction effect is proportional to $a_1$ and $a_2$.
+In fact, for continuous predictors, they demonstrate that when $a_{12} = 0$, the interaction effect can still emerge and is proportional to $a_1$ and $a_2$.
+
+<!--
+For continuous predictors, this can be formalized as the second-order cross partial derivative of the expected value (which is $\frac{\partial^2 E(Y|x_1, x_2)}{\partial x_1 \partial x_2}$), while for discrete predictors, it can be expressed using discrete differences between levels of the predictors.
+-->
 
@@
 
-*“Statistical testing about the model specification is done at this step.
+> *“Statistical testing about the model specification is done at this step.
 Hypothesis tests are about model coefficients and about the structural aspects of the model specifications.
@@
 
-**Clarifying interpretation boundaries.** Interaction interpretation issues only arise when parallel main effects are present, that is, when the coefficients of all interacting factors are non-zero.
-But interpretation issues can also affect main effects.
-In particular, if the interaction coefficient $a_{12}$ is non-zero, we may observe a change in the marginal effect of a factor (e.g., $x_1$) despite the fact that its corresponding model coefficient (e.g., $a_1$) is zero.
+**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects -- that is, when the coefficients of all interacting factors are non-zero.
+Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero.
+In such cases, we may observe a change in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.
 
-We emphasize that the existing literature does not clarify how ART is expected to interpret main and interaction effects in these scenarios.
-Early evaluations of ART did not consider nonlinear models, so the question was not even posed.
-@elkin:2021 who do study such models, use log-transformed values as their baseline for log-normal distributions (see their example on Page 4).
-So it seems that Elkin et al.
-assume ART's conclusions should align with the coefficients of the linear submodel.
-Unfortunately, their evaluations focus on contrasts and do not account for parallel effects, avoiding interpretation issues.
+By contrast, interpretation is unambiguous in these situations:
 
-Throughout the rest of the article, we make every effort to clarify when the interpretation of an effect is ambiguous.
-Because different methods may rely on different assumptions about the null hypothesis, we analyze such situations separately to avoid confusion.
-Importantly, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* and no interaction.
-Similarly, in the majority of our experiments demonstrating ART's deficient behavior, we avoid interpretation issues by varying the model coefficient of a single factor while keeping all other coefficients at zero.
+1. *Main effects*: when the interaction coefficient $a_{12}$ is zero.
+
+2. *Interaction effect*: when either $a_{1}$ or $a_{2}$ is zero.
+
+For example, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* ($a_{1} = 0$) and no interaction ($a_{12}=0$).
+
+Throughout the remainder of the article, we make every effort to clarify when the interpretation of an effect is ambiguous.
+Because different methods may rely on different assumptions about the null hypothesis, we analyze such situations separately.
+However, it is important to note that the existing literature offers no guidance on how ART is expected to interpret main and interaction effects under these conditions.
+Early evaluations of ART did not consider nonlinear models, and therefore this issue was not raised.
+While @elkin:2021 do examine such models, they do not address interpretation issues.
 

### 2025-07-04 text/65-Results-EffectSize.qmd
small correction
final corrections on interpretation issues

@@
 The results clearly demonstrate that the method's alignment mechanism fails in nonlinear models with skewed distributions and confounds effects.
-
-@fig-eta-squared-2 shows the $\eta^2$ estimates when the effect of $X_1$ is no longer null ($a_1 = 1$).
-In this scenario, PAR consistently underestimates effect sizes, indicating its unsuitability for such data.
-RNK also slightly underestimates effects.
-ART tends to underestimate effects when $a_2 = 0$, but as $a_2$ increases, its estimates become more variable and exhibit an upward bias.
-INT provides the most accurate and stable estimates, with only a slight deviation when $a_2 = 8$.
 
@@
 The mean difference between them is $\overline{B} - \overline{A} = 3$, whether calculated from the row values or their ranks.
-However, if the set $B$ becomes ${9, 10, 11\}$, the mean difference of raw values increases to $8$, while the mean difference in ranks remains $3$.
+However, if the set $B$ becomes $\{9, 10, 11\}$, the mean difference of raw values increases to $8$, while the mean difference in ranks remains $3$.
 As effects become larger, this problem becomes more apparent.

### 2025-07-07 appendix.qmd
updated appendix

@@
 source("plotlying.R")
+source("interactions_plot.R")
 ```
@@
 Note that error rates are not proportional to the $\alpha$ level.
+For results from other experiments, we refer readers to our raw result files.
+
+**Main effects**. @fig-ratio-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases.
 
@@
 
+**Interaction effects**.
+@fig-ratio-interaction-1 presents Type I error rates for the interaction effect $X_1 \times X_2$, when the main effect on $X_2$ is zero while the main effect on $X_1$ increases.
+
 ::: {#fig-ratio-interaction-1}
@@
 :::
+
+**Interactions under parallel main effects**.
+We also present results for interaction effects when the two main effects change in parallel.
+As we explain in the main paper, these results require special attention, as error rates largely depend on the way we define the null hypothesis for interactions.
 
@@
 ```
-Type I error rates ($\alpha = .01$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ 
-:::
-
-For results from different experiments, we refer readers to our raw result files.
-
-### Conclusion
-Results for $\alpha = .01$ confirm that ART is not a robust method. It only behaves correctly when distributions are normal.
-
-## Main effects in the presence of interactions {#interactions}
-In all experiments assessing Type I error rates reported in our article, we assumed no interaction effects.
-However, we also need to understand whether weak or strong interaction effects could affect the sensitivity of the methods in detecting main effects.
-This experiment evaluates the Type I error rate of the methods in the presence of an interaction effect alone, or alternatively, in the presence of a simultaneous main effect.
-We focus again on the three two-factor experimental designs we evaluated for our previous experiment, setting the sample size to $n = 20$.
-
-To simulate populations in which interactions emerge in the absence of main effects, we examine perfectly symmetric cross-interactions.
-To this end, we slightly change the method we use to encode the levels of each factor, such that levels are uniformly positioned around 0.
-For a factor with three levels, we numerically encode the levels as $\{-0.5, 0, 0.5\}$.
-For a factor with four levels, we encode them as $\{-0.5, -0.1667, 0.1667, 0.5\}$.
-
-**Interaction effect only**.
-We first test how the interaction effect alone influences the Type I error rate on $X_2$.
-@fig-interactions-main-2 presents our results.
-We observe that both PAR and ART fail for many configurations.
-Error rates are especially high in the case of the $4 \times 3$ within-subjects design and the $2\times 4$ mixed design under the log-normal and exponential distribution.
-ART exhibits the worst performance.
-In contrast, RNK and INT keep error rates close to nominal levels.
-However, when interaction effects become sufficiently large ($a_{12} > 4$), we observe that under the binomial and ordinal scale, they also start inflating errors.
-
-::: {#fig-interactions-main-1}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-alpha <- .05
-
-prefix <- "Appendix_test_growing-interactions"
-distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
-dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
-
-df <- readData(prefix, n = 20, alpha, effectType = 0, distributions)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX1", xvar = "effectX1X2", max = 104)
-
-```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$ ($n=20$)
-:::
-
-::: {#fig-interactions-main-2}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX2", xvar = "effectX1X2", max = 104)
-
-```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$ ($n=20$)
-:::
-
-**Interaction effect combined with main effect**.
-We also evaluate the Type I error rate on $X_2$ when the interaction effect is combined with a main effect on $X_1$.
-@fig-interactions-main-3 presents our results.
-The error rates of ART and PAR now explode for all non-normal distributions and all three designs.
-But the performance of RNK and INT is also affected.
-Their error rates become extremely low under continuous distributions, which suggests a lack of power in detecting small main effects when strong effects of other factors are combined with strong interactions.
-In contrast, the Type I error of the two methods explodes under the binomial and ordinal scale.
-Interestingly, RNK exhibits the best performance in these tests.
-
-::: {#fig-interactions-main-3}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- readData(prefix, n = 20, alpha, effectType = 6, distributions)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of effects", var = "rateX2", xvar = "effectX1X2", max = 104)
-
-```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitudes $a_1 = a_{12}$ of a combined main effect on $X_1$ and an interaction effect $X_1 \times X_2$ ($n=20$)
+Type I error rates ($\alpha = .01$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$.
+*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
+A different definition may lead to different results.*
 :::
@@
 ### Conclusion
-When distributions are skewed, the presence of an interaction effect can cause ART to detect a non-existent main effect.
-ART is more sensitive to such problems than parametric ANOVA.
-The performance of INT and RNK can also be affected by the presence of interaction effects, especially when the interaction effect is combined with a main effect, and distributions are either binomial or ordinal.
-More generally, main effects should be interpreted with caution when strong interactions exist.
+Results for $\alpha = .01$ are consistent with our results for $\alpha = .05$, confirming that ART is not a robust method.
 
@@
 **Interaction effects**.
-@fig-missing-interaction1 and @fig-missing-interaction2 present Type I error rates for the interaction effect in the presence of a single main effect or two parallel main effects.
+We also present Type I error rates for the interaction effect in the presence of a single main effect or two parallel main effects.
 The error levels for all methods, including ART, are now very similar to the ones observed with no missing data.
@@
 
-::: {#fig-missing-interaction2}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- readData(prefix, n = 20, alpha = .05, effectType = 0, distributions)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 103)
-
-```
-Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitudes $a_{1}$ and $a_{2}$ of the main effects ($n=20$)
-:::
-
 ### Conclusion
@@
 Specifically, we found that its Type I error rate for main effects increases further, even with normal distributions.
-The other methods do not seem to be affected by missing data.
+The other methods do not seem to be affected.
 However, we emphasize that we only tested missing data that are randomly drawn from the complete set of observations.
@@
 
-**Interaction effects**.
-We observe similar patterns for the Type I error rate of the interaction effect in the presence of a single main effect (@fig-lognormal-interaction1) or two parallel main effects (@fig-lognormal-interaction2).
-As shown in @fig-lognormal-interaction2, any advantage of ART over RNK and INT for testing interaction disappears even when distributions exhibit light skew levels.
-We also observe again that the performance of RNK and INT remains identical across all skew levels.
+**Interaction effects**. We observe similar patterns for the Type I error rate of the interaction effect in the presence of a single main effect.
 
@@
 
+**Interactions under parallel main effects**.
+We also present results when the two main effects increase in parallel.
+These results require again careful interpretation, as the null hypothesis of interest may be different for each method.
+It is important to note that even slight departures from normality (e.g., when $\sigma = 0.2$) make the interpretation of interactions problematic.
+
 ::: {#fig-lognormal-interaction2}
@@
 ```
-Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitudes $a_{1}$ and $a_{2}$ of the main effects ($n=20$)
+Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitudes $a_{1}$ and $a_{2}$ of the main effects ($n=20$).
+*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
+A different definition may lead to different results.*
 :::
@@
 ### Conclusion
-ART's robustness issues become less severe as log-normal distributions become less skewed and thus closer to normal.
-However, even under distributions with light skew, ART's Type I error rates for interactions reach higher levels than those of both INT and RNK when parallel main effects are present.
+ART's robustness issues become less severe as log-normal distributions become less skewed and approach normality.
+However, interpretation issues for interactions (in the presence of parallel main effects) can arise even when distributions are only slightly skewed.
+
+## Log-normal distributions with equal variances
+As we explain in the article, skewed distributions naturally lead to unequal variances, with the standard deviation of response times typically proportional to the mean.
+Since ART confounds effects under conditions of unequal variance, an important question remains: *Does ART’s difficulty with continuous skewed distributions stem solely from issues of heteroscedasticity?*
+
+To address this, we present results from an additional experiment using log-normal distributions in which we directly control the population parameters of the observed distributions (rather than those of the latent variable) to ensure constant variances.
+Specifically, given a mean $\mu$ and a fixed standard deviation $\sigma$ for the responses, the corresponding parameters of the log-normal distribution are:
+
+$\sigma_{log} = \sqrt{log(1 + \frac{\sigma^2}{\mu^2})}$ and $\mu_{log} = \log(\frac{\mu^2}{\sqrt{\sigma^2 + \mu^2}})$
+
+We again consider a $4 \times 3$ within-subjects design with $n = 20$, setting the main effect of the first factor to $a_1 = 2$.
+We then measure the Power (effect of $X_1$) and Type I error rates (effects $X_2$ and $X_1 \times X_2$) of the methods across three standard deviations: $\sigma = 1$, $\sigma = 2$, and $\sigma = 3$.
+
+The following table presents our results.
+Once again, ART inflates Type I error rates for both the effect of $X_2$ and its interaction with $X_1$, with error rates increasing as the common standard deviation grows.
+In contrast, PAR shows a loss in statistical power, while RNK and INT remain robust.
+
+::: {#tbl-homoscedastic layout-ncol=3}
+|     | PAR | RNK | INT | ART |
+|:-----:|----:|----:|----:|----:|
+|$X_1$| 100 | 100 | 100 | 100 |
+|$X_2$| 5.0 | 4.9 | 5.4 | 7.6 |
+|$X_1\times X_2$| 3.6 | 4.2 | 4.6 | 5.7 |
+
+: $\sigma = 1$
+
+| PAR | RNK | INT | ART |
+|----:|----:|----:|----:|
+| 98.8 | 100 | 100 | 100 |
+| 4.0 | 5.1  | 5.3 | 12.1 |
+| 3.5 | 5.3 |  5.1 | 9.3  |
+
+: $\sigma = 2$
+
+| PAR | RNK | INT | ART |
+|----:|----:|----:|----:|
+| 94.0 | 100 | 100 | 100 |
+| 3.6 | 4.4  | 4.7 | 17.8 |
+| 3.0 | 4.7  | 5.0 | 11.2 |
+
+: $\sigma = 3$
+
+Percentage of positives (%) over 5000 repetitions ($\alpha = .05$). For $X_1$, values reflect statistical power. For $X_2$ and $X_1 \times X_2$, they represent Type I error rates.
+:::
+
+### Conclusion
+Heteroscedasticity alone cannot explain ART's failures under log-normal distributions.
+ART's alignment method is based on the assumption that the shape of the distribution (log-normal here) is identical across all factor levels.
+This is not the case in the above scenarios where distribution shapes change across difficulty levels, causing the method to break down.
 
@@
 
-**Interaction effects**.
-@fig-binom-interaction1 shows similar patterns for the Type I error rate of the interaction effect in the presence of a single main effect.
-When both main effects increase beyond a certain level (see @fig-binom-interaction2), all methods seem to fail to control the error rate.
-ART again demonstrates the worst behavior, systematically inflating error rates even when main effects are absent.
-In several cases, RNK performs better than INT.
+**Interaction effects**. @fig-binom-interaction1 shows similar patterns for the Type I error rate of the interaction effect in the presence of a single main effect.
 
@@
 
+**Interactions under parallel main effects**.
+When both main effects exceed a certain level (see @fig-binom-interaction2), all methods fail to control the Type I error rate.
+Whether this inflation is due to interpretation issues or to a lack of robustness in the methods themselves, these results once again highlight that testing interactions in such scenarios is highly problematic.
+
 ::: {#fig-binom-interaction2}
@@
 ### Conclusion
-ART is extremely problematic under binomial distributions, raising Type I error at very high levels even in the absence of other effects.
-Testing interactions in the presence of parallel main effects can be problematic for all other methods.
+ART is extremely problematic under binomial distributions, raising Type I error at very high levels even when effects are null across all factors.
+We also observe that testing interactions in the presence of parallel main effects can be problematic for all other methods.
 
@@
 **Interaction effects**.
-@fig-missing-interaction1 and @fig-missing-interaction2 present Type I error rates for the interaction effect in the presence of a single main effect or two parallel main effects.
+We present Type I error rates for the interaction effect in the presence of a single main effect.
 These results lead to similar conclusions.
-Even in cases where ART keeps error rates close to nominal levels (e.g., under the between-subjects design with equidistance thresholds), the performance of parametric ANOVA is constantly better.
+Even in cases where ART keeps error rates close to nominal levels (e.g., under the between-subjects design with equidistance thresholds), the performance of PAR is constantly better.
 
@@
 
+**Interactions under parallel main effects**. Finally, we show results for scenarios where main effects increase in parallel.
+
 ::: {#fig-ordinal-designs-interaction-2}
@@
 ```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$)
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$).
+*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
+A different definition may lead to different results.*
 :::
@@
 For the between-subjects and mixed designs, the problem primarily concerns ordinal scales with flexible thresholds.
-However, for within-subjects designs, ART also inflates error rates for scales with equidistant thresholds, particularly when the number of levels is as low as five or seven.
+For the within-subjects designs, ART also inflates error rates for scales with equidistant thresholds, particularly when the number of levels is as low as five or seven.
 Again, all methods may fail to correctly infer interactions when parallel main effects exceed a certain threshold.
+
+## Main effects in the presence of interactions {#removable-main}
+In all experiments assessing Type I error rates reported in our article, we assumed no interaction effects.
+However, we also need to understand whether weak or strong interaction effects could affect the sensitivity of the methods in detecting main effects.
+Unfortunately, the interpretation of main effects may become ambiguous when interactions are present.
+
+### Illustrative example 
+To understand the problem, let us take a [dataset](removable_main.csv) from a fictional experiment (within-participants design with $n = 24$) that evaluates the performance of two techniques (*Tech A* and *Tech B*) under two task difficulty levels (*easy* vs.
+*hard*).
+@fig-example-interactions visualizes the means for each combination of the levels of the factors, using the original scale of measurements (left) and a logarithmic scale (right).
+Note that time measurements were drawn from log-normal distributions.
+
+::: {#fig-example-interactions}
+```{r, echo=FALSE, message=FALSE, fig.height=3.5, fig.width = 9, warning=FALSE}
+df <- read.csv("removable_main.csv", sep=",", header=TRUE, strip.white=TRUE)
+dftime <- aggregate(Time ~ Difficulty+Technique, data = df, mean)
+
+fig1 <- createInteractionPlot(dftime) %>% layout( yaxis = list(hoverformat = '.2f'))
+fig2 <- createInteractionPlot(dftime, logscale = TRUE) %>% layout( yaxis = list(hoverformat = '.2f'))
+
+fig <- subplot(fig1, fig2, titleY = TRUE, titleX = TRUE, margin = 0.06) %>% layout(hovermode = 'x', dragmode='pan')
+
+fig
+```
+The line charts visualize the effects of task difficulty (*easy* vs. *hard*) and technique (*Tech* A vs. *Tech B*) for task completion time. All data points represent group means. 
+:::
+
+We observe a clear main effect of *Difficulty* and a strong interaction effect, regardless of the scale.
+However, the main effect of *Technique* largely depends on the scale used to present the results.
+Can we then conclude that *Technique A* is overall faster than *Technique B*?
+
+Below, we present the results of the analysis using different methods.
+
+|        | PAR  | LOG  | ART | RNK | INT |
+|--------|------|------|-----|-----|-----|
+| Difficulty  | $7.8 \times 10^{-8}$ | $4.3 \times 10^{-16}$  |  $8.6 \times 10^{-12}$ | $2.5 \times 10^{-16}$ | $1.1 \times 10^{-15}$ |
+| Technique   | $.024$ | $.81$  | $.00017$ | $.69$ | $.69$ |
+| Difficulty $\times$ Technique | $7.6 \times 10^{-6}$ | $4.1 \times 10^{-9}$ | $9.4 \times 10^{-8}$ | $1.1 \times 10^{-9}$ | $1.3 \times 10^{-9}$ |
+: *p*-values for main and interaction effects {.sm}
+
+PAR detects a main effect of *Technique* ($\alpha = .05$), as does ART, which yields an even smaller $p$-value.
+In contrast, LOG, RNK, and INT do not find no evidence for such an effect.
+As with *removable interactions*, we can speak about *removable main effects* in this case.
+
+As a general principle, it makes little sense to interpret main effects when crossing patterns emerge due to a strong interaction.
+In this scenario, the researchers should instead compare techniques within each level of *Difficulty*.
+Accordingly, we can conclude that *Technique A* is slower on *easy* tasks but faster on *hard* tasks in this experiment.
+
+### Experiment
+To better understand how the different methods detect main in the presence of interactions, we conduct an experiment.
+
+We focus again on two-factor experimental designs, setting the sample size to $n = 20$.
+To simulate populations in which interactions emerge in the absence of main effects, we examine perfectly symmetric cross-interactions.
+To this end, we slightly change the method we use to encode the levels of each factor, such that levels are uniformly positioned around 0.
+For a factor with three levels, we numerically encode the levels as $\{-0.5, 0, 0.5\}$.
+For a factor with four levels, we encode them as $\{-0.5, -0.1667, 0.1667, 0.5\}$.
+
+Since the interpretation of main effects can be ambiguous, we define Type I errors with respect to the coefficients of the linear model.
+Accordingly, the null hypothesis is either $a_1 = 0$ or $a_2 = 0$, depending on the variable of interest.
+
+**Interaction effect only**.
+We first examine how the interaction effect alone influences the Type I error rate for $X_2$.
+The results are shown in @fig-interactions-main-2.
+We emphasize that interpretation issues arise only in the case of the $4 \times 3$ design.
+In the other two designs, $X_1$ has only two levels, and due to their symmetry, the observed main effects should be null --- regardless of how we define the null hypothesis.
+
+In the $4 \times 3$ design, error rates for PAR and ART grow rapidly as the interaction effect becomes large under the log-normal and exponential distributions.
+This is largely due to interpretation issues, as illustrated in our example.
+However, ART also fails to control the error rate in other scenarios.
+Finally, under the binomial and ordinal scales, we observe that RNK and INT may also inflate Type I error rates when the interaction effect becomes sufficiently large.
+
+::: {#fig-interactions-main-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+alpha <- .05
+
+prefix <- "Appendix_test_growing-interactions"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+
+df <- readData(prefix, n = 20, alpha, effectType = 0, distributions)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX1", xvar = "effectX1X2", max = 104)
+
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$ ($n=20$).
+*Note: Type I errors are defined based on the null hypothesis $a_{1} = 0$.
+For the $4 \times 3$ design, a different definition may lead to different results.
+However, the two other designs, where $X_1$ has only two levels, Type I error rates are the same, regardless of the definition we use.*
+:::
+
+We also examine error rates for the main effect of $X_2$. Now, interpretation issues affect the results across all three designs, leading to similar patterns.
+
+::: {#fig-interactions-main-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX2", xvar = "effectX1X2", max = 104)
+
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$ ($n=20$).
+*Note: Type I errors are defined based on the null hypothesis $a_{2} = 0$.
+A different definition may lead to different results.*
+:::
+
+**Interaction effect combined with main effect**.
+Finally, we evaluate the Type I error rate on $X_2$ when the interaction effect is combined with a main effect on $X_1$.
+@fig-interactions-main-3 presents our results.
+
+The error rates of ART and PAR now increase dramatically across all non-normal distributions and all three designs --- a result that is, in part, due to interpretation issues.
+RNK and INT are also affected: their error rates become extremely low under continuous distributions, suggesting a lack of power to detect small main effects when strong effects of other factors are combined with strong interactions.
+In contrast, under the binomial and ordinal scales, both methods exhibit inflated Type I error rates.
+Interestingly, RNK demonstrates the best overall performance in these tests.
+
+::: {#fig-interactions-main-3}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- readData(prefix, n = 20, alpha, effectType = 6, distributions)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyErrorByDesign2(df, xlab = "magnitude of effects", var = "rateX2", xvar = "effectX1X2", max = 104)
+
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitudes $a_1 = a_{12}$ of a combined main effect on $X_1$ and an interaction effect $X_1 \times X_2$ ($n=20$).
+*Note: Type I errors are defined based on the null hypothesis $a_{2} = 0$.
+A different definition may lead to different results.*
+:::
+
+### Conclusion
+When strong interactions are present, the interpretation of main effects can become ambiguous, especially when the interaction is combined with a non-zero main effect.
+ART is more sensitive to these issues than other methods and may even inflate Type I error rates in scenarios where interpretation problems are not expected.
+More broadly, main effects should be interpreted with caution in the presence of strong interactions.
 
@@
 We emphasize that @Salter:1993 only apply mean and median alignment to interactions.
-Our implementation for main effects is based on the alignment approach of @wobbrock:2011, where we simply replace means by medians.
+Our implementation for main effects is based on the alignment approach of @wobbrock:2011, where we simply replace means by medians --- we are not aware of more adapted methods.
 
 **Main effects**.
-Our results presented in @fig-median-main demonstrate that median alignment (ART-MED), or at least our implementation of the method, is not appropriate for testing main effects.
+Our results presented in @fig-median-main demonstrate that median alignment (ART-MED) --- or at least our implementation of the method --- is not appropriate for testing main effects.
 Although Type I error rates are now lower for the Cauchy distribution compared to the original method, they are still above nominal levels.
@@
 In contrast, median alignment works surprisingly well for interactions, correcting deficiencies of ART, especially when main effects are absent or weak.
-@fig-median-interaction-1 and @fig-median-interaction-2 present our results.
 Despite this improved performance, we cannot recommend using the method because it still cannot compete with INT.
@@
 
+**Interactions under parallel main effects**.
+We also evaluate Type I error rates for interactions in the presence of parallel main effects, where we caution readers about intepretation issues.
+We observe again that median alignment improves ART's bevahior across all distributions.
+However, as with PAR, its definition of the null hypothesis aligns with the scale of responses rather than with the model parameters.
+
 ::: {#fig-median-interaction-2}
@@
 ```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ 
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$.
+*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
+A different definition may lead to different results.*
 :::
@@
 ### Conclusion
-Using median instead of mean alignment with ART significantly improves the method's performance in testing interactions across all the distributions we tested.
-However, we cannot recommend it, as the method is still less robust than INT.
-Furthermore, it is unclear how to apply median alignment for testing main effects --- using medians with the alignment method of @wobbrock:2011 results in extremely high error rates.
+Using median instead of mean alignment in ART significantly improves the method's performance for testing interactions across all the distributions we examined.
+However, we cannot recommend it, as ART remains less robust than INT, and its advantages over PAR are unclear.
+Moreover, it is not evident how to apply median alignment for testing main effects --- using medians with the alignment procedure of @wobbrock:2011 leads to extremely high error rates.
 

### 2025-07-07 sandbox.qmd
updated appendix

@@
 
-**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects -- that is, when the coefficients of all interacting factors are non-zero.
-Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero.
-In such cases, we may observe a change in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.
+## Main effects in the presence of interactions {#interactions}
+In all experiments assessing Type I error rates reported in our article, we assumed no interaction effects.
+However, we also need to understand whether weak or strong interaction effects could affect the sensitivity of the methods in detecting main effects.
+Unfortunately, the interpretation of main effects may become ambiguous when interactions are present.
 
-By contrast, interpretation is unambiguous in these situations:
+To understand the problem, let us take a [dataset](removable_main.csv) from a fictional experiment (within-participants design with $n = 24$) that evaluates the performance of two techniques (*Tech A* and *Tech B*) under two task difficulty levels (*easy* vs.
+*hard*).
+@fig-example-interactions visualizes the means for each combination of the levels of the factors, using the original scale of measurements (left) and a logarithmic scale (right).
+Note that time measurements were drawn from log-normal distributions.
 
-- *Main effects*: when the interaction coefficient $a_{12}$ is zero.
+::: {#fig-example-interactions}
+```{r, echo=FALSE, message=FALSE, fig.height=3.5, fig.width = 9, warning=FALSE}
+df <- read.csv("removable_main.csv", sep=",", header=TRUE, strip.white=TRUE)
+dftime <- aggregate(Time ~ Difficulty+Technique, data = df, mean)
 
-- *Interaction effect*: when either $a_{1}$ or $a_{2}$ is zero.
+fig1 <- createInteractionPlot(dftime) %>% layout( yaxis = list(hoverformat = '.2f'))
+fig2 <- createInteractionPlot(dftime, logscale = TRUE) %>% layout( yaxis = list(hoverformat = '.2f'))
 
-For example, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* ($a_{1} = 0$) and no interaction ($a_{12}=0$).
+fig <- subplot(fig1, fig2, titleY = TRUE, titleX = TRUE, margin = 0.06) %>% layout(hovermode = 'x', dragmode='pan')
 
-Throughout the remainder of the article, we make every effort to clarify when the interpretation of an effect is ambiguous.
-Because different methods may rely on different assumptions about the null hypothesis, we analyze such situations separately.
-However, it is important to note that the existing literature offers no guidance on how ART is expected to interpret main and interaction effects under these conditions.
-Early evaluations of ART did not consider nonlinear models, and therefore this issue was not raised.
-While @elkin:2021 does examine such models, they avoid these scenarios and do not address these problems.
+fig
+```
+The line charts visualize the effects of task difficulty (*easy* vs. *hard*) and technique (*Tech* A vs. *Tech B*) for task completion time. All data points represent group means. 
+:::
+
+We observe a clear main effect of *Difficulty* and a strong interaction effect, regardless of the scale.
+However, the main effect of *Technique* largely depends on the scale used to present the results.
+Can we then conclude that *Technique A* is overall faster than *Technique B*?
+
+Below, we present the results of the analysis using different methods.
+
+|        | PAR  | LOG  | ART | RNK | INT |
+|--------|------|------|-----|-----|-----|
+| Difficulty  | $7.8 \times 10^{-8}$ | $4.3 \times 10^{-16}$  |  $8.6 \times 10^{-12}$ | $2.5 \times 10^{-16}$ | $1.1 \times 10^{-15}$ |
+| Technique   | $.024$ | $.81$  | $.00017$ | $.69$ | $.69$ |
+| Difficulty $\times$ Technique | $7.6 \times 10^{-6}$ | $4.1 \times 10^{-9}$ | $9.4 \times 10^{-8}$ | $1.1 \times 10^{-9}$ | $1.3 \times 10^{-9}$ |
+: *p*-values for main and interaction effects {.sm}
+
+PAR detects a main effect of *Technique* ($\alpha = .05$), as does ART, which yields an even smaller $p$-value.
+In contrast, LOG, RNK, and INT do not find no evidence for such an effect.
+As with *removable interactions*, we can speak about *removable main effects* in this case.
+
+As a general principle, it makes little sense to interpret main effects when crossing patterns emerge due to a strong interaction.
+In this scenario, the researchers should instead compare techniques within each level of *Difficulty*.
+Accordingly, we can conclude that *Technique A* is slower on *easy* tasks but faster on *hard* tasks in this experiment.
+
+In the $4 \times 3$ design, error rates for PAR and ART grow rapidly as the interaction effect becomes large under the log-normal and exponential distributions.
+This is largely due to interpretation issues, as illustrated in our example.
+However, ART also fails to control the error rate in other scenarios.
+Finally, under the binomial and ordinal scales, we observe that RNK and INT may also inflate Type I error rates when the interaction effect becomes sufficiently large.

### 2025-07-07 text/10-intro.qmd
updated appendix

@@
 Our findings suggest that the widespread use of ART largely stems from a lack of awareness regarding its underlying assumptions.
-Many researchers treat ART as a general-purpose nonparametric alternative when ANOVA's assumptions are violated, overlooking the fact that ART’s alignment mechanism imposes even stricter assumptions -- including a linear relationship between the response and predictors, equal variances, and continuous data.
+Many researchers treat ART as a general-purpose nonparametric alternative when ANOVA's assumptions are violated, overlooking the fact that ART’s alignment mechanism imposes even stricter assumptions --- including a linear relationship between the response and predictors, equal variances, and continuous data.
 

### 2025-07-07 text/30-assumptions.qmd
updated appendix

@@
 
-logm1 <- meanlog(0.8, 0.4)
-logm2 <- meanlog(1.4, 0.4)
+logm1 <- meanlog(0.8, 1)
+logm2 <- meanlog(1.4, 1)
 
-logsd1 <- sdlog(0.8, 0.4)
-logsd2 <- sdlog(1.4, 0.4)
+logsd1 <- sdlog(0.8, 1)
+logsd2 <- sdlog(1.4, 1)
 
@@
 
-med1 <- 0.68
-med2 <- 1.40
+x1 <- 0.36
+x2 <- 1.30
 
@@
       add_lines(x = xs, y = y2s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot2, hoverinfo = 'text', line=list(color=palette[2])) %>%      
-      add_annotations(x = med1, y = max(y1s), text = "easy", xref = "x", yref = "y", showarrow = F, font = list(color = palette[1]), xanchor = 'left') %>% 
-      add_annotations(x = med2, y = max(y2s), text = "hard", xref = "x", yref = "y", showarrow = F, font = list(color = palette[2]), xanchor = 'left') %>%   
+      add_annotations(x = x1, y = max(y1s), text = "easy", xref = "x", yref = "y", showarrow = F, font = list(color = palette[1]), xanchor = 'left') %>% 
+      add_annotations(x = x2, y = max(y2s), text = "hard", xref = "x", yref = "y", showarrow = F, font = list(color = palette[2]), xanchor = 'left') %>%   
       layout(

### 2025-07-07 text/40-interpretingEffects.qmd
updated appendix

@@
 
-**Dealing with interpretation issues.** To avoid such interpretation issues, we focus -- unless otherwise stated -- on effects that apply monotonic transformations to population distributions.
+**Dealing with interpretation issues.** To avoid such interpretation issues, we focus --- unless otherwise stated --- on effects that apply monotonic transformations to population distributions.
 This also ensures a monotonic relationship between different measures of central tendency such as medians and means (with the exception of the Cauchy distribution, where the mean is undefined).
@@
 In conclusion, focusing solely on the issues with the rank tranformation illustrated in @fig-interactions-rank is akin to missing the forest for the trees.
-When parallel main effects are present, interpreting interactions can be challenging for all methods -- and ART offers no solution to this problem.
+When parallel main effects are present, interpreting interactions can be challenging for all methods --- and ART offers no solution to this problem.
 
@@
 
-GLMs are defined by a link function $g$ that connects a variable -- linearly defined by the predictors -- to a response variable $Y$, which follows an arbitrary distribution.
+GLMs are defined by a link function $g$ that connects a variable --- linearly defined by the predictors --- to a response variable $Y$, which follows an arbitrary distribution.
 For an experimental design with two factors, $x_1$ and $x_2$, the expected value (or mean) of $Y$ conditional on $x_1$ and $x_2$ is expressed as follows:
@@
 
-**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects -- that is, when the coefficients of all interacting factors are non-zero.
+**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects --- that is, when the coefficients of all interacting factors are non-zero.
 Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero.

### 2025-07-07 text/66-Results-Interactions.qmd
updated appendix

@@
 In this scenario, the interpretation of an interaction effect can become ambiguous.
-Here, we define the null hypothesis in terms of the coefficients of the latent linear model, following the recommendation of @Greene:2010 -- that is, $a_{12} = 0$.
+Here, we define the null hypothesis in terms of the coefficients of the latent linear model, following the recommendation of @Greene:2010 --- that is, $a_{12} = 0$.
 
@@
 3.
-The purpose of these results is not to provide precise benchmarks, but rather to illustrate the complexity and pitfalls of interpreting interaction effects under assumption violations -- regardless of the method used.
+The purpose of these results is not to provide precise benchmarks, but rather to illustrate the complexity and pitfalls of interpreting interaction effects under assumption violations --- regardless of the method used.
 Many researchers are unaware of these issues and often misinterpret the $p$-value when testing interactions.

### 2025-07-07 supplementary/rebuttals/higgins/homoscedastic/experiment/data-generators.R

@@ -1,4 +1,7 @@
-# Author: Theophanis Tsandilas, Inria & Université Paris-Saclay
+# Author: Theophanis Tsandilas, 2023-2024
+# Inria & Université Paris-Saclay
+
+# Several functions used by our eperimental code, mainly for data generation and transformation
 
 # A great library for simulating data for a large range of desings: https://debruine.github.io/faux
 library(faux)
@@ -14,6 +17,37 @@ inverseNormalTransform = function(x){
 	qnorm((rank(x) - 0.5)/length(x))
 }
 
+# Transforms a set of values to ordinal based on the given thresholds
+# See: https://doi.org/10.1016/j.jesp.2018.08.009
+toOrdinal = function(values, thresholds = c(-2,-1, 1,2)) {
+	discretizeValue = function(x){
+		for(i in 1:length(thresholds)) {
+			if(x <= thresholds[i]) return(i)
+		}
+		return(length(thresholds) + 1)
+	}
+
+	sapply(values, discretizeValue)
+}
+
+
+# Transforms a set of values to ordinal based on the given thresholds
+# See: https://doi.org/10.1016/j.jesp.2018.08.009
+# Here, the thresholds take into account the variance of the data, trying to better cover the full range of values.
+toOrdinalDistr = function(values, nlevels = 5, equidistant = TRUE) {
+	lim <- 2*sd(values) # I take 2 standard deviations that should contain the largest portion of the data (around 95% of the data values)
+
+	thresholds <- NA
+	if(equidistant) {
+		thresholds <- seq(from=-lim, to=lim, by=2*lim/nlevels)[c(-1, -(nlevels + 1))]
+	} else {
+		thresholds <- sort(runif(nlevels - 1))*2*lim-lim
+	}
+
+	toOrdinal(values, thresholds)
+}
+
+
 # Additional support for other distributions with not conversion support at faux
 # Based on https://github.com/debruine/faux/blob/master/R/distribution_convertors.R
 
@@ -28,28 +62,51 @@ norm2lnorm <- function(x, meanlog = 0, sdlog = 1, mu = mean(x), sd = stats::sd(x
  	 stats::qlnorm(p, meanlog, sdlog, ...)
 }
 
+toDistribution <- function(data, distr = "norm") {
[…]

### 2025-07-07 sandbox.qmd
minor correction

@@
 Finally, under the binomial and ordinal scales, we observe that RNK and INT may also inflate Type I error rates when the interaction effect becomes sufficiently large.
+
+**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects --- that is, when the coefficients of all interacting factors are non-zero.
+Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero.
+In such cases, we may observe a change in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.
+We discuss these scenarios in more depth in the [appendix](appendix.html#removable-main).

### 2025-07-07 text/10-intro.qmd
minor correction

@@
 However, given their substantial impact on research outcomes in our community, these widely adopted methods deserve closer scrutiny.
-As the first author previously argued [@tsandilas:2018], the HCI community often lacks the necessary expertise to critically assess the validity of statistical techniques.
-Our findings suggest that the widespread use of ART largely stems from a lack of awareness regarding its underlying assumptions.
+As the first author previously argued [@tsandilas:2018], the HCI community lacks the necessary expertise to critically assess the validity of statistical techniques.
+Our findings suggest that the widespread use of ART largely stems from a lack of awareness of its underlying assumptions.
 Many researchers treat ART as a general-purpose nonparametric alternative when ANOVA's assumptions are violated, overlooking the fact that ART’s alignment mechanism imposes even stricter assumptions --- including a linear relationship between the response and predictors, equal variances, and continuous data.

### 2025-07-07 sandbox.qmd
minor change

--- 2025-07-04 sandbox.qmd
@@
 ```
-
-**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects -- that is, when the coefficients of all interacting factors are non-zero.
-Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero.
-In such cases, we may observe a change in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.
-
-By contrast, interpretation is unambiguous in these situations:
-
-- *Main effects*: when the interaction coefficient $a_{12}$ is zero.
-
-- *Interaction effect*: when either $a_{1}$ or $a_{2}$ is zero.
-
-For example, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* ($a_{1} = 0$) and no interaction ($a_{12}=0$).
-
-Throughout the remainder of the article, we make every effort to clarify when the interpretation of an effect is ambiguous.
-Because different methods may rely on different assumptions about the null hypothesis, we analyze such situations separately.
-However, it is important to note that the existing literature offers no guidance on how ART is expected to interpret main and interaction effects under these conditions.
-Early evaluations of ART did not consider nonlinear models, and therefore this issue was not raised.
-While @elkin:2021 does examine such models, they avoid these scenarios and do not address these problems.

### 2025-07-07 text/30-assumptions.qmd
minor change

@@
 ## Revisiting ART's distributional assumptions {#assumptions}
-Although ART is often regarded as a nonparametric method, its alignment mechanism relies on a set of strict assumptions.
-For a two-way experimental design with independent observation, @higgins:1990 assume the following model for responses:
+ART is often regarded as a nonparametric method, but its alignment mechanism relies on a set of strict assumptions.
+For a two-way experimental design with independent observation, @higgins:1990 assume the following model for the responses:
 
@@
 We explain the severity of these assumptions in more detail.
-
-<!--
-Thus, the critical question we must address is: how does ART perform when its core assumptions are violated?
--->
 
@@
 Would ART perform correctly in this case?
-While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html) showing that as long as distribution shapes differ across levels, ART tends to confound effects.
-Naturally, as the shapes become more similar, the problem diminishes.
-In conclusion, heteroscedasticity may exacerbate the problem, but it is not its sole cause.
 
@@
 :::
+
+While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html) showing that as long as distribution shapes differ across levels, ART tends to confound effects.
+Naturally, as the shapes become more similar, the problem diminishes.
+In conclusion, heteroscedasticity may exacerbate the problem, but it is not its sole cause.
 

### 2025-07-08 text/30-assumptions.qmd
small corrections

@@
 
-While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html) showing that as long as distribution shapes differ across levels, ART tends to confound effects.
+While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html#log-normal-distributions-with-equal-variances) showing that as long as distribution shapes differ across levels, ART tends to confound effects.
 Naturally, as the shapes become more similar, the problem diminishes.

### 2025-07-08 text/62-Results-AcrossExpDesigns.qmd
small corrections

@@
 ### Type I errors across experimental designs {#extra}
-We investigates the generalizability of the aforementioned results to other experimental designs involving two or three factors.
+We investigate the generalizability of the aforementioned results to other experimental designs involving two or three factors.
 We assess five out of the six ratio scales examined earlier.

### 2025-07-08 text/63-Results-UnequalVariance.qmd
small corrections

@@
 We begin by investigating how the four methods detect main effects on $X_1$.
-The interpretation of results in this scenario is challenging due to differences in the original populations -- although they share the same means (and medians), their variances differ.
+The interpretation of results in this scenario is challenging due to differences in the original populations --- although they share the same means (and medians), their variances differ.
 Depending on the null hypothesis of interest, conclusions may vary.

### 2025-07-08 text/66-Results-Interactions.qmd
small corrections

@@
 This problem arises from the way the rank transformation distorts interaction effects (see @fig-interactions-rank).
-For all continuous distributions, we observe that error rates exhibit the exact same patterns, since ranks are scale-invariant.
+For all continuous distributions, we observe that error rates exhibit the exact same patterns.
 
@@
 For all other distributions, error rates increase rapidly as effect sizes grow.
-Two factors explain this behavior: (i) a lack of statistical robustness to violation of the method's assumptions, as also observed in earlier tests; and (ii) interaction interpretation issues.
+Two factors explain this behavior: (i) a lack of statistical robustness to violation of the method's assumptions, as also observed in our previous experiments; and (ii) interaction interpretation issues.
 These results confirm that ART is not scale-invariant, interpreting interactions relative to the scale of the observed data.
@@
 Many researchers are unaware of these issues and often misinterpret the $p$-value when testing interactions.
-As previously noted, @elkin:2021 compared ART to ANOVA on log-transformed data under the belief that ART tests null hypotheses about parameters of the latent linear model.
+Note that @elkin:2021 compared ART to ANOVA on log-transformed data under the belief that ART tests null hypotheses about parameters of the latent linear model.
 Our findings clearly show that this is not the case.

### 2025-07-08 text/70-CaseStudies.qmd
small corrections

@@
 However, we now test a perfectly balanced design.
-We transform the latent variable to a Bernoulli distribution with a success probability parameter $p=.46$ ---average success rate found by @Martin:2023--- and then assess the Type I error rate of ART in the absence of any effect.
+We transform the latent variable to a Bernoulli distribution with a success probability parameter $p=.46$ --- average success rate found by @Martin:2023 --- and then assess the Type I error rate of ART in the absence of any effect.
 We also assess the Type I error rate of regular ANOVA (LMER model) with no transformation (PAR).

### 2025-07-08 text/80-Recommandations.qmd
small corrections

@@
 
-### Priviledging parametric methods
+### Privileging parametric methods
 The scope of rank-based nonparametric statistics is very limited. @baguley2012 summarizes a fundamental concern regarding their use as follows:
@@
 However, researchers are required to acquaint themselves with concepts of generalized mixed-effects models or Bayesian inference in order to use them correctly.
-They also come with certain limitations -- for example, the need to carefully specify priors to ensure convergence in certain scenarios, or their reliance on strict assumptions about populations variances and thresholds.
+These methods also come with certain limitations --- for example, the need to carefully specify priors to ensure convergence in certain scenarios, or their reliance on strict assumptions about population thresholds.
 

### 2025-07-08 text/abstract.qmd
small corrections

@@
 Surprisingly, ART functions as intended only under the strict assumption of a linear model with continuous responses, equal variances, and no missing data.
-In all other scenarios, it performs worse than simpler rank transformation methods and, in many cases -- such as with ordinal data -- significantly worse than parametric ANOVA.
+In all other scenarios, it performs worse than simpler rank transformation methods and, in many cases --- such as with ordinal data --- significantly worse than parametric ANOVA.
 Our analysis of real-world data further demonstrates that using ART often leads to incorrect conclusions.

### 2025-07-08 text/30-assumptions.qmd
correct figure 7

@@
 
-logm1 <- meanlog(0.8, 1)
-logm2 <- meanlog(1.4, 1)
+sd <- 1 # This is the common standard deviation
+m1 <- 0.8
+m2 <- 1.4
 
-logsd1 <- sdlog(0.8, 1)
-logsd2 <- sdlog(1.4, 1)
+logm1 <- meanlog(m1, sd)
+logm2 <- meanlog(m2, sd)
+
+logsd1 <- sdlog(m1, sd)
+logsd2 <- sdlog(m2, sd)
 
@@
 
-annot1 <- paste("Log-normal: meanlog = ", round(logm1, 2), ", sdlog = ", round(logsd1, 2))
-annot2 <- paste("Log-normal: meanlog = ", round(logm2, 2), ", sdlog = ",  round(logsd1, 2))
+annot1 <- paste("Log-normal: mean = ", m1, ", sd = ", sd)
+annot2 <- paste("Log-normal: mean = ", m2, ", sd = ", sd)
 

### 2025-07-08 text/20-background.qmd
minor correction
corrected links

--- 2025-06-13 text/20-background.qmd
@@
 Interestingly, identical responses (e.g., the two values of $0.3$) can be assigned very different ranks, while responses that differ substantially (e.g., $0.3$ and $3.0$) may receive the same rank.
-More surprisingly, larger values can receive lower ranks —-- revealing that ART is a non-monotonic transformation.
+More surprisingly, larger values can receive lower ranks --- revealing that ART is a non-monotonic transformation.
 For instance, for subject $s_1$ under condition $b_1$, the responses are $0.4$ and $0.8$ for $a_1$ and $a_2$, yet their respective $ART_A$ ranks are $7.5$ and $3.0$.
@@
 Only five papers have publicly available data.
-We revisit the results of three of these papers in [Section 6](#casestudies).
+We revisit the results of three of these papers in [Section 7](#casestudies).
 
@@
 With a focus on clarity, we chose to exclusively examine the three rank-based transformations presented earlier: the pure rank transformation (RNK), ART, and INT.
-While we do not elaborate on the performance of ATS in the main paper, additional experimental results are available in the [appendix](appendix.html).
+While we do not elaborate on the performance of ATS in the main paper, additional experimental results are available in the [appendix](appendix.html#ATS).
 These findings indicate that its performance is comparable to the rank transformation but seems to be inferior to the simpler and more versatile INT.
-The appendix features additional results regarding the performance of the generalized van der Waerden test, as well as the generalized Kruskal-Wallis and Friedman tests.
+The [appendix](appendix.html#generalized) features additional results regarding the performance of the generalized van der Waerden test, as well as the generalized Kruskal-Wallis and Friedman tests.
 Our findings show that, at least in balanced designs, these methods suffer from low power issues in certain situations without demonstrating any clear benefits compared to INT.
@@
 The rationale behind this decision is that we have not encountered any prior claims about ART's suitability for unbalanced data, and the ARTool [@artool] issues a warning in such situations.
-However, the [appendix](appendix.html) presents additional results on experimental designs with missing data, which confirm that in such situations, ART's accuracy may deteriorate further.
+However, the [appendix](appendix.html#missing) presents additional results on experimental designs with missing data, which confirm that in such situations, ART's accuracy may deteriorate further.

### 2025-07-18 sandbox.qmd
small correction

@@
+```         
 ---
 title: "The illusory promise of the Aligned Rank Transform"
-subtitle: "A systematic study of rank transformations"
 author: 
@@
     orcid: 0000-0003-1905-815X
-    email: gery.casiezuniv-lille.fr
     affiliations:
@@
 ---
+```
 
@@
 ## Under Review {.unnumbered}
-This paper is [under review](https://www.journalovi.org/under-review.html) on the experimental track of
-the [Journal of Visualization and Interaction](https://www.journalovi.org/). See the 
-[reviewing process](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues).
+
+This paper is [under review](https://www.journalovi.org/under-review.html) on the experimental track of the [Journal of Visualization and Interaction](https://www.journalovi.org/).
+See the [reviewing process](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues).
 :::
@@
 ```
+
+Hello!

### 2025-07-18 text/40-interpretingEffects.qmd
minor correction
small correction

@@
 
-### Controversy about the interpretation of interactions
-We further discuss interpetation issues in the broader context of generalized liner models (GLMs).
+### Disagreement on the interpretation of effects in GLMs
+We further discuss interpetation issues in the broader context of generalized linear models (GLMs).
 
@@
 In such cases, we may observe a change in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.
+We discuss such scenarios in more depth in the [appendix](appendix.html#removable-main).
 

### 2025-12-01 supplementary/rebuttals/CLeyrat/response.md
revised code

---
title: "Responses to our reviewer"
output: github_document
---

**Important:** @CLeyrat @xxx @Lonni 
We have realized that our previous response to @CLeyrat (Comment 18: "More fundamentally, the authors should clarify how the simulation plan with a latent variable relates to real-life data generation mechanisms") was not correct. After further investigation, we have made an important decision to revise our simulation methodology. Below, we explain the issue.

Our latent-variable approach, which generates data by applying a monotonic transformation to a normal distribution, works as intended for log-normal data, and it also remains valid for ordinal outcomes, where we rely on thresholding a latent normal variable. However, this approach fails for the other distributions used in our simulations: Exponential, Cauchy, Binomial, and Poisson.

In these cases, whenever an effect is present (e.g., $a_1 > 0$ or $a_2 > 0$), the transformed distributions no longer follow the intended theoretical distributions. For instance, under the "Cauchy" condition, the generated data become asymmetric, meaning the data are not truly Cauchy-distributed. Although the Type I error and power results we reported are correct for the data we generated, they unfortunately pertain to distributions that we can neither clearly characterize nor theoretically justify.

Given this, and also acknowledging that the latent-modeling approach caused confusion and raised legitimate concerns, we have decided to replace it with a standard simulation framework based on Generalized Linear Models, using the appropriate link functions as described in Equation 4. The only exception concerns ordinal (Likert-like) data, where our original latent-threshold method remains suitable and will be retained.

Because we must now rerun a large number of simulations and substantially revise our methodology and results sections, we anticipate needing additional time. Since all designs in the main article are balanced, we plan to use ANOVA (via R’s aov() function) instead of linear mixed-effects models, as this greatly improves computational efficiency.

@CLeyrat: please let us know if you have any objections to this change.

We sincerely apologize for not identifying these issues earlier, and we thank the reviewer for prompting this important correction.

We thank our reviewer for her time and constructive comments. We provide detailed clarifications below and address each point in turn. We would **greatly appreciate a direct discussion within this thread** to ensure that all points of doubt are adequately addressed.

A brief summary of some key points:

1. **Comparability of methods**. Our article clearly distinguishes between scenarios where interpretation issues due to different definitions of the null hypothesis arise, and those where they do NOT. For most of the scenarios we test, the Type I error rates and power are perfectly comparable across methods---both for main and interaction effects. Comparability issues arise when testing interactions in the presence of parallel main effects (also [in this scenario (Fig. 29)](https://www.journalovi.org/2024-tsandilas-ranktransforms/#fig-unequal-main-1) and [in Appendix: Fig. 16-19](https://www.journalovi.org/2024-tsandilas-ranktransforms/appendix.html#removable-main)), but the article explicitly identifies and discusses these scenarios and is explicit about the null hypothesis we test.

2. **LMER models**. For the scenarios we test, using *lmer* leads to identical (or nearly identical) results to ANOVA. Any potential issues with mixed models have not affected our results.

3. **Categorical variables**. We treat independent variables as categorical, not numerical. Our coding approach is compatible with dummy coding but offers additional control over effect magnitudes.

4. **Data transformations**. The monotonic transformations we apply do not induce selection bias and do not violate any assumptions of random sampling.

We respond below to all individual reviewer comments (numbered from 1 to 28 for easy reference). Regarding presentation suggestions, we note that our reviewer's disciplinary background differs from ours. We kindly ask that the discussion focus primarily on the validity of our research methods and conclusions rather than presentation preferences.

**Note on terminology:** Our reviewer uses the term *estimand*, whereas we refer to population parameters of interest. We prefer our terminology for conceptual clarity and consistency with the terminology most commonly used in disciplines closer to ours.

### "Overall weakness"
**Comment 1.** *"Nevertheless, what is presented as bias or error (for ART) in the manuscript might be explained by something more fundamental: a discrepancy between the estimands the authors wish to target, and the estimands actually targeted with each estimator. Although the authors clearly state that the methods they investigate test different null hypotheses, the simulation study may not fully capture this. My main concern is that comparing these methods is like comparing apples and oranges, but with the aim to show they all look like apples ..."*

[Section 4](https://www.journalovi.org/2024-tsandilas-ranktransforms/#interpretation) already extensively discusses these points. We have also clarified them through our [our responses to Prof. Higgins](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues/2#issuecomment-2528141692), where we present additional tests to explain why our comparisons are correct.

We repeat that unless we explicitly mention interpretational issues, the Type I error rates and power results we report are identical regardless of whether one tests means, medians, or full distributional shape. This holds for the vast majority of our results (e.g., Figures 15–17, 19–24, 30–34). In all these cases, Type I errors are evaluated for identical populations, meaning all population parameters (means, medians, etc.) are equal across levels---both in the latent and the transformed space. Hence, the null hypothesis definition does not affect these comparisons.

We agree that there are situations where interpretation issues arise, but the article is very clear about when this happens. See captions of Figures 25-27, and 29, as well as the related discussion in these paragraphs. As we further discuss later, we are explicit about which null hypothesis we test for these results.

### Specific comments
**Comment 2.** *"... the current structure of the manuscript could be improved ... This is particularly important because the illustrative example comes before any explanation of the methods ... the illustrative example is the least convincing section.*

Our choice to begin with an illustrative example is motivated by closely related work. Higgins et al. (1990) demonstrated the breakdown of rank transformations for interactions using a concrete data example. Elkin et al. (2021) likewise used a specific running example to show how ART fails to handle contrasts, and Lüpsen (2016) illustrated ART’s problems under discrete distributions with a concrete example as well.

Naturally, the examples in all these works are deliberately chosen. Their purpose is to illustrate the problem and provide intuition for readers---not to serve as a systematic proof. It is the role of Monte Carlo experiments to evaluate such issues systematically over a large number of iterations.

**Comment 3.** *"Was this dataset selected at random (as it should be) or selected to illustrate the point? If it is the latter, it may not be completely honest: based on this example, it may only be fair to argue that ART may be more sensitive to outliers."*

The example was chosen to *illustrate* (so we call it *illustrative*) ART’s erratic behavior.

Our results clearly demonstrate that ART inflates Type I errors. However, in many cases, it still produces results that are numerically close to the ground truth. For this reason, we do not see how randomly drawing a dataset would add meaningful insight.

It is also important to note that ART’s *p*-values in our example are orders of magnitude smaller than those returned by other methods---indicating an erratic behavior that is clearly not normal. We do not present this example as a proof, and we would not have included it without first confirming this problematic pattern through systematic simulations.

Even if one interprets these results as showing that ART is more sensitive to "outliers" than other methods, this is already troubling, since previous literature advocating ART explicitly claims the opposite.

**Comment 4.** *"Given the small sample size, I am not confident in commenting on observed differences between p-values."*

We used a small sample to keep the example simple and easily reproducible. We also chose $n=12$ because it is a very common sample size in HCI experiments following a within-participants design (although we know that such a sample size is considered insufficient in other research domains).

Larger samples do not reduce *p*-value variability (ART actually struggles even more with larger samples). Nonetheless, if preferred, we can provide an example with a larger sample, e.g., see ["A challenge for Prof. Higgins"](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues/2#issuecomment-2564006376), where we discuss results for a sample of $n=1000$.

### General comment about the mixed models and the R implementation

**Comment 5:** *"The authors used a random-effect model as an estimator across methods. Was the correlation between pairs of measures for a participant assumed identical no matters the order of the repetitions?"*

We did not model any order effect; our simulation design includes none.

**Comment 6:** *"Also, by default, the lmer package does not provide p-values for main effects. How were the p-values calculated? (Wald test with REML, LRT with ML, etc.). In particular, the authors provide one p-value per variable (even when there are more than two levels), so it is important to clarify how this overall p-value is obtained for categorical variables, given that they are not provided by default."*

As our reviewer remarks, our simulation designs are balanced (with the exception of [these results in the Appendix](https://www.journalovi.org/2024-tsandilas-ranktransforms/appendix.html#missing)), so Type I, Type II, and Type III tests give the same results (see  [Kuznetsova et al., 2017](https://www.jstatsoft.org/article/view/v082i13)). We verified that results are identical (or nearly so) across all these methods:

1. Conducting ANOVA with R's *aov()* function, where *p*-values are then obtained using R's *summary()* function. Example: 
`mpar <- aov(Time ~ Difficulty*Technique + Error(Participant), data=df)`
`summary(mpar)` 
2. Using *lmer* (fit by REML) and then R's *anova()* function to obtain *p*-values (Type III tests). This is the method we use for our all designs with a subject random effect. Example:
`mpar <- lmer(Time ~ Difficulty*Technique + (1|Participant), data=df)`
`anova(mpar)`
3. Using *lmer* (fit by REML) and then R's *car::Anova()* function to obtain *p*-values (Type II Wald F tests). Example:
`mpar <- lmer(Time ~ Difficulty*Technique + (1|Participant), data=df)`
`car::Anova(mpar, type=2, test.statistic = "F")`

We also obtain practically identical results, whether we use a Kenward-Roger (default for *art()* but more computationally expensive as [Kuznetsova et al., 2017](https://www.jstatsoft.org/article/view/v082i13) report) or a Satterthwaite (default for *anova()*) approximation for calculating the *F* test.

For example, the figures below reproduce [Figure 15](https://www.journalovi.org/2024-tsandilas-ranktransforms/#fig-ratio-main) (Type I error rates for ratio scales) and [Figure 19](https://www.journalovi.org/2024-tsandilas-ranktransforms/#fig-ordinal-main) (Type I error rates for ordinal scales) when $n=10$ (smallest sample size we test), using repeated measures ANOVA: 
![](https://notes.inria.fr/uploads/upload_7b6bd501ee3d7f2c12e7a733ad2ee615.png)

![](https://notes.inria.fr/uploads/upload_963e517256cba4fc9b19ec6fa416b78e.png)

Trends are identical with the ones of the original figures (where we use LMER). We can provide additional results upon request.

**Comment 7:**  *"My concern after reading the simulation plan is that the main effects were included as numeric variables ..."*

We treat all experimental factors as categorical, not numeric. All factor levels in our simulated data are strings that take the form "A1", "A2", ... or "B1", "B2", ..., and we use R's *factor()* function before analysis. Our coding method (sum-to-zero contrasts) (explained [here](https://www.journalovi.org/2024-tsandilas-ranktransforms/#statistical-modeling)) ensures symmetry and a grand mean of $\mu$, while allowing controlled manipulation of effect sizes via parameters like $a_1$, $a_2$, and $a_{12}$.

See examples of distributions [(Figure 14)](https://www.journalovi.org/2024-tsandilas-ranktransforms/#fig-effects) we generate with this approach for two, three, or four **categorical** levels.

**Comment 8:** *"... thus making the unrealistic assumption of a linear effect of the experimentation conditions on the outcome. This assumption is not needed, and the different levels of the fixed effects can be included as dummy variables.  In particular, the authors state that both ART and ANOVA require the linearity assumption, but as far as I understand, the variables considered in this paper are all categorical (e.g. easy, medium, hard), included as dummy variables, and therefore no linearity assumption is required."*

All linear regression and ANOVA models are linear regardless of how categorical variables are encoded. Assumptions about a linear relationship between independent variables and responses do not disappear when using dummy coding.

For example, consider the first factor $X_1$ in Equation 9. Suppose it has three levels. Using dummy coding, the term $x_{1i}$ can be expressed as:

$$x_{1i} = \beta_1​ + \beta_2 d_2 + \beta_3 d_3$$

where $d_2$, and $d_3$ are the dummy variables for the three levels of the categorical variable. Our coding simply enforces desirable constraints (zero-centered effects) for $\beta_1$, $\beta_2$, and $\beta_3$. But linearity is intrinsic to the model, not to the coding scheme.

We would appreciate clarification from the reviewer: how would the linearity assumption disappear when using dummy coding?

**Comment 9:** 
*"Finally, mixed-models can lead to incorrect type I error rates in small samples. Did the authors use any small sample correction?"*

No correction was used. Type I error rates are stable across sample sizes in our simulations, and as our results show, Type I error rates are not affected by small sample sizes (down to $n=10$). ART’s problems, in contrast, worsen with larger $n$ (and this is unrelated to the models we use).

### Illustrative example: in addition to my previous comments

**Comment 10:** 
*"'We observe that ART exaggerates both the effect of Technique and its interaction with Difficulty' : this is incorrect, the p-values quantify the strength of evidence, not the magnitude of the effect."*

Our statement refers to the partial $\eta^2$ values shown in the preceding table, not to *p*-values. As we discuss later, such effect size estimates are commonly reported by authors who make use of ART.

**Comment 11:** 
*"Figure 1, y-axis label: should it be Time (rather than median time) unless each dot is a median across other parameters."*

For the first plot, we aggregate over difficulty levels but not for the second one. We will correct the mistake.

**Comment 12:** 
*"Perceived efficiency: despite having in theory 5 levels, in the sample it is almost binary and any method for continuous or discrete data should be prohibited. This example is too extreme and as a consequence may be quite damaging overall. The authors should only compare the performance of methods that are realistic for the setting."*

The data are not binary, though responses cluster at higher scale points. Such skewed ordinal data are common in practice---precisely where nonparametric methods like ART are typically recommended. Demonstrating ART's instability here is therefore important and relevant.

Furthermore, as we discuss [here](https://www.journalovi.org/2024-tsandilas-ranktransforms/#analyzing-likert-type-data), there is ongoing debate regarding the use of parametric techniques with Likert data, thus we therefore disagree with the term "prohibited."

**Comment 13:** 
*"Table 1 (p-values): it would be useful to report the regression coefficient and CI from the mixed models (even though their interpretation differ), along with the p-values, and be consistent with the number of decimal places."*

Below, we show the 95% CIs of the regression coefficients for the LOG method only:

![](https://notes.inria.fr/uploads/upload_ac1cc1b4dddb9e437b143a5ddc355d13.png)

We are concerned that presenting all of these numbers (and for all methods) would only confuse readers. Note that ART produces three separate models, each corresponding to a different alignment: the first factor (*Difficulty*), the second factor (*Technique*), and their interaction. It is unclear which set of coefficients should be reported, or how such information could be meaningfully interpreted.

However, we can replace the mixed-model analysis with a repeated-measures ANOVA (which yields **identical results**)---if this addresses the reviewer's concern. See [ANOVA analysis](https://www.journalovi.org/2024-tsandilas-ranktransforms/supplementary/examples-case-studies/1-Introduction/illustrative-example-aov.html) vs. [LMER](https://www.journalovi.org/2024-tsandilas-ranktransforms/supplementary/examples-case-studies/1-Introduction/illustrative-example.html).

### Interactions
**Comment 14:** *"In the paper, the methods apply different transformations, thus making the coefficients (and inferences) for the interaction parameters not comparable."*

A [full subsection in our article](https://www.journalovi.org/2024-tsandilas-ranktransforms/#interpreting-interaction-effects) is dedicated to this issue. An [additional subsection](https://www.journalovi.org/2024-tsandilas-ranktransforms/#disagreement-on-the-interpretation-of-effects-in-glms) discusses the debate within the context of GLMs, where we explicitly state when the interpretation of interaction is ambiguous (and thus comparisons are problematic) and **when it is not**. Also see [our older response to Matthew Kay](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues/2#issuecomment-2627759256).

We repeat that when either $a_1 = 0$ or $a_2 = 0$ (see Equation 4), there is no ambiguity in the interpretation of Type I errors for the interaction---no matter which method we use, no matter whether we test means or medians, either on the latent or the observed variable. If $a_{12} = 0$, no interaction should appear under any monotonic transformation---so comparing Type I errors across methods (see Figures 17, 20, 21, 24, 31) is valid.

*Note on power:* As discussed in the article, transformations to discrete distributions are monotonic but not strictly monotonic. Thus, an effect in the latent variable may vanish after transformation. This issue affects all methods equally, so power comparisons remain valid.

**Comment 15:** *"Instead of debating whether one is better than the other, the authors should instead explain when one type of interaction may be more relevant than the other for the research question of interest. Thinking about the different experimental settings, when do we care about interaction, and which type of interaction is relevant?"*

This is precisely what we do for the scenarios in which interpretational issues arise--namely, when the main effects of all interacting factors are non-zero:

- Please see our discussion on [the interpretation of effects in GLMs](https://www.journalovi.org/2024-tsandilas-ranktransforms/#disagreement-on-the-interpretation-of-effects-in-glms).

- Please also see our results on [interactions under parallel main effects](https://www.journalovi.org/2024-tsandilas-ranktransforms/#interactions) and the related discussion in this section. Note that the captions of Figures 25, 26, 27 explicitly state that "Type I errors are defined based on the null hypothesis $a_{12} = 0$. A different definition may lead to different results."

### Simulation study
**Comment 16:** *"The authors should also consider whether the simulations on ordinal scales are important here. Many statisticians would argue for the use of appropriate regression models for this type of outcomes (extensions of logistic regression) ...*"

Our article includes a dedicated [discussion on ordinal data](https://www.journalovi.org/2024-tsandilas-ranktransforms/#analyzing-likert-type-data), where we explicitly advocate for the use of appropriate regressions models. However, these methods also present several practical challenges, and, as we discuss extensively, there is no consensus on whether ANOVAs and linear parametric models can be safely applied to ordinal data.

Moreover, in practice, researchers often use nonparametric methods far more frequently than ordinal regression models. As detailed in [this section](https://www.journalovi.org/2024-tsandilas-ranktransforms/#ART-use), ART is very commonly used for exactly this type of data.

We invite our reviewer to consider:

1. [This decision tree](https://github.com/valentin-schwind/statistics-decision-tree/?tab=readme-ov-file), which provides methodological recommendations for HCI researchers. For discrete ordinal data and multiple independent variables, it explicitly recommends ART---ordered probit models are not even mentioned.

2. A simple query to ChatGPT: *"I want to conduct a multifactorial statistical analysis where responses follow an ordinal scale of 5 levels. Which statistical methods could I use?"* ART appears among the five recommended methods ([see here](https://chatgpt.com/share/6910d909-ae20-8008-bb05-8935f1040404)), despite the fact that no empirical studies have demonstrated that ART is valid for such data.

Our results provide clear evidence that ART is **inappropriate** for ordinal data and poses **greater risks** than commonly used parametric methods. We therefore believe it is both relevant and necessary to present these findings.

**Comment 17:** *"Similarly, it is unlikely statisticians pick methods for continuous data for count (Poisson distribution) or binomial distributions. This applies to the re-analyses as well. It may be useful to reduce the amount of examples and scenarios but focus on the key differences between methods."*

Rank transformations (and the nonparametric statistical tests based on them) are in fact very commonly used for both continuous and discrete data. While this may not be typical practice in our reviewer's field, it is widespread in many others. Importantly, researchers ignore that the alignment step in ART implicitly assumes continuous distributions.

It is true that trained statisticians may not typically choose such methods. However, in many scientific domains, data analyses are performed by non-statisticians (ourselves included) and frequently by researchers with limited statistical training. Our work aims precisely to evaluate how these widely used methods behave under realistic conditions encountered in such applied research contexts.

**Comment 18:** *"More fundamentally, the authors should clarify how the simulation plan with a latent variable relates to real-life data generation mechanisms."*

Our simulation plan produces a broad range of distributions that are commonly used as benchmarks in the related literature. A justification for the various ratio scales employed is provided [here](https://www.journalovi.org/2024-tsandilas-ranktransforms/#ratio). For ordinal scales, we used appropriate ordinal models, following approaches such as those described by Liddell and Kruschke (2018).

As we [also explain here](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues/2#issuecomment-2638043122), the great advantage of our methodology lies in its flexibility: it enables the generation of data for any distribution using the same underlying structure, by changing only a single line of code. This also allows us to:

- present results consistently across different distributions by varying the same parameters;
- define a clear ground truth for effect size estimates (see below); and
- systematically study scenarios where interaction interpretation issues arise---defining the null hypothesis with respect to the parameters of the latent linear model, as is standard in regression analyses.

We also provide [supplementary code and results](https://github.com/journalovi/2024-tsandilas-ranktransforms/tree/main/supplementary/rebuttals/kay) using a more traditional simulation methodology, where data are drawn directly from the target distributions (e.g., using the *rlnorm()* function for log-normal samples). The results are fully consistent with those obtained via our latent-variable approach. However, controlling parameters becomes cumbersome with this approach, and presenting results coherently across multiple distributions is considerably more difficult.

We empahsize that our methodology has many strenghts and was chosen after considering several alternative approaches.

**Comment 19:** *"If the true population distribution is normal, but the observed distribution follows a different distribution, it seems to immediately imply a selection bias and a violation of the assumption of random sampling."*

This is incorrect: there is neither selection bias nor any violation of random sampling. If we draw a random sample of the latent variable $Y$ and then apply a monotonic transform $g$, the transformed observations $g(Y_i)$ are an independent and identically distributed sample from $g(Y)$.

Monotonic transformations preserve the rank order and quantiles of the distribution. Specifically, if $q_k$ denotes the $k^{th}$ quantile of $Y$, then $g(q_k)$ is the $k^{th}$ quantile of $g(Y)$.

**Comment 20:** *"Either the null hypothesis relates to the population but the assumptions for all the methods are violated ... Therefore, my concern is that the design of the simulation study implicitly favours some methods over others.*"

This is not the case. Suppose we have two identical distributions $Y_1 = Y_2$. This implies that their population parameters---means, medians, and so on---are also identical. Because monotonic transformations preserve equality, it follows that $g(Y_1) = g(Y_2)$, and the corresponding means, medians, etc. of $g(Y_1)$ and $g(Y_2)$ are again the same.

If a method rejects the null hypothesis under these conditions, it constitutes a Type I error, regardless of how the method defines the null hypothesis (e.g., in terms of means, medians, or another parameter).

As we noted earlier, our article explicitly clarifies when and why issues related to the definition of the null hypothesis become relevant.

**Comment 21:** *"What may be helpful would be to write down the different possible null hypotheses for the main effects (e.g. No difference in means, No difference in mean log, No difference of location shift …) and for the interaction (e.g. no multiplicative interaction), and compare the statistical performances of each method."*

See our response above.

Note that in the case of parallel main effects, where the definition of the null hypothesis for interaction becomes ambiguous, we explicitly state that the reported Type I error rates are defined with respect to the null hypothesis $a_{12}=0$. As we further discuss at the end of this section: *"It is unclear how to reliably control the null hypothesis at the level of the response scale when both main effects are varied."*

If our reviewer is aware of any method for nonlinear models that allows defining null hypotheses for interactions (for example, based on means) directly at the level of the response scale and comparing methods under this definition, we would be very interested to learn more.

**Comment 22:** *"How come the type I error rate for ART is inflated as compared to the other methods, but the power is similar?"*

We explain in our article: *"Because there is a tradeoff between Type I and Type II errors, high power can simply be the result of a high Type I error rate. Since parallel effects can inflate errors (see our previous results), we focus here on single effects, both main or interaction effects."*

To clarify further: for continuous heavy-tailed distributions (e.g., log-normal and exponential), ART's Type I error inflation for one factor (e.g., $X_1$) is only noticeable when there is also an effect on the other factor (e.g., $X_2$). In the power results we present, the effect on the second factor is zero. This choice avoids confounding Type I and Type II errors. If the effect on $X_2$ were increased, ART would appear more powerful, but such a comparison would be unfair.

For discrete distributions, however, ART inflates Type I error rates even when all effects are null. This explains why ART's power may appear higher for small effects (due to increased Type I errors) but deteriorates relative to other methods as effects grow larger---because the influence of Type I errors diminishes in these scenarios.

**Comment 23:** *"DGM: the authors should explain why the variables x1 and x2 were not generated using dummy indicators."*

See responses to Comments 7-8.

 **Comment 24:** *"It would be interesting to see, for each mixed model, whether the residuals and the random effects follow a normal distribution."*

We are not entirely sure we understand the purpose of this suggestion. Could the reviewer clarify what insight they expect from examining the residuals and random effects distributions?

 **Comment 25:** *"Were there any convergence issues with the mixed models?"*

Our models are relatively simple, and we did not encounter any convergence issues throughout the simulations. Convergence problems can arise in discrete data when all values for a particular factor level coincide, especially with a small number of levels; however, we have excluded such rare cases from our results.

 **Comment 26:** *"Did the empirical standard deviation of the regression coefficients across simulations correspond to the model-based standard errors?*"

We did not perform checks on the regression coefficients and are unsure what specific comparison the reviewer intends with this suggestion.

 **Comment 27:** *"In the presence of heteroscedasticity, sandwich standard errors may be used. Did the authors consider it?"*

Our experiments focus on comparing the robustness of different methods without any additional corrections. In reviewing previous studies using ART, we observed that several authors justify its use as a way to handle heteroscedasticity---even though, as we noted earlier, these authors ignore the method's underlying assumptions.

While we agree that appropriate methods exist to address heteroscedasticity, evaluating them was beyond the scope of our study. However, we have conducted additional tests that we do not report in the article---for example, for single factor designs ([see Appendix](https://www.journalovi.org/2024-tsandilas-ranktransforms/appendix.html#nonparametric-tests)), showing that combining the inverse normal transformation (INT) with sphericity corrections (for within-subjects designs) leads to better results than the Friedman test, which otherwise was shown to be more robust under heteroscedasticity (suffering from lower power though).

 **Comment 28:** *"Can the eta squared be directly compared across methods? Given that the total variance changes based on the transformation, it may impact the percentage explained by each of the variables."*

This is an excellent question. We fully agree that (partial) eta squared is not preserved under transformations, whether monotonic or not. We will clarify this point in a revised version.

It is also important to note that when researchers use ART, they typically report standardized effect sizes, following [Matthew Kay's vignette on effect sizes with ART](https://cran.r-project.org/web/packages/ARTool/vignettes/art-effect-size.html#partial-eta-squared). This raises a broader question: how should standardized effect sizes obtained from rank transformations be interpreted? Are they meaningful?

Our approach, using a latent variable, addresses this question by providing a baseline (ground truth) for comparison (see Figures 35–40). In many scenarios, we observe that the estimates from rank-based methods are close to this baseline, indicating that they can provide useful information in those cases.

However, significant discrepancies occur in certain situations, mainly due to two types of information loss (as discussed in the article):

 1. Rank transformations fail to capture large effects. As we explain: *"Consider, for example, two sets of numbers $A= \{1, 2, 3\}$ and $B= \{4, 5, 6\}$. The mean difference between them is $\overline{B} - \overline{A} = 3$, whether calculated from the row values or their ranks. However, if the set $B$ becomes $\{9, 10, 11\}$, the mean difference of raw values increases to $8$, while the mean difference in ranks remains $3$. As effects become larger, this problem becomes more apparent."*

2. Discretization in ordinal scales leads to information loss. All methods tend to underestimate effect sizes, though some methods still perform better than others.

A natural question arises: why is our ground truth appropriate? The answer is that it represents the effect size estimates of the **ideal** generalized linear regression model, which fully normalizes the data and supports inference at the level of the latent linear model parameters. If the data are log-normal, our baseline represents the effect sizes estimates of the linear model after applying a log transformation. If the data are binomial, it represents the estimates of the ideal logistic regression model. If the data are ordinal, it represents the estimates of the ideal ordinal regression model, and so on.

### 2025-12-01 appendix.qmd
minor correction
revised code

@@
 ART's alignment method is based on the assumption that the shape of the distribution (log-normal here) is identical across all factor levels.
-This is not the case in the above scenarios where distribution shapes change across difficulty levels, causing the method to break down.
+This is not the case in the above scenarios where distribution shapes change across the levels of $X_1$, causing the method to break down.
 
@@
 We first examine how the interaction effect alone influences the Type I error rate for $X_2$.
-The results are shown in @fig-interactions-main-2.
+The results are shown in @fig-interactions-main-1.
 We emphasize that interpretation issues arise only in the case of the $4 \times 3$ design.

### 2025-12-01 conversions_plot.R

@@ -8,6 +8,11 @@ norm2lnorm <- function(x, meanlog = 0, sdlog = 1, mu = mean(x), sd = stats::sd(x
    stats::qlnorm(p, meanlog, sdlog, ...)
 }
 
+norm2cauchy <- function(x, location = 0, scale = 1, mu = mean(x), sd = stats::sd(x), ...) {
+  p <- stats::pnorm(x, mu, sd)
+  stats::qcauchy(p, location, scale, ...)
+}
+
 toOrdinal = function(values, thresholds) {
   discretizeValue = function(x){
     for(i in 1:length(thresholds)) {
@@ -18,6 +23,14 @@ toOrdinal = function(values, thresholds) {
 
   sapply(values, discretizeValue)
 }
+
+
+norm2binom = function(x, size = 1, prob = 0.5, mu = mean(x), sd = stats::sd(x)) 
+{
+    p <- stats::pnorm(x, mu, sd)
+    stats::qbinom(p, size, prob)
+}
+
 
 set.seed(100)
 
@@ -50,6 +63,24 @@ plotConversion <- function(xs, ys, xlab = "Normal scale", ylab, ymin = 0, ymax =
   plot
 }
 
+n <- 5000
 
+xs1 <- c(rnorm(n, mean = -1, sd = 1))
+xs2 <- c(rnorm(n, mean = 1, sd = 1)) 
+xs <- c(xs1, xs2)
 
+#ys1 <- norm2lnorm(xs1, meanlog = 0, sdlog = 1, mu = -4, sd = 1)
+#ys2 <- norm2lnorm(xs2, meanlog = 0, sdlog = 1, mu = -4, sd = 1)
 
+ys1 <- c(rlnorm(n, mean = -1, sd = 1), rlnorm(n, mean = 1, sd = 1))
+plot1 <- plotConversion(xs, ys1, ylab = "Log-normal scale", ymax = 10)
+
+ys2 <- norm2binom(xs, size = 10, prob=.1)
+plot2 <- plotConversion(xs, ys2, ymin = - 0.3, ylab = "Binomial scale", as.density = F)
+
+ys3 <-  norm2gamma(xs, shape = 1, rate = 2, mu = mean(xs), sd = stats::sd(xs))
+plot3 <- plotConversion(xs, ys3, ymin = - 0.3, ylab = "Expnential scale")
+
+fig <- cowplot::plot_grid(plot1, plot2, plot3, ncol= 3) 
+
+fig

### 2025-12-01 supplementary/experimental-code-revised/analytical/test-ART-compare-rnk.R

@@ -1,0 +1,110 @@
+library(dplyr)
+
+#############################################
+### PARAMETERS
+#############################################
+mu     <- 0
+sigma  <- 1            # lognormal sd
+alphas <- seq(0, 4, by=0.5)
+#ns     <- c(10, 30, 50, 200, 1000)
+ns <- c(10, 20, 50, 100)
+R      <- 100         # Monte Carlo replicates
+set.seed(123)
+
+
+#############################################
+### FUNCTION: compute ART B-error (raw + logged)
+#############################################
+compute_art_error <- function(alpha, n){
+
+  # Levels
+  a1 <- 0
+  a2 <- alpha
+
+  # Create 2x2 structure
+  dat <- expand.grid(
+    i = c(1,2),
+    j = c(1,2),
+    k = 1:n
+  )
+  dat$alpha <- ifelse(dat$i==1, a1, a2)
+
+  # ----- Generate data -----
+  dat$Y_raw <- exp(mu + dat$alpha + rnorm(nrow(dat), sd=sigma))
+  dat$Y_log <- log(dat$Y_raw)
+
+
+  # ----- ART ALIGNMENT FOR RAW -----
+  grand.raw  <- mean(dat$Y_raw)
+  alpha.raw  <- tapply(dat$Y_raw, dat$i, mean) - grand.raw
+  dat$YB_raw <- dat$Y_raw - alpha.raw[dat$i] + grand.raw
+
+  e_raw <- abs(mean(dat$YB_raw[dat$j==2]) - mean(dat$YB_raw[dat$j==1]))
+
+  # ----- ART ALIGNMENT FOR LOGGED -----
+  grand.log  <- mean(dat$Y_log)
+  alpha.log  <- tapply(dat$Y_log, dat$i, mean) - grand.log
+  dat$YB_log <- dat$Y_log - alpha.log[dat$i] + grand.log
+
+  e_log <- abs(mean(dat$YB_log[dat$j==2]) -
+               mean(dat$YB_log[dat$j==1]))
+
+  c(error_raw = e_raw, error_log = e_log)
+}
+
+# analytic approximation
+analytic_B <- function(alpha, n, sigma = 1, mu = 0) {
+  S <- exp(mu)
+  m <- exp(sigma^2/2)
+  SD2 <- exp(sigma^2) - 1
+  
+  # approx: E|ΔB| ≈ S m * sqrt( SD2*(1 + e^(2α)) / (π n) )
[…]

### 2025-12-01 supplementary/experimental-code-revised/analytical/test-ART-compare.R

@@ -1,0 +1,196 @@
+library(dplyr)
+
+#############################################
+### PARAMETERS
+#############################################
+mu     <- 0
+sigma  <- 1            # lognormal sd
+alphas <- seq(0, 4, by=0.5)
+#ns     <- c(10, 30, 50, 200, 1000)
+ns <- c(10, 50, 100)
+R      <- 80         # Monte Carlo replicates
+set.seed(123)
+
+
+#############################################
+### FUNCTION: compute ART B-error (raw + logged)
+#############################################
+compute_art_error <- function(alpha, n){
+
+  # Levels
+  a1 <- 0
+  a2 <- alpha
+
+  # Create 2x2 structure
+  dat <- expand.grid(
+    i = c(1,2),
+    j = c(1,2),
+    k = 1:n
+  )
+  dat$alpha <- ifelse(dat$i==1, a1, a2)
+
+  # ----- Generate data -----
+  dat$Y_raw <- exp(mu + dat$alpha + rnorm(nrow(dat), sd=sigma))
+  dat$Y_log <- log(dat$Y_raw)
+
+
+  # ----- ART ALIGNMENT FOR RAW -----
+#  grand.raw  <- mean(dat$Y_raw)
+#  alpha.raw  <- tapply(dat$Y_raw, dat$i, mean) - grand.raw
+#  dat$YB_raw <- dat$Y_raw - alpha.raw[dat$i] + grand.raw
+
+#  e_raw <- abs(mean(dat$YB_raw[dat$j==2]) - mean(dat$YB_raw[dat$j==1]))
+
+  ## ------------- Alternative alignment
+  dat$S <- factor(dat$k)
+  dat$A <- factor(dat$i)
+  dat$B <- factor(dat$j)
+  dat <- dat %>% 
+    group_by(A) %>% mutate(Ai = mean(Y_raw)) %>% # mean for A = Ai
+    group_by(B) %>% mutate(Bj = mean(Y_raw)) %>% # mean for B = Bj
+    # mean for A=Ai, B=Bj and cell mean
+    group_by(A,B) %>% mutate(AiBj = mean(Y_raw)) %>% mutate(Cell = mean(Y_raw)) 
+
+  m <- mean(dat$Y_raw) # the grand mean
+
+  # Aligned observations
+#  dat$alignedA <- rank(dat$Y_raw - dat$Cell + dat$Ai - m)
+  dat$alignedB <- dat$Y_raw - dat$Cell + dat$Bj - m
[…]

### 2025-12-01 supplementary/experimental-code-revised/analytical/test-ART.R

@@ -1,0 +1,76 @@
+library(tidyverse)
+
+## ----------------------------------------------------------
+## Functions
+## ----------------------------------------------------------
+
+# simulate one 2×2 dataset and estimate spurious B effect
+simulate_once <- function(alpha, n, sigma = 1, mu = 0) {
+  # A = 1,2 ; B = 1,2
+  alpha_vals <- c(0, alpha)                # effect only on A2
+  mu_mat <- outer(alpha_vals, c(0,0), "+") # no B-effect
+  
+  # sample log-normal data
+  Y <- matrix(rlnorm(4*n, meanlog = mu_mat, sdlog = sigma),
+              nrow = 4)
+  
+  # cell means
+  mA1B1 <- mean(Y[1,])
+  mA1B2 <- mean(Y[2,])
+  mA2B1 <- mean(Y[3,])
+  mA2B2 <- mean(Y[4,])
+  
+  # estimated marginal B effect:
+  #   B2 - B1
+  est_B <- ((mA1B2 + mA2B2)/2) - ((mA1B1 + mA2B1)/2)
+  return(est_B)
+}
+
+# analytic approximation
+analytic_B <- function(alpha, n, sigma = 1, mu = 0) {
+  S <- exp(mu)
+  m <- exp(sigma^2/2)
+  SD2 <- exp(sigma^2) - 1
+  
+  # approx: E|ΔB| ≈ S m * sqrt( SD2*(1 + e^(2α)) / (π n) )
+  S*m * sqrt( SD2*(1 + exp(2*alpha)) / (pi*n) )
+}
+
+## ----------------------------------------------------------
+## Run simulation across alpha and n
+## ----------------------------------------------------------
+
+alphas <- seq(0, 2, by = 0.25)
+ns <- c(10, 20, 100)
+R <- 2000
+
+results <- expand.grid(alpha = alphas, n = ns, rep = 1:R) %>%
+  group_by(alpha, n) %>%
+  do({
+    ests <- replicate(R, simulate_once(.$alpha[1], .$n[1]))
+    tibble(
+      est_B_abs = abs(ests),
+      analytic  = analytic_B(.$alpha[1], .$n[1])
+    )
+  }) %>%
+  ungroup() %>%
+  group_by(alpha, n) %>%
+  summarise(
+    mean_abs_B = mean(est_B_abs),
+    analytic   = first(analytic)
+  )
+
+## ----------------------------------------------------------
+## Plot
[…]

### 2025-12-01 supplementary/experimental-code-revised/effects_plot.R

@@ -1,0 +1,204 @@
+# Author: Theophanis Tsandilas, Inria & Université Paris-Saclay
+
+# A great library for simulating data for a large range of desings: https://debruine.github.io/faux
+library(faux)
+
+library(dplyr)
+library(tidyr)
+library(egg)
+library(latex2exp)
+
+source("data-generators.R")
+source("simulation-utils.R")
+
+############################################################################################
+# Data generation for various experimental designs
+
+# Encodes the 1:nlevels to numerical values
+# It ensures a fixed distance between the two extreme levels (equal to 1) and random distances in between
+# It also ensures that the sum is zero
+categ2ratio <- function(nlevels) { 
+	levels <- sort(c(0,1, runif(nlevels-2)))
+
+	return(levels - mean(levels))
+}
+
+# Transforms categorical to numverical
+toNumber <- function(factorvalues) {
+	as.numeric(as.character(factorvalues))
+}
+
+# This function produces difference variance ratios for each variable level
+# where the most extreme will be 1 to maxratio
+sd.ratios <- function(nlevels = 4, maxratio = 1.5) {
+	ratios <- c(1, maxratio, 1 + runif(nlevels - 2)*(maxratio - 1))
+	sample(ratios, replace=FALSE) - mean(ratios) + 1
+}
+
+
+getThresholds = function(values, nlevels = 5, equidistant = TRUE) {
+	lim <- 2*sd(values) # I take 2 standard deviations that should contain the largest portion of the data
+
+	thresholds <- NA
+	if(equidistant) {
+		thresholds <- seq(from=-lim, to=lim, by=2*lim/nlevels)[c(-1, -(nlevels + 1))]
+	} else {
+		thresholds <- sort(runif(nlevels - 1))*2*lim-lim
+	}
+
+	thresholds
+}
+
+
+# Distribution parameters used in simulation
+params_normal <- list(sigma_s = 0, sigma_e = 1)
+params_lognormal <- list(sigma_s = 0, sigma_e = 1, mean_target = 1)
+params_cauchy <- list(sigma_s = 0, gamma = 1)
[…]

### 2025-12-01 supplementary/experimental-code-revised/simulation-art-fast.R

@@ -1,0 +1,35 @@
+# Fast implementation of ART - LMER using lmertest and Satterthwaite approximation
+# This is of interest for simulation with large number of subjects, where AOV becomes expensive
+fast_art_p <- function(m) {
+  # Recover ORIGINAL formula (with random-effects preserved)
+  original_formula <- eval(m$formula)
+
+  # aligned columns
+  aligned <- m$aligned.ranks
+  term_names <- colnames(aligned)
+
+  # Build full data frame with predictors INCLUDING 'subject'
+  predictors <- get_all_vars(original_formula, data = m$data)
+
+  df_base <- cbind(aligned, predictors)
+
+  # Extract RHS from original formula
+  rhs <- original_formula[[3]]
+
+  results <- numeric(length(term_names))
+  names(results) <- term_names
+
+  for (term in term_names) {
+    df <- df_base
+    df$.y <- df[[term]]
+    df <- df[,-c(1:length(term_names))] # And remove the columns that could cause confusion 
+
+    f <- reformulate(deparse(rhs), response = ".y")
+
+    model <- lmerTest::lmer(f, data = df)
+    tab <- anova(model, type = 3)
+    results[term] <- tab[term, "Pr(>F)"]
+  }
+
+  as.data.frame(as.list(results))
+}

### 2025-12-01 supplementary/experimental-code-revised/simulation-benchmarks-binom.R

@@ -1,0 +1,38 @@
+rm(list=ls())
+#library(lmerTest)
+#library(ARTool)
+
+library(microbenchmark)
+
+source("simulation-utils.R")
+
+nlevels = c(4, 3)
+nsubjects = 10000
+reps = 1
+a1 = 2
+a2 = 0
+a12 = 0
+
+
+bench <- microbenchmark(
+  poisson = simulate_response(
+    nlevels, nsubjects, reps,
+    a1, a2, a12,
+    sigma_s = 0.3,
+    family = "poisson",
+    mean_target = 3
+  ),
+
+  binom = simulate_response(
+    nlevels, nsubjects, reps,
+    a1, a2, a12,
+    sigma_s = 0.3,
+    family = "binomial",
+    p_target = 0.1,
+    size = 10
+  ),
+#  accurate = mu_binomial_accurate(compute_effects(make_design(m1=2,m2=2),a1=4,a2=0,a12=0), sigma_s = 0.3, p_target = 0.1),
+#  gh = mu_binomial(compute_effects(make_design(m1=2,m2=2),a1=4,a2=0,a12=0), sigma_s = 0.3, p_target = 0.1, size=10, n_nodes = 32),
+  times = 100
+)
+

### 2025-12-01 supplementary/experimental-code-revised/simulation-dispersion.R

@@ -1,0 +1,44 @@
+set.seed(123)
+
+# OD values
+OD_values <- c(1.1, 1.2, 1.5, 2, 5)
+
+# Preallocate
+res <- data.frame(OD = OD_values,
+                  sigma_normal = NA,
+                  sigma_lognormal = NA,
+                  sigma_poisson = NA,
+                  sigma_exponential = NA,
+                  sigma_binomial = NA,
+                  sigma_cauchy = NA)
+
+lambda <- 3
+sigma_e <- 1
+p_target <- 0.1
+n_bin <- 20
+gamma <- 1
+
+for(i in seq_along(OD_values)) {
+  OD <- OD_values[i]
+  # Normal
+  res$sigma_normal[i] <- sqrt((OD-1)*sigma_e^2)
+  # Log-normal
+  res$sigma_lognormal[i] <- sqrt(log(OD + 1))
+  # Poisson
+  res$sigma_poisson[i] <- if(OD <=1) 0 else sqrt(log(1 + (OD-1)/lambda))
+  # Exponential
+  res$sigma_exponential[i] <- if(OD <=1) 0 else sqrt(log(OD))
+  # Binomial (numerical)
+  f_bin <- function(s) {
+    p <- plogis(qlogis(p_target) + rnorm(1e5,0,s))
+    var(p)/(p_target*(1-p_target)) - OD
+  }
+  # check if achievable
+  tryCatch({
+    res$sigma_binomial[i] <- uniroot(f_bin, c(0,5))$root
+  }, error=function(e) { res$sigma_binomial[i] <- NA })
+  # Cauchy (rough)
+  res$sigma_cauchy[i] <- sqrt(OD - 1)
+}
+
+res

### 2025-12-01 supplementary/experimental-code-revised/simulation-test-repeated-4x3.R

@@ -1,0 +1,175 @@
+rm(list=ls())
+library(lmerTest)
+library(ARTool)
+
+library(tidyr)
+library(dplyr)
+library(broom)
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+
+source("simulation-utils.R")
+
+# https://yingji15.github.io/2019-08-17-inverse-normal-transformation/
+# https://www.biostars.org/p/80597/
+# use rankit version where c = 0.5, based on https://www.researchgate.net/publication/286889581_Impact_of_Rank-Based_Normalizing_Transformations_on_the_Accuracy_of_Test_Scores
+INT = function(x){
+  qnorm((rank(x) - 0.5)/length(x))
+}
+
+# Extract the p-values from the result objects returned by summary() or anova()
+get_p_values <- function(model, vars, is_lmer = FALSE) {
+  if(is_lmer) return(anova(model)[vars, "Pr(>F)"])
+  else if(is.null(model$formula)){
+    return((tidy(summary(model)[[2]][[1]]) %>% filter(term %in% vars) %>% select(p.value))[[1]])
+  } else {
+    anov <- anova(model)
+    return(anov[anov$Term %in% vars, "Pr(>F)"])
+  }
+}
+
+compare_p_values_aov <- function(df){ 
+  model_par <- suppressMessages(aov(Y ~ X1*X2 + Error(subject), data=df)) # Parametric
+  model_art <- suppressMessages(art(Y ~ X1*X2 + Error(factor(subject)), data=df)) # ARTool
+  model_rnk <- suppressMessages(aov(rank(Y) ~ X1*X2 + Error(subject), data=df)) # RNK
+  model_int <- suppressMessages(aov(INT(Y) ~ X1*X2 + Error(subject), data=df)) # INT
+
+  vars <- c("X1", "X2", "X1:X2")
+  c(get_p_values(model_par, vars), 
+    get_p_values(model_rnk, vars), 
+    get_p_values(model_art, vars), 
+    get_p_values(model_int, vars)
+  )
+}
+
+
+################################
+# Distribution parameters used in simulation
+params_normal <- list(sigma_s = 0.4, sigma_e = 1)
+params_lognormal <- list(sigma_s = 0.4, sigma_e = 1, mean_target = 1)
+params_cauchy <- list(sigma_s = 0.4, gamma = 1)
+params_exponential <- list(sigma_s = 0.3, mean_target = 0.5)
[…]

### 2025-12-01 supplementary/experimental-code-revised/simulation-test.R

@@ -1,0 +1,99 @@
+rm(list=ls())
+
+source("utils-data-generation.R")
+source("utils-analysis.R")
+
+
+# Distribution parameters used in simulation
+params_normal <- list(sigma_s = 0.4, sigma_e = 1)
+params_ordinal <- list(sigma_s = 0.4, sigma_e = 1, levels = 5, equidistant = FALSE)
+params_lognormal <- list(sigma_s = 0.4, sigma_e = 1, mean_target = 1)
+params_cauchy <- list(sigma_s = 0.4, gamma = 1)
+params_exponential <- list(sigma_s = 0.3, mean_target = 0.5)
+params_poisson <- list(sigma_s = 0.3, mean_target = 3)
+params_binomial <- list(sigma_s = 0.3, size = 10, p_target = 0.1)
+
+nlevels = c(4, 3)
+within = c(1, 1)
+n = 20
+coeffs <- c(X1=2, X2=0, X1X2=0) # a1, a2, a12
+
+df_normal <- simulate_response(
+  nlevels, within, n, coeffs,
+  family = "normal",
+  params = params_normal
+)
+
+df_ordinal <- simulate_response(
+  nlevels, within, n, coeffs,
+  family = "ordinal",
+  params = params_ordinal
+)
+
+df_lognormal <- simulate_response(
+  nlevels, within, n, coeffs,
+  family = "lognormal",
+  params = params_lognormal
+)
+
+df_cauchy <- simulate_response(
+  nlevels, within, n, coeffs,
+  family = "cauchy",
+  params = params_cauchy
+)
+
+df_exp <- simulate_response(
+  nlevels, within, n, coeffs,
+  family = "exponential",
+  params = params_exponential
+)
+
+df_poisson <- simulate_response(
+  nlevels, within, n, coeffs,
+  family = "poisson",
+  params = params_poisson
+)
+
+df_binom <- simulate_response(
+  nlevels, within, n, coeffs,
+  family = "binomial",
+  params = params_binomial
+)
+
+printResults <- function(family, res){
+  cat("Results for ", family, " #################\n")
+  cat("PAR: ", res[grep("PAR", names(res))], "\n")
+  cat("ART: ", res[grep("ART", names(res))], "\n")
+  cat("RNK: ", res[grep("RNK", names(res))], "\n")
+  cat("INT: ", res[grep("INT", names(res))], "\n\n")
+}
+
+vars = c("X1", "X2", "X1:X2")
+formula = Y ~ X1*X2 + (1|subject)
[…]

### 2025-12-01 supplementary/experimental-code-revised/utils-analysis.R

@@ -1,0 +1,62 @@
+library(lmerTest)
+library(ARTool)
+library(broom)
+library(dplyr)
+
+# Implementation of the Inverse Normal Transformation
+# https://yingji15.github.io/2019-08-17-inverse-normal-transformation/
+# https://www.biostars.org/p/80597/
+# use rankit version where c = 0.5, based on https://www.researchgate.net/publication/286889581_Impact_of_Rank-Based_Normalizing_Transformations_on_the_Accuracy_of_Test_Scores
+INT = function(x){
+  qnorm((rank(x) - 0.5)/length(x))
+}
+
+# Extract the p-values from the models (either lmer or aov)
+get_p_values <- function(model, vars, is_aov = TRUE) {
+  if(!is_aov) return(anova(model)[vars, "Pr(>F)"])
+  else if(is.null(model$formula)){ # Handling AOV objects is more complicated
+    summary_object <- summary(model)
+    if(length(summary_object) == 1){ # This occurs when there are only between-subjects factors
+      res <- tidy(summary_object[[1]][[1]])
+      } 
+    else { # When there is at list a within-subjects factor
+      res <- bind_rows(tidy(summary_object[[1]][[1]]), tidy(summary_object[[2]][[1]])) 
+    }
+
+    res <- res  %>% # Get both the between- and within-subjects terms
+        filter(term %in% vars) %>% # Filter the relevant rows
+        arrange(match(term, vars)) %>% # Order them
+        select(p.value) # And select the p-value
+    return(res[[1]])
+  } else {
+    anov_object <- anova(model)
+    return(anov_object[anov_object$Term %in% vars, "Pr(>F)"])
+  }
+}
+
+# Returns a vector with all p-values for all methods
+compare_p_values <- function(df, 
+  formula = Y ~ X1*X2 + (1|subject), 
+  vars = c("X1", "X2", "X1:X2") # Specify the terms of interest. P-values will be returned with the specified order
+){  
+  is_aov <- "Error" %in% all.names(formula) # Identifies if it should use lmer or aov
+
+  if(is_aov) { # Use aov
+    model_par <- suppressMessages(aov(formula, data=df)) # Parametric
+    model_art <- suppressMessages(art(formula, data=df)) # ARTool
+    model_rnk <- suppressMessages(aov(update(formula, rank(.) ~ .), data=df)) # RNK
[…]

### 2025-12-01 supplementary/experimental-code-revised/utils-data-generation.R

@@ -1,0 +1,240 @@
+# Code for generating experimental designs (within-, between-subjects, and mixed) and respones for a range of distributions
+# Author: Theophanis Tsandilas, Dec 2025
+
+require("statmod")
+
+##################################################################################
+# Encodes the 1:nlevels to numerical values
+# It ensures a fixed distance between the two extreme levels (equal to 1) and random distances in between
+# It also ensures that the sum is zero
+coding <- function(nlevels) { 
+  levels <- sort(c(0, 1, runif(nlevels - 2)))
+  return(levels - mean(levels))
+}
+
+# Generate the basic design matrix (balanced)
+build_structure <- function(nlevels = c(2,2)) {
+  k <- length(nlevels)
+
+  # 1. Create factor labels and numeric codings, e.g., for X1: A1, A2, ... and for X2: B1, B2, ...  
+  Xlabs <- lapply(seq_len(k), function(j) paste0(LETTERS[j], seq_len(nlevels[j])))
+  xcoding <- lapply(nlevels, coding)
+
+  # 2: Create a frame with the data variables
+  df <- expand.grid(
+    lapply(1:k, function(j) Xlabs[[j]])
+  )
+  names(df) <- paste0("X", 1:k)
+
+  # 3: Match numeric coding for each level
+  for (j in seq_len(k)) {
+    df[[paste0("x", j)]] <- xcoding[[j]][ match(df[[paste0("X", j)]], Xlabs[[j]]) ]
+  }
+  
+  df
+}
+
+# Create the full design, adding the participants
+# n is the cell sample - the number of times a specific combination of factor levels will appear 
+# For within-subjects designs, it is also the number of subjects
+make_design <- function(nlevels = c(4,3), within = c(1,1), n = 20) {
+  # 1. Create the basic full factorial structure
+  df <- build_structure(nlevels)
+
+  k <- length(nlevels)
+  if (length(within) != k)
+    stop("within must have same length as nlevels")
+
+  # 2. Identify between-subject factors
+  between_idx <- which(within == 0)
+
+  if (length(between_idx) == 0) { # Fully within-subjects
+    df$subject <- 1
+  }
+  else {
+    # 3. Define unique between-subject combinations
+    between_vars <- paste0("X", between_idx)
+    df_between <- unique(df[between_vars])
+
+    # 4. Assign subject IDs per between-subject cell
+    df_between$subject <- seq_len(nrow(df_between))
+
[…]

### 2025-12-10 sandbox.qmd
starting new revisions

--- 2025-07-07 sandbox.qmd
@@
 
-::: {.callout-important appearance="simple"}
-## Under Review {.unnumbered}
-This paper is [under review](https://www.journalovi.org/under-review.html) on the experimental track of
-the [Journal of Visualization and Interaction](https://www.journalovi.org/). See the 
-[reviewing process](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues).
-:::
-
 ```{r, echo=FALSE, warning=FALSE, message=FALSE}
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
-# Source code for reading data from experimental results and for plotting
-source("dataReaders.R")
-source("plotting.R")
-source("plotlying.R")
-source("effects_plot.R")
-source("interactions_plot.R")
-```
-
 ```{=html}
@@
 ```
+
+::: {#fig-ratio-main}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+library(dplyr)
+library(tidyverse)
+source("utils-data-reading.R")
+source("utils-plotlying.R")
+```
+:::
+
+::: {#fig-power-main-1}
+<div id="plot-power-1.1">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Power"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+
+data <- read_data(prefix, alpha = .05, effectType = 3, distributions)
+df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX1", max = 104,  ytitle = 'Power (%)')
+```
+</div>
+
+<div id="plot-power-1.2">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- data %>% arrange(design,family,effectX1,rateX1)  %>% group_by(design,family,effectX1) %>% mutate(rank = rank(rateX1))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power - ranking')
+```
+</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-1.2").classList.add("hidden"); }); </script>
+
+<button id="btn1" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
+    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
+  onclick="const plot = document.getElementById('plot-power-1.1');
+  plot.classList.toggle('hidden');
+  document.getElementById('plot-power-1.2').classList.toggle('hidden');
+  btn1.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
+">
+  show ranking
+</button>
+
+Power ($\alpha = .05$) for detecting the **main effect of $X_1$** as a function of its magnitude $a_1$, where $n=20$.
+:::
+
+@fig-power-main-2 presents our results for $X_2$.
+Given the different number of levels for this factor, we now assess a different range of power for each design.
+Nevertheless, we observe similar patterns.
+INT stands out as the most powerful method.
+Although ART surpasses PAR under the log-normal and exponential distributions, it does not demonstrate any other notable advantages compared to the other methods.
+
+::: {#fig-power-main-2}
+<div id="plot-power-2.1">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+data <- read_data(prefix, alpha=.05, effectType = 4, distributions) # CHANGE n TO 20
+
+df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX2", max = 104,  ytitle = 'Power (%)')
+```
+</div>
+
+<div id="plot-power-2.2">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- data %>% arrange(design,family,effectX2,rateX2)  %>% group_by(design,family,effectX2) %>% mutate(rank = rank(rateX2))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+#df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power - ranking')
+```
+</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-2.2").classList.add("hidden"); }); </script>
+
+<button id="btn2" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
+    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
+  onclick="const plot = document.getElementById('plot-power-2.1');
+  plot.classList.toggle('hidden');
+  document.getElementById('plot-power-2.2').classList.toggle('hidden');
+  btn2.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
+">
+  show ranking
+</button>
+
+Power ($\alpha = .05$) for detecting the **main effect of $X_2$** as a function of its magnitude $a_2$, where $n=20$. 
+:::
+
+**Interaction effects**.
+We also vary the magnitude of the interaction effect $X_1 \times X_2$ by manipulating the parameter $a_{12}$ across the range of $0.5$ to $2.0$.
+We then measure the power to detect this interaction effect.
+Our results are summarized in @fig-power-interaction, aligning with findings for main effects.
+ART's relative advantage diminishes later under the binomial and ordinal distributions.
+Once again, the method's inflated Type I error rates could account for this observed trend.
+
+::: {#fig-power-interaction}
+<div id="plot-power-3.1">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+data <- read_data(prefix, alpha=.05, effectType = 5, distributions)
+
+df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of interaction effect", var = "rateX1X2", xvar = "effectX1X2", max = 104,  ytitle = 'Power (%)')
+```
+</div>
+
+<div id="plot-power-3.2">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- data %>% arrange(design,family,effectX1X2,rateX1X2)  %>% group_by(design,family,effectX1X2) %>% mutate(rank = rank(rateX1X2))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+#df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power - ranking')
+```
+</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-3.2").classList.add("hidden"); }); </script>
+
+<button id="btn3" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
+    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
+  onclick="const plot = document.getElementById('plot-power-3.1');
+  plot.classList.toggle('hidden');
+  document.getElementById('plot-power-3.2').classList.toggle('hidden');
+  btn3.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
+">
+  show ranking
+</button>
+
+Power ($\alpha = .05$) for detecting the **interaction effect $X_1 \times X_2$** as a function of its magnitude $a_{12}$, where $n=20$.
+:::

### 2025-12-10 text/60-Results-RatioScales.qmd
starting new revisions

@@
 library(tidyverse)
+source("utils-data-reading.R")
+source("utils-plotlying.R")
+
+# To remote those: 
 source("dataReaders.R")
@@
 
-prefix <- "1_test_4x3_Ratio"
+prefix <- "Type_I_4x3_ratio"
 alpha <- .05
@@
 dnames <- c("Normal", "Log-normal", "Exponential", "Cauchy", "Poisson", "Binomial")
-df <- readlyData(prefix, alpha, 1, distributions, dnames)
-plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 64)
+df <- readly_data(prefix, alpha, 1, distributions, dnames)
+plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 102)
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "1_test_4x3_Contrasts"
+prefix <- "Type_I_4x3_ratio_contrasts"
 
-df <- readlyData(prefix, alpha, 1, distributions, dnames)
-plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 52)
+df <- readly_data(prefix, alpha, 1, distributions, dnames)
+plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 101)
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "1_test_4x3_Ratio"
-df <- readlyData(prefix, alpha, 1, distributions, dnames)
+prefix <- "Type_I_4x3_ratio"
+df <- readly_data(prefix, alpha, 1, distributions, dnames)
 
-plotlyError(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 52)
+plotly_error(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 100)
 ```
@@
 Since there can be different possible definitions of an interaction effect in these cases, we address them in a separate subsection.
+
+**Analysis for large samples.** Let us further explore ART's error rates when sample sizes become larger.
+
+::: {#fig-ratio-large-samples}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Type_I_4x3_ratio_large_samples"
+
+distributions <- c("norm", "lnorm", "exp", "cauchy", "poisson", "binom")
+dnames <- c("Normal", "Log-normal", "Exponential", "Cauchy", "Poisson", "Binomial")
+
+df <- readly_data_1(prefix, alpha, 1, distributions, dnames)
+plotly_error_by_effect_1(df, xlab = "n", max = 86)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** and the **interaction** $X_1 \times X_2$ as a function of the size $n$ of the sample.
+The effect on $X_1$ is non-null, where we set $a_1 = 1$.
+:::

### 2025-12-10 text/61-Results-OrdinalScales.qmd
starting new revisions

@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "2_test_4x3_Ordinal"
+prefix <- "Type_I_ordinal"
 
-distributions <- c("likert5", "likert5B", "likert7", "likert7B", "likert11", "likert11B")
+distributions <- c("likert_5", "likert_5_flex", "likert_7", "likert_7_flex", "likert_11", "likert_11_flex")
 dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
-df <- readlyData(prefix, alpha, 1, distributions, dnames)
+df <- readly_data(prefix, alpha, 1, distributions, dnames)
 
-plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 52)
+plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 60)
 ```
@@
 
-plotlyError(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 45)
+plotly_error(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 50)
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "2_test_4x3_Ordinal_sample_size"
+prefix <- "Type_I_4x3_ordinal_large_samples"
 
-distributions <- c("likert5", "likert5B", "likert7", "likert7B", "likert11", "likert11B")
-dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
-
-df <- readlyData1(prefix, alpha, 0, distributions, dnames)
-plotlyErrorByEffect(df, xlab = "n", max = 100)
+df <- readly_data_1(prefix, alpha, 0, distributions, dnames)
+plotly_error_by_effect_2(df, xlab = "n", max = 100)
 ```
@@
 :::
-
-We will observe similar trends if we test other discrete distributions, such as binomial distributions. These results confirm that ART is inappropriate for discrete data.

### 2025-12-10 text/62-Results-AcrossExpDesigns.qmd
starting new revisions

@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "3_test-Designs"
+prefix <- "Type_I_designs"
 alpha = 0.05
 
-distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
 dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
 
-df <- readData(prefix, n = 20, alpha, effectType = 1, distributions)
-df <- reshapeByDesign(df, dnames)
+df <- read_data(prefix, alpha, effectType = 1, distributions)
+df <- reshape_by_design(df, dnames)
 
-plotlyErrorByDesign(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 62)
+plotly_error_by_design(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 101)
 ```
 Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
-:::
-
-We provide additional results for the between-subjects design in @fig-2x3-main, where we vary the effect of $X_2$ and measure Type I error rates on $X_1$.
-ART's error rates for the three discrete distributions appear higher than those observed in @fig-designs-main, indicating a potential dependence on the number of levels of the factors.
-Again, we observe a consistent trend where error rates in all discrete distributions increase with the sample size.
-
-::: {#fig-2x3-main}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix2 <- "3_test-Designs_2x3"
-
-df2x3 <- readlyData2(prefix2, alpha, effectType = 2, distributions, dnames)
-
-plotlyError(df2x3, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX2", max = 62)
-```
-Additional results for the $2 \times 3$ between-subjects design, where we measure Type I error rates ($\alpha = .05$) for the **main effect of $X_1$.** We now vary the magnitude $a_2$ of the main effect of $X_2$.
 :::
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-plotlyErrorByDesign(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 62)
+plotly_error_by_design(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 101)
 ```

### 2025-12-10 text/63-Results-UnequalVariance.qmd
starting new revisions

@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "4_test_Unequal_Variances"
-distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
-dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+prefix <- "Type_I_heteroscedacity"
+distributions = c("norm", "likert_5", "likert_5_flex", "likert_7", "likert_7_flex")
+dnames = c("Normal", "Equidistant (5 levels)", "Flexible (5 levels)", "Equidistant (7 levels)", "Flexible (7 levels)")
 
-df <- readData(prefix, n = 20, alpha, effectType = 1, distributions)
-df <- reshapeByDesign(df, dnames, effectvars = c("sd_ratio", "effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1", xvar = "sd_ratio", max = 82, ytitle = 'Positives (%)')
+df <- read_data(prefix, alpha = 0.05, effectType = 1, distributions)
+df <- reshape_by_design(df, dnames, effectvars = c("sdratio", "effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1", xvar = "sdratio", max = 41)
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-plotlyErrorByDesign2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX2", xvar = "sd_ratio", max = 52)
+plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX2", xvar = "sdratio", max = 41)
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-plotlyErrorByDesign2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1X2", xvar = "sd_ratio", max = 52)
+plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1X2", xvar = "sdratio", max = 42)
 ```

### 2025-12-10 text/64-Results-StatPower.qmd
starting new revisions

@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "5_test-Power"
-distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+prefix <- "Power"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
 dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
 
-data <- readData(prefix, n = 20, alpha = .05, effectType = 3, distributions)
-df <- reshapeByDesign(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX1", max = 104,  ytitle = 'Power (%)')
+data <- read_data(prefix, alpha = .05, effectType = 3, distributions)
+df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX1", max = 104,  ytitle = 'Power (%)')
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- data %>% arrange(design,distr,effectX1,rateX1)  %>% group_by(design,distr,effectX1) %>% mutate(rank = rank(rateX1))
-df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyPowerByDesign(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power - ranking')
+df <- data %>% arrange(design,family,effectX1,rateX1)  %>% group_by(design,family,effectX1) %>% mutate(rank = rank(rateX1))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power - ranking')
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-data <- readData(prefix, n = 20, alpha, effectType = 4, distributions) # CHANGE n TO 20
+data <- read_data(prefix, alpha=.05, effectType = 4, distributions) # CHANGE n TO 20
 
-df <- reshapeByDesign(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX2", max = 104,  ytitle = 'Power (%)')
+df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX2", max = 104,  ytitle = 'Power (%)')
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- data %>% arrange(design,distr,effectX2,rateX2)  %>% group_by(design,distr,effectX2) %>% mutate(rank = rank(rateX2))
-df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+df <- data %>% arrange(design,family,effectX2,rateX2)  %>% group_by(design,family,effectX2) %>% mutate(rank = rank(rateX2))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
 
-#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyPowerByDesign(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power - ranking')
+#df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power - ranking')
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-data <- readData(prefix, n = 20, alpha, effectType = 5, distributions)
+data <- read_data(prefix, alpha=.05, effectType = 5, distributions)
 
-df <- reshapeByDesign(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX1X2", xvar = "effectX1X2", max = 104,  ytitle = 'Power (%)')
+df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of interaction effect", var = "rateX1X2", xvar = "effectX1X2", max = 104,  ytitle = 'Power (%)')
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- data %>% arrange(design,distr,effectX1X2,rateX1X2)  %>% group_by(design,distr,effectX1X2) %>% mutate(rank = rank(rateX1X2))
-df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+df <- data %>% arrange(design,family,effectX1X2,rateX1X2)  %>% group_by(design,family,effectX1X2) %>% mutate(rank = rank(rateX1X2))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
 
-#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyPowerByDesign(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power - ranking')
+#df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power - ranking')
 ```

### 2025-12-10 text/66-Results-Interactions.qmd
starting new revisions

@@
 alpha <- .05
-prefix <- "1_test_4x3_Ratio"
-df <- readlyData(prefix, alpha, 0, distributions, dnames)
+prefix <- "Type_I_4x3_ratio"
+df <- readly_data(prefix, alpha, 0, distributions, dnames)
 
 #xlab <- TeX("\\text{main effect of factors }X_1\\text{ and }X_2\\text{ }(a_1 = a_2)")
-plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
+plotly_error(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "2_test_4x3_Ordinal"
-distributions <- c("likert5", "likert5B", "likert7", "likert7B", "likert11", "likert11B")
+prefix <- "Type_I_ordinal"
+distributions <- c("likert_5", "likert_5_flex", "likert_7", "likert_7_flex", "likert_11", "likert_11_flex")
 dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
 
-df <- readlyData(prefix, alpha, 0, distributions, dnames)
-plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
-
+df <- readly_data(prefix, alpha, 0, distributions, dnames)
+plotly_error(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "3_test-Designs"
-distributions <- c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+prefix <- "Type_I_designs"
+distributions <- c("norm", "lnorm", "exp", "poisson", "binom", "likert")
 dnames <- c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
 
-df <- readData(prefix, n = 20, alpha, effectType = 0, distributions)
-df <- reshapeByDesign(df, dnames)
-plotlyErrorByDesign(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
+df <- read_data(prefix, alpha, effectType = 0, distributions)
+df <- reshape_by_design(df, dnames)
+plotly_error_by_design(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
 ```

### 2025-12-10 supplementary/experiments-code/experiment-effect-sizes.R

@@ -43,13 +43,15 @@ distributions <- c("norm", "lnorm", "likert")
 
 # Various combinations of effects
 effects <- matrix(c(0, 0, 0,
-            0.5, 0, 0,
+            0, 1, 0,
+            0, 2, 0,
+            0, 4, 0,
+            0, 8, 0,
             1, 0, 0,
-            2, 0, 0,
-            4, 0, 0,
-            8, 0, 0,
-            0.5, 0.5, 0,
             1, 1, 0,
+            1, 2, 0,
+            1, 4, 0,
+            1, 8, 0,
             2, 2, 0,
             4, 4, 0,
             8, 8, 0), 

### 2025-12-10 supplementary/experiments-code/utils-data-generation.R

@@ -230,7 +230,10 @@ simulate_response <- function(
   # -------------------------------------------------
   Y <- switch(family,
     norm     = eta + rnorm(length(eta), 0, params$sigma_e),
-    likert   = toOrdinal(eta + rnorm(length(eta), 0, params$sigma_e), levels = params$levels, flexible = params$flexible),
+    likert   = { # For ordinal, I need eta to store the latent values (that include the trial error)
+      eta = eta + rnorm(length(eta), 0, params$sigma_e)
+      toOrdinal(eta, levels = params$levels, flexible = params$flexible)
+    },
     poisson  = rpois(length(eta), exp(eta)),
     exp      = rexp(length(eta), rate = 1/exp(eta)),
     lnorm    = rlnorm(length(eta), eta, params$sigma_e),
@@ -306,9 +309,11 @@ simulate_heteroscedastic_response <- function(
   # -------------------------------------------------
   # Generate Y
   # -------------------------------------------------
-  Y <- eta + sapply(1:length(eta), function(i){ rnorm(1,0, sigmas_e[i]) })
+  eta <- eta + sapply(1:length(eta), function(i){ rnorm(1,0, sigmas_e[i]) })
   if(family == "likert") {
-    Y <- toOrdinal(Y, levels = params$levels, flexible = params$flexible)
+    Y <- toOrdinal(eta, levels = params$levels, flexible = params$flexible)
+  } else {
+    Y <- eta
   }
 
   # Ensure that categorical variables and the random effect are coded as factors

### 2025-12-10 supplementary/experiments-code/utils-effect-size.R

@@ -5,6 +5,8 @@ library(lmerTest)
 library(ARTool)
 library(effectsize) # Effect size functions
 
+library(broom)
+library(dplyr)
 
 # Implementation of the Inverse Normal Transformation
 INT = function(x){
@@ -57,6 +59,30 @@ compare_effect_sizes <- function(df,
   )
 }
 
+# This is to add the results of groundtruth method-simple anova over the correct data
+get_groundtruth_effect_sizes <- function(df,
+    formula = Y ~ X1*X2 + (1|subject), 
+    vars = c("X1", "X2", "X1:X2"), # Specify the terms of interest. P-values will be returned with the specified order
+    measure = c("eta", "cohens")
+  ){
+    
+  measure <- match.arg(measure)
+  effect_size_function <- {
+    if(measure == "eta") eta_values
+    else cohensf_values 
+  }
+
+  is_aov <- "Error" %in% all.names(formula) # Identifies if it should use lmer or aov
+  if(is_aov) { # Use aov
+    model <- suppressMessages(do.call(aov, list(formula, df))) 
+  } 
+  else { # Use lmer
+    model <- suppressMessages(do.call(lmer, list(formula, data=df))) 
+  }
+
+  c(GRT = suppressMessages(effect_size_function(model, vars)))
+}
+
 ################################################################
 # Tests over a large number of iterations 
 ################################################################
@@ -68,13 +94,16 @@ repeat_test_effect_size <- function(
   within = c(1,1),
   n=20, 
   coeffs=c("X1"=0, "X2"=0, "X1:X2"=0),
-  family="norm",
+  family=c("norm", "lnorm", "likert"),
   params,
   formula,
   vars,
   measure = "eta",
-  iterations = 1000
+  iterations = 100
 ) {
+  
+  family <- match.arg(family)
+
   results <- foreach(rid = 1:iterations, .combine=rbind) %dopar% {
     tryCatch(
       {
@@ -86,7 +115,21 @@ repeat_test_effect_size <- function(
           data <- removeCells(data, params$ratio_missing)
         }
         
-        compare_effect_sizes(data, formula, vars, measure)
+        eff_sizes <- compare_effect_sizes(data, formula, vars, measure)
+        if(family=="norm") { # The ground truth is the PAR
+          groundtruth <- eff_sizes[startsWith(names(eff_sizes), "PAR")]
+          names(groundtruth) <- paste0("GRT", seq_along(vars))  
[…]

### 2025-12-10 utils-data-reading.R

@@ -1,0 +1,116 @@
+# Author: Theophanis Tsandilas, Inria & Université Paris-Saclay
+
+# the effectType parameter determines if I'm interested in results where there are effects on both factors (effectType = 0) 
+# or only on the first factor (effectType = 1), second only factor (effectType = 2) or in all the results (effectType = 3)
+read_data <- function(prefix, alpha = .05, effectType = 0, 
+		distributions=c("norm", "lnorm", "exp", "cauchy", "poisson", "binom"),
+		methods = c("PAR", "RNK", "INT", "ART")) {
+
+	df <- read.csv(paste("results/", prefix, ".csv" , sep=""), sep=",", header=TRUE, strip.white=TRUE)
+
+	df <- df[df$family %in% distributions,]
+	df$family <- factor(df$family, levels=distributions)
+	df$method <- factor(df$method, levels=methods)
+
+	if(effectType == 0) {
+		df <- df[(df$effectX2 > 0 & df$effectX1 > 0) | (df$effectX1 == 0 & df$effectX2 == 0),]
+	} else if(effectType == 1) {
+		df <- df[df$effectX2 == 0,]
+	} else if(effectType == 2) {
+		df <- df[df$effectX1 == 0,]
+	} else if(effectType == 3) {
+		df <- df[df$effectX1 > 0,]
+	} else if(effectType == 4) {
+		df <- df[df$effectX2 > 0,]
+	} 
+	else if(effectType == 5) {
+		df <- df[df$effectX1X2 > 0,]
+	} else if(effectType == 6) {
+		df <- df[(df$effectX1X2 > 0 & df$effectX1 > 0) | (df$effectX1X2 == 0 & df$effectX1 == 0),]
+	}
+
+
+	if(is.na(alpha)) return(df)
+	else return(df[df$alpha == alpha,])
+}
+
+# the effectType parameter determines if I'm interested in results where there are effects on both factors (effectType = 0) 
+# or only on the first factor (effectType = 1) or in all the results (effectType = 2)
+readly_data <- function(prefix, alpha = .05, effectType = 0,
+		distributions=c("norm", "lnorm", "exp", "cauchy", "poisson", "binom"),
+		dnames = c("Normal", "Log-normal", "Exponential", "Cauchy", "Poisson", "Binomial"),
+		methods = c("PAR", "RNK", "INT", "ART"), effectvars = c("effectX1","effectX2","effectX1X2")) {
+
[…]

### 2025-12-10 utils-plotlying.R

@@ -1,0 +1,327 @@
+
+library(plotly)
+
+plotly_error <- function(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", ytitle = 'Type I errors (%)', max = 100, 
+	cbPalette = c("#888888", "#E69F00", "#009E73", "#FF5E00"), symbols = c("asterisk", "x", "star-diamond", "star-triangle-up")){
+	# aesthetics
+	
+	margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
+
+	dnames <- levels(df$family)
+	# To be used as labels on the subplots
+	annots <- lapply(1:length(dnames), function(index){list(x = (index-0.5)/length(dnames), y = 1, text = dnames[index], xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F)})
+
+	menuItems <- c("n = 10", "n = 20", "n = 30")
+	yvarnames <- c(paste(var,"10", sep="."), paste(var,"20", sep="."), paste(var,"30", sep="."))
+	
+	createPlot <- function(data, dnames, symbols) {
+	  p <- plot_ly(data,  x = ~factor(eval(as.symbol(xvar))), y = ~100*eval(as.symbol(yvarnames[1])), color = ~method, symbol = ~method, symbols = symbols, colors = cbPalette, type = 'scatter', mode = 'lines+markers', marker = list(line = list(width = 2)), legendgroup = ~method, showlegend = unique(data$family == dnames[1]), visible = F) 
+	  p <- add_trace(p, data = data, x = ~factor(eval(as.symbol(xvar))), y = ~100*eval(as.symbol(yvarnames[2])), color = ~method, symbol = ~method, symbols = symbols, colors = cbPalette, type = 'scatter', mode = 'lines+markers', marker = list(line = list(width = 2)), legendgroup = ~method, showlegend = unique(data$family == dnames[1]), visible = T) 
+	  p <- add_trace(p, data = data, x = ~factor(eval(as.symbol(xvar))), y = ~100*eval(as.symbol(yvarnames[3])), color = ~method, symbol = ~method, symbols = symbols, colors = cbPalette, type = 'scatter', mode = 'lines+markers', marker = list(line = list(width = 2)), legendgroup = ~method, showlegend = unique(data$family == dnames[1]), visible = F) %>%
+	  layout(
[…]

### 2025-12-12 appendix_I.qmd
working on appendix I

---
title: "The illusory promise of the Aligned Rank Transform"
subtitle: "Appendix I. Mathematical explanations and proofs"
author: 
  - name: Theophanis Tsandilas
    orcid: 0000-0002-0158-228X
    email: theophanis.tsandilas@inria.fr
    affiliations:
      - name: Université Paris-Saclay, CNRS, Inria, LISN
        country: France
  - name: Géry Casiez
    orcid: 0000-0003-1905-815X
    email: gery.casiezuniv-lille.fr
    affiliations:
      - name: Univ. Lille, CNRS, Inria, Centrale Lille, UMR 9189 CRIStAL
        country: France
bibliography: bibliography.bib
csl: canadian-journal-of-philosophy.csl

tbl-cap-location: top
---

## Comparing Type I error rates under non-linear models
Consider again the general equation that connects the expected value of a response variable $Y$, which may follow a non-normal distribution, to the predictors in an experimental design with two factors $x_1$ and $x_2$:

$$
E(Y|x_1, x_2) = g^{-1}(a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2)
$$
where $g^{-1}$ is the inverse link function.

We explain why in the main scenarios in this article, the interpretation of effects is not affected by the way the null hypothesis is defined. Our analysis is largely based on related analyses by @Ai:2003, @Greene:2010, and @McCabe:2022.

To simplify our analysis, we use $\eta(x_1, x_2)$ to denote the linear part of the model:

$$
\eta(x_1, x_2) = a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2
$$

We also use $\mu(x_1, x_2)$ to denote the expected value as a function of $x_1$ and $x_2$:

$$
\mu(x_1, x_2) = E(Y|x_1, x_2) = g^{-1}(\eta(x_1, x_2))
$$

### Main effects
We first show that there is no ambiguity in the interpretation of main effects when $a_{12}=0$.

Without loss of generality, we assume that $a_1$ is non-zero and focus on the observed effect on the second factor $x_2$ when $a_2 = 0$. Then: 
$$
\begin{aligned}
\eta(x_1, x_2) = a_0 + a_1 x_1, \qquad
\mu(x_1, x_2) = g^{-1}(a_0 + a_1 x_1)
\end{aligned}
$$

We consider two different situations:

(a) *Continuous $x_2$ --- partial-derivative definition of the effect*. In this case, we can define the effect of $x_2$ on the response as the partial derivative $\frac{\partial \mu}{\partial x_2}$. This becomes:
$$\frac{\partial \mu(x_1, x_2)}{\partial x_2}  = \frac{\partial}{\partial x_2} g^{-1}(a_0 + a_1 x_1) = 0$$
since the term $a_0 + a_1 x_1$ does not contain $x_2$. Thus, $x_2$ has no effect on the responses.

(b) *Categorical $x_2$ --- contrast definition of the effect*. This case is closer to the scenarios we consider in our simulations, where all independent variables are categorical. Let $L_2$ be the categorical levels of $x_2$. Then, the effect of $x_2$ can be defined as the difference in means across its levels for a fixed $x_1$:
$$\Delta_{u,\upsilon}E(Y|x_1, x_2) = \mu(x_1, u) - \mu(x_1, \upsilon) = g^{-1}(a_0 + a_1 x_1) - g^{-1}(a_0 + a_1 x_1) = 0$$
for any two levels $u, \upsilon \in L_2$.

**Conclusion:** Let $a_{12} = 0$. If $a_2 = 0$ then the effect on the mean of responses is also zero. Thus, whether we define Type I errors in terms of the coefficient $a_2$ or in terms of the means of the responses, results should not change. We emphasize that this result **is not specific to means**. If $Y$ depends on $x_1$, $x_2$ only through the predictor $\eta(x_1, x_2)$ --- that is, variances or other distribution shape parameters do not depend on these variables --- then no effect of $x_2$ can appear on the mean, the median, or any other quantile of the distribution when both $a_{12}$ and $a_2$ are zero.

**Special note:** However, if $a_{12}$ is non-zero, it is possible to observe an effect of $x_2$ on responses even if $a_2 = 0$.

### Interactions
We similarly show that there is no ambiguity in the interpretation of the interaction effect when either $a_1 = 0$ or $a_2 =0$. Without loss of generality, we assume below that $a_2 = 0$ while $a_1$ is non-zero. For $a_{12}=0$: 
$$
\begin{aligned}
\eta(x_1, x_2) = a_0 + a_1 x_1, \qquad
\mu(x_1, x_2) = g^{-1}(a_0 + a_1 x_1)
\end{aligned}
$$

A common definition of interaction on the response scale [@Ai:2003] is that the *marginal effect* of $x_1$ depends on $x_2$. Again, we consider two situations:

(a) *Continuous $x_1$ and $x_2$ --- definition of interaction through second-order cross-partial derivative*. For continuous $x_1$, this is expressed by the second-order cross-partial derivative $\frac{\partial^2 \mu}{\partial x_2 \partial x_1}$, which becomes:
 $$\frac{\partial^2 \mu(x_1, x_2)}{\partial x_2 \partial x_1}  = \frac{\partial}{\partial x_2} [\dot{g}^{-1}(\eta(x_1, x_2)) \cdot a_1] = \ddot{g}^{-1}(\eta(x_1, x_2)) \cdot \frac{\partial \eta(x_1, x_2)}{\partial x_2} \cdot a_1$$ 
 based on the rule of chain, where $\dot{g}^{-1}$ and $\ddot{g}^{-1}$ are the first and second derivatives of the inverse link function (assuming that $g^{-1}$ is twice differentiable). But we have:
 $$\frac{\partial \eta(x_1, x_2)}{\partial x_2} = a_2 + a_{12}x_1 = 0$$ because $a_2$ and $a_{12}$ are both zero. Thus, $\frac{\partial^2 \mu}{\partial x_2 \partial x_1}$ becomes zero, which means that the marginal effect (slope) of $x_1$ does not change with $x_2$ (and vice versa). There is therefore no interaction on the response scale.

(b) *Categorical $x_1$ and $x_2$ --- definition of interaction through double discrete difference*. Let $L_1$ and $L_2$ be the categorical levels of $x_1$ and $x_2$, respectively. Then, the interaction effect can be defined as the double difference in means across all combinations of levels in $L_1$ and $L_2$:
$$\Delta_{u_1, \upsilon_1} \Delta_{u_2, \upsilon_2}E(Y|x_1, x_2) = [\mu(u_1,u_2) - \mu(u_1, \upsilon_2)] - [\mu(\upsilon_1,u_2) - \mu(\upsilon_1, \upsilon_2)] $$
for any levels $u_1, \upsilon_1 \in L_1$ and $u_2, \upsilon_2 \in L_2$.

**Conclusion:** Let $a_{12} = 0$. If $a_2 = 0$ then the effect on the mean of responses is also zero. Thus, whether we define Type I errors in terms of the coefficient $a_2$ or in terms of the means of the responses, results should not change.

## References {.unnumbered}

::: {#refs}
:::

### 2025-12-12 text/10-intro.qmd
working on appendix I

@@
 This paper aims to clarify the severity of these issues and understand the potential risks of the method.
-We present the outcomes of a series of Monte Carlo experiments, following a distinctive simulation methodology grounded in latent variable modeling.
-This approach enables us to simulate effects consistently across a broad spectrum of distributions, both discrete and continuous, and allows us to address issues in interpreting main and interaction effects.
+We present the results of a series of Monte Carlo experiments in which we simulate a broad spectrum of distributions, both discrete and continuous.
+Throughout our analysis, we pay special attention to definition-related issues in the null hypothesis, as these can affect the interpretation of the simulation results.
 To ensure the clarity of our findings and facilitate their reproducibility, we divide the experimental process into multiple smaller experiments, where each experiment focuses on a distinct variable (e.g., sample size, experimental design, variance ratio) or measure (Type I error rate, power, precision of effect size estimates).
@@
 Given these new insights, we conclude that ART is not a viable analysis method and advocate for its abandonment.
-We provide recommendations for alternative analysis methods, while we also raise warnings about the interpretation of interaction effects.
+We provide recommendations for alternative analysis methods, while we also raise warnings about the analysis and interpretation of interaction effects.
 

### 2025-12-12 text/20-background.qmd
working on appendix I

@@
 
-Our goal is to verify Lüpsen's findings by employing an alternative set of experimental configurations and a distinctly different methodology.
-In particular, we adopt a latent variable modeling approach that establishes a unified framework for data generation across all distributions.
-This approach allows us to identify interpretation issues regarding the definition of the null hypothesis, especially for interaction effects, and better explain when and why each method fails.
-Moreover, it facilitates the assessment of the methods through more suitable generation procedures for ordinal data, as suggested by @Liddell:2018.
-Our simulation strategy is similar to that of @elkin:2021 but employs a simpler method for generating data from a latent parameter space.
-In addition, we evaluate critical scenarios that @elkin:2021 did not consider, particularly the influence of no-null effects and discrete data.
+Our goal is to verify Lüpsen's findings by employing an alternative set of experimental configurations and a distinctly different methodology based on generalized linear models.
+While our simulation strategy is similar to that of @elkin:2021, we investigate a larger range of experimental settings and critical scenarios the authors did not consider, particularly the influence of no-null effects and discrete data.
 

### 2025-12-12 text/40-interpretingEffects.qmd
working on appendix I

@@
 
-Statistical procedures based on GLMs define interaction in terms of the interaction coefficient $a_{12}$ in the linear component of the model.
-That is, the null hypothesis should be rejected when the coefficient is non-zero.
+Statistical procedures based on GLMs define effects in terms of the coefficients $a_{1}$, $a_{2}$, and $a_{12}$ in the linear component of the model.
+That is, the null hypothesis should be rejected when the corresponding coefficient (e.g., the coefficient $a_{12}$ for the interaction) is non-zero.
 However, there is no consensus among researchers on this approach.
@@
 
-**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects --- that is, when the coefficients of all interacting factors are non-zero.
+Nevertheless, in most of our experiments we select scenarios that avoid these interpretational complications, ensuring that the Type I error rates of the methods we compare do not depend on how one defines interactions.
+We clarify this point below.
+
+**Clarifying when the interpretation of effects becomes ambiguous.** Ambiguities in interpreting interaction effects arise only when all relevant main effects are present --- that is, when the coefficients of all interacting factors are non-zero.
 Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero.
-In such cases, we may observe a change in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.
-We discuss such scenarios in more depth in the [appendix](appendix.html#removable-main).
+In such cases, we may observe changes in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.
+We discuss such scenarios in more detail in the [appendix](appendix.html#removable-main).
 
-By contrast, interpretation is unambiguous in these situations:
+By contrast, interpretation is unambiguous in the following situations:
 
@@
 
+In these cases, error rates (Type I or Type II) are comparable across all methods, regardless of whether the null hypothesis is defined in terms of the coefficients of the linear model or the observed effects on the response variable $Y$.
 For example, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* ($a_{1} = 0$) and no interaction ($a_{12}=0$).
@@
 
+<!--
 ### Our approach {#approach}
@@
 Consequently, an effect on the latent variable may not be observable in the responses, or the observed magnitude of the effect may appear reduced.
+
+-->

### 2025-12-12 text/65-Results-EffectSize.qmd
working on appendix I

@@
 ```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
-data <- read.csv("data/effect-sizes-lognormal.csv", sep=",", header=TRUE, strip.white=TRUE)
+data <- read.csv("results/eta_squared.csv", sep=",", header=TRUE, strip.white=TRUE)
+data <- data[data$family == "lnorm", ]
 data$effectX2 <- factor(data$effectX2)
@@
 
-fig1 <- plot_ly(df, x = ~effectX2, y = ~X1, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig1 <- plot_ly(df, x = ~effectX2, y = ~grtX1, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
 fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
@@
 ```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
-data <- read.csv("data/effect-sizes-normal-interactions.csv", sep=",", header=TRUE, strip.white=TRUE)
-data$effectX2 <- factor(data$effectX2)
+data <- read.csv("results/eta_squared.csv", sep=",", header=TRUE, strip.white=TRUE)
+df <- data[data$family == "norm" & data$effectX1 == data$effectX2, ]
+df$effectX2 <- factor(df$effectX2)
 
@@
 
-df <- data
-
-fig1 <- plot_ly(df, x = ~effectX2, y = ~X1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig1 <- plot_ly(df, x = ~effectX2, y = ~grtX1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
 fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1X2, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
@@
 ```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
-data <- read.csv("data/effect-sizes-ordinal-interactions.csv", sep=",", header=TRUE, strip.white=TRUE)
-data$effectX2 <- factor(data$effectX2)
+df <- data[data$family == "likert" & data$effectX1 == data$effectX2, ]
+df$effectX2 <- factor(df$effectX2)
 
@@
 
-df <- data
-#df <- data[data$effectX1 == 0,]
-
-fig1 <- plot_ly(df, x = ~effectX2, y = ~X1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig1 <- plot_ly(df, x = ~effectX2, y = ~grtX1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
 fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1X2, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))

### 2025-12-12 supplementary/experiments-code/utils-effect-size.R

@@ -144,6 +144,6 @@ repeat_test_effect_size <- function(
   colnames(results) <- c(paste0("par", vars), paste0("rnk", vars), paste0("art", vars), paste0("int", vars), paste0("grt", vars)) 
   coeffs_prefixed <- setNames(coeffs, paste0("effect", names(coeffs)))
 
-  cbind(n = n, designStr=designStr, family = family, as.data.frame(as.list(coeffs_prefixed)), round(results,digits=4))
+  cbind(n = n, design=designStr, family = family, as.data.frame(as.list(coeffs_prefixed)), round(results,digits=4))
 }
 

### 2025-12-13 appendix_I.qmd
added proof of equivalence of null hypothesis definitions

@@
 ## Comparing Type I error rates under non-linear models
-Consider again the general equation that connects the expected value of a response variable $Y$, which may follow a non-normal distribution, to the predictors in an experimental design with two factors $x_1$ and $x_2$:
+Consider again the general relationship between the expected value of a response variable $Y$, which may follow a non-normal distribution, and the predictors in an experimental design with two independent variables $x_1$ and $x_2$:
 
@@
 
-We explain why in the main scenarios in this article, the interpretation of effects is not affected by the way the null hypothesis is defined.
-Our analysis is largely based on related analyses by @Ai:2003, @Greene:2010, and @McCabe:2022.
+We show that, in the main scenarios considered in this article, the interpretation of effects is not affected by whether the null hypothesis is defined on the coefficient scale or on the response-mean scale.
+Our analysis follows related discussions in @Ai:2003, @Greene:2010, and @McCabe:2022.
 
-To simplify our analysis, we use $\eta(x_1, x_2)$ to denote the linear part of the model:
-
+For clarity, let
 $$
@@
 $$
-
-We also use $\mu(x_1, x_2)$ to denote the expected value as a function of $x_1$ and $x_2$:
-
+denote the linear predictor, and
 $$
@@
 $$
+denote the expected value (or mean) as a function of $x_1$ and $x_2$.
 
@@
 
-Without loss of generality, we assume that $a_1$ is non-zero and focus on the observed effect on the second factor $x_2$ when $a_2 = 0$. Then: 
+Without loss of generality, assume that $a_1 \neq 0$ an examine the effect on the second factor $x_2$ when $a_2 = 0$. In that case,
+$$
+\begin{aligned}
+\eta(x_1, x_2) = a_0 + a_1 x_1, \qquad
+\mu(x_1, x_2) = g^{-1}(a_0 + a_1 x_1)
+\end{aligned}
+$$
+so $\mu$ is a function of $x_1$ only.
+
+We consider two situations:
+
+**(a) Continuous $x_2$: effect defined by a partial derivative**
+
+The effect of $x_2$ on the response can be defined as $\frac{\partial \mu}{\partial x_2}$. Because the argument of $g^{-1}$ contains no $x_2$:
+$$\frac{\partial \mu(x_1, x_2)}{\partial x_2}  = \frac{\partial}{\partial x_2} g^{-1}(a_0 + a_1 x_1) = 0$$
+Thus, $x_2$ has no effect on the mean response.
+
+**(b) Categorical $x_2$: effect defined by contrasts**
+
+This is the setting relevant to our simulations. Let $L_2$ denote the levels of $x_2$. For any two levels $u, \nu \in L_2$, the contrast is:
+$$\Delta_{u,\upsilon}\mu(x_1, x_2) = \mu(x_1, u) - \mu(x_1, \upsilon) = g^{-1}(a_0 + a_1 x_1) - g^{-1}(a_0 + a_1 x_1) = 0.$$
+
+#### Conclusion (main effects)
+
+If $a_{12} = 0$ and $a_2 = 0$, then $x_2$ has no effect on the mean of the responses.
+Thus, whether Type I error is defined in terms of the coefficient $a_2$ or in terms of the equality of observed means across levels of $x_2$ makes no difference.
+
+This conclusion **is not limited to means**.
+If the distribution of $Y | (x_1, x_2)$ depends on $x_1$, $x_2$ only through the predictor $\eta(x_1, x_2)$ --- that is, if scale or shape parameters do not depend on the factors --- then no effect of $x_2$ can appear on the mean, the median, or any other quantile of the distribution whenever $a_{12}=0$ and $a_2=0$.
+
+### Interactions
+Similarly, we show that there is no ambiguity in the interpretation of interaction effects when either $a_1 = 0$ or $a_2 =0$.
+Without loss of generality, assume $a_2 = 0$ and $a_1 \neq 0$.
+When $a_{12}=0$,
 $$
@@
 
-We consider two different situations:
+A common definition of interaction on the response scale [@Ai:2003] is that the marginal effect of $x_1$ depends on $x_2$. We examine two situations:
 
-(a) *Continuous $x_2$ --- partial-derivative definition of the effect*.
-In this case, we can define the effect of $x_2$ on the response as the partial derivative $\frac{\partial \mu}{\partial x_2}$.
-This becomes:
-$$\frac{\partial \mu(x_1, x_2)}{\partial x_2}  = \frac{\partial}{\partial x_2} g^{-1}(a_0 + a_1 x_1) = 0$$
-since the term $a_0 + a_1 x_1$ does not contain $x_2$. Thus, $x_2$ has no effect on the responses.
+**(a) Continuous predictors: interaction defined by a cross-partial derivative.**
 
-(b) *Categorical $x_2$ --- contrast definition of the effect*.
-This case is closer to the scenarios we consider in our simulations, where all independent variables are categorical.
-Let $L_2$ be the categorical levels of $x_2$.
-Then, the effect of $x_2$ can be defined as the difference in means across its levels for a fixed $x_1$:
-$$\Delta_{u,\upsilon}E(Y|x_1, x_2) = \mu(x_1, u) - \mu(x_1, \upsilon) = g^{-1}(a_0 + a_1 x_1) - g^{-1}(a_0 + a_1 x_1) = 0$$
-for any two levels $u, \upsilon \in L_2$.
+Assuming $g^{-1}$ is twice differentiable, an interaction is present if $\frac{\partial^2 \mu}{\partial x_2 \partial x_1} \neq 0$.
 
-**Conclusion:** Let $a_{12} = 0$.
-If $a_2 = 0$ then the effect on the mean of responses is also zero.
-Thus, whether we define Type I errors in terms of the coefficient $a_2$ or in terms of the means of the responses, results should not change.
-We emphasize that this result **is not specific to means**.
-If $Y$ depends on $x_1$, $x_2$ only through the predictor $\eta(x_1, x_2)$ --- that is, variances or other distribution shape parameters do not depend on these variables --- then no effect of $x_2$ can appear on the mean, the median, or any other quantile of the distribution when both $a_{12}$ and $a_2$ are zero.
+We compute:
+$$\frac{\partial \mu}{\partial x_1} = \dot{g}^{-1}(\eta) \cdot a_1$$
 
-**Special note:** However, if $a_{12}$ is non-zero, it is possible to observe an effect of $x_2$ on responses even if $a_2 = 0$.
+and
 
-### Interactions
-We similarly show that there is no ambiguity in the interpretation of the interaction effect when either $a_1 = 0$ or $a_2 =0$.
-Without loss of generality, we assume below that $a_2 = 0$ while $a_1$ is non-zero.
-For $a_{12}=0$:
-$$
-\begin{aligned}
-\eta(x_1, x_2) = a_0 + a_1 x_1, \qquad
-\mu(x_1, x_2) = g^{-1}(a_0 + a_1 x_1)
-\end{aligned}
-$$
+$$\frac{\partial^2 \mu}{\partial x_2 \partial x_1} = \ddot{g}^{-1}(\eta) \cdot \frac{\partial \eta}{\partial x_2} \cdot a_1$$
 
-A common definition of interaction on the response scale [@Ai:2003] is that the *marginal effect* of $x_1$ depends on $x_2$. Again, we consider two situations:
+But for our settings ($a_2=a_{12}=0$)
 
-(a) *Continuous $x_1$ and $x_2$ --- definition of interaction through second-order cross-partial derivative*.
-For continuous $x_1$, this is expressed by the second-order cross-partial derivative $\frac{\partial^2 \mu}{\partial x_2 \partial x_1}$, which becomes:
- $$\frac{\partial^2 \mu(x_1, x_2)}{\partial x_2 \partial x_1}  = \frac{\partial}{\partial x_2} [\dot{g}^{-1}(\eta(x_1, x_2)) \cdot a_1] = \ddot{g}^{-1}(\eta(x_1, x_2)) \cdot \frac{\partial \eta(x_1, x_2)}{\partial x_2} \cdot a_1$$
- based on the rule of chain, where $\dot{g}^{-1}$ and $\ddot{g}^{-1}$ are the first and second derivatives of the inverse link function (assuming that $g^{-1}$ is twice differentiable). But we have:
- $$\frac{\partial \eta(x_1, x_2)}{\partial x_2} = a_2 + a_{12}x_1 = 0$$ because $a_2$ and $a_{12}$ are both zero.
-Thus, $\frac{\partial^2 \mu}{\partial x_2 \partial x_1}$ becomes zero, which means that the marginal effect (slope) of $x_1$ does not change with $x_2$ (and vice versa).
-There is therefore no interaction on the response scale.
+$$\frac{\partial \eta}{\partial x_2} = a_2 + a_{12}x_1 = 0$$
 
-(b) *Categorical $x_1$ and $x_2$ --- definition of interaction through double discrete difference*.
-Let $L_1$ and $L_2$ be the categorical levels of $x_1$ and $x_2$, respectively.
-Then, the interaction effect can be defined as the double difference in means across all combinations of levels in $L_1$ and $L_2$:
-$$\Delta_{u_1, \upsilon_1} \Delta_{u_2, \upsilon_2}E(Y|x_1, x_2) = [\mu(u_1,u_2) - \mu(u_1, \upsilon_2)] - [\mu(\upsilon_1,u_2) - \mu(\upsilon_1, \upsilon_2)] $$
-for any levels $u_1, \upsilon_1 \in L_1$ and $u_2, \upsilon_2 \in L_2$.
+and therefore
 
-**Conclusion:** Let $a_{12} = 0$.
-If $a_2 = 0$ then the effect on the mean of responses is also zero.
-Thus, whether we define Type I errors in terms of the coefficient $a_2$ or in terms of the means of the responses, results should not change.
+$$\frac{\partial^2 \mu}{\partial x_2 \partial x_1} = 0$$
+
+ Thus the marginal effect of $x_1$ does not vary with $x_2$, and there is no interaction.
+
+**(b) Categorical predictors: interaction defined by double discrete differences**
+
+Let $L_1$ and $L_2$ be the levels of $x_1$ and $x_2$, respectively. The categorical analogue of a cross-partial derivative is the double difference:
+
+$$\Delta_{u_1, \nu_1} \Delta_{u_2, \nu_2}\mu(x_1, x_2) = [\mu(u_1,u_2) - \mu(u_1, \nu_2)] - [\mu(\nu_1,u_2) - \mu(\nu_1, \nu_2)]$$
+
+But when $a_2 = a_{12} = 0$, $\mu(u_1,u_2) = \mu(u_1,\nu_2)$ and $\mu(\nu_1,u_2) = \mu(\nu_1,\nu_2)$, so the double difference is always zero for any levels $u_1, \nu_1 \in L_1$ and $u_2, \nu_2 \in L_2$.
+
+#### Conclusion (interaction effects)
+
+If either $a_1 = 0$ or $a_2 = 0$ and $a_{12} = 0$, then no interaction can be observed on the response scale.
+Thus Type I error definitions based on the coefficient $a_{12}$ or on observed interaction contrasts are equivalent in these settings.
 

### 2025-12-15 appendix_I.qmd
appendix i

@@
 title: "The illusory promise of the Aligned Rank Transform"
-subtitle: "Appendix I. Mathematical explanations and proofs"
+subtitle: "Appendix I. Analytical derivations and mathematical proofs"
 author: 
@@
 
-## Comparing Type I error rates under non-linear models
-Consider again the general relationship between the expected value of a response variable $Y$, which may follow a non-normal distribution, and the predictors in an experimental design with two independent variables $x_1$ and $x_2$:
-
-$$
-E(Y|x_1, x_2) = g^{-1}(a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2)
-$$
-where $g^{-1}$ is the inverse link function.
-
-We show that, in the main scenarios considered in this article, the interpretation of effects is not affected by whether the null hypothesis is defined on the coefficient scale or on the response-mean scale.
-Our analysis follows related discussions in @Ai:2003, @Greene:2010, and @McCabe:2022.
-
-For clarity, let
+## Absence of spurious effects when $a_2=0$ and $a_{12} = 0${#spurious-absence}
+Let $x_1$ and $x_2$ be two independent variables, and suppose the conditional distribution of the response $Y | (x_1, x_2)$ depends on these variables only through the linear predictor:
 $$
@@
 $$
-denote the linear predictor, and
+
+Assume that the conditional mean can be written as:
 $$
-\mu(x_1, x_2) = E(Y|x_1, x_2) = g^{-1}(\eta(x_1, x_2))
+\mu(x_1, x_2) = E(Y|x_1, x_2) = h(\eta(x_1, x_2))
 $$
-denote the expected value (or mean) as a function of $x_1$ and $x_2$.
+where $h$ is a monotone function.
+This function may coincide with the inverse link function $g^{-1}$ in a GLM, but it may also arise from other frameworks, including transformation models or latent-variable constructions.
+Thus it does not necessarily need to be smooth or invertible; for example, this assumption is not required for the ordinal example discussed later.
 
-### Main effects
-We first show that there is no ambiguity in the interpretation of main effects when $a_{12}=0$.
+Suppose $a_2 = 0$ and $a_{12} = 0$. Then:
 
-Without loss of generality, assume that $a_1 \neq 0$ an examine the effect on the second factor $x_2$ when $a_2 = 0$. In that case,
-$$
-\begin{aligned}
-\eta(x_1, x_2) = a_0 + a_1 x_1, \qquad
-\mu(x_1, x_2) = g^{-1}(a_0 + a_1 x_1)
-\end{aligned}
-$$
-so $\mu$ is a function of $x_1$ only.
+1. The independent variable $x_2$ has no effect on the response scale.
+2. There is no interaction between $x_1$ and $x_2$ on the response scale.
 
-We consider two situations:
+Thus, whether Type I errors are defined in terms of the model coefficients $a_2$ (when $a_{12} = 0$) or $a_{12}$ (when $a_2 = 0$) or in terms of the patterns of observed means (and alternatively medians or other quantiles) on the response scale makes no difference.
 
-**(a) Continuous $x_2$: effect defined by a partial derivative**
+### Proof
+Under $a_2 = a_{12} = 0$, the linear predictor reduces to
 
-The effect of $x_2$ on the response can be defined as $\frac{\partial \mu}{\partial x_2}$. Because the argument of $g^{-1}$ contains no $x_2$:
-$$\frac{\partial \mu(x_1, x_2)}{\partial x_2}  = \frac{\partial}{\partial x_2} g^{-1}(a_0 + a_1 x_1) = 0$$
-Thus, $x_2$ has no effect on the mean response.
+$$\eta(x_1, x_2) = a_0 + a_1x_1$$
 
-**(b) Categorical $x_2$: effect defined by contrasts**
+which does not depend on $x_2$.
+Hence, the conditional distribution $Y | (x_1, x_2)$, including its mean, median, and all quantiles, is invariant to $x_2$.
+This establishes the absence of a main effect of $x_2$ or an effect of its interaction $x_2$.
 
-This is the setting relevant to our simulations. Let $L_2$ denote the levels of $x_2$. For any two levels $u, \nu \in L_2$, the contrast is:
-$$\Delta_{u,\upsilon}\mu(x_1, x_2) = \mu(x_1, u) - \mu(x_1, \upsilon) = g^{-1}(a_0 + a_1 x_1) - g^{-1}(a_0 + a_1 x_1) = 0.$$
+We further explain this results for special cases, where we largely rely on the relevant discussions of @Ai:2003, @Greene:2010, and @McCabe:2022.
 
-#### Conclusion (main effects)
+#### Main effect of $x_2$
+If $x_2$ is continuous and $h$ is differentiable, the effect of $x_2$ on the response can be defined as:
 
-If $a_{12} = 0$ and $a_2 = 0$, then $x_2$ has no effect on the mean of the responses.
-Thus, whether Type I error is defined in terms of the coefficient $a_2$ or in terms of the equality of observed means across levels of $x_2$ makes no difference.
+$$\frac{\partial \mu(x_1, x_2)}{\partial x_2}  = h'(\eta) \frac{\partial \eta}{\partial x_2} = 0$$
+where $h'$ denotes the first derivative of $h$.
 
-This conclusion **is not limited to means**.
-If the distribution of $Y | (x_1, x_2)$ depends on $x_1$, $x_2$ only through the predictor $\eta(x_1, x_2)$ --- that is, if scale or shape parameters do not depend on the factors --- then no effect of $x_2$ can appear on the mean, the median, or any other quantile of the distribution whenever $a_{12}=0$ and $a_2=0$.
+If $x_2$ is categorical, all contrasts across its levels are zero because the mean response is identical at each level:
+$$\Delta_{u,\nu}\mu(x_1, x_2) = \mu(x_1, u) - \mu(x_1, \nu) = h(a_0 + a_1 x_1) - h(a_0 + a_1 x_1) = 0$$
+where $u, \nu$ are any two levels of $x_2$.
 
-### Interactions
-Similarly, we show that there is no ambiguity in the interpretation of interaction effects when either $a_1 = 0$ or $a_2 =0$.
-Without loss of generality, assume $a_2 = 0$ and $a_1 \neq 0$.
-When $a_{12}=0$,
-$$
-\begin{aligned}
-\eta(x_1, x_2) = a_0 + a_1 x_1, \qquad
-\mu(x_1, x_2) = g^{-1}(a_0 + a_1 x_1)
-\end{aligned}
-$$
+#### Interaction between $x_1$ and $x_2$
+An interaction on the response scale is present if the marginal effect of $x_1$ depends on $x_2$.
+For continuous predictors and assuming that $h$ is twice differentiable, this is characterized by the mixed partial derivative $\frac{\partial^2 \mu}{\partial x_2 \partial x_1}$.
 
-A common definition of interaction on the response scale [@Ai:2003] is that the marginal effect of $x_1$ depends on $x_2$. We examine two situations:
+Using the chain rule:
 
-**(a) Continuous predictors: interaction defined by a cross-partial derivative.**
-
-Assuming $g^{-1}$ is twice differentiable, an interaction is present if $\frac{\partial^2 \mu}{\partial x_2 \partial x_1} \neq 0$.
-
-We compute:
-$$\frac{\partial \mu}{\partial x_1} = \dot{g}^{-1}(\eta) \cdot a_1$$
-
-and
-
-$$\frac{\partial^2 \mu}{\partial x_2 \partial x_1} = \ddot{g}^{-1}(\eta) \cdot \frac{\partial \eta}{\partial x_2} \cdot a_1$$
-
-But for our settings ($a_2=a_{12}=0$)
-
-$$\frac{\partial \eta}{\partial x_2} = a_2 + a_{12}x_1 = 0$$
-
-and therefore
-
-$$\frac{\partial^2 \mu}{\partial x_2 \partial x_1} = 0$$
+$$\frac{\partial^2 \mu(x_1, x_2)}{\partial x_2 \partial x_1} = h''(\eta) \frac{\partial \eta}{\partial x_2} \frac{\partial \eta}{\partial x_1} + h'(\eta)\frac{\partial^2\eta}{\partial x_2 \partial x_1} = 0$$
 
@@
 
-**(b) Categorical predictors: interaction defined by double discrete differences**
+For categorical predictors, the analogue of a mixed partial derivative is the double difference:
 
-Let $L_1$ and $L_2$ be the levels of $x_1$ and $x_2$, respectively. The categorical analogue of a cross-partial derivative is the double difference:
+$$\Delta_{u_1, \nu_1} \Delta_{u_2, \nu_2}\mu(x_1, x_2) = [\mu(u_1,u_2) - \mu(u_1, \nu_2)] - [\mu(\nu_1,u_2) - \mu(\nu_1, \nu_2)]=$$ 
+$$= [h(a_0 + a_1 u_1) - h(a_0 + a_1 u_1)] - [h(a_0 + a_1 \nu_2) - h(a_0 + a_1 \nu_1)] = 0$$
 
-$$\Delta_{u_1, \nu_1} \Delta_{u_2, \nu_2}\mu(x_1, x_2) = [\mu(u_1,u_2) - \mu(u_1, \nu_2)] - [\mu(\nu_1,u_2) - \mu(\nu_1, \nu_2)]$$
+for any levels $u_1, \nu_1$ of $x_1$ and $u_2, \nu_2$ of $x_2$. Again, there is no observed interaction.
 
-But when $a_2 = a_{12} = 0$, $\mu(u_1,u_2) = \mu(u_1,\nu_2)$ and $\mu(\nu_1,u_2) = \mu(\nu_1,\nu_2)$, so the double difference is always zero for any levels $u_1, \nu_1 \in L_1$ and $u_2, \nu_2 \in L_2$.
+### Illustrative examples
+We discuss two examples.
 
-#### Conclusion (interaction effects)
+#### Lognormal data
+Let $x_1$ and $x_2$ be two experimental factors. Suppose the true model of the data is
+$$
+\begin{aligned}
+log Y = a_0 + a_1 x_1 + \varepsilon, \qquad
+\varepsilon \sim \mathcal{N}(0, \sigma^2)
+\end{aligned} 
+$$
+so that 
+$$Y | (x_1, x_2) \sim Lognormal(a_0 + a_1 x_1, \sigma^2)$$
 
-If either $a_1 = 0$ or $a_2 = 0$ and $a_{12} = 0$, then no interaction can be observed on the response scale.
-Thus Type I error definitions based on the coefficient $a_{12}$ or on observed interaction contrasts are equivalent in these settings.
+Since the conditional distribution of $Y$ does not depend on $x_2$, there is no main effect of $x_2$ on the mean, median, or any quantile of the response distribution.
+In particular, the conditional mean
+$$\mu(x_1,x_2) = E(Y|x_1, x_2) = exp(a_0 + a_1 x_1 + \frac{1}{2}\sigma^2)$$
+remains identical across all levels of $x_2$.
+
+For the same reasons, there is no interaction on the response scale.
+
+#### Ordinal responses
+Suppose there exists a latent continuous variable:
+
+$$Y^* = a_0 + a_1 x_1 + \varepsilon$$
+
+and the observed response $Y$ is obtained by applying fixed thresholds: 
+$$
+\begin{aligned}
+Y = k\ \ \text{if}\ \ \tau_{k-1} < Y^* \leq \tau_k, 
+\qquad k = 1,2 ... K
+\end{aligned} 
+$$ 
+Again, the conditional distribution $Y| (x_1, x_2)$ depends on the independent variables only through $\eta(x_1, x_2)=a_0+a_1 x_1$, and there is no effect of $x_2$ on either the latent or the observed ordinal scale.
+
+In this case, the mapping from $\eta$ to the observed response is monotone but not strictly monotone and is therefore non-invertible.
+As a result, a true effect of $x_2$ on the latent variable (i.e., $a_2 \neq 0$) may be attenuated or entirely unobservable on the ordinal response scale.
+This does not affect Type I error rates, provided the null hypothesis is defined on the latent scale and the thresholds are fixed, because the absence of a latent effect implies the absence of an effect on the observed scale.
+However, power and effect sizes can be reduced when effects are assessed on the ordinal scale, reflecting the loss of information due to discretization.
+
+<!--
+### Invariance under monotone transformations 
+Suppose we apply a monotone transformation $T$ directly to the observed responses---such as the rank transformation (RNK) or the inverse normal transformation (INT).
+Because $T$ is monotone, it preserves the ordering of observations within each condition.
+The transformed conditional distribution $T(Y) | (x_1, x_2)$ therefore depends on the independent variables only through $\eta(x_1, x_2) = a_0+a_1 x_1$.
+Consequently, the mean, median, and all quantiles of $T(Y) | (x_1, x_2)$ remain invariant to $x_2$, and marginal effects of $x_1$ do not depend on $x_2$.
+
+In conclusion, $T$ does not introduce a main effect of $x_2$ or an interaction between $x_1$ and $x_2$ when $a_2=0$ and $a_{12}=0$.
+
+### ART introducing spurious effects 
+ART differs fundamentally from simple monotone transformations.
+Instead of transforming $Y$ directly, ART first constructs aligned responses by **subtracting and adding sample means on the response scale**, for example:
+
+$$Y^{(B)} = Y - \bar{Y}_{ij} + \bar{Y}_{.j} - \bar{Y}_{..}$$
+or analogous expressions for interactions, and only then applies ranking.
+The conditional distribution $Y^{(B)} | (x_1, x_2)$ now depends on $x_2$ through the arithmetic mean components even though $a_2 = 0$ and $a_{12}=0$.
+
+These alignment terms can vary across levels of $x_2$ due to sampling variability alone, and depending on $h$, they can acquire systematic shifts across levels of $x_2$ or combinations of $(x_1, x_2)$, even when the true conditional distribution does not.
+in particular, the term $\bar{Y}_{.j} - \bar{Y}_{..}$ is constant within each level of $x_2$ and shifts all aligned observations for that level either upward or downward.
+
+After ranking, observations from one level of $x_2$ tend to receive systematically higher or lower ranks. Our results confirm this problem.
+-->
 

### 2025-12-18 appendix_I.qmd
methodology..

--- 2025-12-13 appendix_I.qmd
@@
 title: "The illusory promise of the Aligned Rank Transform"
-subtitle: "Appendix I. Mathematical explanations and proofs"
+subtitle: "Appendix I. Additional experimental results"
 author: 
@@
 
-## Comparing Type I error rates under non-linear models
-Consider again the general relationship between the expected value of a response variable $Y$, which may follow a non-normal distribution, and the predictors in an experimental design with two independent variables $x_1$ and $x_2$:
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+# Useful libraries
+library(crosstalk)
+library(kableExtra)
+library(gridExtra)
+library(lmerTest)
+library(tidyverse)
+library(plotly)
+```
 
-$$
-E(Y|x_1, x_2) = g^{-1}(a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2)
-$$
-where $g^{-1}$ is the inverse link function.
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+# Source code for reading data from experimental results and for plotting
+source("dataReaders.R")
+source("plotting.R")
 
-We show that, in the main scenarios considered in this article, the interpretation of effects is not affected by whether the null hypothesis is defined on the coefficient scale or on the response-mean scale.
-Our analysis follows related discussions in @Ai:2003, @Greene:2010, and @McCabe:2022.
+library(dplyr)
+library(tidyverse)
+source("plotlying.R")
+source("interactions_plot.R")
+```
 
-For clarity, let
-$$
-\eta(x_1, x_2) = a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2
-$$
-denote the linear predictor, and
-$$
-\mu(x_1, x_2) = E(Y|x_1, x_2) = g^{-1}(\eta(x_1, x_2))
-$$
-denote the expected value (or mean) as a function of $x_1$ and $x_2$.
+```{=html}
+<style>
+.math.inline .MathJax  {
+  font-size: 105% !important;
+}
+</style>
+```
 
-### Main effects
-We first show that there is no ambiguity in the interpretation of main effects when $a_{12}=0$.
+We present complementary results and new experiments that investigate additional scenarios.
+We also compare INT and RNK with other nonparametric methods.
+Unless explicitly mentioned in each section, we follow the experimental methodology presented in the main article.
+At the end of each section, we summarize our conclusions.
 
-Without loss of generality, assume that $a_1 \neq 0$ an examine the effect on the second factor $x_2$ when $a_2 = 0$. In that case,
-$$
-\begin{aligned}
-\eta(x_1, x_2) = a_0 + a_1 x_1, \qquad
-\mu(x_1, x_2) = g^{-1}(a_0 + a_1 x_1)
-\end{aligned}
-$$
-so $\mu$ is a function of $x_1$ only.
+## Results for $\alpha = .01$ {#alpha}
+Although we only presented results for $\alpha = .05$ in the main article, we observe the same trends for other significance levels.
+The following figures show the Type I error rates of the methods for the $4 \times 3$ within-subjects design when $\alpha = .01$.
+Note that error rates are not proportional to the $\alpha$ level.
+For results from other experiments, we refer readers to our raw result files.
 
-We consider two situations:
+**Main effects**. @fig-ratio-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases.
 
-**(a) Continuous $x_2$: effect defined by a partial derivative**
+::: {#fig-ratio-main}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix1 <- "1_test_4x3_Ratio"
+alpha <- .01
+distributions1 <- c("norm", "lnorm", "exp", "poisson", "binom")
+dnames1 <- c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial")
+df1 <- readlyData(prefix1, alpha, 1, distributions1, dnames1)
 
-The effect of $x_2$ on the response can be defined as $\frac{\partial \mu}{\partial x_2}$. Because the argument of $g^{-1}$ contains no $x_2$:
-$$\frac{\partial \mu(x_1, x_2)}{\partial x_2}  = \frac{\partial}{\partial x_2} g^{-1}(a_0 + a_1 x_1) = 0$$
-Thus, $x_2$ has no effect on the mean response.
+prefix2 <- "2_test_4x3_Ordinal"
+distributions2 <- c("likert5B")
+dnames2 <- c("Ordinal (5 levels)")
+df2 <- readlyData(prefix2, alpha, 1, distributions2, dnames2)
 
-**(b) Categorical $x_2$: effect defined by contrasts**
+df <- rbind(df1,df2)
 
-This is the setting relevant to our simulations. Let $L_2$ denote the levels of $x_2$. For any two levels $u, \nu \in L_2$, the contrast is:
-$$\Delta_{u,\upsilon}\mu(x_1, x_2) = \mu(x_1, u) - \mu(x_1, \upsilon) = g^{-1}(a_0 + a_1 x_1) - g^{-1}(a_0 + a_1 x_1) = 0.$$
+plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 27.1)
+```
+Type I error rates ($\alpha = .01$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
+:::
 
-#### Conclusion (main effects)
+**Interaction effects**.
+@fig-ratio-interaction-1 presents Type I error rates for the interaction effect $X_1 \times X_2$, when the main effect on $X_2$ is zero while the main effect on $X_1$ increases.
 
-If $a_{12} = 0$ and $a_2 = 0$, then $x_2$ has no effect on the mean of the responses.
-Thus, whether Type I error is defined in terms of the coefficient $a_2$ or in terms of the equality of observed means across levels of $x_2$ makes no difference.
+::: {#fig-ratio-interaction-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+plotlyError(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 27.1)
+```
+Type I error rates ($\alpha = .01$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
+:::
 
-This conclusion **is not limited to means**.
-If the distribution of $Y | (x_1, x_2)$ depends on $x_1$, $x_2$ only through the predictor $\eta(x_1, x_2)$ --- that is, if scale or shape parameters do not depend on the factors --- then no effect of $x_2$ can appear on the mean, the median, or any other quantile of the distribution whenever $a_{12}=0$ and $a_2=0$.
+**Interactions under parallel main effects**.
+We also present results for interaction effects when the two main effects change in parallel.
+As we explain in the main paper, these results require special attention, as error rates largely depend on the way we define the null hypothesis for interactions.
 
-### Interactions
-Similarly, we show that there is no ambiguity in the interpretation of interaction effects when either $a_1 = 0$ or $a_2 =0$.
-Without loss of generality, assume $a_2 = 0$ and $a_1 \neq 0$.
-When $a_{12}=0$,
-$$
-\begin{aligned}
-\eta(x_1, x_2) = a_0 + a_1 x_1, \qquad
-\mu(x_1, x_2) = g^{-1}(a_0 + a_1 x_1)
-\end{aligned}
-$$
+::: {#fig-ratio-interaction-2}
+```{r, echo=FALSE, message=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df1 <- readlyData(prefix1, alpha, 0, distributions1, dnames1)
+df2 <- readlyData(prefix2, alpha, 0, distributions2, dnames2)
+df <- rbind(df1,df2)
 
-A common definition of interaction on the response scale [@Ai:2003] is that the marginal effect of $x_1$ depends on $x_2$. We examine two situations:
+plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 104)
+```
+Type I error rates ($\alpha = .01$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$.
+*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
+A different definition may lead to different results.*
+:::
 
-**(a) Continuous predictors: interaction defined by a cross-partial derivative.**
+### Conclusion
+Results for $\alpha = .01$ are consistent with our results for $\alpha = .05$, confirming that ART is not a robust method.
 
-Assuming $g^{-1}$ is twice differentiable, an interaction is present if $\frac{\partial^2 \mu}{\partial x_2 \partial x_1} \neq 0$.
+## Missing data {#missing}
+We evaluate how missing data can affect the performance of the four methods.
+Specifically, we study a scenario, where a random sample of $10\%$ of the observations is missing.
+Missing observations lead to unbalanced designs.
+However, we emphasize that our scenario does not cover systematic imbalances due to missing data for specific levels of a factor.
 
-We compute:
-$$\frac{\partial \mu}{\partial x_1} = \dot{g}^{-1}(\eta) \cdot a_1$$
+**Main effects**.
+@fig-missing-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases.
+We observe that missing data make the error rate for ART to increase even further, with a larger increase for the mixed design.
+This is also the case for the normal distribution.
+In contrast, the accuracy of the three other methods does not seem to be affected.
 
-and
+::: {#fig-missing-main}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Appendix_test_missing"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
 
-$$\frac{\partial^2 \mu}{\partial x_2 \partial x_1} = \ddot{g}^{-1}(\eta) \cdot \frac{\partial \eta}{\partial x_2} \cdot a_1$$
+df <- readData(prefix, n = 20, alpha = .05, effectType = 1, distributions)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 98)
 
-But for our settings ($a_2=a_{12}=0$)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
+:::
 
-$$\frac{\partial \eta}{\partial x_2} = a_2 + a_{12}x_1 = 0$$
+**Interaction effects**.
+We also present Type I error rates for the interaction effect in the presence of a single main effect or two parallel main effects.
+The error levels for all methods, including ART, are now very similar to the ones observed with no missing data.
 
-and therefore
+::: {#fig-missing-interaction1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 98)
 
-$$\frac{\partial^2 \mu}{\partial x_2 \partial x_1} = 0$$
+```
+Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
+:::
 
- Thus the marginal effect of $x_1$ does not vary with $x_2$, and there is no interaction.
+### Conclusion
+ART is sensitive to the presence of missing data when at least $10\%$ of the observations are missing.
+Specifically, we found that its Type I error rate for main effects increases further, even with normal distributions.
+The other methods do not seem to be affected.
+However, we emphasize that we only tested missing data that are randomly drawn from the complete set of observations.
+Results might be different if there were systematic bias in the data imbalance structure.
 
-**(b) Categorical predictors: interaction defined by double discrete differences**
+## Log-normal distributions {#log-normal}
+In a different experiment, we evaluate log-normal distributions with a wider range of $\sigma$ parameters (see @fig-lognormal-distributions), in particular distributions with less variance, which exhibit a lower degree of skew.
 
-Let $L_1$ and $L_2$ be the levels of $x_1$ and $x_2$, respectively. The categorical analogue of a cross-partial derivative is the double difference:
+::: {#fig-lognormal-distributions}
+```{r, echo=FALSE, fig.height=1.3, fig.width = 9, warning=FALSE,  message=FALSE}
+source("effects_plot.R")
+plotEffectsLognormal(nlevels = 2, effectSize = 2)
+```
+Log-normal distributions that we evaluate, shown here for a factor with two levels and a magnitude of effect $a_{i} = 2$. 
+:::
 
-$$\Delta_{u_1, \nu_1} \Delta_{u_2, \nu_2}\mu(x_1, x_2) = [\mu(u_1,u_2) - \mu(u_1, \nu_2)] - [\mu(\nu_1,u_2) - \mu(\nu_1, \nu_2)]$$
+**Main effects**.
+@fig-lognormal-main presents our results on Type I error rates for main effects.
+As expected, ART's inflation of error rates is less serious when distributions are closer to normal, while the problem becomes worse as distributions are more skewed.
 
-But when $a_2 = a_{12} = 0$, $\mu(u_1,u_2) = \mu(u_1,\nu_2)$ and $\mu(\nu_1,u_2) = \mu(\nu_1,\nu_2)$, so the double difference is always zero for any levels $u_1, \nu_1 \in L_1$ and $u_2, \nu_2 \in L_2$.
+::: {#fig-lognormal-main}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Appendix_test_lognormal"
+distributions = c("lnorm-0.2", "lnorm-0.4", "lnorm-0.6", "lnorm-0.8", "lnorm-1.0", "lnorm-1.2")
+dnames = c("sd = 0.2", "sd = 0.4", "sd = 0.6", "sd = 0.8", "sd = 1.0", "sd = 1.2")
 
-#### Conclusion (interaction effects)
+df <- readData(prefix, n = 20, alpha = .05, effectType = 1, distributions)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 69)
 
-If either $a_1 = 0$ or $a_2 = 0$ and $a_{12} = 0$, then no interaction can be observed on the response scale.
-Thus Type I error definitions based on the coefficient $a_{12}$ or on observed interaction contrasts are equivalent in these settings.
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
+:::
+
+**Interaction effects**. We observe similar patterns for the Type I error rate of the interaction effect in the presence of a single main effect.
+
+::: {#fig-lognormal-interaction1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 69)
+
+```
+Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
+:::
+
+**Interactions under parallel main effects**.
+We also present results when the two main effects increase in parallel.
+These results require again careful interpretation, as the null hypothesis of interest may be different for each method.
+It is important to note that even slight departures from normality (e.g., when $\sigma = 0.2$) make the interpretation of interactions problematic.
+
+::: {#fig-lognormal-interaction2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- readData(prefix, n = 20, alpha = .05, effectType = 0, distributions)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 103)
+
+```
+Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitudes $a_{1}$ and $a_{2}$ of the main effects ($n=20$).
+*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
+A different definition may lead to different results.*
+:::
+
+### Conclusion
+ART's robustness issues become less severe as log-normal distributions become less skewed and approach normality.
+However, interpretation issues for interactions (in the presence of parallel main effects) can arise even when distributions are only slightly skewed.
+
+## Log-normal distributions with equal variances
+As we explain in the article, skewed distributions naturally lead to unequal variances, with the standard deviation of response times typically proportional to the mean.
+Since ART confounds effects under conditions of unequal variance, an important question remains: *Does ART’s difficulty with continuous skewed distributions stem solely from issues of heteroscedasticity?*
+
+To address this, we present results from an additional experiment using log-normal distributions in which we directly control the population parameters of the observed distributions (rather than those of the latent variable) to ensure constant variances.
+Specifically, given a mean $\mu$ and a fixed standard deviation $\sigma$ for the responses, the corresponding parameters of the log-normal distribution are:
+
+$\sigma_{log} = \sqrt{log(1 + \frac{\sigma^2}{\mu^2})}$ and $\mu_{log} = \log(\frac{\mu^2}{\sqrt{\sigma^2 + \mu^2}})$
+
+We again consider a $4 \times 3$ within-subjects design with $n = 20$, setting the main effect of the first factor to $a_1 = 2$.
+We then measure the Power (effect of $X_1$) and Type I error rates (effects $X_2$ and $X_1 \times X_2$) of the methods across three standard deviations: $\sigma = 1$, $\sigma = 2$, and $\sigma = 3$.
+
+The following table presents our results.
+Once again, ART inflates Type I error rates for both the effect of $X_2$ and its interaction with $X_1$, with error rates increasing as the common standard deviation grows.
+In contrast, PAR shows a loss in statistical power, while RNK and INT remain robust.
+
+::: {#tbl-homoscedastic layout-ncol=3}
+|     | PAR | RNK | INT | ART |
+|:-----:|----:|----:|----:|----:|
+|$X_1$| 100 | 100 | 100 | 100 |
+|$X_2$| 5.0 | 4.9 | 5.4 | 7.6 |
+|$X_1\times X_2$| 3.6 | 4.2 | 4.6 | 5.7 |
+
+: $\sigma = 1$
+
+| PAR | RNK | INT | ART |
+|----:|----:|----:|----:|
+| 98.8 | 100 | 100 | 100 |
+| 4.0 | 5.1  | 5.3 | 12.1 |
+| 3.5 | 5.3 |  5.1 | 9.3  |
+
+: $\sigma = 2$
+
+| PAR | RNK | INT | ART |
+|----:|----:|----:|----:|
+| 94.0 | 100 | 100 | 100 |
+| 3.6 | 4.4  | 4.7 | 17.8 |
+| 3.0 | 4.7  | 5.0 | 11.2 |
+
+: $\sigma = 3$
+
+Percentage of positives (%) over 5000 repetitions ($\alpha = .05$). For $X_1$, values reflect statistical power. For $X_2$ and $X_1 \times X_2$, they represent Type I error rates.
+:::
+
+### Conclusion
+Heteroscedasticity alone cannot explain ART's failures under log-normal distributions.
+ART's alignment method is based on the assumption that the shape of the distribution (log-normal here) is identical across all factor levels.
+This is not the case in the above scenarios where distribution shapes change across the levels of $X_1$, causing the method to break down.
+
+## Binomial distributions {#binomial}
+We also evaluate a wider range of parameters for the binomial distribution.
+We focus on the lower range of probabilities $p$.
+However, we expect results to be identical for their symmetric probabilities $1-p$.
+Specifically, we test $p=.05$, $.1$, and $.2$, and for each, we consider $k=5$ and $10$ task repetitions (Bernoulli trials).
+
+**Main effects**.
+We present our results for the main effect in @fig-binom-main.
+We observe that ART's Type I error rates increase as the number of repetitions decreases and the probability of success approaches zero, reaching very high levels when $k=5$ and $p=.05$.
+This trend is consistent across designs.
+The other methods maintain low error rates.
+However, their error rates fall below nominal levels when the magnitude of the effect on $X_1$ grows beyond a certain threshold, indicating a loss of power in these cases.
+
+::: {#fig-binom-main}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Appendix_test_binom"
+distributions = c("binom-5-05", "binom-5-10", "binom-5-20", "binom-10-05", "binom-10-10", "binom-10-20")
+dnames = c("k=5, p=5%", "k=5, p=10%", "k=5, p=20%", "k=10, p=5%", "k=10, p=10%", "k=10, p=20%")
+
+df <- readData(prefix, n = 20, alpha = .05, effectType = 1, distributions)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 98)
+
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
+:::
+
+**Interaction effects**. @fig-binom-interaction1 shows similar patterns for the Type I error rate of the interaction effect in the presence of a single main effect.
+
+::: {#fig-binom-interaction1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 98)
+
+```
+Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
+:::
+
+**Interactions under parallel main effects**.
+When both main effects exceed a certain level (see @fig-binom-interaction2), all methods fail to control the Type I error rate.
+Whether this inflation is due to interpretation issues or to a lack of robustness in the methods themselves, these results once again highlight that testing interactions in such scenarios is highly problematic.
+
+::: {#fig-binom-interaction2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- readData(prefix, n = 20, alpha = .05, effectType = 0, distributions)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 103)
+
+```
+Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitudes $a_{1}$ and $a_{2}$ of the main effects ($n=20$).
+*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
+A different definition may lead to different results.*
+:::
+
+### Conclusion
+ART is extremely problematic under binomial distributions, raising Type I error at very high levels even when effects are null across all factors.
+We also observe that testing interactions in the presence of parallel main effects can be problematic for all other methods.
+
+## Ordinal data {#ordinal}
+Given the frequent use of ART with ordinal data, we evaluate our complete set of ordinal scales, based on both equidistant and flexible thresholds, with additional experimental designs.
+
+**Main effects**.
+@fig-ordinal-designs-main presents Type I error rates for the main effect.
+ART preserves error rates at nominal levels under the $2 \times 3$ between-subjects design and the $2 \times 3$ mixed design, as long as thresholds are equidistant.
+Under the two within-subject designs, it inflates error rates, especially when there are fewer ordinal levels with flexible thresholds.
+
+::: {#fig-ordinal-designs-main}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Appendix_test_ordinal"
+alpha = 0.05
+
+distributions <- c("likert5", "likert5B", "likert7", "likert7B", "likert11", "likert11B")
+dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
+
+df <- readData(prefix, n = 20, alpha, effectType = 1, distributions)
+df <- reshapeByDesign(df, dnames)
+
+plotlyErrorByDesign(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 62)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
+:::
+
+**Interaction effects**.
+We present Type I error rates for the interaction effect in the presence of a single main effect.
+These results lead to similar conclusions.
+Even in cases where ART keeps error rates close to nominal levels (e.g., under the between-subjects design with equidistance thresholds), the performance of PAR is constantly better.
+
+::: {#fig-ordinal-designs-interaction-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+plotlyErrorByDesign(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 62)
+```
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$) 
+:::
+
+**Interactions under parallel main effects**. Finally, we show results for scenarios where main effects increase in parallel.
+
+::: {#fig-ordinal-designs-interaction-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- readData(prefix, n = 20, alpha, effectType = 0, distributions)
+df <- reshapeByDesign(df, dnames)
+plotlyErrorByDesign(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
+```
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$).
+*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
+A different definition may lead to different results.*
+:::
+
+### Conclusion
+ART's inflation of Type I error rates with ordinal data is confirmed across a range of designs.
+For the between-subjects and mixed designs, the problem primarily concerns ordinal scales with flexible thresholds.
+For the within-subjects designs, ART also inflates error rates for scales with equidistant thresholds, particularly when the number of levels is as low as five or seven.
+Again, all methods may fail to correctly infer interactions when parallel main effects exceed a certain threshold.
+
+## Main effects in the presence of interactions {#removable-main}
+In all experiments assessing Type I error rates reported in our article, we assumed no interaction effects.
+However, we also need to understand whether weak or strong interaction effects could affect the sensitivity of the methods in detecting main effects.
+Unfortunately, the interpretation of main effects may become ambiguous when interactions are present.
+
+### Illustrative example 
+To understand the problem, let us take a [dataset](removable_main.csv) from a fictional experiment (within-participants design with $n = 24$) that evaluates the performance of two techniques (*Tech A* and *Tech B*) under two task difficulty levels (*easy* vs.
+*hard*).
+@fig-example-interactions visualizes the means for each combination of the levels of the factors, using the original scale of measurements (left) and a logarithmic scale (right).
+Note that time measurements were drawn from log-normal distributions.
+
+::: {#fig-example-interactions}
+```{r, echo=FALSE, message=FALSE, fig.height=3.5, fig.width = 9, warning=FALSE}
+df <- read.csv("removable_main.csv", sep=",", header=TRUE, strip.white=TRUE)
+dftime <- aggregate(Time ~ Difficulty+Technique, data = df, mean)
+
+fig1 <- createInteractionPlot(dftime) %>% layout( yaxis = list(hoverformat = '.2f'))
+fig2 <- createInteractionPlot(dftime, logscale = TRUE) %>% layout( yaxis = list(hoverformat = '.2f'))
+
+fig <- subplot(fig1, fig2, titleY = TRUE, titleX = TRUE, margin = 0.06) %>% layout(hovermode = 'x', dragmode='pan')
+
+fig
+```
+The line charts visualize the effects of task difficulty (*easy* vs. *hard*) and technique (*Tech* A vs. *Tech B*) for task completion time. All data points represent group means. 
+:::
+
+We observe a clear main effect of *Difficulty* and a strong interaction effect, regardless of the scale.
+However, the main effect of *Technique* largely depends on the scale used to present the results.
+Can we then conclude that *Technique A* is overall faster than *Technique B*?
+
+Below, we present the results of the analysis using different methods.
+
+|        | PAR  | LOG  | ART | RNK | INT |
+|--------|------|------|-----|-----|-----|
+| Difficulty  | $7.8 \times 10^{-8}$ | $4.3 \times 10^{-16}$  |  $8.6 \times 10^{-12}$ | $2.5 \times 10^{-16}$ | $1.1 \times 10^{-15}$ |
+| Technique   | $.024$ | $.81$  | $.00017$ | $.69$ | $.69$ |
+| Difficulty $\times$ Technique | $7.6 \times 10^{-6}$ | $4.1 \times 10^{-9}$ | $9.4 \times 10^{-8}$ | $1.1 \times 10^{-9}$ | $1.3 \times 10^{-9}$ |
+: *p*-values for main and interaction effects {.sm}
+
+PAR detects a main effect of *Technique* ($\alpha = .05$), as does ART, which yields an even smaller $p$-value.
+In contrast, LOG, RNK, and INT do not find no evidence for such an effect.
+As with *removable interactions*, we can speak about *removable main effects* in this case.
+
+As a general principle, it makes little sense to interpret main effects when crossing patterns emerge due to a strong interaction.
+In this scenario, the researchers should instead compare techniques within each level of *Difficulty*.
+Accordingly, we can conclude that *Technique A* is slower on *easy* tasks but faster on *hard* tasks in this experiment.
+
+### Experiment
+To better understand how the different methods detect main in the presence of interactions, we conduct an experiment.
+
+We focus again on two-factor experimental designs, setting the sample size to $n = 20$.
+To simulate populations in which interactions emerge in the absence of main effects, we examine perfectly symmetric cross-interactions.
+To this end, we slightly change the method we use to encode the levels of each factor, such that levels are uniformly positioned around 0.
+For a factor with three levels, we numerically encode the levels as $\{-0.5, 0, 0.5\}$.
+For a factor with four levels, we encode them as $\{-0.5, -0.1667, 0.1667, 0.5\}$.
+
+Since the interpretation of main effects can be ambiguous, we define Type I errors with respect to the coefficients of the linear model.
+Accordingly, the null hypothesis is either $a_1 = 0$ or $a_2 = 0$, depending on the variable of interest.
+
+**Interaction effect only**.
+We first examine how the interaction effect alone influences the Type I error rate for $X_2$.
+The results are shown in @fig-interactions-main-1.
+We emphasize that interpretation issues arise only in the case of the $4 \times 3$ design.
+In the other two designs, $X_1$ has only two levels, and due to their symmetry, the observed main effects should be null --- regardless of how we define the null hypothesis.
+
+In the $4 \times 3$ design, error rates for PAR and ART grow rapidly as the interaction effect becomes large under the log-normal and exponential distributions.
+This is largely due to interpretation issues, as illustrated in our example.
+However, ART also fails to control the error rate in other scenarios.
+Finally, under the binomial and ordinal scales, we observe that RNK and INT may also inflate Type I error rates when the interaction effect becomes sufficiently large.
+
+::: {#fig-interactions-main-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+alpha <- .05
+
+prefix <- "Appendix_test_growing-interactions"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+
+df <- readData(prefix, n = 20, alpha, effectType = 0, distributions)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX1", xvar = "effectX1X2", max = 104)
+
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$ ($n=20$).
+*Note: Type I errors are defined based on the null hypothesis $a_{1} = 0$.
+For the $4 \times 3$ design, a different definition may lead to different results.
+However, the two other designs, where $X_1$ has only two levels, Type I error rates are the same, regardless of the definition we use.*
+:::
+
+We also examine error rates for the main effect of $X_2$. Now, interpretation issues affect the results across all three designs, leading to similar patterns.
+
+::: {#fig-interactions-main-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX2", xvar = "effectX1X2", max = 104)
+
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$ ($n=20$).
+*Note: Type I errors are defined based on the null hypothesis $a_{2} = 0$.
+A different definition may lead to different results.*
+:::
+
+**Interaction effect combined with main effect**.
+Finally, we evaluate the Type I error rate on $X_2$ when the interaction effect is combined with a main effect on $X_1$.
+@fig-interactions-main-3 presents our results.
+
+The error rates of ART and PAR now increase dramatically across all non-normal distributions and all three designs --- a result that is, in part, due to interpretation issues.
+RNK and INT are also affected: their error rates become extremely low under continuous distributions, suggesting a lack of power to detect small main effects when strong effects of other factors are combined with strong interactions.
+In contrast, under the binomial and ordinal scales, both methods exhibit inflated Type I error rates.
+Interestingly, RNK demonstrates the best overall performance in these tests.
+
+::: {#fig-interactions-main-3}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- readData(prefix, n = 20, alpha, effectType = 6, distributions)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyErrorByDesign2(df, xlab = "magnitude of effects", var = "rateX2", xvar = "effectX1X2", max = 104)
+
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitudes $a_1 = a_{12}$ of a combined main effect on $X_1$ and an interaction effect $X_1 \times X_2$ ($n=20$).
+*Note: Type I errors are defined based on the null hypothesis $a_{2} = 0$.
+A different definition may lead to different results.*
+:::
+
+### Conclusion
+When strong interactions are present, the interpretation of main effects can become ambiguous, especially when the interaction is combined with a non-zero main effect.
+ART is more sensitive to these issues than other methods and may even inflate Type I error rates in scenarios where interpretation problems are not expected.
+More broadly, main effects should be interpreted with caution in the presence of strong interactions.
+
+## ART with median alignment {#median-alignment}
+We evaluate a modified implementation of ART (ART-MED), where we use medians instead of means to align ranks.
+This approach draws inspiration from results by @Salter:1993, showing that median alignment corrects ART's instable behavior under the Cauchy distribution.
+We only test the $4 \times 3$ within-participants design for sample sizes $n=10$, $20$, and $30$.
+For this experiment, we omit the RNK method and only present results for non-normal distributions.
+
+We emphasize that @Salter:1993 only apply mean and median alignment to interactions.
+Our implementation for main effects is based on the alignment approach of @wobbrock:2011, where we simply replace means by medians --- we are not aware of more adapted methods.
+
+**Main effects**.
+Our results presented in @fig-median-main demonstrate that median alignment (ART-MED) --- or at least our implementation of the method --- is not appropriate for testing main effects.
+Although Type I error rates are now lower for the Cauchy distribution compared to the original method, they are still above nominal levels.
+In addition, they are significantly higher for all other distributions.
+
+::: {#fig-median-main}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Appendix_test_4x3_ART-Median"
+alpha <- .05
+
+methods = c("PAR", "INT", "ART", "ART-MED")
+palette = c("#888888", "#009E73", "#FF5E00", "#FFA27F")
+symbols = c("asterisk",  "star-diamond", "star-triangle-up", "star-triangle-down")
+
+distributions <- c("lnorm", "exp", "cauchy", "poisson", "binom", "likert5B")
+dnames <- c("Log-normal", "Exponential", "Cauchy", "Poisson", "Binomial", "Ordinal (5 levels)")
+df <- readlyData(prefix, alpha, 1, distributions, dnames, methods = methods)
+
+plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 64, cbPalette = palette, symbols = symbols)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
+:::
+
+**Interaction effects**.
+In contrast, median alignment works surprisingly well for interactions, correcting deficiencies of ART, especially when main effects are absent or weak.
+Despite this improved performance, we cannot recommend using the method because it still cannot compete with INT.
+Additionally, its advantages over parametric ANOVA are only apparent for the Cauchy distribution.
+
+::: {#fig-median-interaction-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+
+plotlyError(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 64, cbPalette = palette, symbols = symbols)
+```
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
+:::
+
+**Interactions under parallel main effects**.
+We also evaluate Type I error rates for interactions in the presence of parallel main effects, where we caution readers about intepretation issues.
+We observe again that median alignment improves ART's bevahior across all distributions.
+However, as with PAR, its definition of the null hypothesis aligns with the scale of responses rather than with the model parameters.
+
+::: {#fig-median-interaction-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- readlyData(prefix, alpha, 0, distributions, dnames, methods = methods)
+plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105, cbPalette = palette, symbols = symbols)
+
+```
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$.
+*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
+A different definition may lead to different results.*
+:::
+
+### Conclusion
+Using median instead of mean alignment in ART significantly improves the method's performance for testing interactions across all the distributions we examined.
+However, we cannot recommend it, as ART remains less robust than INT, and its advantages over PAR are unclear.
+Moreover, it is not evident how to apply median alignment for testing main effects --- using medians with the alignment procedure of @wobbrock:2011 leads to extremely high error rates.
+
+## Nonparametric tests in single-factor designs {#nonparametric-tests}
+We compare PAR, RNK, and INT to nonparamatric tests for within- and between-subjects single-factor designs, where the factor has two, three, or four levels.
+Depending on the design, we use different nonparametric tests.
+For within-subjects designs, we use the Wilcoxon sign-rank test if the factor has two levels (*2 within*) and the Friedman test if the factor has three (*3 within*) or four (*4 within*) levels.
+For between-subjects designs (*2 between*, *3 between*, and *4 between*), we use the Kruskal–Wallis test.
+
+**Power**.
+@fig-nonparametric-power compares the power of the various methods as the magnitude of the main effect increases, where we use the abbreviation *NON* to designate a nonparametric test.
+We observe that primarily INT, but also RNK, generally exhibit better power than the nonparametric tests.
+Differences are more pronounced for within-subjects designs, corroborating Conover's [-@Conover:2012] observation that the rank transformation results in a test that is superior to the Friedman test under certain conditions.
+
+::: {#fig-nonparametric-power}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+source("plotlying-v3.R")
+
+readSingle <- function(prefix, n, alpha, distributions, methods, POWER = FALSE) {
+  	df <- read.csv(paste("data/", prefix, "_", n, ".csv" , sep=""), sep=",", header=TRUE, strip.white=TRUE)
+	  df$distr <- factor(df$distr, levels=distributions)
+	  df$method <- factor(df$method, levels=methods)
+
+    if(POWER) {
+      df <- df[df$effect > 0,]
+    }
+    else {
+      df <- df[df$effect == 0,]
+    }
+
+    if(is.na(alpha)) return(df)
+    else return(df[df$alpha == alpha,])
+}
+
+prefix <- "Appendix_test-One-Factor"
+alpha = 0.05
+
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+methods = c("PAR", "RNK", "INT", "NON")
+palette = c("#888888", "#E69F00", "#009E73", "#FF70AB")
+
+df <- readSingle(prefix, n = 20, alpha, distributions, methods = methods, POWER = TRUE)
+#df <- reshapeByDesign(df, dnames, effectvars = c("effect"))
+df <- df %>% arrange(design,distr,effect,rates)  %>% group_by(design,distr,effect) %>% mutate(rank = rank(rates))
+
+df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effect"))
+plotlyPowerByDesign_v3(df, xlab = "magnitude of effect", var = "rank", xvar = "effect", max = 4.3, cbPalette = palette)
+```
+Ranking of methods comparing their power ($\alpha = .05$) as a function of the magnitude of effect ($n = 20$). 
+:::
+
+We expect that the accuracy of ANOVA on rank-transformed values will decrease with smaller samples.
+However, our tests with smaller samples of $n=10$ show that INT remains robust and still outperforms other nonparametric methods.
+Although it is possibe to couple INT with permutation testing for higher accuracy [@Beasley:2009], we have not explored this possibility here.
+
+**Type I error rate under equal and unequal variances**.
+@fig-nonparametric-unequal-var presents the rate of positives under conditions of equal ($r_{sd} = 0$) and unequal variances ($r_{sd} > 0$).
+While this rate can be considered a Type I error rate when variances are equal, interpreting it under other conditions requires special attention because the hypothesis of interest may differ among methods.
+Parametric ANOVA is particularly sensitive to unequal variances when distributions are skewed because it tests differences among means.
+While the normal distributions of the latent space have the same means, this is not the case with the skewed distributions of the transformed variable, which have the same median but different means.
+All nonparametric methods we tested use ranks, which preserve medians and mitigate this problem.
+However, their rate of positives can still exceed $5\%$ under certain conditions.
+
+For between-subjects designs, we observe that the Kruskal–Wallis test and RNK yield very similar results.
+This is not surprising, as RNK is known to be a good approximation of the Kruskal–Wallis test [@Conover:2012].
+INT's positive rates are similar, although slightly higher under the binomial distribution.
+For within-subjects designs, differences among methods are more pronounced.
+The Wilcoxon sign-rank test (*2 within*) inflates rates well above $5\%$, demonstrating that the test is not a pure test of medians.
+In contrast, the Friedman test (*3 within* and *4 within*) provides the best control among all methods.
+
+::: {#fig-nonparametric-unequal-var}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Appendix_test-One-Factor-Unequal-Var"
+
+df <- readSingle(prefix, n = 20, alpha, distributions, methods = methods)
+df <- reshapeByDesign(df, dnames, effectvars = c("sd_ratio", "effect"))
+
+plotlyErrorByDesign_v3(df, xlab = "max ratio of standard deviations", var = "rates", xvar = "sd_ratio", ytitle = 'Positives (%)', max = 24.2, nticks=8, cbPalette = palette)
+```
+Positives rates ($\alpha = .05$) as a function of the maximum ratio $r_{sd}$ of standard deviations between levels ($n = 20$). 
+:::
+
+@fig-nonparametric-unequal-var-alpha01 presents the same results but for $\alpha = .01$.
+Discrepancies among different methods are now more pronounced, and we notice again that the Friedman test keeps rates closer to the nominal level of $1\%$ compared to INT and RNK.
+Nevertheless, in addition to their greater power compared to the Friedman test, RNK or INT present other advantages, such as the possibility to use common ANOVA-based procedures to partly correct issues associated with unequal variances.
+For instance, we ran an experiment, where we applied a Greenhouse–Geisser correction following sphericity violations detected with the Mauchly's sphericity test ($\alpha = .05$).
+We found that this correction brings the error rates of INT close to nominal levels for continuous distributions, such as the normal and log-normal distributions.
+In the case of the binomial and ordinal distributions, error rates are also significantly reduced, well below those of the Friedman test, although not reaching nominal levels.
+
+::: {#fig-nonparametric-unequal-var-alpha01}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- readSingle(prefix, n = 20, alpha= 0.01, distributions, methods = methods)
+df <- reshapeByDesign(df, dnames, effectvars = c("sd_ratio", "effect"))
+
+plotlyErrorByDesign_v3(df, xlab = "max ratio of standard deviations", var = "rates", xvar = "sd_ratio", ytitle = 'Positives (%)', max = 7.2, nticks=8, cbPalette = palette)
+```
+Positives rates ($\alpha = .01$) as a function of the maximum ratio $r_{sd}$ of standard deviations between levels ($n = 20$). 
+:::
+
+### Conclusion
+We do not see significant benefits in using dedicated nonparametric tests over RNK or INT.
+INT can replace nonparametric tests even for single-factor designs.
+If, after transforming the data, the assumptions of homoscedasticity or sphericity are still not met, applying common correction procedures (e.g., a Greenhouse–Geisser correction for sphericity violations) on the transformed data can reduce the risk of Type I errors.
+
+## ANOVA-type statistic (ATS) {#ATS}
+We compare PAR, RNK, and INT to the ANOVA-type statistic (ATS) [@Brunner_ATS:2001] for two-factor designs.
+We use its implementation in the R package *nparLD* [@nparLD], which does not support between-subjects designs.
+Thus, we only evaluate it for the $4 \times 3$ within-subjects and the $2 \times 4$ mixed designs.
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+source("plotlying-v2.R")
+```
+
+**Type I error rates: Main effects**.
+@fig-ATS-designs-main presents Type I error rates for the main effect of $X_2$.
+Under the mixed design, RNK, INT, and ATS exhibit very similar error rates, which are close to nominal levels.
+In the within-subjects design, the error rates of ATS tend to be slightly above $5\%$.
+Additionally, unlike the other methods whose error rates drop significantly below $5\%$ when the effect of $X_1$ becomes stronger under binomial and ordinal scales, the power of ATS does not seem to be affected in these cases.
+
+::: {#fig-ATS-designs-main}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Appendix_test-ATS"
+alpha = 0.05
+
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+methods = c("PAR", "RNK", "INT", "ATS")
+palette = c("#888888", "#E69F00", "#009E73", "#FF70AB")
+
+df <- readData(prefix, n = 20, alpha, effectType = 1, distributions, methods = methods)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+plotlyErrorByDesign_v2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 7.1, nticks=8, cbPalette = palette)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
+:::
+
+@fig-ATS-designs-main-2 presents results for the main effect of $X_1$.
+The error rates of ATS are now slightly inflated under the mixed design.
+The other methods exhibit the same trends as for the other factor.
+
+::: {#fig-ATS-designs-main-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df2 <- readData(prefix, n = 20, alpha, effectType = 2, distributions, methods = methods)
+df2 <- reshapeByDesign(df2, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+plotlyErrorByDesign_v2(df2, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX2", max = 7.1, nticks=8, cbPalette = palette)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as a function of the magnitude $a_2$ of the main effect of $X_2$ ($n = 20$)
+:::
+
+**Type I error rates: Interactions**.
+@fig-ATS-designs-interaction-1 presents Type I error rates for the interaction in the presence of a single main effect.
+Results are again very similar for all three nonparametric methods under the mixed design.
+In contrast, the error rates of ATS tend to be lower than nominal levels under the within-subjects design, often falling below $4\%$.
+
+::: {#fig-ATS-designs-interaction-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+plotlyErrorByDesign_v2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 7.1, nticks=8, cbPalette = palette)
+```
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$) 
+:::
+
+ When two parallel main effects are present, ATS and RNK lead to very similar trends (see @fig-ATS-designs-interaction-2).
+Overall, INT appears to be a more robust method with the exception of the binomial distribution, for which error rates are higher for this method.
+
+::: {#fig-ATS-designs-interaction-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+
+df <- readData(prefix, n = 20, alpha, effectType = 0, distributions, methods = methods)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyErrorByDesign_v2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 104, nticks=8, cbPalette = palette)
+```
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$)
+:::
+
+**Power: Main effects**.
+As shown in @fig-ATS-power-main-1 and @fig-ATS-power-main-2, ATS appears as the most powerful method for detecting effects of $X_1$ for the mixed design.
+However, in all other situations, it has less power than INT and offer less power than RNK.
+
+::: {#fig-ATS-power-main-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+alpha <- .05
+
+prefix <- "Appendix_test-ATS-Power"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+
+df <- readData(prefix, n = 20, alpha, effectType = 3, distributions, methods = methods)
+
+df <- df %>% arrange(design,distr,effectX1,rateX1)  %>% group_by(design,distr,effectX1) %>% mutate(rank = rank(rateX1))
+df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyPowerByDesign_v2(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
+```
+Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_1$** as a function of the magnitude of effect $a_1$ ($n=20$). Hover over the charts to inspect their power ($\%$).
+:::
+
+::: {#fig-ATS-power-main-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- readData(prefix, n = 20, alpha, effectType = 4, distributions, methods = methods)
+
+df <- df %>% arrange(design,distr,effectX2,rateX2)  %>% group_by(design,distr,effectX2) %>% mutate(rank = rank(rateX2))
+df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyPowerByDesign_v2(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
+```
+Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_2$** as a function of the magnitude of effect $a_2$ ($n=20$). Hover over the charts to inspect their power ($\%$).
+:::
+
+**Power: Interactions**.
+@fig-ATS-power-interaction shows results on power for interactions.
+INT emerges again as the most powerful method.
+The power of ATS is particularly low under the within-subjects design.
+
+::: {#fig-ATS-power-interaction}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- readData(prefix, n = 20, alpha, effectType = 5, distributions, methods = methods)
+
+df <- df %>% arrange(design,distr,effectX1X2,rateX1X2)  %>% group_by(design,distr,effectX1X2) %>% mutate(rank = rank(rateX1X2))
+df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyPowerByDesign_v2(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
+```
+Ranking of methods for their power ($\alpha = .05$) to detect the **interaction effect $X_1 \times X_2$** as a function of the magnitude of effect $a_{12}$ ($n=20$).
+Hover over the charts to inspect their power ($\%$).
+:::
+
+### Conclusion
+Although ATS appears to be a valid alternative, it does not offer clear performance advantages over INT, which is also simpler and more versatile.
+
+## Generalizations of nonparametric tests {#generalized}
+
+Finally, we evaluate the generalizations of nonparametric tests recommended by Lüpsen [-@luepsen:2018; -@luepsen:2023] as implemented in his *np.anova* function [@luepsen_R].
+Specifically, we evaluate the generalization of the van der Waerden test (VDW), and the generalization of the Kruskal-Wallis and Friedman test (KWF).
+Their implementation uses R's *aov* function and requires considering random slopes in the error term of the model, that is, using ``Error(s/(x1*x2))`` for the two-factor within-participants design and ``Error(s/x2)`` for the mixed design, where ``s`` is the subject identifier variable.
+We also used the *aov* function and the same formulation of the error term for all other methods.
+
+**Type I error rates: Main effects**.
+@fig-vdWaerden-designs-main presents Type I error rates for the main effect of $X_2$.
+While all methods perform well under the within-subjects and mixed designs, the error rates of VDW and KWF drop drastically as the effect of $X_1$ grows under the between-subjects design.
+We will see below that the power of these methods evaporates in these cases.
+
+::: {#fig-vdWaerden-designs-main}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Appendix_test-vdWaerden"
+alpha = 0.05
+
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+methods = c("RNK", "INT", "VDW", "KWF")
+palette = c("#E69F00", "#009E73", "#FF70AB", "#888888")
+
+df <- readData(prefix, n = 20, alpha, effectType = 1, distributions, methods = methods)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 7.1, nticks=8, cbPalette = palette)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
+:::
+
+**Type I error rates: Interactions**.
+@fig-vdWaerden-designs-interaction-1 presents the Type I error rates for the interaction, in the presence of a single main effect.
+Once again, the error rates of VDW and KWF decrease rapidly in both the between-subjects and mixed designs.
+@fig-vdWaerden-designs-interaction-2 shows the results when both main effects are present.
+Under the within-subjects design, KWF is the poorest-performing technique.
+While VDW outperforms RNK, it remains inferior to INT.
+For the between-subjects and mixed designs, KFW and VDW yield low or near-zero error rates when main effects are large, likely due to their extreme loss of power in these scenarios.
+
+::: {#fig-vdWaerden-designs-interaction-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- readData(prefix, n = 20, alpha, effectType = 2, distributions, methods = methods)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX2", max = 7.1, nticks=8, cbPalette = palette)
+```
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_2$ of the main effect of $X_2$ ($n = 20$) 
+:::
+
+::: {#fig-vdWaerden-designs-interaction-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+
+df <- readData(prefix, n = 20, alpha, effectType = 0, distributions, methods = methods)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 104, nticks=8, cbPalette = palette)
+```
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ ($n = 20$)
+:::
+
+**Power: Main effects**.
+@fig-vdWaerden-power-main-1 presents power results for detecting the main effect of $X_1$.
+For the within-subjects and mixed design, KFW and VDW exhibit a lower power than RNK and INT, but differences are generally small.
+Although VDW appears as powerful as INT under the between-subjects design, this only occurs when the effect of the second factor is zero.
+As shown in @fig-vdWaerden-power-main-1B, the power of both KFD and VDW drops drastically as the effect of $X_2$ increases for this design.
+
+::: {#fig-vdWaerden-power-main-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+alpha <- .05
+
+prefix <- "Appendix_test-vdWaerden-Power"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+
+df <- readData(prefix, n = 20, alpha, effectType = 3, distributions, methods = methods)
+
+df <- df %>% arrange(design,distr,effectX1,rateX1)  %>% group_by(design,distr,effectX1) %>% mutate(rank = rank(rateX1))
+df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyPowerByDesign(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
+```
+Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_1$** as a function of the magnitude of effect $a_1$ ($n=20$). Hover over the charts to inspect their power ($\%$).
+:::
+
+::: {#fig-vdWaerden-power-main-1B}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+alpha <- .05
+prefix <- "Appendix_test-vdWaerden-Power-Multieffect"
+
+df <- readData(prefix, n = 20, alpha, effectType = 3, distributions, methods = methods)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX2", ytitle ="Power (%)", max = 104, nticks=6, cbPalette = palette)
+```
+Power ($\alpha = .05$) to detect the **main effect of $X_1$** ($a_1 = 0.8$) as a function of the magnitude of effect $a_2$ of the main effect of $X_2$ ($n=20$).
+:::
+
+@fig-vdWaerden-power-main-2 also presents results for the main effect of $X_1$, where we observe once again that the generalized tests cannot compete with the more powerful INT, or even with RNK.
+
+::: {#fig-vdWaerden-power-main-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Appendix_test-vdWaerden-Power"
+df <- readData(prefix, n = 20, alpha, effectType = 4, distributions, methods = methods)
+
+df <- df %>% arrange(design,distr,effectX2,rateX2)  %>% group_by(design,distr,effectX2) %>% mutate(rank = rank(rateX2))
+df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyPowerByDesign(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
+```
+Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_2$** as a function of the magnitude of effect $a_2$ ($n=20$). Hover over the charts to inspect their power ($\%$).
+:::
+
+**Power: Interactions**. We also present results on the power of methods to detect interactions in @fig-vdWaerden-power-interaction, confirming the advantage of INT across all designs.
+@fig-vdWaerden-power-interactionB provides a clearer picture of how power is affected by the presence of a main effect.
+We observe that the power of all methods drops as the main effect of $X_2$ increases, but this trend is more pronounced for KFD and VDW, particularly for the between-subjects and mixed designs.
+
+::: {#fig-vdWaerden-power-interaction}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- readData(prefix, n = 20, alpha, effectType = 5, distributions, methods = methods)
+
+df <- df %>% arrange(design,distr,effectX1X2,rateX1X2)  %>% group_by(design,distr,effectX1X2) %>% mutate(rank = rank(rateX1X2))
+df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotlyPowerByDesign(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
+```
+Ranking of methods for their power ($\alpha = .05$) to detect the **interaction effect $X_1 \times X_2$** as a function of the magnitude of effect $a_{12}$ ($n=20$).
+Hover over the charts to inspect their power ($\%$).
+:::
+
+::: {#fig-vdWaerden-power-interactionB}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+alpha <- .05
+prefix <- "Appendix_test-vdWaerden-Power-Multieffect"
+
+df <- readData(prefix, n = 20, alpha, effectType = 5, distributions, methods = methods)
+df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX2", ytitle ="Power (%)", max = 104, nticks=6, cbPalette = palette)
+```
+Power ($\alpha = .05$) to detect the **interaction effect** ($a_{12} = 1.5$) as a function of the magnitude of effect $a_2$ of the main effect of $X_2$ ($n=20$).
+:::
+
+### Conclusion
+Our results do not support Lüpsen's [-@luepsen:2018; -@luepsen:2023] conclusions.
+The behavior of the generalized nonparametric tests presents issues in numerous scenarios.
+While these tests exhibit lower error rates under specific conditions, this is due to a significant loss of power when other effects are at play.
+Therefore, we advise against the use of these methods.
 

### 2025-12-18 appendix_II.qmd
methodology..

---
title: "The illusory promise of the Aligned Rank Transform"
subtitle: "Appendix II. Analytical derivations and mathematical proofs"
author: 
  - name: Theophanis Tsandilas
    orcid: 0000-0002-0158-228X
    email: theophanis.tsandilas@inria.fr
    affiliations:
      - name: Université Paris-Saclay, CNRS, Inria, LISN
        country: France
  - name: Géry Casiez
    orcid: 0000-0003-1905-815X
    email: gery.casiezuniv-lille.fr
    affiliations:
      - name: Univ. Lille, CNRS, Inria, Centrale Lille, UMR 9189 CRIStAL
        country: France
bibliography: bibliography.bib
csl: canadian-journal-of-philosophy.csl

tbl-cap-location: top
---

## Absence of spurious effects when $a_2=0$ and $a_{12} = 0${#effects-proof}
Let $x_1$ and $x_2$ be two independent variables, and suppose the conditional distribution of the response $Y | (x_1, x_2)$ depends on these variables only through the linear predictor:
$$
\eta(x_1, x_2) = a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2
$$

Assume that the conditional mean can be written as:
$$
\mu(x_1, x_2) = E(Y|x_1, x_2) = h(\eta(x_1, x_2))
$$
where $h$ is a monotone function. This function may coincide with the inverse link function $g^{-1}$ in a GLM, but it may also arise from other frameworks, including transformation models or latent-variable constructions. Thus it does not necessarily need to be smooth or invertible; for example, this assumption is not required for the ordinal example discussed later.

Suppose $a_2 = 0$ and $a_{12} = 0$. Then:

1. The independent variable $x_2$ has no effect on the response scale.
2. There is no interaction between $x_1$ and $x_2$ on the response scale.

Thus, whether Type I errors are defined in terms of the model coefficients $a_2$ (when $a_{12} = 0$) or $a_{12}$ (when $a_2 = 0$) or in terms of the patterns of observed means (and alternatively medians or other quantiles) on the response scale makes no difference.

### Proof
Under $a_2 = a_{12} = 0$, the linear predictor reduces to

$$\eta(x_1, x_2) = a_0 + a_1x_1$$

which does not depend on $x_2$. Hence, the conditional distribution $Y | (x_1, x_2)$, including its mean, median, and all quantiles, is invariant to $x_2$. This establishes the absence of a main effect of $x_2$ or an effect of its interaction $x_2$.

We further explain this results for special cases, where we largely rely on the relevant discussions of @Ai:2003, @Greene:2010, and @McCabe:2022.

#### Main effect of $x_2$
If $x_2$ is continuous and $h$ is differentiable, the effect of $x_2$ on the response can be defined as:

$$\frac{\partial \mu(x_1, x_2)}{\partial x_2}  = h'(\eta) \frac{\partial \eta}{\partial x_2} = 0$$
where $h'$ denotes the first derivative of $h$.

If $x_2$ is categorical, all contrasts across its levels are zero because the mean response is identical at each level:
$$\Delta_{u,\nu}\mu(x_1, x_2) = \mu(x_1, u) - \mu(x_1, \nu) = h(a_0 + a_1 x_1) - h(a_0 + a_1 x_1) = 0$$
where $u, \nu$ are any two levels of $x_2$.

#### Interaction between $x_1$ and $x_2$
An interaction on the response scale is present if the marginal effect of $x_1$ depends on $x_2$. For continuous predictors and assuming that $h$ is twice differentiable, this is characterized by the mixed partial derivative $\frac{\partial^2 \mu}{\partial x_2 \partial x_1}$.

Using the chain rule:

$$\frac{\partial^2 \mu(x_1, x_2)}{\partial x_2 \partial x_1} = h''(\eta) \frac{\partial \eta}{\partial x_2} \frac{\partial \eta}{\partial x_1} + h'(\eta)\frac{\partial^2\eta}{\partial x_2 \partial x_1} = 0$$

Thus the marginal effect of $x_1$ does not vary with $x_2$, and there is no interaction.

For categorical predictors, the analogue of a mixed partial derivative is the double difference:

$$\Delta_{u_1, \nu_1} \Delta_{u_2, \nu_2}\mu(x_1, x_2) = [\mu(u_1,u_2) - \mu(u_1, \nu_2)] - [\mu(\nu_1,u_2) - \mu(\nu_1, \nu_2)]=$$ 
$$= [h(a_0 + a_1 u_1) - h(a_0 + a_1 u_1)] - [h(a_0 + a_1 \nu_2) - h(a_0 + a_1 \nu_1)] = 0$$

for any levels $u_1, \nu_1$ of $x_1$ and $u_2, \nu_2$ of $x_2$. Again, there is no observed interaction.

### Illustrative examples
We discuss two examples.

#### Lognormal data
Let $x_1$ and $x_2$ be two experimental factors. Suppose the true model of the data is
$$
\begin{aligned}
log Y = a_0 + a_1 x_1 + \varepsilon, \qquad
\varepsilon \sim \mathcal{N}(0, \sigma^2)
\end{aligned} 
$$
so that 
$$Y | (x_1, x_2) \sim Lognormal(a_0 + a_1 x_1, \sigma^2)$$

Since the conditional distribution of $Y$ does not depend on $x_2$, there is no main effect of $x_2$ on the mean, median, or any quantile of the response distribution. In particular, the conditional mean
$$\mu(x_1,x_2) = E(Y|x_1, x_2) = exp(a_0 + a_1 x_1 + \frac{1}{2}\sigma^2)$$
remains identical across all levels of $x_2$.

For the same reasons, there is no interaction on the response scale.

#### Ordinal responses
Suppose there exists a latent continuous variable:

$$Y^* = a_0 + a_1 x_1 + \varepsilon$$

and the observed response $Y$ is obtained by applying fixed thresholds: 
$$
\begin{aligned}
Y = k\ \ \text{if}\ \ \tau_{k-1} < Y^* \leq \tau_k, 
\qquad k = 1,2 ... K
\end{aligned} 
$$ 
Again, the conditional distribution $Y| (x_1, x_2)$ depends on the independent variables only through $\eta(x_1, x_2)=a_0+a_1 x_1$, and there is no effect of $x_2$ on either the latent or the observed ordinal scale.

In this case, the mapping from $\eta$ to the observed response is monotone but not strictly monotone and is therefore non-invertible. As a result, a true effect of $x_2$ on the latent variable (i.e., $a_2 \neq 0$) may be attenuated or entirely unobservable on the ordinal response scale. This does not affect Type I error rates, provided the null hypothesis is defined on the latent scale and the thresholds are fixed, because the absence of a latent effect implies the absence of an effect on the observed scale. However, power and effect sizes can be reduced when effects are assessed on the ordinal scale, reflecting the loss of information due to discretization.

### Invariance under monotone transformations 
Suppose we apply a monotone transformation $T$ directly to the observed responses---such as the rank transformation (RNK) or the inverse normal transformation (INT). Because $T$ is monotone, it preserves the ordering of observations within each condition. The transformed conditional distribution $T(Y) | (x_1, x_2)$ therefore depends on the independent variables only through $\eta(x_1, x_2) = a_0+a_1 x_1$. Consequently, the mean, median, and all quantiles of $T(Y) | (x_1, x_2)$ remain invariant to $x_2$, and marginal effects of $x_1$ do not depend on $x_2$.

We emphasize that ART is not a monotone transformation. Instead of transforming $Y$ directly, ART first constructs aligned responses by subtracting and adding sample means on the response scale. These alignment terms can vary across levels of $x_2$ or combinations of $(x_1, x_2)$. Therefore, ART is not invariant to nonlinear mean–response relationships.

<!--
In conclusion, $T$ does not introduce a main effect of $x_2$ or an interaction between $x_1$ and $x_2$ when $a_2=0$ and $a_{12}=0$.

### ART introducing spurious effects 
ART differs fundamentally from simple monotone transformations. Instead of transforming $Y$ directly, ART first constructs aligned responses by **subtracting and adding sample means on the response scale**, for example:

$$Y^{(B)} = Y - \bar{Y}_{ij} + \bar{Y}_{.j} - \bar{Y}_{..}$$
or analogous expressions for interactions, and only then applies ranking. The conditional distribution $Y^{(B)} | (x_1, x_2)$ now depends on $x_2$ through the arithmetic mean components even though $a_2 = 0$ and $a_{12}=0$.

These alignment terms can vary across levels of $x_2$ due to sampling variability alone, and depending on $h$, they can acquire systematic shifts across levels of $x_2$ or combinations of $(x_1, x_2)$, even when the true conditional distribution does not. in particular, the term $\bar{Y}_{.j} - \bar{Y}_{..}$ is constant within each level of $x_2$ and shifts all aligned observations for that level either upward or downward.

After ranking, observations from one level of $x_2$ tend to receive systematically higher or lower ranks. Our results confirm this problem.
-->

## References {.unnumbered}

::: {#refs}
:::

### 2025-12-18 sandbox.qmd
methodology..

@@
 ```
+
+## Experimental method {#methodology}
+We can now detail our experimental method.
+We evaluate the standard parametric approach (*PAR*) and the three rank-transformation methods (*RNK*, *INT*, and *ART*) that we introduced earlier.
+We conduct a series of Monte Carlo experiments that assess their performance under a variety of experimental configurations:
+
+**Distributions.** We evaluate both *ratio* and *ordinal* data, drawn from the following distributions:
+
+1. *Normal distribution* $\mathcal{N}(\mu, \sigma)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation. For most experiments where variances are equal, we set $\sigma = 1$.
+2.
+*Log-normal distribution* $\mathcal{LogN}(\mu, \sigma)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation at the logarithmic scale.
+The log-normal distribution is a good model for various measures bounded by zero, such as task-completion times.
+We set $\sigma = 1$ in our main experiments but evaluate a wider range of parameters in [Appendix I](appendix_I.html#lognormal).
+3.
+*Exponential distribution* $Exp(\lambda)$, where $\lambda$ is its rate.
+The exponential distribution naturally emerges when describing the time elapsed between events.
+For example, we could use it to model the time a random person spends with a public display, or the waiting time before a new person approaches to interact with the display, when the average waiting time is $\frac{1}{\lambda}$.
+4.
+*Cauchy distribution* $\mathcal{Cauchy}(x_0, \gamma)$, where $x_0$ defines its location (median) and $\gamma$ defines its scale.
+The Cauchy distribution is the distribution of the ratio of two independent normally distributed random variables.
+It rarely emerges in practice.
+However, it is commonly used in statistics to test the robustness of statistical procedures because both its mean and variance are undefined.
+As we discussed earlier, past evaluations of ART [@Mansouri:1995; @elkin:2021] show that the method fails under the Cauchy distribution.
+We set $\gamma = 1$.
+5.
+*Poisson distribution* $\mathcal{Pois}(\lambda)$, where $\lambda$ is its rate.
+It expresses the probability of a given number of events in a fixed interval of time.
+For example, we could use it to model the number of people who interact with a public display in an hour, when the average rate is $\lambda$ people per hour.
+6.
+*Binomial distribution* $\mathcal{B}(\kappa, p)$, where $\kappa$ defines the number of independent Bernoulli trials and $p$ is the probability of error or success of each trial.
+It frequently appears in HCI research, as it can model the number of successes and failures in a series of experimental tasks.
+We set $\kappa = 10$ in our main experiments but evaluate a wider range of parameters in [Appendix I](appendix_I.html#binomial).
+7. Distributions of *Likert-item* responses with 5, 7, and 11 levels.
+
+**Experimental designs.** We present results for five experimental designs.
+To simplify our presentation, we start with a 4 $\times$ 3 within-subjects factorial design.
+We then show how our conclusions generalize to four additional designs: (i) a 2 $\times$ 3 between-subjects design; (ii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a within-subjects factor; (iii) a 2 $\times$ 2 $\times$ 2 within-subjects design, and (iv) a 3 $\times$ 3 $\times$ 3 within-subjects design.
+
+**Sample sizes.** We focus on three sample sizes, $n=10$, $n=20$, and $n=30$, where $n$ represents the cell size in an experimental design.
+However, for certain scenarios, we also report results for larger sample sizes, up to $n = 512$.
+In within-subjects designs, where all factors are treated as repeated measures, $n$ corresponds to the number of subjects $N$ (commonly human participants in HCI research).
+In contrast, in a 2 $\times$ 3 between-subjects design, a cell size of $n = 20$, implies a total of $N = 120$ subjects.
+
+**Equal vs. unequal variances.** We test the robustness of the methods when the variances on the latent variable are unequal.
+
+**Evaluation measures.** In addition to Type I error rates, we compare the statistical power of the methods and assess the accuracy of their effect size estimates.
+
+**Magnitude of effects.** We analyze both main and interaction effects, examining how increasing the magnitude of effect of one factor influences the Type I error rates of other factors and their interactions.
+In addition, we examine how increasing two main effects in parallel influences the Type I error rate of their interaction, while emphasizing that the definition of the null hypothesis in this case is ambiguous.
+
+<!--
+Furthermore, we assess how a growing interaction effect, either in isolation or in combination with a main effect, can affect the inference of main effects. -->
+
+Previous evaluations of rank transformation methods [@Beasley:2009; @luepsen:2018] have also examined unbalanced designs, where cell sizes vary across the levels of a factor.
+When combined with unequal variances, such designs often pose challenges for both parametric procedures [@blanca:2018] and rank transformation methods [@Beasley:2009; @luepsen:2018].
+As noted earlier, we do not consider unbalanced designs in our main article.
+However, we provide additional experimental results on missing data in [Appendix I](appendix_I.html#missing).
+
+### Statistical modeling
+For simplicity, we explain here our modeling approach for two factors, but its extension to three factors is straightforward.
+
+**Linear component.** The base of our data generation process for all distributions is the following linear term:
+
+$$ 
+\eta_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j}
+$${#eq-linear-model}
+
+ - $\mu$ is the grand mean
+
+ - $s_k \sim \mathcal{N}(0, \sigma_s^2)$ is the random intercept effect of the *k*-th subject, where $k = 1..n$
+
+ - $x_{1i}$ is a numerical encoding of the *i*-th level of factor $X_1$, where $i = 1..m_1$
+
+ - $x_{2j}$ is a numerical encoding of the *j*-th level of factor $X_2$, where $j = 1..m_2$
+
+ - $a_1$, $a_2$, and $a_{12}$ express the magnitude of main and interaction effects
+
+<!--
+ - $\varepsilon_{ijk}$ is the experimental error effect
+ -->
+
+ While random slope effects can have an impact on real experimental data [@barr:2013], we do not consider them here for two main reasons: (1) to be consistent with previous evaluations of the ART procedure [@elkin:2021]; and (2) because mixed-effects procedures with random slope effects are computationally demanding, adding strain to simulation resources.
+However, there is no good reason to believe that adding random slope effects would impact our findings and conclusions.
+
+**Encoding the levels of factors.** To encode the levels of the two factors $x_{1i} \in X_1$ and $x_{2j} \in X_2$ we proceed as follows:
+
+1.
+We normalize the distance between their first and their second levels such that $x_{12} - x_{11} = 1$ and $x_{22} - x_{21} = 1$.
+This approach enables us to conveniently control for the main and interaction effects by simply varying the parameters $a_1$, $a_2$, and $a_{12}$.
+
+2.
+For the remaining levels, we randomly sample from a uniform distribution that spans the range between these two extreme levels, i.e., between $x_{11}$ and $x_{12}$ for $X_1$, and between $x_{21}$ and $x_{22}$ for $X_2$.
+This approach allows us to generate and evaluate a substantial variety of configurations, each representing different relative effects between levels.
+
+3. We require all levels to sum up to 0, or $\sum\limits_{i=1}^{m_1}  x_{1i} = 0$ and $\sum\limits_{j=1}^{m_2}  x_{2j} = 0$, which ensures that the grand mean is $\mu$.
+
+For example, we can encode a factor with four levels as $\{-.6, .4, .1, .1\}$ or as $\{-.5, .5, .3, -.3\}$.
+
+**Generating ratio data.** We adopt generalized linear modeling for ratio data, where we use an inverse link function $g^{-1}$ to make the connection between $\eta_{ijk}$ and the location parameter of the distribution of interest:
+$$
+Y_{ijk} \sim \mathcal{F}(location=g^{-1}(\eta_{ijk}), ...)
+$${#eq-glm-ratio}
+
+where $\mathcal{F}$ denotes the probability density (or mass) function from which we draw samples.
+For example, the mean of the exponential distribution is equal to $1 / \lambda$, where $\lambda$ is its rate parameter.
+This mean is linked to the linear term through the function $g^{-1}(x) = e^{x}$, and therefore, $\lambda = e^{-\eta_{ijk}}$.
+
+ @tbl-distributions presents how the parameters of the other distributions are linked to the linear term.
+
+| distribution | parameters | description               |
+|--------------|------------|---------------------------|
+| Normal       |$\mu = \eta_{ijk}$ <br> $\sigma^2 = \sigma_{\varepsilon}^2$ | mean <br> standard deviation |
+| Log-normal    |$\mu = \eta_{ijk}$ <br> $\sigma^2 = \sigma_{\varepsilon}^2$ | mean at logarithmic scale (logarithm of median) <br> standard deviation at logarithmic scale|
+| Exponential  |$\lambda = e^{-\eta_{ijk}}$ | rate (1 / mean) |
+| Cauchy       |$x_0 = \eta_{ijk}$ <br> $\gamma = 1$ | location (median) <br> scale 
+| Poisson      |$\lambda = e^{\eta_{ijk}}$ | rate (mean) |
+| Binomial     |$p = Logistic(\eta_{ijk})$ <br> $\kappa = 10$ | probability of error <br> number of Bernoulli trials |
+
+: Distribution parameters and their links to the linear term $\eta_{ijk}$ {#tbl-distributions}
+
+**Generating ordinal data.**
+
+### Population control and distribution conversions 
+To simplify our simulations, we fix the following population parameters: $\mu = 0$, $\sigma = 1$, and $\sigma_s = 1$.
+We then control the magnitude of effects by varying $a_1$, $a_2$, and, for some experiments $a_{12}$.
+@fig-effects presents the range of values that we test for $a_1$ and $a_2$.
+We also visualize their effects on the latent variable for factors with two, three, and four categorical levels.
+
+::: {#fig-effects}
+```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
+source("effects_plot.R")
+plotEffects()
+```
+Varying $a_1$ or $a_2$ to control the magnitude of main effects.
+The plots show examples of population distributions on the latent variable $\mathcal{Y}$ for factors ($X_1$ or $X_2$) with two (top), three (middle), or four categorical levels (bottom).
+:::
+
+We follow the approach of @DeBruine:2021 and use the R package *faux* v1.2.1 [@faux] to simulate data for our mixed-effects models.
+We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\varepsilon_{ijk} \sim N(0,\sigma)$.
+To simulate the observed variable $Y$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach).
+A key advantage of this method is that we can generate responses for any distribution, while we control effects on the latent variable in the exact same way.
+
+### Implementation of rank transformation methods
+For the aligned rank transformation (ART), we use the R implementation of ARTool v0.11.1 [@artool].
+For the pure rank transformation (RNK), we use R's *rank()* function.
+We use the *Rankit* formulation [@Bliss:1956] for the inverse normal transformation (INT), as explained earlier.
+Unless explicitly mentioned otherwise, we use the formula ``Y' ~ X1*X2 + (1|S)`` with R's *lmer* function, except for the between-subjects design where we use the formula ``Y' ~ X1*X2 + Error(S)`` with R's *aov* function.
+
+### Evaluation measures
+Significance tests have two types of errors.
+*Type I errors*, or false positives, are mistaken rejections of the null hypothesis.
+Type II errors, of false negatives, are failures to reject a null hypothesis that is actually true.
+In our illustrative example in @fig-example, a Type I error is finding that there is an effect of the choice of the technique on time performance.
+A *Type II error*, in turn, is finding that the task difficulty has no effect on time performance.
+
+Statistical significance testing requires setting a significance threshold known as significance or $\alpha$ (alpha) level, with typical values $\alpha = .05$ and $\alpha = .01$.
+The Type I error rate of a well-behaved significance test should be close to this nominal alpha level.
+An error rate clearly above this level suggests that the significance test is too liberal, while an error rate clearly below this level suggests that the test is too conservative.
+Four of our experiments specifically assess the Type I error rate of the methods.
+We test two significance levels: $\alpha = .05$ and $\alpha = .01$.
+For brevity, we only report results for $\alpha = .05$ in the main paper and include additional results in our supplementary material.
+
+We do not directly evaluate Type II errors.
+Instead, we report on statistical *power* defined as $Power = 1 - \beta$, where $\beta$ is the rate of Type II errors.
+Significance tests do not provide any power guarantees.
+However, we can compare the power of different methods to evaluate their relative performance.
+In addition to power, we assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted on the latent variable $\mathcal{Y}$.
+While partial $\eta^2$ is the most commonly used effect size measure, we also evaluate Cohen's $f$, as its expected value is proportional to the real magnitude of effect.
+However, $\eta^2$ can be directly derived from Cohen's $f$ as follows:
+
+$$ 
+\eta^2 = \frac{f^2}{1 + f^2}
+$${#eq-cohensf}
+
+As explained earlier, interaction effects are distorted when the latent variable $\mathcal{Y}$ is transformed to produce the observed responses.
+These transformations also influence the Type I error rates and statistical power we observe.
+In this analysis, we focus on evaluating these measures with respect to the effects applied to the latent variable $\mathcal{Y}$.
+We explicitly make this distinction whenever it is relevant to our discussion.
+
+### Hardware platform and iterations
+Our experimental R code is available in our supplementary material.
+We ran our experiments separately in a cluster of 8 machines Dell R640 Intel Xeon Silver 4112 2.6GHz with 4 cores and 64 GB memory.
+Our R code was parallelized to use all four cores of each machine.
+Some experiments took a few hours to complete, while others took several days.
+
+To estimate the power and Type I error rates of the four methods with enough precision, we ran $5000$ iterations for each population configuration and each sample size.

### 2025-12-18 text/20-background.qmd
methodology..

@@
 
-![Example showing the use of the rank, ART, and INT transformation methods (a). Construction of ART for a two-factor design (b,c)](images/art-explained.svg){#fig-ART-explained width=80%}
+![Example showing the use of the rank, ART, and INT transformation methods (a).
+Construction of ART for a two-factor design (b,c), where $Y_{ijk}$ represents the response of the $k^{th}$ subject for $A=a_i$ and $B=b_j$.](images/art-explained.svg){#fig-ART-explained width=80%}
 
@@
 With a focus on clarity, we chose to exclusively examine the three rank-based transformations presented earlier: the pure rank transformation (RNK), ART, and INT.
-While we do not elaborate on the performance of ATS in the main paper, additional experimental results are available in the [appendix](appendix.html#ATS).
+While we do not elaborate on the performance of ATS in the main paper, additional experimental results are available in [Appendix I](appendix_I.html#ATS).
 These findings indicate that its performance is comparable to the rank transformation but seems to be inferior to the simpler and more versatile INT.
-The [appendix](appendix.html#generalized) features additional results regarding the performance of the generalized van der Waerden test, as well as the generalized Kruskal-Wallis and Friedman tests.
+[Appendix I](appendix_I.html#generalized) features additional results regarding the performance of the generalized van der Waerden test, as well as the generalized Kruskal-Wallis and Friedman tests.
 Our findings show that, at least in balanced designs, these methods suffer from low power issues in certain situations without demonstrating any clear benefits compared to INT.
@@
 The rationale behind this decision is that we have not encountered any prior claims about ART's suitability for unbalanced data, and the ARTool [@artool] issues a warning in such situations.
-However, the [appendix](appendix.html#missing) presents additional results on experimental designs with missing data, which confirm that in such situations, ART's accuracy may deteriorate further.
+However, [Appendix I](appendix_I.html#missing) presents additional results on experimental designs with missing data, which confirm that in such situations, ART's accuracy may deteriorate further.

### 2025-12-18 text/30-assumptions.qmd
methodology..

@@
 $$ 
-Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \epsilon_{ijk}
+Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \varepsilon_{ijk}
 $${#eq-higgins-model}
 
-where $\mu$ is the grand mean, $\alpha_i$ is the main effect of the first factor for level $i$, $\beta_j$ is the main effect of the second factor for level $j$, $(\alpha \beta)_{ij}$ is the interaction effect, and $\epsilon_{ijk}$ is the residual error term, assumed to have a common variance.
+where $\mu$ is the grand mean, $\alpha_i$ is the main effect of the first factor for level $i$, $\beta_j$ is the main effect of the second factor for level $j$, $(\alpha \beta)_{ij}$ is the interaction effect, and $\varepsilon_{ijk}$ is the residual error term, assumed to have a common variance.
 The index $k = 1, ..., n$ denotes the subject number.
@@
 $$ 
-log(Y_{ijk} - \theta) = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \epsilon_{ijk}
+log(Y_{ijk} - \theta) = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \varepsilon_{ijk}
 $${#eq-generalized-model}
 
-where $\epsilon_{ijk}$ is normally distributed, and $\theta$ represents a threshold below which response values cannot occur.
+where $\varepsilon_{ijk}$ is normally distributed, and $\theta$ represents a threshold below which response values cannot occur.
 Unfortunately, ART's alignment mechanism is not designed to handle such cases.
@@
 The model we presented earlier (see @eq-generalized-model) is in line with these observations.
-Even when the standard deviation of the error term $\epsilon_{ijk}$ is constant, the standard deviation of observed log-normal distributions increases linearly with their mean.
+Even when the standard deviation of the error term $\varepsilon_{ijk}$ is constant, the standard deviation of observed log-normal distributions increases linearly with their mean.
 @fig-example (right) shows this pattern in our illustrative example, where the standard deviation of time responses increases proportionally with task difficulty.
@@
 
-While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html#log-normal-distributions-with-equal-variances) showing that as long as distribution shapes differ across levels, ART tends to confound effects.
+While we do not directly address this question in the main paper, we include additional results in [Appendix I](appendix_I.html#log-normal-distributions-with-equal-variances) showing that as long as distribution shapes differ across levels, ART tends to confound effects.
 Naturally, as the shapes become more similar, the problem diminishes.

### 2025-12-18 text/40-interpretingEffects.qmd
methodology..

@@
 
-Nevertheless, in most of our experiments we select scenarios that avoid these interpretational complications, ensuring that the Type I error rates of the methods we compare do not depend on how one defines interactions.
+Nevertheless, in most of our experiments we select scenarios that avoid these interpretation problems, ensuring that the Type I error rates of the methods we compare do not depend on how one defines interactions.
 We clarify this point below.
 
-**Clarifying when the interpretation of effects becomes ambiguous.** Ambiguities in interpreting interaction effects arise only when all relevant main effects are present --- that is, when the coefficients of all interacting factors are non-zero.
+**Clarifying when the interpretation of effects becomes ambiguous.** In nonlinear models, ambiguities in interpreting interaction effects arise when all relevant main effects are present --- that is, when the coefficients of all interacting factors are non-zero.
 Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero.
-In such cases, we may observe changes in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.
-We discuss such scenarios in more detail in the [appendix](appendix.html#removable-main).
+In such cases, the marginal effect of a factor (e.g., $x_1$) may vary across levels of another factor even when its associated coefficient (e.g., $a_1$) is zero.
 
-By contrast, interpretation is unambiguous in the following situations:
+By contrast, interpretation is unambiguous in the following siutations:
 
-1. *Main effects*: when the interaction coefficient $a_{12}$ is zero.
+1.
+*Main effects*, when the interaction coefficient $a_{12}$ is zero.
+In this case, if $a_1 = 0$, then the factor $x_1$ has no effect on the response scale.
+Likewise, if $a_2 = 0$, then the factor $x_2$ has no effect on the response scale.
 
-2. *Interaction effect*: when either $a_{1}$ or $a_{2}$ is zero.
+2. *Interaction effect*, when either $a_{1}$ or $a_{2}$ is zero. In this case, if $a_{12}=0$, then there is no interaction between $x_1$ and $x_2$ on the response scale.
 
-In these cases, error rates (Type I or Type II) are comparable across all methods, regardless of whether the null hypothesis is defined in terms of the coefficients of the linear model or the observed effects on the response variable $Y$.
+Proofs of these statements are provided in [Appendix II](appendix_II.html#spurious-absence).
+In these scenarios, error rates are comparable across methods, regardless of whether the null hypothesis is defined in terms of the coefficients of the linear model or the observed effects on the response scale.
 For example, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* ($a_{1} = 0$) and no interaction ($a_{12}=0$).
 
-Throughout the remainder of the article, we make every effort to clarify when the interpretation of an effect is ambiguous.
-Because different methods may rely on different assumptions about the null hypothesis, we analyze such situations separately.
-However, it is important to note that the existing literature offers no guidance on how ART is expected to interpret main and interaction effects under these conditions.
-Early evaluations of ART did not consider nonlinear models, and therefore this issue was not raised.
-While @elkin:2021 do examine such models, they do not address interpretation issues.
+Throughout the remainder of the article, we explicitly distinguish between situations in which effect interpretation is unambiguous and those in which it is not.
+Because different methods may rely on different assumptions about the null hypothesis, we analyze ambiguous cases separately.
+However, we emphasize that the existing literature offers no guidance on how ART should be expected to behave when the definition of a main or interaction effect is ambiguous.
+Early evaluations of ART focused on linear models.
+Although @elkin:2021 consider nonlinear models, they do not address how main and interaction effects should be interpreted when the definition of the null hypothesis depends on the scale of responses.
 

### 2025-12-18 text/50-ExperimentalMethod.qmd
corrected links
methodology..

@@
 We present results for five experimental designs.
-To simplify our presentation, we start with a 4 $\times$ 3 repeated-measures factorial design.
-We then show how our conclusions generalize to four additional designs: (i) a 2 $\times$ 3 between-subjects design; (ii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a repeated-measures factor; (iii) a 2 $\times$ 2 $\times$ 2 repeated-measures design, and (iv) a 3 $\times$ 3 $\times$ 3 repeated-measures design.
+To simplify our presentation, we start with a 4 $\times$ 3 within-subjects factorial design.
+We then show how our conclusions generalize to four additional designs: (i) a 2 $\times$ 3 between-subjects design; (ii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a within-subjects factor; (iii) a 2 $\times$ 2 $\times$ 2 within-subjects design, and (iv) a 3 $\times$ 3 $\times$ 3 within-subjects design.
 
@@
 We focus on three sample sizes, $n=10$, $n=20$, and $n=30$, where $n$ represents the cell size in an experimental design.
+However, for certain scenarios, we also report results for larger sample sizes, up to $n = 512$.
 In within-subjects designs, where all factors are treated as repeated measures, $n$ corresponds to the number of subjects $N$ (commonly human participants in HCI research).
 In contrast, in a 2 $\times$ 3 between-subjects design, a cell size of $n = 20$, implies a total of $N = 120$ subjects.
-For ordinal data only, we also report results for a larger range of sample sizes, up to $n = 512$.
 
@@
 When combined with unequal variances, such designs often pose challenges for both parametric procedures [@blanca:2018] and rank transformation methods [@Beasley:2009; @luepsen:2018].
-As noted earlier, we do not include unbalanced designs in our evaluation.
-However, we provide additional experimental results on missing data in the [appendix](appendix.html).
+As noted earlier, we do not consider unbalanced designs in our main article.
+However, we provide additional experimental results on missing data in [Appendix I](appendix_I.html#missing).
 
@@
 $$ 
-y_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j} + \epsilon_{ijk}
+y_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j} + \varepsilon_{ijk}
 $${#eq-linear-model}
@@
 
- - $\epsilon_{ijk}$ is the experimental error effect
+ - $\varepsilon_{ijk}$ is the experimental error effect
 
@@
 We follow the approach of @DeBruine:2021 and use the R package *faux* v1.2.1 [@faux] to simulate data for our mixed-effects models.
-We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\epsilon_{ijk} \sim N(0,\sigma)$.
+We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\varepsilon_{ijk} \sim N(0,\sigma)$.
 To simulate the observed variable $Y$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach).

### 2025-12-19 appendix_II.qmd
finished methodology

@@
 $$
-\eta(x_1, x_2) = a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2
+\ell(x_1, x_2) = a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2
 $$
@@
 $$
-\mu(x_1, x_2) = E(Y|x_1, x_2) = h(\eta(x_1, x_2))
+\mu(x_1, x_2) = E(Y|x_1, x_2) = h(\ell(x_1, x_2))
 $$
@@
 
-$$\eta(x_1, x_2) = a_0 + a_1x_1$$
+$$\ell(x_1, x_2) = a_0 + a_1x_1$$
 
@@
 
-$$\frac{\partial \mu(x_1, x_2)}{\partial x_2}  = h'(\eta) \frac{\partial \eta}{\partial x_2} = 0$$
+$$\frac{\partial \mu(x_1, x_2)}{\partial x_2}  = h'(\ell) \frac{\partial \ell}{\partial x_2} = 0$$
 where $h'$ denotes the first derivative of $h$.
@@
 
-$$\frac{\partial^2 \mu(x_1, x_2)}{\partial x_2 \partial x_1} = h''(\eta) \frac{\partial \eta}{\partial x_2} \frac{\partial \eta}{\partial x_1} + h'(\eta)\frac{\partial^2\eta}{\partial x_2 \partial x_1} = 0$$
+$$\frac{\partial^2 \mu(x_1, x_2)}{\partial x_2 \partial x_1} = h''(\ell) \frac{\partial \ell}{\partial x_2} \frac{\partial \ell}{\partial x_1} + h'(\ell)\frac{\partial^2\ell}{\partial x_2 \partial x_1} = 0$$
 
@@
 
-### Illustrative examples
+### Examples
 We discuss two examples.
 
-#### Lognormal data
+#### Log-normal distributions
 Let $x_1$ and $x_2$ be two experimental factors. Suppose the true model of the data is
@@
 
-$$Y^* = a_0 + a_1 x_1 + \varepsilon$$
+$$Y^* = a_0 + a_1 x_1 + \varepsilon, \qquad
+\varepsilon \sim \mathcal{N}(0, \sigma^2)$$
 
@@
 \begin{aligned}
-Y = k\ \ \text{if}\ \ \tau_{k-1} < Y^* \leq \tau_k, 
-\qquad k = 1,2 ... K
+Y = d\ \ \text{if}\ \ \tau_{d-1} < Y^* \leq \tau_k, 
+\qquad d = 1,2 ... D
 \end{aligned} 
 $$ 
-Again, the conditional distribution $Y| (x_1, x_2)$ depends on the independent variables only through $\eta(x_1, x_2)=a_0+a_1 x_1$, and there is no effect of $x_2$ on either the latent or the observed ordinal scale.
+Again, the conditional distribution $Y| (x_1, x_2)$ depends on the independent variables only through $\ell(x_1, x_2)=a_0+a_1 x_1$, and there is no effect of $x_2$ on either the latent or the observed ordinal scale.
 
-In this case, the mapping from $\eta$ to the observed response is monotone but not strictly monotone and is therefore non-invertible.
+In this case, the mapping from $\ell$ to the observed response is monotone but not strictly monotone and is therefore non-invertible.
 As a result, a true effect of $x_2$ on the latent variable (i.e., $a_2 \neq 0$) may be attenuated or entirely unobservable on the ordinal response scale.
@@
 Because $T$ is monotone, it preserves the ordering of observations within each condition.
-The transformed conditional distribution $T(Y) | (x_1, x_2)$ therefore depends on the independent variables only through $\eta(x_1, x_2) = a_0+a_1 x_1$.
+The transformed conditional distribution $T(Y) | (x_1, x_2)$ therefore depends on the independent variables only through $\ell(x_1, x_2) = a_0+a_1 x_1$.
 Consequently, the mean, median, and all quantiles of $T(Y) | (x_1, x_2)$ remain invariant to $x_2$, and marginal effects of $x_1$ do not depend on $x_2$.

### 2025-12-19 sandbox.qmd
finished methodology

@@
 
-1. *Normal distribution* $\mathcal{N}(\mu, \sigma)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation. For most experiments where variances are equal, we set $\sigma = 1$.
+1. *Normal distribution* $\mathcal{N}(\mu, \sigma^2)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation. For most experiments where variances are equal, we set $\sigma^2 = 1$.
 2.
-*Log-normal distribution* $\mathcal{LogN}(\mu, \sigma)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation at the logarithmic scale.
+*Log-normal distribution* $\mathcal{LogN}(\mu, \sigma^2)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation at the logarithmic scale.
 The log-normal distribution is a good model for various measures bounded by zero, such as task-completion times.
@@
 We set $\kappa = 10$ in our main experiments but evaluate a wider range of parameters in [Appendix I](appendix_I.html#binomial).
-7. Distributions of *Likert-item* responses with 5, 7, and 11 levels.
+7. Distributions of *Likert-item* responses with 5, 7, and 11 discrete levels.
 
 **Experimental designs.** We present results for five experimental designs.
-To simplify our presentation, we start with a 4 $\times$ 3 within-subjects factorial design.
-We then show how our conclusions generalize to four additional designs: (i) a 2 $\times$ 3 between-subjects design; (ii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a within-subjects factor; (iii) a 2 $\times$ 2 $\times$ 2 within-subjects design, and (iv) a 3 $\times$ 3 $\times$ 3 within-subjects design.
+To simplify our presentation, we start with (i) a 4 $\times$ 3 within-subjects factorial design.
+We then show how our conclusions generalize to four additional designs: (ii) a 2 $\times$ 3 between-subjects design; (iii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a within-subjects factor; (iv) a 2 $\times$ 2 $\times$ 2 within-subjects design; and (v) a 3 $\times$ 3 $\times$ 3 within-subjects design.
 
@@
 
-**Equal vs. unequal variances.** We test the robustness of the methods when the variances on the latent variable are unequal.
+**Equal vs. unequal variances.** For normal and ordinal distributions, we test the robustness of the methods when variances are unequal.
 
-**Evaluation measures.** In addition to Type I error rates, we compare the statistical power of the methods and assess the accuracy of their effect size estimates.
+**Evaluation measures.** In addition to Type I error rates, we compare the statistical power of the methods and compare their effect size estimates with ground-truth estimates.
 
@@
 $$ 
-\eta_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j}
+\ell_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j}
 $${#eq-linear-model}
@@
 
-**Generating ratio data.** We adopt generalized linear modeling for ratio data, where we use an inverse link function $g^{-1}$ to make the connection between $\eta_{ijk}$ and the location parameter of the distribution of interest:
+**Generating ratio data.** We adopt generalized linear modeling [@Bolker2009glm] for ratio data, where we use an inverse link function $g^{-1}$ to make the connection between $\ell_{ijk}$ and the location parameter of the distribution of interest:
 $$
-Y_{ijk} \sim \mathcal{F}(location=g^{-1}(\eta_{ijk}), ...)
+Y_{ijk} \sim \mathcal{F}(location=g^{-1}(\ell_{ijk}), ...)
 $${#eq-glm-ratio}
 
-where $\mathcal{F}$ denotes the probability density (or mass) function from which we draw samples.
+where $\mathcal{F}$ denotes the probability density (or mass) function from which we draw the responses $Y_{ijk}$.
 For example, the mean of the exponential distribution is equal to $1 / \lambda$, where $\lambda$ is its rate parameter.
-This mean is linked to the linear term through the function $g^{-1}(x) = e^{x}$, and therefore, $\lambda = e^{-\eta_{ijk}}$.
+This mean is linked to the linear term through the function $g^{-1}(x) = e^{x}$, and therefore, $\lambda = e^{-\ell_{ijk}}$.
+Or the link function commonly used for the binomial distribution is the *logit* function, but its inverse if the *logistic* function $g^{-1}(x) = 1 / (1 + e^{-x})$.
 
- @tbl-distributions presents how the parameters of the other distributions are linked to the linear term.
+ @tbl-distributions presents how the location parameters of the other distributions are linked to the linear term.
+We emphasize that the Cauchy distribution does not define a GLM (since it does not belong to the exponential family), but we use a model that connects its location (median) through the identity link.
 
-| distribution | parameters | description               |
+| distribution | location parameter | description               |
 |--------------|------------|---------------------------|
-| Normal       |$\mu = \eta_{ijk}$ <br> $\sigma^2 = \sigma_{\varepsilon}^2$ | mean <br> standard deviation |
-| Log-normal    |$\mu = \eta_{ijk}$ <br> $\sigma^2 = \sigma_{\varepsilon}^2$ | mean at logarithmic scale (logarithm of median) <br> standard deviation at logarithmic scale|
-| Exponential  |$\lambda = e^{-\eta_{ijk}}$ | rate (1 / mean) |
-| Cauchy       |$x_0 = \eta_{ijk}$ <br> $\gamma = 1$ | location (median) <br> scale 
-| Poisson      |$\lambda = e^{\eta_{ijk}}$ | rate (mean) |
-| Binomial     |$p = Logistic(\eta_{ijk})$ <br> $\kappa = 10$ | probability of error <br> number of Bernoulli trials |
+| Normal       |$\mu = \ell_{ijk}$ | mean |
+| Log-normal    |$\mu = \ell_{ijk}$ | mean at logarithmic scale (logarithm of median) |
+| Exponential  |$\lambda = e^{-\ell_{ijk}}$ | rate (1 / mean) |
+| Cauchy       |$x_0 = \ell_{ijk}$ | median |
+| Poisson      |$\lambda = e^{\ell_{ijk}}$ | rate (mean) |
+| Binomial     |$p = 1 / (1 + e^{-\ell_{ijk}})$ | probability of success or error |
 
-: Distribution parameters and their links to the linear term $\eta_{ijk}$ {#tbl-distributions}
+: Distribution parameters and their links to the linear term $\ell_{ijk}$ {#tbl-distributions}
 
-**Generating ordinal data.**
+**Generating ordinal data.** For generating ordinal data, we follow a latent-variable approach as explained by @Liddell:2018.
+Specifically, we consider that there is a continuous latent variable that follows a normal distribution:
+$$
+Y_{ijk}^* \sim \mathcal{N}(\mu = \ell_{ijk}, \sigma^2 = 1)
+$${#eq-ordinal-latent}
+From this, we derive discrete responses $d = 1,2 ... D$ by applying fixed thresholds $\tau_d \in (-\infty, \infty)$ as follows:
+$$
+\begin{aligned}
+Y_{ijk} = d\ \ \text{if}\ \ \tau_{d-1} < Y_{ijk}^* \leq \tau_d
+\end{aligned} 
+$${#eq-ordinal-thresholds}
+Since, in practice, $\tau_0 = -\infty$ and $\tau_D = \infty$, generating responses with $D$ discrete levels requires $D-1$ thresholds.
 
-### Population control and distribution conversions 
-To simplify our simulations, we fix the following population parameters: $\mu = 0$, $\sigma = 1$, and $\sigma_s = 1$.
-We then control the magnitude of effects by varying $a_1$, $a_2$, and, for some experiments $a_{12}$.
-@fig-effects presents the range of values that we test for $a_1$ and $a_2$.
-We also visualize their effects on the latent variable for factors with two, three, and four categorical levels.
+### Effects and population control
+We next detail how we select the parameters of the linear component (see @eq-linear-model) for the various distributions, as well as the thresholds of the ordinal model (see @eq-ordinal-thresholds).
 
-::: {#fig-effects}
-```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
-source("effects_plot.R")
-plotEffects()
+**Magnitude of effects.** We control the magnitude of effects by varying $a_1$, $a_2$, and, for some experiments (those evaluating power for interactions), $a_{12}$.
+@fig-effects-normal presents normal distributions ($\sigma^2 = 1$) generated across the range of values of $a_1$ used to evaluate Type I error rates.
+In these examples, we illustrate effects for a factor with three categorical levels and, for clarity, disregard between-subject variability ($\sigma_k^2 = 0$).
+
+::: {#fig-effects-normal}
+```{r, echo=FALSE, message=FALSE, fig.height=1.2, fig.width = 9, warning=FALSE}
+source("distributions_plots.R")
+plot_effects_normal()
 ```
-Varying $a_1$ or $a_2$ to control the magnitude of main effects.
-The plots show examples of population distributions on the latent variable $\mathcal{Y}$ for factors ($X_1$ or $X_2$) with two (top), three (middle), or four categorical levels (bottom).
+Varying $a_1$ to control the magnitude of the main effect of $X_1$. The plots show examples of normal distributions ($\sigma^2=1$) for a factor with three categorical levels. 
 :::
 
-We follow the approach of @DeBruine:2021 and use the R package *faux* v1.2.1 [@faux] to simulate data for our mixed-effects models.
-We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\varepsilon_{ijk} \sim N(0,\sigma)$.
-To simulate the observed variable $Y$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach).
-A key advantage of this method is that we can generate responses for any distribution, while we control effects on the latent variable in the exact same way.
+**Grand mean and variance parameters.** We select the mean term $\mu$ to match a target location of the responses.
+@tbl-locations reports the target location for each distribution, which is typically the overall mean (or the median in the case of the Cauchy distribution).
+We describe how $\mu$ is derived from these parameters in [Appendix II](appendix_II.html#mu).
+
+| Normal | Log-normal | Exponential | Cauchy | Poisson | Binomial | Ordinal |
+|:------:|:----------:|:---------:|:-------:|:--------:|:--------:|:--------:|
+| 0 | 1 | 0.5 | 0 | 3 | 0.1 (per trial) | 0 |
+
+: Target mean (median for the Cauchy distribution) of responses $Y_{ijk}$ {#tbl-locations}
+
+@fig-effects-distributions shows distributions generated with these location values when $a_1 = 2$ for a factor with three categorical levels.
+
+::: {#fig-effects-distributions}
+```{r, echo=FALSE, message=FALSE, fig.height=1.2, fig.width = 9, warning=FALSE}
+source("distributions_plots.R")
+plot_effects_distr()
+```
+Different ratio distributions generated with $a_1 = 2$ when $X_1$ has three categorical levels (here, $\sigma_k^2 = 0$).
+:::
+
+In practice, we randomly choose the standard deviation $\sigma_k^2$ --- which determines between-subject variability --- from the interval $[0.1, 0.5]$, drawing each value from a uniform distribution.
+
+**Thresholds of ordinal scales.** For ordinal data, we evaluate $D =$ 5, 7, or 11 discrete levels.
+To derive the discretization thresholds $\tau_d$, we first consider the range $[-2SD, 2SD]$ , where $SD$ is the overall standard deviation of the latent variable $Y^{*}_{ijk}$.
+We then divide this range into 5, 7, or 11 intervals, following two different strategies: (i) setting thresholds to be *equidistant*; or (ii) considering *flexible* thresholds, randomly drawing their position in the above range.
+
+ @fig-ordinal presents examples of equidistant and flexible thresholds for a 5-level scale when the magnitude of main effect of $X_1$ is either $a_1 = 2$ or $a_1 = 8$, while all other effects are zero.
+
+::: {#fig-ordinal}
+```{r, echo=FALSE, message=FALSE, fig.height=2.5, fig.width = 9, warning=FALSE}
+plot_thresholds()
+```
+The four vertical lines in each plot represent thresholds defining 5-level ordinal scales.
+Thresholds are either equidistant (top) or flexible (bottom) within a range of $\pm 2$ standard deviations around the grand mean $\mu = 0$.
+:::
 
@@
 We use the *Rankit* formulation [@Bliss:1956] for the inverse normal transformation (INT), as explained earlier.
-Unless explicitly mentioned otherwise, we use the formula ``Y' ~ X1*X2 + (1|S)`` with R's *lmer* function, except for the between-subjects design where we use the formula ``Y' ~ X1*X2 + Error(S)`` with R's *aov* function.
+
+Unless stated otherwise, we analyze the data using ANOVA with R’s *aov()* function and the model specification ``Y' ~ X1 * X2 + Error(S)``.
+For balanced designs, this approach yields results that are identical or nearly identical to those obtained using R's *lmer()* function with the model specification ``Y' ~ X1 * X2 + (1 | S)``.
 

### 2025-12-19 text/50-ExperimentalMethod.qmd
finished methodology

@@
 We evaluate the standard parametric approach (*PAR*) and the three rank-transformation methods (*RNK*, *INT*, and *ART*) that we introduced earlier.
-We conduct a series of Monte Carlo experiments that assess their performance under a variety of experimental configurations:
+We conduct a series of Monte Carlo experiments that assess their performance under a variety of experimental configurations.
 
-1.
-We evaluate *ratio* and *ordinal* data.
-For ratio data, we examine four representative *continuous* distributions (normal, log-normal, exponential, and Cauchy distribution) and two *discrete* distributions (Poisson and binomial distribution).
-For ordinal data, we examine distributions for 5-level, 7-level, and 11-level Likert item responses.
+**Distributions.** We evaluate both *ratio* and *ordinal* data, drawn from the following distributions:
 
+1. *Normal distribution* $\mathcal{N}(\mu, \sigma^2)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation. For most experiments where variances are equal, we set $\sigma^2 = 1$.
 2.
-We present results for five experimental designs.
-To simplify our presentation, we start with a 4 $\times$ 3 within-subjects factorial design.
-We then show how our conclusions generalize to four additional designs: (i) a 2 $\times$ 3 between-subjects design; (ii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a within-subjects factor; (iii) a 2 $\times$ 2 $\times$ 2 within-subjects design, and (iv) a 3 $\times$ 3 $\times$ 3 within-subjects design.
+*Log-normal distribution* $\mathcal{LogN}(\mu, \sigma^2)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation at the logarithmic scale.
+The log-normal distribution is a good model for various measures bounded by zero, such as task-completion times.
+We set $\sigma = 1$ in our main experiments but evaluate a wider range of parameters in [Appendix I](appendix_I.html#lognormal).
+3.
+*Exponential distribution* $Exp(\lambda)$, where $\lambda$ is its rate.
+The exponential distribution naturally emerges when describing the time elapsed between events.
+For example, we could use it to model the time a random person spends with a public display, or the waiting time before a new person approaches to interact with the display, when the average waiting time is $\frac{1}{\lambda}$.
+4.
+*Cauchy distribution* $\mathcal{Cauchy}(x_0, \gamma)$, where $x_0$ defines its location (median) and $\gamma$ defines its scale.
+The Cauchy distribution is the distribution of the ratio of two independent normally distributed random variables.
+It rarely emerges in practice.
+However, it is commonly used in statistics to test the robustness of statistical procedures because both its mean and variance are undefined.
+As we discussed earlier, past evaluations of ART [@Mansouri:1995; @elkin:2021] show that the method fails under the Cauchy distribution.
+We set $\gamma = 1$.
+5.
+*Poisson distribution* $\mathcal{Pois}(\lambda)$, where $\lambda$ is its rate.
+It expresses the probability of a given number of events in a fixed interval of time.
+For example, we could use it to model the number of people who interact with a public display in an hour, when the average rate is $\lambda$ people per hour.
+6.
+*Binomial distribution* $\mathcal{B}(\kappa, p)$, where $\kappa$ defines the number of independent Bernoulli trials and $p$ is the probability of error or success of each trial.
+It frequently appears in HCI research, as it can model the number of successes and failures in a series of experimental tasks.
+We set $\kappa = 10$ in our main experiments but evaluate a wider range of parameters in [Appendix I](appendix_I.html#binomial).
+7. Distributions of *Likert-item* responses with 5, 7, and 11 discrete levels.
 
-3.
-We focus on three sample sizes, $n=10$, $n=20$, and $n=30$, where $n$ represents the cell size in an experimental design.
+**Experimental designs.** We present results for five experimental designs.
+To simplify our presentation, we start with (i) a 4 $\times$ 3 within-subjects factorial design.
+We then show how our conclusions generalize to four additional designs: (ii) a 2 $\times$ 3 between-subjects design; (iii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a within-subjects factor; (iv) a 2 $\times$ 2 $\times$ 2 within-subjects design; and (v) a 3 $\times$ 3 $\times$ 3 within-subjects design.
+
+**Sample sizes.** We focus on three sample sizes, $n=10$, $n=20$, and $n=30$, where $n$ represents the cell size in an experimental design.
 However, for certain scenarios, we also report results for larger sample sizes, up to $n = 512$.
@@
 
-4. We test the robustness of the methods when the variances on the latent variable are unequal.
+**Equal vs. unequal variances.** For normal and ordinal distributions, we test the robustness of the methods when variances are unequal.
 
-5. In addition to Type I error rates, we compare the statistical power of the methods and assess the accuracy of their effect size estimates.
+**Evaluation measures.** In addition to Type I error rates, we compare the statistical power of the methods and compare their effect size estimates with ground-truth estimates.
 
-6. We analyze both main and interaction effects, examining how increasing the effect size of one or two factors influences the Type I error rates of other factors and their interactions.
+**Magnitude of effects.** We analyze both main and interaction effects, examining how increasing the magnitude of effect of one factor influences the Type I error rates of other factors and their interactions.
+In addition, we examine how increasing two main effects in parallel influences the Type I error rate of their interaction, while emphasizing that the definition of the null hypothesis in this case is ambiguous.
 
@@
 ### Statistical modeling
-To model the latent variable $\mathcal{Y}$, we use a two-way (two factors) or a three-way (three factors) mixed-effects model.
-For simplicity, we explain here the model for two factors.
-Its extension to three factors is straightforward.
-The model has the following form:
+For simplicity, we explain here our modeling approach for two factors, but its extension to three factors is straightforward.
+
+**Linear component.** The base of our data generation process for all distributions is the following linear term:
 
 $$ 
-y_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j} + \varepsilon_{ijk}
+\ell_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j}
 $${#eq-linear-model}
@@
 
- - $s_k$ is the random intercept effect of the *k*-th subject, where $k = 1..n$
+ - $s_k \sim \mathcal{N}(0, \sigma_s^2)$ is the random intercept effect of the *k*-th subject, where $k = 1..n$
 
@@
 
+<!--
  - $\varepsilon_{ijk}$ is the experimental error effect
+ -->
 
-To encode the levels of the two factors $x_{1i} \in X_1$ and $x_{2j} \in X_2$ we proceed as follows:
+ While random slope effects can have an impact on real experimental data [@barr:2013], we do not consider them here for two main reasons: (1) to be consistent with previous evaluations of the ART procedure [@elkin:2021]; and (2) because mixed-effects procedures with random slope effects are computationally demanding, adding strain to simulation resources.
+However, there is no good reason to believe that adding random slope effects would impact our findings and conclusions.
+
+**Encoding the levels of factors.** To encode the levels of the two factors $x_{1i} \in X_1$ and $x_{2j} \in X_2$ we proceed as follows:
 
@@
 
-While random slope effects can have an impact on real experimental data [@barr:2013], we do not consider them here for two main reasons: (1) to be consistent with previous evaluations of the ART procedure [@elkin:2021]; and (2) because mixed-effects procedures with random slope effects are computationally demanding, adding strain to simulation resources.
-However, there is no good reason to believe that adding random slope effects would impact our findings and conclusions.
+**Generating ratio data.** We adopt generalized linear modeling [@Bolker2009glm] for ratio data, where we use an inverse link function $g^{-1}$ to make the connection between $\ell_{ijk}$ and the location parameter of the distribution of interest:
+$$
+Y_{ijk} \sim \mathcal{F}(location=g^{-1}(\ell_{ijk}), ...)
+$${#eq-glm-ratio}
 
-### Population control and distribution conversions 
-To simplify our simulations, we fix the following population parameters: $\mu = 0$, $\sigma = 1$, and $\sigma_s = 1$.
-We then control the magnitude of effects by varying $a_1$, $a_2$, and, for some experiments $a_{12}$.
-@fig-effects presents the range of values that we test for $a_1$ and $a_2$.
-We also visualize their effects on the latent variable for factors with two, three, and four categorical levels.
+where $\mathcal{F}$ denotes the probability density (or mass) function from which we draw the responses $Y_{ijk}$.
+For example, the mean of the exponential distribution is equal to $1 / \lambda$, where $\lambda$ is its rate parameter.
+This mean is linked to the linear term through the function $g^{-1}(x) = e^{x}$, and therefore, $\lambda = e^{-\ell_{ijk}}$.
+Or the link function commonly used for the binomial distribution is the *logit* function, but its inverse if the *logistic* function $g^{-1}(x) = 1 / (1 + e^{-x})$.
 
-::: {#fig-effects}
-```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
-source("effects_plot.R")
-plotEffects()
+ @tbl-distributions presents how the location parameters of the other distributions are linked to the linear term.
+We emphasize that the Cauchy distribution does not define a GLM (since it does not belong to the exponential family), but we use a model that connects its location (median) through the identity link.
+
+| distribution | location parameter | description               |
+|--------------|------------|---------------------------|
+| Normal       |$\mu = \ell_{ijk}$ | mean |
+| Log-normal    |$\mu = \ell_{ijk}$ | mean at logarithmic scale (logarithm of median) |
+| Exponential  |$\lambda = e^{-\ell_{ijk}}$ | rate (1 / mean) |
+| Cauchy       |$x_0 = \ell_{ijk}$ | median |
+| Poisson      |$\lambda = e^{\ell_{ijk}}$ | rate (mean) |
+| Binomial     |$p = 1 / (1 + e^{-\ell_{ijk}})$ | probability of success or error |
+
+: Distribution parameters and their links to the linear term $\ell_{ijk}$ {#tbl-distributions}
+
+**Generating ordinal data.** For generating ordinal data, we follow a latent-variable approach as explained by @Liddell:2018.
+Specifically, we consider that there is a continuous latent variable that follows a normal distribution:
+$$
+Y_{ijk}^* \sim \mathcal{N}(\mu = \ell_{ijk}, \sigma^2 = 1)
+$${#eq-ordinal-latent}
+From this, we derive discrete responses $d = 1,2 ... D$ by applying fixed thresholds $\tau_d \in (-\infty, \infty)$ as follows:
+$$
+\begin{aligned}
+Y_{ijk} = d\ \ \text{if}\ \ \tau_{d-1} < Y_{ijk}^* \leq \tau_d
+\end{aligned} 
+$${#eq-ordinal-thresholds}
+Since, in practice, $\tau_0 = -\infty$ and $\tau_D = \infty$, generating responses with $D$ discrete levels requires $D-1$ thresholds.
+
+### Effects and population control
+We next detail how we select the parameters of the linear component (see @eq-linear-model) for the various distributions, as well as the thresholds of the ordinal model (see @eq-ordinal-thresholds).
+
+**Magnitude of effects.** We control the magnitude of effects by varying $a_1$, $a_2$, and, for some experiments (those evaluating power for interactions), $a_{12}$.
+@fig-effects-normal presents normal distributions ($\sigma^2 = 1$) generated across the range of values of $a_1$ used to evaluate Type I error rates.
+In these examples, we illustrate effects for a factor with three categorical levels and, for clarity, disregard between-subject variability ($\sigma_k^2 = 0$).
+
+::: {#fig-effects-normal}
+```{r, echo=FALSE, message=FALSE, fig.height=1.2, fig.width = 9, warning=FALSE}
+source("distributions_plots.R")
+plot_effects_normal()
 ```
-Varying $a_1$ or $a_2$ to control the magnitude of main effects.
-The plots show examples of population distributions on the latent variable $\mathcal{Y}$ for factors ($X_1$ or $X_2$) with two (top), three (middle), or four categorical levels (bottom).
+Varying $a_1$ to control the magnitude of the main effect of $X_1$. The plots show examples of normal distributions ($\sigma^2=1$) for a factor with three categorical levels. 
 :::
 
-We follow the approach of @DeBruine:2021 and use the R package *faux* v1.2.1 [@faux] to simulate data for our mixed-effects models.
-We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\varepsilon_{ijk} \sim N(0,\sigma)$.
-To simulate the observed variable $Y$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach).
-A key advantage of this method is that we can generate responses for any distribution, while we control effects on the latent variable in the exact same way.
+**Grand mean and variance parameters.** We select the mean term $\mu$ to match a target location of the responses.
+@tbl-locations reports the target location for each distribution, which is typically the overall mean (or the median in the case of the Cauchy distribution).
+We describe how $\mu$ is derived from these parameters in [Appendix II](appendix_II.html#mu).
+
+| Normal | Log-normal | Exponential | Cauchy | Poisson | Binomial | Ordinal |
+|:------:|:----------:|:---------:|:-------:|:--------:|:--------:|:--------:|
+| 0 | 1 | 0.5 | 0 | 3 | 0.1 (per trial) | 0 |
+
+: Target mean (median for the Cauchy distribution) of responses $Y_{ijk}$ {#tbl-locations}
+
+@fig-effects-distributions shows distributions generated with these location values when $a_1 = 2$ for a factor with three categorical levels.
+
+::: {#fig-effects-distributions}
+```{r, echo=FALSE, message=FALSE, fig.height=1.2, fig.width = 9, warning=FALSE}
+source("distributions_plots.R")
+plot_effects_distr()
+```
+Different ratio distributions generated with $a_1 = 2$ when $X_1$ has three categorical levels (here, $\sigma_k^2 = 0$).
+:::
+
+In practice, we randomly choose the standard deviation $\sigma_k^2$ --- which determines between-subject variability --- from the interval $[0.1, 0.5]$, drawing each value from a uniform distribution.
+
+**Thresholds of ordinal scales.** For ordinal data, we evaluate $D =$ 5, 7, or 11 discrete levels.
+To derive the discretization thresholds $\tau_d$, we first consider the range $[-2SD, 2SD]$ , where $SD$ is the overall standard deviation of the latent variable $Y^{*}_{ijk}$.
+We then divide this range into 5, 7, or 11 intervals, following two different strategies: (i) setting thresholds to be *equidistant*; or (ii) considering *flexible* thresholds, randomly drawing their position in the above range.
+
+ @fig-ordinal presents examples of equidistant and flexible thresholds for a 5-level scale when the magnitude of main effect of $X_1$ is either $a_1 = 2$ or $a_1 = 8$, while all other effects are zero.
+
+::: {#fig-ordinal}
+```{r, echo=FALSE, message=FALSE, fig.height=2.5, fig.width = 9, warning=FALSE}
+plot_thresholds()
+```
+The four vertical lines in each plot represent thresholds defining 5-level ordinal scales.
+Thresholds are either equidistant (top) or flexible (bottom) within a range of $\pm 2$ standard deviations around the grand mean $\mu = 0$.
+:::
 
@@
 Our R code was parallelized to use all four cores of each machine.
-Some experiments took a few hours to complete, while others took several days.
 
 To estimate the power and Type I error rates of the four methods with enough precision, we ran $5000$ iterations for each population configuration and each sample size.
-
-<!--
-We evaluate the Type I error rate, that is the rate of false positives of each method.
-To evaluate the Type I error rate for the interaction effect, we set $a_{12} = 0$.
-Likewise, to evaluate the Type I error rate for the second factor $X_2$, we set $a_{2} = 0$.
-For each population configuration and sample size $n$, we run $5000$ iterations and test two significance levels: $\alpha = .05$ and $\alpha = .01$.
-For brevity, we only report results for $\alpha = .05$ in the main paper, while additional results are presented in supplementary materialsxa
--->

### 2025-12-19 text/60-Results-RatioScales.qmd
finished methodology

@@
 ### Type I error rates in ratio scales {#ratio}
-We first evaluate Type I error rates for ratio scales.
-We test a 4 $\times$ 3 repeated-measures design, where we refer to the 4-level factor as $X_1$ and the 3-level factor as $X_2$.
-For the observed response variable $Y = \mathcal{T}(\mathcal{Y})$, we evaluate transformations to four continuous and two discrete distributions:
-
-1. No transformation, or $Y = \mathcal{Y}$. Distributions are normal with a grand mean $\mu = 0$ (e.g., see @fig-effects).
-
-2.
-Log-normal distribution $LogN(\mu, \sigma)$ with global parameters $\mu = 0$ and $\sigma = 1$.
-As discussed in [the introduction](#intro), the log-normal distribution is a good model for various measures bounded by zero, such as task-completion times.
-
-3.
-Exponential distribution $Exp(\lambda)$ with a global parameter $\lambda = 2$.
-The exponential distribution naturally emerges when describing the time elapsed between events.
-For example, we could use it to model the time a random person spends with a public display, or the waiting time before a new person approaches to interact with the display, when the average waiting time is $\frac{1}{\lambda}$.
-
-4.
-Cauchy distribution $Cauchy(x_0,\gamma)$ with global parameters $x_0 = 0$ and $\gamma = 1$.
-The Cauchy distribution is the distribution of the ratio of two independent normally distributed random variables.
-It rarely emerges in practice.
-However, it is commonly used in statistics to test the robustness of statistical procedures because both its mean and variance are undefined.
-As we discussed earlier, past evaluations of ART [@Mansouri:1995; @elkin:2021] show that the method fails under the Cauchy distribution.
-
-5.
-Poisson distribution $Pois(\lambda)$ with a single parameter $\lambda = 3$.
-It expresses the probability of a given number of events in a fixed interval of time.
-For example, we could use it to model the number of people who interact with a public display in an hour, when the average rate is $\lambda = 3$ people per hour.
-
-6.
-Binomial distribution $B(k,p)$ with parameters $k = 10$ and $p=.1$.
-It frequently appears in HCI research, as it can model the number of successes and failures in a series of experimental tasks.
-For example, we could use it to model the number of errors that participants make in a series of $k = 10$ repetitions of a memorization task, when the average error rate is $10\%$, thus the average error probability is $p=.1$.
-
-Specifically for the log-normal and binomial distributions, we present results for a wider range of their parameters in the [appendix](appendix.html).
+We first evaluate Type I error rates for ratio scales. We test a 4 $\times$ 3 repeated-measures design, where we refer to the 4-level factor as $X_1$ and the 3-level factor as $X_2$.
 

### 2025-12-19 text/61-Results-OrdinalScales.qmd
finished methodology

@@
 ### Type I error rates in ordinal scales {#ordinal}
-We also evaluate Type I error rates for ordinal scales.
-We test again a $4 \times 3$ repeated-measures design.
-We focus on individual Likert items levels and implement an ordered-probit method [@Liddell:2018] to discretize the latent variable $Y$ into 5, 7, or 11 ordinal levels.
-To derive the discretization thresholds, we first consider the range $[-2SD, 2SD]$ , where $SD$ is the overall standard deviation of the responses $y_{ijk}$.
-We then divide this range into 5, 7, or 11 intervals, following two different strategies: (i) setting thresholds to be *equidistant*; or (ii) considering *flexible* thresholds, randomly drawing their position in the above range.
-@fig-ordinal presents examples of equidistant and flexible thresholds for a 5-level scale when the magnitude of main effect of $X_1$ is either $a_1 = 2$ or $a_1 = 8$, while all other effects are zero.
-
-::: {#fig-ordinal}
-```{r, echo=FALSE, message=FALSE, fig.height=2.5, fig.width = 9, warning=FALSE}
-plotThresholds()
-```
-The four vertical lines in each plot represent thresholds defining 5-level ordinal scales.
-Thresholds are either equidistant (top) or flexible (bottom) within a range of $\pm 2$ standard deviations around the grand mean $\mu = 0$.
-:::
+We also evaluate Type I error rates for ordinal scales. We test again a $4 \times 3$ repeated-measures design.
 

### 2025-12-19 distributions_plots.R

@@ -1,0 +1,177 @@
+# Author: Theophanis Tsandilas, Inria & Université Paris-Saclay
+
+library(dplyr)
+library(tidyr)
+library(egg)
+library(latex2exp)
+
+source("utils-data-generation.R")
+
+# Distribution parameters used in simulation
+params_norm   <- list(sigma_e = 1, sigma_s = c(0, 0))
+params_lnorm  <- list(sigma_e = 1, mean_target = 1, sigma_s = c(0, 0))
+params_cauchy <- list(gamma = 1, sigma_s = c(0, 0))
+params_exp    <- list(mean_target = 0.5, sigma_s = c(0, 0))
+params_poisson <- list(mean_target = 3, sigma_s = c(0, 0))
+params_binom  <- list(size = 10, p_target = 0.1, sigma_s = c(0, 0))
+
+plot_densities <- function(
+			n = 10000, nlevels = 2, effectSize = 4, family = "norm", params = params_norm, 
+			alpha = 0.5, breaks = seq(-8, 8, by=4), xlim = c(-10, 10)
+) {
+  data <- simulate_response(nlevels = nlevels, within = 1, n = n, coeffs = c("X1"=effectSize), family = family, params = params) 
+
+  palette <- c("#888888", "#4E84C4", "red", "#FF5E00")
+
+  plot <- ggplot(data) + geom_density(aes(x = Y, fill = X1, color = X1), alpha = alpha) + scale_color_manual(values = rep("#FFFFFF00",nlevels)) + scale_fill_manual(values = palette) + 
+  theme_classic() +
+  theme(plot.margin=unit(c(0,0,0,0), "cm"), legend.position = "none", axis.title.x  = element_blank(), axis.line.y  = element_blank(), 
+  	axis.title.y = element_blank(), axis.ticks.y  = element_blank(), axis.text.y  = element_blank()) +
+  scale_x_continuous(name =element_blank(), breaks = breaks, limits = xlim)
+
+  plot
+}
+
+plot_densities_discrete <- function(
+			n = 10000, nlevels = 2, effectSize = 4, family = "binom", params = params_binom, 
+			alpha = 0.5, breaks = seq(0, 10, by=1), xlim = c(0, 10)
+) {
[…]

### 2025-12-19 utils-data-generation.R

@@ -1,0 +1,328 @@
+# Code for generating experimental designs (within-, between-subjects, and mixed) and respones for a range of distributions
+# Author: Theophanis Tsandilas, Dec 2025
+
+##################################################################################
+# Encodes the 1:nlevels to numerical values
+# It ensures a fixed distance between the two extreme levels (equal to 1) and random distances in between
+# It also ensures that the sum is zero
+coding <- function(nlevels) { 
+  levels <- sort(c(0, 1, runif(nlevels - 2)))
+  return(levels - mean(levels))
+}
+
+# Generate the basic design matrix (balanced)
+build_structure <- function(nlevels = c(2,2)) {
+  k <- length(nlevels)
+
+  # 1. Create factor labels and numeric codings, e.g., for X1: A1, A2, ... and for X2: B1, B2, ...  
+  Xlabs <- lapply(seq_len(k), function(j) paste0(LETTERS[j], seq_len(nlevels[j])))
+  xcoding <- lapply(nlevels, coding)
+
+  # 2: Create a frame with the data variables
+  df <- expand.grid(
+    lapply(1:k, function(j) Xlabs[[j]])
+  )
+  names(df) <- paste0("X", 1:k)
+
+  # 3: Match numeric coding for each level
+  for (j in seq_len(k)) {
+    df[[paste0("x", j)]] <- xcoding[[j]][ match(df[[paste0("X", j)]], Xlabs[[j]]) ]
+  }
+  
+  df
+}
+
+# Create the full design, adding the participants
+# n is the cell sample - the number of times a specific combination of factor levels will appear 
+# For within-subjects designs, it is also the number of subjects
+make_design <- function(nlevels = c(4,3), within = c(1,1), n = 20) {
+  # 1. Create the basic full factorial structure
+  df <- build_structure(nlevels)
+
+  k <- length(nlevels)
+  if (length(within) != k)
+    stop("within must have same length as nlevels")
+
+  # 2. Identify between-subject factors
+  between_idx <- which(within == 0)
+
+  if (length(between_idx) == 0) { # Fully within-subjects
+    df$subject <- 1
+  }
+  else {
+    # 3. Define unique between-subject combinations
+    between_vars <- paste0("X", between_idx)
+    df_between <- unique(df[between_vars])
+
+    # 4. Assign subject IDs per between-subject cell
+    df_between$subject <- seq_len(nrow(df_between))
+
+    # 5. Merge subject IDs back onto the full design
[…]

### 2025-12-20 sandbox.qmd
added several result subsections

--- 2025-12-12 sandbox.qmd
@@
 ```
+### Type I error rates in ordinal scales {#ordinal}
+We also evaluate Type I error rates for ordinal scales. We test again a $4 \times 3$ within-subjects design.
+
+**Main effects**. @fig-ordinal-main present Type I errors for $X_2$'s main effect, as we vary $X_1$'s magnitude of main effect.
+
+::: {#fig-ordinal-main}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Type_I_ordinal"
+
+distributions <- c("likert_5", "likert_5_flex", "likert_7", "likert_7_flex", "likert_11", "likert_11_flex")
+dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
+df <- readly_data(prefix, alpha, 1, distributions, dnames)
+
+plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 60)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
+:::
+
+Our results indicate that PAR, RNK, and INT consistently maintain error rates close to $5\%$ across all tested ordinal scales.
+In contrast, ART exhibits a significant inflation of error rates, although this issue becomes less severe when the number of levels increases.
+Notably, error rates are more pronounced for flexible thresholds and tend to increase with sample size.
+
+**Interaction effects**.
+@fig-ordinal-interaction-1 displays error rates for the interaction $X_1 \times X_2$ with a single main effect applied to $X_1$.
+The observed patterns align with our previous findings; ART consistently inflates error rates.
+
+::: {#fig-ordinal-interaction-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+
+plotly_error(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 50)
+```
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
+:::
+
+**Analysis for large samples.** Let us further explore ART's error rates when sample sizes become larger.
+As shown in @fig-ordinal-large-samples, error rates explode even when all population effects are null.
+The patterns are similar for all three effects.
+
+::: {#fig-ordinal-large-samples}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Type_I_4x3_ordinal_large_samples"
+
+df <- readly_data_1(prefix, alpha, 0, distributions, dnames)
+plotly_error_by_effect_2(df, xlab = "n", max = 100)
+```
+Type I error rates ($\alpha = .05$) as a function of the size $n$ of the sample, where all population effects are null.
+:::

### 2025-12-20 text/50-ExperimentalMethod.qmd
added several result subsections

@@
 
-**Thresholds of ordinal scales.** For ordinal data, we evaluate $D =$ 5, 7, or 11 discrete levels.
+**Thresholds for ordinal scales.** For ordinal data, we evaluate $D =$ 5, 7, or 11 discrete levels.
 To derive the discretization thresholds $\tau_d$, we first consider the range $[-2SD, 2SD]$ , where $SD$ is the overall standard deviation of the latent variable $Y^{*}_{ijk}$.
@@
 However, we can compare the power of different methods to evaluate their relative performance.
-In addition to power, we assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted on the latent variable $\mathcal{Y}$.
+For log-normal and ordinal distributions, we also assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted either on the logarithmic scale of the data (for log-normal) or latent variable $Y^*$ (for ordinal).
 While partial $\eta^2$ is the most commonly used effect size measure, we also evaluate Cohen's $f$, as its expected value is proportional to the real magnitude of effect.
@@
 
-As explained earlier, interaction effects are distorted when the latent variable $\mathcal{Y}$ is transformed to produce the observed responses.
-These transformations also influence the Type I error rates and statistical power we observe.
-In this analysis, we focus on evaluating these measures with respect to the effects applied to the latent variable $\mathcal{Y}$.
-We explicitly make this distinction whenever it is relevant to our discussion.
-
 ### Hardware platform and iterations

### 2025-12-20 text/60-Results-RatioScales.qmd
added several result subsections

@@
 ### Type I error rates in ratio scales {#ratio}
-We first evaluate Type I error rates for ratio scales. We test a 4 $\times$ 3 repeated-measures design, where we refer to the 4-level factor as $X_1$ and the 3-level factor as $X_2$.
+We first evaluate Type I error rates for ratio scales.
+We test a 4 $\times$ 3 within-subjects design, where we refer to the 4-level factor as $X_1$ and the 3-level factor as $X_2$.
+We encourage readers to use the interactive tools provided in the figures to explore the trends in greater detail.
 
@@
 
- The results show a very good behavior for RNK and INT across all distributions.
-The regular parametric ANOVA (PAR) keeps error rates below $5\%$.
-However, error rates become extremely low for some distributions, suggesting a loss in statistical power.
-We confirm previous results that ART fails to control the Type I error rate under the Cauchy distribution [@Mansouri:1995; @elkin:2021] but we also show that the method is problematic with other non-normal distributions.
-As the main effect on the first factor $X_1$ increases, Type I errors on the second factor $X_2$ grow and reach high levels.
-This pattern is particularly pronounced under the log-normal distribution.
-We also observe that for the binomial distribution, error rates are high ($\approx 11\%$ for $n = 20$) even when effects on $X_1$ are zero.
-In addition, error rates further grow when the sample size increases.
+Overall, RNK and INT exhibit very good behavior across all distributions.
+The standard parametric ANOVA (PAR) maintains error rates below $5\%$, but for some distributions (log-normal, Cauchy, and Poisson) the error rates are noticeably lower, suggesting a loss of statistical power.
+Consistent with previous findings [@Mansouri:1995; @elkin:2021], ART fails to control the Type I error rate under the Cauchy distribution.
+However, we further show that ART is also problematic for other non-normal distributions.
+As the main effect of the first factor $X_1$ increases, Type I errors on the second factor $X_2$ increase dramatically, approaching $100\%$ when $a_1=8$.
+For the binomial distribution, error rates are already elevated ($\approx 9\%$ for $n = 20$) even when $a_1=0$, and increase further as the sample size grows.
 
 **Contrasts**.
-The same problems appear when we run ART's procedure for contrasts [@elkin:2021].
-@fig-ratio-contrasts shows our results, where we report average error rates for three pairwise comparisons (since $X_2$ has three levels).
-In the rest of the paper, we only show results for overall effects, since results for contrasts exhibit the same patterns.
+The same issues arise when running ART's contrast procedure [@elkin:2021].
+@fig-ratio-contrasts shows the corresponding results, where we report average error rates for three pairwise comparisons (since $X_2$ has three levels).
+In the rest of the paper, we focus on overall effects, since the results for contrasts exhibit the same patterns.
 
@@
 **Interaction effects**.
-@fig-ratio-interaction-1 presents Type I error rates for the interaction effect $X_1 \times X_2$, when the main effect on $X_2$ is zero while the main effect on $X_1$ increases.
-Overall, we observe the same trends as for main effects.
-Again, ART fails in similar ways, although its error rates are now slightly lower.
+@fig-ratio-interaction-1 presents Type I error rates for the $X_1 \times X_2$ interaction when $a_2=0$ and $a_1$ increases.
+Overall, we observe trends similar to those for main effects.
+ART again fails in comparable ways, although its error rates are slightly lower in this case.
+In addition, PAR tends to inflate error rates under the log-normal and exponential distributions as $a_1$ increases.
 
@@
 
-The scenarios above do not capture the failure cases of the rank transformation discussed earlier (see @fig-interactions-rank), as those arise only when both coefficients $a_1$ and $a_2$ are non-zero.
-Since there can be different possible definitions of an interaction effect in these cases, we address them in a separate subsection.
+The scenarios above do not capture the failure cases of the rank transformation discussed earlier (see @fig-interactions-rank), as these arise only when both coefficients $a_1$ and $a_2$ are non-zero.
+Because multiple definitions of interaction effects are possible in such cases, we address them in a separate subsection.
 
-**Analysis for large samples.** Let us further explore ART's error rates when sample sizes become larger.
+**Analysis for large samples.** We also examine Type I error rates for larger sample sizes (up to $n=512$), fixing $a_1=1$.
+As shown in @fig-ratio-large-samples, ART's error rates increase only slightly for log-normal data but rise substantially under the Cauchy, Poisson, and binomial distributions.
+These patters are consistent for both the main effect of $X_2$ and the interaction effect.
+As we later observe for ordinal scales, ART's tendency for increasing Type I error rates with sample size appears to be a more general issue for discrete distributions.
 

### 2025-12-20 text/62-Results-AcrossExpDesigns.qmd
added several result subsections

@@
 ### Type I errors across experimental designs {#extra}
-We investigate the generalizability of the aforementioned results to other experimental designs involving two or three factors.
-We assess five out of the six ratio scales examined earlier.
-We exclude the Cauchy distribution and replace it by a 5-level ordinal scale with flexible thresholds.
+We examine the generalizability of the preceding results across alternative experimental designs involving two or three factors.
+We assess five of the six ratio-scale distributions considered earlier, excluding the Cauchy distribution and replacing it with a 5-level ordinal scale with flexible thresholds.
 
 **Main effects**.
-@fig-designs-main illustrates Type I errors for the main effect of $X_2$ while varying the magnitude of the main effect of $X_1$, with a focus on $n = 20$.
-Across all cases, RNK and INT consistently maintain error rates close to $5\%$.
-PAR's error rate remains overall close to $5\%$ but reaches notably low levels for specific combinations of designs ($2 \times 2 \times 2$ and $3 \times 3 \times 3$ repeated-measures) and distributions (log-normal and exponential).
-In contrast, ART inflates error rates for all non-normal distributions, with discrepancies across different designs.
-We observe that ART is particularly problematic in discrete distributions when applied to the $2 \times 2 \times 2$ and $3 \times 3 \times 3$ designs.
+@fig-designs-main illustrates Type I error rates for the main effect of $X_2$ while varying the magnitude of the main effect of $X_1$, focusing on $n = 20$.
+RNK and INT consistently maintain error rates close to $5\%$, with the exception of the Poisson distribution, for which error rates become low in the two within-subjects designs (particularly for INT).
+Similarly, PAR generally maintains error rates near $5\%$, but they drop to notably low levels for certain combinations of designs ($2 \times 2 \times 2$ and $3 \times 3 \times 3$ within-subjects) and distributions (log-normal and exponential).
+
+In contrast, ART inflates error rates across all non-normal distributions, with the extent of inflation varying by design.
+For example, ART performs particularly poorly for the ordinal scale when applied to the $2 \times 2 \times 2$ and $3 \times 3 \times 3$ designs.
+In these designs, we also observe inflated error rates for log-normal data, even when $a_1 = 0$.
 

### 2025-12-20 text/66-Results-Interactions.qmd
added several result subsections

@@
 Let us now examine interaction effects when both coefficients $a_1$ and $a_2$ increase in parallel.
-In this scenario, the interpretation of an interaction effect can become ambiguous.
-Here, we define the null hypothesis in terms of the coefficients of the latent linear model, following the recommendation of @Greene:2010 --- that is, $a_{12} = 0$.
+In this scenario, the interpretation of an interaction effect becomes ambiguous in nonlinear models.
+Here, we define the null hypothesis in terms of the coefficients of the linear component of the model (see @eq-linear-model), following the recommendation of @Greene:2010 --- that is, $a_{12} = 0$.
 
@@
 
-Error rates become exceptionally high in some cases, reaching up to 100%.
-However, these results require careful interpretation.
-As discussed in [Section 4](#interpretation), interaction effects in nonlinear models can be interpreted in two distinct ways, and the results from the different methods are not consistent with our definition.
-Let us examine the outcomes for each method:
+Error rates become exceptionally high in many cases, reaching up to 100%.
+These results require careful interpretation.
+As discussed in [Section 4](#interpretation), interaction effects in nonlinear models can be interpreted in two distinct ways, and the results produced by the different methods are not alwyas consistent with our chosen definition.
+We therefore examine the outcomes for each method in turn:
 
 - *PAR*.
-Error rates are high for all non-normal distributions and become even higher when the sample size increases.
-However, this behavior clearly stems from the method's sensitivity to the scale of the observations, rather than to the model parameters.
+Error rates are high for all non-normal distributions (with the exception of the Cauchy distribution) and increase further as sample size grows.
+This behavior arises because the null hypothesis for this method is defined on the scale of the response variable, rather than on the scale of the linear model.
 
-- *RNK.* Error rates explode when both main effects exceed a certain level (e.g., when $a_1, a_2 \ge 4$ and $n = 20$).
-This problem arises from the way the rank transformation distorts interaction effects (see @fig-interactions-rank).
-For all continuous distributions, we observe that error rates exhibit the exact same patterns.
+- *RNK.* Error rates increase dramatically once main effects exceed a certain magnitude.
+This well-known issue stems from the way rank transformations distort interaction effects (see @fig-interactions-rank), even when the underlying distributions are normal.
 
-- *INT.* It performs better than RNK.
-Errors begin to increase across all distributions only when the main effects become large.
-An exception is the binomial distribution, for which the error rates of INT and RNK are similar.
+- *INT.* It performs better than RNK in continuous distributions, with errors rates increasing only when $a_1$ and $a_2$ become large.
+However, under the Poisson and binomial distributions, INT performs worse than RNK.
+Interestingly, INT is particularly robust under the Cauchy distribution.
 
-- *ART.* It maintains correct error rates as long as population distributions are normal.
+- *ART.* ART maintains correct error rates only when the population distributions are normal.
 For all other distributions, error rates increase rapidly as effect sizes grow.
-Two factors explain this behavior: (i) a lack of statistical robustness to violation of the method's assumptions, as also observed in our previous experiments; and (ii) interaction interpretation issues.
-These results confirm that ART is not scale-invariant, interpreting interactions relative to the scale of the observed data.
+Two factors may account for this behavior: (i) a lack of robustness to violations of the method’s assumptions, as also observed in our earlier experiments; and (ii) issues related to interaction interpretation.
+These findings confirm that ART is not scale-invariant and interprets interactions relative to the scale of the observed data.
 
@@
 
-**Other designs.** Our findings in @fig-designs-interaction-2 further demonstrate that all methods struggle to accurately detect interactions for different reasons.
-Interestingly, INT shows unusually low error rates in certain configurations (e.g., $2 \times 3$ and $2 \times 2 \times 2$), indicating potential difficulty in detecting subtle interaction effects when main effects are large.
-RNK follows a similar trend for the $2 \times 2 \times 2$ design.
-Both RNK and INT also exhibit higher error rates under ordinal and binomial scales, though these remain lower than those observed for PAR and ART.
+**Other designs.** Our findings in @fig-designs-interaction-2 further demonstrate that all methods struggle to accurately detect interactions.
+INT shows unusually low error rates in certain configurations (e.g., $2 \times 3$ and $2 \times 2 \times 2$), suggesting potential difficulty in detecting subtle interaction effects when main effects are large.
+RNK displays a similar pattern for the $2 \times 2 \times 2$ design.
+In addition, both RNK and INT exhibit higher error rates under discrete distributions, in some cases exceeding those observed for ART.
 
@@
 
-Some readers may argue that evaluating PAR and ART using an interaction definition that accounts for the scale of the observed variable might be more appropriate.
-We offer three responses to this criticism:
-
-1. It is unclear how to reliably control the null hypothesis at the level of the response scale when both main effects are varied.
-
-2.
-Our earlier experiments demonstrate that ART fails to maintain nominal Type I error rates for interactions even when only a single main effect is present.
-Introducing a second main effect could only exacerbate the problem.
-
-3.
-The purpose of these results is not to provide precise benchmarks, but rather to illustrate the complexity and pitfalls of interpreting interaction effects under assumption violations --- regardless of the method used.
-Many researchers are unaware of these issues and often misinterpret the $p$-value when testing interactions.
-Note that @elkin:2021 compared ART to ANOVA on log-transformed data under the belief that ART tests null hypotheses about parameters of the latent linear model.
-Our findings clearly show that this is not the case.
+These results illustrate the complexity and pitfalls of interpreting interaction effects under assumption violations --- regardless of the method used.

### 2025-12-20 text/50-ExperimentalMethod.qmd
and some changes

@@
 
-**Linear component.** The base of our data generation process for all distributions is the following linear term:
+**Linear component.** The base of our data generation process for all distributions is the following linear predictor:
 
@@
 
-**Generating ratio data.** We adopt generalized linear modeling [@Bolker2009glm] for ratio data, where we use an inverse link function $g^{-1}$ to make the connection between $\ell_{ijk}$ and the location parameter of the distribution of interest:
+**Generating ratio data.** We generate ratio-scaled data using generalized linear modeling [@Bolker2009glm].
+Specifically, we use an inverse link function $g^{-1}$ to relate the linear predictor $\ell_{ijk}$ to the location parameter of the target distribution:
 $$
@@
 
-where $\mathcal{F}$ denotes the probability density (or mass) function from which we draw the responses $Y_{ijk}$.
+where $\mathcal{F}$ denotes the probability density (or mass) function from which the responses $Y_{ijk}$ are drawn.
 For example, the mean of the exponential distribution is equal to $1 / \lambda$, where $\lambda$ is its rate parameter.
-This mean is linked to the linear term through the function $g^{-1}(x) = e^{x}$, and therefore, $\lambda = e^{-\ell_{ijk}}$.
-Or the link function commonly used for the binomial distribution is the *logit* function, but its inverse if the *logistic* function $g^{-1}(x) = 1 / (1 + e^{-x})$.
+This mean is linked to the linear predictor through the inverse link function $g^{-1}(x) = e^{x}$, which implies $\lambda = e^{-\ell_{ijk}}$.
+Similarly, for the binomial distribution, the canonical link function is the *logit* function, whose inverse is the *logistic* function $g^{-1}(x) = 1 / (1 + e^{-x})$.
 
- @tbl-distributions presents how the location parameters of the other distributions are linked to the linear term.
-We emphasize that the Cauchy distribution does not define a GLM (since it does not belong to the exponential family), but we use a model that connects its location (median) through the identity link.
+ @tbl-distributions summarizes how the location parameters of all the distributions are linked to the linear predictor.
+We emphasize that the Cauchy distribution does not define a generalized linear model  --- because it does not belong to the exponential family --- but we nevertheless use a model that links its location (the median) to the linear predictor via the identity link.
 
@@
 
-: Distribution parameters and their links to the linear term $\ell_{ijk}$ {#tbl-distributions}
+: Distribution parameters and their links to the linear predictor $\ell_{ijk}$ {#tbl-distributions}
 

### 2025-12-20 sandbox.qmd
results on unequal variances completed

@@
 ```
-### Type I error rates in ordinal scales {#ordinal}
-We also evaluate Type I error rates for ordinal scales. We test again a $4 \times 3$ within-subjects design.
+### Type I errors under unequal variances
+Many statistical procedures assume equal variances across all levels of each independent variable, or more strictly, across all possible pairs of factor levels (the latter is known as the sphericity assumption in repeated-measures ANOVA).
+Nonparametric tests are often mistakenly assumed to be free of such assumptions; however, as discussed earlier, this is generally incorrect.
 
-**Main effects**. @fig-ordinal-main present Type I errors for $X_2$'s main effect, as we vary $X_1$'s magnitude of main effect.
+Our next experiment evaluates the behavior of the four methods under conditions of unequal variances, focusing on the normal distribution and four ordinal distributions.
+For ordinal outcomes, we control variances at the level of the latent variable $Y^*$.
 
-::: {#fig-ordinal-main}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Type_I_ordinal"
+We examine three two-factor designs: (i) a $4 \times 3$ within-subjects design; (ii) a $2 \times 3$ between-subjects design; and (iii) a $2 \times 4$ mixed design.
+We set all effects to zero and vary the ratio $r_{sd}$ of standard deviations between the levels of the first factor $X_1$.
+@fig-unequal shows the five ratios tested when $X_1$ has two levels.
+When $X_1$ has additional levels (as in the $4 \times 3$ design), their standard deviations are randomly drawn from a range bounded by the standard deviations of the first two levels.
 
-distributions <- c("likert_5", "likert_5_flex", "likert_7", "likert_7_flex", "likert_11", "likert_11_flex")
-dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
-df <- readly_data(prefix, alpha, 1, distributions, dnames)
-
-plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 60)
+::: {#fig-unequal}
+```{r, echo=FALSE, message=FALSE, fig.height=1.5, fig.width = 9, warning=FALSE}
+source("distributions_plots.R")
+plot_unequal_variances()
 ```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
+Varying the ratio $r_{sd}$ of standard deviations across the levels of $X_1$. The plots show examples of population distributions for factors with two categorical levels.    
 :::
 
-Our results indicate that PAR, RNK, and INT consistently maintain error rates close to $5\%$ across all tested ordinal scales.
-In contrast, ART exhibits a significant inflation of error rates, although this issue becomes less severe when the number of levels increases.
-Notably, error rates are more pronounced for flexible thresholds and tend to increase with sample size.
+**Main effects**.
+We begin by investigating how the four methods detect main effects of $X_1$.
+Interpreting results in this setting is challenging because the underlying populations differ in variance while sharing the same means (and medians).
+For ordinal scales, differences in variance may also manifest as differences in observed means [@Liddell:2018].
+In this experiment, we define the null hypothesis in terms of mean differences on the latent variable, but we emphasize that the results should be interpreted with caution.
+
+@fig-unequal-main-1 presents the results.
+Under normal distributions, and depending on the experimental design, the different methods either slightly inflate or slightly deflate Type I error rates as $r_{sd}$ increases, with no method clearly outperforming the others.
+For ordinal scales, larger values of $r_{sd}$ primarily affect conditions with flexible thresholds.
+This effect is particularly pronounced for PAR, although all methods are affected.
+
+::: {#fig-unequal-main-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Type_I_heteroscedacity"
+distributions = c("norm", "likert_5", "likert_5_flex", "likert_7", "likert_7_flex")
+dnames = c("Normal", "Equidistant (5 levels)", "Flexible (5 levels)", "Equidistant (7 levels)", "Flexible (7 levels)")
+
+df <- read_data(prefix, alpha = 0.05, effectType = 1, distributions)
+df <- reshape_by_design(df, dnames, effectvars = c("sdratio", "effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1", xvar = "sdratio", max = 41)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as the ratio of standard deviations $r_{sd}$ for $X_1$ increases ($n=20$).
+When $r_{sd} > 1$, the underlying populations are no longer identical, and different interpretations of the null hypothesis may lead to different outcomes.
+:::
+
+We also examine the influence of unequal variances across the levels of $X_1$ on the Type I error rate for $X_2$, as shown in @fig-unequal-main-2.
+In this case, there is no ambiguity in the interpretation of effects because the populations defined by $X_2$ are identical.
+
+The Type I error rates of PAR, RNK, and INT appear largely unaffected by increases in $r_{sd}$.
+In contrast, ART yields high error rates, surpassing acceptable levels even under normal distributions (see, for example, the $2 \times 4$ mixed design).
+Thus, unequal variances pose a more serious problem for ART than for the other methods, as the method's sensitivity propagates across factors.
+This provides further evidence that ART's alignment procedure leads to confounding across effects.
+
+::: {#fig-unequal-main-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX2", xvar = "sdratio", max = 41)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as the ratio of standard deviations $r_{sd}$ across the levels of $X_1$ increases ($n=20$)
+:::
 
 **Interaction effects**.
-@fig-ordinal-interaction-1 displays error rates for the interaction $X_1 \times X_2$ with a single main effect applied to $X_1$.
-The observed patterns align with our previous findings; ART consistently inflates error rates.
+Finally, we evaluate Type I error rates for the interaction, as shown in @fig-unequal-interaction.
+ART exhibits a similar pattern as before, with error rates increasing as $r_{sd}$ grows.
+We also observe some inflation of Type I error rates for the other three methods under the $4 \times 3$ within-subjects design, albeit to a significantly lesser degree.
 
-::: {#fig-ordinal-interaction-1}
+::: {#fig-unequal-interaction}
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-
-plotly_error(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 50)
+plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1X2", xvar = "sdratio", max = 42)
 ```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
+Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as the ratio of standard deviations $r_{sd}$ across the levels of $X_1$ increases ($n=20$)
 :::
-
-**Analysis for large samples.** Let us further explore ART's error rates when sample sizes become larger.
-As shown in @fig-ordinal-large-samples, error rates explode even when all population effects are null.
-The patterns are similar for all three effects.
-
-::: {#fig-ordinal-large-samples}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Type_I_4x3_ordinal_large_samples"
-
-df <- readly_data_1(prefix, alpha, 0, distributions, dnames)
-plotly_error_by_effect_2(df, xlab = "n", max = 100)
-```
-Type I error rates ($\alpha = .05$) as a function of the size $n$ of the sample, where all population effects are null.
-:::

### 2025-12-20 text/50-ExperimentalMethod.qmd
results on unequal variances completed

@@
 ```{r, echo=FALSE, message=FALSE, fig.height=1.2, fig.width = 9, warning=FALSE}
-source("distributions_plots.R")
 plot_effects_distr()

### 2025-12-20 text/63-Results-UnequalVariance.qmd
results on unequal variances completed

@@
 ### Type I errors under unequal variances
-Many statistical procedures assume equal variances among all levels of each independent variable, or more strictly, among all possible pairs of independent variable levels (known as sphericity assumption in repeated-measures ANOVA).
-Nonparametric tests are often mistakenly considered to be free of such as assumptions, but as we discussed earlier, this is generally incorrect.
+Many statistical procedures assume equal variances across all levels of each independent variable, or more strictly, across all possible pairs of factor levels (the latter is known as the sphericity assumption in repeated-measures ANOVA).
+Nonparametric tests are often mistakenly assumed to be free of such assumptions; however, as discussed earlier, this is generally incorrect.
 
-Our next experiment evaluates the behavior of the four methods when populations on the latent variable $\mathcal{Y}$ have unequal variances.
-As we previously discussed, even when variances are equal at the latent space, they may appear as unequal on the response scale if the response distributions are skewed.
-This is the case, for example, with the log-normal and the exponential distributions.
-So, in this experiment, we explore more extreme violations of the equal-variance assumption for these distributions.
+Our next experiment evaluates the behavior of the four methods under conditions of unequal variances, focusing on the normal distribution and four ordinal distributions.
+For ordinal outcomes, we control variances at the level of the latent variable $Y^*$.
 
-We examine three 2-factor designs: (i) a $4 \times 3$ within-subjects design; (ii) a $2 \times 3$ between-subjects design; and (iii) a $2 \times 4$ mixed design.
-We set all effects to zero and then vary the ratio $r_{sd}$ of standard deviations between the levels of the first factor $X_1$.
-@fig-unequal shows the five ratios that we test when $X_1$ has two levels.
-When $X_1$ has additional levels ($4 \times 3$ design), their standard deviation is randomly drawn in a range between the standard deviations of the first two levels.
+We examine three two-factor designs: (i) a $4 \times 3$ within-subjects design; (ii) a $2 \times 3$ between-subjects design; and (iii) a $2 \times 4$ mixed design.
+We set all effects to zero and vary the ratio $r_{sd}$ of standard deviations between the levels of the first factor $X_1$.
+@fig-unequal shows the five ratios tested when $X_1$ has two levels.
+When $X_1$ has additional levels (as in the $4 \times 3$ design), their standard deviations are randomly drawn from a range bounded by the standard deviations of the first two levels.
 
@@
 ```{r, echo=FALSE, message=FALSE, fig.height=1.5, fig.width = 9, warning=FALSE}
-plotVariances()
+source("distributions_plots.R")
+plot_unequal_variances()
 ```
-Varying the ratio $r_{sd}$ of standard deviations between the levels of $X_1$. The plots show examples of population distributions for factors with two categorical levels.    
+Varying the ratio $r_{sd}$ of standard deviations across the levels of $X_1$. The plots show examples of population distributions for factors with two categorical levels.    
 :::
@@
 **Main effects**.
-We begin by investigating how the four methods detect main effects on $X_1$.
-The interpretation of results in this scenario is challenging due to differences in the original populations --- although they share the same means (and medians), their variances differ.
-Depending on the null hypothesis of interest, conclusions may vary.
-Complicating matters further, non-linear transformations can result in distributions with differing means and medians.
-Consequently, the choice of statistical method may lead to different outcomes; one method may be sensitive to mean differences while others to median differences.
+We begin by investigating how the four methods detect main effects of $X_1$.
+Interpreting results in this setting is challenging because the underlying populations differ in variance while sharing the same means (and medians).
+For ordinal scales, differences in variance may also manifest as differences in observed means [@Liddell:2018].
+In this experiment, we define the null hypothesis in terms of mean differences on the latent variable, but we emphasize that the results should be interpreted with caution.
 
-@fig-unequal-main-1 presents the percentage of times each method rejects the null hypothesis ($\%$ positives) at a significance level of $\alpha = .05$.
-If the null hypothesis of interest posits that *"$X_1$ has no effect on the mean of the latent variable $Y$,"* then these instances should be considered as Type I errors.
-Under normal distributions, all methods either moderately inflate or deflate the positives ratio as $r_{sd}$ increases.
-For the $4 \times 3$ within-subjects design ($r_{sd} \ge 2.5$), the ratio for PAR, RNK, and ART reaches levels of around $7-8\%$, while for INT, it drops below $3\%$ under the $2 \times 4$ mixed design ($r_{sd} \ge 2.5$).
+@fig-unequal-main-1 presents the results.
+Under normal distributions, and depending on the experimental design, the different methods either slightly inflate or slightly deflate Type I error rates as $r_{sd}$ increases, with no method clearly outperforming the others.
+For ordinal scales, larger values of $r_{sd}$ primarily affect conditions with flexible thresholds.
+This effect is particularly pronounced for PAR, although all methods are affected.
 
@@
 ```
-The percentage of positives (i.e., rejecting the null hypothesis) for the **main effect of $X_1$** as the ratio of standard deviations $r_{sd}$ on $X_1$ increases ($n=20$).
-Depending on the hypothesis of interest, these percentages can be interpreted as Type I error rates ($\alpha = .05$).
+Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as the ratio of standard deviations $r_{sd}$ for $X_1$ increases ($n=20$).
+When $r_{sd} > 1$, the underlying populations are no longer identical, and different interpretations of the null hypothesis may lead to different outcomes.
 :::
 
-The performance of RNK and INT remains unchanged across continuous distributions, a result expected due to the preservation of medians in such scales.
-Conversely, mean differences grow with increasing $r_{sd}$, which explains why PAR's ratio of positives reaches high levels under these distributions.
-However, this trend is not consistent across all three designs.
-The behavior of ART is less clear, further supporting our argument that its null hypothesis of interest is ill-defined.
-Under discrete distributions, the trends of positives vary across designs, with PAR and ART exhibiting the highest rates.
+We also examine the influence of unequal variances across the levels of $X_1$ on the Type I error rate for $X_2$, as shown in @fig-unequal-main-2.
+In this case, there is no ambiguity in the interpretation of effects because the populations defined by $X_2$ are identical.
 
-We also investigate the influence of unequal variances among the levels of $X_1$ on positive rates for $X_2$, as depicted in @fig-unequal-main-2.
-These can be reliably considered as Type I errors because source populations defined by $X_2$ are identical.
-The error rates of PAR, RNK, and INT do not seem to be affected by an increase of $r_{sd}$.
-In contrast, ART yields high error rates, surpassing acceptable levels even in the case of normal distributions (check results for the $2 \times 4$ mixed design).
-Thus, the problem of unequal variance is more serious for ART than for other methods, as the method's sensitivity propagates across factors.
-This further demonstrates that ART's alignment procedure causes the method to confound effects.
+The Type I error rates of PAR, RNK, and INT appear largely unaffected by increases in $r_{sd}$.
+In contrast, ART yields high error rates, surpassing acceptable levels even under normal distributions (see, for example, the $2 \times 4$ mixed design).
+Thus, unequal variances pose a more serious problem for ART than for the other methods, as the method's sensitivity propagates across factors.
+This provides further evidence that ART's alignment procedure leads to confounding across effects.
 
@@
 ```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as the ratio of standard deviations $r_{sd}$ on $X_1$ increases ($n=20$)
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as the ratio of standard deviations $r_{sd}$ across the levels of $X_1$ increases ($n=20$)
 :::
@@
 **Interaction effects**.
-Finally, we measure Type I error rates for the interaction effect $X_1 \times X_2$, shown in @fig-unequal-interaction.
-The error rates of ART exhibit similar trends, with a somewhat lower level under the $4 \times 3$ within-subjects design (compared to @fig-unequal-main-2).
-Additionally, for this design, we observe error inflation for the three other methods, albeit to a significantly lesser degree.
+Finally, we evaluate Type I error rates for the interaction, as shown in @fig-unequal-interaction.
+ART exhibits a similar pattern as before, with error rates increasing as $r_{sd}$ grows.
+We also observe some inflation of Type I error rates for the other three methods under the $4 \times 3$ within-subjects design, albeit to a significantly lesser degree.
 
@@
 ```
-Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as the ratio of standard deviations $r_{sd}$ on $X_1$ increases ($n=20$)
+Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as the ratio of standard deviations $r_{sd}$ across the levels of $X_1$ increases ($n=20$)
 :::

### 2025-12-20 sandbox.qmd
added power results

--- 2025-12-20 sandbox.qmd
@@
 ```
-### Type I error rates in ordinal scales {#ordinal}
-We also evaluate Type I error rates for ordinal scales. We test again a $4 \times 3$ within-subjects design.
 
-**Main effects**. @fig-ordinal-main present Type I errors for $X_2$'s main effect, as we vary $X_1$'s magnitude of main effect.
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+library(dplyr)
+library(tidyverse)
+source("utils-data-reading.R")
+source("utils-plotlying.R")
+```
 
-::: {#fig-ordinal-main}
+### Statistical power {#power}
+We evaluate the statistical power of the four methods on the three two-factor designs.
+Because there is a tradeoff between Type I and Type II errors, high power may simply reflect an inflated Type I error rate.
+Since a main effect on $X_1$ inflates ART's Type I error rates for $X_2$ and for their interaction (see previous sections), we therefore focus on single-effect scenarios.
+
+**Main effects**.
+We vary the magnitude of effect on $X_1$ and $X_2$ separately and assess each method's power to detect the corresponding effect at $\alpha = .05$.
+Specifically, we manipulate the parameters $a_1$ or $a_2$ over the range $0.2$ to $0.8$.
+Depending on the experimental design, this range allows us to simulate both low-power settings (e.g., power below $30\%$) and high-power settings (e.g., power above $80\%$).
+
+@fig-power-main-1 presents the results for $X_1$.
+In many configurations, differences in power across methods are modest.
+To facilitate comparison, we also report method rankings; readers are encouraged to interact with the figures for precise power estimates.
+
+PAR achieves the highest power under the normal and exponential distributions, but its power is particularly low under the log-normal distribution.
+ART initially appears advantageous under the the log-normal, binomial, and ordinal distributions for small effects, likely reflecting inflated Type I errors under these conditions.
+As $a_1$ grows larger, ART's relative advantage diminishes, and it ultimately becomes the least powerful method.
+Finally, INT consistently outperforms RNK and ranks highly across all conditions.
+
+::: {#fig-power-main-1}
+<div id="plot-power-1.1">
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Type_I_ordinal"
+prefix <- "Power"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
 
-distributions <- c("likert_5", "likert_5_flex", "likert_7", "likert_7_flex", "likert_11", "likert_11_flex")
-dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
-df <- readly_data(prefix, alpha, 1, distributions, dnames)
+data <- read_data(prefix, alpha = .05, effectType = 3, distributions)
+df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX1", max = 104,  ytitle = 'Power (%)')
+```
+</div>
 
-plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 60)
+<div id="plot-power-1.2">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- data %>% arrange(design,family,effectX1,rateX1)  %>% group_by(design,family,effectX1) %>% mutate(rank = rank(rateX1))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power - ranking')
 ```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
+</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-1.2").classList.add("hidden"); }); </script>
+
+<button id="btn1" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
+    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
+  onclick="const plot = document.getElementById('plot-power-1.1');
+  plot.classList.toggle('hidden');
+  document.getElementById('plot-power-1.2').classList.toggle('hidden');
+  btn1.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
+">
+  show ranking
+</button>
+
+Power ($\alpha = .05$) for detecting the **main effect of $X_1$** as a function of its magnitude $a_1$, with $n=20$.
 :::
 
-Our results indicate that PAR, RNK, and INT consistently maintain error rates close to $5\%$ across all tested ordinal scales.
-In contrast, ART exhibits a significant inflation of error rates, although this issue becomes less severe when the number of levels increases.
-Notably, error rates are more pronounced for flexible thresholds and tend to increase with sample size.
+@fig-power-main-2 presents our results for $X_2$.
+Because this factor has a different number of levels across designs, we assess different ranges of power for each design.
+Nevertheless, the same overall patterns emerge.
+
+::: {#fig-power-main-2}
+<div id="plot-power-2.1">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+data <- read_data(prefix, alpha=.05, effectType = 4, distributions) # CHANGE n TO 20
+
+df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX2", max = 104,  ytitle = 'Power (%)')
+```
+</div>
+
+<div id="plot-power-2.2">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- data %>% arrange(design,family,effectX2,rateX2)  %>% group_by(design,family,effectX2) %>% mutate(rank = rank(rateX2))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+#df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power - ranking')
+```
+</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-2.2").classList.add("hidden"); }); </script>
+
+<button id="btn2" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
+    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
+  onclick="const plot = document.getElementById('plot-power-2.1');
+  plot.classList.toggle('hidden');
+  document.getElementById('plot-power-2.2').classList.toggle('hidden');
+  btn2.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
+">
+  show ranking
+</button>
+
+Power ($\alpha = .05$) for detecting the **main effect of $X_2$** as a function of its magnitude $a_2$, with $n=20$. 
+:::
 
 **Interaction effects**.
-@fig-ordinal-interaction-1 displays error rates for the interaction $X_1 \times X_2$ with a single main effect applied to $X_1$.
-The observed patterns align with our previous findings; ART consistently inflates error rates.
+We also vary the magnitude of the interaction effect $X_1 \times X_2$ by manipulating the parameter $a_{12}$ over the range $0.5$ to $2.0$, and then measure the power to detect this interaction.
+The results, summarized in @fig-power-interaction, are consistent with the patterns observed for main effects.
 
-::: {#fig-ordinal-interaction-1}
+::: {#fig-power-interaction}
+<div id="plot-power-3.1">
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+data <- read_data(prefix, alpha=.05, effectType = 5, distributions)
 
-plotly_error(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 50)
+df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of interaction effect", var = "rateX1X2", xvar = "effectX1X2", max = 104,  ytitle = 'Power (%)')
 ```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
+</div>
+
+<div id="plot-power-3.2">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- data %>% arrange(design,family,effectX1X2,rateX1X2)  %>% group_by(design,family,effectX1X2) %>% mutate(rank = rank(rateX1X2))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+#df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power - ranking')
+```
+</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-3.2").classList.add("hidden"); }); </script>
+
+<button id="btn3" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
+    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
+  onclick="const plot = document.getElementById('plot-power-3.1');
+  plot.classList.toggle('hidden');
+  document.getElementById('plot-power-3.2').classList.toggle('hidden');
+  btn3.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
+">
+  show ranking
+</button>
+
+Power ($\alpha = .05$) for detecting the **interaction effect $X_1 \times X_2$** as a function of its magnitude $a_{12}$, with $n=20$.
 :::
 
-**Analysis for large samples.** Let us further explore ART's error rates when sample sizes become larger.
-As shown in @fig-ordinal-large-samples, error rates explode even when all population effects are null.
-The patterns are similar for all three effects.
-
-::: {#fig-ordinal-large-samples}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Type_I_4x3_ordinal_large_samples"
-
-df <- readly_data_1(prefix, alpha, 0, distributions, dnames)
-plotly_error_by_effect_2(df, xlab = "n", max = 100)
-```
-Type I error rates ($\alpha = .05$) as a function of the size $n$ of the sample, where all population effects are null.
-:::
+Overall, there is no clear winner, but we observe that PAR shows poor power performance for log-normal data, while INT consistently outperforms RNK.
+Furthermore, ART provides no substantial power advantage, despite its inflated Type I error rates.

### 2025-12-20 text/64-Results-StatPower.qmd
added power results

@@
 ### Statistical power {#power}
-Our sixth experiment compares the statistical power of the four methods on the three 2-factor designs.
-Because there is a tradeoff between Type I and Type II errors, high power can simply be the result of a high Type I error rate.
-Since parallel effects can inflate errors (see our previous results), we focus here on single effects, both main or interaction effects.
-We only report results for $n=20$ (see supplementary materials for additional results).
+We evaluate the statistical power of the four methods on the three two-factor designs.
+Because there is a tradeoff between Type I and Type II errors, high power may simply reflect an inflated Type I error rate.
+Since a main effect on $X_1$ inflates ART's Type I error rates for $X_2$ and for their interaction (see previous sections), we therefore focus on single-effect scenarios.
 
 **Main effects**.
-We individually vary the magnitude of effect on $X_1$ and $X_2$ and observe the power of each method to detect this effect ($\alpha = .05$).
-Specifically, we manipulate the parameters $a_1$ or $a_2$ within the range of $0.4$ to $1.0$.
-Depending on the experimental design, this range allows us to simulate both low-power experiments (e.g., with less than $30\%$ power) and high-power experiments (e.g., with more than $80\%$ power).
+We vary the magnitude of effect on $X_1$ and $X_2$ separately and assess each method's power to detect the corresponding effect at $\alpha = .05$.
+Specifically, we manipulate the parameters $a_1$ or $a_2$ over the range $0.2$ to $0.8$.
+Depending on the experimental design, this range allows us to simulate both low-power settings (e.g., power below $30\%$) and high-power settings (e.g., power above $80\%$).
 
-@fig-power-main-1 presents our results for $X_1$.
-In several configurations, the differences in power among the methods are marginal.
-To enhance clarity in our comparisons, we also present the method rankings.
-Readers are encouraged to interact with the graphs for precise power values.
-PAR exhibits the highest power under the normal distribution, closely followed by INT.
-INT is also consistently the winning method across all non-normal distributions.
-PAR's power is notably low under the log-normal and exponential distribution.
+@fig-power-main-1 presents the results for $X_1$.
+In many configurations, differences in power across methods are modest.
+To facilitate comparison, we also report method rankings; readers are encouraged to interact with the figures for precise power estimates.
 
-ART initially appears advantageous under the three discrete distributions for small effects, likely due to inflated Type I errors even without an effect on $X_2$ (as observed in @fig-2x3-main).
-However, as effects grow larger, its relative power diminishes, and ART eventually becomes the least effective method.
+PAR achieves the highest power under the normal and exponential distributions, but its power is particularly low under the log-normal distribution.
+ART initially appears advantageous under the the log-normal, binomial, and ordinal distributions for small effects, likely reflecting inflated Type I errors under these conditions.
+As $a_1$ grows larger, ART's relative advantage diminishes, and it ultimately becomes the least powerful method.
+Finally, INT consistently outperforms RNK and ranks highly across all conditions.
 
@@
 
-Power ($\alpha = .05$) for detecting the **main effect of $X_1$** as a function of its magnitude $a_1$, where $n=20$.
+Power ($\alpha = .05$) for detecting the **main effect of $X_1$** as a function of its magnitude $a_1$, with $n=20$.
 :::
@@
 @fig-power-main-2 presents our results for $X_2$.
-Given the different number of levels for this factor, we now assess a different range of power for each design.
-Nevertheless, we observe similar patterns.
-INT stands out as the most powerful method.
-Although ART surpasses PAR under the log-normal and exponential distributions, it does not demonstrate any other notable advantages compared to the other methods.
+Because this factor has a different number of levels across designs, we assess different ranges of power for each design.
+Nevertheless, the same overall patterns emerge.
 
@@
 
-Power ($\alpha = .05$) for detecting the **main effect of $X_2$** as a function of its magnitude $a_2$, where $n=20$. 
+Power ($\alpha = .05$) for detecting the **main effect of $X_2$** as a function of its magnitude $a_2$, with $n=20$. 
 :::
@@
 **Interaction effects**.
-We also vary the magnitude of the interaction effect $X_1 \times X_2$ by manipulating the parameter $a_{12}$ across the range of $0.5$ to $2.0$.
-We then measure the power to detect this interaction effect.
-Our results are summarized in @fig-power-interaction, aligning with findings for main effects.
-ART's relative advantage diminishes later under the binomial and ordinal distributions.
-Once again, the method's inflated Type I error rates could account for this observed trend.
+We also vary the magnitude of the interaction effect $X_1 \times X_2$ by manipulating the parameter $a_{12}$ over the range $0.5$ to $2.0$, and then measure the power to detect this interaction.
+The results, summarized in @fig-power-interaction, are consistent with the patterns observed for main effects.
 
@@
 
-Power ($\alpha = .05$) for detecting the **interaction effect $X_1 \times X_2$** as a function of its magnitude $a_{12}$, where $n=20$.
+Power ($\alpha = .05$) for detecting the **interaction effect $X_1 \times X_2$** as a function of its magnitude $a_{12}$, with $n=20$.
 :::
+
+Overall, there is no clear winner, but we observe that PAR shows poor power performance for log-normal data, while INT consistently outperforms RNK.
+Furthermore, ART provides no substantial power advantage, despite its inflated Type I error rates.

### 2025-12-22 sandbox.qmd
added effect sizes

--- 2025-12-20 sandbox.qmd
@@
 ```
-### Type I error rates in ordinal scales {#ordinal}
-We also evaluate Type I error rates for ordinal scales. We test again a $4 \times 3$ within-subjects design.
 
-**Main effects**. @fig-ordinal-main present Type I errors for $X_2$'s main effect, as we vary $X_1$'s magnitude of main effect.
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+library(dplyr)
+library(tidyverse)
+source("utils-data-reading.R")
+source("utils-plotlying.R")
+```
 
-::: {#fig-ordinal-main}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Type_I_ordinal"
+### Effect size estimates {#effect-sizes}
+Transformations --- even monotonic ones --- alter both the absolute magnitude of effects and the total variance of the data.
+A natural question, therefore, is how to interpret effect size estimates derived from rank transformations and whether such estimates are meaningful and comparable across methods.
+Nevertheless, @Kay:art-effect-sizes provides a tutorial for deriving standardized effect sizes from ART, and many of the papers we reviewed report such estimates.
+This practice motivates a more systematic investigation.
 
-distributions <- c("likert_5", "likert_5_flex", "likert_7", "likert_7_flex", "likert_11", "likert_11_flex")
-dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
-df <- readly_data(prefix, alpha, 1, distributions, dnames)
+**Log-normal distributions.** We begin by examining log-normal responses.
+Consider a $4 \times 3$ within-subjects design with $n = 20$, in which the effect of the first factor $X_1$ is null ($a_1 = 0$) and there is no interaction ($a_{12} = 0$).
+@fig-eta-squared displays the partial $\eta^2$ estimates produced by each method for the effect of $X_1$ as the magnitude of the effect on the second factor $X_2$ increases.
+These estimates are compared to those from a ground-truth analysis, defined here as ANOVA applied to the log-transformed responses.
 
-plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 60)
+::: {#fig-eta-squared}
+```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
+data <- read.csv("results/eta_squared.csv", sep=",", header=TRUE, strip.white=TRUE)
+data <- data[data$family == "lnorm", ]
+data$effectX2 <- factor(data$effectX2)
+
+ylab = "X1 - partial eta squared"
+xlab = "X2 - magnitude of main effect (a2)"
+
+margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
+cbPalette = c("#8888ff", "#888888", "#E69F00", "#009E73", "#FF5E00")
+cbPalette2 = c("#8888ff88", "#88888888", "#E69F0088", "#009E7388", "#FF5E0088")
+
+df <- data[data$effectX1 == 0,]
+
+fig1 <- plot_ly(df, x = ~effectX2, y = ~grtX1, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
+fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
+fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
+fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))
+
+fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
+  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
+	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab)) %>% layout(
+    hovermode = "x unified",  # Shared hover on the x-axis
+    yaxis = list(spikemode = "across", spikesnap = "cursor"),
+    xaxis = list(fixedrange=T)
+  )
+
+fig
 ```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
+Boxplots summarizing the $\eta^2$ estimates obtained with each method from 300 simulated datasets ($n=20$).
+The population effect of $X1$ is null, and responses follow **log-normal distributions**.
+Hover over the plots to compare the methods.
 :::
 
-Our results indicate that PAR, RNK, and INT consistently maintain error rates close to $5\%$ across all tested ordinal scales.
-In contrast, ART exhibits a significant inflation of error rates, although this issue becomes less severe when the number of levels increases.
-Notably, error rates are more pronounced for flexible thresholds and tend to increase with sample size.
+We observe that ART is the only method for which the estimated effect sizes for $X_1$ are influenced by the magnitude of the effect on $X_2$.
+These results clearly demonstrate that ART's alignment mechanism fails in nonlinear models with skewed distributions and leads to confounding across effects.
 
-**Interaction effects**.
-@fig-ordinal-interaction-1 displays error rates for the interaction $X_1 \times X_2$ with a single main effect applied to $X_1$.
-The observed patterns align with our previous findings; ART consistently inflates error rates.
+@fig-eta-squared-2 presents partial $\eta^2$ estimates when the effect of $X_1$ is no longer null ($a_1 = 1$).
+PAR performs poorly for these data, consistently underestimating effect sizes.
+RNK also slightly underestimates effects.
+ART also tends to underestimate effects when $a_2 = 0$; however, as $a_2$ increases, its estimates become more variable and exhibit an overall tendency to increase.
+INT yields the most precise estimates and is only mildly affected when $a_2 = 8$.
 
-::: {#fig-ordinal-interaction-1}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+::: {#fig-eta-squared-2}
+```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
+df <- data[data$effectX1 == 1,]
 
-plotly_error(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 50)
+fig1 <- plot_ly(df, x = ~effectX2, y = ~grtX1, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
+fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
+fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
+fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))
+
+fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
+  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
+	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab)) %>% layout(
+    hovermode = "x unified",  # Shared hover on the x-axis
+    yaxis = list(spikemode = "across", spikesnap = "cursor"),
+    xaxis = list(fixedrange=T)
+  )
+
+fig
 ```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
+Boxplots summarizing partial $\eta^2$ estimates obtained with each method from 300 simulated datasets ($n=20$).
+The population effect of $X1$ is non-null ($a_1 = 1$), and responses again follow **log-normal distributions**.
+Hover over the plots to compare the methods.
 :::
 
-**Analysis for large samples.** Let us further explore ART's error rates when sample sizes become larger.
-As shown in @fig-ordinal-large-samples, error rates explode even when all population effects are null.
-The patterns are similar for all three effects.
+**Systematic comparison to ground truth estimates**.
+We conduct an additional experiment to compare the methods to the ground truth across a wider range of effect magnitudes for the normal, log-normal, and ordinal distributions (the latter with five levels and flexible thresholds).
+For ordinal data, the ground truth is defined as ANOVA applied to the latent variable (i.e., the data prior to discretization).
+We now use Cohen’s $f$ as a measure of effect size, which is expected to be proportional to the true effect magnitude.
+Without loss of generality, we focus on the main effect of $X_1$ and fix the sample size to $n=20$.
 
-::: {#fig-ordinal-large-samples}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Type_I_4x3_ordinal_large_samples"
+@fig-effect-main-scatterplots-1 illustrates the relationship between the Cohen's $f$ estimated by each method and the corresponding ground-truth estimate, as $a_1$ and $a_2$ vary independently within the range $[-2, 2]$.
+This produces Cohen’s $f$ values between $0$ and $1$ for the $4 \times 3$ within-subject and $2 \times 3$ between-subject designs, and a wider range for the $2 \times 4$ mixed design.
+Proximity to the black diagonal indicates closer agreement with the ground truth.
+Points above the diagonal indicate overestimation of the effect size, whereas points below indicate underestimation.
 
-df <- readly_data_1(prefix, alpha, 0, distributions, dnames)
-plotly_error_by_effect_2(df, xlab = "n", max = 100)
+::: {#fig-effect-main-scatterplots-1}
+```{r, echo=FALSE, fig.height=3.4, message=FALSE, warning=FALSE}
+source("cohens_f_plots.R")
+prefix <- "cohens_f_2"
+df <- readly_datapoints(prefix,  distributions=c("norm", "lnorm", "likert"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)"))
+fig <- plotly_scatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "grtX1", max = 1.1)
+
+fig
 ```
-Type I error rates ($\alpha = .05$) as a function of the size $n$ of the sample, where all population effects are null.
+Scatterplots showing the relationship between effect size estimates (Cohen's $f$) obtained with each method and those from the ground-truth analysis for the main effect of $X_1$, when $a_1$ and $a_2$ vary independently within the range $[-2, 2]$.
 :::
+
+Under log-normal distributions, PAR tends to severely underestimate effect sizes, whereas ART produces estimates that are widely dispersed on both sides of the ground-truth values.
+INT yields the most precise estimates, followed by RNK.
+However, neither method provides accurate estimates for ordinal data.
+This limitation largely reflects the restricted resolution of the ordinal scale itself: with only five discrete levels, it cannot adequately represent the full range of variation in the latent variable, leading all methods to underestimate effect sizes.
+The flexible nature of the thresholds in the distributions we tested makes effect size estimation even more challenging.
+
+@fig-effect-main-scatterplots-2 presents the same relationship for a larger range of effects, with $a_1$ and $a_2$ varying within $[-8, 8]$.
+
+::: {#fig-effect-main-scatterplots-2}
+```{r, echo=FALSE, fig.height=3.4, message=FALSE, warning=FALSE}
+prefix <- "cohens_f_8"
+
+df <- readly_datapoints(prefix,  distributions=c("norm", "lnorm", "likert"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)"))
+fig <- plotly_scatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "grtX1", max = 4.2)
+
+fig
+```
+Scatterplots showing the relationship between effect size estimates (Cohen's $f$) obtained with each method and those from the ground-truth analysis for the main effect of $X_1$, when $a_1$ and $a_2$ vary independently within the range $[-8, 8]$.
+:::
+
+In this broader range, all three rank-based methods --- although RNK to a lesser extent --- struggle to recover large effects, even when the underlying distribution is normal.
+However, we acknowledge that accurately estimating effects of such magnitudes (e.g., when Cohen's $f > 2$ or $\eta^2 > 0.8$) is of limited practical importance.
+
+ART fails entirely under log-normal data, regardless of effect magnitude. Finally, all methods produce highly variable effect size estimates for ordinal data.
+
+**Interactions**.
+Finally, we present additional results illustrating how rank-based methods can fail to accurately estimate interaction effects.
+@fig-eta-squared-interaction shows partial $\eta^2$ estimates for the interaction term under normal distributions, as both $a_1$ and $a_2$ increase.
+Once again, RNK substantially distorts the estimated interaction effect.
+Although INT mitigates this distortion, the problem reemerges as the main effects become larger.
+
+::: {#fig-eta-squared-interaction}
+```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
+data <- read.csv("results/eta_squared.csv", sep=",", header=TRUE, strip.white=TRUE)
+df <- data[data$family == "norm" & data$effectX1 == data$effectX2, ]
+df$effectX2 <- factor(df$effectX2)
+
+ylab = "interaction - partial eta squared"
+xlab = "magnitude of main effects: a1 and a2"
+
+margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
+cbPalette = c("#8888ff", "#888888", "#E69F00", "#009E73", "#FF5E00")
+cbPalette2 = c("#8888ff88", "#88888888", "#E69F0088", "#009E7388", "#FF5E0088")
+
+fig1 <- plot_ly(df, x = ~effectX2, y = ~grtX1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1X2, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
+fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
+fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
+fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))
+
+fig <- subplot(fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
+  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
+	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab), xaxis = list(fixedrange=T)) %>% layout(
+    hovermode = "x unified",  # Shared hover on the x-axis
+    yaxis = list(spikemode = "across", spikesnap = "cursor")
+  )
+
+fig
+```
+Boxplots summarizing partial $\eta^2$ estimates for interaction effects from 300 simulated datasets ($n=20$).
+The population interaction effect is null, and responses follow **normal distributions**.
+Hover over the plots to compare the methods.
+:::
+
+@fig-eta-squared-ordinal presents results for an ordinal scale with five flexible levels. In this case, effect size estimates become even more distorted across all methods.
+
+::: {#fig-eta-squared-ordinal}
+```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
+df <- data[data$family == "likert" & data$effectX1 == data$effectX2, ]
+df$effectX2 <- factor(df$effectX2)
+
+ylab = "interaction - partial eta squared"
+xlab = "magnitude of main effects: a1 and a2"
+
+margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
+cbPalette = c("#8888ff", "#888888", "#E69F00", "#009E73", "#FF5E00")
+cbPalette2 = c("#8888ff88", "#88888888", "#E69F0088", "#009E7388", "#FF5E0088")
+
+fig1 <- plot_ly(df, x = ~effectX2, y = ~grtX1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
+fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1X2, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
+fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
+fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
+fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))
+
+fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
+  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
+	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab), xaxis = list(fixedrange=T)) %>% layout(
+    hovermode = "x unified",  # Shared hover on the x-axis
+    yaxis = list(spikemode = "across", spikesnap = "cursor")
+  )
+
+fig
+```
+Boxplots summarizing partial $\eta^2$ estimates for interaction effects from 300 simulated datasets ($n=20$).
+The population interaction effect is null, and responses are **ordinal with 5 flexible levels**.
+Hover over the plots to compare the methods.
+:::

### 2025-12-22 text/65-Results-EffectSize.qmd
added effect sizes

@@
 ### Effect size estimates {#effect-sizes}
+Transformations --- even monotonic ones --- alter both the absolute magnitude of effects and the total variance of the data.
+A natural question, therefore, is how to interpret effect size estimates derived from rank transformations and whether such estimates are meaningful and comparable across methods.
+Nevertheless, @Kay:art-effect-sizes provides a tutorial for deriving standardized effect sizes from ART, and many of the papers we reviewed report such estimates.
+This practice motivates a more systematic investigation.
 
-We expect that discrepancies in the *p*-values produced by different methods will also be reflected in their corresponding effect size estimates.
-We presents results from various experiments, concentrating on either partial $\eta^2$ estimates or Cohen's $f$ estimates.
-
-**Effect sizes estimates in log-normal distributions.** Consider a $4 \times 3$ repeated-measures design with $n = 20$, where the effect of the first factor $X_1$ is null ($a_1 = 0$) and there is no interaction ($a_{12} = 0$).
-We assume that the responses follow log-normal distributions.
-@fig-eta-squared displays the partial $\eta^2$ estimated by each method for the effect of $X_1$ as the magnitude of the effect on the second factor increases.
-Each cell contains 300 data points.
-The ground truth uses ANOVA directly on the latent variable, where distributions are normal.
+**Log-normal distributions.** We begin by examining log-normal responses.
+Consider a $4 \times 3$ within-subjects design with $n = 20$, in which the effect of the first factor $X_1$ is null ($a_1 = 0$) and there is no interaction ($a_{12} = 0$).
+@fig-eta-squared displays the partial $\eta^2$ estimates produced by each method for the effect of $X_1$ as the magnitude of the effect on the second factor $X_2$ increases.
+These estimates are compared to those from a ground-truth analysis, defined here as ANOVA applied to the log-transformed responses.
 
@@
 ```
-Boxplots summarizing the $\eta^2$ estimates obtained with each method for 300 data points ($n=20$).
+Boxplots summarizing the $\eta^2$ estimates obtained with each method from 300 simulated datasets ($n=20$).
 The population effect of $X1$ is null, and responses follow **log-normal distributions**.
@@
 
-ART is the only method whose effect size estimates for $X_1$ are affected by the magnitude of the effect on $X_2$.
-The results clearly demonstrate that the method's alignment mechanism fails in nonlinear models with skewed distributions and confounds effects.
+We observe that ART is the only method for which the estimated effect sizes for $X_1$ are influenced by the magnitude of the effect on $X_2$.
+These results clearly demonstrate that ART's alignment mechanism fails in nonlinear models with skewed distributions and leads to confounding across effects.
 
-@fig-eta-squared-2 presents $\eta^2$ estimates when the effect of $X_1$ is no longer null ($a_1 = 1$).
-We observe that PAR is not ideal for such data, as it consistently underestimates effect sizes.
+@fig-eta-squared-2 presents partial $\eta^2$ estimates when the effect of $X_1$ is no longer null ($a_1 = 1$).
+PAR performs poorly for these data, consistently underestimating effect sizes.
 RNK also slightly underestimates effects.
-ART tends to underestimate effects when $a_2 = 0$, but as $a_2$ increases, its estimates become more variable and show an overall tendency to increase.
-INT provides the most precise estimates but is also slightly affected when $a_2 = 8$.
+ART also tends to underestimate effects when $a_2 = 0$; however, as $a_2$ increases, its estimates become more variable and exhibit an overall tendency to increase.
+INT yields the most precise estimates and is only mildly affected when $a_2 = 8$.
 
@@
 ```
-Boxplots summarizing the $\eta^2$ estimates obtained with each method for 300 data points ($n=20$).
-The population effect of $X1$ is now non-null ($a_1 = 1$).
-Responses follow again **log-normal distributions**.
+Boxplots summarizing partial $\eta^2$ estimates obtained with each method from 300 simulated datasets ($n=20$).
+The population effect of $X1$ is non-null ($a_1 = 1$), and responses again follow **log-normal distributions**.
 Hover over the plots to compare the methods.
@@
 
-Our final experiment assesses the precision of the effect size estimates (Cohen's $f$) of each method.
+**Systematic comparison to ground truth estimates**.
+We conduct an additional experiment to compare the methods to the ground truth across a wider range of effect magnitudes for the normal, log-normal, and ordinal distributions (the latter with five levels and flexible thresholds).
+For ordinal data, the ground truth is defined as ANOVA applied to the latent variable (i.e., the data prior to discretization).
+We now use Cohen’s $f$ as a measure of effect size, which is expected to be proportional to the true effect magnitude.
+Without loss of generality, we focus on the main effect of $X_1$ and fix the sample size to $n=20$.
 
-**Relationship to ground truth estimates**.
-We evaluate the precision of effect size estimates for wider range of effects.
-We now use Cohen's $f$ as a measure of effect size, which is expected to be proportional to the real effect.
-Without loss of generality, we focus on the main effect of $X_1$.
-We also limit our analysis to three distributions: normal, log-normal, and ordinal with five flexible levels.
-
-@fig-effect-main-scatterplots-1 illustrates the relationship between the Cohen's $f$ obtained with each method and the $f$ of the ground-truth method, when $a_1$ and $a_2$ vary within the range $[-2, 2]$.
-The proximity of data points to the black diagonal indicates the closeness of estimates to the ground truth.
-Points above the line are likely to overestimate the effect size, while points under the line are likely to underestimate it.
+@fig-effect-main-scatterplots-1 illustrates the relationship between the Cohen's $f$ estimated by each method and the corresponding ground-truth estimate, as $a_1$ and $a_2$ vary independently within the range $[-2, 2]$.
+This produces Cohen’s $f$ values between $0$ and $1$ for the $4 \times 3$ within-subject and $2 \times 3$ between-subject designs, and a wider range for the $2 \times 4$ mixed design.
+Proximity to the black diagonal indicates closer agreement with the ground truth.
+Points above the diagonal indicate overestimation of the effect size, whereas points below indicate underestimation.
 
@@
 ```{r, echo=FALSE, fig.height=3.4, message=FALSE, warning=FALSE}
-source("effect_sizes_scatterplots.R")
-prefix <- "6_scatter-Effect_Size-g2"
-df <- readlyDataPoints(prefix,  distributions=c("norm", "lnorm", "likert5B"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)")) %>% toCohensf()
-fig <- plotlyScatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "fX1", max = 1.05)
+source("cohens_f_plots.R")
+prefix <- "cohens_f_2"
+df <- readly_datapoints(prefix,  distributions=c("norm", "lnorm", "likert"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)"))
+fig <- plotly_scatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "grtX1", max = 1.1)
 
@@
 ```
-Scatterplots showing the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground-truth method for the main effect of $X_1$, when $a_1$ and $a_2$ vary within the range $[-2, 2]$.
+Scatterplots showing the relationship between effect size estimates (Cohen's $f$) obtained with each method and those from the ground-truth analysis for the main effect of $X_1$, when $a_1$ and $a_2$ vary independently within the range $[-2, 2]$.
 :::
 
-We observe that PAR tends to underestimate effect sizes under log-normal distributions, in consistency with its low power (see @fig-power-main-1).
-ART's estimates are instead spread across both sides.
-We invite the reader to zoom in on the lower range of values ($f < 0.4$), where the method frequently exaggerates effects.
-INT demonstrates the highest precision, followed by RNK.
-However, their performance deteriorates under the ordinal scale.
-Still, ART performs worse than all other methods.
+Under log-normal distributions, PAR tends to severely underestimate effect sizes, whereas ART produces estimates that are widely dispersed on both sides of the ground-truth values.
+INT yields the most precise estimates, followed by RNK.
+However, neither method provides accurate estimates for ordinal data.
+This limitation largely reflects the restricted resolution of the ordinal scale itself: with only five discrete levels, it cannot adequately represent the full range of variation in the latent variable, leading all methods to underestimate effect sizes.
+The flexible nature of the thresholds in the distributions we tested makes effect size estimation even more challenging.
 
-@fig-effect-main-scatterplots-2 shows the same relationship but for a larger range of effects $[-8, 8]$.
+@fig-effect-main-scatterplots-2 presents the same relationship for a larger range of effects, with $a_1$ and $a_2$ varying within $[-8, 8]$.
 
@@
 ```{r, echo=FALSE, fig.height=3.4, message=FALSE, warning=FALSE}
-#source("effect_sizes_scatterplots.R")
-prefix <- "6_scatter-Effect_Size-g8"
+prefix <- "cohens_f_8"
 
-df <- readlyDataPoints(prefix,  distributions=c("norm", "lnorm", "likert5B"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)")) %>% toCohensf()
-fig <- plotlyScatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "fX1", max = 4.05)
+df <- readly_datapoints(prefix,  distributions=c("norm", "lnorm", "likert"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)"))
+fig <- plotly_scatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "grtX1", max = 4.2)
 
@@
 ```
-Scatterplots showing the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground-truth method for the main effect of $X_1$, when $a_1$ and $a_2$ vary within the range $[-8, 8]$.
+Scatterplots showing the relationship between effect size estimates (Cohen's $f$) obtained with each method and those from the ground-truth analysis for the main effect of $X_1$, when $a_1$ and $a_2$ vary independently within the range $[-8, 8]$.
 :::
 
-We observe that all three rank-based methods struggle to detect large effects, even when the underlying distribution is normal.
-This result reflects a fundamental limitation of ranks.
-Consider, for example, two sets of numbers $A= \{1, 2, 3\}$ and $B= \{4, 5, 6\}$.
-The mean difference between them is $\overline{B} - \overline{A} = 3$, whether calculated from the row values or their ranks.
-However, if the set $B$ becomes $\{9, 10, 11\}$, the mean difference of raw values increases to $8$, while the mean difference in ranks remains $3$.
-As effects become larger, this problem becomes more apparent.
+In this broader range, all three rank-based methods --- although RNK to a lesser extent --- struggle to recover large effects, even when the underlying distribution is normal.
+However, we acknowledge that accurately estimating effects of such magnitudes (e.g., when Cohen's $f > 2$ or $\eta^2 > 0.8$) is of limited practical importance.
 
-In the case of ordinal data, all methods yield highly variable estimates.
-This variability is largely due to the limited resolution of the ordinal scale itself: with only five discrete levels, it cannot adequately represent the full range of variation in the latent variable.
-The systematic underestimation of large effects across methods is also due to the limited range of the ordinal scale's extremes, which masks the true magnitude of underlying differences.
+ART fails entirely under log-normal data, regardless of effect magnitude. Finally, all methods produce highly variable effect size estimates for ordinal data.
 
 **Interactions**.
-We also provide additional results illustrating how rank-based methods can fail to accurately estimate interaction effects.
-@fig-eta-squared-interaction shows the estimated values of partial $\eta^2$ for the interaction term under normal distributions, as both main effects increase.
-Once again, we observe that RNK can substantially distort the estimated interaction.
-While INT reduces this distortion, the problem reappears as the main effects grow larger.
+Finally, we present additional results illustrating how rank-based methods can fail to accurately estimate interaction effects.
+@fig-eta-squared-interaction shows partial $\eta^2$ estimates for the interaction term under normal distributions, as both $a_1$ and $a_2$ increase.
+Once again, RNK substantially distorts the estimated interaction effect.
+Although INT mitigates this distortion, the problem reemerges as the main effects become larger.
 
@@
 ```
-Boxplots summarizing the $\eta^2$ estimates for interactions obtained with each method for 300 data points ($n=20$).
+Boxplots summarizing partial $\eta^2$ estimates for interaction effects from 300 simulated datasets ($n=20$).
 The population interaction effect is null, and responses follow **normal distributions**.
@@
 
-If the data take a discrete form, effect size estimates become even more distorted across all methods. @fig-eta-squared-ordinal presents results for an ordinal scale with 5 flexible levels.
+@fig-eta-squared-ordinal presents results for an ordinal scale with five flexible levels. In this case, effect size estimates become even more distorted across all methods.
 
@@
 ```
-Boxplots summarizing the $\eta^2$ estimates for interactions obtained with each method for 300 data points ($n=20$).
+Boxplots summarizing partial $\eta^2$ estimates for interaction effects from 300 simulated datasets ($n=20$).
 The population interaction effect is null, and responses are **ordinal with 5 flexible levels**.
@@
 :::
-
-As discussed extensively earlier, such ordinal scales can distort interaction trends when main effects are non-zero, leading to interpretation issues.
-Although the methods may detect strong interaction effects under these conditions, the effects may not be intrinsic or meaningful.
-
-<!--
-To systematically evaluate the precision of effect size estimates across different methods, we employ the coefficient of determination, denoted as $R^2$ ($R$ squared).
-This metric quantifies the proportion of variation in the $f$ estimates of each method that is predictable from the estimate provided by the reference method, considered as the ground truth.
-We compute $R^2$ over 1000 iterations.
-Within each iteration, we randomly sample the magnitude of the two main effects ($a_1$ and $a_2$) from a range spanning $[-g, g]$, where $g$ takes values from the set ${0, 0.5, 1, 2, 4, 8}$.
-It is worth noting that the $R^2$ measure assesses the proximity of estimates to the ground truth, without distinguishing between underestimates and overestimates.
-We present these results based on a sample size of $n=20$.
-
-@fig-effect-main-1 depicts trends for the first factor $X_1$.
-INT performs exceptionally well across all continuous distributions, but its precision declines when the range of effects becomes large ($g > 4$).
-PAR's performance is strikingly low under the log-normal distribution and, to a lesser degree, under the exponential distribution.
-ART emerges as the second-worst method for these distributions.
-Finally, we observe low $R^2$ values for all methods under the binomial and ordinal distributions.
-ART's performance is particularly problematic when effects sizes are low.
-
-::: {#fig-effect-main-1}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "6_test-Effect-Size-Main"
-distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
-dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
-
-df <- readData(prefix, n = 20, alpha= NA, effectType = -1, distributions)
-
-df <- df %>% arrange(design,distr,effectX1,cohensfX1)  %>% group_by(design,distr,effectX1)
-df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"), groupvars = c("distr","method","n"))
-
-#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "range of effects (g)", var = "cohensfX1", xvar = "effectX1", min = -11, max = 103, ytitle = 'R squared')
-```
-$R^2$ scores measuring the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground-truth method for the main effect on $X_1$.
-We express these scores as a function of the range of the magnitude of effects $g$.
-:::
-
-**Interaction effects**.
-We extend our analysis to examine trends regarding the interaction effect, as depicted in Figure @fig-effect-interaction.
-We now also vary $a_{12}$ within the same range $[-g, g]$.
-$R^2$ approaches zero or becomes negative under certain conditions, indicating that effect size estimates cannot be trusted in these cases.
-RNK and INT emerge as the most precise methods under non-normal distributions, but their performance declines rapidly when $g > 2$.
-Finally, PAR outperforms ART across all discrete distributions, as well as the normal distribution.
-
-::: {#fig-effect-interaction}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "6_test-Effect-Size"
-distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
-dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
-
-df <- readData(prefix, n = 20, alpha= NA, effectType = -1, distributions)
-
-df <- df %>% arrange(design,distr,effectX1X2,cohensfX1X2)  %>% group_by(design,distr,effectX1X2)
-df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"), groupvars = c("distr","method","n"))
-
-#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "range of effects (g)", var = "cohensfX1X2", xvar = "effectX1X2", min = -51, max = 105, ytitle = 'R squared')
-```
-$R^2$ scores measuring the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground truth method for the interaction effect $X_1 \times X_2$.
-We express these scores as a function of the range of the magnitude of effects $g$.
-:::
-
--->

### 2025-12-22 cohens_f_plots.R

@@ -1,0 +1,93 @@
+#rm(list=ls())
+#library(plotly)
+#library(tidyverse)
+
+library(reshape2)
+
+plotly_scatter <- function(df, xlab = "ground truth", ylab= "Cohen's f", xvar = "grtX1", yvar = paste(xvar, "_", sep =""), max = 0.4) {
+  # aesthetics
+
+  symbols <- c("asterisk", "x", "star-diamond", "star-triangle-up")
+  palette <- c("#888888", "#E69F00", "#009E73", "#FF5E00")
+  margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
+  ldata = data.frame(x=c(0,5), y = c(0,5))
+
+  dnames <- levels(df$family)
+
+  # To be used as labels on the subplots
+  annots <- lapply(1:length(dnames), function(index){list(x = (index-0.5)/length(dnames), y = 1, text = dnames[index], xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F)})
+
+  menuItems <- c("4x3 within", "2x3 between", "2x4 mixed")
+  xvarnames <- c(paste(xvar,"4x3",sep="."), paste(xvar,"2x3",sep="."), paste(xvar,"2x4",sep="."))
+  yvarnames <- c(paste(yvar,"4x3",sep="."), paste(yvar,"2x3",sep="."), paste(yvar,"2x4",sep="."))
+
+  createPlot <- function(data, dnames, symbols) {
+    p <- plot_ly(colors = palette, symbols = symbols) %>%  add_trace(data=ldata, x = ~x, y = ~y, type = 'scatter', mode ="lines", line = list(color = "black"), showlegend = FALSE)
+    p <- add_trace(p, data = data, y = ~eval(as.symbol(yvarnames[1])), x = ~eval(as.symbol(xvarnames[1])), color = ~method, symbol = ~method,  type = 'scatter', mode = 'markers', marker = list(line = list(width = 1)), legendgroup = ~method, showlegend = unique(data$family == dnames[1]), visible = T) 
+    p <- add_trace(p, data = data, y = ~eval(as.symbol(yvarnames[2])), x = ~eval(as.symbol(xvarnames[2])), color = ~method, symbol = ~method,  type = 'scatter', mode = 'markers', marker = list(line = list(width = 1)), legendgroup = ~method, showlegend = unique(data$family == dnames[2]), visible = F)  
[…]

### 2025-12-22 supplementary/experiments-code/experiment-effect-sizes-scatter.R

@@ -1,7 +1,7 @@
 # Author: Theophanis Tsandilas, Dec 2025
 # Inria & Université Paris-Saclay
 
-# Experiment evaluating the precision of effect size estimates (eta square)
+# Experiment evaluating the precision of effect size estimates
 
 rm(list=ls())
 
@@ -27,45 +27,31 @@ use_parameters <- function(family){
   )
 
   # Choose a random standard deviation for the random subject effect between 0.1 and 0.5
-  params$sigma_s <- c(0.1, 0.5) # Specifies the min and max of a uniform range
+  params$sigma_s <- c(0.5) # Specifies the min and max of a uniform range
 
   params
 }
 
-# 4 x 3 repeated measures
-design = c(4,3)
-within = c(1,1) # Both factors are within-subjects
+# (1) 4 x 3 within-subjects, (2) 2 x 3 between-subjects, (3) 2 x 4 mixed design
+designs <- list(c(4,3), c(2,3), c(2,4))
+within <- list(c(1,1), c(0,0), c(0,1))
+
 formula = Y ~ X1*X2 + Error(factor(subject)) 
 vars = c("X1", "X2", "X1:X2") 
 
 # Continuous distributions (equal variance, full) and discrete distribution: binom (size = 10, prob = 0.1) and Poisson
 distributions <- c("norm", "lnorm", "likert")
 
-# Various combinations of effects
-effects <- matrix(c(0, 0, 0,
-            0, 1, 0,
-            0, 2, 0,
-            0, 4, 0,
-            0, 8, 0,
-            1, 0, 0,
-            1, 1, 0,
-            1, 2, 0,
-            1, 4, 0,
-            1, 8, 0,
-            2, 2, 0,
-            4, 4, 0,
-            8, 8, 0), 
-           ncol = 3, byrow = TRUE)
-
-colnames(effects) <- vars
+max_effects = c(8,8,0)
+names(max_effects) <- vars
 
 # Cell sizes (n in the paper) -- for within-subject designs, it's also the number of subjects
 Ns <- c(20) 
 
-# 300 iterations
-R <- 300
+# 100 iterations
+R <- 100
 
-filename = "eta_squared"
+filename = "cohens_f_8"
 
 # Set the seed for reproducibility
 set.seed(4567)
@@ -77,19 +63,20 @@ registerDoParallel(CoresNum)  # use multicore, set to the number of our cores
 time <- system.time({ 
[…]

### 2025-12-22 supplementary/experiments-code/utils-effect-size.R

@@ -98,15 +98,21 @@ repeat_test_effect_size <- function(
   params,
   formula,
   vars,
-  measure = "eta",
-  iterations = 100
+  measure = c("eta", "cohens"),
+  iterations = 100,
+  is_max = FALSE # If true, the coeffs define maximum effects (to be randomly chosen)
 ) {
   
   family <- match.arg(family)
+  measure <- match.arg(measure)
 
   results <- foreach(rid = 1:iterations, .combine=rbind) %dopar% {
     tryCatch(
       {
+        if(is_max) {
+          coeffs[1:length(coeffs)] = 
+                sapply(1:length(coeffs), function(x){runif(1, min=-coeffs[x], max=coeffs[x])})
+        } 
         data <- { # Check if variances are equal
           if(is.null(params$ratio_sd)) simulate_response(nlevels, within, n, coeffs, sub("_.*", "", family), params)
           else simulate_heteroscedastic_response(nlevels, within, n, coeffs, sub("_.*", "", family), params)
@@ -129,7 +135,7 @@ repeat_test_effect_size <- function(
           groundtruth <- get_groundtruth_effect_sizes(data, formula, vars, measure)
         }
 
-        c(eff_sizes, groundtruth)
+        c(coeffs, eff_sizes, groundtruth)
       }, 
       error = function(cond) {
         # do nothing
@@ -141,9 +147,9 @@ repeat_test_effect_size <- function(
 
   designStr <- paste(nlevels, collapse="x")
 
-  colnames(results) <- c(paste0("par", vars), paste0("rnk", vars), paste0("art", vars), paste0("int", vars), paste0("grt", vars)) 
-  coeffs_prefixed <- setNames(coeffs, paste0("effect", names(coeffs)))
+  colnames(results) <- c( paste0("effect", names(coeffs)), paste0("par", vars), paste0("rnk", vars), paste0("art", vars), paste0("int", vars), paste0("grt", vars)) 
+  #coeffs_prefixed <- setNames(coeffs, paste0("effect", names(coeffs)))
 
-  cbind(n = n, design=designStr, family = family, as.data.frame(as.list(coeffs_prefixed)), round(results,digits=4))
+  as.data.frame(cbind(n = n, design=designStr, family = family, round(results,digits=4)))
 }
 

### 2025-12-28 appendix_I.qmd
working on appendix

@@
 library(tidyverse)
+
+source("utils-data-reading.R")
+source("utils-plotlying.R")
+
 source("plotlying.R")
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix1 <- "1_test_4x3_Ratio"
+prefix1 <- "Type_I_4x3_ratio"
 alpha <- .01
@@
 dnames1 <- c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial")
-df1 <- readlyData(prefix1, alpha, 1, distributions1, dnames1)
+df1 <- readly_data(prefix1, alpha, 1, distributions1, dnames1)
 
-prefix2 <- "2_test_4x3_Ordinal"
-distributions2 <- c("likert5B")
+prefix2 <- "Type_I_ordinal"
+distributions2 <- c("likert_5_flex")
 dnames2 <- c("Ordinal (5 levels)")
-df2 <- readlyData(prefix2, alpha, 1, distributions2, dnames2)
+df2 <- readly_data(prefix2, alpha, 1, distributions2, dnames2)
 
@@
 
-plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 27.1)
+plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 103)
 ```
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-plotlyError(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 27.1)
+plotly_error(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 103)
 ```
@@
 We also present results for interaction effects when the two main effects change in parallel.
-As we explain in the main paper, these results require special attention, as error rates largely depend on the way we define the null hypothesis for interactions.
+As we explain in the main paper, these results require special attention, as whether positives rates are considered as Type I error rates depends on the way we define the null hypothesis for interactions.
 
@@
 ```{r, echo=FALSE, message=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df1 <- readlyData(prefix1, alpha, 0, distributions1, dnames1)
-df2 <- readlyData(prefix2, alpha, 0, distributions2, dnames2)
+df1 <- readly_data(prefix1, alpha, 0, distributions1, dnames1)
+df2 <- readly_data(prefix2, alpha, 0, distributions2, dnames2)
 df <- rbind(df1,df2)
 
-plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 104)
+plotly_error(df, xlab = "magnitude of main effects", ytitle = 'Positives (%)',var = "rateX1X2", xvar = "effectX1", max = 104)
 ```
-Type I error rates ($\alpha = .01$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$.
-*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
-A different definition may lead to different results.*
+Positive rates ($\alpha = .01$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$.
+*Note: These values correspond to Type I error rates under the null hypothesis $a_{12} = 0$; alternative definitions of the null hypothesis may lead to different results.*
 :::
@@
 ### Conclusion
-Results for $\alpha = .01$ are consistent with our results for $\alpha = .05$, confirming that ART is not a robust method.
+Results for $\alpha = .01$ are consistent with our results for $\alpha = .05$.
 
 ## Missing data {#missing}
-We evaluate how missing data can affect the performance of the four methods.
-Specifically, we study a scenario, where a random sample of $10\%$ of the observations is missing.
-Missing observations lead to unbalanced designs.
-However, we emphasize that our scenario does not cover systematic imbalances due to missing data for specific levels of a factor.
+We evaluate how missing data affect the performance of the four methods.
+Specifically, we consider scenarios in which a random sample of $10\%$, $20\%$, or $30\%$ of the observations is missing.
+Our analysis is restricted to normal distributions and includes two conditions: (i) equal variances and (ii) unequal variances across the levels of the factor to effects are applied ($X_1$ in our experiments), with the maximum ratio of standard deviations fixed to $r_{sd}=2$.
+
+Missing observations lead to unbalanced designs, for which Type I hypothesis tests obtained using the `aov()` function are difficult to interpret.
+Accordingly, we use LMER and conduct Type III tests [@Kuznetsova:2017], focusing on a $4 \times 3$ within-subjects design and a $2 \times 4$ mixed design.
+
+We emphasize, however, that our scenarios do not include systematic imbalances arising from missing data concentrated in specific factor levels.
 
@@
 @fig-missing-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases.
-We observe that missing data make the error rate for ART to increase even further, with a larger increase for the mixed design.
-This is also the case for the normal distribution.
-In contrast, the accuracy of the three other methods does not seem to be affected.
+We find that missing data cause ART's error rates to increase, with a more pronounced effects in the mixed design.
+This additional inflation compounds the error introduced by unequal variances.
+In contrast, the accuracy of the three other methods appears largely unaffected.
 
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Appendix_test_missing"
-distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
-dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+prefix <- "Type_I_missing"
+distributions = c("norm")
 
-df <- readData(prefix, n = 20, alpha = .05, effectType = 1, distributions)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 98)
+dnames <- c("10% - equal", "10% - unequal",  "20% - equal", "20% - unequal", "30% - equal", "30% - unequal")
+df <- read_data(prefix, alpha = .05, effectType = 1, distributions)
+df$family <- paste(df$missingratio*100, "% - ", ifelse(df$sdratio == 1, "equal", "unequal"), sep="") 
+df <- reshape_by_design(df, dnames = dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+plotly_error_by_design_3(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 81)
+
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$, when $10\%$, $20\%$, or $30\%$ of the observations are missing ($n=20$).
+:::
+
+**Interaction effects**.
+We also examine Type I error rates for the interaction effect.
+In this case, ART's error rates are not influenced by missing data; the inflated errors observed in the mixed design are due to unequal variances.
+
+::: {#fig-missing-interaction1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+plotly_error_by_design_3(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 21)
+
+```
+Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$, when $10\%$, $20\%$, or $30\%$ of the observations are missing ($n=20$).
+:::
+
+### Conclusion
+ART is sensitive to the presence of missing data. Even under normal distributions, its Type I error rates for main effects increase.
+
+## Log-normal distributions {#log-normal}
+We evaluate log-normal distributions across a wider range of $\sigma$ values (see @fig-lognormal-distributions), with particular attention to distributions with smaller variance, which exhibit lower degrees of skewness.
+
+::: {#fig-lognormal-distributions}
+```{r, echo=FALSE, fig.height=1.3, fig.width = 9, warning=FALSE,  message=FALSE}
+source("distributions_plots.R")
+plot_effects_lognormal()
+```
+Log-normal distributions that we evaluate, shown here for a factor with two levels and a magnitude of effect $a_{1} = 2$. 
+:::
+
+**Main effects**.
+@fig-lognormal-main presents the Type I error rates for main effects.
+As expected, ART’s inflation of error rates is less severe when the distributions are closer to normal, whereas the problem becomes more pronounced as skewness increases.
+
+::: {#fig-lognormal-main}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Type_I_lognormal"
+distributions = c("lnorm_0.2", "lnorm_0.4", "lnorm_0.6", "lnorm_0.8", "lnorm_1.0", "lnorm_1.2")
+dnames = c("sd = 0.2", "sd = 0.4", "sd = 0.6", "sd = 0.8", "sd = 1.0", "sd = 1.2")
+
+df <- read_data(prefix, alpha = .05, effectType = 1, distributions)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 103)
 
@@
 **Interaction effects**.
-We also present Type I error rates for the interaction effect in the presence of a single main effect or two parallel main effects.
-The error levels for all methods, including ART, are now very similar to the ones observed with no missing data.
-
-::: {#fig-missing-interaction1}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 98)
-
-```
-Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
-:::
-
-### Conclusion
-ART is sensitive to the presence of missing data when at least $10\%$ of the observations are missing.
-Specifically, we found that its Type I error rate for main effects increases further, even with normal distributions.
-The other methods do not seem to be affected.
-However, we emphasize that we only tested missing data that are randomly drawn from the complete set of observations.
-Results might be different if there were systematic bias in the data imbalance structure.
-
-## Log-normal distributions {#log-normal}
-In a different experiment, we evaluate log-normal distributions with a wider range of $\sigma$ parameters (see @fig-lognormal-distributions), in particular distributions with less variance, which exhibit a lower degree of skew.
-
-::: {#fig-lognormal-distributions}
-```{r, echo=FALSE, fig.height=1.3, fig.width = 9, warning=FALSE,  message=FALSE}
-source("effects_plot.R")
-plotEffectsLognormal(nlevels = 2, effectSize = 2)
-```
-Log-normal distributions that we evaluate, shown here for a factor with two levels and a magnitude of effect $a_{i} = 2$. 
-:::
-
-**Main effects**.
-@fig-lognormal-main presents our results on Type I error rates for main effects.
-As expected, ART's inflation of error rates is less serious when distributions are closer to normal, while the problem becomes worse as distributions are more skewed.
-
-::: {#fig-lognormal-main}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Appendix_test_lognormal"
-distributions = c("lnorm-0.2", "lnorm-0.4", "lnorm-0.6", "lnorm-0.8", "lnorm-1.0", "lnorm-1.2")
-dnames = c("sd = 0.2", "sd = 0.4", "sd = 0.6", "sd = 0.8", "sd = 1.0", "sd = 1.2")
-
-df <- readData(prefix, n = 20, alpha = .05, effectType = 1, distributions)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 69)
-
-```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{1}$ of the main effect of $X_1$ ($n=20$)
-:::
-
-**Interaction effects**. We observe similar patterns for the Type I error rate of the interaction effect in the presence of a single main effect.
+We observe similar patterns for the Type I error rates of interaction effects in the presence of a single main effect.
+Notice that PAR's error rates are highly unstable.
+They appears inflated under the within-subjects design and deflated under other configurations.
 
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 69)
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 103)
 
@@
 **Interactions under parallel main effects**.
-We also present results when the two main effects increase in parallel.
-These results require again careful interpretation, as the null hypothesis of interest may be different for each method.
-It is important to note that even slight departures from normality (e.g., when $\sigma = 0.2$) make the interpretation of interactions problematic.
+We also report results for scenarios in which both main effects increase in parallel.
+These results again require careful interpretation, as the null hypothesis of interest may differ across methods.
+Notably, even slight departures from normality (e.g., $\sigma = 0.2$) can make the interpretation of interaction effects problematic.
 
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- readData(prefix, n = 20, alpha = .05, effectType = 0, distributions)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 103)
+df <- read_data(prefix, alpha = .05, effectType = 0, distributions)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effects", ytitle = 'Positives (%)', var = "rateX1X2", xvar = "effectX1", max = 103)
 
 ```
-Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitudes $a_{1}$ and $a_{2}$ of the main effects ($n=20$).
-*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
-A different definition may lead to different results.*
+Positive rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as a function of the magnitudes $a_{1}$ and $a_{2}$ of the main effects ($n=20$).
+*Note: These values correspond to Type I error rates under the null hypothesis $a_{12} = 0$; alternative definitions of the null hypothesis may lead to different results.*
 :::
@@
 ### Conclusion
-ART's robustness issues become less severe as log-normal distributions become less skewed and approach normality.
-However, interpretation issues for interactions (in the presence of parallel main effects) can arise even when distributions are only slightly skewed.
+ART’s robustness issues become less severe as log-normal distributions approach normality. However, even when variance is relatively small, Type I error rates can still become unacceptably large.
 
@@
 ## Binomial distributions {#binomial}
-We also evaluate a wider range of parameters for the binomial distribution.
-We focus on the lower range of probabilities $p$.
-However, we expect results to be identical for their symmetric probabilities $1-p$.
-Specifically, we test $p=.05$, $.1$, and $.2$, and for each, we consider $k=5$ and $10$ task repetitions (Bernoulli trials).
+We examine a broader range of parameter settings for the binomial distribution.
+Although we focus on the lower range of the success probability $p$, we expect analogous results for their symmetric values $1-p$.
+Specifically, we test $p=.05$, $.1$, and $.2$, and for each value we consider $\kappa=5$ and $10$ Bernoulli trials.
 
 **Main effects**.
-We present our results for the main effect in @fig-binom-main.
-We observe that ART's Type I error rates increase as the number of repetitions decreases and the probability of success approaches zero, reaching very high levels when $k=5$ and $p=.05$.
-This trend is consistent across designs.
-The other methods maintain low error rates.
-However, their error rates fall below nominal levels when the magnitude of the effect on $X_1$ grows beyond a certain threshold, indicating a loss of power in these cases.
+The results for main effects are presented in @fig-binom-main.
+For the within-subjects design, we find that ART’s Type I error rates increase as the number of trials decreases and as the probability of success approaches zero, reaching extremely high levels when $\kappa = 5$ and $p=.05$.
+However, trends change across designs.
+The other methods maintain good control of Type I error rates.
 
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Appendix_test_binom"
-distributions = c("binom-5-05", "binom-5-10", "binom-5-20", "binom-10-05", "binom-10-10", "binom-10-20")
-dnames = c("k=5, p=5%", "k=5, p=10%", "k=5, p=20%", "k=10, p=5%", "k=10, p=10%", "k=10, p=20%")
+prefix <- "Type_I_binomial"
+distributions = c("binom_5_05", "binom_5_10", "binom_5_20", "binom_10_05", "binom_10_10", "binom_10_20")
+dnames = c("5 trials, p=5%", "5 trials, p=10%", "5 trials, p=20%", "10 trials, p=5%", "10 trials, p=10%", "10 trials, p=20%")
 
-df <- readData(prefix, n = 20, alpha = .05, effectType = 1, distributions)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 98)
+df <- read_data(prefix, alpha = .05, effectType = 1, distributions)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 101)
 
@@
 
-**Interaction effects**. @fig-binom-interaction1 shows similar patterns for the Type I error rate of the interaction effect in the presence of a single main effect.
+**Interaction effects**.
+@fig-binom-interaction1 shows similar patterns for the Type I error rates of interaction effects in the presence of a single main effect.
+In the within-subjects design, however, PAR, RNK, and INT also tend to inflate error rates when $a_1 > 2$, although to a much smaller extent than ART.
 
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 98)
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 101)
 
@@
 **Interactions under parallel main effects**.
-When both main effects exceed a certain level (see @fig-binom-interaction2), all methods fail to control the Type I error rate.
-Whether this inflation is due to interpretation issues or to a lack of robustness in the methods themselves, these results once again highlight that testing interactions in such scenarios is highly problematic.
+When both main effects exceed a certain magnitude (see @fig-binom-interaction2), positive rates increase rapidilly for all methods.
+Whether this inflation is due to interpretation issues or to limitations in the robustness of the methods themselves, these results once again demonstrate the severe difficulties associated with testing interactions under such conditions.
 
@@
 ### Conclusion
-ART is extremely problematic under binomial distributions, raising Type I error at very high levels even when effects are null across all factors.
-We also observe that testing interactions in the presence of parallel main effects can be problematic for all other methods.
-
-## Ordinal data {#ordinal}
-Given the frequent use of ART with ordinal data, we evaluate our complete set of ordinal scales, based on both equidistant and flexible thresholds, with additional experimental designs.
-
-**Main effects**.
-@fig-ordinal-designs-main presents Type I error rates for the main effect.
-ART preserves error rates at nominal levels under the $2 \times 3$ between-subjects design and the $2 \times 3$ mixed design, as long as thresholds are equidistant.
-Under the two within-subject designs, it inflates error rates, especially when there are fewer ordinal levels with flexible thresholds.
-
-::: {#fig-ordinal-designs-main}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Appendix_test_ordinal"
-alpha = 0.05
-
-distributions <- c("likert5", "likert5B", "likert7", "likert7B", "likert11", "likert11B")
-dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
-
-df <- readData(prefix, n = 20, alpha, effectType = 1, distributions)
-df <- reshapeByDesign(df, dnames)
-
-plotlyErrorByDesign(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 62)
-```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
-:::
-
-**Interaction effects**.
-We present Type I error rates for the interaction effect in the presence of a single main effect.
-These results lead to similar conclusions.
-Even in cases where ART keeps error rates close to nominal levels (e.g., under the between-subjects design with equidistance thresholds), the performance of PAR is constantly better.
-
-::: {#fig-ordinal-designs-interaction-1}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-plotlyErrorByDesign(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 62)
-```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$) 
-:::
-
-**Interactions under parallel main effects**. Finally, we show results for scenarios where main effects increase in parallel.
-
-::: {#fig-ordinal-designs-interaction-2}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- readData(prefix, n = 20, alpha, effectType = 0, distributions)
-df <- reshapeByDesign(df, dnames)
-plotlyErrorByDesign(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
-```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$).
-*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
-A different definition may lead to different results.*
-:::
-
-### Conclusion
-ART's inflation of Type I error rates with ordinal data is confirmed across a range of designs.
-For the between-subjects and mixed designs, the problem primarily concerns ordinal scales with flexible thresholds.
-For the within-subjects designs, ART also inflates error rates for scales with equidistant thresholds, particularly when the number of levels is as low as five or seven.
-Again, all methods may fail to correctly infer interactions when parallel main effects exceed a certain threshold.
+ART performs extremely poorly for binomial data, producing very high Type I error rates even when all population effects are null.
+We also observe that testing interaction effects in the presence of parallel main effects is problematic for all methods.
 
@@
 ### Experiment
-To better understand how the different methods detect main in the presence of interactions, we conduct an experiment.
-
-We focus again on two-factor experimental designs, setting the sample size to $n = 20$.
-To simulate populations in which interactions emerge in the absence of main effects, we examine perfectly symmetric cross-interactions.
-To this end, we slightly change the method we use to encode the levels of each factor, such that levels are uniformly positioned around 0.
-For a factor with three levels, we numerically encode the levels as $\{-0.5, 0, 0.5\}$.
-For a factor with four levels, we encode them as $\{-0.5, -0.1667, 0.1667, 0.5\}$.
-
-Since the interpretation of main effects can be ambiguous, we define Type I errors with respect to the coefficients of the linear model.
-Accordingly, the null hypothesis is either $a_1 = 0$ or $a_2 = 0$, depending on the variable of interest.
+To better understand how the different methods detect main in the presence of interactions, we conduct a dedicated experiment.
+We focus on a $2 \times 2$ within-subjects experimental designs and examine perfectly symmetric cross-interactions.
 
 **Interaction effect only**.
-We first examine how the interaction effect alone influences the Type I error rate for $X_2$.
+We first examine how the interaction effect alone influences the Type I error rate for $X_1$ (or $X_2$, since the design is fully symmetric).
+In this configuration, no interpretation issues arise: the true main effects are null under any reasonable definition of the null hypothesis.
+
 The results are shown in @fig-interactions-main-1.
-We emphasize that interpretation issues arise only in the case of the $4 \times 3$ design.
-In the other two designs, $X_1$ has only two levels, and due to their symmetry, the observed main effects should be null --- regardless of how we define the null hypothesis.
-
-In the $4 \times 3$ design, error rates for PAR and ART grow rapidly as the interaction effect becomes large under the log-normal and exponential distributions.
-This is largely due to interpretation issues, as illustrated in our example.
-However, ART also fails to control the error rate in other scenarios.
-Finally, under the binomial and ordinal scales, we observe that RNK and INT may also inflate Type I error rates when the interaction effect becomes sufficiently large.
+We find that ART is the only method adversely affected under non-normal distributions, with Type I error rates increasing as the magnitude of the interaction effect grows.
 
@@
 
-prefix <- "Appendix_test_growing-interactions"
-distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+prefix <- "Type_I_non_null_interaction"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
 dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
 
-df <- readData(prefix, n = 20, alpha, effectType = 0, distributions)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX1", xvar = "effectX1X2", max = 104)
-
+df <- readly_data(prefix, alpha, effectType = 0, distributions, dnames)
+plotly_error(df, xlab = "magnitude of interaction effect", var = "rateX1", xvar = "effectX1X2", max = 104)
 ```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$ ($n=20$).
-*Note: Type I errors are defined based on the null hypothesis $a_{1} = 0$.
-For the $4 \times 3$ design, a different definition may lead to different results.
-However, the two other designs, where $X_1$ has only two levels, Type I error rates are the same, regardless of the definition we use.*
-:::
-
-We also examine error rates for the main effect of $X_2$. Now, interpretation issues affect the results across all three designs, leading to similar patterns.
-
-::: {#fig-interactions-main-2}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-plotlyErrorByDesign2(df, xlab = "magnitude of interaction effect", var = "rateX2", xvar = "effectX1X2", max = 104)
-
-```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$ ($n=20$).
-*Note: Type I errors are defined based on the null hypothesis $a_{2} = 0$.
-A different definition may lead to different results.*
+Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as a function of the magnitude $a_{12}$ of the interaction effect $X_1 \times X_2$.
 :::
@@
 **Interaction effect combined with main effect**.
-Finally, we evaluate the Type I error rate on $X_2$ when the interaction effect is combined with a main effect on $X_1$.
-@fig-interactions-main-3 presents our results.
+We evaluate the behavior of the methods for the effect of $X_2$ when a nonzero interaction effect is combined with a main effect on $X_1$.
+In this setting, the definition of the null hypothesis is inherently ambiguous, and the results must be interpreted with caution.
+We therefore report *Positives (%)* rather than Type I error rates.
+However, if the null hypothesis is defined as $a_2 = 0$, these values can be interpreted as Type I error rates.
 
-The error rates of ART and PAR now increase dramatically across all non-normal distributions and all three designs --- a result that is, in part, due to interpretation issues.
-RNK and INT are also affected: their error rates become extremely low under continuous distributions, suggesting a lack of power to detect small main effects when strong effects of other factors are combined with strong interactions.
-In contrast, under the binomial and ordinal scales, both methods exhibit inflated Type I error rates.
-Interestingly, RNK demonstrates the best overall performance in these tests.
+The results are shown in @fig-interactions-main-3.
+None of the methods maintain positive rates at nominal levels.
+ART and PAR perform appropriately under normal distributions, but their rates increase dramatically for all non-normal distributions.
+In contrast, RNK and INT exhibit decreasing positive rates under continuous distributions, suggesting a loss of sensitivity to small main effects when strong main and interaction effects are present simultaneously.
+Under discrete distributions, however, both methods show substantial inflation of positive rates.
 
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- readData(prefix, n = 20, alpha, effectType = 6, distributions)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of effects", var = "rateX2", xvar = "effectX1X2", max = 104)
-
+df <- readly_data(prefix, alpha, effectType = 6, distributions, dnames)
+plotly_error(df, xlab = "magnitude of interaction effect", ytitle = 'Positives (%)', var = "rateX2", xvar = "effectX1X2", max = 104)
 ```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitudes $a_1 = a_{12}$ of a combined main effect on $X_1$ and an interaction effect $X_1 \times X_2$ ($n=20$).
-*Note: Type I errors are defined based on the null hypothesis $a_{2} = 0$.
-A different definition may lead to different results.*
+Positive rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitudes $a_1 = a_{12}$ of a combined main effect on $X_1$ and an interaction effect $X_1 \times X_2$.
+*Note: These values correspond to Type I error rates under the null hypothesis $a_{2} = 0$; alternative definitions of the null hypothesis may lead to different results.*
 :::
@@
 ### Conclusion
-When strong interactions are present, the interpretation of main effects can become ambiguous, especially when the interaction is combined with a non-zero main effect.
-ART is more sensitive to these issues than other methods and may even inflate Type I error rates in scenarios where interpretation problems are not expected.
-More broadly, main effects should be interpreted with caution in the presence of strong interactions.
+When strong interaction effects are present, the interpretation of main effects becomes especially problematic, particularly when those interactions co-occur with nonzero main effects.
+ART may even inflate Type I error rates in scenarios where such interpretation issues would not ordinarily be expected.
+More generally, these results underscore that main effects should be interpreted with great caution in the presence of strong interactions.
 
@@
 Our results presented in @fig-median-main demonstrate that median alignment (ART-MED) --- or at least our implementation of the method --- is not appropriate for testing main effects.
-Although Type I error rates are now lower for the Cauchy distribution compared to the original method, they are still above nominal levels.
-In addition, they are significantly higher for all other distributions.
+Although Type I error rates are now lower for certain configurations (e.g., for the Cauchy distribution), they are higher in others.
 
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Appendix_test_4x3_ART-Median"
+prefix <- "Type_I_art_median"
 alpha <- .05
@@
 
-distributions <- c("lnorm", "exp", "cauchy", "poisson", "binom", "likert5B")
+distributions <- c("lnorm", "exp", "cauchy", "poisson", "binom", "likert")
 dnames <- c("Log-normal", "Exponential", "Cauchy", "Poisson", "Binomial", "Ordinal (5 levels)")
-df <- readlyData(prefix, alpha, 1, distributions, dnames, methods = methods)
+df <- readly_data(prefix, alpha, 1, distributions, dnames, methods = methods)
 
-plotlyError(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 64, cbPalette = palette, symbols = symbols)
+plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 103, cbPalette = palette, symbols = symbols)
 ```
@@
 **Interaction effects**.
-In contrast, median alignment works surprisingly well for interactions, correcting deficiencies of ART, especially when main effects are absent or weak.
-Despite this improved performance, we cannot recommend using the method because it still cannot compete with INT.
-Additionally, its advantages over parametric ANOVA are only apparent for the Cauchy distribution.
+In contrast, median alignment performs surprisingly well for interactions, correcting several deficiencies of ART, particularly when main effects are absent or weak.
+It appears to resolve ART's issues with discrete distributions and performs very well under the Cauchy distribution.
+
+Despite this improved performance, we do no recommend the method.
+It continues to struggle with skewed distributions, and its advantages over parametric ANOVA are essentially confined to the Cauchy distribution.
 
@@
 
-plotlyError(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 64, cbPalette = palette, symbols = symbols)
+plotly_error(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 103, cbPalette = palette, symbols = symbols)
 ```
@@
 **Interactions under parallel main effects**.
-We also evaluate Type I error rates for interactions in the presence of parallel main effects, where we caution readers about intepretation issues.
-We observe again that median alignment improves ART's bevahior across all distributions.
-However, as with PAR, its definition of the null hypothesis aligns with the scale of responses rather than with the model parameters.
+We also examine the behavior of the methods for interactions in the presence of parallel main effects, where interpretation issues warrant particular caution.
+Once again, median alignment improves ART’s performance across all distributions, with its positive rates now approaching those of PAR.
 
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- readlyData(prefix, alpha, 0, distributions, dnames, methods = methods)
-plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105, cbPalette = palette, symbols = symbols)
+df <- readly_data(prefix, alpha, 0, distributions, dnames, methods = methods)
+plotly_error(df, xlab = "magnitude of main effects", ytitle = 'Positives (%)', var = "rateX1X2", xvar = "effectX1", max = 105, cbPalette = palette, symbols = symbols)
 
@@
 Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$.
-*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
-A different definition may lead to different results.*
+*Note: These values correspond to Type I error rates under the null hypothesis $a_{12} = 0$; alternative definitions of the null hypothesis may lead to different results.*
 :::
@@
 ### Conclusion
-Using median instead of mean alignment in ART significantly improves the method's performance for testing interactions across all the distributions we examined.
-However, we cannot recommend it, as ART remains less robust than INT, and its advantages over PAR are unclear.
-Moreover, it is not evident how to apply median alignment for testing main effects --- using medians with the alignment procedure of @wobbrock:2011 leads to extremely high error rates.
+Using median rather than mean alignment in ART substantially improves the method's performance for testing interactions across all distributions we examined.
+Nevertheless, we cannot recommend the approach.
+ART remains less robust than other rank-based methods, and its advantages over PAR are unclear.
+Moreover, it is not evident how median alignment should be applied to the testing of main effects --- simply using medians within the alignment procedure of @wobbrock:2011 leads to extremely high error rates.
 

### 2025-12-28 appendix_II.qmd
working on appendix

@@
 
-## Absence of spurious effects when $a_2=0$ and $a_{12} = 0${#effects-proof}
+## On spurious effects when $a_2=0$ and $a_{12} = 0${#effects-proof}
 Let $x_1$ and $x_2$ be two independent variables, and suppose the conditional distribution of the response $Y | (x_1, x_2)$ depends on these variables only through the linear predictor:
@@
 $$
-\mu(x_1, x_2) = E(Y|x_1, x_2) = h(\ell(x_1, x_2))
+\mu(x_1, x_2) = E[Y|x_1, x_2] = h(\ell(x_1, x_2))
 $$
@@
 so that 
-$$Y | (x_1, x_2) \sim Lognormal(a_0 + a_1 x_1, \sigma^2)$$
+$$Y | (x_1, x_2) \sim \mathcal{LogN}(a_0 + a_1 x_1, \sigma^2)$$
 
@@
 In particular, the conditional mean
-$$\mu(x_1,x_2) = E(Y|x_1, x_2) = exp(a_0 + a_1 x_1 + \frac{1}{2}\sigma^2)$$
+$$\mu(x_1,x_2) = E[Y|x_1, x_2] = \text{exp}(a_0 + a_1 x_1 + \frac{1}{2}\sigma^2)$$
 remains identical across all levels of $x_2$.
@@
 \begin{aligned}
-Y = d\ \ \text{if}\ \ \tau_{d-1} < Y^* \leq \tau_k, 
+Y = d\ \ \text{if}\ \ \tau_{d-1} < Y^* \leq \tau_d, 
 \qquad d = 1,2 ... D
@@
 
-<!--
-In conclusion, $T$ does not introduce a main effect of $x_2$ or an interaction between $x_1$ and $x_2$ when $a_2=0$ and $a_{12}=0$.
+## Calculation of location parameters {#mu}
+We recall the linear predictor used to model nonlinear relationships (see the [article](index.html#statistical-modeling)):
 
-### ART introducing spurious effects 
-ART differs fundamentally from simple monotone transformations.
-Instead of transforming $Y$ directly, ART first constructs aligned responses by **subtracting and adding sample means on the response scale**, for example:
+$$ 
+\ell_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j}
+$$
+where $s_k \sim \mathcal{N}(0, \sigma_s^2)$. Our goal is to determine $\mu$ such as the expected value of the response equals a pre-specified target value $E[Y_{ijk}] = M$.
 
-$$Y^{(B)} = Y - \bar{Y}_{ij} + \bar{Y}_{.j} - \bar{Y}_{..}$$
-or analogous expressions for interactions, and only then applies ranking.
-The conditional distribution $Y^{(B)} | (x_1, x_2)$ now depends on $x_2$ through the arithmetic mean components even though $a_2 = 0$ and $a_{12}=0$.
+To simplify notation, we define the fixed-effects component
+$$\phi_{ij} = a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j}$$
 
-These alignment terms can vary across levels of $x_2$ due to sampling variability alone, and depending on $h$, they can acquire systematic shifts across levels of $x_2$ or combinations of $(x_1, x_2)$, even when the true conditional distribution does not.
-in particular, the term $\bar{Y}_{.j} - \bar{Y}_{..}$ is constant within each level of $x_2$ and shifts all aligned observations for that level either upward or downward.
+so that
 
-After ranking, observations from one level of $x_2$ tend to receive systematically higher or lower ranks. Our results confirm this problem.
--->
+$$ 
+\ell_{ijk} = \mu + s_k + \phi_{ij}
+$$
+
+**A useful identity.** Let $X$ be a continuous random variable with probability density function $f_X(x)$. For any function $g$, its expectation is defined as
+
+$$
+E[g(X)] = \int\limits_{-\infty}^\infty g(x) f_X(x) dx
+$$
+
+If $g(x) = e^x$, then
+
+$$
+E[e^X] = \int\limits_{-\infty}^\infty e^x f_X(x) dx
+$$
+
+When $X \sim \mathcal{N}(0, \sigma^2)$, substituting the normal probability density function yields
+
+$$
+E[e^X] = \int\limits_{-\infty}^\infty e^x \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\frac{x^2}{2\sigma^2}} dx = e^{\frac{1}{2}\sigma^2}
+$$ {#eq-useful}
+
+### Log-normal model
+When responses are log-normally distributed,
+
+$$
+\begin{aligned}
+log(Y_{ijk}) = \ell_{ijk} + \varepsilon_{ijk},
+\qquad \varepsilon_{ijk} \sim \mathcal{N}(0, \sigma^2)
+\end{aligned} 
+$$
+
+so that
+
+$$ 
+Y_{ijk} = e^{\ell_{ijk} + \varepsilon_{ijk}} = e^{\mu + s_k + e_{ij} + \varepsilon_{ijk}} = e^{\mu} \cdot e^{\phi_{ij}} \cdot e^{s_k} \cdot e^{\varepsilon_{ijk}} 
+$$
+
+Taking expectations and using independence of components,
+
+$$
+E[Y_{ijk}] = e^\mu \cdot E[e^{\phi_{ij}}] \cdot E[e^{s_k}] \cdot E[e^{\varepsilon_{ijk}}] = M   
+$$
+
+Solving for $\mu$ gives
+
+$$
+\mu = log\ M - log\ E[e^{\phi_{ij}}] - log\ E[e^{s_k}] - log\ E[e^{\varepsilon_{ijk}}]
+$$
+
+Because $s_k \sim \mathcal{N}(0, \sigma_s^2)$ and $\epsilon_{ijk} \sim \mathcal{N}(0, \sigma^2)$, @eq-useful implies
+
+$$
+\begin{aligned}
+E[e^{s_k}] = e^{\frac{1}{2}\sigma_s^2},
+\qquad E[e^{\varepsilon_{ijk}}] = e^{\frac{1}{2}\sigma^2}
+\end{aligned} 
+$$
+
+Furthermore, since $\phi_{ij}$ is deterministic,
+
+$$E[e^{\phi_{ij}}] = \frac{1}{m_1 m_2} \sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2} e^{\phi_{ij}}$$
+where $m_1$ and $m_2$ are the number of levels of $X_1$ and $X_2$, respectively.
+
+Therefore,
+
+$$
+\mu = log\ M - log(\frac{1}{m_1 m_2} \sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2} e^{\phi_{ij}}) - \frac{1}{2}\sigma_s^2 - \frac{1}{2}\sigma^2
+$$ {#eq-mu-lognornal}
+
+### Exponential model
+For exponential responses with a log link,
+
+$$
+Y_{ijk} \sim \text{Exp}(\lambda_{ijk} = e^{-\ell_{ijk}})
+$$
+
+The expected value is
+
+$$
+E[Y_{ijk}] = E[1/\lambda_{ijk}] = E[e^{\ell_{ijk}}]  =  e^{\mu} \cdot E[e^{\phi_{ij}}] \cdot E[e^{s_k}]
+$$
+
+which, as before, it leads to the solution
+
+$$
+\mu = log\ M - log(\frac{1}{m_1 m_2} \sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2} e^{\phi_{ij}}) - \frac{1}{2}\sigma_s^2
+$$ {#eq-mu-exp}
+
+### Poisson model
+For Poisson responses with a log link,
+$$
+Y_{ijk} \sim \mathcal{Pois}(\lambda_{ijk} = e^{\ell_{ijk}})
+$$
+
+The expected value is
+
+$$
+E[Y_{ijk}] = E[\lambda_{ijk}] = E[e^{\ell_{ijk}}]  =  e^{\mu} \cdot E[e^{\phi_{ij}}] \cdot E[e^{s_k}]
+$$
+which leads to the same solution as for the exponential distribution:
+$$
+\mu = log\ M - log(\frac{1}{m_1 m_2} \sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2} e^{\phi_{ij}}) - \frac{1}{2}\sigma_s^2
+$$ {#eq-mu-poisson}
+
+### Binomial model
+For binomial responses with a logit link,
+
+$$
+Y_{ijk} \sim \mathcal{B}(\kappa, p_{ijk} = \frac{1}{1 + e^{-\ell_{ijk}}})
+$$
+
+The expected value is now
+
+$$E[Y_{ijk}] = \kappa \cdot E[p_{ijk}] = 
+\frac{\kappa}{m_1 m_2} \sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2}
+\int_{-\infty}^{\infty}
+\frac{1}{1 + e^{-(\mu + \phi_{ij} + s)}}
+\cdot
+\frac{1}{\sqrt{2\pi}\sigma_s}
+e^{-\frac{s^2}{2\sigma_s^2}}\, ds
+= M
+$$ {#eq-mu-binomial}
+because we cannot separate the sums from the integral. As no close form exists to analytically derive $\mu$, we create the function
+
+$$F(\mu) = \frac{\kappa}{m_1 m_2} \sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2}
+\int_{-\infty}^{\infty}
+\frac{1}{1 + e^{-(\mu + \phi_{ij} + s)}}
+\cdot
+\frac{1}{\sqrt{2\pi}\sigma_s}
+e^{-\frac{s^2}{2\sigma_s^2}}\, ds
+- M$$
+
+and numerically find its root using R's ``uniroot()`` function.
+
+### Cauchy model
+Since the Cauchy distribution has no finite mean, we target the median instead:
+
+$$\text{Median}(Y_{ijk}) = \ell_{ijk} = \mu + \phi_{ij} = M$$
+
+Because our coding is centered, $\sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2} \phi_{ij} = 0$. Therefore,
+
+$$
+\mu  = M
+$$
 

### 2025-12-28 sandbox.qmd
working on appendix

--- 2025-12-20 sandbox.qmd
@@
 ```
-### Type I errors under unequal variances
-Many statistical procedures assume equal variances across all levels of each independent variable, or more strictly, across all possible pairs of factor levels (the latter is known as the sphericity assumption in repeated-measures ANOVA).
-Nonparametric tests are often mistakenly assumed to be free of such assumptions; however, as discussed earlier, this is generally incorrect.
 
-Our next experiment evaluates the behavior of the four methods under conditions of unequal variances, focusing on the normal distribution and four ordinal distributions.
-For ordinal outcomes, we control variances at the level of the latent variable $Y^*$.
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+library(dplyr)
+library(tidyverse)
 
-We examine three two-factor designs: (i) a $4 \times 3$ within-subjects design; (ii) a $2 \times 3$ between-subjects design; and (iii) a $2 \times 4$ mixed design.
-We set all effects to zero and vary the ratio $r_{sd}$ of standard deviations between the levels of the first factor $X_1$.
-@fig-unequal shows the five ratios tested when $X_1$ has two levels.
-When $X_1$ has additional levels (as in the $4 \times 3$ design), their standard deviations are randomly drawn from a range bounded by the standard deviations of the first two levels.
+source("utils-data-reading.R")
+source("utils-plotlying.R")
+```
 
-::: {#fig-unequal}
-```{r, echo=FALSE, message=FALSE, fig.height=1.5, fig.width = 9, warning=FALSE}
-source("distributions_plots.R")
-plot_unequal_variances()
+```{=html}
+<style>
+.math.inline .MathJax  {
+  font-size: 105% !important;
+}
+</style>
 ```
-Varying the ratio $r_{sd}$ of standard deviations across the levels of $X_1$. The plots show examples of population distributions for factors with two categorical levels.    
+
+We present complementary results and new experiments that investigate additional scenarios.
+We also compare INT and RNK with other nonparametric methods.
+Unless explicitly mentioned in each section, we follow the experimental methodology presented in the main article.
+At the end of each section, we summarize our conclusions.
+
+## Generalizations of nonparametric tests {#generalized}
+
+Finally, we evaluate the generalizations of nonparametric tests recommended by Lüpsen [-@luepsen:2018; -@luepsen:2023] as implemented in his *np.anova* function [@luepsen_R].
+Specifically, we examine the generalization of the van der Waerden test (VDW) and the generalization of the Kruskal-Wallis and Friedman tests (KWF).
+
+These implementations require random slopes to be included in the error term of the model.
+Concretely, we use ``Error(Subject/(X1*X2))`` for the two-factor within-participants design and ``Error(Subject/X2)`` for the mixed design.
+We use the same modeling approach for all other methods to ensure comparability.
+
+**Type I error rates: Main effects**.
+@fig-vdWaerden-designs-main shows the Type I error rates for the main effect of $X_2$.
+While all methods perform well under the within-subjects and mixed designs, the error rates of VDW and KWF drop sharply as the effect of $X_1$ increases under the between-subjects design.
+As shown below, this behavior reflects a severe loss of power for these two methods in these conditions.
+
+::: {#fig-vdWaerden-designs-main}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Type_I_vdWaerden"
+alpha = 0.05
+
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+methods = c("RNK", "INT", "VDW", "KWF")
+palette = c("#E69F00", "#009E73", "#FF70AB", "#888888")
+
+df <- read_data(prefix, alpha, effectType = 1, distributions, methods = methods)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 8.1, nticks=8, cbPalette = palette)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
 :::
 
-**Main effects**.
-We begin by investigating how the four methods detect main effects of $X_1$.
-Interpreting results in this setting is challenging because the underlying populations differ in variance while sharing the same means (and medians).
-For ordinal scales, differences in variance may also manifest as differences in observed means [@Liddell:2018].
-In this experiment, we define the null hypothesis in terms of mean differences on the latent variable, but we emphasize that the results should be interpreted with caution.
+**Type I error rates: Interactions**.
+@fig-vdWaerden-designs-interaction-1 presents the Type I error rates for the interaction in the presence of a single main effect.
+Again, the error rates of VDW and KWF decrease rapidly, now in both the between-subjects and mixed designs.
+We also observe inflation of the error rates of RNK and INT for large effects under discrete distributions.
 
-@fig-unequal-main-1 presents the results.
-Under normal distributions, and depending on the experimental design, the different methods either slightly inflate or slightly deflate Type I error rates as $r_{sd}$ increases, with no method clearly outperforming the others.
-For ordinal scales, larger values of $r_{sd}$ primarily affect conditions with flexible thresholds.
-This effect is particularly pronounced for PAR, although all methods are affected.
+::: {#fig-vdWaerden-designs-interaction-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- read_data(prefix, alpha, effectType = 2, distributions, methods = methods)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
 
-::: {#fig-unequal-main-1}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Type_I_heteroscedacity"
-distributions = c("norm", "likert_5", "likert_5_flex", "likert_7", "likert_7_flex")
-dnames = c("Normal", "Equidistant (5 levels)", "Flexible (5 levels)", "Equidistant (7 levels)", "Flexible (7 levels)")
-
-df <- read_data(prefix, alpha = 0.05, effectType = 1, distributions)
-df <- reshape_by_design(df, dnames, effectvars = c("sdratio", "effectX1","effectX2","effectX1X2"))
-plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1", xvar = "sdratio", max = 41)
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX2", max = 8.1, nticks=8, cbPalette = palette)
 ```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as the ratio of standard deviations $r_{sd}$ for $X_1$ increases ($n=20$).
-When $r_{sd} > 1$, the underlying populations are no longer identical, and different interpretations of the null hypothesis may lead to different outcomes.
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_2$ of the main effect of $X_2$ ($n = 20$) 
 :::
 
-We also examine the influence of unequal variances across the levels of $X_1$ on the Type I error rate for $X_2$, as shown in @fig-unequal-main-2.
-In this case, there is no ambiguity in the interpretation of effects because the populations defined by $X_2$ are identical.
+@fig-vdWaerden-designs-interaction-2 shows the results when both $a_1$ and $a_2$ increase.
+In these settings, all four methods struggle as effect magnitudes grow, though their behavior varies across distributions and experimental designs.
+Depending on the configuration, the methods either inflate positive rates or deflate them below nominal levels.
+INT appears to be the most stable method for continuous distributions, but its positive rates increase more rapidly under discrete distributions.
 
-The Type I error rates of PAR, RNK, and INT appear largely unaffected by increases in $r_{sd}$.
-In contrast, ART yields high error rates, surpassing acceptable levels even under normal distributions (see, for example, the $2 \times 4$ mixed design).
-Thus, unequal variances pose a more serious problem for ART than for the other methods, as the method's sensitivity propagates across factors.
-This provides further evidence that ART's alignment procedure leads to confounding across effects.
-
-::: {#fig-unequal-main-2}
+::: {#fig-vdWaerden-designs-interaction-2}
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX2", xvar = "sdratio", max = 41)
+df <- read_data(prefix, alpha, effectType = 0, distributions, methods = methods)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effects", ytitle = 'Positives (%)', var = "rateX1X2", xvar = "effectX1", max = 104, nticks=8, cbPalette = palette)
 ```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as the ratio of standard deviations $r_{sd}$ across the levels of $X_1$ increases ($n=20$)
+Positive rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ ($n = 20$).
+*Note: These values correspond to Type I error rates under the null hypothesis $a_{12} = 0$; alternative definitions of the null hypothesis may lead to different results.*
 :::
 
-**Interaction effects**.
-Finally, we evaluate Type I error rates for the interaction, as shown in @fig-unequal-interaction.
-ART exhibits a similar pattern as before, with error rates increasing as $r_{sd}$ grows.
-We also observe some inflation of Type I error rates for the other three methods under the $4 \times 3$ within-subjects design, albeit to a significantly lesser degree.
+**Power: Main effects**.
+@fig-vdWaerden-power-main-1 presents power results for detecting the main effect of $X_1$.
+Although differences are generally small, INT emerges as the most powerful method in most configurations.
+In several cases (e.g., log-normal distributions in the within-subjects and mixed designs), the power of VDW and KWF lags behind that of RNK and INT.
 
-::: {#fig-unequal-interaction}
+::: {#fig-vdWaerden-power-main-1}
+<div id="plot-power-vd.1">
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1X2", xvar = "sdratio", max = 42)
+prefix <- "Power_vdWaerden"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+
+data <- read_data(prefix, alpha = .05, effectType = 3, distributions, methods = methods)
+df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX1", max = 104,  ytitle = 'Power (%)', cbPalette = palette)
 ```
-Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as the ratio of standard deviations $r_{sd}$ across the levels of $X_1$ increases ($n=20$)
+</div>
+
+<div id="plot-power-vd.2">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- data %>% arrange(design,family,effectX1,rateX1)  %>% group_by(design,family,effectX1) %>% mutate(rank = rank(rateX1))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power - ranking', cbPalette = palette)
+```
+</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-vd.2").classList.add("hidden"); }); </script>
+
+<button id="btnvd1" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
+    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
+  onclick="const plot = document.getElementById('plot-power-vd.1');
+  plot.classList.toggle('hidden');
+  document.getElementById('plot-power-vd.2').classList.toggle('hidden');
+  btnvd1.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
+">
+  show ranking
+</button>
+
+Power ($\alpha = .05$) for detecting the **main effect of $X_1$** as a function of the magnitude of effect $a_1$ ($n=20$).
 :::
+
+We also report power results for the second factor $X_2$, which confirm that VDW and KWF offer no clear advantage, particularly when compared with INT.
+
+::: {#fig-vdWaerden-power-main-2}
+<div id="plot-power-vd.3">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Power_vdWaerden"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+
+data <- read_data(prefix, alpha = .05, effectType = 4, distributions, methods = methods)
+df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX2", max = 104,  ytitle = 'Power (%)', cbPalette = palette)
+```
+</div>
+
+<div id="plot-power-vd.4">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- data %>% arrange(design,family,effectX2,rateX2)  %>% group_by(design,family,effectX2) %>% mutate(rank = rank(rateX2))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power - ranking', cbPalette = palette)
+```
+</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-vd.4").classList.add("hidden"); }); </script>
+
+<button id="btnvd2" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
+    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
+  onclick="const plot = document.getElementById('plot-power-vd.3');
+  plot.classList.toggle('hidden');
+  document.getElementById('plot-power-vd.4').classList.toggle('hidden');
+  btnvd2.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
+">
+  show ranking
+</button>
+
+Power ($\alpha = .05$) for detecting the **main effect of $X_2$** as a function of the magnitude of effect $a_2$ ($n=20$).
+:::
+
+**Power: Interactions**.
+We also report power results for detecting interaction effects in @fig-vdWaerden-power-interaction.
+These results further confirm the overall advantage of INT across all designs.
+
+::: {#fig-vdWaerden-power-interaction}
+<div id="plot-power-vd.5">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+prefix <- "Power_vdWaerden"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+
+data <- read_data(prefix, alpha = .05, effectType = 5, distributions, methods = methods)
+df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1X2", max = 104,  ytitle = 'Power (%)', cbPalette = palette)
+```
+</div>
+
+<div id="plot-power-vd.6">
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- data %>% arrange(design,family,effectX1X2,rateX1X2)  %>% group_by(design,family,effectX1X2) %>% mutate(rank = rank(rateX1X2))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power - ranking', cbPalette = palette)
+```
+</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-vd.6").classList.add("hidden"); }); </script>
+
+<button id="btnvd3" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
+    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
+  onclick="const plot = document.getElementById('plot-power-vd.5');
+  plot.classList.toggle('hidden');
+  document.getElementById('plot-power-vd.6').classList.toggle('hidden');
+  btnvd3.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
+">
+  show ranking
+</button>
+
+Power ($\alpha = .05$) for detecting the **interaction effect $X_1 \times X_2$** as a function of the magnitude of effect $a_{12}$ ($n=20$).
+:::

### 2025-12-28 text/50-ExperimentalMethod.qmd
working on appendix

@@
 
- - $\mu$ is the grand mean
+ - $\mu$ is the intercept
 
@@
 @fig-effects-normal presents normal distributions ($\sigma^2 = 1$) generated across the range of values of $a_1$ used to evaluate Type I error rates.
-In these examples, we illustrate effects for a factor with three categorical levels and, for clarity, disregard between-subject variability ($\sigma_k^2 = 0$).
+In these examples, we illustrate effects for a factor with three categorical levels and, for clarity, disregard between-subject variability ($\sigma_s^2 = 0$).
 
@@
 
-**Grand mean and variance parameters.** We select the mean term $\mu$ to match a target location of the responses.
-@tbl-locations reports the target location for each distribution, which is typically the overall mean (or the median in the case of the Cauchy distribution).
-We describe how $\mu$ is derived from these parameters in [Appendix II](appendix_II.html#mu).
+**Intercept and variance parameters.** Fixing the intercept $\mu$ directly in a nonlinear model generally leads to arbitrary population-level response distributions, because the marginal mean depends on the link function, the response distribution, the variance of the random effects, and the magnitude of the fixed effects.
+We therefore calibrate the intercept $\mu$ so that the induced location of the responses matches a pre-specified target value.
+@tbl-locations reports the target location for each distribution, which is typically the overall mean in all cases except for the Cauchy distribution, where the median is used.
+We detail how $\mu$ is analytically derived from these target values in [Appendix II](appendix_II.html#mu).
 
@@
 ```
-Different ratio distributions generated with $a_1 = 2$ when $X_1$ has three categorical levels (here, $\sigma_k^2 = 0$).
+Different ratio distributions generated with $a_1 = 2$ when $X_1$ has three categorical levels (here, $\sigma_s^2 = 0$).
 :::
 
-In practice, we randomly choose the standard deviation $\sigma_k^2$ --- which determines between-subject variability --- from the interval $[0.1, 0.5]$, drawing each value from a uniform distribution.
+In practice, we randomly choose the standard deviation $\sigma_s^2$ --- which determines between-subject variability --- from the interval $[0.1, 0.5]$, drawing each value from a uniform distribution.
 

### 2025-12-28 text/63-Results-UnequalVariance.qmd
working on appendix

@@
 For ordinal scales, differences in variance may also manifest as differences in observed means [@Liddell:2018].
-In this experiment, we define the null hypothesis in terms of mean differences on the latent variable, but we emphasize that the results should be interpreted with caution.
+To highlight these interpretation issues,we therefor label the observed rates of positives as *Positives (%)*.
+These rates can nonetheless be interpreted as Type I error rates if the null hypothesis is defined in terms of mean differences on the latent variable.
 
 @fig-unequal-main-1 presents the results.
-Under normal distributions, and depending on the experimental design, the different methods either slightly inflate or slightly deflate Type I error rates as $r_{sd}$ increases, with no method clearly outperforming the others.
+Under normal distributions, and depending on the experimental design, the different methods either slightly inflate or slightly deflate the positive rates as $r_{sd}$ increases, with no method clearly outperforming the others.
 For ordinal scales, larger values of $r_{sd}$ primarily affect conditions with flexible thresholds.
@@
 df <- reshape_by_design(df, dnames, effectvars = c("sdratio", "effectX1","effectX2","effectX1X2"))
-plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1", xvar = "sdratio", max = 41)
+plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", ytitle = 'Positives (%)', var = "rateX1", xvar = "sdratio", max = 41)
 ```
-Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as the ratio of standard deviations $r_{sd}$ for $X_1$ increases ($n=20$).
-When $r_{sd} > 1$, the underlying populations are no longer identical, and different interpretations of the null hypothesis may lead to different outcomes.
+Positive rates ($\alpha = .05$) for the **main effect of $X_1$** as the ratio of standard deviations $r_{sd}$ for $X_1$ increases ($n=20$).
+*Note: When $r_{sd} > 1$, the underlying populations are no longer identical, and different interpretations of the null hypothesis may lead to different outcomes.*
 :::

### 2025-12-28 text/65-Results-EffectSize.qmd
working on appendix

@@
 ```
-Boxplots summarizing the $\eta^2$ estimates obtained with each method from 300 simulated datasets ($n=20$).
+Boxplots summarizing the partial $\eta^2$ estimates obtained with each method from 300 simulated datasets ($n=20$).
 The population effect of $X1$ is null, and responses follow **log-normal distributions**.
@@
 In this broader range, all three rank-based methods --- although RNK to a lesser extent --- struggle to recover large effects, even when the underlying distribution is normal.
-However, we acknowledge that accurately estimating effects of such magnitudes (e.g., when Cohen's $f > 2$ or $\eta^2 > 0.8$) is of limited practical importance.
+However, we acknowledge that accurately estimating effects of such magnitudes (e.g., when Cohen's $f > 2$ or partial $\eta^2 > 0.8$) is of limited practical importance.
 

### 2025-12-28 text/66-Results-Interactions.qmd
working on appendix

@@
 In this scenario, the interpretation of an interaction effect becomes ambiguous in nonlinear models.
-Here, we define the null hypothesis in terms of the coefficients of the linear component of the model (see @eq-linear-model), following the recommendation of @Greene:2010 --- that is, $a_{12} = 0$.
+Therefore, we refrain from labeling positive rates as Type I error rates.
+However, if the null hypothesis is defined in terms of the coefficients of the linear component of the model (see @eq-linear-model), following the recommendation of @Greene:2010 --- that is, $a_{12} = 0$ --- then all these values can be interpreted as Type I error rates.
 
-**Ratio scales**. @fig-ratio-interaction-2 presents Type I error rates for ratio scales.
+**Ratio scales**. @fig-ratio-interaction-2 presents results for ratio scales.
 
@@
 #xlab <- TeX("\\text{main effect of factors }X_1\\text{ and }X_2\\text{ }(a_1 = a_2)")
-plotly_error(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
+plotly_error(df, xlab = "magnitude of main effects", ytitle = 'Positives (%)', var = "rateX1X2", xvar = "effectX1", max = 105)
 ```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$.
-*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
-A different definition may lead to different results.*
+Positive rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$.
+*Note: These values correspond to Type I error rates under the null hypothesis $a_{12} = 0$; alternative definitions of the null hypothesis may lead to different results.*
 :::
 
-Error rates become exceptionally high in many cases, reaching up to 100%.
+Positive rates become exceptionally high in many cases, reaching up to 100%.
 These results require careful interpretation.
-As discussed in [Section 4](#interpretation), interaction effects in nonlinear models can be interpreted in two distinct ways, and the results produced by the different methods are not alwyas consistent with our chosen definition.
+As discussed in [Section 4](#interpretation), interaction effects in nonlinear models can be interpreted in two distinct ways, and the results produced by the different methods are not always consistent with our chosen definition.
 We therefore examine the outcomes for each method in turn:
@@
 - *PAR*.
-Error rates are high for all non-normal distributions (with the exception of the Cauchy distribution) and increase further as sample size grows.
+Positive rates are high for all non-normal distributions (with the exception of the Cauchy distribution) and increase further as sample size grows.
 This behavior arises because the null hypothesis for this method is defined on the scale of the response variable, rather than on the scale of the linear model.
 
-- *RNK.* Error rates increase dramatically once main effects exceed a certain magnitude.
+- *RNK.* Positive rates increase dramatically once main effects exceed a certain magnitude.
 This well-known issue stems from the way rank transformations distort interaction effects (see @fig-interactions-rank), even when the underlying distributions are normal.
 
-- *INT.* It performs better than RNK in continuous distributions, with errors rates increasing only when $a_1$ and $a_2$ become large.
+- *INT.* It performs better than RNK in continuous distributions, with rates increasing only when $a_1$ and $a_2$ become large.
 However, under the Poisson and binomial distributions, INT performs worse than RNK.
@@
 
-- *ART.* ART maintains correct error rates only when the population distributions are normal.
-For all other distributions, error rates increase rapidly as effect sizes grow.
+- *ART.* ART maintains correct positive rates only when the population distributions are normal.
+For all other distributions, rates increase rapidly as effect sizes grow.
 Two factors may account for this behavior: (i) a lack of robustness to violations of the method’s assumptions, as also observed in our earlier experiments; and (ii) issues related to interaction interpretation.
@@
 **Ordinal scales.** @fig-ordinal-interaction-2 presents our results for ordinal data.
-No method maintains low error rates under all conditions.
+No method maintains low positive rates under all conditions.
 We note that ART consistently performs worse than PAR.
 INT performs worse than ART and PAR in one specific case ($a_2, a_2 = 8$ in a scale with 11 equidistant levels), but overall, it exhibits a better behavior than all other methods.
-Once again, the high error rates can be partly explained by insufficient statistical robustness and ambiguities in how the methods interpret interaction effects.
+Once again, the high positive rates can be partly explained by insufficient statistical robustness and ambiguities in how the methods interpret interaction effects.
 
@@
 df <- readly_data(prefix, alpha, 0, distributions, dnames)
-plotly_error(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
+plotly_error(df, xlab = "magnitude of main effects", ytitle = 'Positives (%)', var = "rateX1X2", xvar = "effectX1", max = 105)
 ```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$.
-*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
-A different definition may lead to different results.*
+Positive rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$.
+*Note: These values correspond to Type I error rates under the null hypothesis $a_{12} = 0$; alternative definitions of the null hypothesis may lead to different results.*
 :::
@@
 **Other designs.** Our findings in @fig-designs-interaction-2 further demonstrate that all methods struggle to accurately detect interactions.
-INT shows unusually low error rates in certain configurations (e.g., $2 \times 3$ and $2 \times 2 \times 2$), suggesting potential difficulty in detecting subtle interaction effects when main effects are large.
+INT shows unusually low rates in certain configurations (e.g., $2 \times 3$ and $2 \times 2 \times 2$), suggesting potential difficulty in detecting subtle interaction effects when main effects are large.
 RNK displays a similar pattern for the $2 \times 2 \times 2$ design.
-In addition, both RNK and INT exhibit higher error rates under discrete distributions, in some cases exceeding those observed for ART.
+In addition, both RNK and INT exhibit higher rates under discrete distributions, in some cases exceeding those observed for ART.
 
@@
 df <- reshape_by_design(df, dnames)
-plotly_error_by_design(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
+plotly_error_by_design(df, xlab = "magnitude of main effects", ytitle = 'Positives (%)', var = "rateX1X2", xvar = "effectX1", max = 105)
 ```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$).
-*Note: Type I errors are defined based on the null hypothesis $a_{12} = 0$.
-A different definition may lead to different results.*
+Positive rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$).
+*Note: These values correspond to Type I error rates under the null hypothesis $a_{12} = 0$; alternative definitions of the null hypothesis may lead to different results.*
 :::

### 2025-12-28 supplementary/_experimental-code-old/Appendix-ATS-power.R

@@ -13,8 +13,8 @@ library(lmerTest)
 # Also try the ANOVA-type statistic (ATS)
 # https://www.quantargo.com/help/r/latest/packages/nparLD/2.1/nparLD
 library(nparLD)
-# Parallel computation
 
+# Parallel computation
 library(foreach)
 library(doParallel)
 

### 2025-12-28 supplementary/_experimental-code-old/Appendix-ATS.R

@@ -18,7 +18,9 @@ library(nparLD)
 library(foreach)
 library(doParallel)
 
-source("data-generators.R") # Import the data-generation functions
+source("utils-data-generation.R")
+source("utils-analysis.R")
+source("utils-extra.R")
 
 getLmerPValues <- function(model, vars, lmer = TRUE) {
 	return(suppressMessages(anova(model)[vars, ifelse(lmer, 6, 5)])) # For aov models the table is different

### 2025-12-28 supplementary/experiments-code/appendix-experiment-ART-median.R

@@ -86,7 +86,7 @@ time <- system.time({
       foreach(n = Ns, .combine=rbind) %do% {
         foreach(effId = 1:nrow(effects), .combine=rbind) %do% {
           # The test function is parallelized (using multiple cores)
-          repeat_test_median(
+          repeat_test_custom(
             nlevels=design, 
             within=within, 
             n=n, 
@@ -95,7 +95,10 @@ time <- system.time({
             params=use_parameters(family),
             formula=formula,
             vars=vars,
-            iterations = R
+            iterations = R,
+            methods = c("PAR", "ART", "ART-MED", "INT"),
+            methods_alt = c("PAR", "ARTmean", "ARTmedian", "INT"),
+            compare_function = compare_p_values_median
           )
         }
       }

### 2025-12-28 supplementary/experiments-code/appendix-experiment-ATS-Type_I.R

@@ -1,0 +1,126 @@
+# Author: Theophanis Tsandilas, Dec 2025
+# Inria & Université Paris-Saclay
+
+# Experiment evaluating the Type I error rate of the ATS method
+# 4x3 within-subjects and 2x4 mixed-subjects
+# We fix n=20
+
+rm(list=ls())
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+library(tidyr)
+
+# Try the ANOVA-type statistic (ATS)
+# https://www.quantargo.com/help/r/latest/packages/nparLD/2.1/nparLD
+library(nparLD)
+
+source("utils-data-generation.R")
+source("utils-analysis.R")
+source("utils-extra.R")
+
+################################
+# Distribution parameters used in simulation
+params_norm   <- list(sigma_e = 1)
+params_lnorm  <- list(sigma_e = 1, mean_target = 1)
+params_exp    <- list(mean_target = 0.5)
+params_poisson <- list(mean_target = 3)
+params_binom  <- list(size = 10, p_target = 0.1)
+params_likert <- list(sigma_e = 1, levels = 5, flexible=TRUE)
+
+use_parameters <- function(family){
+  params <- switch(family, 
+    norm    = params_norm, 
+    lnorm   = params_lnorm, 
+    cauchy  = params_cauchy, 
+    exp     = params_exp, 
+    poisson = params_poisson, 
+    binom   = params_binom,
+    likert  = params_likert
+  )
+
+  # Choose a random standard deviation for the random subject effect between 0.1 and 0.5
+  params$sigma_s <- c(0.1, 0.5) # Specifies the min and max of a uniform range
+
+  params
+}
+
+# (1) 4 x 3 within-subjects, (2) 2 x 4 mixed design (one between-subjects and one repeated-measures factor)
+designs <- list(c(4,3), c(2,4))
+within <- list(c(1,1), c(0,1))
+
+formula = Y ~ X1*X2 + Error(factor(subject)) 
+vars = c("X1", "X2", "X1:X2") 
+
+# Continuous distributions (equal variance, full) and discrete distribution: binom (size = 10, prob = 0.1) and Poisson
+distributions <- c("norm", "lnorm", "exp", "binom", "poisson", "likert")
+
+# Various combinations of effects
+effects <- matrix(c(0, 0, 0,
+            0.5, 0, 0,
+            1, 0, 0,
+            2, 0, 0,
+            4, 0, 0,
[…]

### 2025-12-28 supplementary/experiments-code/appendix-experiment-ATS-power.R

@@ -1,0 +1,128 @@
+# Author: Theophanis Tsandilas, Dec 2025
+# Inria & Université Paris-Saclay
+
+# Experiment evaluating the Power of the ATS method
+# 4x3 within-subjects and 2x4 mixed-subjects
+# We fix n=20
+
+rm(list=ls())
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+library(tidyr)
+
+# Try the ANOVA-type statistic (ATS)
+# https://www.quantargo.com/help/r/latest/packages/nparLD/2.1/nparLD
+library(nparLD)
+
+source("utils-data-generation.R")
+source("utils-analysis.R")
+source("utils-extra.R")
+
+################################
+# Distribution parameters used in simulation
+params_norm   <- list(sigma_e = 1)
+params_lnorm  <- list(sigma_e = 1, mean_target = 1)
+params_exp    <- list(mean_target = 0.5)
+params_poisson <- list(mean_target = 3)
+params_binom  <- list(size = 10, p_target = 0.1)
+params_likert <- list(sigma_e = 1, levels = 5, flexible=TRUE)
+
+use_parameters <- function(family){
+  params <- switch(family, 
+    norm    = params_norm, 
+    lnorm   = params_lnorm, 
+    cauchy  = params_cauchy, 
+    exp     = params_exp, 
+    poisson = params_poisson, 
+    binom   = params_binom,
+    likert  = params_likert
+  )
+
+  # Choose a random standard deviation for the random subject effect between 0.1 and 0.5
+  params$sigma_s <- c(0.1, 0.5) # Specifies the min and max of a uniform range
+
+  params
+}
+
+# (1) 4 x 3 within-subjects, (2) 2 x 4 mixed design (one between-subjects and one repeated-measures factor)
+designs <- list(c(4,3), c(2,4))
+within <- list(c(1,1), c(0,1))
+
+formula = Y ~ X1*X2 + Error(factor(subject)) 
+vars = c("X1", "X2", "X1:X2") 
+
+# Continuous distributions (equal variance, full) and discrete distribution: binom (size = 10, prob = 0.1) and Poisson
+distributions <- c("norm", "lnorm", "exp", "binom", "poisson", "likert")
+
+# Various combinations of effects
+effects <- matrix(c(
+          0, 0, 0.5,
+          0, 0, 1,
+          0, 0, 1.5,
+          0, 0, 2,
[…]

### 2025-12-28 supplementary/experiments-code/appendix-experiment-VDW-Type_I.R

@@ -1,0 +1,138 @@
+# Author: Theophanis Tsandilas, Dec 2025
+# Inria & Université Paris-Saclay
+
+# Comparison of Type I error rates of RNK and INT methods against the multifactorial generalizations of the van der Waerden test 
+# and the Kruskal-Wallis and Friedman tests: http://www.uni-koeln.de/~luepsen/R/
+# 4x3 within-subjects, 2x3 within-subjects, and 2x4 mixed-subjects
+# We fix n=20
+
+rm(list=ls())
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+library(tidyr)
+
+source("np.anova.R") # From http://www.uni-koeln.de/~luepsen/R/
+
+source("utils-data-generation.R")
+source("utils-analysis.R")
+source("utils-extra.R")
+library("POSSA") # to get access to the get_p function
+
+################################
+# Distribution parameters used in simulation
+params_norm   <- list(sigma_e = 1)
+params_lnorm  <- list(sigma_e = 1, mean_target = 1)
+params_exp    <- list(mean_target = 0.5)
+params_poisson <- list(mean_target = 3)
+params_binom  <- list(size = 10, p_target = 0.1)
+params_likert <- list(sigma_e = 1, levels = 5, flexible=TRUE)
+
+use_parameters <- function(family){
+  params <- switch(family, 
+    norm    = params_norm, 
+    lnorm   = params_lnorm, 
+    cauchy  = params_cauchy, 
+    exp     = params_exp, 
+    poisson = params_poisson, 
+    binom   = params_binom,
+    likert  = params_likert
+  )
+
+  # Choose a random standard deviation for the random subject effect between 0.1 and 0.5
+  params$sigma_s <- c(0.1, 0.5) # Specifies the min and max of a uniform range
+
+  params
+}
+
+# (1) 4 x 3 within-subjects, (2) 2x3 within-subjects, and (3) 2 x 4 mixed design (one between-subjects and one repeated-measures factor)
+designs <- list(c(4,3), c(2,3), c(2,4))
+within <- list(c(1,1), c(0,0), c(0,1))
+
+# Notice that I include slope random effects in my models, which is required by the np.anova functions  
+formula1 = Y ~ X1*X2 + Error(subject/(X1*X2))
+formula2 = Y ~ X1*X2 
+formula3 = Y ~ X1*X2 + Error(subject/X2) 
+formulas <- c(formula1, formula2, formula3)
+
+vars = c("X1", "X2", "X1:X2") 
+
[…]

### 2025-12-28 supplementary/experiments-code/appendix-experiment-VDW-power-multieffect.R

@@ -1,0 +1,134 @@
+# Author: Theophanis Tsandilas, Dec 2025
+# Inria & Université Paris-Saclay
+
+# Comparison of the power of RNK and INT methods against the multifactorial generalizations of the van der Waerden test 
+# and the Kruskal-Wallis and Friedman tests: http://www.uni-koeln.de/~luepsen/R/
+# 4x3 within-subjects, 2x3 within-subjects, and 2x4 mixed-subjects
+# In this condition, we assess how an effect on X2 affects the power for detetign main effects on X1 and interactions X1 x X2
+# We fix n=20
+
+rm(list=ls())
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+library(tidyr)
+
+source("np.anova.R") # From http://www.uni-koeln.de/~luepsen/R/
+
+source("utils-data-generation.R")
+source("utils-analysis.R")
+source("utils-extra.R")
+library("POSSA") # to get access to the get_p function
+
+################################
+# Distribution parameters used in simulation
+params_norm   <- list(sigma_e = 1)
+params_lnorm  <- list(sigma_e = 1, mean_target = 1)
+params_exp    <- list(mean_target = 0.5)
+params_poisson <- list(mean_target = 3)
+params_binom  <- list(size = 10, p_target = 0.1)
+params_likert <- list(sigma_e = 1, levels = 5, flexible=TRUE)
+
+use_parameters <- function(family){
+  params <- switch(family, 
+    norm    = params_norm, 
+    lnorm   = params_lnorm, 
+    cauchy  = params_cauchy, 
+    exp     = params_exp, 
+    poisson = params_poisson, 
+    binom   = params_binom,
+    likert  = params_likert
+  )
+
+  # Choose a random standard deviation for the random subject effect between 0.1 and 0.5
+  params$sigma_s <- c(0.1, 0.5) # Specifies the min and max of a uniform range
+
+  params
+}
+
+# (1) 4 x 3 within-subjects, (2) 2x3 within-subjects, and (3) 2 x 4 mixed design (one between-subjects and one repeated-measures factor)
+designs <- list(c(4,3), c(2,3), c(2,4))
+within <- list(c(1,1), c(0,0), c(0,1))
+
+# Notice that I include slope random effects in my models, which is required by the np.anova functions  
+formula1 = Y ~ X1*X2 + Error(subject/(X1*X2))
+formula2 = Y ~ X1*X2 
[…]

### 2025-12-28 supplementary/experiments-code/appendix-experiment-VDW-power.R

@@ -1,0 +1,135 @@
+# Author: Theophanis Tsandilas, Dec 2025
+# Inria & Université Paris-Saclay
+
+# Comparison of the power of RNK and INT methods against the multifactorial generalizations of the van der Waerden test 
+# and the Kruskal-Wallis and Friedman tests: http://www.uni-koeln.de/~luepsen/R/
+# 4x3 within-subjects, 2x3 within-subjects, and 2x4 mixed-subjects
+# We fix n=20
+
+rm(list=ls())
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+library(tidyr)
+
+source("np.anova.R") # From http://www.uni-koeln.de/~luepsen/R/
+
+source("utils-data-generation.R")
+source("utils-analysis.R")
+source("utils-extra.R")
+library("POSSA") # to get access to the get_p function
+
+################################
+# Distribution parameters used in simulation
+params_norm   <- list(sigma_e = 1)
+params_lnorm  <- list(sigma_e = 1, mean_target = 1)
+params_exp    <- list(mean_target = 0.5)
+params_poisson <- list(mean_target = 3)
+params_binom  <- list(size = 10, p_target = 0.1)
+params_likert <- list(sigma_e = 1, levels = 5, flexible=TRUE)
+
+use_parameters <- function(family){
+  params <- switch(family, 
+    norm    = params_norm, 
+    lnorm   = params_lnorm, 
+    cauchy  = params_cauchy, 
+    exp     = params_exp, 
+    poisson = params_poisson, 
+    binom   = params_binom,
+    likert  = params_likert
+  )
+
+  # Choose a random standard deviation for the random subject effect between 0.1 and 0.5
+  params$sigma_s <- c(0.1, 0.5) # Specifies the min and max of a uniform range
+
+  params
+}
+
+# (1) 4 x 3 within-subjects, (2) 2x3 within-subjects, and (3) 2 x 4 mixed design (one between-subjects and one repeated-measures factor)
+designs <- list(c(4,3), c(2,3), c(2,4))
+within <- list(c(1,1), c(0,0), c(0,1))
+
+# Notice that I include slope random effects in my models, which is required by the np.anova functions  
+formula1 = Y ~ X1*X2 + Error(subject/(X1*X2))
+formula2 = Y ~ X1*X2 
+formula3 = Y ~ X1*X2 + Error(subject/X2) 
+formulas <- c(formula1, formula2, formula3)
+
+vars = c("X1", "X2", "X1:X2") 
+
[…]

### 2025-12-28 supplementary/experiments-code/appendix-experiment-single-factor-nonparametric.R

@@ -1,0 +1,116 @@
+# Author: Theophanis Tsandilas, Dec 2025
+# Inria & Université Paris-Saclay
+
+# Comparing single-factor non-parametric methods 
+# Testing single-factor designs with 2,3 or 4 levels (both within-subjects and between-subjects) 
+# The ratio_sd is the maximum ratio of standard deviations of levels of the factor
+
+rm(list=ls())
+
+# Parallel computation
+library(foreach)
+library(doParallel)
+library(tidyr)
+
+source("utils-data-generation.R")
+source("utils-analysis.R")
+source("utils-extra.R")
+
+################################
+# Distribution parameters used in simulation
+params_norm   <- list(sigma_e = 1)
+params_likert_5       <- list(sigma_e = 1, levels = 5, flexible=FALSE)
+params_likert_5_flex  <- list(sigma_e = 1, levels = 5, flexible=TRUE)
+params_likert_7       <- list(sigma_e = 1, levels = 7, flexible=FALSE)
+params_likert_7_flex  <- list(sigma_e = 1, levels = 7, flexible=TRUE)
+params_likert_11      <- list(sigma_e = 1, levels = 11, flexible=FALSE)
+params_likert_11_flex <- list(sigma_e = 1, levels = 11, flexible=TRUE)
+
+use_parameters <- function(family, ratio){
+  params <- switch(family, 
+    norm          = params_norm, 
+    likert_5      = params_likert_5, 
+    likert_5_flex = params_likert_5_flex, 
+    likert_7      = params_likert_7, 
+    likert_7_flex = params_likert_7_flex,
+    likert_11     = params_likert_11, 
+    likert_11_flex = params_likert_11_flex, 
+  )
+
+  # Choose a random standard deviation for the random subject effect between 0.1 and 0.5
+  params$sigma_s <- c(0.1, 0.5) # Specifies the min and max of a uniform range
+  params$ratio_sd = ratio
+
+  params
+}
+
+# Testing single-factor designs with 2,3 or 4 levels (both within-subjects and between-subjects) 
+designs <- list(2, 3, 4, 2, 3, 4)
+within <- list(1, 1, 1, 0, 0, 0)
+
+formula = Y ~ X1 + Error(factor(subject)) 
+vars = c("X1") 
+
+# Normal distribution + some ordinal scales
[…]

### 2025-12-28 supplementary/experiments-code/np.anova.R

@@ -1,0 +1,223 @@
+# This code was downloaded from http://www.uni-koeln.de/~luepsen/R/
+# As we only call the method with pseudo = F, the pseudorank library is not loaded.
+# Note that this library was removed from CRAN but can be obtained from the archive: https://cran.r-project.org/web/packages/pseudorank/index.html
+
+np.anova <- function(formel,data,method=0,compact=T,pseudo=F)
+{                        # data  :  dataframe 
+                         # formel:  anova-Modell
+                         # method:  0: generalized Kruskal-Wallis-Friedman tests
+                         #             including Iman & Davenport F-tests
+                         #          1: generalized van der Waerden tests
+                         #          2: Puri & Sen tests
+                         #          3: Puri & Sen tests with INT
+                         # compact: T: all tests in one table
+                         #          F: seperate tables for each error term (repeated measures only)
+                         # pseudo   T: use pseudoranks instead of ordinary ranks (for unequal ni)
+
+if (pseudo) library(pseudorank) # See: https://cran.r-project.org/web/packages/pseudorank/index.html
+                                               # check formula
+if (mode(formel)!="call") stop("invalid formula")
+if (class(data[,as.character(formel)[2]])=="factor") stop("invalid type of dependent variable")
+                                               # perform standard aov and check model
+aov.1 <- aov(formel,data)
+if (class(aov.1)[1]=="aov")     repm<-0
+if (class(aov.1)[1]=="aovlist") repm<-1
+                                               # extract varnames
+vars  <- all.vars(formel)
+vdep  <- vars[1]                               # dependent var
+nfact <- length(vars)-1
+if (repm==1) nfact <- length(vars)-2
+nfgrp <- nfact
+if (repm==1)                                   # params for repeated
+ { nam.aov  <- names(aov.1)
+   l.aov    <- length(nam.aov)
+   if (nam.aov[l.aov]=="Within") nam.aov <- nam.aov[-l.aov]
+   nerrterm <- length(nam.aov)-2               # n of error terms for repeated
[…]

### 2025-12-28 supplementary/experiments-code/utils-analysis.R

@@ -31,7 +31,7 @@ get_p_values <- function(model, vars, is_aov = TRUE) {
     res <- res  %>% # Get both the between- and within-subjects terms
         filter(term %in% vars) %>% # Filter the relevant rows
         arrange(match(term, vars)) %>% # Order them
-        select(p.value) # And select the p-value
+        dplyr::select(p.value) # And select the p-value
     return(res[[1]])
   } else {
     anov_object <- anova(model)

### 2025-12-28 utils-data-reading.R

@@ -29,7 +29,6 @@ read_data <- function(prefix, alpha = .05, effectType = 0,
 		df <- df[(df$effectX1X2 > 0 & df$effectX1 > 0) | (df$effectX1X2 == 0 & df$effectX1 == 0),]
 	}
 
-
 	if(is.na(alpha)) return(df)
 	else return(df[df$alpha == alpha,])
 }
@@ -101,16 +100,8 @@ reshape_by_design <- function(df, dnames = c("Normal", "Log-normal", "Exponentia
 	# Different column for each design
 	df <- reshape(df, idvar=c(groupvars, effectvars), timevar = "design", direction = "wide")
 	df <- df %>% group_by(family) %>%  mutate(family=dnames[cur_group_id()])
-	df$family <- factor(df$family, levels = dnames)
+	df$family <- factor(df$family, levels = dnames)		
 
 	df
 }
 
-
-#prefix <- "Type_I_4x3_ratio"
-#distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
-#dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
-
-#df <- read_data(prefix, alpha=0.05, effectType = 3, distributions)
-#df1 <- df %>% arrange(design,family,effectX1,rateX1)  %>% group_by(design,family,effectX1) %>% mutate(rank = rank(rateX1))
-#df <- as.data.frame(df1) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

### 2025-12-28 utils-plotlying.R

@@ -234,6 +234,53 @@ plotly_error_by_design_2 <- function(df, xlab = "magnitude of main effects", var
 	fig
 }
 
+# This is for alterative graphs for a different set of designs : 2x3, 2x4 (writing more generic code with plotly is painful)
+plotly_error_by_design_3 <- function(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", min = 0, max = 100, 
+	    ytitle = 'Type I errors (%)', cbPalette = c("#888888", "#E69F00", "#009E73", "#FF5E00"), nticks=6){
+	# aesthetics
+	symbols <- c("asterisk", "x", "star-diamond", "star-triangle-up")
+	margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
+
+	dnames <- levels(df$family)
+	# To be used as labels on the subplots
+	annots <- lapply(1:length(dnames), function(index){list(x = (index-0.5)/length(dnames), y = 1, text = dnames[index], xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F)})
+
+	menuItems <- c("4x3 within", "2x4 mixed")
+	yvarnames <- c(paste(var,"4x3",sep="."), paste(var,"2x4",sep="."))
+
+	createPlot <- function(data, dnames, symbols) {
+	  p <- plot_ly(data,  x = ~factor(eval(as.symbol(xvar))), y = ~100*eval(as.symbol(yvarnames[1])), color = ~method, symbol = ~method, symbols = symbols, colors = cbPalette, type = 'scatter', mode = 'lines+markers', marker = list(line = list(width = 2)), legendgroup = ~method, showlegend = unique(data$family == dnames[1]), visible = T) 
+	  p <- add_trace(p, data = data, x = ~factor(eval(as.symbol(xvar))), y = ~100*eval(as.symbol(yvarnames[2])), color = ~method, symbol = ~method, symbols = symbols, colors = cbPalette, type = 'scatter', mode = 'lines+markers', marker = list(line = list(width = 2)), legendgroup = ~method, showlegend = unique(data$family == dnames[1]), visible = F) %>%
+	  layout(
+	    legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5),
[…]

### 2025-12-29 appendix_I.qmd
still updating the appendix

@@
 ART’s robustness issues become less severe as log-normal distributions approach normality. However, even when variance is relatively small, Type I error rates can still become unacceptably large.
-
-## Log-normal distributions with equal variances
-As we explain in the article, skewed distributions naturally lead to unequal variances, with the standard deviation of response times typically proportional to the mean.
-Since ART confounds effects under conditions of unequal variance, an important question remains: *Does ART’s difficulty with continuous skewed distributions stem solely from issues of heteroscedasticity?*
-
-To address this, we present results from an additional experiment using log-normal distributions in which we directly control the population parameters of the observed distributions (rather than those of the latent variable) to ensure constant variances.
-Specifically, given a mean $\mu$ and a fixed standard deviation $\sigma$ for the responses, the corresponding parameters of the log-normal distribution are:
-
-$\sigma_{log} = \sqrt{log(1 + \frac{\sigma^2}{\mu^2})}$ and $\mu_{log} = \log(\frac{\mu^2}{\sqrt{\sigma^2 + \mu^2}})$
-
-We again consider a $4 \times 3$ within-subjects design with $n = 20$, setting the main effect of the first factor to $a_1 = 2$.
-We then measure the Power (effect of $X_1$) and Type I error rates (effects $X_2$ and $X_1 \times X_2$) of the methods across three standard deviations: $\sigma = 1$, $\sigma = 2$, and $\sigma = 3$.
-
-The following table presents our results.
-Once again, ART inflates Type I error rates for both the effect of $X_2$ and its interaction with $X_1$, with error rates increasing as the common standard deviation grows.
-In contrast, PAR shows a loss in statistical power, while RNK and INT remain robust.
-
-::: {#tbl-homoscedastic layout-ncol=3}
-|     | PAR | RNK | INT | ART |
-|:-----:|----:|----:|----:|----:|
-|$X_1$| 100 | 100 | 100 | 100 |
-|$X_2$| 5.0 | 4.9 | 5.4 | 7.6 |
-|$X_1\times X_2$| 3.6 | 4.2 | 4.6 | 5.7 |
-
-: $\sigma = 1$
-
-| PAR | RNK | INT | ART |
-|----:|----:|----:|----:|
-| 98.8 | 100 | 100 | 100 |
-| 4.0 | 5.1  | 5.3 | 12.1 |
-| 3.5 | 5.3 |  5.1 | 9.3  |
-
-: $\sigma = 2$
-
-| PAR | RNK | INT | ART |
-|----:|----:|----:|----:|
-| 94.0 | 100 | 100 | 100 |
-| 3.6 | 4.4  | 4.7 | 17.8 |
-| 3.0 | 4.7  | 5.0 | 11.2 |
-
-: $\sigma = 3$
-
-Percentage of positives (%) over 5000 repetitions ($\alpha = .05$). For $X_1$, values reflect statistical power. For $X_2$ and $X_1 \times X_2$, they represent Type I error rates.
-:::
-
-### Conclusion
-Heteroscedasticity alone cannot explain ART's failures under log-normal distributions.
-ART's alignment method is based on the assumption that the shape of the distribution (log-normal here) is identical across all factor levels.
-This is not the case in the above scenarios where distribution shapes change across the levels of $X_1$, causing the method to break down.
 
@@
 
-Finally, we evaluate the generalizations of nonparametric tests recommended by Lüpsen [-@luepsen:2018; -@luepsen:2023] as implemented in his *np.anova* function [@luepsen_R].
-Specifically, we evaluate the generalization of the van der Waerden test (VDW), and the generalization of the Kruskal-Wallis and Friedman test (KWF).
-Their implementation uses R's *aov* function and requires considering random slopes in the error term of the model, that is, using ``Error(s/(x1*x2))`` for the two-factor within-participants design and ``Error(s/x2)`` for the mixed design, where ``s`` is the subject identifier variable.
-We also used the *aov* function and the same formulation of the error term for all other methods.
+Finally, we evaluate the generalizations of nonparametric tests recommended by Lüpsen [-@luepsen:2018; -@luepsen:2023] as implemented in his `np.anova` function [@luepsen_R].
+Specifically, we examine the generalization of the van der Waerden test (VDW) and the generalization of the Kruskal-Wallis and Friedman tests (KWF).
+
+These implementations require random slopes to be included in the error term of the model.
+Concretely, we use ``Error(Subject/(X1*X2))`` for the two-factor within-participants design and ``Error(Subject/X2)`` for the mixed design.
+We use the same modeling approach for all other methods to ensure comparability.
 
 **Type I error rates: Main effects**.
-@fig-vdWaerden-designs-main presents Type I error rates for the main effect of $X_2$.
-While all methods perform well under the within-subjects and mixed designs, the error rates of VDW and KWF drop drastically as the effect of $X_1$ grows under the between-subjects design.
-We will see below that the power of these methods evaporates in these cases.
+@fig-vdWaerden-designs-main shows the Type I error rates for the main effect of $X_2$.
+While all methods perform well under the within-subjects and mixed designs, the error rates of VDW and KWF drop sharply as the effect of $X_1$ increases under the between-subjects design.
+As shown below, this behavior reflects a severe loss of power for these two methods in these conditions.
+We omit results for the other factor ($X_1$ as $a_2$ increases) as we observe very similar trends.
 
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Appendix_test-vdWaerden"
+prefix <- "Type_I_vdWaerden"
 alpha = 0.05
 
-distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
 dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
@@
 
-df <- readData(prefix, n = 20, alpha, effectType = 1, distributions, methods = methods)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+df <- read_data(prefix, alpha, effectType = 1, distributions, methods = methods)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
 
-plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 7.1, nticks=8, cbPalette = palette)
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 8.1, nticks=8, cbPalette = palette)
 ```
@@
 **Type I error rates: Interactions**.
-@fig-vdWaerden-designs-interaction-1 presents the Type I error rates for the interaction, in the presence of a single main effect.
-Once again, the error rates of VDW and KWF decrease rapidly in both the between-subjects and mixed designs.
-@fig-vdWaerden-designs-interaction-2 shows the results when both main effects are present.
-Under the within-subjects design, KWF is the poorest-performing technique.
-While VDW outperforms RNK, it remains inferior to INT.
-For the between-subjects and mixed designs, KFW and VDW yield low or near-zero error rates when main effects are large, likely due to their extreme loss of power in these scenarios.
+@fig-vdWaerden-designs-interaction-1 presents the Type I error rates for the interaction in the presence of a single main effect.
+Again, the error rates of VDW and KWF decrease rapidly, now in both the between-subjects and mixed designs.
+We also observe inflation of the error rates of RNK and INT for large effects under discrete distributions.
 
@@
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- readData(prefix, n = 20, alpha, effectType = 2, distributions, methods = methods)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+df <- read_data(prefix, alpha, effectType = 2, distributions, methods = methods)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
 
-plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX2", max = 7.1, nticks=8, cbPalette = palette)
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX2", max = 8.1, nticks=8, cbPalette = palette)
 ```
@@
 
+@fig-vdWaerden-designs-interaction-2 shows the results when both $a_1$ and $a_2$ increase.
+In these settings, all four methods struggle as effect magnitudes grow, though their behavior varies across distributions and experimental designs.
+Depending on the configuration, the methods either inflate positive rates or deflate them below nominal levels.
+INT appears to be the most stable method for continuous distributions, but its positive rates increase more rapidly under discrete distributions.
+
 ::: {#fig-vdWaerden-designs-interaction-2}
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
-dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
-
-df <- readData(prefix, n = 20, alpha, effectType = 0, distributions, methods = methods)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 104, nticks=8, cbPalette = palette)
+df <- read_data(prefix, alpha, effectType = 0, distributions, methods = methods)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_2(df, xlab = "magnitude of main effects", ytitle = 'Positives (%)', var = "rateX1X2", xvar = "effectX1", max = 104, nticks=8, cbPalette = palette)
 ```
-Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ ($n = 20$)
+Positive rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ ($n = 20$).
+*Note: These values correspond to Type I error rates under the null hypothesis $a_{12} = 0$; alternative definitions of the null hypothesis may lead to different results.*
 :::
@@
 **Power: Main effects**.
-@fig-vdWaerden-power-main-1 presents power results for detecting the main effect of $X_1$.
-For the within-subjects and mixed design, KFW and VDW exhibit a lower power than RNK and INT, but differences are generally small.
-Although VDW appears as powerful as INT under the between-subjects design, this only occurs when the effect of the second factor is zero.
-As shown in @fig-vdWaerden-power-main-1B, the power of both KFD and VDW drops drastically as the effect of $X_2$ increases for this design.
-
-::: {#fig-vdWaerden-power-main-1}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-alpha <- .05
-
-prefix <- "Appendix_test-vdWaerden-Power"
-distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
-dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
-
-df <- readData(prefix, n = 20, alpha, effectType = 3, distributions, methods = methods)
-
-df <- df %>% arrange(design,distr,effectX1,rateX1)  %>% group_by(design,distr,effectX1) %>% mutate(rank = rank(rateX1))
-df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-
-#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyPowerByDesign(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
-```
-Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_1$** as a function of the magnitude of effect $a_1$ ($n=20$). Hover over the charts to inspect their power ($\%$).
-:::
+We examine how power is affected by increasing the magnitude of the effect on the second factor.
+As shown in @fig-vdWaerden-power-main-1B, the power of all methods decreases, but KWF and VDW appear especially problematic under the between-subjects design (across all distributions) and the mixed design (for the log-normal and exponential distributions).
+INT once again emerges as the best-performing method, with the exception of the $2 \times 4$ mixed design under the Poisson and Binomial distributions, for which VDW and KWF exhibit a better behavior.
 
@@
 alpha <- .05
-prefix <- "Appendix_test-vdWaerden-Power-Multieffect"
+prefix <- "Power_vdWaerden_multieffect"
 
-df <- readData(prefix, n = 20, alpha, effectType = 3, distributions, methods = methods)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+df <- read_data(prefix, alpha, effectType = 3, distributions, methods = methods)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
 
-plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX2", ytitle ="Power (%)", max = 104, nticks=6, cbPalette = palette)
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX2", ytitle ="Power (%)", max = 104, nticks=6, cbPalette = palette)
 ```
-Power ($\alpha = .05$) to detect the **main effect of $X_1$** ($a_1 = 0.8$) as a function of the magnitude of effect $a_2$ of the main effect of $X_2$ ($n=20$).
+Power ($\alpha = .05$) for detecting the **main effect of $X_1$** ($a_1 = 0.8$) as a function of the magnitude of effect $a_2$ of the main effect of $X_2$ ($n=20$).
 :::
 
-@fig-vdWaerden-power-main-2 also presents results for the main effect of $X_1$, where we observe once again that the generalized tests cannot compete with the more powerful INT, or even with RNK.
-
-::: {#fig-vdWaerden-power-main-2}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Appendix_test-vdWaerden-Power"
-df <- readData(prefix, n = 20, alpha, effectType = 4, distributions, methods = methods)
-
-df <- df %>% arrange(design,distr,effectX2,rateX2)  %>% group_by(design,distr,effectX2) %>% mutate(rank = rank(rateX2))
-df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-
-#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyPowerByDesign(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
-```
-Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_2$** as a function of the magnitude of effect $a_2$ ($n=20$). Hover over the charts to inspect their power ($\%$).
-:::
-
-**Power: Interactions**. We also present results on the power of methods to detect interactions in @fig-vdWaerden-power-interaction, confirming the advantage of INT across all designs.
-@fig-vdWaerden-power-interactionB provides a clearer picture of how power is affected by the presence of a main effect.
-We observe that the power of all methods drops as the main effect of $X_2$ increases, but this trend is more pronounced for KFD and VDW, particularly for the between-subjects and mixed designs.
-
-::: {#fig-vdWaerden-power-interaction}
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- readData(prefix, n = 20, alpha, effectType = 5, distributions, methods = methods)
-
-df <- df %>% arrange(design,distr,effectX1X2,rateX1X2)  %>% group_by(design,distr,effectX1X2) %>% mutate(rank = rank(rateX1X2))
-df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-
-#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotlyPowerByDesign(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
-```
-Ranking of methods for their power ($\alpha = .05$) to detect the **interaction effect $X_1 \times X_2$** as a function of the magnitude of effect $a_{12}$ ($n=20$).
-Hover over the charts to inspect their power ($\%$).
-:::
+**Power: Interactions**.
+Figure @fig-vdWaerden-power-interactionB shows how the power for detecting the interation effect is affected by the presence of a main effect.
+We observe that the power of all methods decreases as the main effect of $X_2$ increases, but this decline is substantially more pronounced for KWF and VDW, particularly in the between-subjects and mixed designs.
 
@@
 alpha <- .05
-prefix <- "Appendix_test-vdWaerden-Power-Multieffect"
+prefix <- "Power_vdWaerden_multieffect"
 
-df <- readData(prefix, n = 20, alpha, effectType = 5, distributions, methods = methods)
-df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+df <- read_data(prefix, alpha, effectType = 5, distributions, methods = methods)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
 
-plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX2", ytitle ="Power (%)", max = 104, nticks=6, cbPalette = palette)
+plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX2", ytitle ="Power (%)", max = 104, nticks=6, cbPalette = palette)
 ```
-Power ($\alpha = .05$) to detect the **interaction effect** ($a_{12} = 1.5$) as a function of the magnitude of effect $a_2$ of the main effect of $X_2$ ($n=20$).
+Power ($\alpha = .05$) for detecting the **interaction effect** ($a_{12} = 1.5$) as a function of the magnitude of effect $a_2$ of the main effect of $X_2$ ($n=20$).
 :::
@@
 ### Conclusion
-Our results do not support Lüpsen's [-@luepsen:2018; -@luepsen:2023] conclusions.
-The behavior of the generalized nonparametric tests presents issues in numerous scenarios.
-While these tests exhibit lower error rates under specific conditions, this is due to a significant loss of power when other effects are at play.
-Therefore, we advise against the use of these methods.
+Our results do not support the conclusions of Lüpsen [-@luepsen:2018; -@luepsen:2023].
+The generalized nonparametric tests exhibit serious limitations across a wide range of scenarios.
+Although these methods sometimes lead to lower Type I error rates, this behavior is largely due to a substantial loss of statistical power when other effects are present.
+Consequently, we do not recommend the use of these methods.
 

### 2025-12-29 sandbox.qmd
still updating the appendix

--- 2025-12-20 sandbox.qmd
@@
 
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
 library(dplyr)
 library(tidyverse)
+
 source("utils-data-reading.R")
@@
 
-### Statistical power {#power}
-We evaluate the statistical power of the four methods on the three two-factor designs.
-Because there is a tradeoff between Type I and Type II errors, high power may simply reflect an inflated Type I error rate.
-Since a main effect on $X_1$ inflates ART's Type I error rates for $X_2$ and for their interaction (see previous sections), we therefore focus on single-effect scenarios.
+```{=html}
+<style>
+.math.inline .MathJax  {
+  font-size: 105% !important;
+}
+</style>
+```
 
-**Main effects**.
-We vary the magnitude of effect on $X_1$ and $X_2$ separately and assess each method's power to detect the corresponding effect at $\alpha = .05$.
-Specifically, we manipulate the parameters $a_1$ or $a_2$ over the range $0.2$ to $0.8$.
-Depending on the experimental design, this range allows us to simulate both low-power settings (e.g., power below $30\%$) and high-power settings (e.g., power above $80\%$).
+We present complementary results and new experiments that investigate additional scenarios.
+We also compare INT and RNK with other nonparametric methods.
+Unless explicitly mentioned in each section, we follow the experimental methodology presented in the main article.
+At the end of each section, we summarize our conclusions.
 
-@fig-power-main-1 presents the results for $X_1$.
-In many configurations, differences in power across methods are modest.
-To facilitate comparison, we also report method rankings; readers are encouraged to interact with the figures for precise power estimates.
+## ANOVA-type statistic (ATS) {#ATS}
+We compare PAR, RNK, and INT to the ANOVA-type statistic (ATS) [@Brunner_ATS:2001] for two-factor designs.
+We use its implementation in the R package *nparLD* [@nparLD], which does not support between-subjects designs.
+Thus, we only evaluate it for the $4 \times 3$ within-subjects and the $2 \times 4$ mixed designs.
 
-PAR achieves the highest power under the normal and exponential distributions, but its power is particularly low under the log-normal distribution.
-ART initially appears advantageous under the the log-normal, binomial, and ordinal distributions for small effects, likely reflecting inflated Type I errors under these conditions.
-As $a_1$ grows larger, ART's relative advantage diminishes, and it ultimately becomes the least powerful method.
-Finally, INT consistently outperforms RNK and ranks highly across all conditions.
+**Type I error rates: Main effects**.
+@fig-ATS-designs-main presents Type I error rates for the main effect of $X_2$.
+Under the mixed design, RNK, INT, and ATS exhibit very similar error rates, which are close to nominal levels.
+In the within-subjects design, the error rates of ATS tend to be slightly above $5\%$.
+Additionally, unlike the other methods whose error rates drop significantly below $5\%$ when the effect of $X_1$ becomes stronger under binomial and ordinal scales, the power of ATS does not seem to be affected in these cases.
 
-::: {#fig-power-main-1}
-<div id="plot-power-1.1">
+::: {#fig-ATS-designs-main}
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-prefix <- "Power"
+prefix <- "Type_I_ATS"
+alpha = 0.05
+
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
+methods = c("PAR", "RNK", "INT", "ATS")
+palette = c("#888888", "#E69F00", "#009E73", "#FF70AB")
+
+df <- read_data(prefix, alpha, effectType = 1, distributions, methods = methods)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+plotly_error_by_design_3(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 7.1, nticks=8, cbPalette = palette)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
+:::
+
+@fig-ATS-designs-main-2 presents results for the main effect of $X_1$.
+The error rates of ATS are now slightly inflated under the mixed design.
+The other methods exhibit the same trends as for the other factor.
+
+::: {#fig-ATS-designs-main-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df2 <- read_data(prefix, alpha, effectType = 2, distributions, methods = methods)
+df2 <- reshape_by_design(df2, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+plotly_error_by_design_3(df2, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX2", max = 7.1, nticks=8, cbPalette = palette)
+```
+Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as a function of the magnitude $a_2$ of the main effect of $X_2$ ($n = 20$)
+:::
+
+**Type I error rates: Interactions**.
+@fig-ATS-designs-interaction-1 presents Type I error rates for the interaction in the presence of a single main effect.
+Results are again very similar for all three nonparametric methods under the mixed design.
+In contrast, the error rates of ATS tend to be lower than nominal levels under the within-subjects design, often falling below $4\%$.
+
+::: {#fig-ATS-designs-interaction-1}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+plotly_error_by_design_3(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 10.1, nticks=8, cbPalette = palette)
+```
+Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$) 
+:::
+
+ When two parallel main effects are present, ATS and RNK lead to very similar trends (see @fig-ATS-designs-interaction-2).
+Overall, INT appears to be a more robust method with the exception of the binomial distribution, for which error rates are higher for this method.
+
+::: {#fig-ATS-designs-interaction-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
 distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
@@
 
-data <- read_data(prefix, alpha = .05, effectType = 3, distributions)
-df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX1", max = 104,  ytitle = 'Power (%)')
+df <- read_data(prefix, alpha, effectType = 0, distributions, methods = methods)
+df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_error_by_design_3(df, xlab = "magnitude of main effects", ytitle = 'Positives (%)', var = "rateX1X2", xvar = "effectX1", max = 104, nticks=8, cbPalette = palette)
 ```
-</div>
-
-<div id="plot-power-1.2">
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- data %>% arrange(design,family,effectX1,rateX1)  %>% group_by(design,family,effectX1) %>% mutate(rank = rank(rateX1))
-df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power - ranking')
-```
-</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-1.2").classList.add("hidden"); }); </script>
-
-<button id="btn1" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
-    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
-  onclick="const plot = document.getElementById('plot-power-1.1');
-  plot.classList.toggle('hidden');
-  document.getElementById('plot-power-1.2').classList.toggle('hidden');
-  btn1.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
-">
-  show ranking
-</button>
-
-Power ($\alpha = .05$) for detecting the **main effect of $X_1$** as a function of its magnitude $a_1$, with $n=20$.
+Positive rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$).
+*Note: These values correspond to Type I error rates under the null hypothesis $a_{12} = 0$; alternative definitions of the null hypothesis may lead to different results.*
 :::
 
-@fig-power-main-2 presents our results for $X_2$.
-Because this factor has a different number of levels across designs, we assess different ranges of power for each design.
-Nevertheless, the same overall patterns emerge.
+**Power: Main effects**.
+As shown in @fig-ATS-power-main-1 and @fig-ATS-power-main-2, ATS appears as the most powerful method for detecting effects of $X_1$ for the mixed design.
+However, in all other situations, it has less power than INT and offer less power than RNK.
 
-::: {#fig-power-main-2}
-<div id="plot-power-2.1">
+::: {#fig-ATS-power-main-1}
 ```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-data <- read_data(prefix, alpha=.05, effectType = 4, distributions) # CHANGE n TO 20
+alpha <- .05
 
-df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX2", max = 104,  ytitle = 'Power (%)')
-```
-</div>
+prefix <- "Power_ATS"
+distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
+dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
 
-<div id="plot-power-2.2">
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- data %>% arrange(design,family,effectX2,rateX2)  %>% group_by(design,family,effectX2) %>% mutate(rank = rank(rateX2))
+df <- read_data(prefix, alpha, effectType = 3, distributions, methods = methods)
+
+df <- df %>% arrange(design,family,effectX1,rateX1)  %>% group_by(design,family,effectX1) %>% mutate(rank = rank(rateX1))
 df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
 
-#df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power - ranking')
+#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design_3(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
 ```
-</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-2.2").classList.add("hidden"); }); </script>
-
-<button id="btn2" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
-    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
-  onclick="const plot = document.getElementById('plot-power-2.1');
-  plot.classList.toggle('hidden');
-  document.getElementById('plot-power-2.2').classList.toggle('hidden');
-  btn2.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
-">
-  show ranking
-</button>
-
-Power ($\alpha = .05$) for detecting the **main effect of $X_2$** as a function of its magnitude $a_2$, with $n=20$. 
+Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_1$** as a function of the magnitude of effect $a_1$ ($n=20$). Hover over the charts to inspect their power ($\%$).
 :::
 
-**Interaction effects**.
-We also vary the magnitude of the interaction effect $X_1 \times X_2$ by manipulating the parameter $a_{12}$ over the range $0.5$ to $2.0$, and then measure the power to detect this interaction.
-The results, summarized in @fig-power-interaction, are consistent with the patterns observed for main effects.
+::: {#fig-ATS-power-main-2}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- read_data(prefix, alpha, effectType = 4, distributions, methods = methods)
 
-::: {#fig-power-interaction}
-<div id="plot-power-3.1">
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-data <- read_data(prefix, alpha=.05, effectType = 5, distributions)
-
-df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotly_error_by_design_2(df, xlab = "magnitude of interaction effect", var = "rateX1X2", xvar = "effectX1X2", max = 104,  ytitle = 'Power (%)')
-```
-</div>
-
-<div id="plot-power-3.2">
-```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
-df <- data %>% arrange(design,family,effectX1X2,rateX1X2)  %>% group_by(design,family,effectX1X2) %>% mutate(rank = rank(rateX1X2))
+df <- df %>% arrange(design,family,effectX2,rateX2)  %>% group_by(design,family,effectX2) %>% mutate(rank = rank(rateX2))
 df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
 
-#df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
-plotly_power_by_design(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power - ranking')
+#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design_3(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
 ```
-</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-3.2").classList.add("hidden"); }); </script>
-
-<button id="btn3" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
-    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
-  onclick="const plot = document.getElementById('plot-power-3.1');
-  plot.classList.toggle('hidden');
-  document.getElementById('plot-power-3.2').classList.toggle('hidden');
-  btn3.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
-">
-  show ranking
-</button>
-
-Power ($\alpha = .05$) for detecting the **interaction effect $X_1 \times X_2$** as a function of its magnitude $a_{12}$, with $n=20$.
+Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_2$** as a function of the magnitude of effect $a_2$ ($n=20$). Hover over the charts to inspect their power ($\%$).
 :::
 
-Overall, there is no clear winner, but we observe that PAR shows poor power performance for log-normal data, while INT consistently outperforms RNK.
-Furthermore, ART provides no substantial power advantage, despite its inflated Type I error rates.
+**Power: Interactions**.
+@fig-ATS-power-interaction shows results on power for interactions.
+INT emerges again as the most powerful method.
+The power of ATS is particularly low under the within-subjects design.
+
+::: {#fig-ATS-power-interaction}
+```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
+df <- read_data(prefix, alpha, effectType = 5, distributions, methods = methods)
+
+df <- df %>% arrange(design,family,effectX1X2,rateX1X2)  %>% group_by(design,family,effectX1X2) %>% mutate(rank = rank(rateX1X2))
+df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+
+#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
+plotly_power_by_design_3(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
+```
+Ranking of methods for their power ($\alpha = .05$) to detect the **interaction effect $X_1 \times X_2$** as a function of the magnitude of effect $a_{12}$ ($n=20$).
+Hover over the charts to inspect their power ($\%$).
+:::
+
+### Conclusion
+Although ATS appears to be a valid alternative, it does not offer clear performance advantages over INT, which is also simpler and more versatile.

### 2025-12-29 text/30-assumptions.qmd
still updating the appendix

@@
 
-However, a question remains: Is ART’s failure in handling such models due solely to heteroscedasticity issues?
-For example, consider the two time distributions in @fig-homoscedastic-model, which --- despite having different shapes --- share the same standard deviation.
-Would ART perform correctly in this case?
-
-::: {#fig-homoscedastic-model}
-```{r, echo=FALSE, message=FALSE, fig.height=2.5, fig.width = 9, warning=FALSE}
-library(plotly)
-
-palette <- c("#E69F00", "#009E73")
-
-sdlog <- function(mean, sd) {
-	sqrt(log(1 + sd^2/mean^2))	
-}
-
-meanlog <- function(mean, sd) {
-	log(mean^2/sqrt(sd^2 + mean^2))
-}
-
-xs <- seq(0, 10, length.out = 1000)
-
-sd <- 1 # This is the common standard deviation
-m1 <- 0.8
-m2 <- 1.4
-
-logm1 <- meanlog(m1, sd)
-logm2 <- meanlog(m2, sd)
-
-logsd1 <- sdlog(m1, sd)
-logsd2 <- sdlog(m2, sd)
-
-y1s <- dlnorm(xs, logm1, logsd1)
-y2s <- dlnorm(xs, logm2, logsd2)
-
-x1 <- 0.36
-x2 <- 1.30
-
-annot1 <- paste("Log-normal: mean = ", m1, ", sd = ", sd)
-annot2 <- paste("Log-normal: mean = ", m2, ", sd = ", sd)
-
-fig <- plot_ly() %>%  
-      add_lines(x = xs, y = y1s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot1, hoverinfo = 'text', line=list(color=palette[1])) %>%
-      add_lines(x = xs, y = y2s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot2, hoverinfo = 'text', line=list(color=palette[2])) %>%      
-      add_annotations(x = x1, y = max(y1s), text = "easy", xref = "x", yref = "y", showarrow = F, font = list(color = palette[1]), xanchor = 'left') %>% 
-      add_annotations(x = x2, y = max(y2s), text = "hard", xref = "x", yref = "y", showarrow = F, font = list(color = palette[2]), xanchor = 'left') %>%   
-      layout(
-        showlegend = FALSE,
-        xaxis = list(title = "Time (min)", range = c(0, 4.5), showgrid = F, showticks = T, ticks="outside", zeroline = F), 
-        yaxis = list(showgrid = F, showticklabels = F, showline = F, range=c(0, max(y1s) + .05), fixedrange=T)
-      ) %>% 
-	      config(displayModeBar = TRUE, scrollZoom = FALSE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale", "hoverclosest", 'hoverCompare'))  %>% layout(dragmode='pan')
-
-fig
-
-```
-Log-normal distributions with equal variances but different means. Notice that their shapes are distinct. 
-:::
-
-While we do not directly address this question in the main paper, we include additional results in [Appendix I](appendix_I.html#log-normal-distributions-with-equal-variances) showing that as long as distribution shapes differ across levels, ART tends to confound effects.
-Naturally, as the shapes become more similar, the problem diminishes.
-In conclusion, heteroscedasticity may exacerbate the problem, but it is not its sole cause.
+We will show that ART confounds effects under unequal variances regardless of whether the underlying distribution is skewed. Thus, the problem also arises under normal distributions.
 

### 2025-12-29 supplementary/_experimental-code-old/Appendix-homoscedastic-lnorm-4x3.R

@@ -17,7 +17,6 @@ getLmerPValues <- function(model, vars) {
 	return(anova(model)[vars, 6])
 }
 
-
 getARTPValues <- function(model, vars) {
 	return(anova(model)[vars, 5]) # For within-subjects
 }

### 2025-12-29 supplementary/experiments-code/_effects_plot.R

@@ -1,204 +1,0 @@
-# Author: Theophanis Tsandilas, Inria & Université Paris-Saclay
-
-# A great library for simulating data for a large range of desings: https://debruine.github.io/faux
-library(faux)
-
-library(dplyr)
-library(tidyr)
-library(egg)
-library(latex2exp)
-
-source("data-generators.R")
-source("simulation-utils.R")
-
-############################################################################################
-# Data generation for various experimental designs
-
-# Encodes the 1:nlevels to numerical values
-# It ensures a fixed distance between the two extreme levels (equal to 1) and random distances in between
-# It also ensures that the sum is zero
-categ2ratio <- function(nlevels) { 
-	levels <- sort(c(0,1, runif(nlevels-2)))
-
-	return(levels - mean(levels))
-}
-
-# Transforms categorical to numverical
-toNumber <- function(factorvalues) {
-	as.numeric(as.character(factorvalues))
-}
-
-# This function produces difference variance ratios for each variable level
-# where the most extreme will be 1 to maxratio
-sd.ratios <- function(nlevels = 4, maxratio = 1.5) {
-	ratios <- c(1, maxratio, 1 + runif(nlevels - 2)*(maxratio - 1))
-	sample(ratios, replace=FALSE) - mean(ratios) + 1
-}
-
-
-getThresholds = function(values, nlevels = 5, equidistant = TRUE) {
-	lim <- 2*sd(values) # I take 2 standard deviations that should contain the largest portion of the data
-
-	thresholds <- NA
-	if(equidistant) {
-		thresholds <- seq(from=-lim, to=lim, by=2*lim/nlevels)[c(-1, -(nlevels + 1))]
-	} else {
-		thresholds <- sort(runif(nlevels - 1))*2*lim-lim
-	}
-
-	thresholds
-}
-
-
-# Distribution parameters used in simulation
-params_normal <- list(sigma_s = 0, sigma_e = 1)
-params_lognormal <- list(sigma_s = 0, sigma_e = 1, mean_target = 1)
-params_cauchy <- list(sigma_s = 0, gamma = 1)
[…]

### 2025-12-29 supplementary/experiments-code/_simulation-art-fast.R

@@ -1,35 +1,0 @@
-# Fast implementation of ART - LMER using lmertest and Satterthwaite approximation
-# This is of interest for simulation with large number of subjects, where AOV becomes expensive
-fast_art_p <- function(m) {
-  # Recover ORIGINAL formula (with random-effects preserved)
-  original_formula <- eval(m$formula)
-
-  # aligned columns
-  aligned <- m$aligned.ranks
-  term_names <- colnames(aligned)
-
-  # Build full data frame with predictors INCLUDING 'subject'
-  predictors <- get_all_vars(original_formula, data = m$data)
-
-  df_base <- cbind(aligned, predictors)
-
-  # Extract RHS from original formula
-  rhs <- original_formula[[3]]
-
-  results <- numeric(length(term_names))
-  names(results) <- term_names
-
-  for (term in term_names) {
-    df <- df_base
-    df$.y <- df[[term]]
-    df <- df[,-c(1:length(term_names))] # And remove the columns that could cause confusion 
-
-    f <- reformulate(deparse(rhs), response = ".y")
-
-    model <- lmerTest::lmer(f, data = df)
-    tab <- anova(model, type = 3)
-    results[term] <- tab[term, "Pr(>F)"]
-  }
-
-  as.data.frame(as.list(results))
-}

### 2025-12-29 supplementary/experiments-code/_simulation-benchmarks-binom.R

@@ -1,38 +1,0 @@
-rm(list=ls())
-#library(lmerTest)
-#library(ARTool)
-
-library(microbenchmark)
-
-source("simulation-utils.R")
-
-nlevels = c(4, 3)
-nsubjects = 10000
-reps = 1
-a1 = 2
-a2 = 0
-a12 = 0
-
-
-bench <- microbenchmark(
-  poisson = simulate_response(
-    nlevels, nsubjects, reps,
-    a1, a2, a12,
-    sigma_s = 0.3,
-    family = "poisson",
-    mean_target = 3
-  ),
-
-  binom = simulate_response(
-    nlevels, nsubjects, reps,
-    a1, a2, a12,
-    sigma_s = 0.3,
-    family = "binomial",
-    p_target = 0.1,
-    size = 10
-  ),
-#  accurate = mu_binomial_accurate(compute_effects(make_design(m1=2,m2=2),a1=4,a2=0,a12=0), sigma_s = 0.3, p_target = 0.1),
-#  gh = mu_binomial(compute_effects(make_design(m1=2,m2=2),a1=4,a2=0,a12=0), sigma_s = 0.3, p_target = 0.1, size=10, n_nodes = 32),
-  times = 100
-)
-

### 2025-12-29 supplementary/experiments-code/_simulation-benchmarks.R

@@ -1,61 +1,0 @@
-rm(list=ls())
-library(lmerTest)
-library(ARTool)
-
-library(microbenchmark)
-
-library(tidyr)
-
-source("utils-data-generation.R")
-source("utils-analysis.R")
-
-################################
-# Distribution parameters used in simulation
-params_norm   <- list(sigma_s = 0.3, sigma_e = 1)
-params_lnorm  <- list(sigma_s = 0.3, sigma_e = 1, mean_target = 1)
-params_cauchy <- list(sigma_s = 0.3, gamma = 1)
-params_exp    <- list(sigma_s = 0.3, mean_target = 0.5)
-params_poisson <- list(sigma_s = 0.3, mean_target = 3)
-params_binom  <- list(sigma_s = 0.3, size = 10, p_target = 0.1)
-
-use_parameters <- function(family){
-  params <- switch(family, 
-    norm    = params_norm, 
-    lnorm   = params_lnorm, 
-    cauchy  = params_cauchy, 
-    exp     = params_exp, 
-    poisson = params_poisson, 
-    binom   = params_binom
-  )
-
-  # Choose a random standard deviation for the random subject effect between 0.1 and 0.5
-  params$sigma_s <- runif(1, min = 0.1, max = 0.5)
-
-  params
-}
-
-# B. 4 x 3 repeated measures
-design = c(4,3)
-within = c(1,1) # Both factors are within-subjects
-formula1 = Y ~ X1*X2 + Error(factor(subject)) 
-formula2 = Y ~ X1*X2 + (1|subject) 
-vars = c("X1", "X2", "X1:X2") 
-n = 256
-
-data = simulate_response(
-  nlevels = c(4,3),
-  within = c(1,1), 
-  n = n,
-  coeffs = c("X1"=0, "X2"=0, "X1:X2"=0),
-  family = "lnorm",
-  use_parameters("lnorm")
-)
-
-bench <- microbenchmark(
-  aov = compare_p_values(data, formula1, vars),
-  lmer = compare_p_values(data, formula2, vars),
-  times = 1
-)
-
-
-

### 2025-12-29 supplementary/experiments-code/_simulation-dispersion.R

@@ -1,44 +1,0 @@
-set.seed(123)
-
-# OD values
-OD_values <- c(1.1, 1.2, 1.5, 2, 5)
-
-# Preallocate
-res <- data.frame(OD = OD_values,
-                  sigma_normal = NA,
-                  sigma_lognormal = NA,
-                  sigma_poisson = NA,
-                  sigma_exponential = NA,
-                  sigma_binomial = NA,
-                  sigma_cauchy = NA)
-
-lambda <- 3
-sigma_e <- 1
-p_target <- 0.1
-n_bin <- 20
-gamma <- 1
-
-for(i in seq_along(OD_values)) {
-  OD <- OD_values[i]
-  # Normal
-  res$sigma_normal[i] <- sqrt((OD-1)*sigma_e^2)
-  # Log-normal
-  res$sigma_lognormal[i] <- sqrt(log(OD + 1))
-  # Poisson
-  res$sigma_poisson[i] <- if(OD <=1) 0 else sqrt(log(1 + (OD-1)/lambda))
-  # Exponential
-  res$sigma_exponential[i] <- if(OD <=1) 0 else sqrt(log(OD))
-  # Binomial (numerical)
-  f_bin <- function(s) {
-    p <- plogis(qlogis(p_target) + rnorm(1e5,0,s))
-    var(p)/(p_target*(1-p_target)) - OD
-  }
-  # check if achievable
-  tryCatch({
-    res$sigma_binomial[i] <- uniroot(f_bin, c(0,5))$root
-  }, error=function(e) { res$sigma_binomial[i] <- NA })
-  # Cauchy (rough)
-  res$sigma_cauchy[i] <- sqrt(OD - 1)
-}
-
-res

### 2025-12-29 supplementary/experiments-code/_simulation-test-repeated-4x3-progress.R

@@ -1,188 +1,0 @@
-rm(list=ls())
-library(lmerTest)
-library(ARTool)
-
-library(tidyr)
-library(dplyr)
-library(broom)
-
-# Parallel computation
-library(foreach)
-library(doParallel)
-library(progressr)
-
-source("simulation-utils.R")
-
-# https://yingji15.github.io/2019-08-17-inverse-normal-transformation/
-# https://www.biostars.org/p/80597/
-# use rankit version where c = 0.5, based on https://www.researchgate.net/publication/286889581_Impact_of_Rank-Based_Normalizing_Transformations_on_the_Accuracy_of_Test_Scores
-INT = function(x){
-  qnorm((rank(x) - 0.5)/length(x))
-}
-
-# Extract the p-values from the result objects returned by summary() or anova()
-get_p_values <- function(model, vars, is_lmer = FALSE) {
-  if(is_lmer) return(anova(model)[vars, "Pr(>F)"])
-  else if(is.null(model$formula)){
-    return((tidy(summary(model)[[2]][[1]]) %>% filter(term %in% vars) %>% select(p.value))[[1]])
-  } else {
-    anov <- anova(model)
-    return(anov[anov$Term %in% vars, "Pr(>F)"])
-  }
-}
-
-compare_p_values_aov <- function(df){ 
-  model_par <- suppressMessages(aov(Y ~ X1*X2 + Error(subject), data=df)) # Parametric
-  model_art <- suppressMessages(art(Y ~ X1*X2 + Error(factor(subject)), data=df)) # ARTool
-  model_rnk <- suppressMessages(aov(rank(Y) ~ X1*X2 + Error(subject), data=df)) # RNK
-  model_int <- suppressMessages(aov(INT(Y) ~ X1*X2 + Error(subject), data=df)) # INT
-
-  vars <- c("X1", "X2", "X1:X2")
-  c(get_p_values(model_par, vars), 
-    get_p_values(model_rnk, vars), 
-    get_p_values(model_art, vars), 
-    get_p_values(model_int, vars)
-  )
-}
-
-
-################################
-# Distribution parameters used in simulation
-params_normal <- list(sigma_s = 0.4, sigma_e = 1)
-#params_ordinal <- list(sigma_s = 0.4, sigma_e = 1, levels = 5, equidistant = FALSE)
-params_lognormal <- list(sigma_s = 0.4, sigma_e = 1, mean_target = 1)
[…]

### 2025-12-29 supplementary/experiments-code/_testing-lnorm.R

@@ -1,42 +1,0 @@
-rm(list=ls())
-
-source("utils-data-generation.R")
-source("utils-analysis.R")
-
-# Distribution parameters used in simulation
-#params_norm   <- list(sigma_s = 0.4, sigma_e = 1)
-params_lnorm_1  <- list(sigma_s = 0.5, sigma_e = 0.2, mean_target = 1)
-params_lnorm_2  <- list(sigma_s = 0.5, sigma_e = 1, mean_target = 1)
-
-nlevels = c(2)
-within = c(1)
-n = 100000
-coeffs <- c("X1"=2, "X2"=0, "X1:X2"=0) # a1, a2, a12
-
-df1 <- simulate_response(
-  nlevels, within, n, coeffs,
-  family = "lnorm",
-  params = params_lnorm_1,
-  normalize = TRUE
-)
-
-df2 <- simulate_response(
-  nlevels, within, n, coeffs,
-  family = "lnorm",
-  params = params_lnorm_2,
-  normalize = TRUE
-)
-
-printParams <- function(df, sigma){
-  cat("####### sigma = ", sigma, " ########\n")
-  y1 = log(df[df$X1=="A1", ]$Y)
-  y2 = log(df[df$X1=="A2", ]$Y)
-  diff = abs(mean(y1) - mean(y2))
-
-  cat("means = ", mean(y1), ", ", mean(y2), " diff = ", diff, "\n")
-  cat("sds = ", mean(sd(y1), sd(y2)), "\n")
-  cat("effect size = ", diff/mean(sd(y1), sd(y2)), "\n")
-}
-
-printParams(df1, params_lnorm_1$sigma_e)
-printParams(df2, params_lnorm_2$sigma_e)

### 2025-12-29 supplementary/experiments-code/appendix-experiment-ATS-power-multieffect.R

@@ -1,10 +1,8 @@
 # Author: Theophanis Tsandilas, Dec 2025
 # Inria & Université Paris-Saclay
 
-# Comparison of the power of RNK and INT methods against the multifactorial generalizations of the van der Waerden test 
-# and the Kruskal-Wallis and Friedman tests: http://www.uni-koeln.de/~luepsen/R/
-# 4x3 within-subjects, 2x3 within-subjects, and 2x4 mixed-subjects
-# In this condition, we assess how an effect on X2 affects the power for detetign main effects on X1 and interactions X1 x X2
+# Experiment evaluating the Power of the ATS method
+# 4x3 within-subjects and 2x4 mixed-subjects
 # We fix n=20
 
 rm(list=ls())
@@ -14,12 +12,13 @@ library(foreach)
 library(doParallel)
 library(tidyr)
 
-source("np.anova.R") # From http://www.uni-koeln.de/~luepsen/R/
+# Try the ANOVA-type statistic (ATS)
+# https://www.quantargo.com/help/r/latest/packages/nparLD/2.1/nparLD
+library(nparLD)
 
 source("utils-data-generation.R")
 source("utils-analysis.R")
 source("utils-extra.R")
-library("POSSA") # to get access to the get_p function
 
 ################################
 # Distribution parameters used in simulation
@@ -47,16 +46,11 @@ use_parameters <- function(family){
   params
 }
 
-# (1) 4 x 3 within-subjects, (2) 2x3 within-subjects, and (3) 2 x 4 mixed design (one between-subjects and one repeated-measures factor)
-designs <- list(c(4,3), c(2,3), c(2,4))
-within <- list(c(1,1), c(0,0), c(0,1))
+# (1) 4 x 3 within-subjects, (2) 2 x 4 mixed design (one between-subjects and one repeated-measures factor)
+designs <- list(c(4,3), c(2,4))
+within <- list(c(1,1), c(0,1))
 
-# Notice that I include slope random effects in my models, which is required by the np.anova functions  
-formula1 = Y ~ X1*X2 + Error(subject/(X1*X2))
-formula2 = Y ~ X1*X2 
-formula3 = Y ~ X1*X2 + Error(subject/X2) 
-formulas <- c(formula1, formula2, formula3)
-
+formula = Y ~ X1*X2 + Error(factor(subject)) 
 vars = c("X1", "X2", "X1:X2") 
 
 # Continuous distributions (equal variance, full) and discrete distribution: binom (size = 10, prob = 0.1) and Poisson
[…]

### 2025-12-29 supplementary/experiments-code/appendix-experiment-lognormal-homoskedastic.R

@@ -1,171 +1,0 @@
-# Author: Theophanis Tsandilas, Dec 2025
-# Inria & Université Paris-Saclay
-
-# Experiment evaluating the Type I error rates of PAR, ART, RNK, and INT on special lognormal distributions with homoskedastic responses
-
-rm(list=ls())
-
-# Parallel computation
-library(foreach)
-library(doParallel)
-library(tidyr)
-
-source("utils-data-generation.R")
-source("utils-analysis.R")
-
-##################################
-# These are some extra functions 
-
-# See https://en.wikipedia.org/wiki/Log-normal_distribution
-# We control the standard deviation of the generated data so that it is equal
-sdlog <- function(mean, sd) {
-  sqrt(log(1 + sd^2/mean^2))  
-}
-
-meanlog <- function(mean, sd) {
-  log(mean^2/sqrt(sd^2 + mean^2))
-}
-
-
-simulate_response_equalvar <- function(
-  nlevels = c(4,3), within = c(1, 1), n = 1,
-  coeffs = c("X1"=0, "X2"=0, "X1:X2"=0),
-  family = c("lnorm"),
-  params
-) {
-  family <- match.arg(family)
-  sigma_s <- runif(1, min = params$sigma_s[1], max = params$sigma_s[length(params$sigma_s)])
-
-  # -------------------------------------------------
-  # Create design with correct within/between logic
-  # -------------------------------------------------
-  design <- make_design(nlevels, within, n)
-
-  # -------------------------------------------------
-  # Build effects based on number of factors
-  # -------------------------------------------------
-  k <- length(nlevels)
-  # Currently supporting up to three factors
-  effects <- switch(
-    k,
-    with(design, coeffs["X1"]*x1),
-    with(design, coeffs["X1"]*x1 + coeffs["X2"]*x2 + coeffs["X1:X2"]*x1*x2),
-    with(design, coeffs["X1"]*x1 + coeffs["X2"]*x2 + coeffs["X3"]*x3 + coeffs["X1:X2"]*x1*x2)
-  )
-
-  mu <- mu_lognormal(effects, sigma_s, params$sigma_e, params$mean_target)
-
-  # -------------------------------------------------
-  # Random subject effects
-  # -------------------------------------------------
-  subjects <- unique(design$subject)
-  s <- rnorm(length(subjects), 0, sigma_s)
[…]

### 2025-12-29 supplementary/experiments-code/appendix-experiment-single-factor-power.R

@@ -1,9 +1,8 @@
 # Author: Theophanis Tsandilas, Dec 2025
 # Inria & Université Paris-Saclay
 
-# Experiment evaluating the Type I error rate of the ATS method
-# 4x3 within-subjects and 2x4 mixed-subjects
-# We fix n=20
+# Comparing single-factor non-parametric methods 
+# Testing single-factor designs with 2,3 or 4 levels (both within-subjects and between-subjects) 
 
 rm(list=ls())
 
@@ -11,10 +10,6 @@ rm(list=ls())
 library(foreach)
 library(doParallel)
 library(tidyr)
-
-# Try the ANOVA-type statistic (ATS)
-# https://www.quantargo.com/help/r/latest/packages/nparLD/2.1/nparLD
-library(nparLD)
 
 source("utils-data-generation.R")
 source("utils-analysis.R")
@@ -46,48 +41,29 @@ use_parameters <- function(family){
   params
 }
 
-# (1) 4 x 3 within-subjects, (2) 2 x 4 mixed design (one between-subjects and one repeated-measures factor)
-designs <- list(c(4,3), c(2,4))
-within <- list(c(1,1), c(0,1))
+# Testing single-factor designs with 2,3 or 4 levels (both within-subjects and between-subjects) 
+designs <- list(2, 3, 4, 2, 3, 4)
+within <- list(1, 1, 1, 0, 0, 0)
 
-formula = Y ~ X1*X2 + Error(factor(subject)) 
-vars = c("X1", "X2", "X1:X2") 
+formula = Y ~ X1 + Error(factor(subject)) 
+vars = c("X1") 
 
-# Continuous distributions (equal variance, full) and discrete distribution: binom (size = 10, prob = 0.1) and Poisson
 distributions <- c("norm", "lnorm", "exp", "binom", "poisson", "likert")
 
-# Various combinations of effects
-effects <- matrix(c(0, 0, 0,
-            0.5, 0.5, 0,
-            1, 1, 0,
-            2, 2, 0,
-            4, 4, 0,
-            8, 8, 0,
-            0.5, 0, 0,
-            1, 0, 0,
-            2, 0, 0,
-            4, 0, 0,
-            8, 0, 0,
-            0, 0.5, 0,
-            0, 1, 0,
-            0, 2, 0,
-            0, 4, 0,
-            0, 8, 0), 
-           ncol = 3, byrow = TRUE)
-
[…]

### 2025-12-29 supplementary/experiments-code/utils-extra.R

@@ -180,7 +180,12 @@ repeat_test_custom <- function(
   res.05 <- round(colMeans(results<.05, na.rm = TRUE), digits = 4) # alpha = .05
   res.01 <- round(colMeans(results<.01, na.rm = TRUE), digits = 4) # alpha = .01
 
-  designStr <- paste(nlevels, collapse="x")
+  if(length(nlevels) == 1) {
+    designStr <- paste(ifelse(within, "within", "between"), nlevels, sep="-")
+  }
+  else {
+    designStr <- paste(nlevels, collapse="x")
+  }
 
   ndummies <- length(vars) - length(res.05)/4
   dummies <- rep(0, ndummies) # Complete with zeros irrelevant rate results. Useful when producing results for multiple designs with different numbers of factors

### 2025-12-29 utils-plotlying.R

@@ -329,3 +329,50 @@ plotly_power_by_design <- function(df, xlab = "magnitude of main effects", var =
 
 	fig
 }
+
+# This is for alterative graphs for a different set of designs : 2x3, 2x4, 4x3 specific to Power charts (writing more generic code with plotly is painful)
+plotly_power_by_design_3 <- function(df, xlab = "magnitude of main effects", var = "rank", hovervar="rateX1", xvar = "effectX1", max = 100, ytitle = 'Type I errors (%)', cbPalette = c("#888888", "#E69F00", "#009E73", "#FF5E00")){
+	# aesthetics
+	symbols <- c("asterisk", "x", "star-diamond", "star-triangle-up")
+	margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
+
+	dnames <- levels(df$family)
+	# To be used as labels on the subplots
+	annots <- lapply(1:length(dnames), function(index){list(x = (index-0.5)/length(dnames), y = 1, text = dnames[index], xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F)})
+
+	menuItems <- c("4x3 within", "2x4 mixed")
+	yvarnames <- c(paste(var,"4x3",sep="."), paste(var,"2x4",sep="."))
+	hvarnames <- c(paste(hovervar,"4x3",sep="."), paste(hovervar,"2x4",sep="."))
+
+	createPlot <- function(data, dnames, symbols) {
+	  p <- plot_ly(data,  x = ~factor(eval(as.symbol(xvar))), y = ~eval(as.symbol(yvarnames[1])), text = ~round(100*eval(as.symbol(hvarnames[1])), digits = 2), hoverinfo ="text", color = ~method, symbol = ~method, symbols = symbols, colors = cbPalette, type = 'scatter', mode = 'lines+markers', marker = list(line = list(width = 2)), legendgroup = ~method, showlegend = unique(data$family == dnames[1]), visible = T) 
[…]
